---
title: "Lesson 14: P-Values; Power and Sample Size Determination"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w7d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 14 - P-Values; Power and Sample Size Determination
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 14: P-Values; Power and Sample Size Determination {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 13: Tests About a Population Mean and Proportion**

-   One-sample z test ($\sigma$ known) and t test ($\sigma$ unknown)
-   Score test and exact binomial test for proportions
-   Seven-step hypothesis testing procedure
-   CI-test duality: a $100(1-\alpha)\%$ CI contains values not rejected at level $\alpha$

. . .

**Today's Goals:**

1. Understand P-values as a continuous measure of evidence against $H_0$
2. Compute P-values for z and t tests
3. Understand the distribution of P-values under $H_0$ and $H_a$
4. Use the noncentral t distribution for exact power calculations
5. Determine sample sizes for planning clinical trials using R

## Roadmap for Today

| Topic | Key Idea | Textbook |
|-------|----------|----------|
| P-values | Probability of data as extreme or more extreme than observed | Devore 9.4 |
| P-value distribution | Uniform under $H_0$; concentrated near 0 under $H_a$ | Devore 9.4 |
| Power for the t test | Noncentral t distribution | Devore 9.2 |
| Sample size determination | `power.t.test()` in R | Devore 9.2 |

. . .

::: callout-note
## Why This Matters
In medical research, P-values appear in virtually every published paper. Understanding what they mean --- and what they do **not** mean --- is essential for interpreting clinical evidence.
:::

# P-Values (Devore 9.4) {background-color="#3498db"}

## From Rejection Regions to P-Values

**Recall the rejection region approach:**

1. Fix $\alpha$ (e.g., 0.05)
2. Find critical value(s)
3. Reject $H_0$ if the test statistic falls in the rejection region

. . .

**Limitation:** The rejection region approach gives a binary decision (reject / fail to reject), but tells us nothing about **how strongly** the data contradict $H_0$.

. . .

**Two scenarios with the same decision:**

- $\alpha = 0.05$, P-value = 0.049 $\Rightarrow$ barely reject $H_0$
- $\alpha = 0.05$, P-value = 0.00001 $\Rightarrow$ overwhelming evidence against $H_0$

The rejection region approach treats these identically!

## Definition of the P-Value

::: callout-important
## Definition (Devore 9.4)
The **P-value** is the probability, calculated assuming that the null hypothesis is true, of obtaining a value of the test statistic at least as contradictory to $H_0$ as the value calculated from the available sample.
:::

. . .

**Key points:**

- The P-value **is** a probability
- It is calculated **assuming $H_0$ is true**
- It is a function of the **sample data** (and therefore a random variable!)
- We need to determine which values are "at least as contradictory" to $H_0$

## Formal Definition of the P-Value

Let $T = T(X_1, \ldots, X_n)$ be a test statistic and let $t_{obs}$ be its observed value. For a test of $H_0: \theta = \theta_0$ vs $H_a$:

::: callout-important
## Definition (Formal)
The **P-value** is defined as:

$$\text{P-value} = \begin{cases} P_{\theta_0}(T \geq t_{obs}) & \text{upper-tailed test } (H_a: \theta > \theta_0) \\[6pt] P_{\theta_0}(T \leq t_{obs}) & \text{lower-tailed test } (H_a: \theta < \theta_0) \\[6pt] P_{\theta_0}(|T| \geq |t_{obs}|) & \text{two-tailed test } (H_a: \theta \neq \theta_0) \end{cases}$$

where $P_{\theta_0}(\cdot)$ denotes probability calculated under $H_0$.
:::

. . .

**Interpretation:** The P-value measures how "extreme" or "surprising" the observed data would be if $H_0$ were true. Smaller P-values indicate data that is less compatible with $H_0$.

## The P-Value as a Tail Probability

Consider an upper-tailed test with test statistic $T$ having CDF $F_T(t)$ under $H_0$.

. . .

**The P-value is the "survival function" evaluated at $t_{obs}$:**

$$\text{P-value} = P_{H_0}(T \geq t_{obs}) = 1 - F_T(t_{obs})$$

. . .

**Key insight:** The P-value is the probability of observing a test statistic *at least as extreme* as what we actually observed, computed under the assumption that $H_0$ is true.

. . .

**Why "at least as extreme"?** 

If $t_{obs}$ is unlikely under $H_0$, then values *even more extreme* than $t_{obs}$ must also be unlikely. The P-value captures the total probability of all outcomes at least as unfavorable to $H_0$.

## The Probability Integral Transformation

A fundamental result from probability theory underlies the distribution of P-values.

::: callout-important
## Theorem (Probability Integral Transformation)
Let $X$ be a continuous random variable with CDF $F_X$. Then the random variable 

$$U = F_X(X)$$

has a $\text{Uniform}(0, 1)$ distribution.
:::

. . .

**Proof:** For $u \in (0, 1)$:

$$P(U \leq u) = P(F_X(X) \leq u) = P(X \leq F_X^{-1}(u)) = F_X(F_X^{-1}(u)) = u$$

The second equality uses the fact that $F_X$ is monotone increasing (for continuous $X$). Since $P(U \leq u) = u$ for all $u \in (0,1)$, we have $U \sim \text{Uniform}(0,1)$. $\blacksquare$

## Corollary: The Survival Function is Also Uniform

::: callout-note
## Corollary
If $X$ is a continuous random variable with CDF $F_X$, then 

$$V = 1 - F_X(X)$$

also has a $\text{Uniform}(0, 1)$ distribution.
:::

. . .

**Proof:** If $U \sim \text{Uniform}(0,1)$, then $1 - U \sim \text{Uniform}(0,1)$ as well. 

This is easily verified: $P(1 - U \leq v) = P(U \geq 1 - v) = 1 - (1 - v) = v$. $\blacksquare$

. . .

**Why this matters:** For an upper-tailed test, the P-value is $1 - F_T(T)$, which is exactly the survival function evaluated at the test statistic!

## P-Values Are Uniform Under $H_0$

We can now state and prove the central result:

::: callout-important
## Theorem (Distribution of the P-Value Under $H_0$)
Let $T$ be a test statistic with continuous CDF $F_T$ under $H_0$. For an upper-tailed test, define the P-value as 

$$P = 1 - F_T(T)$$

Then under $H_0$, the P-value satisfies $P \sim \text{Uniform}(0, 1)$.
:::

. . .

**Proof:** Under $H_0$, the test statistic $T$ has CDF $F_T$. By the probability integral transformation, $F_T(T) \sim \text{Uniform}(0,1)$. By the corollary, $P = 1 - F_T(T) \sim \text{Uniform}(0,1)$. $\blacksquare$

## Consequence: P-Value Tests Maintain the Correct Size

This uniformity result has a crucial consequence for hypothesis testing.

::: callout-note
## Definition: Size of a Test
The **size** of a hypothesis test is the probability of rejecting $H_0$ when $H_0$ is true---i.e., the Type I error rate:

$$\text{Size} = P_{H_0}(\text{Reject } H_0) = P(\text{Type I Error})$$

A test has **exact size** $\alpha$ if this probability equals $\alpha$ exactly. A test has size **at most** $\alpha$ if this probability is $\leq \alpha$.
:::

. . .

::: callout-important
## Theorem (Size of P-Value Tests)
If the P-value $P$ has a $\text{Uniform}(0,1)$ distribution under $H_0$, then the test that rejects $H_0$ when $P \leq \alpha$ has **exact size** $\alpha$.
:::

. . .

**Proof:** The Type I error rate is

$$P_{H_0}(\text{Reject } H_0) = P_{H_0}(P \leq \alpha) = \alpha$$

where the last equality follows because $P \sim \text{Uniform}(0,1)$ under $H_0$. $\blacksquare$

. . .

**Key insight:** The P-value approach **automatically** guarantees that rejecting when P-value $\leq \alpha$ produces a test with Type I error rate exactly equal to $\alpha$. This is not a coincidence---it's a direct consequence of the probability integral transformation!

## Why This Matters: Equivalence of Approaches

**Rejection region approach:**

- Choose $\alpha$, find critical value $c$ such that $P_{H_0}(T \geq c) = \alpha$
- Reject $H_0$ if $T \geq c$

. . .

**P-value approach:**

- Compute P-value $= P_{H_0}(T \geq t_{obs})$
- Reject $H_0$ if P-value $\leq \alpha$

. . .

**These are equivalent!**

$$T \geq c \iff 1 - F_T(T) \leq 1 - F_T(c) = \alpha \iff \text{P-value} \leq \alpha$$

. . .

The P-value encodes *all possible rejection decisions* for any choice of $\alpha$:

- If P-value $= 0.03$: reject at $\alpha = 0.05$ but not at $\alpha = 0.01$
- If P-value $= 0.008$: reject at both $\alpha = 0.05$ and $\alpha = 0.01$

## P-Values for Discrete Test Statistics

::: callout-warning
## Technical Note
The probability integral transformation requires $T$ to be **continuous**. For discrete test statistics (e.g., the binomial), the P-value may not be exactly Uniform$(0,1)$.
:::

. . .

**Example:** For the exact binomial test with $H_0: p = 0.5$ and $n = 10$, the P-value can only take finitely many values. The CDF of the P-value is a step function, not the identity.

. . .

**Consequence:** For discrete tests, $P(P \leq \alpha) \leq \alpha$, with equality only when $\alpha$ happens to coincide with one of the achievable P-values. The test is **conservative** (actual Type I error $<$ nominal $\alpha$).

. . .

**Resolution:** Randomized tests can achieve exact size $\alpha$, but are rarely used in practice. Instead, we accept the slight conservatism.

# Conditional P-Values and Fisher's Exact Test {background-color="#8e44ad"}

## The Problem of Nuisance Parameters

In many testing problems, the null hypothesis specifies a relationship but leaves some parameters unspecified.

. . .

**Example: Testing independence in a 2×2 table**

|              | Disease+ | Disease− | Total |
|:-------------|:--------:|:--------:|:-----:|
| Exposed      |    a     |    b     | a + b |
| Not Exposed  |    c     |    d     | c + d |
| **Total**    |  a + c   |  b + d   |   n   |

. . .

**Hypotheses:**

- $H_0$: Exposure and disease are independent (no association)
- $H_a$: Exposure and disease are associated

. . .

**The problem:** Under $H_0$, the marginal probabilities $p_1 = P(\text{Exposed})$ and $p_2 = P(\text{Disease+})$ are **nuisance parameters**---unknown quantities not of direct interest.

## Why Nuisance Parameters Complicate P-Values

To compute a P-value, we need to calculate $P_{H_0}(T \geq t_{obs})$.

. . .

But under $H_0: p_{11} = p_1 \cdot p_2$ (independence), the distribution of any test statistic depends on the unknown values of $p_1$ and $p_2$!

. . .

**Options:**

1. **Estimate** the nuisance parameters and use asymptotic approximations (chi-squared test)
2. **Eliminate** the nuisance parameters by conditioning on sufficient statistics

. . .

Fisher's exact test takes approach (2), which yields an **exact** P-value that is valid for any sample size.

## The Conditionality Principle

::: callout-important
## Conditionality Principle (Informal)
If a statistic $S$ is **ancillary** (its distribution doesn't depend on the parameter of interest), then inference should be conducted **conditional** on the observed value of $S$.
:::

. . .

**Intuition:** Ancillary statistics carry no information about the parameter of interest, but they do affect the "precision" of the experiment. Conditioning on them focuses inference on the relevant variation.

. . .

**For nuisance parameters:** If $S$ is a sufficient statistic for the nuisance parameter $\eta$, then conditioning on $S$ eliminates the dependence on $\eta$.

## Fisher's Exact Test: Setup

Consider a $2 \times 2$ table with fixed total $n$. Let:

- $X$ = number in cell (1,1) (Exposed and Disease+)
- Row totals $r_1 = a + b$ and $r_2 = c + d$ (fixed by design or conditioning)
- Column totals $c_1 = a + c$ and $c_2 = b + d$

. . .

Under $H_0$ (independence) with both margins fixed, the cell count $X$ follows a **hypergeometric distribution**:

$$P(X = x \mid r_1, r_2, c_1, c_2) = \frac{\binom{c_1}{x}\binom{c_2}{r_1 - x}}{\binom{n}{r_1}}$$

. . .

**Key observation:** This distribution has **no unknown parameters**! The nuisance parameters have been eliminated.

## Why Condition on the Margins?

::: callout-note
## Theorem (Sufficiency of Margins)
Under $H_0$ (independence), the row and column marginal totals $(r_1, c_1)$ are **jointly sufficient** for the nuisance parameters $(p_1, p_2)$.
:::

. . .

**Proof sketch:** Under independence, the joint probability of a $2 \times 2$ table is:

$$P(a, b, c, d) = \frac{n!}{a!b!c!d!} p_1^{r_1}(1-p_1)^{r_2} p_2^{c_1}(1-p_2)^{c_2}$$

. . .

By the factorization theorem, since the likelihood factors into a term depending on $(p_1, p_2)$ only through $(r_1, c_1)$, the margins are sufficient for the nuisance parameters.

## The Conditional Distribution Eliminates Nuisance Parameters

Conditioning on the sufficient statistic $(r_1, c_1)$:

$$P(X = x \mid r_1, c_1) = \frac{P(X = x, r_1, c_1)}{P(r_1, c_1)}$$

. . .

After simplification (the nuisance parameters cancel!):

$$P(X = x \mid r_1, c_1) = \frac{\binom{c_1}{x}\binom{n - c_1}{r_1 - x}}{\binom{n}{r_1}}$$

. . .

This is the hypergeometric distribution with parameters $n$, $r_1$, and $c_1$---all **known** from the observed data!

. . .

::: callout-tip
## Key Insight
By conditioning on the margins, we transform a problem with nuisance parameters into one where the null distribution is completely specified.
:::

## Fisher's Exact Test: The Conditional P-Value

The **conditional P-value** is computed using the hypergeometric distribution:

$$\text{P-value} = P(X \geq x_{obs} \mid r_1, c_1) = \sum_{x = x_{obs}}^{\min(r_1, c_1)} \frac{\binom{c_1}{x}\binom{n - c_1}{r_1 - x}}{\binom{n}{r_1}}$$

. . .

(For a two-sided test, we sum probabilities for outcomes as extreme or more extreme in either direction.)

. . .

**Properties:**

- **Exact:** Valid for any sample size, including small samples
- **Free of nuisance parameters:** No need to estimate marginal probabilities  
- **Discrete:** P-value can only take finitely many values (test is slightly conservative)

## Example: Fisher's Exact Test in R

**Setting:** A clinical trial investigates whether a new treatment reduces adverse events.

```{r}
#| echo: true
# Observed 2x2 table
#              Adverse Event   No Event
# Treatment         2            18       (20)
# Placebo           8            12       (20)
#                 (10)          (30)      (40)

adverse_table <- matrix(c(2, 8, 18, 12), nrow = 2, byrow = TRUE,
                        dimnames = list(Group = c("Treatment", "Placebo"),
                                       Outcome = c("Adverse", "None")))
adverse_table
```

## Example: Fisher's Exact Test Output

```{r}
#| echo: true
# Fisher's exact test
fisher_result <- fisher.test(adverse_table)
fisher_result
```

. . .

```{r}
#| echo: true
cat("P-value:", round(fisher_result$p.value, 4), "\n")
cat("Odds ratio estimate:", round(fisher_result$estimate, 3), "\n")
cat("95% CI for odds ratio:", round(fisher_result$conf.int[1], 3), "to", 
    round(fisher_result$conf.int[2], 3), "\n")
```

## Visualizing the Hypergeometric Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 12
# Parameters from our example
n <- 40; r1 <- 20; c1 <- 10
x_obs <- 2  # Observed count in treatment/adverse cell

# Possible values of X
x_vals <- 0:min(r1, c1)
probs <- dhyper(x_vals, m = c1, n = n - c1, k = r1)

# Create data frame
hyper_data <- tibble(
  x = x_vals,
  prob = probs,
  extreme = x <= x_obs | x >= (r1 - (c1 - x_obs))  
  # Two-sided: as or more extreme than observed
)

# P-value components
pval_lower <- sum(probs[x_vals <= x_obs])
pval_upper <- sum(probs[probs <= probs[x_obs + 1]])  # Fisher's method

ggplot(hyper_data, aes(x = x, y = prob, fill = extreme)) +
  geom_col(width = 0.7, color = "black") +
  geom_vline(xintercept = x_obs, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(values = c("gray70", "coral"), 
                    labels = c("Not extreme", "Extreme (contributes to P-value)")) +
  annotate("text", x = x_obs + 0.3, y = max(probs) * 0.9, hjust = 0,
           label = paste0("Observed: X = ", x_obs), color = "red", size = 5) +
  labs(x = "X (count in Treatment/Adverse cell)",
       y = "Probability",
       title = "Hypergeometric Distribution Under H0 (Conditional on Margins)",
       subtitle = paste0("n = ", n, ", r1 = ", r1, ", c1 = ", c1, 
                        "; P-value = ", round(fisher_result$p.value, 4)),
       fill = "") +
  theme(legend.position = "top")
```

## Why Conditioning? A Deeper Look

**The philosophical justification:**

1. **Sufficiency Principle:** Inference should depend on the data only through sufficient statistics
2. **Conditionality Principle:** Inference should condition on ancillary statistics

. . .

When combined, these principles suggest conditioning on sufficient statistics for nuisance parameters.

. . .

**The practical justification:**

- Conditioning produces a test with **exact** size $\alpha$
- No large-sample approximations needed
- Valid even for sparse tables or small samples

. . .

::: callout-warning
## Caveat
Conditioning can reduce power compared to unconditional tests. This is the price of exactness. For large samples, the chi-squared test (which estimates nuisance parameters) may be more powerful.
:::

## Connection to Randomization Tests

Fisher's exact test can also be viewed as a **randomization test**:

. . .

**Setup:** Fix the margins (total exposed, total diseased). Under $H_0$, each possible table consistent with these margins is equally likely given the hypergeometric distribution.

. . .

**The P-value** is the proportion of possible tables that show an association at least as extreme as observed.

. . .

**This connects to permutation testing:** Under $H_0$, "Disease" labels are exchangeable among subjects. The conditional P-value equals the proportion of permutations yielding a test statistic as extreme as observed.

```{r}
#| echo: true
# Verify: Fisher's P-value via simulation
set.seed(551)
n_perm <- 10000
disease_labels <- c(rep(1, 10), rep(0, 30))  # 10 diseased, 30 not
treatment <- c(rep(1, 20), rep(0, 20))       # 20 in each group

# Observed statistic
obs_stat <- sum(treatment[disease_labels == 1])  # = 2

# Permutation distribution
perm_stats <- replicate(n_perm, {
  sum(sample(treatment)[disease_labels == 1])
})

cat("Simulated P-value:", mean(perm_stats <= obs_stat), "\n")
cat("Fisher's exact P-value (one-sided):", 
    fisher.test(adverse_table, alternative = "less")$p.value, "\n")
```

## Decision Rule Based on the P-Value

::: callout-important
## P-Value Decision Rule (Devore 9.4)

Select a significance level $\alpha$. Then:

$$\text{Reject } H_0 \text{ if P-value} \leq \alpha; \qquad \text{Do not reject } H_0 \text{ if P-value} > \alpha$$
:::

. . .

**Equivalence to the rejection region method:**

- If the test statistic falls in the rejection region, then P-value $\leq \alpha$
- If the test statistic falls outside the rejection region, then P-value $> \alpha$

. . .

**Advantage:** The P-value encodes more information than a binary reject/fail-to-reject decision. It conveys the **strength** of evidence against $H_0$.

## P-Values for z Tests

For a z test (test statistic $Z$ with $Z \sim N(0,1)$ under $H_0$):

$$\text{P-value} = \begin{cases} 1 - \Phi(z) & \text{for an upper-tailed test } (H_a: \mu > \mu_0) \\[6pt] \Phi(z) & \text{for a lower-tailed test } (H_a: \mu < \mu_0) \\[6pt] 2[1 - \Phi(|z|)] & \text{for a two-tailed test } (H_a: \mu \neq \mu_0) \end{cases}$$

where $\Phi(z)$ is the standard normal CDF and $z$ is the observed test statistic.

. . .

```{r}
#| echo: true
# Quick R reminders for P-value computation
z_obs <- 2.15
cat("Upper-tailed P-value:", 1 - pnorm(z_obs), "\n")
cat("Lower-tailed P-value:", pnorm(z_obs), "\n")
cat("Two-tailed P-value:  ", 2 * (1 - pnorm(abs(z_obs))), "\n")
```

## Visualizing P-Values for z Tests

```{r}
#| code-fold: true
#| fig-height: 7
#| fig-width: 18
z_obs <- 2.15
z_seq <- seq(-4, 4, length.out = 500)
y <- dnorm(z_seq)

p1 <- tibble(z = z_seq, y = y) |>
  ggplot(aes(x = z, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(z = z_seq[z_seq >= z_obs], 
                          y = dnorm(z_seq[z_seq >= z_obs])),
            aes(x = z, y = y), fill = "coral", alpha = 0.6) +
  geom_vline(xintercept = z_obs, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = z_obs + 0.15, y = 0.06, hjust = 0, size = 5, color = "red",
           label = paste0("z = ", z_obs)) +
  annotate("text", x = 3.2, y = 0.12, size = 5, color = "coral",
           label = paste0("P = ", round(1 - pnorm(z_obs), 4))) +
  labs(title = expression("Upper-tailed: " * H[a]*": " * mu > mu[0]),
       x = "z", y = "Density")

p2 <- tibble(z = z_seq, y = y) |>
  ggplot(aes(x = z, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(z = z_seq[z_seq <= -z_obs], 
                          y = dnorm(z_seq[z_seq <= -z_obs])),
            aes(x = z, y = y), fill = "coral", alpha = 0.6) +
  geom_vline(xintercept = -z_obs, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = -z_obs - 0.15, y = 0.06, hjust = 1, size = 5, color = "red",
           label = paste0("z = ", -z_obs)) +
  annotate("text", x = -3.2, y = 0.12, size = 5, color = "coral",
           label = paste0("P = ", round(pnorm(-z_obs), 4))) +
  labs(title = expression("Lower-tailed: " * H[a]*": " * mu < mu[0]),
       x = "z", y = "Density")

p3 <- tibble(z = z_seq, y = y) |>
  ggplot(aes(x = z, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(z = z_seq[z_seq >= z_obs], 
                          y = dnorm(z_seq[z_seq >= z_obs])),
            aes(x = z, y = y), fill = "coral", alpha = 0.6) +
  geom_area(data = tibble(z = z_seq[z_seq <= -z_obs], 
                          y = dnorm(z_seq[z_seq <= -z_obs])),
            aes(x = z, y = y), fill = "coral", alpha = 0.6) +
  geom_vline(xintercept = c(-z_obs, z_obs), linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 3.2, y = 0.12, size = 5, color = "coral",
           label = paste0("P/2 = ", round(1 - pnorm(z_obs), 4))) +
  annotate("text", x = -3.2, y = 0.12, size = 5, color = "coral",
           label = paste0("P/2 = ", round(pnorm(-z_obs), 4))) +
  annotate("text", x = 0, y = 0.3, size = 5, color = "coral",
           label = paste0("Total P = ", round(2*(1 - pnorm(z_obs)), 4))) +
  labs(title = expression(paste("Two-tailed: ", H[a], ": ", mu != mu[0])),
       x = "z", y = "Density")

p1 + p2 + p3
```

## P-Values for t Tests

For the one-sample t test with $n - 1$ degrees of freedom, the P-value formulas are analogous:

$$\text{P-value} = \begin{cases} P(T_{n-1} \geq t) & \text{upper-tailed} \\[6pt] P(T_{n-1} \leq t) & \text{lower-tailed} \\[6pt] 2 \cdot P(T_{n-1} \geq |t|) & \text{two-tailed} \end{cases}$$

. . .

```{r}
#| echo: true
# P-values for t tests in R
t_obs <- 2.15
df <- 14

cat("Upper-tailed P-value:", pt(t_obs, df = df, lower.tail = FALSE), "\n")
cat("Lower-tailed P-value:", pt(t_obs, df = df, lower.tail = TRUE), "\n")
cat("Two-tailed P-value:  ", 2 * pt(abs(t_obs), df = df, lower.tail = FALSE), "\n")
```

. . .

::: callout-tip
## R Functions
`t.test()` and `prop.test()` automatically compute P-values. Manual computation is needed only for understanding and for non-standard situations.
:::

# Medical Examples: Computing P-Values {background-color="#9b59b6"}

## Example: Hemoglobin A1c Levels in Diabetic Patients

**Setting:** A diabetes clinic is evaluating whether a new management protocol changes mean hemoglobin A1c (HbA1c) levels. The standard target is 7.0%. A sample of $n = 20$ patients on the new protocol gives $\bar{x} = 7.42\%$ and $s = 0.95\%$.

**Question:** Is there evidence that the new protocol produces different mean HbA1c than the target?

. . .

```{r}
#| echo: true
# Data summary
n <- 20
xbar <- 7.42
s <- 0.95
mu0 <- 7.0

# Test statistic
t_stat <- (xbar - mu0) / (s / sqrt(n))
cat("Test statistic: t =", round(t_stat, 3), "\n")
cat("Degrees of freedom:", n - 1, "\n")

# P-value (two-tailed)
p_value <- 2 * pt(abs(t_stat), df = n - 1, lower.tail = FALSE)
cat("P-value (two-tailed):", round(p_value, 4), "\n")
```

## HbA1c Example: Visualizing the P-Value

```{r}
#| code-fold: true
#| fig-height: 6
df <- 19
t_seq <- seq(-5, 5, length.out = 500)
y_t <- dt(t_seq, df = df)

tibble(t = t_seq, y = y_t) |>
  ggplot(aes(x = t, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(t = t_seq[t_seq >= abs(t_stat)], 
                          y = dt(t_seq[t_seq >= abs(t_stat)], df)),
            aes(x = t, y = y), fill = "coral", alpha = 0.6) +
  geom_area(data = tibble(t = t_seq[t_seq <= -abs(t_stat)], 
                          y = dt(t_seq[t_seq <= -abs(t_stat)], df)),
            aes(x = t, y = y), fill = "coral", alpha = 0.6) +
  geom_vline(xintercept = c(-abs(t_stat), abs(t_stat)), 
             linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = t_stat + 0.2, y = 0.15, hjust = 0, size = 5, color = "red",
           label = paste0("t = ", round(t_stat, 2))) +
  annotate("text", x = 0, y = 0.25, size = 6, color = "coral",
           label = paste0("P-value = ", round(p_value, 4))) +
  labs(x = "t", y = "Density",
       title = expression("Two-Tailed t Test: HbA1c Levels (" * t[19] * " distribution)"),
       subtitle = expression(paste(H[0], ": ", mu == 7.0, "% vs ", H[a], ": ", mu != 7.0, "%")))
```

## HbA1c Example: Complete Analysis with `t.test()`

```{r}
#| echo: true
# Simulate individual patient data consistent with our summary statistics
set.seed(2026)
hba1c <- rnorm(20, mean = 7.42, sd = 0.95)
# Adjust to match desired summary statistics closely
hba1c <- (hba1c - mean(hba1c)) / sd(hba1c) * 0.95 + 7.42

# Full t.test output
hba1c_test <- t.test(hba1c, mu = 7.0, alternative = "two.sided")
hba1c_test
```

. . .

**Interpretation:** At $\alpha = 0.05$, the P-value of `r round(hba1c_test$p.value, 4)` is less than 0.05, so we reject $H_0$. There is statistically significant evidence that the mean HbA1c under the new protocol differs from 7.0%.

## Example: Serum Transferrin Receptor in Pregnancy (Devore 9.4, Exercise 55)

**Setting:** The average serum transferrin receptor concentration for all women is known to be 5.63 mg/L. A study reported P-value > 0.10 for testing $H_0: \mu = 5.63$ vs $H_a: \mu \neq 5.63$ based on $n = 176$ pregnant women.

. . .

**Question:** Using $\alpha = 0.01$, what would you conclude?

. . .

```{r}
#| echo: true
alpha <- 0.01
p_value_reported <- 0.10  # P-value > 0.10

cat("Decision rule: Reject H0 if P-value <=", alpha, "\n")
cat("Observed: P-value > 0.10\n")
cat("Since 0.10 >", alpha, ", we fail to reject H0.\n\n")
cat("Conclusion: There is insufficient evidence at the 0.01\n")
cat("significance level to conclude that the mean serum\n")
cat("transferrin receptor concentration for pregnant women\n")
cat("differs from the value of 5.63 mg/L for all women.")
```

## Example: Post-Surgical Recovery Time

**Setting:** The standard mean recovery time after a particular surgery is 14 days. A surgeon has adopted a new minimally invasive technique and records recovery times for 15 patients.

```{r}
#| echo: true
recovery <- c(11.2, 13.5, 10.8, 12.1, 14.3, 9.7, 11.9, 12.8,
              10.5, 13.1, 11.4, 12.6, 10.2, 13.8, 11.1)

# One-sided test: H0: mu = 14 vs Ha: mu < 14
recovery_test <- t.test(recovery, mu = 14, alternative = "less")
recovery_test
```

. . .

```{r}
#| echo: true
cat("P-value =", format(recovery_test$p.value, scientific = TRUE), "\n\n")
cat("This extremely small P-value provides overwhelming evidence\n")
cat("that the new surgical technique reduces mean recovery time\n")
cat("below the standard 14 days.")
```

# The Distribution of P-Values (Devore Example 9.22) {background-color="#e67e22"}

## P-Values Are Random Variables

The P-value is computed from the test statistic, which is computed from sample data. Different samples yield different test statistics and therefore **different P-values**.

. . .

**Key theoretical results:**

- When $H_0$ is **true**: the P-value has a **Uniform(0, 1)** distribution
- When $H_0$ is **false**: the P-value distribution is **concentrated near 0**

. . .

::: callout-important
## Why Uniform Under $H_0$?
If $H_0$ is true, then $P(\text{P-value} \leq \alpha) = \alpha$ for any $\alpha \in [0, 1]$. This is precisely the definition of a Uniform(0, 1) distribution!
:::

## Recap: P-Value Uniformity via the Probability Integral Transform

Recall our earlier result: if $T$ has continuous CDF $F_T$ under $H_0$, then $P = 1 - F_T(T) \sim \text{Uniform}(0,1)$.

. . .

**For the z test specifically:**

Let $Z = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}} \sim N(0,1)$ under $H_0$. The P-value for an upper-tailed test is:

$$P = 1 - \Phi(Z)$$

. . .

Since $\Phi$ is the CDF of $Z$, this is exactly the survival function form, and the probability integral transformation gives us $P \sim \text{Uniform}(0,1)$.

## What Happens Under the Alternative?

Under $H_a: \mu > \mu_0$, the test statistic $Z = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}}$ is **not** standard normal.

. . .

Instead, $Z$ has mean $\frac{\mu - \mu_0}{\sigma/\sqrt{n}} > 0$, so $Z$ tends to be positive and large.

. . .

**Consequence for the P-value:**

$$P = 1 - \Phi(Z) = P_{H_0}(Z' \geq Z)$$

When $Z$ tends to be large and positive, $\Phi(Z) \to 1$, so $P \to 0$.

. . .

**The farther $\mu$ is from $\mu_0$:**

- The more $Z$ is shifted to the right
- The larger $\Phi(Z)$ becomes
- The smaller the P-value $P = 1 - \Phi(Z)$
- The more P-values concentrate near 0

## The Distribution of P-Values: A Unified View

Let $\delta = \frac{\mu - \mu_0}{\sigma/\sqrt{n}}$ be the **noncentrality parameter** (standardized effect size).

. . .

Under the true parameter value:

- $\delta = 0$ (i.e., $H_0$ true): $P \sim \text{Uniform}(0, 1)$
- $\delta > 0$ (upper-tailed $H_a$ true): $P$ has CDF $F_P(p) = 1 - \Phi(\Phi^{-1}(1-p) - \delta)$

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
# Theoretical CDF of P-value for different delta values
p_seq <- seq(0.001, 0.999, length.out = 500)
deltas <- c(0, 1, 2, 3)

pval_cdf <- expand_grid(p = p_seq, delta = deltas) |>
  mutate(
    F_p = 1 - pnorm(qnorm(1 - p) - delta),
    delta_label = paste0("delta == ", delta)
  )

ggplot(pval_cdf, aes(x = p, y = F_p, color = factor(delta))) +
  geom_line(linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  scale_color_viridis_d(
    labels = c(expression(delta == 0 ~ "(Uniform)"), 
               expression(delta == 1), 
               expression(delta == 2), 
               expression(delta == 3))
  ) +
  labs(x = "p", y = expression(F[P](p)),
       title = "CDF of P-Value Under Different Alternatives (Upper-Tailed z Test)",
       subtitle = "Larger delta concentrates the P-value distribution near 0",
       color = "Effect Size") +
  annotate("text", x = 0.5, y = 0.45, label = "Uniform(0,1) reference", 
           color = "gray50", size = 5) +
  theme(legend.position = c(0.85, 0.35),
        legend.text.align = 0)
```

## Power in Terms of P-Value Distribution

The **power** at a given alternative $\mu$ is the probability of rejecting $H_0$:

$$\text{Power}(\mu) = P_\mu(\text{Reject } H_0) = P_\mu(P \leq \alpha) = F_P(\alpha)$$

. . .

Using our formula for the P-value CDF under the alternative:

$$\text{Power}(\mu) = 1 - \Phi\left(\Phi^{-1}(1-\alpha) - \delta\right) = \Phi\left(\delta - z_{1-\alpha}\right)$$

where $\delta = \frac{\mu - \mu_0}{\sigma/\sqrt{n}}$ and $z_{1-\alpha} = \Phi^{-1}(1-\alpha)$.

. . .

**This connects P-value theory to power calculations!**

## Simulation: P-Value Distribution (Devore Example 9.22)

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 16
set.seed(551)
n_sims <- 10000
n <- 4
sigma <- 2

# Function to compute P-value from one-sample t test (upper-tailed)
sim_pvalue <- function(n, mu_true, mu0 = 30, sigma = 2) {
  x <- rnorm(n, mean = mu_true, sd = sigma)
  t_stat <- (mean(x) - mu0) / (sd(x) / sqrt(n))
  pt(t_stat, df = n - 1, lower.tail = FALSE)
}

# Simulate under three scenarios
pvals_H0_true   <- replicate(n_sims, sim_pvalue(n, mu_true = 30))  # H0 true
pvals_mu31      <- replicate(n_sims, sim_pvalue(n, mu_true = 31))  # H0 false, mu=31
pvals_mu32      <- replicate(n_sims, sim_pvalue(n, mu_true = 32))  # H0 false, mu=32

sim_data <- bind_rows(
  tibble(pvalue = pvals_H0_true, scenario = "H0 true (mu = 30)"),
  tibble(pvalue = pvals_mu31,    scenario = "H0 false (mu = 31)"),
  tibble(pvalue = pvals_mu32,    scenario = "H0 false (mu = 32)")
) |>
  mutate(scenario = fct_inorder(scenario))

ggplot(sim_data, aes(x = pvalue)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, 
                 fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 0.08, y = Inf, vjust = 2, hjust = 0,
           label = expression(alpha == 0.05), color = "red", size = 5) +
  facet_wrap(~scenario, ncol = 3) +
  labs(x = "P-value", y = "Density",
       title = "Distribution of P-Values: 10,000 Simulated t Tests (n = 4)",
       subtitle = expression(paste("Testing ", H[0], ": ", mu == 30, " vs ", 
                             H[a], ": ", mu > 30, " with ", sigma == 2))) +
  theme(strip.text = element_text(size = 14, face = "bold"))
```

## What the Simulation Shows

```{r}
#| echo: true
# Proportion of P-values <= 0.05 in each scenario
cat("Proportion rejecting H0 at alpha = 0.05:\n")
cat("  H0 true (mu = 30): ", round(mean(pvals_H0_true <= 0.05), 3), 
    " (should be ~0.05)\n")
cat("  H0 false (mu = 31):", round(mean(pvals_mu31 <= 0.05), 3), 
    " (power at mu = 31)\n")
cat("  H0 false (mu = 32):", round(mean(pvals_mu32 <= 0.05), 3), 
    " (power at mu = 32)\n")
```

. . .

**Key insights (Devore Example 9.22):**

- When $H_0$ is true, the P-value histogram is approximately **flat** (Uniform)
- Only ~5% of P-values fall below 0.05, confirming $\alpha$ controls the Type I error rate
- As $\mu$ moves farther from $\mu_0 = 30$, P-values concentrate closer to 0
- Even at $\mu = 32$, power is less than 50% because $n = 4$ is very small!

## Common Misinterpretations of P-Values

::: callout-warning
## What the P-Value is NOT

1. **Not** the probability that $H_0$ is true: P-value $\neq P(H_0 \text{ is true} \mid \text{data})$
2. **Not** the probability of making an error: it is $P(\text{data this extreme} \mid H_0 \text{ true})$
3. **Not** a measure of effect size: a small P-value can arise from a tiny effect with large $n$
4. A P-value of 0.05 does **not** mean there is a 5% chance $H_0$ is true
5. Failing to reject $H_0$ does **not** mean $H_0$ is true (absence of evidence $\neq$ evidence of absence)
:::

. . .

::: callout-note
## ASA Statement on P-Values (2016)
The American Statistical Association emphasized: "a P-value does not provide a good measure of evidence regarding a model or hypothesis." P-values should be supplemented with effect sizes, confidence intervals, and clinical judgment.
:::

## Statistical vs Clinical Significance

```{r}
#| echo: true
#| fig-height: 5
#| code-fold: true
# Large sample can detect trivially small effects
set.seed(551)
n_large <- 5000
bp_data <- rnorm(n_large, mean = 120.3, sd = 15)  # Mean barely above 120

bp_test <- t.test(bp_data, mu = 120)
cat("n =", n_large, "\n")
cat("Sample mean:", round(mean(bp_data), 2), "mmHg\n")
cat("P-value:", format(bp_test$p.value, digits = 4), "\n")
cat("95% CI: (", round(bp_test$conf.int[1], 2), ",", 
    round(bp_test$conf.int[2], 2), ")\n\n")
cat("Statistically significant (P < 0.05)? ", bp_test$p.value < 0.05, "\n")
cat("Clinically meaningful difference? A 0.3 mmHg difference\n")
cat("in systolic BP has NO clinical relevance!")
```

. . .

::: callout-tip
## Practical Advice for Medical Research
Always report **both** the P-value **and** a confidence interval. The CI gives a range of plausible values for the effect size, which is what clinicians actually need for decision-making.
:::

# Power for the t Test: The Noncentral t Distribution {background-color="#16a085"}

## Review: Power for the z Test (Lesson 13)

For the one-sample z test ($\sigma$ known), we derived closed-form expressions:

| Alternative | $\beta(\mu')$ |
|:---:|:---:|
| $H_a: \mu > \mu_0$ | $\Phi\!\left(z_\alpha + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right)$ |
| $H_a: \mu < \mu_0$ | $1 - \Phi\!\left(-z_\alpha + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right)$ |
| $H_a: \mu \neq \mu_0$ | $\Phi\!\left(z_{\alpha/2} + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right) - \Phi\!\left(-z_{\alpha/2} + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right)$ |

. . .

**Limitation:** These formulas require $\sigma$ to be known. In practice, we use the t test, and the analysis of power becomes more complex.

## The Noncentral t Distribution (Devore 9.2)

When $H_0$ is **false** (i.e., $\mu = \mu' \neq \mu_0$), the t test statistic

$$T = \frac{\bar{X} - \mu_0}{S/\sqrt{n}}$$

does **not** follow a $t_{n-1}$ distribution. Instead, it follows a **noncentral t distribution**.

. . .

::: callout-important
## Definition (Devore 9.2)
Let $Z \sim N(0,1)$ and $Y \sim \chi^2_m$ be independent. For any real number $\delta$, the random variable

$$\frac{Z + \delta}{\sqrt{Y/m}}$$

has a **noncentral t distribution** with $m$ degrees of freedom and **noncentrality parameter** $\delta$.

When $\delta = 0$, this reduces to the ordinary (central) $t_m$ distribution.
:::

## Noncentrality Parameter for the One-Sample t Test

When $\mu = \mu'$ (the true mean under $H_a$), the one-sample t statistic has a noncentral $t$ distribution with:

- **Degrees of freedom:** $n - 1$
- **Noncentrality parameter:** $\delta = \frac{\mu' - \mu_0}{\sigma/\sqrt{n}}$

. . .

The noncentrality parameter $\delta$ measures how far the true mean is from the null value, in units of the standard error.

. . .

```{r}
#| echo: true
# Example: mu_0 = 120, mu' = 125, sigma = 15, n = 25
mu0 <- 120; mu_prime <- 125; sigma <- 15; n <- 25
delta <- (mu_prime - mu0) / (sigma / sqrt(n))
cat("Noncentrality parameter: delta =", round(delta, 3))
```

## Visualizing the Noncentral t Distribution

```{r}
#| code-fold: true
#| fig-height: 7
t_seq <- seq(-4, 8, length.out = 500)
df <- 9
deltas <- c(0, 1, 2, 3)

nct_data <- map_dfr(deltas, \(d) {
  tibble(
    t = t_seq,
    density = dt(t_seq, df = df, ncp = d),
    delta = paste0("delta == ", d)
  )
})

ggplot(nct_data, aes(x = t, y = density, color = delta)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(
    values = c("steelblue", "darkgreen", "darkorange", "red"),
    labels = c(
      expression(delta == 0 ~ "(central t)"),
      expression(delta == 1),
      expression(delta == 2),
      expression(delta == 3)
    )
  ) +
  labs(x = "t", y = "Density", color = "Noncentrality",
       title = expression("Noncentral t Distribution (df = 9)"),
       subtitle = expression("Larger " * delta * " shifts the distribution right" *
                             " " * symbol("\256") * " more power for upper-tailed tests")) +
  theme(legend.text.align = 0, legend.position = "right")
```

## Computing Power with the Noncentral t

**Power for an upper-tailed test** $H_a: \mu > \mu_0$:

$$\text{Power} = P\!\left(T \geq t_{\alpha, n-1}\right) = 1 - F\!\left(t_{\alpha, n-1};\, n-1,\, \delta\right)$$

where $F(x; m, \delta)$ is the CDF of the noncentral $t$ distribution.

. . .

**In R:** `pt(t_crit, df, ncp = delta, lower.tail = FALSE)`

. . .

```{r}
#| echo: true
# Example (Devore Example 9.14 adapted to medical context):
# Testing whether mean systolic BP exceeds 120 mmHg
# H0: mu = 120, Ha: mu > 120, alpha = 0.05, n = 10, sigma = 10
# What is the power when mu' = 125?
n <- 10; mu0 <- 120; mu_prime <- 125; sigma <- 10; alpha <- 0.05

delta <- (mu_prime - mu0) / (sigma / sqrt(n))
t_crit <- qt(1 - alpha, df = n - 1)

power <- pt(t_crit, df = n - 1, ncp = delta, lower.tail = FALSE)
cat("Noncentrality parameter: delta =", round(delta, 3), "\n")
cat("Critical value: t_{0.05, 9} =", round(t_crit, 3), "\n")
cat("Power =", round(power, 4), "\n")
cat("Beta (Type II error) =", round(1 - power, 4))
```

## Power Curves for the t Test: Medical Example

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
alpha <- 0.05
mu0 <- 120  # null value for systolic BP
sigma <- 10

# Power curves for different sample sizes
sample_sizes <- c(10, 25, 50, 100)
mu_range <- seq(120, 130, by = 0.25)

power_data <- expand_grid(n = sample_sizes, mu_prime = mu_range) |>
  mutate(
    delta = (mu_prime - mu0) / (sigma / sqrt(n)),
    t_crit = qt(1 - alpha, df = n - 1),
    power = pt(t_crit, df = n - 1, ncp = delta, lower.tail = FALSE),
    n_label = paste0("n = ", n)
  )

ggplot(power_data, aes(x = mu_prime, y = power, color = n_label)) +
  geom_line(linewidth = 1.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = alpha, linetype = "dotted", color = "gray60") +
  annotate("text", x = 129.5, y = 0.82, label = "80% power", 
           color = "gray40", size = 5) +
  annotate("text", x = 129.5, y = 0.07, 
           label = expression(alpha == 0.05), color = "gray60", size = 4) +
  scale_color_brewer(palette = "Set1") +
  labs(x = expression("True " * mu * " (mmHg)"),
       y = "Power",
       color = "Sample Size",
       title = "Power Curves: One-Sided t Test for Systolic Blood Pressure",
       subtitle = expression(paste(H[0], ": ", mu == 120, " vs ", H[a], ": ", 
                             mu > 120, ",  ", sigma == 10, ",  ", alpha == 0.05)))
```

## Key Features of Power Curves

**Two intuitive features:**

1. For any **fixed** departure from $H_0$, power **increases** with sample size
2. For any **fixed** sample size, power **increases** as $\mu'$ moves farther from $\mu_0$

. . .

**Factors that increase power:**

- **Larger $n$:** more data $\Rightarrow$ smaller SE $\Rightarrow$ easier to detect departures
- **Larger effect size** $|\mu' - \mu_0|$: bigger departures are easier to detect
- **Smaller $\sigma$:** less noise $\Rightarrow$ easier to detect signal
- **Larger $\alpha$:** easier to reject (but at the cost of more Type I errors)

. . .

::: callout-tip
## Clinical Trial Implication
The power analysis must be conducted **before** collecting data. It ensures the study has sufficient sample size to detect a **clinically meaningful** effect.
:::

# Sample Size Determination {background-color="#e74c3c"}

## Why Determine Sample Size?

**Clinical scenario:** You are designing a randomized trial to test whether a new antihypertensive drug lowers systolic blood pressure more than a placebo.

. . .

**Key questions:**

- How many patients do we need to **detect** a meaningful effect?
- What effect size is **clinically important**?
- What balance of Type I and Type II errors is acceptable?

. . .

::: callout-important
## The Three Ingredients for Sample Size
1. **Significance level** $\alpha$ (usually 0.05)
2. **Desired power** $1 - \beta$ (usually 0.80 or 0.90)
3. **Clinically meaningful effect size** $\Delta = |\mu' - \mu_0|$ (determined by clinical judgment)
4. **(Estimated) standard deviation** $\sigma$
:::

## Sample Size Formula for the z Test (Review)

For the one-sample z test, the required sample size for power $1 - \beta$ at alternative $\mu'$:

$$n = \begin{cases} \left[\dfrac{\sigma(z_\alpha + z_\beta)}{\mu_0 - \mu'}\right]^2 & \text{one-tailed test} \\[12pt] \left[\dfrac{\sigma(z_{\alpha/2} + z_\beta)}{\mu_0 - \mu'}\right]^2 & \text{two-tailed test (approximate)} \end{cases}$$

. . .

```{r}
#| echo: true
# How many patients to detect a 5 mmHg reduction with 80% power?
delta <- 5  # clinically meaningful difference (mmHg)
sigma <- 15  # estimated SD
alpha <- 0.05
beta <- 0.20  # power = 0.80

# Two-tailed z test
n_z <- (sigma * (qnorm(1 - alpha/2) + qnorm(1 - beta)) / delta)^2
cat("z test sample size (two-tailed):", ceiling(n_z), "patients")
```

## R's `power.t.test()` Function

R provides `power.t.test()` for sample size and power calculations for the one-sample and two-sample t tests. It accounts for the noncentral t distribution automatically.

. . .

```{r}
#| echo: true
# Same example using power.t.test()
result <- power.t.test(
  delta = 5,           # difference to detect
  sd = 15,             # estimated standard deviation
  sig.level = 0.05,    # alpha
  power = 0.80,        # desired power
  type = "one.sample", # one-sample test
  alternative = "two.sided"
)
result
```

. . .

```{r}
#| echo: true
cat("Required sample size:", ceiling(result$n), "patients")
```

## Example: Planning a Clinical Trial for Blood Pressure

**Setting:** Researchers are planning a trial to test whether a new drug reduces systolic blood pressure. Historical data suggests $\sigma \approx 12$ mmHg. A reduction of 5 mmHg is considered clinically meaningful.

```{r}
#| echo: true
# One-sided test: H0: mu >= 0 vs Ha: mu < 0 (reduction in BP)
power_result <- power.t.test(
  delta = 5,
  sd = 12,
  sig.level = 0.05,
  power = 0.80,
  type = "one.sample",
  alternative = "one.sided"
)
cat("For 80% power:", ceiling(power_result$n), "patients needed\n")

# What about 90% power?
power_90 <- power.t.test(
  delta = 5, sd = 12, sig.level = 0.05,
  power = 0.90, type = "one.sample", alternative = "one.sided"
)
cat("For 90% power:", ceiling(power_90$n), "patients needed\n")

# What about a smaller effect (3 mmHg)?
power_small <- power.t.test(
  delta = 3, sd = 12, sig.level = 0.05,
  power = 0.80, type = "one.sample", alternative = "one.sided"
)
cat("For 3 mmHg effect:", ceiling(power_small$n), "patients needed")
```

## Sample Size as a Function of Effect Size and Power

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 14
sigma <- 12
alpha <- 0.05

# Vary effect size
delta_range <- seq(2, 10, by = 0.5)
power_levels <- c(0.80, 0.90, 0.95)

ss_data <- expand_grid(delta = delta_range, power = power_levels) |>
  rowwise() |>
  mutate(
    n = ceiling(power.t.test(
      delta = delta, sd = sigma, sig.level = alpha,
      power = power, type = "one.sample", alternative = "two.sided"
    )$n),
    power_label = paste0("Power = ", power * 100, "%")
  ) |>
  ungroup()

p_ss <- ggplot(ss_data, aes(x = delta, y = n, color = power_label)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(x = expression("Clinically Meaningful Difference " * Delta * " (mmHg)"),
       y = "Required Sample Size (n)",
       color = "",
       title = "Sample Size vs Effect Size for One-Sample t Test",
       subtitle = expression(paste(sigma == 12, " mmHg, ", alpha == 0.05, 
                             ", two-tailed"))) +
  theme(legend.position = "top")

# Power as a function of n for fixed delta
n_range <- seq(5, 200, by = 5)
delta_levels <- c(3, 5, 8)

pow_data <- expand_grid(n = n_range, delta = delta_levels) |>
  rowwise() |>
  mutate(
    power = power.t.test(
      n = n, delta = delta, sd = sigma, sig.level = alpha,
      type = "one.sample", alternative = "two.sided"
    )$power,
    delta_label = paste0("Delta = ", delta, " mmHg")
  ) |>
  ungroup()

p_pow <- ggplot(pow_data, aes(x = n, y = power, color = delta_label)) +
  geom_line(linewidth = 1.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Sample Size (n)", y = "Power",
       color = "",
       title = "Power vs Sample Size",
       subtitle = expression(paste(sigma == 12, " mmHg, ", alpha == 0.05, 
                             ", two-tailed"))) +
  theme(legend.position = "top")

p_ss + p_pow
```

## Example: Neonatal ICU Glucose Monitoring

**Setting:** A neonatal ICU has historically maintained mean blood glucose at 80 mg/dL (sd = 20 mg/dL) in premature infants. A new feeding protocol is introduced. The clinical team wants to detect any shift in mean glucose of at least 8 mg/dL with 90% power.

```{r}
#| echo: true
# How many infants are needed?
nicu_power <- power.t.test(
  delta = 8,
  sd = 20,
  sig.level = 0.05,
  power = 0.90,
  type = "one.sample",
  alternative = "two.sided"
)

cat("=== NICU Study Design ===\n")
cat("Effect to detect: 8 mg/dL\n")
cat("Assumed SD: 20 mg/dL\n")
cat("Significance level: 0.05\n")
cat("Desired power: 0.90\n")
cat("Required sample size:", ceiling(nicu_power$n), "infants\n\n")

# Verify: what is the actual power at this sample size?
actual_power <- power.t.test(
  n = ceiling(nicu_power$n),
  delta = 8,
  sd = 20,
  sig.level = 0.05,
  type = "one.sample",
  alternative = "two.sided"
)
cat("Actual power at n =", ceiling(nicu_power$n), ":", 
    round(actual_power$power, 4))
```

## Manual Power Calculation with the Noncentral t

Let's verify `power.t.test()` using the noncentral t distribution directly:

```{r}
#| echo: true
# NICU example: n = 70, delta = 8, sigma = 20, alpha = 0.05, two-tailed
n <- ceiling(nicu_power$n)
mu0 <- 80; mu_prime <- 88; sigma <- 20
alpha <- 0.05

# Noncentrality parameter
delta_ncp <- (mu_prime - mu0) / (sigma / sqrt(n))
cat("Noncentrality parameter: delta =", round(delta_ncp, 3), "\n")

# Critical values for two-tailed test
t_crit <- qt(1 - alpha/2, df = n - 1)
cat("Critical value: t_{0.025,", n-1, "} =", round(t_crit, 3), "\n")

# Power = P(|T| >= t_crit) under noncentral t
power_upper <- pt(t_crit, df = n - 1, ncp = delta_ncp, lower.tail = FALSE)
power_lower <- pt(-t_crit, df = n - 1, ncp = delta_ncp, lower.tail = TRUE)
power_total <- power_upper + power_lower

cat("Power (upper tail):", round(power_upper, 4), "\n")
cat("Power (lower tail):", round(power_lower, 6), " (negligible)\n")
cat("Total power:", round(power_total, 4))
```

# Connecting P-Values, Power, and Sample Size {background-color="#8e44ad"}

## The Complete Picture

```{r}
#| code-fold: true
#| fig-height: 8
#| fig-width: 16
# Comprehensive example: How P-values, power, and n interact
set.seed(551)
mu0 <- 120; mu_true <- 125; sigma <- 15

# Simulate P-values for different sample sizes
sim_comprehensive <- function(n, mu_true, mu0 = 120, sigma = 15, n_sims = 5000) {
  replicate(n_sims, {
    x <- rnorm(n, mean = mu_true, sd = sigma)
    t_stat <- (mean(x) - mu0) / (sd(x) / sqrt(n))
    2 * pt(abs(t_stat), df = n - 1, lower.tail = FALSE)  # two-tailed
  })
}

pvals_n10  <- sim_comprehensive(10,  mu_true)
pvals_n30  <- sim_comprehensive(30,  mu_true)
pvals_n100 <- sim_comprehensive(100, mu_true)

comp_data <- bind_rows(
  tibble(pvalue = pvals_n10,  n_label = "n = 10"),
  tibble(pvalue = pvals_n30,  n_label = "n = 30"),
  tibble(pvalue = pvals_n100, n_label = "n = 100")
) |>
  mutate(n_label = fct_inorder(n_label))

p_hist <- ggplot(comp_data, aes(x = pvalue)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", linewidth = 1) +
  facet_wrap(~n_label, ncol = 3, scales = "free_y") +
  labs(x = "P-value", y = "Count",
       title = expression("P-value Distributions When " * H[0] * 
                          " Is False (" * mu == 125 * ")"),
       subtitle = expression(paste(H[0], ": ", mu == 120, ", two-tailed, ",
                             sigma == 15))) +
  theme(strip.text = element_text(size = 13, face = "bold"))

# Power summary
power_summary <- comp_data |>
  group_by(n_label) |>
  summarise(
    prop_reject = mean(pvalue <= 0.05),
    median_pval = median(pvalue),
    .groups = "drop"
  )

p_table <- ggplot(power_summary, aes(x = n_label, y = prop_reject)) +
  geom_col(fill = "steelblue", alpha = 0.7, width = 0.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_text(aes(label = paste0(round(prop_reject * 100, 1), "%")), 
            vjust = -0.5, size = 6) +
  annotate("text", x = 3.4, y = 0.82, label = "80% power", color = "red", size = 5) +
  scale_y_continuous(limits = c(0, 1.05), labels = scales::percent) +
  labs(x = "", y = "Proportion Rejecting H0\n(Simulated Power)",
       title = "Power Increases with Sample Size")

p_hist / p_table + plot_layout(heights = c(2, 1.2))
```

## Summary: P-Values and Their Relationship to Testing

| Concept | What It Tells You |
|---------|-------------------|
| **P-value** | Strength of evidence against $H_0$ (smaller = stronger) |
| **$\alpha$** | Threshold for "significant" evidence (set in advance) |
| **$\beta(\mu')$** | Probability of missing a true effect of size $\mu'$ |
| **Power** $= 1 - \beta$ | Probability of detecting a true effect |
| **Sample size** $n$ | Number of observations needed for desired power |

. . .

**The chain of reasoning:**

$$\text{Larger } n \implies \text{Smaller SE} \implies \text{Larger } |T| \implies \text{Smaller P-values} \implies \text{More Power}$$

## Practical Guidelines for Medical Research

::: callout-tip
## Best Practices

1. **Always report P-values AND confidence intervals** --- the CI tells you about effect size
2. **Conduct power analysis BEFORE the study** --- not after!
3. **Use clinically meaningful effect sizes** --- not just "any" difference
4. **Don't p-hack:** pre-register hypotheses and analysis plans
5. **Interpret in context:** a P-value of 0.048 vs 0.052 should not lead to radically different conclusions
6. **Consider multiple comparisons:** if testing many hypotheses, adjust $\alpha$ (Bonferroni, FDR, etc.)
:::

. . .

::: callout-important
## FDA Guidance
Clinical trials typically require $\alpha = 0.05$ (two-sided) and power $\geq 0.80$. Phase III trials often require power $\geq 0.90$ to ensure adequate evidence for drug approval.
:::

# Key Takeaways and Looking Ahead {background-color="#2c3e50"}

## Key Takeaways

::: callout-important
## Today's Main Points

1. **P-value definition:** probability of observing data as extreme or more extreme, assuming $H_0$ is true
2. **P-value decision rule:** reject $H_0$ when P-value $\leq \alpha$ (equivalent to rejection region method)
3. **P-value formulas:** depend on whether the test is upper-tailed, lower-tailed, or two-tailed
4. **P-value distribution:** Uniform(0, 1) under $H_0$; concentrated near 0 under $H_a$
5. **Noncentral t:** when $H_0$ is false, $T$ follows a noncentral $t$ with $\delta = (\mu' - \mu_0)/(\sigma/\sqrt{n})$
6. **Power in R:** `pt(t_crit, df, ncp = delta)` for manual; `power.t.test()` for convenience
7. **Sample size:** determined by $\alpha$, power, $\sigma$, and the clinically meaningful effect size
:::

. . .

**Common mistakes to avoid:**

- Interpreting P-value as $P(H_0 \text{ true})$
- Confusing statistical and clinical significance
- Computing power **after** seeing the data (post hoc power)
- Using $S$ (from data) in place of $\sigma$ in the z test power formula

## Looking Ahead

**Next class (Lesson 15):** Neyman-Pearson Lemma; Likelihood Ratio Tests (Devore 9.5)

- Most powerful tests for simple hypotheses
- The Neyman-Pearson framework
- Likelihood ratio test statistic
- Connection to tests we've already learned

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Sections 9.2, 9.4
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Sections 8.1--8.2
-   Wasserstein, R. L., & Lazar, N. A. (2016). The ASA Statement on p-Values: Context, Process, and Purpose. *The American Statistician*, 70(2), 129--133.
:::

## Questions? {.center background-color="#3498db"}

Thank you!


```{bash}
#| eval: false
#| include: false
# For PDF export
decktape reveal --fragments docs/lessons/14_PValues_Power_SampleSize/14_PValues_Power_SampleSize.html lessons/14_PValues_Power_SampleSize/14_PValues_Power_SampleSize.pdf
```
