---
title: "Lesson 7: Confidence Intervals"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w4d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 7 - Confidence Intervals
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 7: From Point Estimates to Intervals {background-color="#2c3e50"}

## Review: What We've Learned About Estimation

**Point Estimation (Lessons 1-6):**

-   Estimators: statistics that estimate parameters
-   Properties: bias, variance, MSE, efficiency
-   Construction: Method of Moments, Maximum Likelihood
-   Quality: MVUE, sufficient statistics, Cramér-Rao bound

. . .

**The Limitation of Point Estimates:**

A single number ($\hat{\theta} = 0.75$) tells us nothing about **precision** or **uncertainty**.

## Today's Goals

::: incremental
1.  Understand why we need interval estimates
2.  Derive confidence intervals using pivotal quantities
3.  Construct and interpret z-intervals (known $\sigma$)
4.  Construct and interpret t-intervals (unknown $\sigma$)
5.  Understand confidence intervals for proportions
6.  Apply these methods to medical research examples
:::

## Motivating Example: Clinical Trial

A new antihypertensive drug is tested in $n = 36$ patients. The mean reduction in systolic blood pressure is $\bar{x} = 12.3$ mmHg.

. . .

**Point estimate:** $\hat{\mu} = 12.3$ mmHg

. . .

**Questions we can't answer with just the point estimate:**

-   How close is 12.3 to the true mean reduction?
-   Is the true mean likely to be at least 10 mmHg?
-   How much could $\bar{X}$ vary in repeated trials?

. . .

::: callout-note
## The Solution
A **confidence interval** provides a range of plausible values for the parameter, along with a measure of confidence.
:::

# The Fundamental Idea {background-color="#27ae60"}

## From Point to Interval

```{r}
#| echo: false
#| fig-width: 14
#| fig-height: 6
tibble(x = seq(5, 20, by = 0.1)) |>
  mutate(y = dnorm(x, mean = 12.3, sd = 2.5)) |>
  ggplot(aes(x = x, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_vline(xintercept = 12.3, color = "red", linetype = "dashed", linewidth = 1.2) +
  geom_segment(aes(x = 7.4, xend = 17.2, y = 0.02, yend = 0.02), 
               color = "darkgreen", linewidth = 2,
               arrow = arrow(ends = "both", length = unit(0.2, "cm"))) +
  annotate("text", x = 12.3, y = 0.18, label = expression(bar(x) == 12.3), size = 7, color = "red") +
  annotate("text", x = 12.3, y = 0.05, label = "95% CI: (7.4, 17.2)", size = 6, color = "darkgreen") +
  labs(x = "Blood Pressure Reduction (mmHg)", y = "Density",
       title = "Point Estimate with Confidence Interval") +
  theme_minimal(base_size = 18)
```

**Key insight:** The interval captures our uncertainty about the true parameter value.

## What Is a Confidence Interval?

::: callout-important
## Definition: Confidence Interval
A **confidence interval (CI)** for a parameter $\theta$ is a random interval $(L, U)$ computed from sample data such that:
$$P(L < \theta < U) = 1 - \alpha$$
for a specified confidence level $1 - \alpha$.
:::

. . .

**Important:**

-   $L$ and $U$ are **random** (they depend on the sample)
-   $\theta$ is **fixed** (but unknown)
-   The probability statement is about the **procedure**, not about $\theta$

## Correct Interpretation of Confidence Intervals

::: callout-warning
## Common Misconception
"There is a 95% probability that $\mu$ lies in the interval (7.4, 17.2)."

**This is WRONG!** After computing the interval, $\mu$ either is or is not in the interval. There's no probability involved.
:::

. . .

::: callout-tip
## Correct Interpretation
"If we repeated this procedure many times, 95% of the resulting intervals would contain the true $\mu$."

Or more simply: "We are 95% confident that the true mean lies between 7.4 and 17.2."
:::

## Visualizing Confidence Level: Simulation

```{r}
#| echo: true
#| fig-height: 6
# Simulate 100 confidence intervals
set.seed(551)
true_mu <- 10; sigma <- 4; n <- 25

ci_data <- tibble(sample_id = 1:100) |>
  rowwise() |>
  mutate(
    xbar = mean(rnorm(n, true_mu, sigma)),
    lower = xbar - 1.96 * sigma / sqrt(n),
    upper = xbar + 1.96 * sigma / sqrt(n),
    contains_mu = lower < true_mu & upper > true_mu
  )

# Count how many contain the true mean
sum(ci_data$contains_mu)
```

## Visualizing Confidence Level: The Plot

```{r}
#| echo: false
#| fig-height: 8
ci_data |>
  slice_head(n = 50) |>
  ggplot(aes(y = sample_id, color = contains_mu)) +
  geom_segment(aes(x = lower, xend = upper, yend = sample_id), linewidth = 0.8) +
  geom_point(aes(x = xbar), size = 2) +
  geom_vline(xintercept = true_mu, color = "red", linetype = "dashed", linewidth = 1.2) +
  scale_color_manual(values = c("FALSE" = "red", "TRUE" = "steelblue"),
                     labels = c("Misses μ", "Contains μ")) +
  labs(x = "Value", y = "Sample Number", 
       title = "50 Confidence Intervals (95% Level)",
       subtitle = paste("True μ =", true_mu, "(red dashed line)"),
       color = "") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "top")
```

## The R Code for CI Visualization

```{r}
#| eval: false
#| echo: true
# Simulate 100 confidence intervals
set.seed(551)
true_mu <- 10; sigma <- 4; n <- 25

ci_data <- tibble(sample_id = 1:100) |>
  rowwise() |>
  mutate(
    xbar = mean(rnorm(n, true_mu, sigma)),
    lower = xbar - 1.96 * sigma / sqrt(n),
    upper = xbar + 1.96 * sigma / sqrt(n),
    contains_mu = lower < true_mu & upper > true_mu
  )

# Proportion containing true mu (should be ~95%)
mean(ci_data$contains_mu)
```

This is how the confidence level works: about 95% of intervals capture $\mu$.

# Pivotal Quantities {background-color="#e74c3c"}

## The Key to Constructing CIs

::: callout-important
## Definition: Pivotal Quantity
A **pivotal quantity** is a function $h(X_1, \ldots, X_n; \theta)$ of the data and parameter such that:

1.  It depends on both the sample and $\theta$
2.  Its **distribution does not depend on $\theta$** (or any other unknown parameters)
:::

. . .

**Why pivotal quantities matter:**

-   We can find critical values that don't depend on unknowns
-   We can "solve for $\theta$" to get confidence limits

## Example: The Standard Pivotal Quantity

Let $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with $\sigma$ known.

. . .

Consider:
$$Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}}$$

. . .

**Check the properties:**

1.  $Z$ depends on $\bar{X}$ (data) and $\mu$ (parameter) ✓
2.  $Z \sim N(0, 1)$ — the distribution doesn't depend on $\mu$ ✓

. . .

$Z$ is a **pivotal quantity**!

## From Pivotal Quantity to CI

Since $Z \sim N(0, 1)$, we know:
$$P(-z_{\alpha/2} < Z < z_{\alpha/2}) = 1 - \alpha$$

. . .

Substitute $Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}}$:
$$P\left(-z_{\alpha/2} < \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} < z_{\alpha/2}\right) = 1 - \alpha$$

. . .

**Solve for $\mu$:**
$$P\left(\bar{X} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}} < \mu < \bar{X} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right) = 1 - \alpha$$

. . .

This gives us the **confidence interval formula**!

# CI for Mean: $\sigma$ Known {background-color="#3498db"}

## The Z-Interval

::: callout-important
## Z Confidence Interval for Normal Mean ($\sigma$ Known)
If $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$ with $\sigma$ known, a $100(1-\alpha)\%$ CI for $\mu$ is:
$$\bar{x} \pm z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$
where $z_{\alpha/2}$ is the critical value with $P(Z > z_{\alpha/2}) = \alpha/2$.
:::

. . .

**Common critical values:**

| Confidence Level | $\alpha$ | $\alpha/2$ | $z_{\alpha/2}$ |
|------------------|----------|------------|----------------|
| 90% | 0.10 | 0.05 | 1.645 |
| 95% | 0.05 | 0.025 | 1.960 |
| 99% | 0.01 | 0.005 | 2.576 |

## Example: Tensile Strength (Devore Example 8.1)

Synthetic fibers are tested for tensile strength. From past data, $\sigma = 55$ MPa is known. A sample of $n = 31$ fibers has $\bar{x} = 1064$ MPa.

. . .

**95% Confidence Interval:**

```{r}
xbar <- 1064
sigma <- 55
n <- 31
alpha <- 0.05

z_crit <- qnorm(1 - alpha/2)
margin <- z_crit * sigma / sqrt(n)

c(xbar - margin, xbar + margin)
```

. . .

**Interpretation:** We are 95% confident that the true mean tensile strength is between 1044.6 and 1083.4 MPa.

## Medical Example: Blood Glucose Levels

A hospital monitors fasting blood glucose in diabetic patients. Historical data shows $\sigma = 18$ mg/dL. A sample of $n = 40$ patients has $\bar{x} = 142$ mg/dL.

**Construct a 99% CI for the population mean glucose level:**

```{r}
xbar <- 142
sigma <- 18
n <- 40
alpha <- 0.01

z_crit <- qnorm(1 - alpha/2)
cat("z critical value:", round(z_crit, 3), "\n")

margin <- z_crit * sigma / sqrt(n)
cat("Margin of error:", round(margin, 2), "mg/dL\n")

cat("99% CI: (", round(xbar - margin, 1), ",", round(xbar + margin, 1), ")")
```

. . .

**Clinical interpretation:** We can be 99% confident that the true mean fasting glucose for this patient population is between 134.7 and 149.3 mg/dL.

## Anatomy of a Confidence Interval

$$\bar{x} \pm z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$

```{r}
#| echo: false
#| fig-height: 5
tibble(x = c(1, 2, 3)) |>
  ggplot() +
  annotate("text", x = 1.5, y = 0.8, label = "Point Estimate", size = 7, hjust = 0.5) +
  annotate("text", x = 1.5, y = 0.6, label = expression(bar(x)), size = 10, hjust = 0.5, color = "red") +
  annotate("text", x = 2.5, y = 0.8, label = "Margin of Error", size = 7, hjust = 0.5) +
  annotate("text", x = 2.5, y = 0.55, label = expression(z[alpha/2] %.% frac(sigma, sqrt(n))), 
           size = 9, hjust = 0.5, color = "steelblue") +
  annotate("text", x = 2, y = 0.3, label = "↑ Confidence Level    ↑ Precision    ↑ Sample Size", 
           size = 6, hjust = 0.5) +
  xlim(1, 3) + ylim(0, 1) +
  theme_void()
```

**Three factors affect the margin of error:**

1. **Confidence level** ($z_{\alpha/2}$): Higher confidence → wider interval
2. **Population variability** ($\sigma$): More variation → wider interval
3. **Sample size** ($n$): Larger sample → narrower interval

## Interactive: Effect of Confidence Level

```{r}
#| echo: true
# Same data, different confidence levels
xbar <- 1064; sigma <- 55; n <- 31

confidence_levels <- c(0.90, 0.95, 0.99)

tibble(conf_level = confidence_levels) |>
  mutate(
    z_crit = qnorm(1 - (1 - conf_level)/2),
    margin = z_crit * sigma / sqrt(n),
    lower = xbar - margin,
    upper = xbar + margin,
    width = upper - lower
  )
```

**Takeaway:** Higher confidence requires wider intervals!

## Sample Size Determination

**Question:** How large should $n$ be to achieve a desired precision?

. . .

If we want the margin of error to be at most $b$:
$$z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}} \leq b$$

. . .

**Solving for $n$:**
$$n \geq \left(\frac{z_{\alpha/2} \cdot \sigma}{b}\right)^2$$

## Sample Size Example: Clinical Trial Design

A pharmaceutical company wants to estimate mean drug clearance time. From pilot data, $\sigma \approx 25$ minutes. They want a 95% CI with margin of error ≤ 5 minutes.

```{r}
sigma <- 25
b <- 5  # desired margin of error
alpha <- 0.05
z_crit <- qnorm(1 - alpha/2)

n_required <- (z_crit * sigma / b)^2
cat("Required sample size:", ceiling(n_required))
```

. . .

**Note:** Always round **up** to ensure the desired precision!

## One-Sided Confidence Bounds

Sometimes we only care about one direction:

-   **Lower bound:** We want to know the minimum plausible value
-   **Upper bound:** We want to know the maximum plausible value

. . .

::: callout-tip
## One-Sided Bounds
**Lower Confidence Bound:** $\bar{x} - z_{\alpha} \cdot \frac{\sigma}{\sqrt{n}} < \mu$

**Upper Confidence Bound:** $\mu < \bar{x} + z_{\alpha} \cdot \frac{\sigma}{\sqrt{n}}$

Note: Use $z_\alpha$ (not $z_{\alpha/2}$) for one-sided bounds at confidence level $1 - \alpha$.
:::

## Medical Example: Drug Half-Life

A pharmacologist measures half-life of a new drug. With $n = 20$ subjects, $\bar{x} = 4.2$ hours, and known $\sigma = 0.8$ hours.

**Find a 95% lower confidence bound:**

```{r}
xbar <- 4.2
sigma <- 0.8
n <- 20
alpha <- 0.05

z_crit <- qnorm(1 - alpha)  # Note: z_alpha, not z_alpha/2
lower_bound <- xbar - z_crit * sigma / sqrt(n)

cat("95% lower confidence bound:", round(lower_bound, 2), "hours\n")
cat("Interpretation: We are 95% confident that the true half-life",
    "\nis at least", round(lower_bound, 2), "hours")
```

# CI for Mean: $\sigma$ Unknown {background-color="#9b59b6"}

## The Realistic Case

In practice, we **rarely know** the population standard deviation $\sigma$.

. . .

**Natural idea:** Replace $\sigma$ with the sample standard deviation $s$.

. . .

But what happens to our pivotal quantity?

$$Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim N(0, 1)$$

becomes

$$T = \frac{\bar{X} - \mu}{S/\sqrt{n}} \sim \; ???$$

## The t Distribution

```{r}
#| echo: false
#| fig-height: 6
tibble(x = seq(-4, 4, by = 0.01)) |>
  mutate(
    normal = dnorm(x),
    t_5 = dt(x, df = 5),
    t_15 = dt(x, df = 15),
    t_30 = dt(x, df = 30)
  ) |>
  pivot_longer(-x, names_to = "distribution", values_to = "density") |>
  mutate(distribution = factor(distribution, 
                               levels = c("normal", "t_30", "t_15", "t_5"),
                               labels = c("N(0,1)", "t(df=30)", "t(df=15)", "t(df=5)"))) |>
  ggplot(aes(x = x, y = density, color = distribution)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("black", "#3498db", "#e74c3c", "#27ae60")) +
  labs(x = "x", y = "Density", color = "Distribution",
       title = "Student's t Distribution vs. Standard Normal") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "right")
```

The t distribution has **heavier tails** than the normal — accounting for extra uncertainty from estimating $\sigma$.

## Student's t Distribution

::: callout-important
## Gosset's Theorem
If $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$, then:
$$T = \frac{\bar{X} - \mu}{S/\sqrt{n}} \sim t_{n-1}$$
where $t_{n-1}$ denotes the t distribution with $n-1$ **degrees of freedom**.
:::

. . .

**Key properties of $t_\nu$:**

-   Symmetric around 0
-   Heavier tails than $N(0, 1)$
-   As $\nu \to \infty$, $t_\nu \to N(0, 1)$

## Why $(n-1)$ Degrees of Freedom?

The sample variance involves:
$$S^2 = \frac{1}{n-1}\sum_{i=1}^n(X_i - \bar{X})^2$$

. . .

Since $\sum(X_i - \bar{X}) = 0$, knowing any $n-1$ deviations determines the last one.

. . .

**Intuition:** We "use up" one degree of freedom by estimating $\mu$ with $\bar{X}$.

## The One-Sample t Interval

::: callout-important
## t Confidence Interval for Normal Mean ($\sigma$ Unknown)
If $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$, a $100(1-\alpha)\%$ CI for $\mu$ is:
$$\bar{x} \pm t_{\alpha/2, n-1} \cdot \frac{s}{\sqrt{n}}$$
where $t_{\alpha/2, n-1}$ is the critical value with $P(T > t_{\alpha/2, n-1}) = \alpha/2$ for $T \sim t_{n-1}$.
:::

. . .

The quantity $s/\sqrt{n}$ is called the **standard error of the mean** (SEM or SE).

## t Critical Values

```{r}
# Compare z and t critical values for 95% CI
z_crit <- qnorm(0.975)

df_values <- c(5, 10, 20, 30, 50, 100, Inf)
t_crits <- qt(0.975, df = df_values)

tibble(
  df = c("5", "10", "20", "30", "50", "100", "∞ (z)"),
  t_critical = round(c(head(t_crits, -1), z_crit), 3)
)
```

**Key insight:** For small samples, $t$ critical values are substantially larger than $z$ critical values.

## Example: Porsche Odometer Readings (Devore Example 8.8)

A sample of $n = 16$ used Porsche Boxsters has odometer readings (miles):

```{r}
odometer <- c(80000, 30100, 97500, 58551, 73787, 51800, 69267, 44530,
              42192, 104920, 41442, 27418, 43436, 77219, 5991, 14362)

n <- length(odometer)
xbar <- mean(odometer)
s <- sd(odometer)

cat("Sample size:", n, "\n")
cat("Sample mean:", round(xbar, 1), "miles\n")
cat("Sample SD:", round(s, 1), "miles\n")
```

## Example: Porsche (continued)

**95% Confidence Interval:**

```{r}
alpha <- 0.05
t_crit <- qt(1 - alpha/2, df = n - 1)

cat("t critical value (df=15):", round(t_crit, 3), "\n")

se <- s / sqrt(n)
margin <- t_crit * se

cat("Standard error:", round(se, 1), "\n")
cat("Margin of error:", round(margin, 1), "\n\n")

cat("95% CI: (", round(xbar - margin, 1), ",", round(xbar + margin, 1), ")")
```

## Checking the Normality Assumption

The t interval assumes normally distributed data. Always check!

```{r}
#| fig-height: 5
par(mfrow = c(1, 2))
hist(odometer, main = "Histogram", xlab = "Odometer (miles)", col = "lightblue")
qqnorm(odometer, main = "Normal Q-Q Plot", pch = 19)
qqline(odometer, col = "red", lwd = 2)
```

The Q-Q plot looks reasonably linear — normality assumption is plausible.

## Medical Example: Systolic Blood Pressure

A clinical study measures systolic blood pressure in $n = 25$ patients after a new treatment:

```{r}
# Simulated BP data (mmHg)
bp <- c(118, 124, 131, 127, 122, 119, 128, 135, 121, 126,
        130, 124, 129, 133, 120, 125, 127, 132, 123, 128,
        126, 131, 122, 129, 125)

n <- length(bp)
xbar <- mean(bp)
s <- sd(bp)

# 95% CI
t_crit <- qt(0.975, df = n - 1)
se <- s / sqrt(n)
ci <- c(xbar - t_crit * se, xbar + t_crit * se)

cat("Sample mean:", round(xbar, 1), "mmHg\n")
cat("Sample SD:", round(s, 1), "mmHg\n")
cat("95% CI: (", round(ci[1], 1), ",", round(ci[2], 1), ") mmHg")
```

## Using t.test() in R

```{r}
# The easy way to compute t confidence intervals
t.test(bp, conf.level = 0.95)
```

. . .

**Note:** R's `t.test()` also performs a hypothesis test (which we'll cover later).

# Confidence Intervals for Proportions {background-color="#e67e22"}

## The Setup

**Situation:** We observe $X$ successes in $n$ independent Bernoulli trials.

-   $X \sim \text{Binomial}(n, p)$
-   Point estimate: $\hat{p} = X/n$

. . .

**Examples:**

-   Proportion of patients responding to treatment
-   Proportion of adverse events
-   Disease prevalence in a population

## The Wald Interval (Traditional)

For large $n$, $\hat{P} = X/n$ is approximately normal:
$$\hat{P} \approx N\left(p, \frac{p(1-p)}{n}\right)$$

. . .

**The Wald (traditional) 95% CI:**
$$\hat{p} \pm z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

. . .

::: callout-warning
## Problems with the Wald Interval
-   Poor coverage probability for $p$ near 0 or 1
-   Can produce bounds outside [0, 1]
-   Not recommended for small samples
:::

## Coverage Problems with Wald

```{r}
#| echo: false
#| fig-height: 6
# Simulate coverage probability for Wald interval
set.seed(551)
n_sim <- 10000
n <- 100

coverage_wald <- function(p, n, n_sim = 10000) {
  mean(replicate(n_sim, {
    x <- rbinom(1, n, p)
    p_hat <- x/n
    se <- sqrt(p_hat * (1 - p_hat) / n)
    lower <- p_hat - 1.96 * se
    upper <- p_hat + 1.96 * se
    lower < p & p < upper
  }))
}

p_values <- seq(0.05, 0.95, by = 0.05)
coverage <- sapply(p_values, coverage_wald, n = 100)

tibble(p = p_values, coverage = coverage) |>
  ggplot(aes(x = p, y = coverage)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red", linewidth = 1) +
  labs(x = "True Proportion p", y = "Coverage Probability",
       title = "Coverage Probability of 95% Wald Interval (n = 100)",
       subtitle = "Dashed line shows nominal 95% level") +
  theme_minimal(base_size = 18) +
  ylim(0.85, 1)
```

The Wald interval often has coverage **below** the nominal level!

## The Score (Wilson) Interval

A better approach: solve for $p$ in the pivotal quantity directly.

::: callout-important
## Score Confidence Interval for a Proportion
$$\frac{\hat{p} + \frac{z_{\alpha/2}^2}{2n} \pm z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n} + \frac{z_{\alpha/2}^2}{4n^2}}}{1 + \frac{z_{\alpha/2}^2}{n}}$$
:::

. . .

**Advantages:**

-   Coverage probability much closer to nominal level
-   Handles extreme proportions well
-   Always produces bounds in [0, 1]

## The Agresti-Coull Interval

A simpler approximation that works well in practice:

::: callout-tip
## Agresti-Coull ("Add Two Successes and Two Failures")
Let $\tilde{n} = n + 4$ and $\tilde{p} = \frac{X + 2}{n + 4}$

Then use:
$$\tilde{p} \pm z_{\alpha/2}\sqrt{\frac{\tilde{p}(1-\tilde{p})}{\tilde{n}}}$$
:::

. . .

**Why it works:** Adding 2 successes and 2 failures "shrinks" extreme estimates toward 0.5.

## Comparing Intervals: Clinical Trial Example

In a Phase II trial, 23 of 40 patients respond to treatment.

```{r}
x <- 23; n <- 40; p_hat <- x/n
cat("Sample proportion:", p_hat, "\n\n")

# Wald interval
se_wald <- sqrt(p_hat * (1 - p_hat) / n)
wald_ci <- p_hat + c(-1, 1) * 1.96 * se_wald

# Agresti-Coull interval
n_tilde <- n + 4
p_tilde <- (x + 2) / n_tilde
se_ac <- sqrt(p_tilde * (1 - p_tilde) / n_tilde)
ac_ci <- p_tilde + c(-1, 1) * 1.96 * se_ac

cat("Wald 95% CI:          (", round(wald_ci[1], 3), ",", round(wald_ci[2], 3), ")\n")
cat("Agresti-Coull 95% CI: (", round(ac_ci[1], 3), ",", round(ac_ci[2], 3), ")")
```

## Using prop.test() for Score Interval

```{r}
# Score (Wilson) interval in R
prop.test(x = 23, n = 40, conf.level = 0.95, correct = FALSE)
```

## Medical Example: Adverse Event Rate

In a safety study, 8 out of 200 patients experience a serious adverse event.

```{r}
x <- 8; n <- 200

# Score interval
result <- prop.test(x, n, conf.level = 0.95, correct = FALSE)
cat("Point estimate:", x/n, "\n")
cat("95% Score CI:", round(result$conf.int, 4))
```

. . .

**Clinical interpretation:** We are 95% confident that the true rate of serious adverse events is between 1.8% and 7.7%.

## Sample Size for Proportions

To achieve a margin of error $b$ with confidence level $1 - \alpha$:

$$n = \frac{z_{\alpha/2}^2 \cdot p(1-p)}{b^2}$$

. . .

**Problem:** We don't know $p$ before the study!

. . .

**Conservative approach:** Use $p = 0.5$ (maximizes $p(1-p)$):
$$n = \frac{z_{\alpha/2}^2 \cdot 0.25}{b^2}$$

## Sample Size Example: Disease Prevalence

An epidemiologist wants to estimate disease prevalence with a margin of error ≤ 3% with 95% confidence.

```{r}
b <- 0.03  # margin of error
z <- qnorm(0.975)

# Conservative (p = 0.5)
n_conservative <- z^2 * 0.25 / b^2
cat("Conservative sample size:", ceiling(n_conservative), "\n")

# If we believe p ≈ 0.10
n_informed <- z^2 * 0.10 * 0.90 / b^2
cat("Informed sample size (p ≈ 0.10):", ceiling(n_informed))
```

## Summary: CI Methods Comparison

| Method | When to Use | Notes |
|--------|-------------|-------|
| Z-interval | Normal data, $\sigma$ known | Rarely applicable |
| t-interval | Normal data, $\sigma$ unknown | Check normality |
| Wald (proportion) | Large $n$, $p$ not extreme | Poor coverage |
| Score/Wilson | General | Best for proportions |
| Agresti-Coull | Quick calculation | Good approximation |

# Interpreting CIs in Practice {background-color="#16a085"}

## What CIs Tell Us

::: callout-tip
## Valid Interpretations
-   "We are 95% confident that $\mu$ lies between L and U"
-   "The interval (L, U) is a plausible range for $\mu$"
-   "In repeated sampling, 95% of such intervals would contain $\mu$"
:::

. . .

::: callout-warning
## Invalid Interpretations
-   "There is a 95% probability that $\mu$ is in (L, U)" ✗
-   "95% of the data falls between L and U" ✗
-   "The true mean varies between L and U" ✗
:::

## Common Mistakes in Medical Literature

```{r}
#| echo: false
tribble(
  ~Mistake, ~`Correct Interpretation`,
  "The probability that the true effect is in the CI is 95%", 
    "We are 95% confident the true effect is in the CI",
  "95% of patients have values in this range", 
    "We are 95% confident the mean is in this range",
  "If we repeat the study, results will be in the CI", 
    "95% of future CIs would contain the true mean"
) |>
  knitr::kable()
```

## Clinical Decision Making

**Example:** A drug reduces blood pressure by an estimated 12 mmHg with 95% CI (8, 16).

. . .

**Questions we can answer:**

-   Is the effect clinically meaningful? (If minimum important difference is 5 mmHg, yes!)
-   Is zero a plausible value? (No — CI excludes 0)
-   What's the worst-case scenario? (Lower bound suggests at least 8 mmHg effect)

## Relationship to Hypothesis Testing

::: callout-note
## Preview of Coming Material
A 95% CI contains all parameter values that would **not** be rejected by a two-sided test at $\alpha = 0.05$.

-   If CI excludes the null value → reject $H_0$
-   If CI includes the null value → fail to reject $H_0$
:::

. . .

**Example:** 95% CI for mean difference is (8, 16).

-   Test $H_0: \mu = 0$ → Reject (0 not in CI)
-   Test $H_0: \mu = 10$ → Fail to reject (10 in CI)

# Summary and Looking Ahead {background-color="#2c3e50"}

## Lesson 7 Summary

**Key Concepts:**

1.  **Confidence Intervals** provide a range of plausible values with quantified uncertainty

2.  **Pivotal Quantities** allow us to construct CIs without knowing unknown parameters

3.  **Z-intervals** ($\sigma$ known): $\bar{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$

4.  **t-intervals** ($\sigma$ unknown): $\bar{x} \pm t_{\alpha/2, n-1}\frac{s}{\sqrt{n}}$

5.  **Proportion intervals**: Score/Wilson preferred over Wald

## Key Formulas

| Parameter | CI Formula | Conditions |
|-----------|------------|------------|
| $\mu$ ($\sigma$ known) | $\bar{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$ | Normal data |
| $\mu$ ($\sigma$ unknown) | $\bar{x} \pm t_{\alpha/2, n-1}\frac{s}{\sqrt{n}}$ | Normal data |
| $p$ (Wald) | $\hat{p} \pm z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$ | Large $n$ |
| $p$ (Agresti-Coull) | $\tilde{p} \pm z_{\alpha/2}\sqrt{\frac{\tilde{p}(1-\tilde{p})}{\tilde{n}}}$ | General |

## Lesson 7 Practice Problems

1.  **(Devore 8.1)** Consider a normal population with $\sigma$ known. 
    a. What is the confidence level for the interval $\bar{x} \pm 2.81\sigma/\sqrt{n}$?
    b. What is the confidence level for the interval $\bar{x} \pm 1.44\sigma/\sqrt{n}$?

2.  **(Devore 8.9)** A CI for true average breaking strength is (1142.4, 1257.6). 
    a. What assumptions must hold for this to be valid?
    b. What sample size was used?

3.  **(Devore 8.17)** Compute a 95% CI for mean tensile strength based on $n = 20$, $\bar{x} = 13.71$, $s = 3.55$.

4.  **(Devore 8.43)** In a survey, 859 of 1044 adults support mandatory GMO labeling. Construct a 95% CI for the true proportion.

## Lesson 7 Practice Problems (continued)

5.  **(Medical Application)** A clinical trial tests a new cholesterol medication. In $n = 48$ patients, the mean LDL reduction is $\bar{x} = 35$ mg/dL with $s = 18$ mg/dL.
    a. Construct a 95% CI for the mean LDL reduction.
    b. Construct a 99% CI. How does it compare?
    c. The company claims a mean reduction of at least 30 mg/dL. Is this claim supported?

6.  **(Chihara 7.1)** Explain why the statement "There is a 95% probability that $\mu$ lies in the interval (29.5, 33.4)" is incorrect.

7.  **(Sample Size)** How many subjects are needed to estimate the proportion of patients responding to a treatment within $\pm 5\%$ with 95% confidence?

## Looking Ahead

**Next Session (Lesson 7 Part 2): Bootstrap Confidence Intervals**

-   What if normality doesn't hold?
-   Bootstrap percentile intervals
-   Bootstrap t intervals
-   When to use bootstrap vs. classical methods

. . .

**Coming Soon:**

-   Hypothesis testing (Lessons 9-12)
-   Two-sample inference (Lesson 13)
-   Paired data analysis (Lesson 14)

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 8.1-8.4
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 7.
-   Agresti, A. and Coull, B. (1998). Approximate is better than "exact" for interval estimation of binomial proportions. *The American Statistician*, 52:119-126.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/07_Confidence_Intervals/07_Confidence_Intervals.html lessons/07_Confidence_Intervals/07_Confidence_Intervals.pdf
```
