---
title: "Lesson 7: Introduction to Confidence Intervals"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w3d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 7 - Introduction to CIs
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 7: From Point Estimates to Intervals {background-color="#2c3e50"}

## Review: What We've Learned About Estimation

**Point Estimation (Lessons 1-6):**

-   Estimators: statistics that estimate parameters
-   Properties: bias, variance, MSE, efficiency
-   Construction: Method of Moments, Maximum Likelihood
-   Quality: MVUE, sufficient statistics, Cramér-Rao bound

. . .

**The Limitation of Point Estimates:**

A single number ($\hat{\theta} = 0.75$) tells us nothing about **precision** or **uncertainty** of the estimate.

## Today's Goals

::: incremental
1.  Understand why we need interval estimates
2.  Learn the correct interpretation of confidence intervals
3.  Derive confidence intervals using pivotal quantities
4.  Construct and interpret z-intervals (known $\sigma$)
5.  Determine sample sizes for desired precision
6.  Understand one-sided confidence bounds
:::

## Motivating Example: Clinical Trial

A new antihypertensive drug is tested in $n = 36$ patients. The mean reduction in systolic blood pressure is $\bar{x} = 12.3$ mmHg.

. . .

**Point estimate:** $\hat{\mu} = 12.3$ mmHg

. . .

**Questions we can't answer with just the point estimate:**

-   How close is 12.3 to the true mean reduction?
-   Is the true mean likely to be at least 10 mmHg?
-   How much could $\bar{X}$ vary in repeated trials?

. . .

::: callout-note
## The Solution
A **confidence interval** provides a range of plausible values for the parameter, along with a measure of confidence.
:::

# The Fundamental Idea {background-color="#27ae60"}

## From Point to Interval

```{r}
#| echo: false
#| fig-width: 14
#| fig-height: 6
tibble(x = seq(5, 20, by = 0.1)) |>
  mutate(y = dnorm(x, mean = 12.3, sd = 2.5)) |>
  ggplot(aes(x = x, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_vline(xintercept = 12.3, color = "red", linetype = "dashed", linewidth = 1.2) +
  geom_segment(aes(x = 7.4, xend = 17.2, y = 0.02, yend = 0.02), 
               color = "darkgreen", linewidth = 2,
               arrow = arrow(ends = "both", length = unit(0.2, "cm"))) +
  annotate("text", x = 12.3, y = 0.18, label = expression(bar(x) == 12.3), size = 7, color = "red") +
  annotate("text", x = 12.3, y = 0.05, label = "95% CI: (7.4, 17.2)", size = 6, color = "darkgreen") +
  labs(x = "Blood Pressure Reduction (mmHg)", y = "Density",
       title = "Point Estimate with Confidence Interval") +
  theme_minimal(base_size = 18)
```

**Key insight:** The interval captures our uncertainty about the true parameter value.

## What Is a Confidence Interval?

::: callout-important
## Definition: Confidence Interval
A **confidence interval (CI)** for a parameter $\theta$ is a random interval $(L, U)$ computed from sample data such that:
$$P(L < \theta < U) = 1 - \alpha$$
for a specified confidence level $1 - \alpha$.
:::

. . .

**Important:**

-   $L$ and $U$ are **random** (they depend on the sample)
-   $\theta$ is **fixed** (but unknown)
-   The probability statement is about the **random event** that the lower and upper bound cover $\theta$, not about any randomness in $\theta$ itself.
-   The probability statement only makes sense when thinking about defining the estimators $L$ and $U$ and their sampling distributions, it does not make sense when talking about one sample and the observed values for $L$ and $U$

## Correct Interpretation of Confidence Intervals

::: callout-warning
## Common Misconception
"There is a 95% probability that $\mu$ lies in the interval (7.4, 17.2)."

**This is WRONG!** After computing the interval, $\mu$ either is or is not in the interval. There's no probability involved.
:::

. . .

::: callout-tip
## Correct Interpretation
"If we repeated this procedure many times, 95% of the resulting intervals would contain the true $\mu$."

Or more simply/loosely: "We are 95% confident that the true mean lies between 7.4 and 17.2."
:::

## Common Mistakes to Avoid (Chihara Example 7.2)

Suppose a 95% CI for mean gas mileage is (29.5, 33.4) mpg. Evaluate:

. . .

1. ❌ "We are 95% confident that gas mileage for cars in this company is between 29.5 and 33.4 mpg."
   - **Wrong:** CI is for population *mean*, not individual values.

. . .

2. ❌ "95% of all samples will give an average mileage between 29.5 and 33.4 mpg."
   - **Wrong:** Each sample gives a *different* CI; 95% of *intervals* contain $\mu$.

. . .

3. ❌ "There is a 95% chance that the true mean is between 29.5 and 33.4 mpg."
   - **Wrong:** $\mu$ is fixed, not random. It's either in the interval or not.

## Visualizing Confidence Level: Simulation

```{r}
#| echo: true
#| fig-height: 5
# Simulate 100 confidence intervals
set.seed(551)
true_mu <- 10; sigma <- 4; n <- 25

ci_data <- tibble(sample_id = 1:100) |>
  rowwise() |>
  mutate(
    xbar = mean(rnorm(n, true_mu, sigma)),
    lower = xbar - 1.96 * sigma / sqrt(n),
    upper = xbar + 1.96 * sigma / sqrt(n),
    contains_mu = lower < true_mu & upper > true_mu
  )

# Count how many contain the true mean
sum(ci_data$contains_mu)
```

## Visualizing Confidence Level: The Plot

```{r}
#| echo: false
#| fig-height: 8
ci_data |>
  slice_head(n = 50) |>
  ggplot(aes(y = sample_id, color = contains_mu)) +
  geom_segment(aes(x = lower, xend = upper, yend = sample_id), linewidth = 0.8) +
  geom_point(aes(x = xbar), size = 2) +
  geom_vline(xintercept = true_mu, color = "red", linetype = "dashed", linewidth = 1.2) +
  scale_color_manual(values = c("FALSE" = "red", "TRUE" = "steelblue"),
                     labels = c("Misses μ", "Contains μ")) +
  labs(x = "Value", y = "Sample Number", 
       title = "50 Confidence Intervals (95% Level)",
       subtitle = paste("True μ =", true_mu, "(red dashed line)"),
       color = "") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "top")
```

## The R Code for CI Visualization

```{r}
#| eval: true
#| echo: true
# Simulate *5000* confidence intervals
set.seed(551)
true_mu <- 10; sigma <- 4; n <- 25

ci_data <- tibble(sample_id = 1:5000) |>
  rowwise() |>
  mutate(
    xbar = mean(rnorm(n, true_mu, sigma)),
    lower = xbar - 1.96 * sigma / sqrt(n),
    upper = xbar + 1.96 * sigma / sqrt(n),
    contains_mu = lower < true_mu & upper > true_mu
  )

# Proportion containing true mu (should be ~95%)
mean(ci_data$contains_mu)
```

This is how the confidence level works: about 95% of intervals capture $\mu$.

## Your Turn: CI Interpretation

**Exercise:** A researcher reports a 95% CI for mean hospital stay after surgery as (3.2, 4.8) days. Which statements are correct?

A. "95% of patients stay between 3.2 and 4.8 days"

B. "We are 95% confident the true mean stay is between 3.2 and 4.8 days"

C. "There is a 95% probability the mean is in this interval"

D. "If we repeated this study many times, about 95% of the CIs would contain the true mean"

. . .

**Answer:** B and D are correct interpretations.

# Pivotal Quantities {background-color="#e74c3c"}

## The Key to Constructing CIs

::: callout-important
## Definition: Pivotal Quantity
A **pivotal quantity** is a function $h(X_1, \ldots, X_n; \theta)$ of the data and parameter such that:

1.  It depends on both the sample and $\theta$
2.  Its **distribution does not depend on $\theta$** (or any other unknown parameters)
:::

. . .

**Why pivotal quantities matter:**

-   We can find critical values that don't depend on unknowns
-   We can "solve for $\theta$" to get confidence limits

## Example: The Standard Pivotal Quantity

Let $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with $\sigma$ known.

. . .

Consider:
$$Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}}$$

. . .

**Check the properties:**

1.  $Z$ depends on $\bar{X}$ (data) and $\mu$ (parameter) ✓
2.  $Z \sim N(0, 1)$ — the distribution doesn't depend on $\mu$ ✓

. . .

$Z$ is a **pivotal quantity**!

## From Pivotal Quantity to CI

Since $Z \sim N(0, 1)$, we know:
$$P(-z_{\alpha/2} < Z < z_{\alpha/2}) = 1 - \alpha$$

. . .

Substitute $Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}}$:
$$P\left(-z_{\alpha/2} < \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} < z_{\alpha/2}\right) = 1 - \alpha$$

. . .

**Solve for $\mu$:**
$$P\left(\bar{X} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}} < \mu < \bar{X} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}\right) = 1 - \alpha$$

. . .

This gives us the **confidence interval formula**!

. . .

What is $L$ and $U$ here?

## The General Pivotal Quantity Method

::: callout-tip
## Recipe for Constructing CIs
1. Find a pivotal quantity $h(X_1, \ldots, X_n; \theta)$
2. Find constants $a$ and $b$ such that $P(a < h < b) = 1 - \alpha$
3. Solve the inequalities for $\theta$ to get $(L, U)$
:::

. . .

**Example (Devore 8.1):** For normal data with known $\sigma$:

- Pivotal: $Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim N(0,1)$
- Constants: $a = -z_{\alpha/2}$, $b = z_{\alpha/2}$
- Solve: $\bar{X} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}} < \mu < \bar{X} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$

# CI for Mean: $\sigma$ Known {background-color="#3498db"}

## The Z-Interval

::: callout-important
## Z Confidence Interval for Normal Mean ($\sigma$ Known)
If $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$ with $\sigma$ known, a $100(1-\alpha)\%$ CI for $\mu$ is:
$$\bar{x} \pm z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$
where $z_{\alpha/2}$ is the critical value with $P(Z > z_{\alpha/2}) = \alpha/2$.
:::

. . .

**Common critical values:**

| Confidence Level | $\alpha$ | $\alpha/2$ | $z_{\alpha/2}$ |
|------------------|----------|------------|----------------|
| 90% | 0.10 | 0.05 | 1.645 |
| 95% | 0.05 | 0.025 | 1.960 |
| 99% | 0.01 | 0.005 | 2.576 |

## Example: Tensile Strength (Devore Example 8.1)

Synthetic fibers are tested for tensile strength. From past data, $\sigma = 55$ MPa is known. A sample of $n = 31$ fibers has $\bar{x} = 1064$ MPa.

. . .

**95% Confidence Interval:**

```{r}
xbar <- 1064
sigma <- 55
n <- 31
alpha <- 0.05

z_crit <- qnorm(1 - alpha/2)
margin <- z_crit * sigma / sqrt(n)

c(xbar - margin, xbar + margin)
```

. . .

**Interpretation:** We are 95% confident that the true mean tensile strength is between 1044.6 and 1083.4 MPa.

## Medical Example: Blood Glucose Levels

A hospital monitors fasting blood glucose in diabetic patients. Historical data shows $\sigma = 18$ mg/dL. A sample of $n = 40$ patients has $\bar{x} = 142$ mg/dL.

**Construct a 99% CI for the population mean glucose level:**

```{r}
xbar <- 142
sigma <- 18
n <- 40
alpha <- 0.01

z_crit <- qnorm(1 - alpha/2)
cat("z critical value:", round(z_crit, 3), "\n")

margin <- z_crit * sigma / sqrt(n)
cat("Margin of error:", round(margin, 2), "mg/dL\n")

cat("99% CI: (", round(xbar - margin, 1), ",", round(xbar + margin, 1), ")")
```

. . .

**Clinical interpretation:** We can be 99% confident that the true mean fasting glucose for this patient population is between 134.7 and 149.3 mg/dL.

## Anatomy of a Confidence Interval

$$\bar{x} \pm z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$

```{r}
#| echo: false
#| fig-height: 4
tibble(x = c(1, 2, 3)) |>
  ggplot() +
  annotate("text", x = 1.5, y = 0.8, label = "Point Estimate", size = 15, hjust = 0.5) +
  annotate("text", x = 1.5, y = 0.6, label = expression(bar(x)), size = 20, hjust = 0.5, color = "red") +
  annotate("text", x = 2.5, y = 0.8, label = "Margin of Error", size = 15, hjust = 0.5) +
  annotate("text", x = 2.5, y = 0.55, label = expression(z[alpha/2] %.% frac(sigma, sqrt(n))), 
           size = 15, hjust = 0.5, color = "steelblue") +
  xlim(1, 3) + ylim(0, 1) +
  theme_void()
```

**Three factors affect the margin of error:**

1. **Confidence level** ($z_{\alpha/2}$): Higher confidence → wider interval → less precise
2. **Population variability** ($\sigma$): More variation → wider interval → less precise
3. **Sample size** ($n$): Larger sample → narrower interval → more precise

## Interactive: Effect of Confidence Level

```{r}
#| echo: true
# Same data, different confidence levels
xbar <- 1064; sigma <- 55; n <- 31

confidence_levels <- c(0.90, 0.95, 0.99)

tibble(conf_level = confidence_levels) |>
  mutate(
    z_crit = qnorm(1 - (1 - conf_level)/2),
    margin = z_crit * sigma / sqrt(n),
    lower = xbar - margin,
    upper = xbar + margin,
    width = upper - lower
  )
```

**Takeaway:** Higher confidence requires wider intervals!

## Visualizing the Trade-off

```{r}
#| echo: false
#| fig-height: 6
tibble(conf = seq(0.80, 0.99, by = 0.01)) |>
  mutate(
    z = qnorm(1 - (1 - conf)/2),
    width = 2 * z * 55 / sqrt(31)
  ) |>
  ggplot(aes(x = conf, y = width)) +
  geom_line(linewidth = 1.5, color = "steelblue") +
  geom_point(data = tibble(conf = c(0.90, 0.95, 0.99), 
                           width = 2 * qnorm(c(0.95, 0.975, 0.995)) * 55 / sqrt(31)),
             size = 4, color = "red") +
  labs(x = "Confidence Level", y = "CI Width (MPa)",
       title = "Trade-off: Confidence Level vs. Precision") +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal(base_size = 18)
```

## Sample Size Determination

**Question:** How large should $n$ be to achieve a desired precision?

. . .

If we want the margin of error to be at most $b$:
$$z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}} \leq b$$

. . .

**Solving for $n$:**
$$n \geq \left(\frac{z_{\alpha/2} \cdot \sigma}{b}\right)^2$$

## Sample Size Example: Clinical Trial Design

A pharmaceutical company wants to estimate mean drug clearance time. From pilot data, $\sigma \approx 25$ minutes. They want a 95% CI with margin of error ≤ 5 minutes.

```{r}
sigma <- 25
b <- 5  # desired margin of error
alpha <- 0.05
z_crit <- qnorm(1 - alpha/2)

n_required <- (z_crit * sigma / b)^2
cat("Minimum sample size:", ceiling(n_required))
```

. . .

**Note:** Always round **up** to ensure the desired precision!

## Sample Size: Effect of Requirements

```{r}
#| echo: true
# How does required n change with margin of error and confidence?
expand_grid(
  margin = c(3, 5, 10),
  conf_level = c(0.90, 0.95, 0.99)
) |>
  mutate(
    z = qnorm(1 - (1 - conf_level)/2),
    n_required = ceiling((z * 25 / margin)^2)
  ) |>
  pivot_wider(names_from = margin, values_from = n_required, 
              names_prefix = "margin_") |>
  select(-z)
```

**Key insight:** Halving the margin of error (from 10 to 5) quadruples the sample size!

## One-Sided Confidence Bounds

Sometimes we only care about one direction:

-   **Lower bound:** We want to know the minimum plausible value
-   **Upper bound:** We want to know the maximum plausible value

. . .

::: callout-tip
## One-Sided Bounds
**Lower Confidence Bound:** $\bar{x} - z_{\alpha} \cdot \frac{\sigma}{\sqrt{n}} < \mu$

**Upper Confidence Bound:** $\mu < \bar{x} + z_{\alpha} \cdot \frac{\sigma}{\sqrt{n}}$

Note: Use $z_\alpha$ (not $z_{\alpha/2}$) for one-sided bounds at confidence level $1 - \alpha$.
:::

. . .

## Why $z_\alpha$ instead of $z_{\alpha/2}$?

For a **two-sided** CI, we split $\alpha$ into two tails ($\alpha/2$ each):
$$P\left(-z_{\alpha/2} < \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} < z_{\alpha/2}\right) = 1 - \alpha$$

. . .

For a **one-sided lower bound**, all $\alpha$ goes in one tail:
$$P\left(\frac{\bar{X} - \mu}{\sigma/\sqrt{n}} < z_{\alpha}\right) = 1 - \alpha$$

Rearranging for $\mu$:
$$P\left(\bar{X} - z_{\alpha} \cdot \frac{\sigma}{\sqrt{n}} < \mu\right) = 1 - \alpha$$

. . .

So the lower bound is: $L = \bar{X} - z_{\alpha} \cdot \frac{\sigma}{\sqrt{n}}$

**Key insight:** One-sided bounds use $z_\alpha$ because we put all the "error probability" $\alpha$ on one side, not split between two sides.

## Medical Example: Drug Half-Life

A pharmacologist measures half-life of a new drug. With $n = 20$ subjects, $\bar{x} = 4.2$ hours, and known $\sigma = 0.8$ hours.

**Find a 95% lower confidence bound:**

```{r}
xbar <- 4.2
sigma <- 0.8
n <- 20
alpha <- 0.05

z_crit <- qnorm(1 - alpha)  # Note: z_alpha, not z_alpha/2
lower_bound <- xbar - z_crit * sigma / sqrt(n)

cat("95% lower confidence bound:", round(lower_bound, 2), "hours\n")
cat("Interpretation: We are 95% confident that the true half-life",
    "\nis at least", round(lower_bound, 2), "hours")
```

## Why Lower vs. Upper Bounds?

**Use a LOWER bound when:**

- Testing if a treatment effect is "at least" some value
- Establishing minimum efficacy
- Setting safety margins

. . .

**Use an UPPER bound when:**

- Estimating maximum risk or adverse event rate  
- Establishing worst-case scenarios
- Quality control (maximum defect rate)

## Example: Remission Time (Devore 8.7)

A surgeon wants a lower bound for mean remission time after cancer surgery. From $n = 25$ patients, $\bar{x} = 42$ months, $\sigma = 12$ months (from historical data).

```{r}
xbar <- 42
sigma <- 12
n <- 25
alpha <- 0.05

# 95% lower confidence bound
z_alpha <- qnorm(1 - alpha)
lower <- xbar - z_alpha * sigma / sqrt(n)

cat("95% lower confidence bound:", round(lower, 1), "months\n")
cat("\nWe are 95% confident that true mean remission time\n")
cat("is at least", round(lower, 1), "months.")
```

# Pivotal Quantities for Other Distributions {background-color="#9b59b6"}

## Example: Exponential Distribution (Devore 8.5)

Time-to-breakdown of insulating fluid follows $\text{Exp}(\lambda)$ with mean $1/\lambda$ and variance $1/\lambda^2$. Given $n = 10$ observations with $\sum x_i = 550.87$ min.

. . .

**The pivotal quantity:**
$$h = 2\lambda \sum_{i=1}^n X_i \sim \chi^2_{2n}$$

. . .

**Why is this pivotal?**

- Depends on data ($\sum X_i$) and parameter ($\lambda$)
- Distribution ($\chi^2_{2n}$) doesn't depend on $\lambda$

## Exponential CI Derivation

Let $\chi^2_{.025, 20}$ and $\chi^2_{.975, 20}$ be the critical values. Then:

$$P\left(\chi^2_{.975, 20} < 2\lambda \sum X_i < \chi^2_{.025, 20}\right) = 0.95$$

. . .

**Solving for $\lambda$:**
$$P\left(\frac{\chi^2_{.975, 20}}{2\sum X_i} < \lambda < \frac{\chi^2_{.025, 20}}{2\sum X_i}\right) = 0.95$$

```{r}
sum_x <- 550.87
lower <- qchisq(0.025, df = 20) / (2 * sum_x)
upper <- qchisq(0.975, df = 20) / (2 * sum_x)
cat("95% CI for λ: (", round(lower, 4), ",", round(upper, 4), ")")
```

## CI for Mean of Exponential

Since the mean of $\text{Exp}(\lambda)$ is $\mu = 1/\lambda$:

```{r}
# CI for mean breakdown time
mean_lower <- 1/upper
mean_upper <- 1/lower

cat("95% CI for mean breakdown time:\n")
cat("(", round(mean_lower, 1), ",", round(mean_upper, 1), ") minutes")
```

. . .

**Note:** The interval is quite wide — reflecting variability in the data and small sample size.

# Summary and Looking Ahead {background-color="#2c3e50"}

## Lesson 7 Summary

**Key Concepts:**

1.  **Confidence Intervals** provide a range of plausible values with quantified uncertainty

2.  **Correct Interpretation:** "95% of intervals constructed this way contain $\theta$"

3.  **Pivotal Quantities** allow CI construction without knowing unknown parameters

4.  **Z-intervals** ($\sigma$ known): $\bar{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$

5.  **Sample Size:** $n \geq \left(\frac{z_{\alpha/2} \cdot \sigma}{b}\right)^2$

6.  **One-sided bounds** use $z_\alpha$ instead of $z_{\alpha/2}$

## Key Formulas for Z-Intervals

| Quantity | Formula |
|----------|---------|
| Two-sided CI | $\bar{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$ |
| Lower bound | $\bar{x} - z_{\alpha}\frac{\sigma}{\sqrt{n}}$ |
| Upper bound | $\bar{x} + z_{\alpha}\frac{\sigma}{\sqrt{n}}$ |
| Sample size | $n = \left(\frac{z_{\alpha/2} \cdot \sigma}{b}\right)^2$ |
| Width | $w = 2 z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$ |

## Looking Ahead: Lesson 8

**Next Lesson: One-Sample t-Intervals and Proportions**

-   What if $\sigma$ is unknown? (The t-interval)
-   The Student's t distribution
-   Confidence intervals for population proportions
-   Wald vs. Score intervals
-   Checking assumptions

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 8.1
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 7.1-7.2
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/07_Confidence_Intervals/07_Confidence_Intervals.html lessons/07_Confidence_Intervals/07_Confidence_Intervals.pdf
```
