---
title: "Lesson 5: Sufficient Statistics"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w2d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 5
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 5: Sufficient Statistics

## Review: Where We Left Off

**Key concepts from Lessons 1-4:**

-   **Point estimation:** $\hat{\theta}$ estimates unknown parameter $\theta$
-   **Bias and MSE:** $\text{Bias}(\hat{\theta}) = E(\hat{\theta}) - \theta$, $\text{MSE} = \text{Var} + \text{Bias}^2$
-   **MVUE:** Unbiased estimator with minimum variance
-   **Fisher Information:** $I(\theta) = E[-\ell''(\theta)]$
-   **Cramér-Rao Lower Bound:** $\text{Var}(T) \geq 1/I_n(\theta)$
-   **MLE and MME:** Systematic methods for constructing estimators

. . .

**Today's Goals:**

-   Understand sufficient statistics: data reduction without information loss
-   Learn the Neyman Factorization Theorem
-   Connect sufficiency to finding MVUEs via Rao-Blackwell
-   Define completeness and its role in optimal estimation

## Motivating Question: Data Reduction

Suppose we observe $X_1, \ldots, X_n$ from a distribution with unknown $\theta$.

**Question:** Can we summarize the data with a simpler statistic $T(X_1, \ldots, X_n)$ without losing information about $\theta$?

. . .

::: callout-note
## The Key Insight
If a statistic captures **all** the information about $\theta$ contained in the data, we call it **sufficient** for $\theta$.
:::

. . .

**Why this matters:**

-   Simplifies analysis and computation
-   Guides us to optimal estimators
-   Fundamental for understanding MVUEs

# The Concept of Sufficiency {background-color="#3498db"}

## Intuition: What Does "Sufficient" Mean?

Imagine you're a consulting statistician. A colleague ran an experiment and wants your help estimating $\theta$.

**Scenario 1:** They give you all raw data: $x_1 = 1, x_2 = 0, x_3 = 3$

**Scenario 2:** They only tell you the total: $T = x_1 + x_2 + x_3 = 4$

. . .

**Question:** Does Scenario 1 give you more information about $\theta$ than Scenario 2?

. . .

If the answer is **no**—if knowing $T$ tells you everything about $\theta$—then $T$ is **sufficient** for $\theta$.

## Example: Poisson Defects (Devore 7.25)

**Setup:** The number of major defects on new vehicles follows $\text{Poisson}(\mu)$. We observe $n = 3$ vehicles with defects $X_1 = 1, X_2 = 0, X_3 = 3$.

**Key Question:** Would knowing only $T = \sum X_i = 4$ provide the same information about $\mu$ as knowing all individual $X_i$'s?

. . .

**Consider:** Given that $T = 4$, what's the conditional distribution of $(X_1, X_2, X_3)$?

For the specific triple $(2, 1, 1)$:

$$P(X_1 = 2, X_2 = 1, X_3 = 1 | T = 4) = \frac{P(X_1 = 2, X_2 = 1, X_3 = 1)}{P(T = 4)}$$

## Example: Poisson (continued)

$$P(X_1 = 2, X_2 = 1, X_3 = 1 | T = 4) = \frac{\frac{e^{-\mu}\mu^2}{2!} \cdot \frac{e^{-\mu}\mu^1}{1!} \cdot \frac{e^{-\mu}\mu^1}{1!}}{\frac{e^{-3\mu}(3\mu)^4}{4!}}$$

. . .

$$= \frac{e^{-3\mu}\mu^4 / 2}{e^{-3\mu}(3\mu)^4/24} = \frac{24}{2 \cdot 81} = \frac{4}{27}$$

. . .

::: callout-important
## Critical Observation
The $\mu$ terms **cancel out**! The conditional probability doesn't depend on $\mu$.
:::

This means: once we know $T = 4$, the specific $(X_1, X_2, X_3)$ values provide **no additional information** about $\mu$.

## Formal Definition of Sufficiency

::: callout-important
## Definition (Devore 7.3)

A statistic $T = t(X_1, \ldots, X_n)$ is **sufficient for $\theta$** if the conditional distribution of $X_1, \ldots, X_n$ given $T = t$ does not depend on $\theta$, for every possible value $t$.
:::

. . .

**In other words:** Once you know the value of $T$, the distribution of the data no longer involves $\theta$—so no additional information about $\theta$ can be extracted.

. . .

::: callout-tip
## Intuition
$T$ is sufficient if it "captures" all the information in the data relevant to $\theta$. It's sufficient to know $T$; the raw data adds nothing more.
:::

## Why is Sufficiency Important?

**Practical Implications:**

1.  **Data Reduction:** Sufficient statistics compress data without information loss

2.  **Estimator Selection:** Any good estimator should depend on the data only through sufficient statistics

3.  **Path to MVUE:** Combined with other properties, leads to minimum variance estimators

. . .

**Historical Note:** The concept was developed by R.A. Fisher in the 1920s as part of the foundations of modern statistical inference.

# The Neyman Factorization Theorem {background-color="#3498db"}

## Finding Sufficient Statistics: The Challenge

The definition of sufficiency requires computing conditional distributions—which can be quite difficult!

. . .

**We need a shortcut...**

. . .

Fortunately, the **Neyman Factorization Theorem** provides an elegant criterion based on the likelihood function.

## The Neyman Factorization Theorem

::: callout-important
## Theorem (Neyman-Fisher Factorization)

Let $f(x_1, \ldots, x_n; \theta)$ denote the joint pmf or pdf of $X_1, \ldots, X_n$. Then $T = t(X_1, \ldots, X_n)$ is a **sufficient statistic for $\theta$** if and only if there exist functions $g$ and $h$ such that:

$$f(x_1, \ldots, x_n; \theta) = g(t(x_1, \ldots, x_n); \theta) \cdot h(x_1, \ldots, x_n)$$

where:

-   $g$ depends on the data **only through** $t(x_1, \ldots, x_n)$ and involves $\theta$
-   $h$ depends on the data but **not on $\theta$**
:::

## Understanding the Factorization

$$f(x_1, \ldots, x_n; \theta) = \underbrace{g(t(x_1, \ldots, x_n); \theta)}_{\text{Involves } \theta \text{ and data through } T} \cdot \underbrace{h(x_1, \ldots, x_n)}_{\text{Depends on data, not } \theta}$$

. . .

**Key insight:** All information about $\theta$ is in the $g$ factor, which depends on $x_i$'s only through $T$.

. . .

::: callout-tip
## Strategy for Finding Sufficient Statistics

1.  Write down the joint pmf/pdf
2.  Identify the part that involves $\theta$
3.  Factor it so $\theta$ appears only with a function of the data
4.  That function is your sufficient statistic!
:::

## Example 1: Bernoulli Distribution

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Bernoulli}(p)$

**Joint pmf:**
$$f(x_1, \ldots, x_n; p) = \prod_{i=1}^n p^{x_i}(1-p)^{1-x_i} = p^{\sum x_i}(1-p)^{n - \sum x_i}$$

. . .

**Factorization:**
$$= \underbrace{p^t(1-p)^{n-t}}_{g(t; p)} \cdot \underbrace{1}_{h(x_1, \ldots, x_n)}$$

where $t = \sum_{i=1}^n x_i$.

. . .

::: callout-important
## Result
$T = \sum_{i=1}^n X_i$ (equivalently, $\bar{X}$) is **sufficient for $p$**.
:::

## Example 2: Poisson Distribution

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Poisson}(\mu)$

**Joint pmf:**
$$f(x_1, \ldots, x_n; \mu) = \prod_{i=1}^n \frac{e^{-\mu}\mu^{x_i}}{x_i!} = \frac{e^{-n\mu}\mu^{\sum x_i}}{\prod_{i=1}^n x_i!}$$

. . .

**Factorization:**
$$= \underbrace{e^{-n\mu}\mu^t}_{g(t; \mu)} \cdot \underbrace{\frac{1}{\prod_{i=1}^n x_i!}}_{h(x_1, \ldots, x_n)}$$

where $t = \sum_{i=1}^n x_i$.

. . .

::: callout-important
## Result
$T = \sum_{i=1}^n X_i$ is **sufficient for $\mu$**.
:::

## Example 3: Exponential Distribution

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Exponential}(\lambda)$ with pdf $f(x; \lambda) = \lambda e^{-\lambda x}$

**Joint pdf:**
$$f(x_1, \ldots, x_n; \lambda) = \prod_{i=1}^n \lambda e^{-\lambda x_i} = \lambda^n e^{-\lambda \sum x_i}$$

. . .

**Factorization:**
$$= \underbrace{\lambda^n e^{-\lambda t}}_{g(t; \lambda)} \cdot \underbrace{1}_{h(x_1, \ldots, x_n)}$$

where $t = \sum_{i=1}^n x_i$.

. . .

::: callout-important
## Result
$T = \sum_{i=1}^n X_i$ (equivalently, $\bar{X}$) is **sufficient for $\lambda$**.
:::

## Medical Application: Time to Disease Relapse

Cancer relapse times are often modeled with exponential distributions.

```{r}
# Simulated relapse times (months) for 15 patients
set.seed(123)
true_lambda <- 0.1  # Rate parameter (relapses per month)
relapse_times <- rexp(15, rate = true_lambda)

# Sufficient statistic
T_sum <- sum(relapse_times)
T_mean <- mean(relapse_times)

cat("Individual relapse times:\n")
cat(round(relapse_times, 2), "\n\n")
cat("Sufficient statistic T = Σx_i:", round(T_sum, 2), "months\n")
cat("Equivalently, x̄ =", round(T_mean, 2), "months\n")
cat("MLE of λ = 1/x̄ =", round(1/T_mean, 4), "per month")
```

## Example 4: Uniform Distribution (Devore 7.27)

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Uniform}[0, \theta]$

**Joint pdf:**
$$f(x_1, \ldots, x_n; \theta) = \prod_{i=1}^n \frac{1}{\theta} \cdot I(0 \leq x_i \leq \theta) = \frac{1}{\theta^n} \cdot I(\max(x_i) \leq \theta) \cdot I(\min(x_i) \geq 0)$$

. . .

**Factorization:** Using indicator notation:
$$= \underbrace{\frac{1}{\theta^n} I(\max(x_i) \leq \theta)}_{g(t; \theta)} \cdot \underbrace{I(\min(x_i) \geq 0)}_{h(x_1, \ldots, x_n)}$$

where $t = \max(x_1, \ldots, x_n)$.

. . .

::: callout-important
## Result
$T = \max(X_1, \ldots, X_n)$ is **sufficient for $\theta$**. Note: $\bar{X}$ is NOT sufficient here!
:::

## Uniform Example: Visualization

```{r}
#| echo: false
#| fig-height: 5.5
set.seed(456)
theta_true <- 10
n <- 20
uniform_data <- runif(n, 0, theta_true)

tibble(x = uniform_data) |> 
  ggplot(aes(x = x)) +
  geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = max(uniform_data), color = "red", linewidth = 1.5, linetype = "dashed") +
  geom_vline(xintercept = theta_true, color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = max(uniform_data) - 0.5, y = 4, 
           label = paste0("T = max = ", round(max(uniform_data), 2)), 
           color = "red", hjust = 1, size = 5) +
  annotate("text", x = theta_true + 0.3, y = 4, 
           label = paste0("θ = ", theta_true), 
           color = "darkgreen", hjust = 0, size = 5) +
  labs(title = "Uniform[0, θ]: The Maximum is Sufficient",
       subtitle = "All information about θ is captured by max(X)",
       x = "Value", y = "Count") +
  theme(plot.title = element_text(size = 18))
```

Only the maximum observation matters for estimating $\theta$!

## Why Isn't $\bar{X}$ Sufficient for Uniform?

Consider two datasets with the same mean but different maxima:

```{r}
# Two samples with same mean but different maxima
data_A <- c(2, 4, 6, 8)  # mean = 5, max = 8
data_B <- c(1, 4, 5, 10) # mean = 5, max = 10

cat("Dataset A: mean =", mean(data_A), ", max =", max(data_A), "\n")
cat("Dataset B: mean =", mean(data_B), ", max =", max(data_B), "\n")
```

. . .

These have the same $\bar{X} = 5$, but:

-   Dataset A tells us $\theta \geq 8$
-   Dataset B tells us $\theta \geq 10$

. . .

**Knowing only $\bar{X}$ loses information about $\theta$!**

The maximum captures information that the mean cannot.

## Your Turn: Finding Sufficient Statistics

**Exercise:** Let $X_1, \ldots, X_n \sim \text{Geometric}(p)$ with pmf $f(x; p) = (1-p)^{x-1}p$ for $x = 1, 2, 3, \ldots$

Find a sufficient statistic for $p$.

. . .

**Solution:**

$$f(x_1, \ldots, x_n; p) = \prod_{i=1}^n (1-p)^{x_i - 1}p = p^n (1-p)^{\sum x_i - n}$$

$$= \underbrace{p^n (1-p)^{t - n}}_{g(t; p)} \cdot \underbrace{1}_{h}$$

where $t = \sum x_i$.

. . .

**Result:** $T = \sum_{i=1}^n X_i$ is sufficient for $p$.

# Jointly Sufficient Statistics {background-color="#3498db"}

## When One Statistic Isn't Enough

When there are **multiple parameters**, we typically need **multiple sufficient statistics**.

. . .

::: callout-important
## Definition: Jointly Sufficient Statistics

Statistics $T_1, \ldots, T_k$ are **jointly sufficient** for parameters $\theta_1, \ldots, \theta_m$ if the conditional distribution of $X_1, \ldots, X_n$ given $T_1 = t_1, \ldots, T_k = t_k$ does not depend on any of the $\theta_j$'s.
:::

. . .

**Factorization Theorem (Extended):**
$$f(x_1, \ldots, x_n; \theta_1, \ldots, \theta_m) = g(t_1, \ldots, t_k; \theta_1, \ldots, \theta_m) \cdot h(x_1, \ldots, x_n)$$

## Example: Normal Distribution (Both Parameters Unknown)

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with both $\mu$ and $\sigma^2$ unknown.

**Joint pdf:**
$$f(x_1, \ldots, x_n; \mu, \sigma^2) = \left(\frac{1}{\sqrt{2\pi\sigma^2}}\right)^n \exp\left(-\frac{\sum(x_i - \mu)^2}{2\sigma^2}\right)$$

. . .

Expanding $\sum(x_i - \mu)^2 = \sum x_i^2 - 2\mu \sum x_i + n\mu^2$:

$$= \frac{1}{(2\pi)^{n/2}\sigma^n} \exp\left(-\frac{\sum x_i^2 - 2\mu \sum x_i + n\mu^2}{2\sigma^2}\right)$$

## Normal Distribution (continued)

$$f = \underbrace{\frac{1}{\sigma^n} \exp\left(-\frac{\sum x_i^2 - 2\mu \sum x_i + n\mu^2}{2\sigma^2}\right)}_{g(t_1, t_2; \mu, \sigma^2)} \cdot \underbrace{\left(\frac{1}{2\pi}\right)^{n/2}}_{h}$$

. . .

::: callout-important
## Result
$T_1 = \sum X_i$ (or $\bar{X}$) and $T_2 = \sum X_i^2$ are **jointly sufficient** for $(\mu, \sigma^2)$.

Equivalently, $(\bar{X}, S^2)$ are jointly sufficient.
:::

. . .

**Intuition:** For normal data, the sample mean and sample variance capture all information about $\mu$ and $\sigma^2$.

## Medical Application: Blood Pressure Measurements

```{r}
# Systolic blood pressure measurements (mmHg) from 30 hypertension patients
set.seed(789)
bp_data <- rnorm(30, mean = 145, sd = 15)

# Jointly sufficient statistics for (μ, σ²)
T1 <- sum(bp_data)
T2 <- sum(bp_data^2)

# Equivalent sufficient statistics
x_bar <- mean(bp_data)
s_squared <- var(bp_data)

cat("Jointly sufficient statistics:\n")
cat("T₁ = Σxᵢ =", round(T1, 1), "\n")
cat("T₂ = Σxᵢ² =", round(T2, 1), "\n\n")
cat("Equivalent form:\n")
cat("x̄ =", round(x_bar, 2), "mmHg\n")
cat("S² =", round(s_squared, 2), "mmHg²")
```

## Summary: Sufficient Statistics for Common Distributions

| Distribution | Parameter(s) | Sufficient Statistic(s) |
|--------------|--------------|-------------------------|
| Bernoulli($p$) | $p$ | $\sum X_i$ |
| Poisson($\mu$) | $\mu$ | $\sum X_i$ |
| Exponential($\lambda$) | $\lambda$ | $\sum X_i$ |
| Normal($\mu$, known $\sigma^2$) | $\mu$ | $\bar{X}$ |
| Normal($\mu$, $\sigma^2$) | $(\mu, \sigma^2)$ | $(\bar{X}, S^2)$ or $(\sum X_i, \sum X_i^2)$ |
| Uniform$[0, \theta]$ | $\theta$ | $\max(X_i)$ |
| Gamma$(\alpha, \beta)$ | $(\alpha, \beta)$ | $(\sum X_i, \sum \ln X_i)$ |

# Minimal Sufficiency {background-color="#3498db"}

## The Problem: Non-Uniqueness of Sufficient Statistics

Sufficient statistics are **not unique**!

. . .

**Trivial example:** The entire dataset $(X_1, \ldots, X_n)$ is always sufficient.

**Less trivial:** If $T$ is sufficient, so is any one-to-one function of $T$ (e.g., if $\sum X_i$ is sufficient, so is $\bar{X}$).

. . .

**Question:** What is the "best" or "most efficient" sufficient statistic?

. . .

::: callout-note
## Idea
We want the sufficient statistic that achieves the **greatest data reduction** while still capturing all information about $\theta$.
:::

## Definition: Minimal Sufficiency

::: callout-important
## Definition (Devore 7.3)

A sufficient statistic $T$ is **minimal sufficient** if it is a function of every other sufficient statistic.

Equivalently: $T$ is minimal sufficient if, for any other sufficient statistic $T'$, we can write $T = f(T')$ for some function $f$.
:::

. . .

**Intuition:** A minimal sufficient statistic represents the maximum possible data reduction without losing information about $\theta$.

. . .

**Example:** For $N(\mu, \sigma^2)$ with both unknown, $(\bar{X}, S^2)$ are jointly minimal sufficient—you cannot reduce the data any further without losing information.

## Finding Minimal Sufficient Statistics

::: callout-tip
## Practical Criterion (Devore Exercise 7.74)

$T = t(X_1, \ldots, X_n)$ is minimal sufficient if for any two sample points $\mathbf{x} = (x_1, \ldots, x_n)$ and $\mathbf{y} = (y_1, \ldots, y_n)$:

$$\frac{f(\mathbf{x}; \theta)}{f(\mathbf{y}; \theta)} \text{ doesn't depend on } \theta \iff t(\mathbf{x}) = t(\mathbf{y})$$
:::

. . .

**In words:** The likelihood ratio is constant in $\theta$ if and only if the two samples have the same value of $T$.

## Example: Minimal Sufficiency for Exponential

For $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Exponential}(\lambda)$:

$$\frac{f(\mathbf{x}; \lambda)}{f(\mathbf{y}; \lambda)} = \frac{\lambda^n e^{-\lambda \sum x_i}}{\lambda^n e^{-\lambda \sum y_i}} = e^{-\lambda(\sum x_i - \sum y_i)}$$

. . .

This ratio doesn't depend on $\lambda$ $\iff$ $\sum x_i = \sum y_i$.

. . .

**Therefore:** $T = \sum X_i$ is **minimal sufficient** for $\lambda$.

## A Cautionary Example: Cauchy Distribution

**The Cauchy distribution** with location parameter $\theta$ has pdf:
$$f(x; \theta) = \frac{1}{\pi[1 + (x - \theta)^2]}$$

. . .

For a random sample from Cauchy:

-   The **ordered values** $(X_{(1)}, \ldots, X_{(n)})$ are jointly sufficient
-   Remarkably, the ordered values are also **minimal sufficient**!

. . .

::: callout-warning
## Surprising Result
For Cauchy, there is **no reduction** from the ordered sample to a simpler statistic. The full ordered data is needed to capture all information about $\theta$.
:::

This shows that the number of minimal sufficient statistics doesn't always equal the number of parameters.

# The Rao-Blackwell Theorem {background-color="#3498db"}

## Why Sufficiency Matters for Estimation

We've established that sufficient statistics capture all information about $\theta$.

**Key Question:** How do we use this to find better estimators?

. . .

**The Rao-Blackwell Theorem** provides the answer: it shows how to improve any estimator by conditioning on sufficient statistics.

## The Rao-Blackwell Theorem

::: callout-important
## Theorem (Rao-Blackwell)

Suppose:

-   $T$ is sufficient for $\theta$
-   $U$ is any unbiased estimator for $h(\theta)$

Then the estimator $U^* = E(U | T)$ satisfies:

1.  $U^*$ is unbiased for $h(\theta)$
2.  $\text{Var}(U^*) \leq \text{Var}(U)$

with strict inequality unless $U$ is already a function of $T$.
:::

. . .

**In words:** Conditioning an unbiased estimator on a sufficient statistic **never increases variance** and often **decreases it**.

## Proof Sketch: Why Rao-Blackwell Works

**Step 1: $U^*$ is an estimator (doesn't depend on $\theta$)**

Since $T$ is sufficient, the conditional distribution of $U | T$ doesn't involve $\theta$. Therefore $E(U | T)$ doesn't involve $\theta$. ✓

. . .

**Step 2: $U^*$ is unbiased**

By the Law of Total Expectation:
$$E(U^*) = E[E(U | T)] = E(U) = h(\theta) \checkmark$$

. . .

**Step 3: $U^*$ has smaller variance**

By the Law of Total Variance:
$$\text{Var}(U) = \text{Var}(E[U|T]) + E[\text{Var}(U|T)] = \text{Var}(U^*) + E[\text{Var}(U|T)]$$

Since $E[\text{Var}(U|T)] \geq 0$, we have $\text{Var}(U) \geq \text{Var}(U^*)$. ✓

## Example: Poisson Zero-Defect Probability (Devore 7.30)

**Problem:** Estimate $e^{-\mu} = P(\text{no defects})$ for $X_i \sim \text{Poisson}(\mu)$.

**Start with a crude unbiased estimator:**
$$U = I(X_1 = 0) = \begin{cases} 1 & \text{if } X_1 = 0 \\ 0 & \text{if } X_1 > 0 \end{cases}$$

. . .

Check: $E(U) = P(X_1 = 0) = e^{-\mu}$ ✓ (Unbiased!)

. . .

**Apply Rao-Blackwell:** The sufficient statistic is $T = \sum_{i=1}^n X_i$.

We need: $U^* = E(U | T) = P(X_1 = 0 | T = t)$

## Poisson Example (continued)

$$P(X_1 = 0 | T = t) = \frac{P(X_1 = 0 \cap T = t)}{P(T = t)} = \frac{P(X_1 = 0) \cdot P(\sum_{i=2}^n X_i = t)}{P(T = t)}$$

. . .

Using the fact that $\sum_{i=2}^n X_i \sim \text{Poisson}((n-1)\mu)$:

$$= \frac{e^{-\mu} \cdot \frac{e^{-(n-1)\mu}[(n-1)\mu]^t}{t!}}{\frac{e^{-n\mu}(n\mu)^t}{t!}} = \frac{(n-1)^t}{n^t} = \left(1 - \frac{1}{n}\right)^t$$

. . .

::: callout-important
## Improved Estimator
$$U^* = \left(1 - \frac{1}{n}\right)^T$$
is unbiased for $e^{-\mu}$ with variance $\leq$ the variance of the crude indicator!
:::

## Simulation: Comparing Estimators

```{r}
#| code-fold: true
set.seed(42)
mu_true <- 1.5
n <- 10
n_sims <- 5000

# Simulate Poisson samples and compute estimators
poisson_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) rpois(n, mu_true)),
    T_sum = map_dbl(data, sum),
    crude_U = map_dbl(data, \(d) as.numeric(d[1] == 0)),  # I(X_1 = 0)
    improved_U = (1 - 1/n)^T_sum,  # Rao-Blackwell improved
    mle = exp(-T_sum/n)  # MLE of e^(-μ)
  )

# True value
true_prob <- exp(-mu_true)

# Compare estimators
comparison <- poisson_sim |> 
  summarize(
    `E(Crude U)` = mean(crude_U),
    `E(Improved U*)` = mean(improved_U),
    `E(MLE)` = mean(mle),
    `Var(Crude U)` = var(crude_U),
    `Var(Improved U*)` = var(improved_U),
    `Var(MLE)` = var(mle)
  )

cat("True P(X = 0) = e^(-μ) =", round(true_prob, 4), "\n\n")
print(comparison |> pivot_longer(everything()) |> mutate(value = round(value, 4)))
```

## Visualization: Rao-Blackwell Improvement

```{r}
#| echo: false
#| fig-height: 5.5
poisson_sim |> 
  select(crude_U, improved_U) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = case_when(
    Estimator == "crude_U" ~ "Crude: I(X₁ = 0)",
    Estimator == "improved_U" ~ "Rao-Blackwell: (1-1/n)^T"
  )) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = true_prob, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Rao-Blackwell Theorem: Variance Reduction Through Sufficiency",
       subtitle = paste0("True e^(-μ) = ", round(true_prob, 3), " (red line); Improved estimator has smaller variance"),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("orange", "steelblue")) +
  theme(legend.position = "top")
```

## Key Implication: Best Estimators Use Sufficient Statistics

::: callout-important
## Corollary of Rao-Blackwell

Any estimator that is **not** based purely on sufficient statistics is necessarily **inferior** to some other estimator.
:::

. . .

**Example:** For Uniform$[0, \theta]$, $\max(X_i)$ is sufficient. Any estimator using $\bar{X}$ or $S$ alone is suboptimal:

-   $\hat{\theta}_1 = 2\bar{X}$ has variance $\theta^2/(3n)$
-   $\hat{\theta}_2 = \frac{n+1}{n}\max(X_i)$ has variance $\theta^2/[n(n+2)]$

Since $\bar{X}$ isn't a function of $\max(X_i)$, estimator $\hat{\theta}_1$ loses information!

## MLEs and Sufficient Statistics

::: callout-tip
## Theorem (MLEs Use Sufficient Statistics)

If $T_1, \ldots, T_k$ are jointly sufficient for $\theta_1, \ldots, \theta_m$, then the MLEs $\hat{\theta}_1, \ldots, \hat{\theta}_m$ can be expressed as functions of $T_1, \ldots, T_k$ alone.
:::

. . .

**Why?** The likelihood factors as $L(\theta) = g(T; \theta) \cdot h(\mathbf{x})$. Maximizing $L$ is equivalent to maximizing $g$, which depends on the data only through $T$.

. . .

**Practical implication:** MLEs automatically "use" sufficient statistics—this is one reason they have such good properties!

# Completeness {background-color="#3498db"}

## The Final Piece: Completeness

We've seen that:

-   Sufficient statistics capture all information about $\theta$
-   Rao-Blackwell lets us improve estimators by conditioning on sufficient statistics

. . .

**Question:** Given a sufficient statistic $T$, is there a **unique** best unbiased estimator based on $T$?

. . .

The answer involves **completeness**—a property that ensures uniqueness.

## Definition of Completeness

::: callout-important
## Definition: Complete Statistic

A statistic $T$ is **complete** for $\theta$ if for any function $g$:

$$E_\theta[g(T)] = 0 \text{ for all } \theta \implies P_\theta(g(T) = 0) = 1 \text{ for all } \theta$$

In other words, the only function of $T$ with expected value 0 for all $\theta$ is the zero function (almost surely).
:::

. . .

**Intuition:** Completeness means there are "no redundancies" in $T$—you can't construct a non-trivial unbiased estimator of zero using $T$.

## Why Completeness Matters

::: callout-important
## Lehmann-Scheffé Theorem

If $T$ is a **complete sufficient statistic** and $U = h(T)$ is an unbiased estimator of $\tau(\theta)$, then $U$ is the **unique MVUE** of $\tau(\theta)$.
:::

. . .

**Proof Sketch:** 

Suppose $U_1$ and $U_2$ are both unbiased for $\tau(\theta)$ and are functions of $T$.

Then $E[U_1 - U_2] = 0$ for all $\theta$.

By completeness, $U_1 - U_2 = 0$ almost surely.

Therefore $U_1 = U_2$—uniqueness! ✓

## Completeness in Exponential Families

::: callout-tip
## Theorem

For distributions in the **exponential family** (which includes Normal, Poisson, Exponential, Binomial, Gamma), the natural sufficient statistics are typically **complete**.
:::

. . .

**Examples of complete sufficient statistics:**

| Distribution | Complete Sufficient Statistic |
|--------------|-------------------------------|
| Bernoulli$(p)$ | $\sum X_i$ |
| Poisson$(\mu)$ | $\sum X_i$ |
| Exponential$(\lambda)$ | $\sum X_i$ |
| Normal$(\mu, \sigma^2)$ | $(\bar{X}, S^2)$ jointly |

## Example: Normal Mean (Known Variance)

For $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$ with $\sigma^2$ known:

-   $\bar{X}$ is sufficient for $\mu$
-   $\bar{X}$ is complete for $\mu$
-   $\bar{X}$ is unbiased for $\mu$

. . .

By **Lehmann-Scheffé**: $\bar{X}$ is the **unique MVUE** of $\mu$.

. . .

::: callout-tip
## The Power of Completeness + Sufficiency
Once we find a complete sufficient statistic, any unbiased function of it is automatically the MVUE!
:::

## The Road to MVUE: Summary

```{r}
#| echo: false
#| fig-height: 4
library(DiagrammeR)

grViz("
digraph MVUE_path {
  graph [rankdir = LR, bgcolor = transparent]
  node [shape = box, style = filled, fillcolor = lightblue]
  
  A [label = 'Find Sufficient\\nStatistic T']
  B [label = 'Check if T\\nis Complete']
  C [label = 'Find Unbiased\\nFunction h(T)']
  D [label = 'h(T) is the\\nUnique MVUE!', fillcolor = lightgreen]
  
  A -> B -> C -> D
}
")
```

**Alternative path:** Use Cramér-Rao bound to verify an estimator achieves minimum variance.

# Medical Research Applications {background-color="#3498db"}

## Application 1: Disease Incidence Rate

**Scenario:** Epidemiologists count disease cases in $n$ geographic regions.

Let $X_i \sim \text{Poisson}(\mu)$ = cases in region $i$.

```{r}
# Disease cases in 8 regions
disease_cases <- c(3, 5, 2, 7, 4, 6, 3, 5)
n <- length(disease_cases)

# Sufficient statistic
T_total <- sum(disease_cases)

# MVUE of μ
mu_hat <- mean(disease_cases)

# MVUE of P(no cases) = e^(-μ)
prob_zero_hat <- (1 - 1/n)^T_total

cat("Sufficient statistic T = Σxᵢ =", T_total, "\n")
cat("MVUE of μ:", mu_hat, "cases per region\n")
cat("Improved estimate of P(no cases):", round(prob_zero_hat, 4))
```

## Application 2: Drug Response Proportion

**Scenario:** In a clinical trial, we observe patient responses to treatment.

Let $X_i \sim \text{Bernoulli}(p)$ where $p$ = response rate.

```{r}
# 50 patients: 1 = responded, 0 = no response
set.seed(111)
responses <- rbinom(50, 1, prob = 0.7)

# Sufficient statistic
T_sum <- sum(responses)

# MVUE of p (sample proportion)
p_hat <- mean(responses)

# 95% CI for p (using normal approximation)
se <- sqrt(p_hat * (1 - p_hat) / length(responses))
ci <- p_hat + c(-1.96, 1.96) * se

cat("Sufficient statistic T = Σxᵢ =", T_sum, "responders\n")
cat("MVUE of p:", p_hat, "\n")
cat("95% CI: (", round(ci[1], 3), ",", round(ci[2], 3), ")")
```

## Application 3: Survival Time Analysis

**Scenario:** Patient survival times follow Exponential$(\lambda)$.

```{r}
# Survival times (months) for 20 patients
survival_times <- c(14.2, 8.5, 23.1, 11.7, 19.8, 6.3, 15.4, 12.9, 
                    25.6, 9.1, 18.3, 7.8, 21.4, 13.2, 16.7, 10.5,
                    22.8, 8.9, 17.6, 11.3)

n <- length(survival_times)
T_sum <- sum(survival_times)

# MLE/MVUE of λ
lambda_hat <- 1 / mean(survival_times)

# MVUE of median survival time (= log(2)/λ)
median_hat <- log(2) * mean(survival_times)

# MVUE of 1-year survival probability P(X > 12) = e^(-12λ)
surv_12 <- exp(-12 * lambda_hat)

cat("Sufficient statistic T =", round(T_sum, 1), "months\n")
cat("MVUE of λ:", round(lambda_hat, 4), "per month\n")
cat("Estimated median survival:", round(median_hat, 1), "months\n")
cat("Estimated P(survive > 12 months):", round(surv_12, 3))
```

# Summary and Looking Ahead {background-color="#2c3e50"}

## Lesson 5 Summary: Key Concepts

**1. Sufficiency**

-   A statistic $T$ is sufficient if it captures all information about $\theta$
-   Formal criterion: conditional distribution of data given $T$ doesn't depend on $\theta$

. . .

**2. Neyman Factorization Theorem**

-   $f(\mathbf{x}; \theta) = g(T; \theta) \cdot h(\mathbf{x})$
-   Practical tool for identifying sufficient statistics

. . .

**3. Rao-Blackwell Theorem**

-   Conditioning on sufficient statistics improves estimators
-   Any estimator not based on sufficient statistics is suboptimal

. . .

**4. Completeness**

-   Combined with sufficiency, guarantees unique MVUE (Lehmann-Scheffé)

## Sufficient Statistics: Quick Reference

| Distribution | Parameter(s) | Sufficient Statistic(s) | Complete? |
|--------------|--------------|-------------------------|-----------|
| Bernoulli$(p)$ | $p$ | $\sum X_i$ | Yes |
| Poisson$(\mu)$ | $\mu$ | $\sum X_i$ | Yes |
| Exponential$(\lambda)$ | $\lambda$ | $\sum X_i$ | Yes |
| Normal$(\mu, \sigma^2)$ | $\mu, \sigma^2$ | $(\bar{X}, S^2)$ | Yes |
| Uniform$[0, \theta]$ | $\theta$ | $\max(X_i)$ | Yes |
| Cauchy$(\theta)$ | $\theta$ | Order statistics | No reduction |

## Key Theorems Flowchart

**To find MVUE:**

1.  **Find sufficient statistic $T$** (Factorization Theorem)
2.  **Check completeness** (exponential family → typically complete)
3.  **Find unbiased function of $T$**
4.  **Result: MVUE** (by Lehmann-Scheffé)

. . .

**Alternative:**

1.  **Compute Fisher Information** $I(\theta)$
2.  **Find CRLB** = $1/[nI(\theta)]$
3.  **Find estimator achieving CRLB** → MVUE

# Homework Problems {background-color="#2c3e50"}

## Homework: Conceptual Questions

1.  **(Devore 7.38)** Let $X_i = 1$ if vehicle $i$ passes emissions test, 0 otherwise, with $P(X_i = 1) = p$. For $n = 3$ vehicles, let $T = X_1 + X_2 + X_3$. Use the definition of sufficiency to show $T$ is sufficient for $p$ by finding the conditional distribution of $(X_1, X_2, X_3)$ given $T = t$.

2.  **(Devore 7.39)** Components in batches of size $k$ have proportion $p$ satisfactory. For $n$ batches, statistician A knows all $X_i$'s (satisfactory count per batch), while statistician B knows only $T = \sum X_i$. Does A have more information about $p$ than B?

3.  **Conceptual:** Explain in your own words why the sample mean $\bar{X}$ is NOT sufficient for $\theta$ when sampling from Uniform$[0, \theta]$, even though it contains information about $\theta$.

## Homework: Finding Sufficient Statistics

4.  **(Devore 7.40)** Let $X_1, \ldots, X_n$ be lifetimes from Exponential$(\lambda)$. Use the factorization theorem to show $\sum X_i$ is sufficient for $\lambda$.

5.  **(Devore 7.41)** For a random sample from $f(x; \theta) = e^{-(x-\theta)}$ for $x \geq \theta$, find a sufficient statistic for $\theta$.

6.  **(Devore 7.43)** Show that for Normal$(\mu, \sigma^2)$ with both parameters unknown, $\sum X_i$ and $\sum X_i^2$ are jointly sufficient.

7.  **(Devore 7.70)** For $0 < \theta < 1$, consider a uniform distribution on $[\theta, 1/\theta]$. Identify a sufficient statistic for $\theta$.

## Homework: Rao-Blackwell Applications

8.  **(Devore 7.47)** For Bernoulli$(p)$ data, we want to estimate $p(1-p)$. Start with $U = I(X_1 = 1, X_2 = 0)$ (indicator that first succeeds, second fails). Find the improved estimator $U^* = E(U|T)$ where $T = \sum X_i$.

9.  **(Devore 7.48)** In Example 7.30, we started with $U = I(X_1 = 0)$ to estimate $e^{-\mu}$ for Poisson data. Now start with $U = \frac{1}{n}\sum I(X_i = 0)$. Show that the Rao-Blackwell improved estimator is identical to the one in the example.

10. **(Devore 7.49)** For Poisson$(\mu)$, both $\bar{X}$ and $S^2$ are unbiased for $\mu$. Consider $\hat{\mu} = (\bar{X} + S^2)/2$. Which of $\bar{X}$, $S^2$, or $\hat{\mu}$ is the best estimator? Justify using sufficiency.

## Homework: Challenging Problems

11. **(Devore 7.72)** The fraction of a bottle filled follows $f(x; \theta) = \theta x^{\theta-1}$ for $0 < x < 1$.
    a. Find the method of moments estimator for $\theta$.
    b. Is this estimator sufficient? If not, find a sufficient statistic and an estimator based on it.

12. **(Devore 7.71)** An investigator tests individuals until finding $r$ with an allergy, where $P(\text{allergy}) = p$. Let $Y$ = number tested. Use the definition of sufficiency to show $Y$ is sufficient for $p$.

13. **Medical Application:** In a vaccine trial, adverse events follow Poisson$(\mu)$ per 1000 vaccinations. With data from $n = 12$ groups of 1000 people:
    a. Identify the sufficient statistic for $\mu$
    b. Find the MVUE for $\mu$
    c. Find an improved unbiased estimator for $P(\text{no adverse events}) = e^{-\mu}$

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 7.3
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 6.
-   Casella, G. and Berger, R.L. *Statistical Inference* (Duxbury). Chapter 6 (for advanced treatment of completeness).
:::

## Questions? {.center background-color="#3498db"}

Thank you!

Next Lesson: Fisher Information and Large-Sample Properties

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/05_Sufficient_Statistics/05_Sufficient_Statistics.html lessons/05_Sufficient_Statistics/05_Sufficient_Statistics.pdf
```
