---
title: "Lesson 5: Sufficient Statistics"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w2d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 5
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 5: Sufficient Statistics

## Review: Where We Left Off

**Key concepts from Lessons 1-4:**

-   **Point estimation:** $\hat{\theta}$ estimates unknown parameter $\theta$
-   **Bias and MSE:** $\text{Bias}(\hat{\theta}) = E(\hat{\theta}) - \theta$, $\text{MSE} = \text{Var} + \text{Bias}^2$
-   **MVUE:** Unbiased estimator with minimum variance
-   **Fisher Information:** $I(\theta) = E[-\ell''(\theta)]$
-   **Cramér-Rao Lower Bound:** $\text{Var}(T) \geq 1/I_n(\theta)$
-   **MLE and MME:** Systematic methods for constructing estimators

. . .

**Today's Goals:**

-   Understand sufficient statistics: data reduction without information loss
-   Learn the Neyman Factorization Theorem
-   Connect sufficiency to finding MVUEs via Rao-Blackwell
-   Define completeness and its role in optimal estimation

## Motivating Question: Data Reduction

Suppose we observe $X_1, \ldots, X_n$ from a distribution with unknown $\theta$.

**Question:** Can we summarize the data with a simpler statistic $T(X_1, \ldots, X_n)$ without losing information about $\theta$?

. . .

::: callout-note
## The Key Insight
If a statistic captures **all** the information about $\theta$ contained in the data, we call it **sufficient** for $\theta$.
:::

. . .

**Why this matters:**

-   Simplifies analysis and computation
-   Guides us to optimal estimators
-   Fundamental for understanding MVUEs

# The Concept of Sufficiency {background-color="#3498db"}

## Intuition: What Does "Sufficient" Mean?

Imagine you're a consulting statistician. A colleague ran an experiment and wants your help estimating $\theta$.

**Scenario 1:** They give you all raw data: $x_1 = 1, x_2 = 0, x_3 = 3$

**Scenario 2:** They only tell you the total: $T = x_1 + x_2 + x_3 = 4$

. . .

**Question:** Does Scenario 1 give you more information about $\theta$ than Scenario 2?

. . .

If the answer is **no**—if knowing $T$ tells you everything about $\theta$—then $T$ is **sufficient** for $\theta$.

## Example: Poisson Defects (Devore 7.25)

**Setup:** The number of major defects on new vehicles follows $\text{Poisson}(\mu)$. We observe $n = 3$ vehicles with defects $X_1 = 1, X_2 = 0, X_3 = 3$.

**Key Question:** Would knowing only $T = \sum X_i = 4$ provide the same information about $\mu$ as knowing all individual $X_i$'s?

. . .

**Consider:** Given that $T = 4$, what's the conditional probability of $(X_1, X_2, X_3) = (2, 1, 1)$?

$$P(X_1 = 2, X_2 = 1, X_3 = 1 | T = 4) = \frac{\frac{e^{-\mu}\mu^2}{2!} \cdot \frac{e^{-\mu}\mu^1}{1!} \cdot \frac{e^{-\mu}\mu^1}{1!}}{\frac{e^{-3\mu}(3\mu)^4}{4!}} = \frac{4}{27}$$

. . .

::: callout-important
## Critical Observation
The $\mu$ terms **cancel out**! The conditional probability doesn't depend on $\mu$.
:::

## Formal Definition of Sufficiency

::: callout-important
## Definition (Devore 7.3)

A statistic $T = t(X_1, \ldots, X_n)$ is **sufficient for $\theta$** if the conditional distribution of $X_1, \ldots, X_n$ given $T = t$ does not depend on $\theta$, for every possible value $t$.
:::

. . .

**In other words:** Once you know the value of $T$, the distribution of the data no longer involves $\theta$—so no additional information about $\theta$ can be extracted.

. . .

::: callout-tip
## Intuition
$T$ is sufficient if it "captures" all the information in the data relevant to $\theta$. It's sufficient to know $T$; the raw data adds nothing more.
:::

# The Neyman Factorization Theorem {background-color="#3498db"}

## Finding Sufficient Statistics: The Factorization Theorem

Computing conditional distributions is difficult. The **Neyman Factorization Theorem** provides an elegant shortcut.

::: callout-important
## Theorem (Neyman-Fisher Factorization)

Let $f(x_1, \ldots, x_n; \theta)$ denote the joint pmf or pdf. Then $T = t(X_1, \ldots, X_n)$ is **sufficient for $\theta$** if and only if:

$$f(x_1, \ldots, x_n; \theta) = g(t(x_1, \ldots, x_n); \theta) \cdot h(x_1, \ldots, x_n)$$

where:

-   $g$ depends on the data **only through** $T$ and involves $\theta$
-   $h$ depends on the data but **not on $\theta$**
:::

## Understanding the Factorization

$$f(x_1, \ldots, x_n; \theta) = \underbrace{g(T; \theta)}_{\text{Involves } \theta \text{ and data through } T} \cdot \underbrace{h(x_1, \ldots, x_n)}_{\text{Depends on data, not } \theta}$$

. . .

::: callout-tip
## Strategy for Finding Sufficient Statistics

1.  Write down the joint pmf/pdf
2.  Identify the part that involves $\theta$
3.  Factor it so $\theta$ appears only with a function of the data
4.  That function is your sufficient statistic!
:::

## Example 1: Poisson Distribution

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Poisson}(\mu)$

**Joint pmf:**
$$f(x_1, \ldots, x_n; \mu) = \prod_{i=1}^n \frac{e^{-\mu}\mu^{x_i}}{x_i!} = \frac{e^{-n\mu}\mu^{\sum x_i}}{\prod_{i=1}^n x_i!}$$

. . .

**Factorization:**
$$= \underbrace{e^{-n\mu}\mu^t}_{g(t; \mu)} \cdot \underbrace{\frac{1}{\prod_{i=1}^n x_i!}}_{h(x_1, \ldots, x_n)}$$

. . .

::: callout-important
## Result
$T = \sum_{i=1}^n X_i$ is **sufficient for $\mu$**.
:::

## Example 2: Uniform Distribution (Devore 7.27)

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Uniform}[0, \theta]$

**Joint pdf:**
$$f(x_1, \ldots, x_n; \theta) = \frac{1}{\theta^n} \cdot I(\max(x_i) \leq \theta) \cdot I(\min(x_i) \geq 0)$$

. . .

**Factorization:**
$$= \underbrace{\frac{1}{\theta^n} I(\max(x_i) \leq \theta)}_{g(t; \theta)} \cdot \underbrace{I(\min(x_i) \geq 0)}_{h(x_1, \ldots, x_n)}$$

. . .

::: callout-important
## Result
$T = \max(X_1, \ldots, X_n)$ is **sufficient for $\theta$**. 

Note: $\bar{X}$ is NOT sufficient here!
:::

## Why Isn't $\bar{X}$ Sufficient for Uniform?

```{r}
# Two samples with same mean but different maxima
data_A <- c(2, 4, 6, 8)  # mean = 5, max = 8
data_B <- c(1, 4, 5, 10) # mean = 5, max = 10

cat("Dataset A: mean =", mean(data_A), ", max =", max(data_A), "\n")
cat("Dataset B: mean =", mean(data_B), ", max =", max(data_B), "\n")
```

. . .

These have the same $\bar{X} = 5$, but:

-   Dataset A tells us $\theta \geq 8$
-   Dataset B tells us $\theta \geq 10$

**Knowing only $\bar{X}$ loses information about $\theta$!**

## Jointly Sufficient Statistics

When there are **multiple parameters**, we need **multiple sufficient statistics**.

. . .

**Example: Normal$(\mu, \sigma^2)$** with both unknown:

$$f = \frac{1}{\sigma^n} \exp\left(-\frac{\sum x_i^2 - 2\mu \sum x_i + n\mu^2}{2\sigma^2}\right) \cdot \left(\frac{1}{2\pi}\right)^{n/2}$$

. . .

::: callout-important
## Result
$(\bar{X}, S^2)$ or equivalently $(\sum X_i, \sum X_i^2)$ are **jointly sufficient** for $(\mu, \sigma^2)$.
:::

## Summary: Sufficient Statistics for Common Distributions

| Distribution | Parameter(s) | Sufficient Statistic(s) |
|--------------|--------------|-------------------------|
| Bernoulli($p$) | $p$ | $\sum X_i$ |
| Poisson($\mu$) | $\mu$ | $\sum X_i$ |
| Exponential($\lambda$) | $\lambda$ | $\sum X_i$ |
| Normal($\mu$, known $\sigma^2$) | $\mu$ | $\bar{X}$ |
| Normal($\mu$, $\sigma^2$) | $(\mu, \sigma^2)$ | $(\bar{X}, S^2)$ |
| Uniform$[0, \theta]$ | $\theta$ | $\max(X_i)$ |

# Minimal Sufficiency {background-color="#3498db"}

## Minimal Sufficient Statistics

Sufficient statistics are **not unique**! The entire dataset is always (trivially) sufficient.

. . .

::: callout-important
## Definition

A sufficient statistic $T$ is **minimal sufficient** if it is a function of every other sufficient statistic.
:::

. . .

**Intuition:** Minimal sufficient statistics achieve the **greatest data reduction** without losing information.

. . .

::: callout-tip
## Practical Criterion (Devore 7.74)
$T$ is minimal sufficient if the likelihood ratio $f(\mathbf{x}; \theta)/f(\mathbf{y}; \theta)$ doesn't depend on $\theta$ **if and only if** $t(\mathbf{x}) = t(\mathbf{y})$.
:::

# The Rao-Blackwell Theorem {background-color="#3498db"}

## Why Sufficiency Matters for Estimation

**Key Question:** How do we use sufficient statistics to find better estimators?

. . .

::: callout-important
## Rao-Blackwell Theorem

If $T$ is sufficient for $\theta$ and $U$ is any unbiased estimator for $h(\theta)$, then:

$$U^* = E(U | T)$$

satisfies:

1.  $U^*$ is unbiased for $h(\theta)$
2.  $\text{Var}(U^*) \leq \text{Var}(U)$ (with equality only if $U$ is already a function of $T$)
:::

. . .

**In words:** Conditioning on sufficient statistics **never increases variance** and often **decreases it**.

## Proof of Rao-Blackwell

**Step 1:** $U^*$ doesn't depend on $\theta$ (since $T$ is sufficient)

**Step 2:** By Law of Total Expectation: 
$$E(U^*) = E[E(U|T)] = E(U) = h(\theta) \checkmark$$

**Step 3:** By Law of Total Variance:
$$\text{Var}(U) = \text{Var}(E[U|T]) + E[\text{Var}(U|T)] = \text{Var}(U^*) + E[\text{Var}(U|T)]$$

Since $E[\text{Var}(U|T)] \geq 0$, we have $\text{Var}(U) \geq \text{Var}(U^*)$ ✓

## Example: Poisson Zero-Defect Probability (Devore 7.30)

**Problem:** Estimate $e^{-\mu} = P(\text{no defects})$ for $X_i \sim \text{Poisson}(\mu)$.

**Crude estimator:** $U = I(X_1 = 0)$ is unbiased since $E(U) = P(X_1 = 0) = e^{-\mu}$

. . .

**Apply Rao-Blackwell:** With sufficient statistic $T = \sum X_i$:

$$U^* = P(X_1 = 0 | T = t) = \frac{P(X_1 = 0) \cdot P(\sum_{i=2}^n X_i = t)}{P(T = t)} = \left(1 - \frac{1}{n}\right)^t$$

. . .

::: callout-important
## Improved Estimator
$U^* = \left(1 - \frac{1}{n}\right)^T$ is unbiased with smaller variance!
:::

## Simulation: Rao-Blackwell Improvement

```{r}
#| code-fold: true
#| fig-height: 4.5
set.seed(42)
mu_true <- 1.5; n <- 10; n_sims <- 5000

poisson_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) rpois(n, mu_true)),
    T_sum = map_dbl(data, sum),
    crude_U = map_dbl(data, \(d) as.numeric(d[1] == 0)),
    improved_U = (1 - 1/n)^T_sum
  )

poisson_sim |> 
  select(crude_U, improved_U) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = ifelse(Estimator == "crude_U", "Crude: I(X₁=0)", "Improved: (1-1/n)^T")) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = exp(-mu_true), color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Rao-Blackwell: Variance Reduction Through Sufficiency",
       subtitle = paste0("True e^(-μ) = ", round(exp(-mu_true), 3)), x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("orange", "steelblue")) +
  theme(legend.position = "top")
```

## Key Implication

::: callout-important
## Corollary

Any estimator that is **not** based purely on sufficient statistics is necessarily **inferior** to some other estimator.
:::

. . .

**Also:** MLEs always depend on data only through sufficient statistics (because maximizing $L(\theta) = g(T;\theta) \cdot h(\mathbf{x})$ is equivalent to maximizing $g$).

# Completeness {background-color="#3498db"}

## The Final Piece: Completeness

**Question:** Given a sufficient statistic $T$, is there a **unique** best unbiased estimator based on $T$?

. . .

::: callout-important
## Definition: Complete Statistic

A statistic $T$ is **complete** if for any function $g$:

$$E_\theta[g(T)] = 0 \text{ for all } \theta \implies P_\theta(g(T) = 0) = 1 \text{ for all } \theta$$
:::

. . .

**Intuition:** The only unbiased estimator of zero based on $T$ is the zero function itself.

## Lehmann-Scheffé Theorem

::: callout-important
## Theorem (Lehmann-Scheffé)

If $T$ is a **complete sufficient statistic** and $U = h(T)$ is unbiased for $\tau(\theta)$, then $U$ is the **unique MVUE** of $\tau(\theta)$.
:::

. . .

**Key fact:** For exponential family distributions (Normal, Poisson, Exponential, Binomial, Gamma), the natural sufficient statistics are typically **complete**.

. . .

::: callout-tip
## The Power of Completeness + Sufficiency
Once we find a complete sufficient statistic, any unbiased function of it is automatically the MVUE!
:::

## The Road to MVUE: Summary

**Method 1: Via Sufficiency**

1.  Find sufficient statistic $T$ (Factorization Theorem)
2.  Check completeness (exponential family → typically complete)
3.  Find unbiased function of $T$ → **MVUE** (Lehmann-Scheffé)

. . .

**Method 2: Via Cramér-Rao**

1.  Compute Fisher Information $I(\theta)$
2.  Find CRLB = $1/[nI(\theta)]$
3.  Find estimator achieving CRLB → **MVUE**

# Summary {background-color="#2c3e50"}

## Lesson 5: Key Takeaways

**1. Sufficiency:** $T$ captures all information about $\theta$ in the data

**2. Neyman Factorization:** $f(\mathbf{x}; \theta) = g(T; \theta) \cdot h(\mathbf{x})$ identifies sufficient statistics

**3. Rao-Blackwell:** Conditioning on sufficient statistics improves estimators

**4. Completeness + Sufficiency → Unique MVUE** (Lehmann-Scheffé)

. . .

| Distribution | Sufficient Statistic | Complete? |
|--------------|---------------------|-----------|
| Poisson$(\mu)$ | $\sum X_i$ | Yes |
| Exponential$(\lambda)$ | $\sum X_i$ | Yes |
| Normal$(\mu, \sigma^2)$ | $(\bar{X}, S^2)$ | Yes |
| Uniform$[0, \theta]$ | $\max(X_i)$ | Yes |

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 7.3
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 6.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

Next Lesson: Fisher Information and Large-Sample Properties

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/05_Sufficient_Statistics/05_Sufficient_Statistics.html lessons/05_Sufficient_Statistics/05_Sufficient_Statistics.pdf
```
