---
title: "Midterm Exam"
subtitle: "BSTA 551: Theory of Statistical Inference"
date-modified: "2026-02-02"
description: "Take-home exam covering Lessons 1-11; Due: `r library(here); source(here('class_dates.R')); mid_close %>% format('%m/%d')` at 11pm"
format: 
  html:
    link-external-newwindow: true
    toc: true
  pdf: default 
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Exam Instructions

**This is a take-home exam. You have approximately 2 weeks to complete it.**

The qmd file can be found [here](https://github.com/jminnier/BSTA551_W26/blob/main/homework/Midterm_Exam.qmd) or on Sakai.

- **Time estimate:** This exam is designed to take approximately 2 hours.
- **Resources:** You may use your notes, textbook (Devore, Berk, Carlton), and R/RStudio. You may NOT consult with other students or use AI assistants.
- **Submission:** Submit your work as an html or PDF (rendered from Quarto/RMarkdown). Include all R code with `echo = TRUE`. Include your qmd or rmd file.
- **Show your work:** Full credit requires complete derivations and explanations. For computational problems, include your R code.
- **Seed:** For all simulation problems, use `set.seed(551)` for reproducibility.
- **Academic integrity:** By submitting this exam, you affirm that all work is your own.

**Point values are indicated for each problem. Total: 100 points.**

\newpage

## Part I: Conceptual Questions (45 points)

### Problem 1: Estimator Properties (15 points)

A pharmaceutical company is developing a new assay to measure drug concentration in blood plasma. Let $X_1, X_2, \ldots, X_n$ be independent measurements from a distribution with mean $\mu$ (the true concentration) and variance $\sigma^2$.

Three estimators for $\mu$ are proposed:

- $\hat{\mu}_1 = \bar{X}$ (sample mean)
- $\hat{\mu}_2 = \frac{X_1 + X_n}{2}$ (average of first and last observations)
- $\hat{\mu}_3 = \frac{1}{n+1}\sum_{i=1}^{n} X_i$ (weighted sample mean)

**(a)** (3 points) For each estimator, determine whether it is unbiased for $\mu$. If biased, calculate the bias.

**(b)** (4 points) Calculate the variance of each estimator in terms of $\sigma^2$ and $n$. Assume the observations are independent.

**(c)** (4 points) Calculate the Mean Squared Error (MSE) of each estimator. Which estimator has the smallest MSE for $n = 10$?

**(d)** (4 points) As $n \to \infty$, which estimator(s) are consistent for $\mu$? Justify your answer using the definition of consistency.

\newpage

### Problem 2: Fisher Information and the Cramér-Rao Lower Bound (15 points)

A biostatistician models the total time (in weeks) for a patient to complete a two-phase post-operative rehabilitation protocol after hip replacement surgery. Each phase has an independent, identically distributed duration, so the total rehabilitation time follows a Gamma distribution with known shape $\alpha = 2$ and unknown rate parameter $\beta > 0$. The pdf is:

$$f(x; \beta) = \beta^2 x \, e^{-\beta x}, \quad x > 0$$

For this distribution, $E(X) = 2/\beta$ and $\text{Var}(X) = 2/\beta^2$.

**(a)** (4 points) Calculate the Fisher Information $I(\beta)$ for a single observation. Show your derivation using $I(\beta) = E\left[-\frac{d^2}{d\beta^2}\ell(\beta)\right]$.

**(b)** (3 points) State the Cramér-Rao Lower Bound (CRLB) for unbiased estimators of $\beta$ based on $n$ independent observations.

**(c)** (4 points) The MLE of $\beta$ can be shown to be $\hat{\beta}_{MLE} = 2/\bar{X}$. It can also be shown that $\sum_{i=1}^n X_i \sim \text{Gamma}(2n, \beta)$, and for $Y \sim \text{Gamma}(a, b)$, we have $E(1/Y) = b/(a-1)$ when $a > 1$.

   - Use these facts to show that $\hat{\beta}_{MLE}$ is biased, and find the bias.
   - Propose a modified estimator $\tilde{\beta}$ that IS unbiased for $\beta$.

**(d)** (4 points) It can be shown that for $Y \sim \text{Gamma}(a, b)$, $\text{Var}(1/Y) = \frac{b^2}{(a-1)^2(a-2)}$ when $a > 2$. Use this to find $\text{Var}(\tilde{\beta})$ and determine whether the unbiased estimator achieves the CRLB. What does this result tell us?

\newpage

### Problem 3: Confidence Interval Interpretation (15 points)

A clinical trial investigates the effectiveness of a new statin drug in lowering LDL cholesterol. In a sample of $n = 64$ patients, the mean reduction in LDL cholesterol is $\bar{x} = 28$ mg/dL with sample standard deviation $s = 16$ mg/dL.

**(a)** (4 points) Construct a 95% confidence interval for the true mean reduction $\mu$. Justify your choice of interval (z vs. t) and state any assumptions.

**(b)** (4 points) For each statement below, indicate whether it is a CORRECT or INCORRECT interpretation of your confidence interval. Briefly explain why.

   i. "There is a 95% probability that $\mu$ lies in this interval."
   ii. "If we repeated this study many times, 95% of the resulting confidence intervals would contain $\mu$."
   iii. "95% of patients in this study had LDL reductions within this interval."
   iv. "We are 95% confident that the true mean LDL reduction is in this interval."

**(c)** (4 points) The drug company wants to claim that the mean LDL reduction is at least 25 mg/dL. Based on your confidence interval, can they make this claim at the 95% confidence level? Explain.

**(d)** (3 points) If the company wanted a 95% CI with margin of error no greater than 3 mg/dL, what sample size would be required? (Use $s = 16$ as your estimate of $\sigma$.)

\newpage

## Part II: Computational Problems (55 points)

### Problem 4: Maximum Likelihood Estimation (18 points)

The number of emergency room visits per month at a rural hospital follows a Poisson distribution with unknown rate $\mu$. Over 12 months, the following counts were recorded:

$$X = \{47, 53, 61, 42, 58, 55, 49, 64, 52, 45, 59, 51\}$$

**(a)** (4 points) Write the likelihood function $L(\mu)$ and log-likelihood function $\ell(\mu)$ for the observed data.

**(b)** (4 points) Derive the MLE $\hat{\mu}$ of $\mu$ by maximizing the log-likelihood. Calculate the numerical value of $\hat{\mu}$ for this data.

**(c)** (4 points) Using the Neyman Factorization Theorem, show that $T = \sum_{i=1}^{n} X_i$ is a sufficient statistic for $\mu$.

**(d)** (6 points) Using R:

   - Calculate the MLE $\hat{\mu}$
   - Calculate the standard error of $\hat{\mu}$ using the observed Fisher Information
   - Construct an approximate 95% Wald confidence interval for $\mu$

```{r}
#| label: prob4-template
#| eval: false
# ER visit data
er_visits <- c(47, 53, 61, 42, 58, 55, 49, 64, 52, 45, 59, 51)

# Your code here

```

\newpage

### Problem 5: Method of Moments vs. MLE (15 points)

The survival time (in years) of patients after a certain cardiac procedure is modeled using a Weibull distribution with shape parameter $k = 2$ (known) and unknown scale parameter $\lambda > 0$. The pdf is:

$$f(x; \lambda) = \frac{2x}{\lambda^2} \exp\left(-\frac{x^2}{\lambda^2}\right), \quad x > 0$$

For this distribution: $E(X) = \lambda \cdot \frac{\sqrt{\pi}}{2}$ and $E(X^2) = \lambda^2$

The following survival times (in years) were observed for $n = 10$ patients:

$$X = \{3.2, 4.8, 2.1, 5.5, 3.9, 4.2, 6.1, 3.5, 4.6, 2.8\}$$

**(a)** (5 points) Derive the Method of Moments estimator $\tilde{\lambda}$ for $\lambda$ using the first moment. Calculate its numerical value for this data.

**(b)** (5 points) Derive the MLE $\hat{\lambda}$ for $\lambda$. Calculate its numerical value for this data.

**(c)** (5 points) Using R simulation with 5000 replications, compare the bias, variance, and MSE of the two estimators assuming the true parameter is $\lambda = 4$. Which estimator performs better? Use $n = 10$.

```{r}
#| label: prob5-template
#| eval: false
set.seed(551)

# Simulation parameters
n <- 10
lambda_true <- 4
n_sims <- 5000

# Your simulation code here

```

\newpage

### Problem 6: Confidence Intervals for Proportions (12 points)

In a vaccine efficacy trial, 312 out of 400 vaccinated individuals developed protective antibody levels after receiving a new COVID-19 booster.

**(a)** (3 points) Calculate the sample proportion $\hat{p}$ and its standard error (using the Wald formula).

**(b)** (4 points) Compute three different 95% confidence intervals for the true proportion $p$:

   - Wald interval
   - Score (Wilson) interval (use `prop.test()` in R)
   - Agresti-Coull interval

**(c)** (3 points) Which interval would you recommend for reporting in a medical journal, and why?

**(d)** (2 points) The FDA requires that the lower bound of a 95% CI for vaccine efficacy exceed 70% for approval. Based on your preferred interval, would this vaccine meet the FDA threshold?

```{r}
#| label: prob6-template
#| eval: false
# Vaccine trial data
x <- 312  # successes
n <- 400  # total

# Your code here

```

\newpage

### Problem 7: Bootstrap Confidence Intervals (10 points)

Researchers measured the time-to-recovery (in days) for $n = 15$ patients receiving a new physical therapy protocol after knee surgery:

$$X = \{18, 22, 25, 19, 31, 24, 20, 28, 23, 45, 21, 26, 24, 22, 27\}$$

Notice the potential outlier at 45 days.

**(a)** (2 points) Calculate the sample mean, sample median, and sample standard deviation. Create a histogram or Q-Q plot to assess normality.

**(b)** (4 points) Using B = 10,000 bootstrap replications, compute:

   - The bootstrap standard error of the sample mean
   - The 95% bootstrap percentile confidence interval for the population mean
   - The 95% bootstrap percentile confidence interval for the population median

**(c)** (2 points) Calculate the classical t-interval for the mean. How does it compare to the bootstrap percentile interval?

**(d)** (2 points) For this data, would you recommend using the mean or median to summarize the center of the distribution? Which confidence interval (bootstrap or classical) would you report? Justify your answers.

```{r}
#| label: prob7-template
#| eval: false
set.seed(551)

# Recovery time data
recovery <- c(18, 22, 25, 19, 31, 24, 20, 28, 23, 45, 21, 26, 24, 22, 27)

# Your code here

```

\newpage

## Bonus Question (5 points)

**(Optional - for extra credit)**

A biostatistician is estimating the prevalence of a rare genetic mutation. In a sample of $n = 500$ individuals, only $X = 3$ carry the mutation.

**(a)** (2 points) Calculate the Wald, Score, and Agresti-Coull 95% confidence intervals. What problem do you observe with the Wald interval?

**(b)** (3 points) Explain why the Score and Agresti-Coull intervals are more appropriate for rare events. What property of these intervals addresses the limitation of the Wald interval?

```{r}
#| label: bonus-template
#| eval: false
# Rare mutation data
x <- 3
n <- 500

# Your code here

```

---

**End of Exam**

Good luck!
