---
title: "Lesson 20: Bootstrap and Permutation Methods for Two Samples"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w10d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 20 - Bootstrap and Permutation for Two Samples
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 20: Bootstrap and Permutation Methods for Two Samples {background-color="#2c3e50"}

## Review: Lessons 18–19

**Parametric two-sample methods so far:**

-   **Lesson 18:** Welch $t$ / pooled $t$ for $\mu_1 - \mu_2$; $z$-test for $p_1 - p_2$; $F$-test for $\sigma_1^2/\sigma_2^2$
-   **Lesson 19:** Paired $t$-test; variance reduction from positive within-pair correlation

All require **normality** (exact) or rely on the **CLT** (approximate). What if:

-   Samples are small and distributions are visibly skewed?
-   The parameter of interest is a **median, IQR, ratio**, or other non-mean statistic?
-   Distributional assumptions are questionable?

. . .

**Today's Goals:**

1.  Construct **two-sample bootstrap CIs** by resampling each group independently (C&H Ch. 3; DBC 10.6)
2.  Implement the **two-sample permutation test** by permuting group labels (C&H Ch. 5)
3.  Apply bootstrap to **paired data** by resampling whole pairs
4.  Articulate the precise conceptual difference between bootstrap and permutation
5.  Compare resampling to parametric methods through simulation

## Roadmap

| Method | Goal | Resamples From | Reference |
|-----------------|-----------------|----------------------|-----------------|
| Two-sample bootstrap CI | CI for $\theta_1 - \theta_2$ | Each group separately | C&H Ch. 3; DBC 10.6 |
| Two-sample permutation test | Test $H_0: F_1 = F_2$ | Pooled, relabeling | C&H Ch. 5 |
| Paired bootstrap CI | CI for $\mu_D$ | Pairs with replacement | C&H Ch. 3 |

. . .

::: callout-note
## Motivating Example

**Trauma registry:** Compare hospital length of stay (LOS, days) between blunt and penetrating trauma patients. LOS is strongly right-skewed and $n$ is moderate: the Welch $t$-test CI may be unreliable. Bootstrap and permutation give valid inference with no distributional assumptions.
:::

# Two-Sample Bootstrap CI {background-color="#3498db"}

## The Two-Sample Bootstrap Algorithm (C&H 3.2)

**Setup:** Independent samples $x_1, \ldots, x_{n_1}$ from $F_1$ and $y_1, \ldots, y_{n_2}$ from $F_2$.

**Goal:** CI for $\theta = g(F_1, F_2)$, e.g., $\mu_1 - \mu_2$, $\text{median}_1 - \text{median}_2$, $\sigma_1 / \sigma_2$.

**Algorithm:** for $b = 1, \ldots, B$:

1.  Draw $x_1^{*b}, \ldots, x_{n_1}^{*b}$ **with replacement** from $\{x_1, \ldots, x_{n_1}\}$
2.  Draw $y_1^{*b}, \ldots, y_{n_2}^{*b}$ **with replacement** from $\{y_1, \ldots, y_{n_2}\}$
3.  Compute $\hat\theta^{*b} = g(\mathbf{x}^{*b}, \mathbf{y}^{*b})$

**Percentile CI:** $\bigl(\hat\theta^*_{(\alpha/2)},\; \hat\theta^*_{(1-\alpha/2)}\bigr)$

**Basic (reflection) CI:** $\bigl(2\hat\theta - \hat\theta^*_{(1-\alpha/2)},\; 2\hat\theta - \hat\theta^*_{(\alpha/2)}\bigr)$

. . .

::: callout-important
## Key Point

Resample each group **separately and independently** — this preserves the two-group structure and approximates the sampling distribution of $\hat\theta_1 - \hat\theta_2$. Pooling before resampling would impose $H_0$ (that is permutation, not bootstrap).
:::

## Medical Example: Trauma Length of Stay

```{r}
#| echo: true
#| code-fold: true

set.seed(551)
los_blunt <- rgamma(40, shape = 2.0, rate = 0.28)  # mean ~ 7.1 days
los_penet <- rgamma(30, shape = 1.4, rate = 0.22)  # mean ~ 6.4 days

cat(sprintf("Blunt:       n = %d   mean = %.2f   median = %.2f   sd = %.2f\n",
            length(los_blunt), mean(los_blunt),
            median(los_blunt), sd(los_blunt)))
cat(sprintf("Penetrating: n = %d   mean = %.2f   median = %.2f   sd = %.2f\n",
            length(los_penet), mean(los_penet),
            median(los_penet), sd(los_penet)))
```

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5
#| fig-width: 12

tibble(
  LOS   = c(los_blunt, los_penet),
  Group = rep(c("Blunt", "Penetrating"),
              c(length(los_blunt), length(los_penet)))
) |>
  ggplot(aes(x = LOS, fill = Group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 18,
                 color = "white", alpha = 0.7) +
  geom_density(aes(color = Group), linewidth = 1.2, fill = NA) +
  scale_fill_manual(values  = c("Blunt" = "#006a4e",
                                 "Penetrating" = "#e74c3c")) +
  scale_color_manual(values = c("Blunt" = "#006a4e",
                                 "Penetrating" = "#e74c3c")) +
  facet_wrap(~Group) +
  labs(title    = "Hospital Length of Stay by Trauma Type",
       subtitle = "Right-skewed distributions — parametric CIs may be unreliable for moderate n",
       x = "LOS (days)", y = "Density") +
  theme(legend.position = "none")
```

## Computing the Bootstrap CI

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
B        <- 5000
n1       <- length(los_blunt)
n2       <- length(los_penet)
obs_diff <- mean(los_blunt) - mean(los_penet)

boot_diffs <- replicate(B, {
  mean(sample(los_blunt, n1, replace = TRUE)) -
  mean(sample(los_penet, n2, replace = TRUE))
})

ci_pct   <- quantile(boot_diffs, c(0.025, 0.975))
ci_basic <- 2 * obs_diff - quantile(boot_diffs, c(0.975, 0.025))
ci_welch <- t.test(los_blunt, los_penet)$conf.int

cat(sprintf("Observed difference (Blunt - Penet):  %.3f days\n", obs_diff))
cat(sprintf("Bootstrap SE:                         %.3f\n", sd(boot_diffs)))
cat(sprintf("Percentile 95%% CI:  (%.3f, %.3f)\n", ci_pct[1],   ci_pct[2]))
cat(sprintf("Basic     95%% CI:  (%.3f, %.3f)\n", ci_basic[1], ci_basic[2]))
cat(sprintf("Welch t   95%% CI:  (%.3f, %.3f)\n", ci_welch[1], ci_welch[2]))
```

## Visualizing the Bootstrap Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6.5
#| fig-width: 13

tibble(boot_diff = boot_diffs) |>
  ggplot(aes(x = boot_diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 55,
                 fill = "#3498db", color = "white", alpha = 0.7) +
  geom_density(color = "#2c3e50", linewidth = 1.1) +
  geom_vline(xintercept = obs_diff,
             color = "#e74c3c", linewidth = 1.5) +
  geom_vline(xintercept = ci_pct,
             color = "#006a4e", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = ci_welch,
             color = "#e74c3c", linewidth = 1.0, linetype = "dotted") +
  annotate("text", x = obs_diff + 0.1, y = 0.52,
           label = bquote(hat(Delta) == .(round(obs_diff, 2))),
           color = "#e74c3c", size = 5.5, hjust = 0) +
  annotate("text", x = ci_pct[2] + 0.1, y = 0.45,
           label = "Bootstrap\npercentile CI", color = "#006a4e",
           size = 4.5, hjust = 0) +
  annotate("text", x = ci_welch[2] + 0.1, y = 0.35,
           label = "Welch t CI", color = "#e74c3c",
           size = 4.5, hjust = 0) +
  labs(
    title    = "Bootstrap Distribution: Mean LOS Difference (Blunt \u2212 Penetrating)",
    subtitle = "Green dashed = 95% bootstrap percentile CI   |   Red dotted = Welch t CI",
    x        = expression(bar(x)[Blunt]^"*" - bar(x)[Penet]^"*" ~ "(days)"),
    y        = "Density"
  )
```

## Bootstrap CI for Any Statistic

A key advantage over parametric methods: the same algorithm works for **any statistic**.

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Difference in medians (no closed-form parametric CI exists)
boot_med <- replicate(B, {
  median(sample(los_blunt, n1, replace = TRUE)) -
  median(sample(los_penet, n2, replace = TRUE))
})

# Ratio of standard deviations
boot_sdr <- replicate(B, {
  sd(sample(los_blunt, n1, replace = TRUE)) /
  sd(sample(los_penet, n2, replace = TRUE))
})

cat("--- Difference in MEDIANS ---\n")
cat(sprintf("  Observed: %.3f days\n",
            median(los_blunt) - median(los_penet)))
cat(sprintf("  Bootstrap 95%% CI: (%.3f, %.3f)\n\n",
            quantile(boot_med, 0.025), quantile(boot_med, 0.975)))

cat("--- Ratio of SDs ---\n")
cat(sprintf("  Observed: %.3f\n", sd(los_blunt) / sd(los_penet)))
cat(sprintf("  Bootstrap 95%% CI: (%.3f, %.3f)\n",
            quantile(boot_sdr, 0.025), quantile(boot_sdr, 0.975)))
```

# Two-Sample Permutation Test {background-color="#e74c3c"}

## The Permutation Test: Core Idea (C&H Ch. 5)

**Null hypothesis:** $H_0: F_1 = F_2$ — the two groups have the **same distribution**.

Under $H_0$, the group label on each observation is **arbitrary** (exchangeable). Any assignment of the $N = n_1 + n_2$ pooled observations into groups of size $n_1$ and $n_2$ is equally likely.

. . .

**Algorithm:** for $b = 1, \ldots, B$:

1.  Pool all $N$ observations
2.  Randomly permute labels: assign the first $n_1$ to group 1, the rest to group 2
3.  Compute $T^{*b}$ (the test statistic under the permuted labeling)

**Two-sided p-value:** $\;\hat{p} = \frac{1}{B}\sum_{b=1}^B \mathbf{1}\bigl(|T^{*b}| \geq |T_\text{obs}|\bigr)$

. . .

::: callout-important
## Bootstrap vs. Permutation — The Essential Distinction

|   | Bootstrap CI | Permutation test |
|-----------------|-----------------------|--------------------------------|
| **Purpose** | Estimate variability | Test $H_0$ |
| **Resamples from** | Each group separately | Pooled data (relabeling) |
| **Approximates** | Sampling distribution of $\hat\theta$ | Null distribution of $T$ |
| $H_0$ imposed? | No | Yes ($F_1 = F_2$) |
:::

## Permutation Test: Implementation

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
B_perm  <- 9999
all_los <- c(los_blunt, los_penet)
N       <- length(all_los)
t_obs   <- mean(los_blunt) - mean(los_penet)

perm_diffs <- replicate(B_perm, {
  idx <- sample(N)
  mean(all_los[idx[seq_len(n1)]]) - mean(all_los[idx[(n1 + 1):N]])
})

p_perm  <- mean(abs(perm_diffs) >= abs(t_obs))
p_welch <- t.test(los_blunt, los_penet)$p.value

cat(sprintf("Observed mean difference:   %.3f days\n", t_obs))
cat(sprintf("Permutation p-value:        %.4f\n",      p_perm))
cat(sprintf("Welch t-test p-value:       %.4f\n",      p_welch))
```

## Visualizing the Permutation Null Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6.5
#| fig-width: 13

tibble(perm_diff = perm_diffs) |>
  ggplot(aes(x = perm_diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "steelblue", color = "white", alpha = 0.65) +
  geom_density(color = "#2c3e50", linewidth = 1.1) +
  geom_vline(xintercept = c(-abs(t_obs), abs(t_obs)),
             color = "#e74c3c", linewidth = 1.5) +
  annotate("text", x = abs(t_obs) + 0.05, y = 0.65,
           label = bquote("|" * T[obs] * "| =" ~ .(round(abs(t_obs), 2))),
           color = "#e74c3c", size = 5.5, hjust = 0) +
  annotate("text", x = 0, y = 0.72,
           label = sprintf("p = %.4f", p_perm),
           size = 5.5, color = "#2c3e50") +
  labs(
    title    = "Permutation Null Distribution: Difference in Mean LOS",
    subtitle = "Red lines mark \u00b1|T_obs|. Fraction of permuted statistics beyond = p-value.",
    x        = expression(bar(x)[1]^"*" - bar(x)[2]^"*" ~ "(days)"),
    y        = "Density"
  )
```

## Permutation Tests Work with Any Statistic

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Permutation test for difference in MEDIANS
t_obs_med <- median(los_blunt) - median(los_penet)
perm_med  <- replicate(9999, {
  idx <- sample(N)
  median(all_los[idx[seq_len(n1)]]) - median(all_los[idx[(n1 + 1):N]])
})
p_perm_med <- mean(abs(perm_med) >= abs(t_obs_med))

cat(sprintf("Median difference: %.3f   permutation p = %.4f\n",
            t_obs_med, p_perm_med))

# Permutation test for VARIANCE RATIO (vs F-test)
t_obs_var <- var(los_blunt) / var(los_penet)
perm_var  <- replicate(9999, {
  idx <- sample(N)
  var(all_los[idx[seq_len(n1)]]) / var(all_los[idx[(n1 + 1):N]])
})
p_perm_var <- mean(perm_var >= t_obs_var | perm_var <= 1 / t_obs_var)
p_ftest    <- var.test(los_blunt, los_penet)$p.value

cat(sprintf("Variance ratio:    %.3f   permutation p = %.4f   F-test p = %.4f\n",
            t_obs_var, p_perm_var, p_ftest))
```

. . .

The permutation test for the variance ratio is valid regardless of distribution shape, unlike the $F$-test, which requires normality.

# Bootstrap for Paired Data {background-color="#27ae60"}

## Paired Bootstrap CI

For paired data, the bootstrap resamples **entire pairs** $(D_i)$ with replacement which preserves the pairing structure.

**Algorithm:** for $b = 1, \ldots, B$:

1.  Draw $D_1^{*b}, \ldots, D_n^{*b}$ **with replacement** from $\{D_1, \ldots, D_n\}$
2.  Compute $\bar{D}^{*b} = \frac{1}{n}\sum D_i^{*b}$

**Percentile CI:** $\bigl(\bar{D}^*_{(\alpha/2)},\; \bar{D}^*_{(1-\alpha/2)}\bigr)$

No normality assumption on $D_i$.

```{r}
#| echo: true
#| code-fold: false

# Crossover trial differences from Lesson 19
set.seed(42)
n_pair <- 20; rho_p <- 0.75
L_p    <- chol(10^2 * matrix(c(1, rho_p, rho_p, 1), 2))
xy_p   <- matrix(rnorm(2 * n_pair), n_pair, 2) %*% L_p +
          matrix(c(0, -10), n_pair, 2, byrow = TRUE)
d_cross <- xy_p[, 2] - xy_p[, 1]

set.seed(551)
boot_D         <- replicate(5000, mean(sample(d_cross, n_pair, replace = TRUE)))
ci_boot_paired <- quantile(boot_D, c(0.025, 0.975))
ci_t_paired    <- t.test(d_cross)$conf.int

cat(sprintf("Paired t-test 95%% CI:  (%.3f, %.3f)\n",
            ci_t_paired[1], ci_t_paired[2]))
cat(sprintf("Bootstrap    95%% CI:  (%.3f, %.3f)\n",
            ci_boot_paired[1], ci_boot_paired[2]))
```

With approximately normal differences and moderate $n$, the two CIs are very similar. The bootstrap shines when $D_i$ are clearly non-normal or $n$ is very small.

# Comparing Methods: Simulation {background-color="#9b59b6"}

## Type I Error and Power: Welch t-test vs. Permutation Test

The permutation test derives its null distribution empirically rather than from the $t$-distribution. For the **mean difference** statistic, how much does it matter?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 15

set.seed(551)
B_sim  <- 3000
B_perm <- 499

# chi-squared(2): mean = 2, heavily right-skewed (skewness = 2*sqrt(2) ~ 2.8)
# shifted to mean 0 for fair comparison
sim_scenario <- function(dist_fn, n = 10, effect = 0) {
  replicate(B_sim, {
    x1  <- dist_fn(n)
    x2  <- dist_fn(n) + effect
    Ns  <- 2 * n
    all <- c(x1, x2)
    t0  <- mean(x1) - mean(x2)
    p_welch <- t.test(x1, x2)$p.value
    p_perm  <- {
      perms <- replicate(B_perm, {
        idx <- sample(Ns)
        mean(all[idx[seq_len(n)]]) - mean(all[idx[(n + 1):Ns]])
      })
      mean(abs(perms) >= abs(t0))
    }
    c(Welch = p_welch, Permutation = p_perm)
  }) |> t() |> as_tibble() |>
    summarise(across(everything(), \(x) mean(x < 0.05)))
}

norm_fn <- \(n) rnorm(n, 0, 1)
skew_fn <- \(n) rchisq(n, df = 2) - 2   # shifted chi-sq(2), mean = 0

results <- bind_rows(
  bind_rows(
    sim_scenario(norm_fn, n = 10, effect = 0.0) |> mutate(Scenario = "Type I Error (effect = 0)"),
    sim_scenario(norm_fn, n = 10, effect = 1.0) |> mutate(Scenario = "Power (effect = 1 SD)")
  ) |> mutate(Distribution = "Normal(0,1) — symmetric"),
  bind_rows(
    sim_scenario(skew_fn, n = 10, effect = 0.0) |> mutate(Scenario = "Type I Error (effect = 0)"),
    sim_scenario(skew_fn, n = 10, effect = 1.0) |> mutate(Scenario = "Power (effect = 1 SD)")
  ) |> mutate(Distribution = "chi-sq(2) shifted — heavily right-skewed")
)

results |>
  pivot_longer(c(Welch, Permutation),
               names_to = "Method", values_to = "Rate") |>
  mutate(
    Scenario     = factor(Scenario,
                          levels = c("Type I Error (effect = 0)",
                                     "Power (effect = 1 SD)")),
    Distribution = factor(Distribution,
                          levels = c("Normal(0,1) — symmetric",
                                     "chi-sq(2) shifted — heavily right-skewed"))
  ) |>
  ggplot(aes(x = Method, y = Rate, fill = Method)) +
  geom_col(width = 0.5) +
  geom_hline(
    data = tibble(
      Scenario = factor("Type I Error (effect = 0)",
                        levels = c("Type I Error (effect = 0)", "Power (effect = 1 SD)")),
      y = 0.05
    ),
    aes(yintercept = y),
    linetype = "dashed", linewidth = 1.2, color = "black"
  ) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 0.1)),
            vjust = -0.4, size = 4.5, fontface = "bold") +
  facet_grid(Distribution ~ Scenario) +
  scale_fill_manual(values = c("Welch" = "#006a4e", "Permutation" = "#e74c3c")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.95)) +
  labs(
    title    = "Welch t-test vs. Permutation Test  (n = 10 per group, 3000 simulations)",
    subtitle = "Both methods control Type I error near 5% and have similar power — permutation's advantage is flexibility.",
    x = NULL, y = "Rejection Rate", fill = NULL
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12))
```

::: callout-note
The two-sided Welch $t$-test is **robust to skewness for Type I error** even at small $n$: skewness makes each tail individually wrong, but in a two-sided test the errors in the two tails largely cancel. The simulation below confirms this — the two methods agree closely. Permutation's advantage is **generality**: it works with *any* test statistic, not just means.
:::


# Key Takeaways {background-color="#2c3e50"}

## Key Takeaways (C&H Ch. 3, 5; DBC 10.6)

::: callout-important
## Bootstrap and Permutation for Two Samples

1.  **Two-sample bootstrap CI:** Resample each group **independently** with replacement; construct percentile or basic CI. Works for any statistic — means, medians, SD ratios, etc. No distributional assumption

2.  **Permutation test:** Pool all $N$ observations, **relabel** groups at random; fraction of permuted statistics $\geq |T_\text{obs}|$ gives a valid p-value under $H_0: F_1 = F_2$ (exchangeability)

3.  **Bootstrap** $\neq$ permutation: Bootstrap approximates the **sampling distribution** of $\hat\theta$ (for CIs); permutation approximates the **null distribution** of $T$ (for testing); they solve different problems and use data differently

4.  **Paired bootstrap:** Resample **whole pairs** $(D_i)$ to preserve the pairing structure; produces CI for $\mu_D$ without normality assumption

5.  **When to prefer resampling:** Small $n$ with non-normal data; non-standard statistics with no closed-form SE; any setting where parametric assumptions are doubtful
:::

## Looking Ahead

**Next class (Lesson 21): Nonparametric Inference (DBC Ch. 14)**

-   Sign test: based on the binomial distribution, zero distributional assumptions
-   Wilcoxon signed-rank test: uses ranks of $|D_i|$, requires symmetric differences
-   Wilcoxon rank-sum test (Mann–Whitney): independent samples, ranks all $N$ observations
-   Asymptotic relative efficiency vs. parametric counterparts

## References

::: nonincremental
-   Chihara, L. M. and Hesterberg, T. C. *Mathematical Statistics with Resampling and R* (Wiley). Chapters 3 and 5.
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 10.6.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/20_BootstrapPermutation/20_BootstrapPermutation.html lessons/20_BootstrapPermutation/20_BootstrapPermutation.pdf
```

::::::::
