---
title: "Lesson 20: Bootstrap and Permutation Methods for Two Samples"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w10d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 20 - Bootstrap and Permutation for Two Samples
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
library(infer)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 20: Bootstrap and Permutation Methods for Two Samples {background-color="#2c3e50"}

## Review: Lessons 18–19

**Parametric two-sample methods so far:**

-   **Lesson 18:** Welch $t$ / pooled $t$ for $\mu_1 - \mu_2$; $z$-test for $p_1 - p_2$; $F$-test for $\sigma_1^2/\sigma_2^2$
-   **Lesson 19:** Paired $t$-test; variance reduction from positive within-pair correlation

All require **normality** (exact) or rely on the **CLT** (approximate). What if:

-   Samples are small and distributions are visibly skewed?
-   The parameter of interest is a **median, IQR, ratio**, or other non-mean statistic?
-   Distributional assumptions are questionable?

. . .

**Today's Goals:**

1.  Construct **two-sample bootstrap CIs** by resampling each group independently (C&H Ch. 3; DBC 10.6)
2.  Perform **two-sample bootstrap hypothesis tests** by shifting to satisfy $H_0$ (C&H Ch. 8)
3.  Implement the **two-sample permutation test** and understand exactly when it is valid (C&H Ch. 5)
4.  Apply bootstrap CI and bootstrap test to **paired data**
5.  Use the **`infer` package** for tidy, pipeline-based resampling inference

## Roadmap

| Method | Goal | Resamples From | Reference |
|---|---|---|---|
| Two-sample bootstrap CI | CI for $\theta_1 - \theta_2$ | Each group separately | C&H Ch. 3; DBC 10.6 |
| Two-sample bootstrap test | p-value for $H_0:\theta_1=\theta_2$ | Shifted/centred samples | C&H Ch. 8 |
| Two-sample permutation test | Test $H_0: F_1 = F_2$ | Pooled, relabeling | C&H Ch. 5 |
| Paired bootstrap CI | CI for $\mu_D$ | Pairs with replacement | C&H Ch. 3 |
| Paired bootstrap test | p-value for $H_0:\mu_D=0$ | Shifted differences | C&H Ch. 8 |
| `infer` package | Tidy pipeline for all of the above | — | infer vignettes |

. . .

::: callout-note
## Motivating Example

**Trauma registry:** Compare hospital length of stay (LOS, days) between blunt and penetrating trauma patients. LOS is strongly right-skewed and $n$ is moderate: the Welch $t$-test CI may be unreliable. Bootstrap and permutation give valid inference with no distributional assumptions.
:::

# Two-Sample Bootstrap CI {background-color="#3498db"}

## The Two-Sample Bootstrap Algorithm (C&H 3.2)

**Setup:** Independent samples $x_1, \ldots, x_{n_1}$ from $F_1$ and $y_1, \ldots, y_{n_2}$ from $F_2$.

**Goal:** CI for $\theta = g(F_1, F_2)$, e.g., $\mu_1 - \mu_2$, $\text{median}_1 - \text{median}_2$, $\sigma_1 / \sigma_2$.

**Algorithm:** for $b = 1, \ldots, B$:

1.  Draw $x_1^{*b}, \ldots, x_{n_1}^{*b}$ **with replacement** from $\{x_1, \ldots, x_{n_1}\}$
2.  Draw $y_1^{*b}, \ldots, y_{n_2}^{*b}$ **with replacement** from $\{y_1, \ldots, y_{n_2}\}$
3.  Compute $\hat\theta^{*b} = g(\mathbf{x}^{*b}, \mathbf{y}^{*b})$

**Percentile CI:** $\bigl(\hat\theta^*_{(\alpha/2)},\; \hat\theta^*_{(1-\alpha/2)}\bigr)$

**Basic (reflection) CI:** $\bigl(2\hat\theta - \hat\theta^*_{(1-\alpha/2)},\; 2\hat\theta - \hat\theta^*_{(\alpha/2)}\bigr)$

**Bootstrap SE CI:** $\hat\theta \;\pm\; z_{\alpha/2} \cdot \widehat{\text{SE}}_{\text{boot}}$, where $\widehat{\text{SE}}_{\text{boot}} = \text{sd}(\hat\theta^{*1}, \ldots, \hat\theta^{*B})$

. . .

::: callout-important
## Key Point

Resample each group **separately and independently** — this preserves the two-group structure and approximates the sampling distribution of $\hat\theta_1 - \hat\theta_2$. Pooling before resampling would impose $H_0$: that is permutation, not bootstrap.
:::

## Medical Example: Trauma Length of Stay


```{r}
#| echo: true
#| code-fold: true

# (shape/rate = mean; sqrt(shape)/rate = SD)
set.seed(551)
los_blunt <- rgamma(40, shape = 2.8, rate = 0.33)  # mean ~ 8.5 days, SD ~ 5.1
los_penet <- rgamma(30, shape = 2.0, rate = 0.38)  # mean ~ 5.3 days, SD ~ 3.7

cat(sprintf("Blunt:       n = %d   mean = %.2f   median = %.2f   sd = %.2f\n",
            length(los_blunt), mean(los_blunt),
            median(los_blunt), sd(los_blunt)))
cat(sprintf("Penetrating: n = %d   mean = %.2f   median = %.2f   sd = %.2f\n",
            length(los_penet), mean(los_penet),
            median(los_penet), sd(los_penet)))
```

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5
#| fig-width: 12

tibble(
  LOS   = c(los_blunt, los_penet),
  Group = rep(c("Blunt", "Penetrating"),
              c(length(los_blunt), length(los_penet)))
) |>
  ggplot(aes(x = LOS, fill = Group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 18,
                 color = "white", alpha = 0.7) +
  geom_density(aes(color = Group), linewidth = 1.2, fill = NA) +
  scale_fill_manual(values  = c("Blunt" = "#006a4e",
                                 "Penetrating" = "#e74c3c")) +
  scale_color_manual(values = c("Blunt" = "#006a4e",
                                 "Penetrating" = "#e74c3c")) +
  facet_wrap(~Group) +
  labs(title    = "Hospital Length of Stay by Trauma Type",
       subtitle = "Right-skewed distributions — blunt trauma stays appear longer; evidence is borderline significant",
       x = "LOS (days)", y = "Density") +
  theme(legend.position = "none")
```

## Computing the Bootstrap CI: Difference in Means

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
B        <- 5000
n1       <- length(los_blunt)
n2       <- length(los_penet)
obs_diff <- mean(los_blunt) - mean(los_penet)

boot_diffs <- replicate(B, {
  mean(sample(los_blunt, n1, replace = TRUE)) -
  mean(sample(los_penet, n2, replace = TRUE))
})

ci_pct   <- quantile(boot_diffs, c(0.025, 0.975))
ci_basic <- 2 * obs_diff - quantile(boot_diffs, c(0.975, 0.025))
ci_se    <- obs_diff + c(-1, 1) * qnorm(0.975) * sd(boot_diffs)
ci_welch <- t.test(los_blunt, los_penet)$conf.int

cat(sprintf("Observed difference (Blunt - Penet):  %.3f days\n", obs_diff))
cat(sprintf("Bootstrap SE:                         %.3f\n", sd(boot_diffs)))
cat(sprintf("Bootstrap SE  95%% CI: (%.3f, %.3f)\n", ci_se[1],    ci_se[2]))
cat(sprintf("Percentile    95%% CI: (%.3f, %.3f)\n", ci_pct[1],   ci_pct[2]))
cat(sprintf("Basic         95%% CI: (%.3f, %.3f)\n", ci_basic[1], ci_basic[2]))
cat(sprintf("Welch t       95%% CI: (%.3f, %.3f)\n", ci_welch[1], ci_welch[2]))
```

## Visualizing the Bootstrap Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6.5
#| fig-width: 13

tibble(boot_diff = boot_diffs) |>
  ggplot(aes(x = boot_diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 55,
                 fill = "#3498db", color = "white", alpha = 0.7) +
  geom_density(color = "#2c3e50", linewidth = 1.1) +
  geom_vline(xintercept = obs_diff,
             color = "#e74c3c", linewidth = 1.5) +
  geom_vline(xintercept = ci_pct,
             color = "#006a4e", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = ci_welch,
             color = "#e74c3c", linewidth = 1.0, linetype = "dotted") +
  annotate("text", x = obs_diff + 0.1, y = 0.52,
           label = bquote(hat(Delta) == .(round(obs_diff, 2))),
           color = "#e74c3c", size = 5.5, hjust = 0) +
  annotate("text", x = ci_pct[2] + 0.1, y = 0.45,
           label = "Bootstrap\npercentile CI", color = "#006a4e",
           size = 4.5, hjust = 0) +
  annotate("text", x = ci_welch[2] + 0.1, y = 0.35,
           label = "Welch t CI", color = "#e74c3c",
           size = 4.5, hjust = 0) +
  labs(
    title    = "Bootstrap Distribution: Mean LOS Difference (Blunt \u2212 Penetrating)",
    subtitle = "Green dashed = 95% bootstrap percentile CI   |   Red dotted = Welch t CI",
    x        = expression(bar(x)[Blunt]^"*" - bar(x)[Penet]^"*" ~ "(days)"),
    y        = "Density"
  )
```

## Bootstrap CI for Any Statistic

A key advantage over parametric methods: the same algorithm works for **any statistic**.

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Difference in medians (no closed-form parametric CI exists)
boot_med <- replicate(B, {
  median(sample(los_blunt, n1, replace = TRUE)) -
  median(sample(los_penet, n2, replace = TRUE))
})

# Ratio of standard deviations
boot_sdr <- replicate(B, {
  sd(sample(los_blunt, n1, replace = TRUE)) /
  sd(sample(los_penet, n2, replace = TRUE))
})

cat("--- Difference in MEDIANS ---\n")
cat(sprintf("  Observed: %.3f days\n",
            median(los_blunt) - median(los_penet)))
cat(sprintf("  Bootstrap 95%% CI: (%.3f, %.3f)\n\n",
            quantile(boot_med, 0.025), quantile(boot_med, 0.975)))

cat("--- Ratio of SDs ---\n")
cat(sprintf("  Observed: %.3f\n", sd(los_blunt) / sd(los_penet)))
cat(sprintf("  Bootstrap 95%% CI: (%.3f, %.3f)\n",
            quantile(boot_sdr, 0.025), quantile(boot_sdr, 0.975)))
```

# Two-Sample Bootstrap Hypothesis Test {background-color="#8e44ad"}

## Bootstrap Testing: The Core Challenge

Recall from Lesson 17: for a **one-sample** bootstrap test of $H_0: \mu = \mu_0$, we **shift** the sample to satisfy $H_0$ before resampling, then compare the observed statistic to the shifted null distribution.

For **two samples**, we test $H_0: \theta_1 = \theta_2$ (e.g., $\mu_1 = \mu_2$). The same logic applies: we must resample in a way that **imposes $H_0$**.

. . .

::: callout-warning
## Why Not Just Use the Bootstrap Distribution?

If we simply computed the distribution of $\hat\theta^* = \hat\mu_1^* - \hat\mu_2^*$ by resampling each group, we would get the **sampling distribution** centered at the **observed** $\hat\Delta = \hat\mu_1 - \hat\mu_2$.

This gives us a confidence interval — but **not a p-value under $H_0$**. The null distribution must be centered at $0$.
:::

## Two-Sample Bootstrap Test: The Shift Method (C&H 8.2)

**Setup:** Test $H_0: \mu_1 = \mu_2$ vs $H_a: \mu_1 \neq \mu_2$.

**Step 1: Impose $H_0$** by centering both samples at the same mean:

$$\tilde{x}_i = x_i - \bar{x} + \bar{X}_{\text{pool}}, \qquad \tilde{y}_j = y_j - \bar{y} + \bar{X}_{\text{pool}}$$

where $\bar{X}_{\text{pool}} = \frac{n_1 \bar{x} + n_2 \bar{y}}{n_1 + n_2}$ is the pooled mean.

Now $\tilde{x}$ and $\tilde{y}$ have the **same mean** — $H_0$ is satisfied.

. . .

**Step 2: Generate the null distribution** by resampling from the shifted samples:

for $b = 1, \ldots, B$:

1. Draw $\tilde{x}^{*b}$ (size $n_1$) from $\{\tilde{x}_i\}$ with replacement
2. Draw $\tilde{y}^{*b}$ (size $n_2$) from $\{\tilde{y}_j\}$ with replacement
3. Compute $T^{*b} = \bar{\tilde{x}}^{*b} - \bar{\tilde{y}}^{*b}$

**p-value:** $\hat{p} = \frac{1}{B}\sum_{b=1}^{B} \mathbf{1}\bigl(|T^{*b}| \geq |T_\text{obs}|\bigr)$

where $T_\text{obs} = \bar{x} - \bar{y}$.

## Two-Sample Bootstrap Test: Implementation

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
B_test <- 9999

# Observed test statistic
t_obs <- mean(los_blunt) - mean(los_penet)

# Step 1: Shift samples to satisfy H0 (equal means)
grand_mean <- (sum(los_blunt) + sum(los_penet)) / (n1 + n2)
x_shift    <- los_blunt - mean(los_blunt) + grand_mean
y_shift    <- los_penet - mean(los_penet) + grand_mean

cat(sprintf("Before shifting:  mean(blunt) = %.3f, mean(penet) = %.3f\n",
            mean(los_blunt), mean(los_penet)))
cat(sprintf("After  shifting:  mean(x_shift)= %.3f, mean(y_shift)= %.3f\n",
            mean(x_shift), mean(y_shift)))

# Step 2: Resample from shifted samples
boot_null <- replicate(B_test, {
  mean(sample(x_shift, n1, replace = TRUE)) -
  mean(sample(y_shift, n2, replace = TRUE))
})

p_boot  <- mean(abs(boot_null) >= abs(t_obs))
p_welch <- t.test(los_blunt, los_penet)$p.value

cat(sprintf("\nObserved difference:         %.3f days\n", t_obs))
cat(sprintf("Bootstrap test p-value:      %.4f\n", p_boot))
cat(sprintf("Welch t-test p-value:        %.4f\n", p_welch))
```

## Visualizing the Bootstrap Null Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 13

tibble(null_diff = boot_null) |>
  ggplot(aes(x = null_diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#8e44ad", color = "white", alpha = 0.65) +
  geom_density(color = "#2c3e50", linewidth = 1.1) +
  geom_vline(xintercept = c(-abs(t_obs), abs(t_obs)),
             color = "#e74c3c", linewidth = 1.5) +
  annotate("text", x = abs(t_obs) + 0.05, y = 0.6,
           label = bquote("|" * T[obs] * "| =" ~ .(round(abs(t_obs), 2))),
           color = "#e74c3c", size = 5.5, hjust = 0) +
  annotate("text", x = 0, y = 0.7,
           label = sprintf("Bootstrap p = %.4f", p_boot),
           size = 5.5, color = "#2c3e50") +
  labs(
    title    = "Bootstrap Null Distribution: Difference in Mean LOS (Shift Method)",
    subtitle = "Null distribution centered at 0 by shifting both samples to grand mean",
    x        = expression(bar(x)[shift]^"*" - bar(y)[shift]^"*" ~ "(days)"),
    y        = "Density"
  )
```

## Bootstrap Test for Any Statistic (e.g., Median)

The shift method generalizes to **any location statistic** — shift both samples to equal medians (or equal means) and resample.

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Bootstrap test for MEDIAN difference
t_obs_med <- median(los_blunt) - median(los_penet)

# Shift both samples to have equal medians (use grand median or grand mean)
x_shift_med <- los_blunt - median(los_blunt) + grand_mean
y_shift_med <- los_penet - median(los_penet) + grand_mean

boot_null_med <- replicate(B_test, {
  median(sample(x_shift_med, n1, replace = TRUE)) -
  median(sample(y_shift_med, n2, replace = TRUE))
})

p_boot_med <- mean(abs(boot_null_med) >= abs(t_obs_med))

cat(sprintf("Median difference: %.3f days\n", t_obs_med))
cat(sprintf("Bootstrap test p (median):  %.4f\n", p_boot_med))
cat("(No standard parametric test for medians exists for this skewed setting)\n")
```

. . .

::: callout-tip
## Practical Note

For testing mean differences specifically, the **permutation test** (next section) is often preferred over the shift-bootstrap — it is simpler, exact under exchangeability, and avoids the need to carefully define "imposing $H_0$." The bootstrap test shines for non-location statistics where permutation is not straightforwardly applicable.
:::

# Two-Sample Permutation Test {background-color="#e74c3c"}

## The Permutation Test: Core Idea (C&H Ch. 5)

**Null hypothesis:** $H_0: F_1 = F_2$ — the two groups have the **same distribution**.

Under $H_0$, the group label on each observation is **arbitrary** (exchangeable). Any assignment of the $N = n_1 + n_2$ pooled observations into groups of size $n_1$ and $n_2$ is equally likely.

. . .

**Algorithm:** for $b = 1, \ldots, B$:

1.  Pool all $N$ observations
2.  Randomly permute labels: assign the first $n_1$ to group 1, the rest to group 2
3.  Compute $T^{*b}$ (the test statistic under the permuted labeling)

**Two-sided p-value:** $\;\hat{p} = \frac{1}{B}\sum_{b=1}^B \mathbf{1}\bigl(|T^{*b}| \geq |T_\text{obs}|\bigr)$

. . .

::: callout-important
## Bootstrap vs. Permutation — The Essential Distinction

|   | Bootstrap CI | Bootstrap Test | Permutation Test |
|---|---|---|---|
| **Purpose** | Estimate variability | Test $H_0$ | Test $H_0$ |
| **Resamples from** | Each group separately | Shifted samples | Pooled (relabeling) |
| **Approximates** | Sampling dist. of $\hat\theta$ | Null dist. of $T$ | Null dist. of $T$ |
| **$H_0$ imposed?** | No | Yes (via shift) | Yes ($F_1 = F_2$) |
| **Assumption** | Independence | Independence | **Exchangeability** |
:::

## When Is the Permutation Test Valid?

The permutation test is valid (exact Type I error control) when the observations are **exchangeable** under $H_0$.

::: callout-important
## Exchangeability: What It Means

Observations $X_1, \ldots, X_{n_1}, Y_1, \ldots, Y_{n_2}$ are **exchangeable** under $H_0$ if permuting group labels does not change the joint distribution. This holds when:

$$H_0: F_1 = F_2 \quad \text{(identical distributions)}$$

If $H_0$ is true, any relabeling of the pooled data is just as likely as any other — permuting is valid.
:::

. . .

**When exchangeability holds:**

- Randomized experiments: treatment assigned at random, so under $H_0$ labels are arbitrary
- Observational data: valid if groups differ **only** in the parameter of interest (strong assumption)
- Testing $H_0: F_1 = F_2$ exactly (not just equal means)

**When exchangeability fails (permutation test is invalid):**

- Groups differ in **variance or shape** but not in means: permuting is not valid for a test of means because $F_1 \neq F_2$ even under the mean-equality $H_0$
- Paired or clustered data: pairing creates dependence — don't ignore the structure
- Confounding: if groups also differ on covariates, labels are not exchangeable

## Exchangeability in Practice: What Hypothesis Are We Testing?

::: callout-warning
## A Subtle Point

The permutation test is an **exact test** of $H_0: F_1 = F_2$. When used with the mean difference as the test statistic, it **also** has good power to detect $\mu_1 \neq \mu_2$.

However, if the groups differ in **variance** but not in mean, the permutation test (using the mean difference as the statistic) can have **inflated Type I error** because the null $F_1 = F_2$ includes equal variances. Rejecting $H_0: F_1 = F_2$ does not tell us *why* it was rejected.

**Solutions:**

- If you care only about means: use the Welch $t$-test (robust to unequal variances) or the bootstrap test (shift method)
- If you want a permutation test for means when variances may differ: use a standardized test statistic (Welch-type) but permute — this has approximately valid Type I error
- Reserve permutation tests for settings where $F_1 = F_2$ is a plausible and scientifically meaningful null
:::

## Exchangeability: Randomized vs. Observational Studies

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5.5
#| fig-width: 14

scenarios <- tibble(
  Setting = c(
    "RCT: treatment vs. placebo\n(randomized assignment)",
    "Observational: blunt vs.\npenetrating trauma",
    "Groups differ in variance\nbut not in mean",
    "Paired data treated as\nindependent (wrong!)"
  ),
  Exchangeable = c("Yes", "Approximately", "No", "No"),
  `Permutation Valid?` = c("Yes (exact)", "Approximately", "No (inflated Type I)", "No (ignores structure)"),
  Recommendation = c(
    "Permutation valid",
    "Permutation OK;\nWelch t also reasonable",
    "Bootstrap test or Welch t",
    "Pair the data; use paired\nbootstrap or paired t-test"
  )
)

scenarios |>
  ggplot() +
  theme_void(base_size = 14) +
  annotate("text", x = 0.5, y = 0.97, fontface = "bold", size = 6,
           label = "Permutation Test: Validity by Setting") +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 0.93,
           fill = "grey97", color = "grey60") +
  annotate("text", x = c(0.1, 0.36, 0.6, 0.88), y = 0.90,
           label = c("Setting", "Exchangeable?", "Valid?", "Recommendation"),
           fontface = "bold", size = 4.8, hjust = 0.5) +
  {
    ys <- c(0.75, 0.56, 0.37, 0.18)
    cols <- c("#27ae60", "#f39c12", "#e74c3c", "#e74c3c")
    pmap(list(scenarios$Setting, scenarios$Exchangeable,
              scenarios$`Permutation Valid?`, scenarios$Recommendation,
              ys, cols), function(s, e, v, r, y, co) {
      list(
        annotate("text", x = 0.1, y = y, label = s, size = 4, hjust = 0.5, color = "grey20"),
        annotate("text", x = 0.36, y = y, label = e, size = 4, hjust = 0.5, color = co, fontface = "bold"),
        annotate("text", x = 0.60, y = y, label = v, size = 4, hjust = 0.5, color = co),
        annotate("text", x = 0.88, y = y, label = r, size = 4, hjust = 0.5, color = "grey30")
      )
    })
  } +
  xlim(0, 1) + ylim(0, 1)
```

## Permutation Test: Implementation

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
B_perm  <- 9999
all_los <- c(los_blunt, los_penet)
N       <- length(all_los)
t_obs   <- mean(los_blunt) - mean(los_penet)

perm_diffs <- replicate(B_perm, {
  idx <- sample(N)
  mean(all_los[idx[seq_len(n1)]]) - mean(all_los[idx[(n1 + 1):N]])
})

p_perm  <- mean(abs(perm_diffs) >= abs(t_obs))
p_welch <- t.test(los_blunt, los_penet)$p.value

cat(sprintf("Observed mean difference:   %.3f days\n", t_obs))
cat(sprintf("Permutation p-value:        %.4f\n",      p_perm))
cat(sprintf("Welch t-test p-value:       %.4f\n",      p_welch))
cat(sprintf("Bootstrap test p-value:     %.4f\n",      p_boot))
```

## Visualizing the Permutation Null Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6.5
#| fig-width: 13

tibble(perm_diff = perm_diffs) |>
  ggplot(aes(x = perm_diff)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "steelblue", color = "white", alpha = 0.65) +
  geom_density(color = "#2c3e50", linewidth = 1.1) +
  geom_vline(xintercept = c(-abs(t_obs), abs(t_obs)),
             color = "#e74c3c", linewidth = 1.5) +
  annotate("text", x = abs(t_obs) + 0.05, y = 0.65,
           label = bquote("|" * T[obs] * "| =" ~ .(round(abs(t_obs), 2))),
           color = "#e74c3c", size = 5.5, hjust = 0) +
  annotate("text", x = 0, y = 0.72,
           label = sprintf("p = %.4f", p_perm),
           size = 5.5, color = "#2c3e50") +
  labs(
    title    = "Permutation Null Distribution: Difference in Mean LOS",
    subtitle = "Red lines mark \u00b1|T_obs|. Fraction of permuted statistics beyond red = p-value.",
    x        = expression(bar(x)[1]^"*" - bar(x)[2]^"*" ~ "(days)"),
    y        = "Density"
  )
```

## Permutation Tests Work with Any Statistic

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Permutation test for difference in MEDIANS
t_obs_med <- median(los_blunt) - median(los_penet)
perm_med  <- replicate(9999, {
  idx <- sample(N)
  median(all_los[idx[seq_len(n1)]]) - median(all_los[idx[(n1 + 1):N]])
})
p_perm_med <- mean(abs(perm_med) >= abs(t_obs_med))

cat(sprintf("Median difference: %.3f   permutation p = %.4f\n",
            t_obs_med, p_perm_med))

# Permutation test for VARIANCE RATIO (vs F-test)
t_obs_var <- var(los_blunt) / var(los_penet)
perm_var  <- replicate(9999, {
  idx <- sample(N)
  var(all_los[idx[seq_len(n1)]]) / var(all_los[idx[(n1 + 1):N]])
})
p_perm_var <- mean(perm_var >= t_obs_var | perm_var <= 1 / t_obs_var)
p_ftest    <- var.test(los_blunt, los_penet)$p.value

cat(sprintf("Variance ratio:    %.3f   permutation p = %.4f   F-test p = %.4f\n",
            t_obs_var, p_perm_var, p_ftest))
```

. . .

The permutation test for the variance ratio is valid regardless of distribution shape, unlike the $F$-test, which requires normality.

# Bootstrap for Paired Data {background-color="#27ae60"}

## Paired Data: Setup and the Crossover Trial Example

**Paired data** arises when each subject provides two measurements:

-   Crossover trials: same patient receives both treatments (in random order)
-   Before-after studies: same patient measured at two time points
-   Matched cohort studies: pairs matched on confounders

The natural estimand is $\mu_D = E[D_i]$ where $D_i = X_{i,\text{post}} - X_{i,\text{pre}}$ (or Treatment $-$ Control).


```{r}
#| echo: true
#| code-fold: true

# True mean difference = -3 mmHg; SD of differences ≈ 10.5; n = 25
set.seed(551)
n_pair <- 25; rho_p <- 0.45
L_p    <- chol(5^2 * matrix(c(1, rho_p, rho_p, 1), 2))
xy_p   <- matrix(rnorm(2 * n_pair), n_pair, 2) %*% L_p +
          matrix(c(0, -3), n_pair, 2, byrow = TRUE)

bp_data <- tibble(
  patient   = 1:n_pair,
  placebo   = xy_p[, 1],
  drug_A    = xy_p[, 2],
  diff      = drug_A - placebo   # D_i = Drug - Placebo
)

cat(sprintf("Mean difference (Drug - Placebo): %.2f mmHg\n", mean(bp_data$diff)))
cat(sprintf("SD of differences:                %.2f mmHg\n", sd(bp_data$diff)))
cat(sprintf("Paired t-test p-value:            %.4f\n",
            t.test(bp_data$drug_A, bp_data$placebo, paired = TRUE)$p.value))
```

## Paired Bootstrap CI

For paired data, the bootstrap resamples **entire pairs** $(D_i)$ with replacement, which preserves the pairing structure.

**Algorithm:** for $b = 1, \ldots, B$:

1.  Draw $D_1^{*b}, \ldots, D_n^{*b}$ **with replacement** from $\{D_1, \ldots, D_n\}$
2.  Compute $\bar{D}^{*b} = \frac{1}{n}\sum D_i^{*b}$

**Percentile CI:** $\bigl(\bar{D}^*_{(\alpha/2)},\; \bar{D}^*_{(1-\alpha/2)}\bigr)$

No normality assumption on $D_i$.

```{r}
#| echo: true
#| code-fold: false

d_cross <- bp_data$diff

set.seed(551)
boot_D         <- replicate(5000, mean(sample(d_cross, n_pair, replace = TRUE)))
ci_boot_paired <- quantile(boot_D, c(0.025, 0.975))
ci_t_paired    <- t.test(d_cross)$conf.int

cat(sprintf("Observed mean difference:    %.3f mmHg\n", mean(d_cross)))
cat(sprintf("Bootstrap SE:                %.3f\n", sd(boot_D)))
cat(sprintf("Paired t-test  95%% CI: (%.3f, %.3f)\n",
            ci_t_paired[1], ci_t_paired[2]))
cat(sprintf("Bootstrap      95%% CI: (%.3f, %.3f)\n",
            ci_boot_paired[1], ci_boot_paired[2]))
```

With approximately normal differences and moderate $n$, the two CIs are very similar. The bootstrap shines when $D_i$ are clearly non-normal or $n$ is very small.

## Visualizing the Paired Bootstrap Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 15

p_ci <- tibble(boot_mean_D = boot_D) |>
  ggplot(aes(x = boot_mean_D)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#27ae60", color = "white", alpha = 0.7) +
  geom_density(color = "#2c3e50", linewidth = 1.1) +
  geom_vline(xintercept = mean(d_cross), color = "#e74c3c", linewidth = 1.5) +
  geom_vline(xintercept = ci_boot_paired, color = "#006a4e",
             linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = ci_t_paired, color = "#e74c3c",
             linewidth = 1.0, linetype = "dotted") +
  labs(
    title    = "Paired Bootstrap Distribution: Mean BP Reduction",
    subtitle = "Green dashed = bootstrap percentile CI\nRed dotted = paired t CI",
    x        = expression(bar(D)^"*" ~ "(mmHg)"),
    y        = "Density"
  )

p_pairs <- bp_data |>
  pivot_longer(c(placebo, drug_A), names_to = "Treatment", values_to = "BP_reduction") |>
  mutate(Treatment = factor(Treatment, levels = c("placebo", "drug_A"),
                            labels = c("Placebo", "Drug A"))) |>
  ggplot(aes(x = Treatment, y = BP_reduction, group = patient)) +
  geom_line(alpha = 0.3, color = "grey50") +
  geom_point(aes(color = Treatment), size = 3) +
  scale_color_manual(values = c("Placebo" = "#e74c3c", "Drug A" = "#27ae60")) +
  labs(title = "Crossover Trial: BP Reduction per Patient",
       x = NULL, y = "BP Reduction (mmHg)") +
  theme(legend.position = "none")

p_pairs + p_ci
```

## Paired Bootstrap Test

To **test** $H_0: \mu_D = 0$, we must shift the differences to satisfy $H_0$:

$$\tilde{D}_i = D_i - \bar{D}$$

Now $\{\tilde{D}_i\}$ have mean exactly 0 — $H_0$ is imposed. We then resample from the shifted differences.

**Algorithm:** for $b = 1, \ldots, B$:

1.  Draw $\tilde{D}_1^{*b}, \ldots, \tilde{D}_n^{*b}$ with replacement from $\{\tilde{D}_i\}$
2.  Compute $T^{*b} = \bar{\tilde{D}}^{*b}$

**p-value:** $\hat{p} = \frac{1}{B}\sum_{b=1}^B \mathbf{1}\bigl(|T^{*b}| \geq |\bar{D}_\text{obs}|\bigr)$

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
B_paired <- 9999
D_obs    <- mean(d_cross)

# Shift differences to mean 0 (imposing H0)
d_shifted <- d_cross - mean(d_cross)

cat(sprintf("Before shifting: mean(D) = %.4f\n", mean(d_cross)))
cat(sprintf("After  shifting: mean(D_shift) = %.6f\n", mean(d_shifted)))

# Resample from shifted differences
boot_null_D <- replicate(B_paired,
                         mean(sample(d_shifted, n_pair, replace = TRUE)))

p_boot_paired <- mean(abs(boot_null_D) >= abs(D_obs))
p_t_paired    <- t.test(d_cross)$p.value

cat(sprintf("\nObserved mean difference:       %.3f mmHg\n", D_obs))
cat(sprintf("Bootstrap test p-value:         %.4f\n",       p_boot_paired))
cat(sprintf("Paired t-test p-value:          %.4f\n",       p_t_paired))
```

## Paired Bootstrap Null vs. Sampling Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14

tibble(
  value = c(boot_null_D, boot_D),
  type  = rep(c("Null distribution (shifted, for testing)",
                "Sampling distribution (for CI)"),
              times = c(length(boot_null_D), length(boot_D)))
) |>
  ggplot(aes(x = value, fill = type)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 color = "white", alpha = 0.7, position = "identity") +
  geom_vline(xintercept = D_obs, color = "#e74c3c", linewidth = 1.5) +
  geom_vline(xintercept = -abs(D_obs), color = "#e74c3c",
             linewidth = 1.5, linetype = "dashed") +
  scale_fill_manual(values = c(
    "Null distribution (shifted, for testing)"  = "#e74c3c",
    "Sampling distribution (for CI)"            = "#27ae60"
  )) +
  annotate("text", x = D_obs + 0.3, y = 0.28,
           label = bquote(bar(D)[obs] == .(round(D_obs, 2))),
           color = "#e74c3c", size = 5.5, hjust = 0) +
  labs(
    title    = "Paired Bootstrap: Null Distribution vs. Sampling Distribution",
    subtitle = "Red = shifted to H0 (for p-value)  |  Green = resampled as-is (for CI)",
    x        = expression(bar(D)^"*" ~ "(mmHg)"),
    y        = "Density",
    fill     = NULL
  ) +
  theme(legend.position = "bottom")
```

::: callout-note
The null distribution (red) is centered at $0$ — it was shifted to impose $H_0: \mu_D = 0$. The sampling distribution (green) is centered at $\bar{D}_\text{obs}$ and is used for the CI, not the test. Because the true effect is modest (~$-3$ mmHg with substantial within-subject variability), the two distributions overlap considerably, mirroring the borderline p-value.
:::

# The `infer` Package {background-color="#1a5276"}

## The `infer` Package: Tidy Resampling Inference

The `infer` package (part of the `tidymodels` ecosystem) provides a **unified, pipe-friendly framework** for hypothesis testing and interval estimation via simulation.

The workflow follows four consistent verbs:

| Step | Verb | What it does |
|---|---|---|
| 1 | `specify()` | Declare the response (and optional explanatory) variable |
| 2 | `hypothesize()` | State the null hypothesis (for tests only) |
| 3 | `generate()` | Generate bootstrap samples or permutations |
| 4 | `calculate()` | Compute the statistic of interest |

Then use `get_p_value()`, `get_confidence_interval()`, or `visualize()` to summarize.

. . .

::: callout-note
## Why Use `infer`?
- Consistent syntax across one-sample, two-sample, and paired settings
- Integrates naturally with `dplyr` pipelines and tidy data
- Makes the conceptual steps explicit: specify → hypothesize → generate → calculate
- Excellent for teaching and reproducible research workflows
:::

## Setting Up the Data for `infer`

The `infer` package requires **tidy (long) data** — one row per observation.

```{r}
#| echo: true
#| code-fold: false

# Two-sample LOS data in tidy format
los_tidy <- tibble(
  LOS   = c(los_blunt, los_penet),
  Group = rep(c("Blunt", "Penetrating"),
              c(length(los_blunt), length(los_penet)))
)

los_tidy |> slice_sample(n = 6)

# Paired blood pressure data in tidy format
bp_tidy <- bp_data |>
  select(patient, diff)   # use the already-computed differences

bp_tidy |> slice_head(n = 6)
```

## Two-Sample Permutation Test with `infer`

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Step 1-2: Specify and hypothesize
# Step 3: Generate permutation null distribution
null_dist_perm <- los_tidy |>
  specify(LOS ~ Group) |>
  hypothesize(null = "independence") |>
  generate(reps = 9999, type = "permute") |>
  calculate(stat = "diff in means", order = c("Blunt", "Penetrating"))

# Observed statistic
obs_stat <- los_tidy |>
  specify(LOS ~ Group) |>
  calculate(stat = "diff in means", order = c("Blunt", "Penetrating"))

obs_stat

# p-value
pval_perm <- get_p_value(null_dist_perm, obs_stat, direction = "two-sided")
pval_perm
```

## Visualizing the `infer` Permutation Distribution

```{r}
#| echo: true
#| code-fold: false
#| fig-height: 5.5
#| fig-width: 12

visualize(null_dist_perm) +
  shade_p_value(obs_stat, direction = "two-sided") +
  labs(
    title    = "Permutation Null Distribution (via infer)",
    subtitle = sprintf("p-value = %.4f  |  Observed difference = %.3f days",
                       pval_perm$p_value, obs_stat$stat),
    x        = "Difference in Mean LOS (Blunt - Penetrating, days)",
    y        = "Count"
  )
```

## Two-Sample Bootstrap CI with `infer`

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Bootstrap sampling distribution (no hypothesize step)
boot_dist_ci <- los_tidy |>
  specify(LOS ~ Group) |>
  generate(reps = 5000, type = "bootstrap") |>
  calculate(stat = "diff in means", order = c("Blunt", "Penetrating"))

# Percentile confidence interval
ci_infer <- get_confidence_interval(boot_dist_ci,
                                    level = 0.95,
                                    type  = "percentile")
ci_infer

# SE-based CI
ci_infer_se <- get_confidence_interval(boot_dist_ci,
                                       level = 0.95,
                                       type  = "se",
                                       point_estimate = obs_stat)
ci_infer_se
```

## Visualizing the `infer` Bootstrap CI

```{r}
#| echo: true
#| code-fold: false
#| fig-height: 5.5
#| fig-width: 12

visualize(boot_dist_ci) +
  shade_confidence_interval(ci_infer) +
  geom_vline(xintercept = obs_stat$stat,
             color = "#e74c3c", linewidth = 1.5) +
  labs(
    title    = "Bootstrap Sampling Distribution with 95% CI (via infer)",
    subtitle = sprintf("Percentile CI: (%.3f, %.3f)  |  Observed = %.3f",
                       ci_infer$lower_ci, ci_infer$upper_ci, obs_stat$stat),
    x        = "Difference in Mean LOS (Blunt - Penetrating, days)",
    y        = "Count"
  )
```

## Two-Sample Bootstrap Test with `infer`: A Limitation

::: callout-warning
## `infer` does NOT implement the shift-bootstrap for two-sample tests

`generate(type = "bootstrap")` with `hypothesize(null = "independence")` resamples each group **separately with replacement** — exactly the same as the CI bootstrap. The null distribution is centered at the **observed difference**, not at 0. Calling `get_p_value()` on this distribution is essentially asking "what fraction of bootstrap samples are at least as extreme as the observed value?" — against a distribution centered *at* the observed value — which returns a meaningless p-value near 0.5 or higher.

The **shift-bootstrap** (C&H Ch. 8) must be implemented manually (as shown earlier in these slides). For two-sample hypothesis testing in `infer`, always use `type = "permute"`.
:::

```{r}
#| echo: true
#| code-fold: false

# Demonstrate the issue: bootstrap + hypothesize gives wrong null distribution
set.seed(551)
null_dist_boot_wrong <- los_tidy |>
  specify(LOS ~ Group) |>
  hypothesize(null = "independence") |>
  generate(reps = 9999, type = "bootstrap") |>
  calculate(stat = "diff in means", order = c("Blunt", "Penetrating"))

cat(sprintf("Center of bootstrap 'null' distribution: %.3f\n",
            mean(null_dist_boot_wrong$stat)))
cat(sprintf("  (should be ~0 under H0; instead equals observed diff)\n"))
cat(sprintf("Observed difference:                     %.3f\n", obs_stat$stat))
cat(sprintf("'p-value' from this (invalid):           %.4f\n",
            get_p_value(null_dist_boot_wrong, obs_stat,
                        direction = "two-sided")$p_value))
cat(sprintf("Permutation p-value (correct):           %.4f\n", pval_perm$p_value))
```

## Paired Bootstrap with `infer`

For paired data, work directly with the **differences** as a one-sample problem:

```{r}
#| echo: true
#| code-fold: false

set.seed(551)

# Bootstrap CI for mean difference (one-sample on differences)
boot_paired_infer <- bp_tidy |>
  specify(response = diff) |>
  generate(reps = 5000, type = "bootstrap") |>
  calculate(stat = "mean")

ci_paired_infer <- get_confidence_interval(boot_paired_infer,
                                           level = 0.95,
                                           type  = "percentile")

# Bootstrap test: H0: mu_D = 0
obs_mean_D <- bp_tidy |>
  specify(response = diff) |>
  calculate(stat = "mean")

null_paired_infer <- bp_tidy |>
  specify(response = diff) |>
  hypothesize(null = "point", mu = 0) |>
  generate(reps = 9999, type = "bootstrap") |>
  calculate(stat = "mean")

pval_paired_infer <- get_p_value(null_paired_infer, obs_mean_D, direction = "two-sided")

cat(sprintf("Observed mean difference:   %.3f mmHg\n", obs_mean_D$stat))
cat(sprintf("Bootstrap CI (percentile):  (%.3f, %.3f)\n",
            ci_paired_infer$lower_ci, ci_paired_infer$upper_ci))
cat(sprintf("Bootstrap test p-value:     %.4f\n", pval_paired_infer$p_value))
cat(sprintf("Paired t-test p-value:      %.4f\n", p_t_paired))
```

## `infer` Workflow Summary

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5.5
#| fig-width: 14

summary_tbl <- tribble(
  ~Goal,                        ~`specify()`,          ~`hypothesize()`,             ~`generate()`,             ~`calculate()`,
  "Two-sample CI",              "Y ~ Group",           "(none)",                     "bootstrap",               "diff in means/medians",
  "Two-sample perm. test",      "Y ~ Group",           "independence",               "permute",                 "diff in means/medians",
  "Paired CI",                  "response = diff",     "(none)",                     "bootstrap",               "mean",
  "Paired boot. test",          "response = diff",     "point, mu = 0",              "bootstrap",               "mean"
)

summary_tbl |>
  knitr::kable(format = "html") |>
  kableExtra::kable_styling(font_size = 18, full_width = TRUE,
                            bootstrap_options = c("striped", "hover"))
```

. . .

::: callout-tip
## When to Choose What

- **`type = "permute"`:** Two-sample hypothesis testing. Generates the null distribution by relabeling groups — the only correct `infer` approach for two-sample tests.
- **`type = "bootstrap"` without `hypothesize()`:** Constructing CIs (sampling distribution centered at observed statistic).
- **`type = "bootstrap"` with `hypothesize(null = "point")`:** One-sample bootstrap tests (shifts distribution to $H_0$ value). Works correctly for one-sample settings only.
- **Two-sample shift-bootstrap test:** Must be coded manually (see earlier section) — `infer` does not support it.
:::

# Comparing Methods: Simulation {background-color="#9b59b6"}

## Type I Error and Power: Welch t-test vs. Permutation Test

The permutation test derives its null distribution empirically rather than from the $t$-distribution. For the **mean difference** statistic, how much does it matter?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 15

set.seed(551)
B_sim  <- 3000
B_perm <- 499

sim_scenario <- function(dist_fn, n = 10, effect = 0) {
  replicate(B_sim, {
    x1  <- dist_fn(n)
    x2  <- dist_fn(n) + effect
    Ns  <- 2 * n
    all <- c(x1, x2)
    t0  <- mean(x1) - mean(x2)
    p_welch <- t.test(x1, x2)$p.value
    p_perm  <- {
      perms <- replicate(B_perm, {
        idx <- sample(Ns)
        mean(all[idx[seq_len(n)]]) - mean(all[idx[(n + 1):Ns]])
      })
      mean(abs(perms) >= abs(t0))
    }
    c(Welch = p_welch, Permutation = p_perm)
  }) |> t() |> as_tibble() |>
    summarise(across(everything(), \(x) mean(x < 0.05)))
}

norm_fn <- \(n) rnorm(n, 0, 1)
skew_fn <- \(n) rchisq(n, df = 2) - 2

results <- bind_rows(
  bind_rows(
    sim_scenario(norm_fn, n = 10, effect = 0.0) |> mutate(Scenario = "Type I Error (effect = 0)"),
    sim_scenario(norm_fn, n = 10, effect = 1.0) |> mutate(Scenario = "Power (effect = 1 SD)")
  ) |> mutate(Distribution = "Normal(0,1) \u2014 symmetric"),
  bind_rows(
    sim_scenario(skew_fn, n = 10, effect = 0.0) |> mutate(Scenario = "Type I Error (effect = 0)"),
    sim_scenario(skew_fn, n = 10, effect = 1.0) |> mutate(Scenario = "Power (effect = 1 SD)")
  ) |> mutate(Distribution = "chi-sq(2) shifted \u2014 heavily right-skewed")
)

results |>
  pivot_longer(c(Welch, Permutation),
               names_to = "Method", values_to = "Rate") |>
  mutate(
    Scenario     = factor(Scenario,
                          levels = c("Type I Error (effect = 0)", "Power (effect = 1 SD)")),
    Distribution = factor(Distribution,
                          levels = c("Normal(0,1) \u2014 symmetric",
                                     "chi-sq(2) shifted \u2014 heavily right-skewed"))
  ) |>
  ggplot(aes(x = Method, y = Rate, fill = Method)) +
  geom_col(width = 0.5) +
  geom_hline(
    data = tibble(
      Scenario = factor("Type I Error (effect = 0)",
                        levels = c("Type I Error (effect = 0)", "Power (effect = 1 SD)")),
      y = 0.05
    ),
    aes(yintercept = y),
    linetype = "dashed", linewidth = 1.2, color = "black"
  ) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 0.1)),
            vjust = -0.4, size = 4.5, fontface = "bold") +
  facet_grid(Distribution ~ Scenario) +
  scale_fill_manual(values = c("Welch" = "#006a4e", "Permutation" = "#e74c3c")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.95)) +
  labs(
    title    = "Welch t-test vs. Permutation Test  (n = 10 per group, 3000 simulations)",
    subtitle = "Both methods control Type I error near 5% and have similar power \u2014 permutation's advantage is flexibility.",
    x = NULL, y = "Rejection Rate", fill = NULL
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12))
```

::: callout-note
The two-sided Welch $t$-test is **robust to skewness for Type I error** even at small $n$: skewness makes each tail individually wrong, but in a two-sided test the errors largely cancel. Permutation's advantage is **generality**: it works with *any* test statistic, not just means.
:::

# Key Takeaways {background-color="#2c3e50"}

## Key Takeaways (C&H Ch. 3, 5, 8; DBC 10.6)

::: callout-important
## Bootstrap and Permutation for Two Samples

1.  **Two-sample bootstrap CI:** Resample each group **independently** with replacement; construct percentile, basic, or SE CI. Works for any statistic. No distributional assumption.

2.  **Two-sample bootstrap test:** Shift both samples to satisfy $H_0$ (equal grand mean), then resample to generate null distribution. Use when permutation assumptions are not met.

3.  **Permutation test:** Pool all $N$ observations, **relabel** groups at random; fraction of permuted statistics $\geq |T_\text{obs}|$ gives a valid p-value under $H_0: F_1 = F_2$ (**exchangeability**). Most natural for randomized experiments.

4.  **Permutation validity:** Requires **exchangeability** — groups can only differ in label, not in variance or shape, under $H_0$. Fails for paired/clustered data and when groups differ on nuisance parameters.

5.  **Paired bootstrap CI:** Resample **whole pairs/differences** to preserve pairing; CI for $\mu_D$ with no normality assumption.

6.  **Paired bootstrap test:** Shift differences to mean 0 (imposing $H_0: \mu_D = 0$), then resample; gives valid p-value without normality.

7.  **`infer` package:** Unified specify → hypothesize → generate → calculate pipeline. Use `type = "permute"` for two-sample tests and `type = "bootstrap"` for CIs and one-sample tests. The shift-bootstrap for two-sample testing must be coded manually.
:::

## Method Decision Guide

```{r}
#| echo: false
#| fig-height: 6
#| fig-width: 14

decision_tbl <- tribble(
  ~Setting,              ~Estimand,        ~CI_Method,            ~Test_Method,
  "Two independent\nsamples (any stat)", "Any\n(mean, median, SD ratio)", "Two-sample\nbootstrap", "Permutation (if\nexchangeable)\nor bootstrap test",
  "Two independent\nsamples (mean only)", expression(mu[1]-mu[2]),    "Welch t or\nbootstrap",     "Welch t or\npermutation",
  "Paired / matched\ndata",              expression(mu[D]),           "Paired t or\npaired bootstrap", "Paired t or\npaired bootstrap test",
  "Non-normal,\nsmall n",               "Any",                    "Bootstrap\n(percentile)", "Bootstrap\ntest"
)

tibble(
  x     = c(1, 2, 3, 4),
  label = c("Two independent\nsamples\n(any statistic)",
            "Two independent\nsamples\n(mean only)",
            "Paired /\nmatched data",
            "Non-normal\nor small n"),
  ci    = c("Two-sample\nbootstrap CI",
            "Welch t CI or\nbootstrap CI",
            "Paired t CI or\npaired bootstrap CI",
            "Bootstrap CI\n(percentile)"),
  test  = c("Permutation test\n(if exchangeable)\nor bootstrap test",
            "Welch t-test or\npermutation test",
            "Paired t-test or\npaired bootstrap test",
            "Bootstrap test")
) |>
  ggplot() +
  theme_void(base_size = 15) +
  # Setting boxes
  geom_tile(aes(x = x, y = 3), width = 0.85, height = 0.75,
            fill = "#2c3e50", color = "white") +
  geom_text(aes(x = x, y = 3, label = label), color = "white",
            size = 4.2, lineheight = 1.1) +
  # CI row
  geom_tile(aes(x = x, y = 1.8), width = 0.85, height = 0.75,
            fill = "#3498db", color = "white") +
  geom_text(aes(x = x, y = 1.8, label = ci), color = "white",
            size = 4, lineheight = 1.1) +
  # Test row
  geom_tile(aes(x = x, y = 0.6), width = 0.85, height = 0.75,
            fill = "#e74c3c", color = "white") +
  geom_text(aes(x = x, y = 0.6, label = test), color = "white",
            size = 4, lineheight = 1.1) +
  # Row labels
  annotate("text", x = -0.05, y = 3,   label = "Setting",     fontface = "bold", size = 5, color = "#2c3e50", hjust = 1) +
  annotate("text", x = -0.05, y = 1.8, label = "CI Method",   fontface = "bold", size = 5, color = "#3498db", hjust = 1) +
  annotate("text", x = -0.05, y = 0.6, label = "Test Method", fontface = "bold", size = 5, color = "#e74c3c", hjust = 1) +
  xlim(-1.1, 4.5) + ylim(0, 3.5) +
  labs(title = "Method Decision Guide for Two-Sample Inference")
```

## Looking Ahead

**Next class (Lesson 21): Nonparametric Inference (DBC Ch. 14)**

-   Sign test: based on the binomial distribution, zero distributional assumptions
-   Wilcoxon signed-rank test: uses ranks of $|D_i|$, requires symmetric differences
-   Wilcoxon rank-sum test (Mann–Whitney): independent samples, ranks all $N$ observations
-   Asymptotic relative efficiency vs. parametric counterparts

## References

::: nonincremental
-   Chihara, L. M. and Hesterberg, T. C. *Mathematical Statistics with Resampling and R* (Wiley). Chapters 3, 5, and 8.
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 10.6.
-   Couch, S., Bray, A., et al. *infer: An R package for tidyverse-friendly statistical inference*. JOSS, 2021. <https://infer.tidymodels.org>
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/20_BootstrapPermutation/20_BootstrapPermutation.html lessons/20_BootstrapPermutation/20_BootstrapPermutation.pdf
```
