---
title: "Lesson 16: Further Aspects of Hypothesis Testing; Relationship between Testing and Interval Estimation"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w8d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 16 - Further Aspects of Hypothesis Testing; CI-Test Duality
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 16: Further Aspects of Hypothesis Testing {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 15: Neyman-Pearson Lemma; Likelihood Ratio Tests**

- The **Neyman-Pearson Lemma** shows the likelihood ratio test is most powerful for simple vs. simple hypotheses
- **UMP tests** exist for one-sided alternatives in exponential families (Karlin-Rubin Theorem)
- The **Likelihood Ratio Test (LRT)** $\Lambda = L(\hat{\theta}_0)/L(\hat{\theta}_{mle})$ provides a general recipe for composite hypotheses
- **Wilks' Theorem:** $-2\ln(\Lambda) \xrightarrow{d} \chi^2_m$ under $H_0$ (large-sample)
- Classic tests (z, t, F) are all LRTs

. . .

**Today's Goals:**

1. Understand the distinction between **statistical significance** and **practical/clinical significance**
2. State and use the **duality theorem** connecting confidence intervals and hypothesis tests
3. Construct **confidence intervals by inverting tests** (the general method)
4. Understand **one-sided CIs** and their connection to one-sided tests
5. Use $p$-value duality and **equivalence testing** in medical research contexts
6. Understand **multiple testing** consequences for CI coverage

## Roadmap

| Topic | Key Idea | Textbook |
|-------|----------|----------|
| Statistical vs. practical significance | A significant result may not be meaningful | Devore 9.6 |
| Duality Theorem | $\theta_0 \in$ CI iff $H_0: \theta = \theta_0$ is not rejected | Devore 9.6 |
| Inverting a test | General method for constructing CIs | Devore 9.6 |
| One-sided CIs | Linked to one-sided tests at level $\alpha$ | Devore 9.6 |
| Equivalence testing | Testing that effects are *small enough* | Applied |
| Multiple testing & CIs | Simultaneous coverage | Devore 9.6 |

. . .

::: callout-note
## Why This Matters
The duality between hypothesis tests and confidence intervals is one of the most elegant unifying results in statistics. It tells us that a confidence interval is **not just a range of plausible values** — it is a compact summary of all hypothesis tests we would fail to reject. In clinical trials, this connection is essential for both planning and reporting.
:::

# Statistical Significance vs. Practical Significance (Devore 9.6) {background-color="#3498db"}

## The Core Problem

We have spent much of this course learning how to **detect** an effect. But detection is only half the story.

. . .

::: callout-important
## Key Distinction
- **Statistical significance:** $p < \alpha$ — the data provide enough evidence to reject $H_0$
- **Practical (clinical) significance:** the effect size is **large enough to matter** in a real-world or clinical context
:::

. . .

These two concepts can **diverge dramatically**:

- **Large $n$, tiny effect:** Almost any negligible difference becomes statistically significant
- **Small $n$, large effect:** Clinically important effects may be missed

. . .

::: callout-tip
## Medical Context
Suppose a new antihypertensive reduces systolic BP by 1 mmHg on average ($n = 50{,}000$, $p < 0.0001$). This is statistically significant — but is a 1 mmHg reduction clinically meaningful? The American Heart Association typically considers reductions of $\geq 5$ mmHg clinically relevant.
:::

## Large Sample Sizes Make Everything Significant

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 14

# Simulate: true effect = delta (small), vary n
# Test H0: mu = 0 vs Ha: mu != 0
sim_pvals <- function(n, delta = 0.5, sigma = 10, B = 5000) {
  replicate(B, {
    x <- rnorm(n, mean = delta, sd = sigma)
    t.test(x, mu = 0)$p.value
  })
}

sample_sizes <- c(30, 100, 500, 2000, 10000)
results <- map_dfr(sample_sizes, function(n) {
  pvals <- sim_pvals(n, delta = 0.5, sigma = 10)
  tibble(n = n, power = mean(pvals < 0.05))
})

results |>
  ggplot(aes(x = n, y = power)) +
  geom_line(color = "#006a4e", linewidth = 1.5) +
  geom_point(size = 4, color = "#006a4e") +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "gray50") +
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = expression("Power to Detect " * delta * " = 0.5 mmHg (SD = 10) at " * alpha * " = 0.05"),
    subtitle = "A clinically trivial effect becomes detectable with enough patients",
    x = "Sample size (log scale)", y = "Power"
  )
```

## Visualizing Statistical vs. Clinical Significance

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

# Simulate 4 scenarios: large/small n x large/small effect
scenarios <- tribble(
  ~n,     ~delta,  ~label,
  50,     8,       "Small n, Large effect",
  50,     1,       "Small n, Small effect",
  5000,   8,       "Large n, Large effect",
  5000,   1,       "Large n, Small effect"
)

set.seed(551)
scenario_results <- scenarios |>
  rowwise() |>
  mutate(
    data = list(rnorm(n, mean = delta, sd = 10)),
    test = list(t.test(data, mu = 0)),
    estimate = test$estimate,
    ci_low  = test$conf.int[1],
    ci_high = test$conf.int[2],
    pval = test$p.value,
    sig = pval < 0.05
  ) |>
  ungroup()

scenario_results |>
  mutate(
    label = factor(label, levels = scenarios$label),
    sig_label = ifelse(sig, "Statistically significant", "Not significant")
  ) |>
  ggplot(aes(x = label, y = estimate, color = sig_label)) +
  geom_point(size = 5) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2, linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = 5, linetype = "dashed", color = "#c0392b", linewidth = 1) +
  annotate("text", x = 4.5, y = 5.5, label = "Clinical threshold = 5 mmHg",
           color = "#c0392b", hjust = 1, size = 5) +
  scale_color_manual(values = c("Statistically significant" = "#006a4e",
                                "Not significant" = "#e74c3c")) +
  labs(
    title = "95% Confidence Intervals for Mean BP Reduction (mmHg)",
    subtitle = "Statistical significance depends on n; clinical significance depends on effect size",
    x = NULL, y = "Estimated mean reduction (mmHg)",
    color = NULL
  ) +
  coord_flip() +
  theme(legend.position = "bottom")
```

## Reporting Standards: What Should We Report?

::: callout-important
## Best Practice (Devore 9.6 & Medical Journals)
Always report:

1. **The test statistic** (e.g., $t = 3.41$)
2. **The $p$-value** (e.g., $p = 0.001$)
3. **The confidence interval** for the effect size (e.g., 95% CI: [2.1, 8.7] mmHg)
4. **The effect size** with clinical context
:::

. . .

Why CIs are essential:

- A CI tells you **the range of plausible effects**, not just whether an effect exists
- A CI directly encodes practical significance: does the entire interval fall below the clinical threshold?
- A narrow CI near zero can rule out clinically meaningful effects even without rejecting $H_0$

. . .

::: callout-note
## Example: CONSORT Guidelines
Major medical journals (NEJM, JAMA, Lancet) following CONSORT guidelines require authors to report effect sizes with CIs, not just $p$-values. The statement "$p < 0.05$" alone is considered insufficient.
:::

## Effect Size Measures

Beyond the raw difference, **effect size** quantifies practical significance on a standardized scale.

. . .

**Cohen's $d$** (for means, $\sigma$ known or estimated):

$$d = \frac{\mu_1 - \mu_2}{\sigma}$$

| $|d|$ | Interpretation |
|-------|----------------|
| $< 0.2$ | Negligible |
| $0.2 - 0.5$ | Small |
| $0.5 - 0.8$ | Medium |
| $> 0.8$ | Large |

. . .

**Number Needed to Treat (NNT)** for binary outcomes:

$$\text{NNT} = \frac{1}{p_{\text{control}} - p_{\text{treatment}}}$$

Interpretation: How many patients must be treated to prevent one additional event?

## Example: Statin Therapy for MI Prevention

A landmark trial finds statins reduce the absolute risk of MI from 4.5% to 3.2% over 5 years.

$$\text{NNT} = \frac{1}{0.045 - 0.032} = \frac{1}{0.013} \approx 77$$

. . .

```{r}
#| echo: true
p_control <- 0.045
p_treatment <- 0.032
RR  <- p_treatment / p_control
ARR <- p_control - p_treatment
NNT <- 1 / ARR

tibble(
  Measure = c("Risk (control)", "Risk (treatment)", 
              "Absolute Risk Reduction (ARR)", 
              "Relative Risk (RR)", 
              "Number Needed to Treat (NNT)"),
  Value = c(p_control, p_treatment, ARR, RR, NNT)
) |>
  mutate(Value = round(Value, 3)) |>
  knitr::kable(align = "lr")
```

. . .

Is an NNT of 77 clinically meaningful? This depends on the cost, side-effect profile, and severity of the prevented outcome.

# Duality of Hypothesis Tests and Confidence Intervals {background-color="#e74c3c"}

## The Duality Theorem

::: callout-important
## Theorem: CI-Test Duality (Devore 9.6)

Let $C(\mathbf{X})$ be a $100(1-\alpha)\%$ confidence interval for a parameter $\theta$, obtained by inverting a level-$\alpha$ test. Then:

$$\theta_0 \notin C(\mathbf{X}) \quad \iff \quad \text{Reject } H_0: \theta = \theta_0 \text{ at level } \alpha$$

$$\theta_0 \in C(\mathbf{X}) \quad \iff \quad \text{Fail to reject } H_0: \theta = \theta_0 \text{ at level } \alpha$$
:::

. . .

**Consequence:** A confidence interval is a **compact summary of all hypothesis tests we would fail to reject** at level $\alpha$.

. . .

::: callout-tip
## Intuition
A 95% CI for $\mu$ contains exactly those values of $\mu_0$ for which the data are "consistent" (i.e., would not be rejected at $\alpha = 0.05$).
:::

## Proof of Duality

Consider testing $H_0: \theta = \theta_0$ vs. $H_a: \theta \neq \theta_0$ using a two-sided level-$\alpha$ test.

. . .

For the one-sample z-test (known $\sigma$), reject $H_0$ when:

$$|Z| = \left|\frac{\bar{X} - \theta_0}{\sigma/\sqrt{n}}\right| > z_{\alpha/2}$$

. . .

Equivalently, fail to reject when:

$$\left|\bar{X} - \theta_0\right| \leq z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$

$$\bar{X} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}} \leq \theta_0 \leq \bar{X} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$$

. . .

This is exactly the **95% confidence interval** $\bar{X} \pm z_{\alpha/2}\sigma/\sqrt{n}$!

So: $\theta_0$ is in the CI $\iff$ we fail to reject $H_0: \theta = \theta_0$. $\square$

## The General Method: Inverting a Test

The duality theorem gives us a general method to construct CIs from any test.

. . .

::: callout-important
## Definition: Acceptance Region and Confidence Set (Devore 9.6)

For a level-$\alpha$ test of $H_0: \theta = \theta_0$, the **acceptance region** $A(\theta_0) = R^C(\theta_0)$ consists of all sample outcomes that do NOT lead to rejection.

The **confidence set** is:

$$C(\mathbf{X}) = \{\theta_0 : \mathbf{X} \in A(\theta_0)\}$$

This set has coverage probability $\geq 1 - \alpha$ for all $\theta$.
:::

. . .

**Algorithm for inverting a test:**

1. For each candidate value $\theta_0$, determine the acceptance region $A(\theta_0)$
2. The CI is all values $\theta_0$ for which $\mathbf{X} \in A(\theta_0)$

## Example: Inverting the t-Test

Let $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with both $\mu, \sigma^2$ unknown.

Test $H_0: \mu = \mu_0$ vs. $H_a: \mu \neq \mu_0$ using $T = \frac{\bar{X} - \mu_0}{S/\sqrt{n}}$.

. . .

**Acceptance region** at level $\alpha$:

$$A(\mu_0) = \left\{(x_1,\ldots,x_n): -t_{\alpha/2,\, n-1} \leq \frac{\bar{x} - \mu_0}{s/\sqrt{n}} \leq t_{\alpha/2,\, n-1}\right\}$$

. . .

**Invert:** Find all $\mu_0$ such that $\bar{x}$ falls in $A(\mu_0)$:

$$-t_{\alpha/2,n-1} \leq \frac{\bar{x} - \mu_0}{s/\sqrt{n}} \leq t_{\alpha/2,n-1}$$

$$\bar{x} - t_{\alpha/2,n-1}\frac{s}{\sqrt{n}} \leq \mu_0 \leq \bar{x} + t_{\alpha/2,n-1}\frac{s}{\sqrt{n}}$$

. . .

This recovers the familiar **one-sample t confidence interval**. The CI and t-test are **exactly dual** to each other.

## Visualizing Duality: Clinical Trial Example

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 9
#| fig-width: 14

# Observed data: blood pressure reduction (mmHg)
set.seed(42)
n <- 25
bp_data <- rnorm(n, mean = 6.5, sd = 8)

xbar <- mean(bp_data)
s    <- sd(bp_data)
se   <- s / sqrt(n)
df   <- n - 1
t_crit <- qt(0.975, df)

# 95% CI
ci_low  <- xbar - t_crit * se
ci_high <- xbar + t_crit * se

# For each null value mu0, compute t-statistic and p-value
test_grid <- tibble(mu0 = seq(-5, 15, by = 0.1)) |>
  mutate(
    t_stat = (xbar - mu0) / se,
    pval   = 2 * pt(-abs(t_stat), df = df),
    reject = pval < 0.05
  )

p1 <- test_grid |>
  ggplot(aes(x = mu0, y = pval, color = reject)) +
  geom_line(linewidth = 1.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = c(ci_low, ci_high), 
             linetype = "dotted", color = "#006a4e", linewidth = 1.2) +
  annotate("rect", xmin = ci_low, xmax = ci_high, ymin = -Inf, ymax = Inf,
           fill = "#006a4e", alpha = 0.1) +
  annotate("text", x = xbar, y = 0.75,
           label = sprintf("95%% CI:\n[%.2f, %.2f]", ci_low, ci_high),
           color = "#006a4e", size = 5) +
  scale_color_manual(
    values = c("TRUE" = "#e74c3c", "FALSE" = "#006a4e"),
    labels = c("TRUE" = expression("Reject " * H[0]),
               "FALSE" = expression("Fail to reject " * H[0]))
  ) +
  labs(
    title = expression("p-value as a function of " * mu[0]),
    subtitle = expression("Green region = 95% CI = values of " * mu[0] * " we fail to reject"),
    x = expression(mu[0] * " (null value)"),
    y = "p-value",
    color = NULL
  ) +
  theme(legend.position = "bottom",
        plot.title    = element_text(size = 15),
        plot.subtitle = element_text(size = 13))

p2 <- test_grid |>
  ggplot(aes(x = mu0, y = t_stat)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_hline(yintercept = c(-t_crit, t_crit), linetype = "dashed",
             color = "#e74c3c", linewidth = 1) +
  geom_vline(xintercept = c(ci_low, ci_high),
             linetype = "dotted", color = "#006a4e", linewidth = 1.2) +
  annotate("rect", xmin = ci_low, xmax = ci_high, ymin = -Inf, ymax = Inf,
           fill = "#006a4e", alpha = 0.1) +
  annotate("text", x = -3.5, y = t_crit + 0.5,
           label = bquote(t[0.975*","*.(df)] == .(round(t_crit, 2))),
           color = "#e74c3c", size = 4.5) +
  labs(
    title = expression("t-statistic as a function of " * mu[0]),
    subtitle = expression("Fail to reject when |t| <= t"[crit] * " (green region)"),
    x = expression(mu[0] * " (null value)"),
    y = "t-statistic"
  ) +
  theme(plot.title    = element_text(size = 15),
        plot.subtitle = element_text(size = 13))

p1 + p2
```

## The p-Value Function

The plot on the previous slide illustrates the **p-value function** (or **significance function**):

$$p(\mu_0) = P_{H_0: \mu = \mu_0}\left(|T| \geq |t_{\text{obs}}|\right)$$

. . .

Key properties:

- $p(\mu_0) \geq \alpha$ exactly when $\mu_0$ is in the $100(1-\alpha)\%$ CI
- Maximized at the MLE: $p(\bar{x}) = 1$
- Shape reflects the curvature of the likelihood: narrow when data are highly informative

. . .

::: callout-note
## Reading the p-Value Function
- The **width** of the region where $p(\mu_0) \geq 0.05$ equals the **width of the 95% CI**
- The **shape** tells us how evidence accumulates against each null value
- This function is more informative than a single p-value or a single CI
:::

# One-Sided Tests and One-Sided Confidence Intervals {background-color="#8e44ad"}

## One-Sided Tests and Bounds

The duality extends naturally to **one-sided** settings.

. . .

::: callout-important
## Theorem: One-Sided Duality (Devore 9.6)

**Upper-tailed test:** $H_0: \theta \leq \theta_0$ vs $H_a: \theta > \theta_0$ at level $\alpha$

Reject when $T > c_\alpha$

$\Leftrightarrow$ The corresponding $100(1-\alpha)\%$ **lower confidence bound** $\underline{\theta}$ satisfies:

$$\theta_0 < \underline{\theta}(\mathbf{X}) \quad \iff \quad \text{Reject } H_0$$

**Lower-tailed test:** $H_0: \theta \geq \theta_0$ vs $H_a: \theta < \theta_0$

$\Leftrightarrow$ Upper confidence bound: reject $H_0$ iff $\theta_0 > \bar{\theta}(\mathbf{X})$
:::

. . .

**For the one-sample t-test:**

Upper bound: $\bar{X} + t_{\alpha, n-1}\cdot S/\sqrt{n}$

Lower bound: $\bar{X} - t_{\alpha, n-1}\cdot S/\sqrt{n}$

Note: uses $t_{\alpha, n-1}$ (one-sided critical value), NOT $t_{\alpha/2, n-1}$

## Summary: Two-Sided vs One-Sided Duality

| Test Type | Reject $H_0$ when | Dual CI |
|-----------|-------------------|---------|
| Two-sided: $H_a: \theta \neq \theta_0$ | $\|\hat{\theta} - \theta_0\| / SE > z_{\alpha/2}$ | Two-sided: $\theta_0 \notin [\hat{\theta} \pm z_{\alpha/2}\cdot SE]$ |
| Upper-tailed: $H_a: \theta > \theta_0$ | $(\hat{\theta} - \theta_0)/SE > z_\alpha$ | Lower bound: $\theta_0 < \hat{\theta} - z_\alpha\cdot SE$ |
| Lower-tailed: $H_a: \theta < \theta_0$ | $(\hat{\theta} - \theta_0)/SE < -z_\alpha$ | Upper bound: $\theta_0 > \hat{\theta} + z_\alpha\cdot SE$ |

. . .

::: callout-warning
## Common Error
A 95% **two-sided** CI uses $z_{0.025} = 1.96$, while a 95% **one-sided** bound uses $z_{0.05} = 1.645$. These are different! Don't use a two-sided CI to conduct a one-sided test at $\alpha = 0.05$.
:::

# Equivalence Testing and Non-Inferiority Trials {background-color="#27ae60"}

## Equivalence Testing: The Reverse Problem

Most hypothesis tests ask: **Is there an effect?**

Equivalence tests ask: **Is the effect small enough to be negligible?**

. . .

::: callout-important
## Setting
Define a **margin of clinical equivalence** $\delta > 0$.

We want to show the new treatment is **equivalent** to the standard, meaning the true difference $\mu_1 - \mu_2$ falls in $(-\delta, \delta)$.

$$H_0: |\mu_1 - \mu_2| \geq \delta \quad \text{vs} \quad H_a: |\mu_1 - \mu_2| < \delta$$
:::

. . .

**TOST Procedure (Two One-Sided Tests):**

Conduct two one-sided tests at level $\alpha$:

- $H_{01}: \mu_1 - \mu_2 \leq -\delta$ vs $H_{a1}: \mu_1 - \mu_2 > -\delta$
- $H_{02}: \mu_1 - \mu_2 \geq \delta$ vs $H_{a2}: \mu_1 - \mu_2 < \delta$

Conclude equivalence if **both** tests reject.

## TOST is Equivalent to a CI Check

::: callout-important
## Theorem (Schuirmann 1987)
The TOST procedure at level $\alpha$ rejects both $H_{01}$ and $H_{02}$ if and only if the $100(1-2\alpha)\%$ CI for $\mu_1 - \mu_2$ lies entirely within $(-\delta, \delta)$.
:::

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 13

# Bioequivalence example: generic vs. brand-name drug pharmacokinetics
# FDA requires AUC ratio within 80%-125% (log-scale: delta = log(1.25))
# Simulate log(AUC) differences
set.seed(123)
n_bio <- 30
delta <- log(1.25)  # equivalence margin on log scale

diff_data <- rnorm(n_bio, mean = 0.05, sd = 0.20)  # small true difference

xbar_d <- mean(diff_data)
s_d    <- sd(diff_data)
se_d   <- s_d / sqrt(n_bio)

# 90% CI (standard for bioequivalence, corresponding to TOST at alpha=0.05)
t90 <- qt(0.95, df = n_bio - 1)
ci90_low  <- xbar_d - t90 * se_d
ci90_high <- xbar_d + t90 * se_d

scenarios <- tibble(
  Scenario = c("Equivalent (this study)",
               "Possibly not equivalent",
               "Clearly not equivalent"),
  center = c(xbar_d, 0.18, 0.35),
  half_width = c(xbar_d - ci90_low, 0.12, 0.08)
) |>
  mutate(lo = center - half_width, hi = center + half_width,
         equiv = (lo > -delta) & (hi < delta))

ggplot(scenarios, aes(y = Scenario, x = center, color = equiv)) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.25, linewidth = 1.3) +
  geom_vline(xintercept = c(-delta, delta), linetype = "dashed",
             color = "#e74c3c", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "gray40") +
  annotate("rect", xmin = -delta, xmax = delta, ymin = -Inf, ymax = Inf,
           fill = "#27ae60", alpha = 0.08) +
  annotate("text", x = delta + 0.01, y = 3.4,
           label = "+delta == log(1.25)", parse = TRUE,
           color = "#e74c3c", hjust = 0, size = 4.5) +
  annotate("text", x = -delta - 0.01, y = 3.4,
           label = "-delta == log(0.80)", parse = TRUE,
           color = "#e74c3c", hjust = 1, size = 4.5) +
  scale_color_manual(values = c("TRUE" = "#27ae60", "FALSE" = "#e74c3c"),
                     labels = c("TRUE" = "Equivalent (CI inside margin)",
                                "FALSE" = "Not equivalent")) +
  labs(
    title = "Bioequivalence Assessment: 90% CI Must Fall Within Equivalence Margins",
    subtitle = "FDA standard: log(AUC ratio) CI within log(0.80) to log(1.25)",
    x = "Log(AUC ratio): Generic vs. Brand-name",
    y = NULL, color = NULL
  ) +
  theme(legend.position = "bottom")
```

## Non-Inferiority Testing

A **non-inferiority trial** asks: Is the new treatment **no worse** than the control by more than a specified margin?

. . .

$$H_0: \mu_{\text{new}} - \mu_{\text{ctrl}} \leq -\delta \quad \text{vs} \quad H_a: \mu_{\text{new}} - \mu_{\text{ctrl}} > -\delta$$

. . .

This is a **one-sided** equivalence test. It is widely used when:

- The new treatment has fewer side effects or is less costly
- Randomization to placebo is unethical (active-control trials)
- Establishing superiority is not the primary goal

. . .

**By duality:** Reject $H_0$ (conclude non-inferiority) if and only if the **lower bound** of a $100(1-\alpha)\%$ one-sided CI exceeds $-\delta$.

. . .

::: callout-note
## Example
The EINSTEIN-PE trial compared rivaroxaban (once-daily oral) vs. enoxaparin/warfarin for pulmonary embolism. The non-inferiority margin was an HR of 2.0. The upper bound of the 95% CI for the HR was 1.57 < 2.0, establishing non-inferiority.
:::

# Constructing CIs by Inverting Tests: The Pivot Method {background-color="#16a085"}

## The Pivot as the Bridge

The algebraic connection between a CI and a test often works through a **pivotal quantity** — a function of data and parameter whose distribution does not depend on the unknown parameter.

. . .

::: callout-important
## Definition (Devore)
A random variable $Q(X_1, \ldots, X_n; \theta)$ is a **pivotal quantity** (or pivot) if its distribution is the same for all $\theta$.
:::

. . .

**Key pivots we have used:**

| Situation | Pivot | Distribution |
|-----------|-------|-------------|
| $\mu$, $\sigma$ known | $Z = \frac{\bar{X}-\mu}{\sigma/\sqrt{n}}$ | $N(0,1)$ |
| $\mu$, $\sigma$ unknown | $T = \frac{\bar{X}-\mu}{S/\sqrt{n}}$ | $t_{n-1}$ |
| $\sigma^2$ | $\chi^2 = \frac{(n-1)S^2}{\sigma^2}$ | $\chi^2_{n-1}$ |
| Large-sample LRT | $-2\ln\Lambda$ | $\approx \chi^2_m$ |

## Inverting the Chi-Squared Pivot for $\sigma^2$

**Step 1:** The pivot $(n-1)S^2/\sigma^2 \sim \chi^2_{n-1}$.

**Step 2:** The acceptance region for testing $H_0: \sigma^2 = \sigma_0^2$ is:

$$\chi^2_{1-\alpha/2,\, n-1} \leq \frac{(n-1)S^2}{\sigma_0^2} \leq \chi^2_{\alpha/2,\, n-1}$$

. . .

**Step 3:** Solve for $\sigma_0^2$ to invert:

$$\frac{(n-1)S^2}{\chi^2_{\alpha/2,\, n-1}} \leq \sigma_0^2 \leq \frac{(n-1)S^2}{\chi^2_{1-\alpha/2,\, n-1}}$$

. . .

This is the **chi-squared CI for $\sigma^2$**, derived by inverting the chi-squared test.

Note: $\chi^2_{\alpha/2, n-1} > \chi^2_{1-\alpha/2, n-1}$, so dividing flips the inequality.

## Example: Variability in Drug Concentration

In pharmacokinetics, it's important to control **variability** in drug plasma concentrations. A drug passes QC if $\sigma < 5$ ng/mL (so $\sigma^2 < 25$).

```{r}
#| echo: true
# 15 measurements of drug plasma concentration (ng/mL)
set.seed(551)
conc <- rnorm(15, mean = 50, sd = 3.8)  # SD ≈ 3.8 (under QC limit)

n <- length(conc)
s2 <- var(conc)
df <- n - 1

# 95% two-sided CI for sigma^2
chi_upper <- qchisq(0.975, df)  # upper tail => lower bound of CI
chi_lower <- qchisq(0.025, df)  # lower tail => upper bound of CI

ci_var_low  <- df * s2 / chi_upper
ci_var_high <- df * s2 / chi_lower
ci_sd_low   <- sqrt(ci_var_low)
ci_sd_high  <- sqrt(ci_var_high)

cat("Sample variance:", round(s2, 3), "\n")
cat("Sample SD:", round(sqrt(s2), 3), "\n")
cat("95% CI for sigma^2: [", round(ci_var_low, 3), ",", round(ci_var_high, 3), "]\n")
cat("95% CI for sigma:   [", round(ci_sd_low, 3),  ",", round(ci_sd_high, 3),  "]\n")
cat("Upper bound for sigma:", round(ci_sd_high, 3), 
    "-- QC limit:", 5, "\n")
cat("Conclusion:", 
    ifelse(ci_sd_high < 5, "Pass QC (upper CI bound < 5)", "Fail QC"))
```

## Inverting the LRT: Profile Likelihood CIs

For more complex models, we can invert the LRT using Wilks' theorem:

$$C(\mathbf{X}) = \left\{\theta: -2\ln\left(\frac{L(\theta)}{L(\hat{\theta})}\right) \leq \chi^2_{\alpha, 1}\right\}$$

. . .

This is the **profile likelihood confidence interval**, valid for any one-dimensional parameter.

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 13
# Profile likelihood CI for exponential rate lambda
# Data: hospital length of stay (days)
set.seed(99)
los <- rexp(40, rate = 0.2)  # mean LOS = 5 days

log_lik <- function(lambda) {
  sum(dexp(los, rate = lambda, log = TRUE))
}

lambda_hat <- 1 / mean(los)
lmax       <- log_lik(lambda_hat)

lam_grid <- seq(0.05, 0.45, by = 0.002)
ll_vals  <- map_dbl(lam_grid, log_lik)
rel_ll   <- -2 * (ll_vals - lmax)

# chi-squared cutoff for 95% CI
cutoff <- qchisq(0.95, df = 1)

ci_lam <- range(lam_grid[rel_ll <= cutoff])

ggplot(tibble(lambda = lam_grid, rl = rel_ll), aes(x = lambda, y = rl)) +
  geom_line(color = "#006a4e", linewidth = 1.5) +
  geom_hline(yintercept = cutoff, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  geom_vline(xintercept = ci_lam, linetype = "dotted", color = "#3498db", linewidth = 1) +
  annotate("rect", xmin = ci_lam[1], xmax = ci_lam[2], ymin = -Inf, ymax = cutoff,
           fill = "#006a4e", alpha = 0.1) +
  annotate("text", x = lambda_hat, y = cutoff + 0.15,
           label = sprintf("chi^2[0.05,1] = %.2f", cutoff), parse = TRUE,
           color = "#e74c3c", size = 4.5) +
  annotate("text", x = mean(ci_lam), y = -0.2,
           label = sprintf("95%% Profile LI CI: [%.3f, %.3f]", ci_lam[1], ci_lam[2]),
           color = "#3498db", size = 4.5) +
  labs(
    title = expression("Profile Likelihood CI for Hospital LOS Rate " * lambda),
    subtitle = "95% CI = values of lambda where -2*log(LR) < chi-squared(0.95, df=1)",
    x = expression(lambda * " (discharge rate per day)"),
    y = expression(-2 * log(L(lambda) / L(hat(lambda))))
  )
```

# Multiple Testing and Confidence Intervals {background-color="#f39c12"}

## Multiplicity: When We Test Many Hypotheses

When we conduct $m$ tests simultaneously, even if all null hypotheses are true, we expect roughly $m \times \alpha$ false rejections by chance.

. . .

**By duality**, if we construct $m$ confidence intervals each with coverage $1-\alpha$:

- Each individual CI has $95\%$ coverage
- But the probability that **all $m$ CIs simultaneously contain their true parameter** is less than $95\%$

$$P(\text{at least one CI fails}) = 1 - P(\text{all cover}) \leq 1 - (1-\alpha)^m \approx m\alpha \text{ (Bonferroni)}$$

. . .

::: callout-important
## Simultaneous Confidence Intervals
To ensure that **all $m$ CIs simultaneously** contain their true values with probability $\geq 1-\alpha$:

Use individual level $\alpha^* = \alpha/m$ for each CI (Bonferroni correction).

This produces **wider** intervals that account for multiplicity.
:::

## Example: Multiple Endpoints in a Clinical Trial

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

# Clinical trial with 5 endpoints; simulate estimates and SEs
set.seed(551)
m <- 5
endpoints <- c("Systolic BP", "Diastolic BP", "LDL Cholesterol",
                "HbA1c", "6-min Walk Distance")
estimates <- c(-8.2, -4.1, -12.5, -0.4, 35.0)
se_vals   <- c(2.5, 1.8, 4.2, 0.15, 12.0)
units     <- c("mmHg", "mmHg", "mg/dL", "%", "meters")

# Unadjusted 95% CIs (z-based for simplicity)
z_ind <- qnorm(0.975)
# Bonferroni-adjusted 95% simultaneous CIs
z_bon <- qnorm(1 - 0.05 / (2 * m))

ci_df <- tibble(
  Endpoint = factor(endpoints, levels = rev(endpoints)),
  Est = estimates,
  SE  = se_vals,
  Unit = units,
  Est_std = estimates / se_vals  # standardize for display
) |>
  mutate(
    ci_lo_ind = Est_std - z_ind,
    ci_hi_ind = Est_std + z_ind,
    ci_lo_bon = Est_std - z_bon,
    ci_hi_bon = Est_std + z_bon
  ) |>
  pivot_longer(cols = starts_with("ci"),
               names_to = c("bound", "type"),
               names_pattern = "ci_(lo|hi)_(ind|bon)") |>
  pivot_wider(names_from = bound, values_from = value) |>
  mutate(Method = ifelse(type == "ind", "Individual 95% CI", "Bonferroni 95% Simultaneous CI"))

ggplot(ci_df, aes(y = Endpoint, x = Est_std, color = Method)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.25, linewidth = 1,
                 position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("Individual 95% CI" = "#006a4e",
                                "Bonferroni 95% Simultaneous CI" = "#e74c3c")) +
  labs(
    title = "Individual vs. Bonferroni-Adjusted CIs for 5 Trial Endpoints",
    subtitle = "Standardized estimates (Est/SE). Bonferroni CIs are wider to control familywise coverage.",
    x = "Standardized estimate (Est / SE)",
    y = NULL, color = NULL
  ) +
  theme(legend.position = "bottom")
```

# Devore 9.6 Textbook Examples {background-color="#2c3e50"}

## Devore Example 9.6: Testing Hypotheses About a Binomial Proportion

A clinical study examines whether a new screening test has sensitivity greater than 80%.

Let $p$ = true sensitivity. We test $H_0: p = 0.80$ vs $H_a: p > 0.80$.

In a sample of $n = 200$ subjects with disease, $X = 170$ test positive.

. . .

**By exact binomial test and CI:**

```{r}
#| echo: true
# Exact binomial test
n_screen <- 200
x_pos    <- 170

binom_test <- binom.test(x_pos, n_screen, p = 0.80, alternative = "greater")
binom_test_two <- binom.test(x_pos, n_screen, p = 0.80, alternative = "two.sided")

cat("One-sided p-value:", round(binom_test$p.value, 5), "\n")
cat("95% lower bound for p:", round(binom_test$conf.int[1], 4), "\n")
cat("\nTwo-sided 95% CI for p: [",
    round(binom_test_two$conf.int[1], 4), ",",
    round(binom_test_two$conf.int[2], 4), "]\n")
cat("Duality check: 0.80 in two-sided CI?",
    binom_test_two$conf.int[1] <= 0.80 & 0.80 <= binom_test_two$conf.int[2])
```

. . .

**Duality interpretation:** Since the one-sided lower bound (95%) exceeds 0.80, we reject $H_0$ at $\alpha = 0.05$. The 95% lower bound provides the same information as the one-sided p-value.

## CI-Test Duality Illustrated (Textbook Example)

Consider $X \sim \text{Poisson}(\mu)$. We observe $X = 3$ and wish to test $H_0: \mu = 5$ vs $H_a: \mu < 5$.

. . .

```{r}
#| echo: true
# Exact Poisson test
x_obs <- 3
mu_0  <- 5.0

# One-sided: lower-tailed (alternative: mu < 5)
p_lower <- ppois(x_obs, lambda = mu_0)
cat("P(X <= 3 | mu = 5):", round(p_lower, 5), "\n\n")

# 95% two-sided CI for mu via exact Poisson intervals
# Lower bound: qchisq(0.025, 2*x) / 2
# Upper bound: qchisq(0.975, 2*(x+1)) / 2
ci_low_pois  <- qchisq(0.025, 2 * x_obs) / 2
ci_high_pois <- qchisq(0.975, 2 * (x_obs + 1)) / 2

cat("Exact 95% CI for mu: [", round(ci_low_pois, 4), ",",
    round(ci_high_pois, 4), "]\n")
cat("mu_0 = 5 in CI?", ci_low_pois <= mu_0 & mu_0 <= ci_high_pois)
```

. . .

Notice: $\mu_0 = 5$ is *outside* the 95% CI, consistent with the significant (one-sided) p-value of $p = 0.265$... wait, let's check: actually $p_{\text{lower}} = 0.265 > 0.05$, so we fail to reject. Let's verify the CI also contains $\mu_0 = 5$.

This duality confirms: **if $p \geq \alpha$, then $\mu_0$ is in the CI; if $p < \alpha$, it is not.**

# Key Takeaways and Looking Ahead {background-color="#2c3e50"}

## Key Takeaways

::: callout-important
## Today's Main Points

1. **Statistical significance** $\neq$ **practical/clinical significance**: always report effect sizes and CIs alongside p-values

2. **CI-Test Duality (Devore 9.6):** $\theta_0 \in$ CI(X) $\iff$ fail to reject $H_0: \theta = \theta_0$ at the same $\alpha$

3. **Inverting a test** provides a general, principled method for constructing CIs from any hypothesis test

4. **One-sided duality:** a $100(1-\alpha)\%$ one-sided bound corresponds to a one-sided test at level $\alpha$ (using $z_\alpha$, not $z_{\alpha/2}$)

5. **Equivalence/non-inferiority testing:** FDA bioequivalence uses 90% CI (TOST at $\alpha = 0.05$); non-inferiority uses one-sided lower bound

6. **Multiple testing:** simultaneous CIs must be widened (e.g., Bonferroni) to maintain joint coverage

7. **Profile likelihood CIs** invert the LRT using Wilks' theorem: a general tool for any parameter
:::

## Looking Ahead

**Next class (Lesson 17): Bootstrap Hypothesis Testing (C&H Chapter 8)**

- The bootstrap hypothesis test: resampling under $H_0$
- One-sample and two-sample bootstrap tests
- Permutation tests: exact null distribution by permuting labels
- Comparison: parametric vs. bootstrap vs. permutation tests
- Applications to clinical trial data

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 9.6.
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 8.
-   Schuirmann, D. J. (1987). A comparison of the two one-sided tests procedure and the power approach for assessing the equivalence of average bioavailability. *Journal of Pharmacokinetics and Biopharmaceutics*, 15(6), 657–680.
-   Wasserstein, R. L., & Lazar, N. A. (2016). The ASA statement on p-values: context, process, and purpose. *The American Statistician*, 70(2), 129–133.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
# For PDF export
decktape reveal --fragments docs/lessons/16_FurtherHypothesisTesting/16_FurtherHypothesisTesting.html lessons/16_FurtherHypothesisTesting/16_FurtherHypothesisTesting.pdf
```
