---
title: "Lesson 12: Introduction to Hypothesis Testing"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 12 - Introduction to Hypothesis Testing
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 12: Introduction to Hypothesis Testing {background-color="#2c3e50"}

## Review: Where We Left Off

**Lessons 9-11: Bootstrap Methods and Confidence Intervals**

-   Bootstrap principle: resample from sample to estimate sampling distribution
-   Bootstrap percentile and t intervals
-   Relationship between confidence intervals and plausible parameter values

. . .

**The Next Step:** What if we want to make a **decision** rather than just estimate?

. . .

**Today's Goals:**

1. Understand the hypothesis testing framework
2. Define null and alternative hypotheses
3. Learn about test statistics and rejection regions
4. Understand Type I and Type II errors
5. Apply these concepts to medical research examples

## From Estimation to Testing

::: columns
::: {.column width="50%"}
**Confidence Intervals**

- "What is a plausible range for the parameter?"
- Example: "95% CI for mean reduction in blood pressure is (8, 16) mmHg"
- Provides a range of values
:::

::: {.column width="50%"}
**Hypothesis Testing**

- "Is there evidence that the parameter differs from a specific value?"
- Example: "Does this drug reduce blood pressure at all?"
- Provides a yes/no decision
:::
:::

. . .

::: callout-note
## Key Insight (Devore 9.1)
Hypothesis testing is about deciding which of two contradictory claims about a parameter is correct based on sample data.
:::

# The Hypothesis Testing Framework {background-color="#3498db"}

## What is a Statistical Hypothesis?

::: callout-important
## Definition (Devore 9.1)
A **statistical hypothesis** is a claim or assertion about the value of a single parameter, about the values of several parameters, or about the form of an entire probability distribution.
:::

. . .

**Examples:**

- The claim $\mu = 120$ mmHg, where $\mu$ is the true average systolic blood pressure
- The statement $p < 0.10$, where $p$ is the proportion of patients who experience adverse events
- The assertion $\mu_{\text{treatment}} - \mu_{\text{control}} > 0$

## The Two Competing Hypotheses

In any hypothesis testing problem, we have two contradictory hypotheses:

. . .

::: callout-important
## Definitions (Devore 9.1)

**Null Hypothesis ($H_0$):** The claim that is initially assumed to be true (the "prior belief" or "status quo" claim).

**Alternative Hypothesis ($H_a$ or $H_A$):** The assertion that is contradictory to $H_0$ — what we might conclude if the data provides strong evidence against $H_0$.
:::

. . .

The null hypothesis will be rejected in favor of the alternative **only if** sample evidence strongly suggests that $H_0$ is false.

## The Courtroom Analogy

::: columns
::: {.column width="50%"}
**Criminal Trial (US)**

- **Initial assumption:** Defendant is innocent
- **Burden of proof:** On prosecution
- **Standard:** "Beyond reasonable doubt"
- **Decision:** Guilty or Not Guilty

*Note:* "Not Guilty" $\neq$ "Innocent"
:::

::: {.column width="50%"}
**Hypothesis Testing**

- **Initial assumption:** $H_0$ is true
- **Burden of proof:** On data to contradict $H_0$
- **Standard:** Significance level $\alpha$
- **Decision:** Reject $H_0$ or Fail to Reject $H_0$

*Note:* "Fail to Reject" $\neq$ "$H_0$ is True"
:::
:::

## Visualizing the Analogy

```{r}
#| code-fold: true
#| fig-height: 5
#| fig-width: 12
# Create a visual comparison with plain text that renders reliably
trial_data <- tibble(
  Category = rep(c("Courtroom", "Statistics"), each = 4),
  Element = rep(c("Initial Belief", "Evidence", "Decision 1", "Decision 2"), 2),
  x_pos = rep(1:4, 2),
  y_pos = c(rep(1, 4), rep(2, 4))
)

# Base plot with tiles
p <- trial_data |>
  mutate(Element = factor(Element, levels = c("Initial Belief", "Evidence", "Decision 1", "Decision 2"))) |>
  ggplot(aes(x = Element, y = Category)) +
  geom_tile(aes(fill = Category), alpha = 0.3, color = "white", linewidth = 2) +
  scale_fill_manual(values = c("Courtroom" = "steelblue", "Statistics" = "coral")) +
  labs(title = "The Hypothesis Testing Analogy", x = "", y = "") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none",
        axis.text.y = element_text(face = "bold", size = 14))

# Add courtroom labels (plain text)
p <- p +
  annotate("text", x = 1, y = 1, label = "Innocent until\nproven guilty", size = 5) +
  annotate("text", x = 2, y = 1, label = "Prosecution\npresents evidence", size = 5) +
  annotate("text", x = 3, y = 1, label = "Guilty\n(reject innocence)", size = 5) +
  annotate("text", x = 4, y = 1, label = "Not Guilty\n(fail to reject)", size = 5)

# Add statistics labels using expression() for proper subscripts
p <- p +
  annotate("text", x = 1, y = 2, size = 5, parse = TRUE,
           label = 'atop(H[0]~"assumed", "true")') +
  annotate("text", x = 2, y = 2, label = "Sample data\ncollected", size = 5) +
  annotate("text", x = 3, y = 2, size = 5, parse = TRUE,
           label = 'atop("Reject"~H[0], "(favor"~H[a]*")")') +
  annotate("text", x = 4, y = 2, size = 5, parse = TRUE,
           label = 'atop("Fail to reject"~H[0], "("*H[0]~"plausible)")')

p
```

## Two Possible Conclusions

::: callout-warning
## Important Distinction
A hypothesis test can only lead to one of two conclusions:

1. **Reject $H_0$** — Evidence strongly suggests $H_0$ is false
2. **Fail to reject $H_0$** — Evidence does not strongly contradict $H_0$

We **never** say "Accept $H_0$" because failing to reject is not the same as proving $H_0$ is true!
:::

. . .

**Analogy:** A "Not Guilty" verdict doesn't prove innocence — it means there wasn't enough evidence to convict.

# Formulating Hypotheses {background-color="#8e44ad"}

## Example 1: New Antihypertensive Drug

A pharmaceutical company has developed a new drug to lower blood pressure. The current standard treatment has a mean reduction of 10 mmHg. Let $\mu$ denote the true mean BP reduction with the new drug.

. . .

**Question:** Is the new drug better than the standard?

. . .

**Hypotheses:**
$$H_0: \mu = 10 \quad \text{(no improvement over standard)}$$
$$H_a: \mu > 10 \quad \text{(new drug is better)}$$

. . .

The company will only switch to the new drug if there is **convincing evidence** of improvement.

## Example 2: Is Normal Body Temperature Really 98.6°F?

The widely accepted "normal" human body temperature of 98.6°F (37°C) dates back to a 19th century German study. Recent research suggests this value may be inaccurate.

Let $\mu$ denote the true mean body temperature of healthy adults.

. . .

**Hypotheses:**
$$H_0: \mu = 98.6 \quad \text{(traditional value is correct)}$$
$$H_a: \mu \neq 98.6 \quad \text{(traditional value is wrong — either direction)}$$

. . .

This is a **two-sided** test because departures in either direction would be scientifically important.

## Example 3: Adverse Event Rate

An investigational drug is considered safe if the rate of serious adverse events is at most 5%. In a clinical trial, researchers want to determine if the drug exceeds this threshold.

Let $p$ denote the true proportion of patients experiencing serious adverse events.

. . .

**Hypotheses:**
$$H_0: p = 0.05 \quad \text{(drug meets safety threshold)}$$
$$H_a: p > 0.05 \quad \text{(drug exceeds threshold — unsafe)}$$

. . .

This is a **one-sided** (upper-tailed) test because only a rate higher than 5% would be concerning.

## General Structure of Hypotheses

The null hypothesis is typically stated as an **equality**:
$$H_0: \theta = \theta_0$$
where $\theta_0$ is called the **null value** of the parameter.

. . .

The alternative hypothesis takes one of three forms:

| Alternative | Test Type | Example Context |
|-------------|-----------|-----------------|
| $H_a: \theta > \theta_0$ | Upper-tailed | New treatment better than standard |
| $H_a: \theta < \theta_0$ | Lower-tailed | Reduced error rate |
| $H_a: \theta \neq \theta_0$ | Two-tailed | Testing for any difference |

## Choosing the Alternative: Research vs. Null

::: callout-tip
## Practical Guidance (Devore 9.1)
- $H_a$ is often called the **research hypothesis** — what the investigator wants to demonstrate
- $H_0$ represents the **status quo** or "no effect/no difference"
- The burden of proof is on showing $H_a$ is true

This is **conservative**: we only claim an effect when the data provides strong evidence.
:::

. . .

**Medical Research Context:**

- New treatment must **prove** it's better than standard
- We protect against false claims of efficacy
- FDA approval requires strong evidence that benefits outweigh risks

# Test Procedures {background-color="#e74c3c"}

## Components of a Test Procedure

::: callout-important
## Definition (Devore 9.1)
A **test procedure** is specified by:

1. A **test statistic**: a function of the sample data used to make the decision
2. A **rejection region** (critical region): the set of test statistic values for which $H_0$ will be rejected

The null hypothesis is rejected if and only if the observed test statistic value falls in the rejection region.
:::

## Example: Testing Vaccine Efficacy

A new vaccine is tested. Let $p$ be the true proportion of vaccinated individuals who develop the disease. We test $H_0: p = 0.20$ vs $H_a: p < 0.20$.

In a trial of $n = 200$ vaccinated individuals:

. . .

**Test Statistic:** $X$ = number who develop disease

**Rejection Region:** Reject $H_0$ if $X \leq 30$

. . .

```{r}
#| echo: true
# Under H0: X ~ Binomial(200, 0.20)
# Expected value under H0
n <- 200
p0 <- 0.20
expected_X <- n * p0
cat("Expected X under H0:", expected_X, "\n")
cat("Rejection region: X ≤ 30 (substantially less than 40)")
```

## Visualizing the Rejection Region

```{r}
#| code-fold: true
#| fig-height: 6
# Binomial distribution under H0
x_vals <- 0:80
probs <- dbinom(x_vals, size = 200, prob = 0.20)

tibble(x = x_vals, prob = probs) |>
  mutate(region = ifelse(x <= 30, "Rejection Region", "Non-rejection Region")) |>
  ggplot(aes(x = x, y = prob, fill = region)) +
  geom_col(width = 0.8) +
  geom_vline(xintercept = 40, linetype = "dashed", color = "black", linewidth = 1) +
  scale_fill_manual(values = c("Rejection Region" = "coral", 
                                "Non-rejection Region" = "steelblue")) +
  annotate("text", x = 40, y = 0.06, label = expression(E(X) == 40), hjust = -0.1, size = 5) +
  annotate("text", x = 15, y = 0.03, label = expression("Reject"~H[0]), color = "white", size = 6, fontface = "bold") +
  labs(x = "X (number who develop disease)", y = "Probability",
       title = expression(paste("Distribution of X under ", H[0], ": p = 0.20")),
       subtitle = "Rejection region: X ≤ 30") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

## Example: Testing Engine Fuel Efficiency (Devore Example 9.3)

A manufacturer considers modifying truck engines to improve fuel efficiency. Current mean: 25 mpg, SD: 3 mpg (known). Let $\mu$ = true mean mpg for modified engines.

$$H_0: \mu = 25 \quad \text{vs} \quad H_a: \mu > 25$$

. . .

Test with $n = 10$ trucks:

**Test Statistic:** $\bar{X}$ = sample mean fuel efficiency

**Possible Rejection Region:** Reject $H_0$ if $\bar{X} \geq 27.5$

. . .

**Key Question:** How do we choose the cutoff value (27.5 vs 26 vs 28)?

## The Tradeoff in Choosing the Cutoff

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
# Show two different rejection regions
sigma <- 3
n <- 10
se <- sigma / sqrt(n)
mu0 <- 25

x_seq <- seq(22, 30, length.out = 300)
y_H0 <- dnorm(x_seq, mean = mu0, sd = se)

p1 <- tibble(x = x_seq, y = y_H0) |>
  ggplot(aes(x = x, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(x = x_seq[x_seq >= 26], y = y_H0[x_seq >= 26]),
            aes(x = x, y = y), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = 26, linetype = "dashed", color = "red") +
  labs(title = "Cutoff = 26 mpg", 
       subtitle = bquote("P(Reject"~H[0]~"|"~H[0]~"true) ="~.(round(1 - pnorm(26, mu0, se), 3))),
       x = expression(bar(X)), y = "Density") +
  annotate("text", x = 27.5, y = 0.15, label = "Rejection\nRegion", color = "red", size = 5)

p2 <- tibble(x = x_seq, y = y_H0) |>
  ggplot(aes(x = x, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(x = x_seq[x_seq >= 27.5], y = y_H0[x_seq >= 27.5]),
            aes(x = x, y = y), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = 27.5, linetype = "dashed", color = "red") +
  labs(title = "Cutoff = 27.5 mpg", 
       subtitle = bquote("P(Reject"~H[0]~"|"~H[0]~"true) ="~.(round(1 - pnorm(27.5, mu0, se), 3))),
       x = expression(bar(X)), y = "Density") +
  annotate("text", x = 28.5, y = 0.05, label = "Rejection\nRegion", color = "red", size = 5)

p1 + p2
```

**Lower cutoff:** More likely to reject $H_0$ (even when true!)

**Higher cutoff:** Less likely to reject $H_0$ (even when false!)

# Type I and Type II Errors {background-color="#16a085"}

## Two Types of Errors

When making a decision based on sample data, two types of errors are possible:

. . .

::: callout-important
## Definitions (Devore 9.1)

**Type I Error:** Rejecting $H_0$ when it is actually **true** (false positive)

**Type II Error:** Failing to reject $H_0$ when it is actually **false** (false negative)
:::

. . .

| | $H_0$ True | $H_0$ False |
|---|:---:|:---:|
| Reject $H_0$ | **Type I Error** | Correct Decision |
| Fail to Reject $H_0$ | Correct Decision | **Type II Error** |

## Error Probabilities

::: callout-important
## Notation
$$\alpha = P(\text{Type I Error}) = P(\text{Reject } H_0 \mid H_0 \text{ is true})$$

$$\beta = P(\text{Type II Error}) = P(\text{Fail to reject } H_0 \mid H_0 \text{ is false})$$

**Power** = $1 - \beta$ = $P(\text{Reject } H_0 \mid H_0 \text{ is false})$
:::

. . .

- $\alpha$ is called the **significance level** of the test
- Common choices: $\alpha = 0.05$, $\alpha = 0.01$, $\alpha = 0.10$

## Medical Context: Errors Have Consequences

**Example:** Testing whether a new cancer drug is effective

. . .

**Type I Error (False Positive):**

- Conclude drug works when it doesn't
- Patients receive ineffective treatment
- Resources wasted on worthless drug
- *Consequence:* Harm to patients, wasted resources

. . .

**Type II Error (False Negative):**

- Conclude drug doesn't work when it does
- Effective treatment never reaches patients
- *Consequence:* Missed opportunity to save lives

## Example: Drug Approval Errors (Devore Example 9.4)

**Scenario:** Testing a new antihypertensive drug against standard care.

$H_0: \mu_{\text{new}} = \mu_{\text{standard}}$ (no improvement)

$H_a: \mu_{\text{new}} > \mu_{\text{standard}}$ (new drug is better)

. . .

| Error Type | What Happens | Consequence |
|------------|--------------|-------------|
| Type I | Approve ineffective drug | Patients get worse care; company invested in useless drug |
| Type II | Reject effective drug | Patients miss beneficial treatment; opportunity loss |

. . .

**FDA Approach:** Keep $\alpha$ small (typically 0.05) — protect against approving ineffective drugs.

## Visualizing Type I and Type II Errors

```{r}
#| code-fold: true
#| fig-height: 7
#| fig-width: 14
# Parameters
mu0 <- 25  # Null hypothesis mean
mu_a <- 27  # Alternative (true mean if H0 false)
sigma <- 3
n <- 10
se <- sigma / sqrt(n)
cutoff <- 26.5  # Rejection cutoff

x_seq <- seq(21, 31, length.out = 500)

# Create data for both distributions
df_null <- tibble(
  x = x_seq,
  y = dnorm(x_seq, mu0, se),
  dist = "Under H0 (mu = 25)"
)

df_alt <- tibble(
  x = x_seq,
  y = dnorm(x_seq, mu_a, se),
  dist = "Under Ha (mu = 27)"
)

# Plot
ggplot() +
  # Null distribution
  geom_line(data = df_null, aes(x = x, y = y), color = "steelblue", linewidth = 1.5) +
  geom_area(data = df_null |> filter(x >= cutoff), 
            aes(x = x, y = y), fill = "coral", alpha = 0.5) +
  # Alternative distribution  
  geom_line(data = df_alt, aes(x = x, y = y), color = "darkgreen", linewidth = 1.5) +
  geom_area(data = df_alt |> filter(x < cutoff),
            aes(x = x, y = y), fill = "purple", alpha = 0.3) +
  # Cutoff line
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "black", linewidth = 1.2) +
  # Labels
  annotate("text", x = 28, y = 0.12, label = expression(alpha~"(Type I)"), 
           color = "red", size = 6, fontface = "bold") +
  annotate("text", x = 24.5, y = 0.08, label = expression(beta~"(Type II)"), 
           color = "purple", size = 6, fontface = "bold") +
  annotate("text", x = cutoff, y = 0.45, label = paste0("Cutoff = ", cutoff), 
           hjust = -0.1, size = 5) +
  annotate("text", x = 23, y = 0.35, label = expression(paste("Distribution under ", H[0])), 
           color = "steelblue", size = 5) +
  annotate("text", x = 29, y = 0.35, label = expression(paste("Distribution under ", H[a])), 
           color = "darkgreen", size = 5) +
  labs(x = expression(bar(X)), y = "Density",
       title = "Type I and Type II Errors Visualized",
       subtitle = expression("Fuel efficiency example:"~H[0]*":"~mu==25~"vs"~H[a]*":"~mu>25)) +
  theme(legend.position = "none")
```

## Computing Error Probabilities: Example

For the fuel efficiency test with cutoff $\bar{x} \geq 27.5$:

```{r}
#| echo: true
mu0 <- 25
sigma <- 3
n <- 10
se <- sigma / sqrt(n)
cutoff <- 27.5

# Type I error: P(reject H0 | H0 true)
alpha <- 1 - pnorm(cutoff, mean = mu0, sd = se)
cat("α = P(X̄ ≥", cutoff, "| μ =", mu0, ") =", round(alpha, 4), "\n")

# Type II error: P(fail to reject | H0 false)
# Suppose true mean is μ = 26.5
mu_true <- 26.5
beta <- pnorm(cutoff, mean = mu_true, sd = se)
cat("β = P(X̄ <", cutoff, "| μ =", mu_true, ") =", round(beta, 4), "\n")
cat("Power = 1 - β =", round(1 - beta, 4))
```

## The Fundamental Tradeoff

::: callout-warning
## The Tradeoff (Devore 9.1)
For a **fixed sample size**, decreasing $\alpha$ (Type I error) necessarily **increases** $\beta$ (Type II error), and vice versa.

The only way to decrease **both** error probabilities simultaneously is to **increase the sample size**.
:::

. . .

```{r}
#| echo: true
# Demonstrate tradeoff with different cutoffs
cutoffs <- c(26, 26.5, 27, 27.5, 28)
alpha_vals <- 1 - pnorm(cutoffs, mu0, se)
beta_vals <- pnorm(cutoffs, mean = 27, sd = se)  # assuming true mu = 27

tibble(Cutoff = cutoffs, Alpha = round(alpha_vals, 4), Beta = round(beta_vals, 4)) |>
  mutate(Power = 1 - Beta)
```

## Visualizing the Tradeoff

```{r}
#| code-fold: true
#| fig-height: 6
cutoffs_seq <- seq(24, 29, by = 0.1)
error_data <- tibble(
  cutoff = cutoffs_seq,
  alpha = 1 - pnorm(cutoffs_seq, mu0, se),
  beta = pnorm(cutoffs_seq, mean = 27, sd = se)
) |>
  pivot_longer(cols = c(alpha, beta), names_to = "error_type", values_to = "probability")

ggplot(error_data, aes(x = cutoff, y = probability, color = error_type)) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = c("alpha" = "coral", "beta" = "purple"),
                     labels = c(expression(alpha~"(Type I)"), expression(beta~"(Type II)"))) +
  geom_vline(xintercept = 26.56, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 26.56, y = 0.8, label = "Equal error rates", hjust = -0.1) +
  labs(x = "Cutoff Value", y = "Error Probability",
       title = "The α-β Tradeoff (Fixed n = 10, True μ = 27)",
       color = "Error Type") +
  theme(legend.position = "right")
```

## Simulation: Demonstrating Error Types

```{r}
#| echo: true
#| code-fold: true
set.seed(551)
n_simulations <- 10000
n <- 10
sigma <- 3
cutoff <- 26.5

# Simulate under H0 (mu = 25)
samples_H0 <- replicate(n_simulations, {
  x <- rnorm(n, mean = 25, sd = sigma)
  mean(x)
})
type_I_rate <- mean(samples_H0 >= cutoff)

# Simulate under Ha (mu = 27)
samples_Ha <- replicate(n_simulations, {
  x <- rnorm(n, mean = 27, sd = sigma)
  mean(x)
})
type_II_rate <- mean(samples_Ha < cutoff)

cat("Simulated Type I Error Rate:", round(type_I_rate, 4), "\n")
cat("Theoretical α:", round(1 - pnorm(cutoff, 25, sigma/sqrt(n)), 4), "\n\n")
cat("Simulated Type II Error Rate:", round(type_II_rate, 4), "\n")
cat("Theoretical β:", round(pnorm(cutoff, 27, sigma/sqrt(n)), 4))
```

## Visualizing Simulated Decisions

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
p1 <- tibble(xbar = samples_H0) |>
  mutate(decision = ifelse(xbar >= cutoff, "Reject H0 (Type I Error)", "Correct")) |>
  ggplot(aes(x = xbar, fill = decision)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = cutoff, linetype = "dashed", linewidth = 1.2) +
  scale_fill_manual(values = c("Correct" = "steelblue", "Reject H0 (Type I Error)" = "coral")) +
  labs(x = expression(bar(X)), y = "Count",
       title = expression(paste("Under ", H[0], ": ", mu, " = 25")),
       subtitle = paste0("Type I Error Rate: ", round(type_I_rate * 100, 1), "%"),
       fill = "Decision") +
  theme(legend.position = "bottom")

p2 <- tibble(xbar = samples_Ha) |>
  mutate(decision = ifelse(xbar >= cutoff, "Correct (Reject)", "Fail to Reject (Type II Error)")) |>
  ggplot(aes(x = xbar, fill = decision)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = cutoff, linetype = "dashed", linewidth = 1.2) +
  scale_fill_manual(values = c("Correct (Reject)" = "darkgreen", 
                                "Fail to Reject (Type II Error)" = "purple")) +
  labs(x = expression(bar(X)), y = "Count",
       title = expression(paste("Under ", H[a], ": ", mu, " = 27")),
       subtitle = paste0("Type II Error Rate: ", round(type_II_rate * 100, 1), "%"),
       fill = "Decision") +
  theme(legend.position = "bottom")

p1 + p2
```

# Significance Level and Rejection Regions {background-color="#9b59b6"}

## The Significance Level

::: callout-important
## Definition
The **significance level** $\alpha$ of a test is the probability of making a Type I error — rejecting $H_0$ when it is actually true.

$$\alpha = P(\text{Reject } H_0 \mid H_0 \text{ is true})$$

Common choices: $\alpha = 0.05$, $\alpha = 0.01$, $\alpha = 0.10$
:::

. . .

**Why prioritize controlling Type I error?**

- Type I errors can lead to **false claims** (e.g., ineffective drugs approved)
- Scientific community uses $\alpha = 0.05$ as a standard threshold
- $\alpha$ is chosen **before** collecting data

## Constructing Rejection Regions

Once $\alpha$ is chosen, we determine the rejection region:

**Upper-tailed test** ($H_a: \mu > \mu_0$):
$$\text{Reject } H_0 \text{ if } \bar{X} \geq \mu_0 + z_\alpha \cdot \frac{\sigma}{\sqrt{n}}$$

**Lower-tailed test** ($H_a: \mu < \mu_0$):
$$\text{Reject } H_0 \text{ if } \bar{X} \leq \mu_0 - z_\alpha \cdot \frac{\sigma}{\sqrt{n}}$$

**Two-tailed test** ($H_a: \mu \neq \mu_0$):
$$\text{Reject } H_0 \text{ if } |\bar{X} - \mu_0| \geq z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}}$$

## Example: Calculating Critical Values

For the fuel efficiency test at $\alpha = 0.05$:

```{r}
#| echo: true
alpha <- 0.05
mu0 <- 25
sigma <- 3
n <- 10
se <- sigma / sqrt(n)

# Critical value for upper-tailed test
z_alpha <- qnorm(1 - alpha)
critical_value <- mu0 + z_alpha * se

cat("z_α =", round(z_alpha, 3), "\n")
cat("Critical value: X̄ ≥", round(critical_value, 3), "\n")
cat("Rejection region: X̄ ≥", round(critical_value, 3))
```

## Visualizing Critical Regions by Tail Type

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 16
x_seq <- seq(21, 29, length.out = 500)
y <- dnorm(x_seq, mu0, se)

z_alpha <- qnorm(1 - alpha)
z_alpha2 <- qnorm(1 - alpha/2)

crit_upper <- mu0 + z_alpha * se
crit_lower <- mu0 - z_alpha * se
crit_two_upper <- mu0 + z_alpha2 * se
crit_two_lower <- mu0 - z_alpha2 * se

df <- tibble(x = x_seq, y = y)

p1 <- ggplot(df, aes(x, y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = df |> filter(x >= crit_upper), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = crit_upper, linetype = "dashed", color = "red") +
  annotate("text", x = crit_upper, y = 0.35, label = paste0("c = ", round(crit_upper, 2)), hjust = -0.1) +
  labs(title = expression(paste("Upper-tailed: ", H[a], ": μ > 25")),
       subtitle = "Reject if X̄ ≥ c", x = expression(bar(X)), y = "Density")

p2 <- ggplot(df, aes(x, y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = df |> filter(x <= crit_lower), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = crit_lower, linetype = "dashed", color = "red") +
  annotate("text", x = crit_lower, y = 0.35, label = paste0("c = ", round(crit_lower, 2)), hjust = 1.1) +
  labs(title = expression(paste("Lower-tailed: ", H[a], ": μ < 25")),
       subtitle = "Reject if X̄ ≤ c", x = expression(bar(X)), y = "Density")

p3 <- ggplot(df, aes(x, y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = df |> filter(x >= crit_two_upper | x <= crit_two_lower), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = c(crit_two_lower, crit_two_upper), linetype = "dashed", color = "red") +
  labs(title = expression(paste("Two-tailed: ", H[a], ": μ ≠ 25")),
       subtitle = "Reject if |X̄ - 25| ≥ c", x = expression(bar(X)), y = "Density")

p1 + p2 + p3
```

# Medical Research Example: Clinical Trial {background-color="#2c3e50"}

## Complete Example: Blood Pressure Drug Trial

**Setting:** Testing a new antihypertensive drug.

**Data:** $n = 25$ patients, mean BP reduction $\bar{x} = 12.5$ mmHg, known population SD $\sigma = 8$ mmHg.

**Question:** Is there evidence that the true mean reduction exceeds 10 mmHg?

. . .

**Step 1: State Hypotheses**
$$H_0: \mu = 10 \quad \text{vs} \quad H_a: \mu > 10$$

**Step 2: Choose significance level**
$$\alpha = 0.05$$

## Step 3: Identify Test Statistic and Rejection Region

```{r}
#| echo: true
# Given data
xbar <- 12.5
sigma <- 8
n <- 25
mu0 <- 10
alpha <- 0.05

# Standard error
se <- sigma / sqrt(n)

# Test statistic: Z = (X̄ - μ₀) / (σ/√n)
z_stat <- (xbar - mu0) / se
cat("Test statistic: z =", round(z_stat, 3), "\n")

# Critical value for α = 0.05 (upper-tailed)
z_crit <- qnorm(1 - alpha)
cat("Critical value: z_α =", round(z_crit, 3), "\n")

# Rejection region
cat("Rejection region: z ≥", round(z_crit, 3))
```

## Step 4: Make Decision

```{r}
#| echo: true
# Decision
if (z_stat >= z_crit) {
  cat("Decision: Reject H₀\n\n")
  cat("Interpretation: There is statistically significant evidence at α = 0.05\n")
  cat("that the true mean BP reduction exceeds 10 mmHg.")
} else {
  cat("Decision: Fail to reject H₀\n\n")
  cat("Interpretation: There is insufficient evidence at α = 0.05\n")
  cat("to conclude that the true mean BP reduction exceeds 10 mmHg.")
}
```

## Visualizing the Test Result

```{r}
#| code-fold: true
#| fig-height: 6
z_seq <- seq(-4, 4, length.out = 500)
y <- dnorm(z_seq)

tibble(z = z_seq, y = y) |>
  ggplot(aes(x = z, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(z = z_seq[z_seq >= z_crit], 
                          y = dnorm(z_seq[z_seq >= z_crit])),
            aes(x = z, y = y), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = z_crit, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = z_stat, color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = z_crit + 0.1, y = 0.35, 
           label = expression(z[alpha] == 1.645), hjust = 0, size = 5, color = "red") +
  annotate("text", x = z_stat + 0.1, y = 0.25, 
           label = paste0("z = ", round(z_stat, 2)), hjust = 0, size = 5, color = "darkgreen") +
  annotate("text", x = 2.5, y = 0.05, label = expression(alpha == 0.05), color = "coral", size = 5) +
  labs(x = "z", y = "Density",
       title = "Hypothesis Test for Blood Pressure Drug",
       subtitle = expression("Test statistic falls in rejection region"~symbol("\256")~"Reject"~H[0]))
```

## Error Analysis for This Test

```{r}
#| echo: true
#| code-fold: true
# Type I error (fixed at α)
cat("Type I Error Probability: α =", alpha, "\n\n")

# Type II error for different true values of μ
mu_true_vals <- c(11, 12, 13, 14, 15)
beta_vals <- pnorm(mu0 + z_crit * se, mean = mu_true_vals, sd = se)

error_table <- tibble(
  `True μ` = mu_true_vals,
  `β (Type II)` = round(beta_vals, 4),
  `Power (1-β)` = round(1 - beta_vals, 4)
)

cat("Type II Error and Power for different true μ values:\n")
print(error_table)
```

## Power Curve

```{r}
#| code-fold: true
#| fig-height: 6
mu_range <- seq(10, 18, by = 0.1)
power_vals <- 1 - pnorm(mu0 + z_crit * se, mean = mu_range, sd = se)

tibble(mu = mu_range, power = power_vals) |>
  ggplot(aes(x = mu, y = power)) +
  geom_line(linewidth = 1.5, color = "darkgreen") +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 10, linetype = "dotted", alpha = 0.5) +
  annotate("text", x = 17, y = 0.82, label = "80% Power", color = "red", size = 5) +
  annotate("text", x = 10.2, y = 0.5, label = expression(H[0]: mu == 10), hjust = 0, size = 5) +
  labs(x = expression(paste("True ", mu)), y = "Power",
       title = "Power Function for Blood Pressure Test",
       subtitle = "Power increases as true μ moves away from null value") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2))
```

# Connection to Confidence Intervals {background-color="#3498db"}

## CIs and Tests: Two Sides of the Same Coin

::: callout-important
## Key Relationship
A $100(1-\alpha)\%$ confidence interval contains all parameter values that would **not** be rejected by a two-sided test at level $\alpha$.

- If the CI **excludes** the null value $\theta_0$ → Reject $H_0$
- If the CI **includes** the null value $\theta_0$ → Fail to reject $H_0$
:::

## Example: Connecting CI and Test

For our blood pressure example:

```{r}
#| echo: true
# 95% confidence interval (two-sided)
ci_lower <- xbar - qnorm(0.975) * se
ci_upper <- xbar + qnorm(0.975) * se

cat("95% CI for μ: (", round(ci_lower, 2), ",", round(ci_upper, 2), ")\n")
cat("Null value μ₀ = 10\n\n")

if (ci_lower > 10) {
  cat("Since 10 is below the CI, we would reject H₀: μ = 10\n")
  cat("in favor of Hₐ: μ ≠ 10 at α = 0.05")
} else if (ci_upper < 10) {
  cat("Since 10 is above the CI, we would reject H₀: μ = 10")
} else {
  cat("Since 10 is within the CI, we would fail to reject H₀: μ = 10\n")
  cat("at α = 0.05")
}
```

## CI-Test Duality Visualized

```{r}
#| code-fold: true
#| fig-height: 6
# Show CI and null value
tibble(
  x = xbar,
  lower = ci_lower,
  upper = ci_upper
) |>
  ggplot(aes(y = 1)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.1, linewidth = 2, color = "steelblue") +
  geom_point(aes(x = x), size = 5, color = "steelblue") +
  geom_vline(xintercept = 10, linetype = "dashed", color = "red", linewidth = 1.5) +
  annotate("text", x = 10, y = 1.15, label = expression(H[0]: mu == 10), color = "red", size = 6) +
  annotate("text", x = xbar, y = 0.85, label = paste0("x̄ = ", xbar), size = 5) +
  annotate("text", x = (ci_lower + ci_upper)/2, y = 1.1, 
           label = paste0("95% CI: (", round(ci_lower, 1), ", ", round(ci_upper, 1), ")"), 
           size = 5, color = "steelblue") +
  coord_cartesian(xlim = c(7, 18), ylim = c(0.7, 1.3)) +
  labs(x = expression(mu), y = "",
       title = "Confidence Interval and Hypothesis Test",
       subtitle = expression("Null value (10) is below the CI"~symbol("\256")~"Reject"~H[0])) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

# Summary and Key Takeaways {background-color="#2c3e50"}

## The Hypothesis Testing Framework

::: callout-important
## Core Components
1. **Null Hypothesis ($H_0$):** The status quo claim (initially assumed true)
2. **Alternative Hypothesis ($H_a$):** What we want to show (contradicts $H_0$)
3. **Test Statistic:** Function of sample data for making decisions
4. **Rejection Region:** Values of test statistic leading to rejection
5. **Significance Level ($\alpha$):** Probability of Type I error
:::

## The Two Types of Errors

| | $H_0$ True | $H_0$ False |
|---|:---:|:---:|
| Reject $H_0$ | Type I Error ($\alpha$) | Correct (Power) |
| Fail to Reject | Correct | Type II Error ($\beta$) |

. . .

**Key points:**

- We control $\alpha$ by choosing it before the test
- $\beta$ depends on the true parameter value
- Power = $1 - \beta$ = probability of correctly rejecting false $H_0$
- Tradeoff: Can't minimize both errors without increasing $n$

## Connection to Previous Material

| Topic | Lesson | Connection to Testing |
|-------|--------|----------------------|
| Point Estimation | 2-4 | Test statistics are often MLEs |
| Sufficient Statistics | 5 | Best tests use sufficient statistics |
| Confidence Intervals | 7-8 | CIs give all non-rejected values |
| Bootstrap | 9-11 | Bootstrap hypothesis tests use resampling |

## Looking Ahead

**Coming up in this course:**

- **P-values:** Continuous measure of evidence against $H_0$
- **Power and sample size:** How many subjects do we need?
- **Tests for specific parameters:** Mean (z and t), proportion, variance
- **Two-sample tests:** Comparing populations
- **Bootstrap hypothesis tests:** Resampling-based testing

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 9.1
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 3.1-3.2, 8.1-8.2
-   Wasserman, L. (2004). *All of Statistics*. Chapter 10.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
# For PDF export
decktape reveal --fragments docs/lessons/12_Hypothesis_Testing_Introduction/12_Hypothesis_Testing_Introduction.html lessons/12_Hypothesis_Testing_Introduction/12_Hypothesis_Testing_Introduction.pdf
```
