---
title: "Lesson 15: Neyman-Pearson Lemma; Likelihood Ratio Tests"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w7d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 15 - Neyman-Pearson Lemma; Likelihood Ratio Tests
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 15: Neyman-Pearson Lemma; Likelihood Ratio Tests {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 14: P-Values; Power and Sample Size Determination**

-   P-values are Uniform(0,1) under $H_0$ by the probability integral transformation
-   Power depends on the noncentrality parameter $\delta = (\mu' - \mu_0)/(\sigma/\sqrt{n})$
-   Sample size determination via `power.t.test()` for planning clinical trials
-   Multiple testing adjustments: Bonferroni (FWER) and Benjamini-Hochberg (FDR)

. . .

**Today's Goals:**

1. Understand the concept of a **most powerful** test for simple hypotheses
2. State and prove the **Neyman-Pearson Lemma** 
3. Define **power functions** and **uniformly most powerful (UMP)** tests
4. Develop the **likelihood ratio test (LRT)** for composite hypotheses
5. Connect the LRT to familiar tests (z test, t test)
6. Apply the large-sample chi-squared approximation for LRTs

## Roadmap for Today

| Topic | Key Idea | Textbook |
|-------|----------|----------|
| Simple hypotheses | Both $H_0$ and $H_a$ fully specify the distribution | Devore 9.5 |
| Neyman-Pearson Lemma | The likelihood ratio test is most powerful for simple $H_0$ vs simple $H_a$ | Devore 9.5 |
| Power function | $\pi(\theta') = P_{\theta'}(\text{Reject } H_0)$ as a function of $\theta$ | Devore 9.5 |
| UMP tests | Most powerful for every value in the alternative | Devore 9.5 |
| Likelihood ratio tests | General principle for composite hypotheses | Devore 9.5 |
| LRT $\to$ t test | The one-sample t test is a LRT | Devore 9.5 |
| Large-sample LRT | $-2\ln(\Lambda) \xrightarrow{d} \chi^2_m$ | Devore 9.5 |

. . .

::: callout-note
## Why This Matters
The Neyman-Pearson Lemma provides the **theoretical foundation** for optimal hypothesis testing. It tells us *why* the tests we have been using (z tests, t tests) are the best possible tests in their respective settings. The likelihood ratio principle then gives us a **general recipe** for constructing tests in new situations.
:::

# Simple Hypotheses and Optimal Tests (Devore 9.5) {background-color="#3498db"}

## Simple vs. Composite Hypotheses

Recall from Lesson 12:

::: callout-important
## Definition (Simple and Composite Hypotheses)
A hypothesis is **simple** if it completely specifies the distribution of the sample $X_1, \ldots, X_n$. Otherwise, it is **composite**.
:::

. . .

**Examples:**

- $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Exp}(\lambda)$:
  - $H: \lambda = 5$ is **simple** (distribution fully determined)
  - $H: \lambda < 5$ is **composite** (many possible distributions)

. . .

- $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with $\sigma$ **known**:
  - $H: \mu = 100$ is **simple**
  - $H: \mu \neq 100$ is **composite**

. . .

- $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with $\sigma$ **unknown**:
  - $H: \mu = 100$ is **composite** (because $\sigma$ is unspecified --- it is a *nuisance parameter*)

## The Optimality Question

We have developed many test procedures: z tests, t tests, score tests, exact binomial tests...

. . .

**But are these the *best* tests?**

. . .

For a test of $H_0: \theta = \theta_0$ vs $H_a: \theta = \theta_a$, both simple, there are infinitely many possible rejection regions $R$. Each one gives:

- A **Type I error rate**: $\alpha = P_{\theta_0}((X_1, \ldots, X_n) \in R)$
- A **Type II error rate**: $\beta = P_{\theta_a}((X_1, \ldots, X_n) \notin R)$

. . .

**Goal:** Among all tests with $\alpha \leq \alpha^*$ (controlling Type I error), find the test that **minimizes $\beta$** (equivalently, **maximizes power**).

. . .

::: callout-tip
## Medical Research Motivation
In a clinical trial comparing a new drug to a standard, we want the test that gives us the **best chance of detecting a real treatment effect** (high power) while maintaining the required significance level. The Neyman-Pearson Lemma tells us exactly how to construct such a test.
:::

## The Likelihood Ratio as a Test Statistic

**Intuition:** Given data $x_1, \ldots, x_n$, which hypothesis makes the data more likely?

. . .

Consider the ratio:

$$\Lambda^* = \frac{f(x_1, \ldots, x_n; \theta_a)}{f(x_1, \ldots, x_n; \theta_0)} = \frac{L(\theta_a)}{L(\theta_0)}$$

. . .

- If $\Lambda^* \gg 1$: data much more consistent with $H_a$ $\Rightarrow$ evidence against $H_0$
- If $\Lambda^* \approx 1$: data equally consistent with both $\Rightarrow$ no strong evidence
- If $\Lambda^* \ll 1$: data more consistent with $H_0$ $\Rightarrow$ no evidence against $H_0$

. . .

**Natural test:** Reject $H_0$ when $\Lambda^* \geq k$ for some constant $k > 0$.

. . .

::: callout-note
## Convention Note
Devore defines the NP ratio as $L(\theta_a)/L(\theta_0)$ (reject when large). Chihara defines it as $L(\theta_0)/L(\theta_a)$ (reject when small). Both formulations are equivalent --- just inverted. We will primarily follow Devore's convention in the statement of the lemma.
:::

# The Neyman-Pearson Lemma {background-color="#e74c3c"}

## Statement of the Neyman-Pearson Lemma

::: callout-important
## The Neyman-Pearson Lemma (Devore 9.5)
For testing a simple null hypothesis $H_0: \theta = \theta_0$ versus a simple alternative $H_a: \theta = \theta_a$, let $k$ be a fixed positive number and form the rejection region

$$R^* = \left\{(x_1, \ldots, x_n): \frac{f(x_1, \ldots, x_n; \theta_a)}{f(x_1, \ldots, x_n; \theta_0)} \geq k \right\}$$

Let $\alpha^* = P((X_1, \ldots, X_n) \in R^* \text{ when } \theta = \theta_0)$ and let $\beta^*$ denote the Type II error probability.

Then for **any other** test procedure with Type I error probability $\alpha \leq \alpha^*$, the Type II error probability must satisfy $\beta \geq \beta^*$.
:::

. . .

**In plain language:** The test based on the likelihood ratio

- Has the **smallest** $\beta$ (Type II error) among all tests with $\alpha \leq \alpha^*$
- Equivalently, has the **largest power** at $\theta_a$ among all level-$\alpha^*$ tests
- Is called the **most powerful (MP)** test at level $\alpha^*$

## Proof of the Neyman-Pearson Lemma

Let $R$ be the rejection region of **any** test, so that:

$$\alpha = \sum_R f(x_1, \ldots, x_n; \theta_0), \quad \beta = 1 - \sum_R f(x_1, \ldots, x_n; \theta_a)$$

. . .

For any constant $k > 0$, consider the linear combination $k\alpha + \beta$:

$$k\alpha + \beta = k \sum_R f(\mathbf{x}; \theta_0) + 1 - \sum_R f(\mathbf{x}; \theta_a) = 1 + \sum_R \left[k \cdot f(\mathbf{x}; \theta_0) - f(\mathbf{x}; \theta_a)\right]$$

. . .

**Key insight:** The expression $k \cdot f(\mathbf{x}; \theta_0) - f(\mathbf{x}; \theta_a)$ can be positive or negative. To **minimize** $k\alpha + \beta$, choose $R$ to include exactly the points where this expression is negative:

$$R^* = \left\{\mathbf{x}: k \cdot f(\mathbf{x}; \theta_0) - f(\mathbf{x}; \theta_a) \leq 0\right\} = \left\{\mathbf{x}: \frac{f(\mathbf{x}; \theta_a)}{f(\mathbf{x}; \theta_0)} \geq k\right\}$$

. . .

This is exactly the Neyman-Pearson rejection region! So $R^*$ minimizes $k\alpha + \beta$.

## Proof (continued)

We've shown: $k\alpha^* + \beta^* \leq k\alpha + \beta$ for **all** rejection regions $R$.

. . .

Now, for any test with $\alpha \leq \alpha^*$:

$$k\alpha^* + \beta^* \leq k\alpha + \beta$$

$$\beta^* \leq \beta + k(\alpha - \alpha^*) \leq \beta + k \cdot 0 = \beta$$

. . .

since $\alpha - \alpha^* \leq 0$ and $k > 0$.

. . .

Therefore $\beta^* \leq \beta$ for all tests with $\alpha \leq \alpha^*$. $\blacksquare$

. . .

::: callout-note
## What Does This Mean?
The proof works by showing that the NP rejection region simultaneously minimizes **both** $\alpha$ and $\beta$ in a trade-off sense. You cannot do better on one without doing worse on the other. The NP test is on the **efficient frontier** of the $(\alpha, \beta)$ trade-off.
:::

## Example: Adverse Events in a Drug Trial (Normal, Known $\sigma$)

A pharmaceutical company monitors a biomarker level in patients. Under the current treatment, biomarker levels follow $N(\mu, 1)$. An elevated biomarker signals a potential adverse event.

Test $H_0: \mu = 5$ vs $H_a: \mu = 6$ based on $n$ patients (known $\sigma = 1$).

. . .

**Applying the Neyman-Pearson Lemma:**

$$\frac{L(\mu_a = 6)}{L(\mu_0 = 5)} = \frac{\prod_{i=1}^n \frac{1}{\sqrt{2\pi}} e^{-(x_i - 6)^2/2}}{\prod_{i=1}^n \frac{1}{\sqrt{2\pi}} e^{-(x_i - 5)^2/2}} = e^{\sum (x_i - 6)^2/2 \cdot (-1) + \sum(x_i - 5)^2/2}$$

. . .

Simplifying the exponent:

$$= e^{(6 - 5)\sum x_i - n(6^2 - 5^2)/2} = e^{\sum x_i - 11n/2}$$

## Example: Normal (continued)

The rejection region $\Lambda^* \geq k$ becomes:

$$e^{\sum x_i - 11n/2} \geq k$$

. . .

Taking logarithms (since $\mu_a - \mu_0 = 1 > 0$, the exponential is increasing in $\sum x_i$):

$$\sum x_i \geq k' \quad \iff \quad \bar{x} \geq k'' \quad \iff \quad \frac{\bar{x} - 5}{1/\sqrt{n}} \geq c$$

. . .

Setting $c = z_\alpha$, this is exactly the **one-sample z test**!

. . .

::: callout-important
## Key Result (Devore Example 9.24)
For $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with $\sigma$ known, testing $H_0: \mu = \mu_0$ vs $H_a: \mu = \mu_a$ where $\mu_a > \mu_0$:

The Neyman-Pearson Lemma shows that the **one-sample z test** (reject when $z \geq z_\alpha$) is the **most powerful** test at level $\alpha$. Our familiar z test is optimal!
:::

## Example: Poisson Adverse Event Counts (Devore Example 9.23)

In post-market drug surveillance, the number of adverse events per reporting period is monitored. Suppose $X_1, \ldots, X_5 \stackrel{iid}{\sim} \text{Poisson}(\lambda)$ represent adverse event counts over 5 periods. 

Test $H_0: \lambda = 1$ (baseline rate) vs $H_a: \lambda = 2$ (elevated rate).

. . .

**The Poisson likelihood is:** $f(x_1, \ldots, x_5; \lambda) = e^{-5\lambda} \lambda^{\sum x_i} / \prod x_i!$

. . .

**The likelihood ratio:**

$$\frac{L(\lambda = 2)}{L(\lambda = 1)} = \frac{e^{-10} \cdot 2^{\sum x_i}}{e^{-5} \cdot 1^{\sum x_i}} = e^{-5} \cdot 2^{\sum x_i} \geq k$$

. . .

Taking logarithms: $\sum x_i \geq c$ where $c = \ln(k \cdot e^5)/\ln(2)$.

## Example: Poisson (continued)

**Finding the critical value:** Under $H_0$, $Y = \sum X_i \sim \text{Poisson}(5)$.

. . .

If we use $c = 10$: $\alpha^* = P(Y \geq 10 \mid Y \sim \text{Poisson}(5)) = 1 - 0.968 = 0.032$

If we use $c = 9$: $\alpha^* = P(Y \geq 9 \mid Y \sim \text{Poisson}(5)) = 1 - 0.932 = 0.068$

. . .

For $\alpha \leq 0.05$, we use $c = 10$, giving $\alpha^* = 0.032$.

. . .

**Type II error:** Under $H_a$, $Y \sim \text{Poisson}(10)$.

$$\beta^* = P(Y < 10 \mid Y \sim \text{Poisson}(10)) = 0.458$$

. . .

```{r}
#| echo: true
#| code-fold: true
# Verify with R
cat("alpha* =", 1 - ppois(9, lambda = 5), "\n")
cat("beta*  =", ppois(9, lambda = 10), "\n")
cat("Power  =", 1 - ppois(9, lambda = 10), "\n")
```

## Example: Poisson (continued)

**The Neyman-Pearson guarantee:** Any other test with $\alpha \leq 0.032$ must have $\beta \geq 0.458$.

. . .

The power is only $1 - 0.458 = 0.542$. Why so low?

. . .

**Because $n = 5$ is too small** to discriminate between $\lambda = 1$ and $\lambda = 2$.

For $n = 10$: best test with $\alpha \leq 0.05$ uses $c = 16$, giving $\alpha^* = 0.049$ and $\beta^* = 0.157$ (power = 0.843).

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5
#| fig-width: 14
# Compare rejection regions for n=5 and n=10
tibble(y = 0:30) |>
  mutate(
    `n = 5, Poisson(5)` = dpois(y, 5),
    `n = 5, Poisson(10)` = dpois(y, 10),
    `n = 10, Poisson(10)` = dpois(y, 10),
    `n = 10, Poisson(20)` = dpois(y, 20)
  ) |>
  pivot_longer(-y, names_to = "distribution", values_to = "prob") |>
  separate(distribution, into = c("sample_size", "dist"), sep = ", ") |>
  ggplot(aes(x = y, y = prob, fill = dist)) +
  geom_col(position = "identity", alpha = 0.5, width = 0.8) +
  geom_vline(data = tibble(sample_size = c("n = 5", "n = 10"), 
                            cutoff = c(10, 16)),
             aes(xintercept = cutoff - 0.5), linetype = "dashed", 
             color = "red", linewidth = 1) +
  facet_wrap(~sample_size, scales = "free") +
  scale_fill_manual(values = c("steelblue", "coral", "steelblue", "coral")) +
  labs(x = expression(y == sum(x[i])), y = "Probability",
       title = "Neyman-Pearson Test: Poisson Adverse Event Monitoring",
       subtitle = "Red dashed line = rejection boundary; larger n gives better separation") +
  theme(legend.position = "bottom")
```

## Medical Example: Exponential Survival Times (C&H Example 8.18)

In an oncology trial, patient survival times follow $\text{Exp}(\lambda)$. We want to test whether the hazard rate has increased from a historical baseline.

Test $H_0: \lambda = 8$ vs $H_a: \lambda = 10$ based on $n = 9$ patients.

. . .

**Likelihood ratio:**

$$\frac{L(\lambda = 8)}{L(\lambda = 10)} = \frac{8^9 e^{-8\sum x_i}}{10^9 e^{-10\sum x_i}} = (0.8)^9 e^{2\sum x_i}$$

. . .

(Note: Chihara uses $L(\theta_0)/L(\theta_a)$, rejecting for small values.)

Reject $H_0$ when this ratio is small, i.e., when $\sum x_i < c_2$.

. . .

Under $H_0$: $\sum_{i=1}^9 X_i \sim \text{Gamma}(9, 8)$. For $\alpha = 0.05$:

```{r}
#| echo: true
c2 <- qgamma(0.05, shape = 9, rate = 8)
cat("Critical value c2 =", round(c2, 4), "\n")
```

By the Neyman-Pearson Lemma, this is the **best possible** test at this significance level.

# Power Functions and UMP Tests {background-color="#27ae60"}

## The Power Function

::: callout-important
## Definition (Power Function, Devore 9.5)
Let $\Omega_0$ and $\Omega_a$ be two disjoint sets of possible values of $\theta$. For testing $H_0: \theta \in \Omega_0$ vs $H_a: \theta \in \Omega_a$ with rejection region $R$, the **power function** is:

$$\pi(\theta') = P_{\theta'}((X_1, \ldots, X_n) \in R)$$

the probability of rejecting $H_0$ as a function of $\theta$.
:::

. . .

**Connection to error probabilities:**

$$\pi(\theta') = \begin{cases} P(\text{Type I error}) = \alpha(\theta') & \text{when } \theta' \in \Omega_0 \\[4pt] 1 - P(\text{Type II error}) = 1 - \beta(\theta') & \text{when } \theta' \in \Omega_a \end{cases}$$

. . .

**Ideal power function** (unachievable in practice):

$$\pi(\theta') = \begin{cases} 0 & \text{when } \theta' \in \Omega_0 \\ 1 & \text{when } \theta' \in \Omega_a \end{cases}$$

## Visualizing the Power Function

**Example (Devore 9.25 adapted):** Blood pressure reduction (mmHg) with a new antihypertensive follows $N(\mu, 81)$. Test $H_0: \mu \leq 75$ vs $H_a: \mu < 75$ with $n = 100$, $\alpha = 0.10$.

Rejection region: $z \leq -z_{0.10} = -1.28$, i.e., $\bar{x} \leq 73.848$.

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
# Power function for the z test
sigma <- 9
n <- 100
se <- sigma / sqrt(n)
x_crit <- 75 - 1.28 * se  # 73.848

mu_prime <- seq(68, 78, length.out = 500)
power_actual <- pnorm((x_crit - mu_prime) / se)

power_df <- tibble(mu = mu_prime, power = power_actual)

ggplot(power_df, aes(x = mu, y = power)) +
  geom_line(linewidth = 1.3, color = "steelblue") +
  # Ideal power function
  geom_segment(aes(x = 68, xend = 75, y = 1, yend = 1), 
               linetype = "dashed", color = "coral", linewidth = 1) +
  geom_segment(aes(x = 75, xend = 78, y = 0, yend = 0), 
               linetype = "dashed", color = "coral", linewidth = 1) +
  geom_vline(xintercept = 75, linetype = "dotted", color = "gray50") +
  geom_hline(yintercept = 0.10, linetype = "dotted", color = "gray50") +
  annotate("text", x = 76.5, y = 0.6, label = "Ideal", color = "coral", size = 6) +
  annotate("text", x = 70, y = 0.6, label = "Actual", color = "steelblue", size = 6) +
  annotate("text", x = 77, y = 0.12, 
           label = expression(alpha == 0.10), size = 5, color = "gray30") +
  annotate("text", x = 75.3, y = 0.5, 
           label = expression(mu[0] == 75), size = 5, color = "gray30") +
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = expression(mu * "' (true mean)"), y = expression(pi(mu * "'")),
       title = "Power Function for the Lower-Tailed z Test",
       subtitle = bquote("Testing" ~ H[0]*":" ~ mu <= 75 ~ "vs" ~ H[a]*":" ~ mu < 75 ~ "with n = 100," ~ sigma == 9*"," ~ alpha == 0.10))
```

## Properties of the Power Function

From the figure, we observe several important properties:

. . .

1. **At the boundary** ($\mu = 75$): $\pi(75) = \alpha = 0.10$ by design

. . .

2. **In $\Omega_0$** ($\mu > 75$): $\pi(\mu') < 0.10$ --- power is below $\alpha$ (good!)

. . .

3. **Near the boundary** ($\mu$ slightly below 75): power is barely above $\alpha$ --- small departures are hard to detect

. . .

4. **Far from $H_0$** ($\mu \ll 75$): $\pi(\mu') \to 1$ --- large departures are easily detected

. . .

5. **As $n$ increases:** the actual power function approaches the ideal step function

. . .

::: callout-tip
## Clinical Implication
In designing a clinical trial, we need to choose $n$ large enough so that the power function is acceptably close to 1 at the **minimum clinically important difference**. This connects directly to the sample size calculations from Lesson 14.
:::

## Uniformly Most Powerful (UMP) Tests

The NP Lemma handles simple vs simple. What about composite alternatives?

. . .

::: callout-important
## Definition (Uniformly Most Powerful Test)
A test is **uniformly most powerful (UMP)** at level $\alpha$ if it maximizes $\pi(\theta')$ for **every** $\theta' \in \Omega_a$, subject to $\pi(\theta') \leq \alpha$ for all $\theta' \in \Omega_0$.
:::

. . .

**When do UMP tests exist?** If the NP rejection region **does not depend on the specific alternative** $\theta_a$, then the same test is most powerful for *every* $\theta_a \in \Omega_a$.

## UMP Example: Poisson (Devore Example 9.26)

Returning to the Poisson adverse event example with $n = 5$:

Test $H_0: \lambda \leq 1$ vs $H_a: \lambda > 1$.

. . .

For **any** specific $\lambda_a > 1$, the NP test of $H_0: \lambda = 1$ vs $H_a: \lambda = \lambda_a$ rejects when $\sum x_i \geq c$.

. . .

**Key observation:** The rejection region $\sum x_i \geq 10$ is the **same** regardless of which $\lambda_a > 1$ we choose! (The critical value $c$ is determined by $H_0$ alone.)

. . .

Therefore, this test is **UMP** at level $\alpha^* = 0.032$.

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5
#| fig-width: 14
# Power function for the UMP Poisson test
lambda_prime <- seq(0, 5, length.out = 300)
power_poisson <- 1 - ppois(9, lambda = 5 * lambda_prime)

ggplot(tibble(lambda = lambda_prime, power = power_poisson), 
       aes(x = lambda, y = power)) +
  geom_line(linewidth = 1.3, color = "steelblue") +
  geom_vline(xintercept = 1, linetype = "dotted", color = "gray50") +
  geom_hline(yintercept = 0.032, linetype = "dashed", color = "coral") +
  annotate("text", x = 0.5, y = 0.08, 
           label = expression(alpha * "* = 0.032"), color = "coral", size = 5) +
  annotate("text", x = 1.15, y = 0.5, 
           label = expression(lambda[0] == 1), color = "gray40", size = 5) +
  labs(x = expression(lambda * "' (true Poisson rate)"), 
       y = expression(pi(lambda * "'")),
       title = bquote("Power Function of the UMP Test for" ~ H[0]*":" ~ lambda <= 1),
       subtitle = bquote("Reject when" ~ sum(X[i], i==1, 5) >= 10 ~ "; maximizes power for every" ~ lambda > 1))
```

## When UMP Tests Don't Exist

::: callout-warning
## Important Limitation
UMP tests are **fairly rare**, especially in the commonly encountered situation of testing a two-sided alternative.
:::

. . .

**Example:** Testing $H_0: \mu = \mu_0$ vs $H_a: \mu \neq \mu_0$ with $X_i \stackrel{iid}{\sim} N(\mu, \sigma^2)$, $\sigma$ known.

- For $\mu_a > \mu_0$: NP test rejects when $\bar{x}$ is **large** (upper tail)
- For $\mu_a < \mu_0$: NP test rejects when $\bar{x}$ is **small** (lower tail)
- These are **different rejection regions** --- no single test is best for all $\mu_a \neq \mu_0$

. . .

**Resolution:** Restrict attention to **unbiased tests** (tests where $\pi(\theta') \geq \alpha$ for all $\theta' \in \Omega_a$). Under this restriction:

- The two-sided z test is **UMP unbiased** when $\sigma$ is known
- The one-sample t test is **UMP unbiased** when $\sigma$ is unknown

## Unbiased Tests

::: callout-note
## Definition (Unbiased Test)
A test is **unbiased** if the power under any alternative is at least as large as the power under any null value:

$$\inf_{\theta' \in \Omega_a} \pi(\theta') \geq \sup_{\theta' \in \Omega_0} \pi(\theta')$$

Equivalently: we are at least as likely to reject $H_0$ when it is false as when it is true.
:::

. . .

**Why is this desirable?** A biased test would reject $H_0$ *less* often for some alternatives than it does when $H_0$ is true --- that would defeat the purpose of the test!

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5
#| fig-width: 14
# Biased vs unbiased test illustration
mu_seq <- seq(-4, 4, length.out = 500)
n_ex <- 16; sigma_ex <- 4; se_ex <- sigma_ex / sqrt(n_ex)

# Unbiased (symmetric) rejection region
power_unbiased <- pnorm((-1.96 * se_ex - mu_seq) / se_ex) + 
                  1 - pnorm((1.96 * se_ex - mu_seq) / se_ex)

# Biased (asymmetric) rejection region: reject if x_bar > 2.17 or x_bar < -1.81
power_biased <- pnorm((-1.81 * se_ex - mu_seq) / se_ex) + 
                1 - pnorm((2.17 * se_ex - mu_seq) / se_ex)

tibble(mu = mu_seq, Unbiased = power_unbiased, Biased = power_biased) |>
  pivot_longer(-mu, names_to = "Test", values_to = "Power") |>
  ggplot(aes(x = mu, y = Power, color = Test)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("coral", "steelblue")) +
  annotate("text", x = 3.5, y = 0.08, label = expression(alpha == 0.05), 
           color = "gray50", size = 5) +
  labs(x = expression(mu * "'"), y = expression(pi(mu * "'")),
       title = "Unbiased vs Biased Tests for H0: \u03bc = 0 vs Ha: \u03bc \u2260 0",
       subtitle = "The biased test has power < alpha for some alternatives near mu = 0") +
  theme(legend.position = c(0.85, 0.5))
```

# Likelihood Ratio Tests (Devore 9.5) {background-color="#8e44ad"}

## From Simple to Composite Hypotheses

The NP Lemma handles simple vs simple hypotheses. But in practice, we need tests for:

$$H_0: \theta \in \Omega_0 \quad \text{vs} \quad H_a: \theta \in \Omega_a$$

where one or both hypotheses may be composite.

. . .

**The likelihood ratio principle** provides a **general recipe** for constructing test statistics in virtually any testing situation.

. . .

::: callout-important
## The Likelihood Ratio Test (LRT) Procedure

1. Find $\hat{\theta}_{mle}$, the MLE of $\theta$ over the **full** parameter space $\Omega = \Omega_0 \cup \Omega_a$
2. Find $\hat{\theta}_0$, the **restricted MLE** of $\theta$ over $\Omega_0$ only
3. Form the **likelihood ratio statistic**:

$$\Lambda = \frac{L(\hat{\theta}_0)}{L(\hat{\theta}_{mle})}$$

4. Reject $H_0$ when $\Lambda \leq k$ (i.e., when the restricted MLE fits much worse than the unrestricted MLE)
:::

## Understanding the LRT Statistic

$$\Lambda = \frac{L(\hat{\theta}_0)}{L(\hat{\theta}_{mle})} = \frac{\max_{\theta \in \Omega_0} L(\theta)}{\max_{\theta \in \Omega} L(\theta)}$$

. . .

**Key properties:**

- Always $0 \leq \Lambda \leq 1$ (because $\Omega_0 \subseteq \Omega$, so the restricted maximum $\leq$ unrestricted maximum)
- $\Lambda \approx 1$: the data are about equally consistent with $H_0$ and the overall best fit
- $\Lambda \approx 0$: the data are **much** more consistent with $H_a$ than with $H_0$

. . .

**Connection to NP Lemma:** When both hypotheses are simple ($\Omega_0 = \{\theta_0\}$ and $\Omega_a = \{\theta_a\}$):

$$\Lambda = \frac{L(\theta_0)}{\max\{L(\theta_0), L(\theta_a)\}} = \begin{cases} L(\theta_0)/L(\theta_a) & \text{if } L(\theta_a) \geq L(\theta_0) \\[2pt] 1 & \text{otherwise} \end{cases}$$

This is the reciprocal of the NP statistic (when $H_a$ is more likely), so rejecting for small $\Lambda$ is equivalent to rejecting for large $\Lambda^*$.

## LRT Example 1: Normal Mean, Known Variance (Devore Example 9.27 style)

Systolic blood pressure readings follow $N(\mu, 100)$ (known $\sigma = 10$). A clinical trial tests whether a new treatment changes the population mean from a target value.

$$H_0: \mu = \mu_0 \quad \text{vs} \quad H_a: \mu \neq \mu_0$$

. . .

**Step 1: Unrestricted MLE.** Over $\Omega = \mathbb{R}$: $\hat{\mu}_{mle} = \bar{x}$

. . .

**Step 2: Restricted MLE.** Over $\Omega_0 = \{\mu_0\}$: $\hat{\mu}_0 = \mu_0$

. . .

**Step 3: Likelihood ratio.**

$$\Lambda = \frac{L(\mu_0)}{L(\bar{x})} = \frac{\prod \frac{1}{\sqrt{2\pi}\sigma} e^{-(x_i - \mu_0)^2/(2\sigma^2)}}{\prod \frac{1}{\sqrt{2\pi}\sigma} e^{-(x_i - \bar{x})^2/(2\sigma^2)}} = e^{-\frac{1}{2\sigma^2}\left[\sum(x_i - \mu_0)^2 - \sum(x_i - \bar{x})^2\right]}$$

. . .

Since $\sum(x_i - \mu_0)^2 = \sum(x_i - \bar{x})^2 + n(\bar{x} - \mu_0)^2$:

$$\Lambda = e^{-n(\bar{x} - \mu_0)^2 / (2\sigma^2)}$$

## LRT $\Rightarrow$ z Test

Reject when $\Lambda \leq k$:

$$e^{-n(\bar{x} - \mu_0)^2/(2\sigma^2)} \leq k$$

. . .

Taking $-\ln$ of both sides:

$$\frac{n(\bar{x} - \mu_0)^2}{2\sigma^2} \geq -\ln(k) = c'$$

. . .

$$\left|\frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}}\right| \geq c$$

. . .

This is the **two-sided z test**! Setting $c = z_{\alpha/2}$ gives the test at level $\alpha$.

. . .

::: callout-important
## Key Result
The **likelihood ratio test** for $H_0: \mu = \mu_0$ vs $H_a: \mu \neq \mu_0$ when $\sigma$ is known is exactly the **two-sided z test**. The LRT principle automatically "discovers" the z test.
:::

## LRT Example 2: Normal Mean, Unknown Variance (Devore Example 9.27)

Now suppose $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with $\sigma$ **unknown**.

$$H_0: \mu = \mu_0 \quad \text{vs} \quad H_a: \mu \neq \mu_0$$

. . .

**Step 1: Unrestricted MLE** over $\Omega = \{(\mu, \sigma^2): \mu \in \mathbb{R}, \sigma^2 > 0\}$:

$$\hat{\mu}_{mle} = \bar{x}, \quad \hat{\sigma}^2_{mle} = \frac{1}{n}\sum(x_i - \bar{x})^2$$

. . .

**Step 2: Restricted MLE** over $\Omega_0 = \{(\mu_0, \sigma^2): \sigma^2 > 0\}$:

$$\hat{\mu}_0 = \mu_0, \quad \hat{\sigma}^2_0 = \frac{1}{n}\sum(x_i - \mu_0)^2$$

. . .

Note: under $H_0$, we still estimate $\sigma^2$ by MLE, but with $\mu$ fixed at $\mu_0$.

## LRT $\Rightarrow$ t Test (Devore Example 9.27)

**Step 3: Likelihood ratio.** After substitution and simplification:

$$\Lambda = \left[\frac{\hat{\sigma}^2_{mle}}{\hat{\sigma}^2_0}\right]^{n/2} = \left[\frac{\sum(x_i - \bar{x})^2}{\sum(x_i - \mu_0)^2}\right]^{n/2}$$

. . .

Since $\sum(x_i - \mu_0)^2 = \sum(x_i - \bar{x})^2 + n(\bar{x} - \mu_0)^2$:

$$\Lambda = \left[\frac{1}{1 + \frac{n(\bar{x} - \mu_0)^2}{\sum(x_i - \bar{x})^2}}\right]^{n/2} = \left[\frac{1}{1 + t^2/(n-1)}\right]^{n/2}$$

where $t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}$ is the **t statistic**.

. . .

**Reject** when $\Lambda$ is small $\iff$ $t^2/(n-1)$ is large $\iff$ $|t| \geq c$.

Setting $c = t_{\alpha/2, n-1}$ gives the **one-sample t test**.

. . .

::: callout-important
## Key Result
The **likelihood ratio test** for $H_0: \mu = \mu_0$ vs $H_a: \mu \neq \mu_0$ when $\sigma$ is unknown is exactly the **one-sample t test**. Many classical tests can be derived from the LRT principle.
:::

## LRT and the t Test: Simulation Illustration

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 14
# Demonstrate that the LRT statistic is a monotone function of |t|
set.seed(551)
n_sim <- 25
mu_0 <- 120  # target systolic BP
# Simulated clinical trial data: true mu = 125
x_data <- rnorm(n_sim, mean = 125, sd = 15)

x_bar <- mean(x_data)
s <- sd(x_data)
t_stat <- (x_bar - mu_0) / (s / sqrt(n_sim))

# Compute Lambda for a range of hypothetical t values
t_range <- seq(-5, 5, length.out = 500)
lambda_vals <- (1 / (1 + t_range^2 / (n_sim - 1)))^(n_sim / 2)

p1 <- ggplot(tibble(t = t_range, Lambda = lambda_vals), aes(x = t, y = Lambda)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_vline(xintercept = c(-qt(0.025, n_sim - 1), qt(0.025, n_sim - 1)),
             linetype = "dashed", color = "coral") +
  geom_point(data = tibble(t = t_stat, 
                            Lambda = (1 / (1 + t_stat^2 / (n_sim - 1)))^(n_sim / 2)),
             size = 4, color = "darkred") +
  annotate("text", x = t_stat - 0.3, y = 0.6, 
           label = paste0("Observed t = ", round(t_stat, 2)), 
           color = "darkred", size = 5, hjust = 1) +
  labs(x = "t statistic", y = expression(Lambda),
       title = expression(Lambda == (1 + t^2/(n-1))^{-n/2}),
       subtitle = expression("Smaller " * Lambda * " = stronger evidence against " * H[0]))

# Show the relationship between -2ln(Lambda) and t^2
neg2lnL <- -n_sim * log(1 + t_range^2 / (n_sim - 1))
p2 <- ggplot(tibble(t2 = t_range^2, neg2lnL = -neg2lnL), 
             aes(x = t2, y = neg2lnL)) +
  geom_line(linewidth = 1.2, color = "purple") +
  geom_vline(xintercept = qt(0.975, n_sim - 1)^2, 
             linetype = "dashed", color = "coral") +
  labs(x = expression(t^2), y = expression(-2 * ln(Lambda)),
       title = expression(-2 * ln(Lambda) ~ "vs" ~ t^2),
       subtitle = "Monotone increasing: small Lambda = large |t|")

p1 + p2
```

## Medical Application: Drug Efficacy Trial

A Phase III clinical trial tests whether a new cholesterol-lowering drug changes mean LDL from baseline $\mu_0 = 130$ mg/dL. Data from $n = 30$ patients:

```{r}
#| echo: true
set.seed(42)
n <- 30
mu_0 <- 130
ldl_data <- rnorm(n, mean = 122, sd = 18)

x_bar <- mean(ldl_data)
s_val <- sd(ldl_data)
t_obs <- (x_bar - mu_0) / (s_val / sqrt(n))

cat("Sample mean:", round(x_bar, 2), "mg/dL\n")
cat("Sample SD:  ", round(s_val, 2), "mg/dL\n")
cat("t statistic:", round(t_obs, 3), "\n")
```

. . .

```{r}
#| echo: true
# LRT statistic
Lambda <- (1 / (1 + t_obs^2 / (n - 1)))^(n / 2)
neg2lnLambda <- -2 * log(Lambda)

cat("Lambda:       ", round(Lambda, 6), "\n")
cat("-2 ln(Lambda):", round(neg2lnLambda, 3), "\n")
cat("t^2:         ", round(t_obs^2, 3), "\n")
cat("P-value (t):  ", round(2 * pt(-abs(t_obs), df = n - 1), 4), "\n")
```

. . .

```{r}
#| echo: true
# Verify with t.test
t.test(ldl_data, mu = mu_0) |> tidy() |> 
  select(estimate, statistic, p.value, conf.low, conf.high) |>
  mutate(across(everything(), ~round(., 4)))
```

# Large-Sample LRT Approximation {background-color="#f39c12"}

## Wilks' Theorem: The Chi-Squared Approximation

In many situations, the LRT rejection region $\Lambda \leq k$ cannot be expressed in terms of a simple, familiar statistic. A powerful large-sample result resolves this.

. . .

::: callout-important
## Theorem (Wilks' Theorem, Devore 9.5)
If the sample size $n$ is sufficiently large, then under $H_0$:

$$-2\ln(\Lambda) \xrightarrow{d} \chi^2_m$$

where $m$ is the difference between the number of free parameters in $\Omega$ and the number of free parameters in $\Omega_0$.
:::

. . .

**How to use it:** Since $0 \leq \Lambda \leq 1$, we have $-2\ln(\Lambda) \geq 0$. Reject $H_0$ when:

$$-2\ln(\Lambda) \geq \chi^2_{\alpha, m}$$

. . .

**The degrees of freedom $m$:**

- $m = \dim(\Omega) - \dim(\Omega_0)$ = (number of parameters estimated under $H_a$) minus (number estimated under $H_0$)
- Represents the number of "constraints" imposed by $H_0$

## Computing Degrees of Freedom: Examples

| Setting | $\Omega$ | $\Omega_0$ | $m$ |
|---------|----------|------------|-----|
| $N(\mu, \sigma^2)$, $\sigma$ known, $H_0: \mu = \mu_0$ | 1 ($\mu$) | 0 | 1 |
| $N(\mu, \sigma^2)$, $\sigma$ unknown, $H_0: \mu = \mu_0$ | 2 ($\mu, \sigma^2$) | 1 ($\sigma^2$) | 1 |
| Bivariate normal, $H_0: \mu_1 = \mu_2$ and $\sigma_1 = \sigma_2$ | 5 ($\mu_1, \mu_2, \sigma_1, \sigma_2, \rho$) | 3 ($\mu, \sigma, \rho$) | 2 |
| Multinomial with $k$ categories, $H_0: p_i = p_{i,0}$ | $k - 1$ | 0 | $k - 1$ |

. . .

::: callout-tip
## Practical Rule
Count the number of parameters that are "free" (need to be estimated) under each hypothesis. The difference is $m$.
:::

## Large-Sample LRT Example: Double Exponential (Devore 9.5)

The Laplace (double exponential) distribution with pdf $f(x; \theta) = \frac{1}{2}e^{-|x - \theta|}$ is sometimes used to model measurement errors. Test $H_0: \theta = 0$ vs $H_a: \theta \neq 0$.

. . .

**Under $\Omega$:** The MLE of $\theta$ is the **sample median** $\tilde{x}$ (minimizes $\sum |x_i - \theta|$).

**Under $\Omega_0$:** $\theta = 0$ (no estimation needed).

. . .

$$\Lambda = \frac{(1/2)^n e^{-\sum|x_i|}}{(1/2)^n e^{-\sum|x_i - \tilde{x}|}} = e^{-(\sum|x_i| - \sum|x_i - \tilde{x}|)}$$

. . .

$$-2\ln(\Lambda) = 2\left(\sum|x_i| - \sum|x_i - \tilde{x}|\right)$$

. . .

**Degrees of freedom:** $m = 1 - 0 = 1$, so $-2\ln(\Lambda) \stackrel{\cdot}{\sim} \chi^2_1$ under $H_0$.

Reject $H_0$ when $-2\ln(\Lambda) \geq \chi^2_{\alpha, 1}$.

## Simulation: Verifying the Chi-Squared Approximation

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
# Simulate LRT for normal mean (known variance) and check chi-squared approximation
set.seed(551)
n_obs <- 30
n_rep <- 10000
mu_0 <- 0
sigma_known <- 1

# Simulate under H_0
neg2lnL_sims <- replicate(n_rep, {
  x <- rnorm(n_obs, mean = mu_0, sd = sigma_known)
  x_bar <- mean(x)
  # -2ln(Lambda) = n * (x_bar - mu_0)^2 / sigma^2 for known variance case
  n_obs * (x_bar - mu_0)^2 / sigma_known^2
})

# Compare to chi-squared(1)
chi2_theoretical <- seq(0.01, 12, length.out = 500)

p1 <- ggplot(tibble(stat = neg2lnL_sims), aes(x = stat)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60, 
                 fill = "steelblue", alpha = 0.6) +
  geom_line(data = tibble(x = chi2_theoretical, y = dchisq(chi2_theoretical, df = 1)),
            aes(x = x, y = y), color = "coral", linewidth = 1.2) +
  labs(x = expression(-2 * ln(Lambda)), y = "Density",
       title = expression("Simulated " * -2 * ln(Lambda) * " vs " * chi[1]^2 * 
                         " (n = 30, known " * sigma * ")"),
       subtitle = expression("Red curve: " * chi^2 * "(1) density"))

# Q-Q plot
p2 <- ggplot(tibble(stat = neg2lnL_sims), aes(sample = stat)) +
  stat_qq(distribution = qchisq, dparams = list(df = 1), 
          color = "steelblue", alpha = 0.3, size = 0.8) +
  stat_qq_line(distribution = qchisq, dparams = list(df = 1), 
               color = "coral", linewidth = 1) +
  labs(x = expression("Theoretical " * chi[1]^2 * " quantiles"),
       y = expression("Sample quantiles of " * -2 * ln(Lambda)),
       title = "Q-Q Plot: Chi-Squared Approximation",
       subtitle = "Close alignment confirms Wilks' theorem")

p1 + p2
```

## How Good is the Approximation for Small $n$?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
# Check chi-squared approximation quality for different n
# Normal mean, unknown variance -> -2ln(Lambda) = n*ln(1 + t^2/(n-1))
set.seed(551)
n_rep <- 10000
sample_sizes <- c(5, 10, 30, 100)

sim_results <- map_dfr(sample_sizes, function(n_val) {
  stats <- replicate(n_rep, {
    x <- rnorm(n_val, mean = 0, sd = 1)
    t_stat <- mean(x) / (sd(x) / sqrt(n_val))
    n_val * log(1 + t_stat^2 / (n_val - 1))
  })
  tibble(n = paste0("n = ", n_val), stat = stats)
})

sim_results |>
  mutate(n = factor(n, levels = paste0("n = ", sample_sizes))) |>
  ggplot(aes(x = stat)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, 
                 fill = "steelblue", alpha = 0.5) +
  geom_line(data = tibble(x = rep(chi2_theoretical, 4),
                          y = rep(dchisq(chi2_theoretical, 1), 4),
                          n = rep(paste0("n = ", sample_sizes), 
                                  each = length(chi2_theoretical))) |>
              mutate(n = factor(n, levels = paste0("n = ", sample_sizes))),
            aes(x = x, y = y), color = "coral", linewidth = 1) +
  facet_wrap(~n, scales = "free_y") +
  coord_cartesian(xlim = c(0, 10)) +
  labs(x = expression(-2 * ln(Lambda)), y = "Density",
       title = expression("Quality of " * chi^2 * " Approximation for the t Test LRT"),
       subtitle = expression("Red: " * chi[1]^2 * " density. Approximation improves with n"))
```

# Connecting the Pieces {background-color="#2c3e50"}

## The Big Picture: Hypothesis Testing Framework

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 16
library(DiagrammeR)

grViz("
digraph testing_framework {
  graph [rankdir=TB, bgcolor=transparent, fontsize=20]
  node [shape=box, style=filled, fontsize=16, fontname='Helvetica']
  
  # Top level
  A [label='Hypothesis Testing Problem\\nH₀: θ ∈ Ω₀ vs Hₐ: θ ∈ Ωₐ', 
     fillcolor='#3498db', fontcolor=white, width=4]
  
  # Decision node
  B [label='Are both hypotheses simple?', 
     fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3]
  
  # Simple path
  C [label='Neyman-Pearson Lemma\\nLikelihood ratio: L(θₐ)/L(θ₀) ≥ k\\nMost Powerful Test', 
     fillcolor='#27ae60', fontcolor=white, width=3.5]
  
  # Composite path
  D [label='Likelihood Ratio Test (LRT)\\nΛ = L(θ̂₀) / L(θ̂ₘₗₑ) ≤ k', 
     fillcolor='#8e44ad', fontcolor=white, width=3.5]
  
  # UMP
  E [label='Same rejection region\\nfor all θₐ ∈ Ωₐ?', 
     fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3]
  F [label='UMP Test\\n(one-sided alternatives)', 
     fillcolor='#27ae60', fontcolor=white, width=3]
  G [label='No UMP exists\\n(two-sided alternatives)\\nUse UMP Unbiased or LRT', 
     fillcolor='#e74c3c', fontcolor=white, width=3.5]
  
  # LRT outcomes
  H [label='Exact distribution\\nknown?', 
     fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3]
  I [label='Use exact test\\n(z test, t test, F test)', 
     fillcolor='#27ae60', fontcolor=white, width=3]
  J [label='Large-sample approx.\\n-2ln(Λ) ~ χ²ₘ', 
     fillcolor='#16a085', fontcolor=white, width=3]
  
  A -> B
  B -> C [label='  Yes', fontsize=14]
  B -> D [label='  No', fontsize=14]
  C -> E
  E -> F [label='  Yes', fontsize=14]
  E -> G [label='  No', fontsize=14]
  D -> H
  H -> I [label='  Yes', fontsize=14]
  H -> J [label='  No', fontsize=14]
}
")
```

## Summary: Classical Tests as LRTs

| Test | Hypotheses | LRT Produces | Exact or Approx? |
|------|-----------|-------------|-------------------|
| One-sample z | $H_0: \mu = \mu_0$, $\sigma$ known | $z = \frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}}$ | Exact |
| One-sample t | $H_0: \mu = \mu_0$, $\sigma$ unknown | $t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}$ | Exact |
| Pooled two-sample t | $H_0: \mu_1 = \mu_2$, equal $\sigma$ | Pooled $t$ statistic | Exact |
| Chi-squared goodness-of-fit | $H_0: p_i = p_{i,0}$ | $\sum (O_i - E_i)^2/E_i$ | Large-sample |
| F test (ANOVA) | $H_0: \mu_1 = \cdots = \mu_k$ | $F$ ratio | Exact (normal) |

. . .

::: callout-tip
## Key Insight
The LRT principle is a **unifying framework**. Many of the "named" tests from introductory statistics --- z tests, t tests, F tests, chi-squared tests --- can all be derived as likelihood ratio tests. The LRT gives us a principled way to construct new tests for situations not covered by standard formulas.
:::

## Comparison of Approaches: When to Use What

**Neyman-Pearson Lemma:**

- Applies to **simple vs simple** hypotheses
- Guarantees **optimality** (most powerful test)
- Useful for proving that known tests are optimal

. . .

**UMP Tests:**

- Extend NP to **one-sided composite** alternatives
- Exist mainly for one-parameter exponential families with one-sided alternatives
- Do **not** exist for two-sided alternatives in general

. . .

**Likelihood Ratio Tests:**

- Apply to **any** testing problem (simple or composite, one or multiple parameters)
- **General-purpose recipe** that always yields a reasonable test
- Often coincide with optimal tests when those exist
- Large-sample $\chi^2$ approximation makes them practical even when exact distributions are unknown

# Key Takeaways and Looking Ahead {background-color="#2c3e50"}

## Key Takeaways

::: callout-important
## Today's Main Points

1. **Simple hypotheses** completely specify the distribution; composite hypotheses do not
2. **Neyman-Pearson Lemma:** the likelihood ratio test $L(\theta_a)/L(\theta_0) \geq k$ is the most powerful test for simple $H_0$ vs simple $H_a$
3. **Power function** $\pi(\theta') = P_{\theta'}(\text{Reject } H_0)$ characterizes test performance across all parameter values
4. **UMP tests** maximize power for *every* alternative; they exist for one-sided tests in exponential families
5. **Likelihood ratio test** $\Lambda = L(\hat{\theta}_0)/L(\hat{\theta}_{mle})$ provides a general recipe for composite hypotheses
6. **LRT $\Rightarrow$ classical tests:** the z test and t test are both LRTs
7. **Wilks' theorem:** $-2\ln(\Lambda) \xrightarrow{d} \chi^2_m$ provides a large-sample approximation
:::

. . .

**Common mistakes to avoid:**

- Confusing the NP ratio $L(\theta_a)/L(\theta_0)$ (reject when large) with the LRT ratio $L(\hat{\theta}_0)/L(\hat{\theta}_{mle})$ (reject when small)
- Assuming UMP tests always exist (they don't for two-sided alternatives)
- Using the $\chi^2$ approximation when $n$ is too small
- Forgetting that the LRT degrees of freedom $m$ count *free* parameters, not total parameters

## Looking Ahead

**Next class (Lesson 16):** Further Aspects of Hypothesis Testing; Bootstrap Hypothesis Testing (Devore 9.6, C&H Chapter 8)

- Statistical vs practical significance
- Relationship between confidence intervals and hypothesis tests
- Bootstrap hypothesis testing
- Permutation tests

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 9.5
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Section 8.4
-   Neyman, J. & Pearson, E. S. (1933). On the problem of the most efficient tests of statistical hypotheses. *Philosophical Transactions of the Royal Society A*, 231, 289--337.
-   Wilks, S. S. (1938). The large-sample distribution of the likelihood ratio for testing composite hypotheses. *The Annals of Mathematical Statistics*, 9(1), 60--62.
:::

## Questions? {.center background-color="#3498db"}

Thank you!


```{bash}
#| eval: false
#| include: false
# For PDF export
decktape reveal --fragments docs/lessons/15_NeymanPearson_LRT/15_NeymanPearson_LRT.html lessons/15_NeymanPearson_LRT/15_NeymanPearson_LRT.pdf
```
