---
title: "Lesson 9: Introduction to the Bootstrap"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w4d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 9 - Bootstrap Introduction
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 9: The Bootstrap {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 8: One-Sample t-Intervals and Population Proportions**

-   t-intervals when $\sigma$ is unknown: $\bar{x} \pm t_{\alpha/2, n-1}\frac{s}{\sqrt{n}}$
-   Normality assumption for t-intervals
-   Confidence intervals for proportions (Wald, Score, Agresti-Coull)

. . .

**The Problem:** What if normality doesn't hold and the sample size is small?

. . .

**Today's Goals:**

1. Understand the bootstrap principle and its theoretical foundation
2. Learn why resampling **with replacement** is essential
3. Implement the basic bootstrap algorithm in R
4. Explore asymptotic properties of bootstrap estimates
5. Understand when bootstrap methods fail

## Motivating Example: Drug Response Times

In a clinical trial, 12 patients' response times (days) to a new medication:

```{r}
response_times <- c(8, 12, 15, 7, 22, 45, 11, 9, 14, 18, 10, 13)

tibble(x = response_times) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 8, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = mean(response_times), color = "red", linewidth = 1.2, linetype = "dashed") +
  labs(x = "Response Time (days)", y = "Count",
       title = "Distribution of Drug Response Times") +
  annotate("text", x = mean(response_times) + 5, y = 3, 
           label = paste0("Mean = ", round(mean(response_times), 1)), color = "red", size = 5)
```

## The Problem with Small Samples

```{r}
#| code-fold: true
#| fig-height: 5
# Q-Q plot to check normality
qqnorm(response_times, main = "Normal Q-Q Plot", pch = 19, col = "steelblue")
qqline(response_times, col = "red", lwd = 2)
```

. . .

**Issues:**

- Small sample (n = 12) 
- Evidence of right skew (outlier at 45 days)
- Normal-based t-interval may not have correct coverage

## The Key Question

How can we construct confidence intervals when:

1. The population distribution is **unknown or non-normal**?
2. The sample size is **too small** for CLT to apply reliably?
3. We want to estimate parameters **other than the mean** (median, trimmed mean, ratio)?

. . .

::: callout-tip
## The Bootstrap Solution
**Use the data itself** to approximate the sampling distribution of our statistic!
:::

# The Bootstrap Idea {background-color="#3498db"}

## What is the Bootstrap?

::: callout-important
## Definition (Devore, Berk, Carlton 8.5)
The **bootstrap** is a resampling method that uses the observed sample to approximate the sampling distribution of a statistic.

The method was developed by **Bradley Efron** in the late 1970s and substitutes heavy computation for theoretical derivation.
:::

. . .

**Key insight:** The sample is our best estimate of the population. If we resample *from the sample*, we can learn about sampling variability.

## Traditional vs. Bootstrap Approach

**Traditional Inference:**

1. Assume population distribution (e.g., Normal)
2. Derive sampling distribution theoretically
3. Use theoretical distribution to construct intervals

. . .

**Bootstrap Inference:**

1. Use the sample as a proxy for the population
2. Generate sampling distribution empirically via resampling
3. Use empirical distribution for inference

## The Plug-In Principle (Chihara & Hesterberg 5.2)

::: callout-important
## The Plug-In Principle
To estimate a parameter (a quantity that describes the **population**), use the statistic that is the corresponding quantity for the **sample**.
:::

. . .

**Examples:**

| Population Parameter | Sample Statistic |
|---------------------|------------------|
| Mean $\mu$ | $\bar{x}$ |
| Standard deviation $\sigma$ | $s$ |
| Median $\tilde{\mu}$ | Sample median |
| Standard error $\sigma/\sqrt{n}$ | $s/\sqrt{n}$ |

. . .

**The bootstrap extends this:** Use the **entire sample** as an estimate of the **entire population**.

## The Empirical Distribution Function

::: callout-important
## Definition: Empirical CDF
Given a sample $x_1, x_2, \ldots, x_n$, the **empirical cumulative distribution function** is:
$$\hat{F}_n(x) = \frac{1}{n}\#\{x_i \leq x\} = \frac{1}{n}\sum_{i=1}^{n} \mathbf{1}(x_i \leq x)$$

This assigns probability $1/n$ to each observed data point.
:::

. . .

**Key properties of $\hat{F}_n$:**

- $E_{\hat{F}}[X] = \bar{x}$ (mean under empirical distribution = sample mean)
- $\text{Var}_{\hat{F}}[X] = \frac{1}{n}\sum(x_i - \bar{x})^2$ (slightly different from $s^2$)

## Visualizing the Bootstrap Idea

```{r}
#| code-fold: true
#| fig-height: 7
#| fig-width: 14
set.seed(551)
# Simulated "population"
pop <- rgamma(10000, shape = 2, rate = 0.1)

# Take a sample
n <- 30
samp <- sample(pop, n)

# Generate bootstrap samples
boot_means <- replicate(10000, mean(sample(samp, n, replace = TRUE)))

p1 <- tibble(x = pop) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "darkgreen", alpha = 0.6) +
  geom_vline(xintercept = mean(pop), color = "darkgreen", linewidth = 1.5) +
  labs(title = "Population (Unknown in Practice)", 
       subtitle = bquote(mu == .(round(mean(pop), 2))),
       x = "Value", y = "Count") +
  theme(plot.title = element_text(size = 16))

p2 <- tibble(x = samp) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(samp), color = "steelblue", linewidth = 1.5) +
  labs(title = "Our Sample (What We Have)", 
       subtitle = bquote(bar(x) == .(round(mean(samp), 2))),
       x = "Value", y = "Count") +
  theme(plot.title = element_text(size = 16))

p3 <- tibble(x = boot_means) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  geom_vline(xintercept = mean(boot_means), color = "coral", linewidth = 1.5) +
  labs(title = "Bootstrap Distribution of Means", 
       subtitle = paste0("Mean = ", round(mean(boot_means), 2), ", SE = ", round(sd(boot_means), 2)),
       x = expression(bar(x)^"*"), y = "Count") +
  theme(plot.title = element_text(size = 16))

p1 + p2 + p3
```

## Sampling Distribution vs. Bootstrap Distribution

::: columns
::: {.column width="50%"}
**Sampling Distribution:**

- Draw **many samples** from the **population**
- Calculate statistic for each
- Distribution of these statistics

*In practice:* We can't do this (we only have one sample!)
:::

::: {.column width="50%"}
**Bootstrap Distribution:**

- Draw **many resamples** from the **sample**
- Calculate statistic for each
- Distribution of these statistics

*In practice:* We can do this computationally!
:::
:::

. . .

::: callout-note
## Key Point
The bootstrap distribution **approximates** the sampling distribution, allowing us to estimate standard errors and construct confidence intervals.
:::

# Why Sampling WITH Replacement? {background-color="#8e44ad"}

## The Critical Question

Why must we sample **with replacement** rather than **without replacement** (subsampling)?

. . .

::: callout-warning
## Without Replacement
If we sample **without replacement** from our sample of size $n$, we would get:

- A subsample of size $m < n$: Less information, different properties
- A sample of size $n$: **Exactly the same sample every time!**
:::

. . .

**Sampling with replacement allows:**

1. Observations to appear multiple times (or not at all)
2. Variability in the bootstrap samples
3. Approximation of the sampling process from the population

## Mathematical Justification

::: callout-important
## Why Replacement Matters
When sampling from a population of size $N$, we sample with replacement because:

1. The population is treated as **infinite** (or very large relative to $n$)
2. Each draw is **independent** of previous draws
3. Each observation has the **same probability** of selection

The empirical distribution $\hat{F}_n$ is a **discrete** distribution with $n$ equally likely values. Sampling from it **with replacement** mimics sampling from the true population.
:::

## Demonstrating: With vs. Without Replacement

```{r}
#| echo: true
#| code-fold: true
set.seed(551)
original_sample <- c(3, 5, 7, 9, 11)
n <- length(original_sample)

# With replacement: many possible samples
cat("With replacement (5 examples):\n")
for(i in 1:5) {
  boot_samp <- sample(original_sample, n, replace = TRUE)
  cat("  Sample", i, ":", boot_samp, "-> mean =", mean(boot_samp), "\n")
}

cat("\nWithout replacement:\n")
cat("  Always get:", sample(original_sample, n, replace = FALSE), "\n")
```

## Number of Possible Bootstrap Samples

::: callout-note
## Counting Bootstrap Samples
For a sample of size $n$, the number of possible bootstrap samples (sampling $n$ with replacement) is:
$$\binom{2n-1}{n-1} = \binom{2n-1}{n}$$

This is the number of ways to choose $n$ items from $n$ categories with repetition allowed.
:::

```{r}
#| echo: true
# Number of bootstrap samples for various n
tibble(n = c(5, 10, 15, 20, 30)) |>
  mutate(
    num_bootstrap_samples = choose(2*n - 1, n - 1),
    formatted = format(num_bootstrap_samples, big.mark = ",", scientific = FALSE)
  )
```

Even for $n = 15$, there are over **77 million** possible bootstrap samples!

## The Variance Connection

::: callout-important
## Variance of Bootstrap Mean
For the sample mean, the **theoretical** (exhaustive) bootstrap standard error is:
$$SE_{boot}(\bar{X}^*) = \frac{\hat{\sigma}}{\sqrt{n}} = \sqrt{\frac{1}{n^2}\sum_{i=1}^{n}(x_i - \bar{x})^2}$$

where $\hat{\sigma}^2 = \frac{1}{n}\sum(x_i - \bar{x})^2$ is the population variance under $\hat{F}_n$.
:::

. . .

**Note:** This differs slightly from $s/\sqrt{n}$ by a factor of $\sqrt{(n-1)/n}$:
$$SE_{boot} = \sqrt{\frac{n-1}{n}} \cdot \frac{s}{\sqrt{n}}$$

This is why bootstrap SEs are slightly **narrower** than classical SEs.

## Subsampling: An Alternative Approach

::: callout-note
## Subsampling (Without Replacement)
An alternative to bootstrap is **subsampling**: draw samples of size $m < n$ **without replacement**.

**Properties:**

- Requires choosing subsample size $m$
- Typically $m/n \to 0$ as $n \to \infty$
- Can work when bootstrap fails (heavy tails, dependent data)
- Generally less efficient than bootstrap
:::

. . .

**When to use subsampling:**

- Time series or dependent data
- Estimating extreme quantiles
- When bootstrap consistency fails

# The Bootstrap Method {background-color="#9b59b6"}

## Basic Bootstrap Algorithm (Devore 8.5)

::: callout-important
## Basic Bootstrap Method
Given a sample $x_1, x_2, \ldots, x_n$ and a statistic $\hat{\theta}$:

1. **Resample:** Draw a sample of size $n$ **with replacement** from $x_1, \ldots, x_n$
2. **Compute:** Calculate $\hat{\theta}^*$ from this bootstrap sample
3. **Repeat:** Repeat steps 1-2 many times (typically $B = 1000$ to $10000$)
4. **Analyze:** Use the distribution of $\hat{\theta}^*_1, \hat{\theta}^*_2, \ldots, \hat{\theta}^*_B$ for inference
:::

. . .

**Why sample with replacement?**

Without replacement, we'd get the same sample every time!

## Simple Example: Birth Weights (Chihara 5.1)

Consider just 3 birth weights: 3969, 3204, 2892 grams

```{r}
#| echo: true
small_sample <- c(3969, 3204, 2892)
cat("Original sample:", small_sample, "\n")
cat("Sample mean:", mean(small_sample))
```

. . .

**All possible bootstrap samples** (sampling with replacement):

```{r}
#| code-fold: true
# All 3^3 = 27 bootstrap samples
expand_grid(x1 = small_sample, x2 = small_sample, x3 = small_sample) |>
  mutate(
    sample_id = row_number(),
    mean = (x1 + x2 + x3) / 3
  ) |>
  head(10)
```

## Complete Bootstrap Distribution (Small Example)

```{r}
#| code-fold: true
#| fig-height: 5.5
all_boot_samples <- expand_grid(x1 = small_sample, x2 = small_sample, x3 = small_sample) |>
  mutate(boot_mean = (x1 + x2 + x3) / 3)

all_boot_samples |>
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = mean(small_sample), color = "red", linewidth = 1.2, linetype = "dashed") +
  scale_x_continuous(breaks = sort(unique(all_boot_samples$boot_mean))) +
  labs(x = expression(bar(x)^"*"), y = "Frequency",
       title = "Bootstrap Distribution of Mean (All 27 Samples)",
       subtitle = paste0("Original sample mean = ", round(mean(small_sample), 1))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Bootstrap standard error: `r round(sd(all_boot_samples$boot_mean), 1)` grams

## Bootstrap in R: Step by Step

```{r}
#| echo: true
# Original response time data
response_times <- c(8, 12, 15, 7, 22, 45, 11, 9, 14, 18, 10, 13)
n <- length(response_times)
B <- 10000  # Number of bootstrap samples

# Storage for bootstrap means
boot_means <- numeric(B)

# Bootstrap loop
for (i in 1:B) {
  # Step 1: Resample with replacement
  boot_sample <- sample(response_times, size = n, replace = TRUE)
  
  # Step 2: Calculate statistic
  boot_means[i] <- mean(boot_sample)
}

# Results
cat("Original sample mean:", round(mean(response_times), 2), "\n")
cat("Bootstrap mean:", round(mean(boot_means), 2), "\n")
cat("Bootstrap standard error:", round(sd(boot_means), 2))
```

## Bootstrap Distribution: Drug Response Times

```{r}
#| code-fold: true
#| fig-height: 6
tibble(x = boot_means) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", color = "white", alpha = 0.7) +
  geom_vline(xintercept = mean(response_times), color = "blue", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean(boot_means), color = "red", linewidth = 1, linetype = "dotted") +
  labs(x = expression(bar(x)^"*"), y = "Count",
       title = "Bootstrap Distribution of Mean Response Time",
       subtitle = paste0("B = ", B, " resamples; SE = ", round(sd(boot_means), 2), " days")) +
  annotate("text", x = 12, y = 800, 
           label = paste0("Original mean = ", round(mean(response_times), 1)), 
           color = "blue", size = 5)
```

## Using Tidyverse for Bootstrap

```{r}
#| echo: true
# More elegant approach using tidyverse
set.seed(551)
boot_results <- tibble(
  replicate = 1:B
) |>
  mutate(
    boot_sample = map(replicate, ~sample(response_times, n, replace = TRUE)),
    boot_mean = map_dbl(boot_sample, mean),
    boot_median = map_dbl(boot_sample, median),
    boot_sd = map_dbl(boot_sample, sd))
boot_results |>
  select(-boot_sample) |>
  summarize(
    across(starts_with("boot_"), list(mean = mean, se = sd))
  ) |>
  pivot_longer(everything(), names_to = "statistic", values_to = "value")
```

# Asymptotic Properties of Bootstrap {background-color="#e67e22"}

## Bootstrap Consistency

::: callout-important
## Bootstrap Consistency Theorem
Under regularity conditions, the bootstrap distribution consistently estimates the sampling distribution. Specifically:

Let $\hat{\theta}_n$ be an estimator of $\theta$ based on a sample of size $n$. If:
$$\sqrt{n}(\hat{\theta}_n - \theta) \xrightarrow{d} N(0, \sigma^2)$$

Then under the bootstrap:
$$\sqrt{n}(\hat{\theta}_n^* - \hat{\theta}_n) \xrightarrow{d} N(0, \sigma^2)$$

where convergence is in probability.
:::

. . .

**In plain English:** The bootstrap distribution mimics the true sampling distribution as $n \to \infty$.

## What Bootstrap Estimates Well

::: callout-tip
## Bootstrap Accurately Estimates (Chihara 5.2.2)

1. **Spread:** The spread (standard error) of the sampling distribution
2. **Shape:** The shape (including skewness) of the sampling distribution
3. **Bias:** The bias of the estimator
:::

. . .

::: callout-warning
## Bootstrap Does NOT Estimate

**Center:** The bootstrap distribution is centered at $\hat{\theta}$ (the sample statistic), NOT at $\theta$ (the true parameter).

This means bootstrap **cannot improve point estimates**!
:::

## Visualizing What Bootstrap Captures

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 12
set.seed(551)
# True population: Gamma(2, 0.1)
true_mean <- 20
true_se <- 20 / sqrt(50)

# One sample and its bootstrap
samp <- rgamma(50, shape = 2, rate = 0.1)
sample_mean <- mean(samp)

# Bootstrap distribution
boot_dist <- replicate(10000, mean(sample(samp, 50, replace = TRUE)))

# Sampling distribution (simulated)
sampling_dist <- replicate(10000, mean(rgamma(50, shape = 2, rate = 0.1)))

p1 <- tibble(x = sampling_dist) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "darkgreen", alpha = 0.6) +
  geom_vline(xintercept = true_mean, color = "darkgreen", linewidth = 1.5) +
  labs(title = "True Sampling Distribution",
       subtitle = bquote("Centered at" ~ mu == .(true_mean) * "," ~ SE == .(round(sd(sampling_dist), 2))),
       x = expression(bar(X)), y = "Count")

p2 <- tibble(x = boot_dist) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.6) +
  geom_vline(xintercept = sample_mean, color = "coral", linewidth = 1.5) +
  labs(title = "Bootstrap Distribution",
       subtitle = bquote("Centered at" ~ bar(x) == .(round(sample_mean, 2)) * "," ~ SE == .(round(sd(boot_dist), 2))),
       x = expression(bar(X)^"*"), y = "Count")

p1 + p2
```

**Key observation:** Same spread and shape, but different centers!

## Comparison Table (Chihara Table 5.2)

```{r}
#| echo: false
tibble(
  Quantity = c("Population", "Sampling distribution of X̄", "Sample", "Bootstrap distribution"),
  Mean = c("μ", "μ", "x̄", "≈ x̄"),
  `Standard Deviation` = c("σ", "σ/√n", "s", "≈ s·√((n-1)/n)/√n")
) |>
  knitr::kable(align = "lcc")
```

. . .

**Key insights:**

- Bootstrap mean ≈ sample mean (not population mean)
- Bootstrap SE ≈ sampling distribution SE
- Bootstrap shape ≈ sampling distribution shape

## Rate of Convergence

::: callout-note
## Edgeworth Expansion Refinement
For the sample mean, classical confidence intervals have coverage error of order $O(n^{-1/2})$.

**Bootstrap percentile intervals:** Also $O(n^{-1/2})$ — same as classical

**Bootstrap t intervals:** $O(n^{-1})$ — **one order more accurate!**

This is because the bootstrap t automatically corrects for skewness.
:::

. . .

**Practical implication:** Bootstrap t intervals are preferred for skewed distributions, especially with moderate sample sizes.

# When Bootstrap Fails {background-color="#c0392b"}

## Bootstrap Failure Modes

The bootstrap is **not** a universal solution. It can fail when:

1. **Very small samples** ($n < 10$)
2. **Sample doesn't represent population**
3. **Estimating extreme quantiles**
4. **Discrete statistics** (e.g., median with few unique values)
5. **Heavy-tailed distributions** (infinite variance)
6. **Dependent data** (time series, spatial data)
7. **Lattice distributions** with small support

## Failure Mode 1: Very Small Samples

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
set.seed(551)
# Small sample from normal
small_n <- 6
small_samp <- rnorm(small_n, mean = 10, sd = 3)

# Multiple bootstrap distributions from same sample
boot_dists <- map(1:4, function(seed) {
  set.seed(seed)
  replicate(1000, mean(sample(small_samp, small_n, replace = TRUE)))
})

# True sampling distribution
true_sampling <- replicate(10000, mean(rnorm(small_n, 10, 3)))

plots <- map2(boot_dists, 1:4, function(bd, i) {
  tibble(x = bd) |>
    ggplot(aes(x = x)) +
    geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
    geom_vline(xintercept = mean(small_samp), color = "blue", linewidth = 1) +
    labs(title = paste("Bootstrap", i), 
         subtitle = paste0("SE = ", round(sd(bd), 2)),
         x = expression(bar(x)^"*"), y = "") +
    coord_cartesian(xlim = c(6, 14))
})

wrap_plots(plots, ncol = 4)
```

**Problem:** Bootstrap distributions vary substantially due to small sample size.

## Failure Mode 2: The Median

```{r}
#| code-fold: true
#| fig-height: 5.5
set.seed(551)
# Sample for median bootstrap
med_sample <- rnorm(15, mean = 0, sd = 1)

# Bootstrap the median
boot_medians <- replicate(10000, median(sample(med_sample, 15, replace = TRUE)))

tibble(x = boot_medians) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 30, fill = "purple", alpha = 0.7) +
  geom_vline(xintercept = median(med_sample), color = "blue", linewidth = 1.2) +
  labs(x = expression(tilde(x)^"*"), y = "Count",
       title = "Bootstrap Distribution of Median (n = 15)",
       subtitle = "Notice the discrete, irregular shape")
```

**Problem:** Bootstrap medians only take values from the original sample!

## Why Median Bootstrap Fails

::: callout-warning
## The Median Problem (Chihara 5.8.3)
For odd sample size $n$, the bootstrap median can only take values that are **observations in the original sample**.

**Consequences:**

- Bootstrap distribution is **discrete**, not continuous
- Shape depends heavily on **gaps between observations**
- May poorly approximate the continuous sampling distribution
:::

. . .

**Remedies:**

- **Smoothed bootstrap:** Add small noise to observations
- **Larger samples:** Problem diminishes as $n$ increases
- Use percentile intervals rather than bootstrap t

## Failure Mode 3: Heavy-Tailed Distributions

::: callout-important
## Cauchy Distribution Example
The Cauchy distribution has **no mean** (undefined). For such distributions:

- Sample mean is **not consistent**
- Bootstrap cannot "fix" an inconsistent estimator
- Bootstrap distribution doesn't converge properly
:::

```{r}
#| echo: true
# Cauchy has no mean - bootstrap can't help
set.seed(551)
cauchy_sample <- rcauchy(50)
boot_cauchy <- replicate(1000, mean(sample(cauchy_sample, 50, replace = TRUE)))

cat("Sample mean:", round(mean(cauchy_sample), 2), "\n")
cat("Bootstrap SE:", round(sd(boot_cauchy), 2), " (meaningless!)")
```

## Visualizing Heavy-Tail Failure

```{r}
#| code-fold: true
#| fig-height: 5
# Compare bootstrap for normal vs Cauchy
set.seed(551)
normal_boot <- replicate(5000, mean(sample(rnorm(30), 30, replace = TRUE)))
cauchy_boot <- replicate(5000, mean(sample(rcauchy(30), 30, replace = TRUE)))

p1 <- tibble(x = normal_boot) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Bootstrap Means: Normal(0,1)", x = expression(bar(x)^"*")) +
  coord_cartesian(xlim = c(-1, 1))

p2 <- tibble(x = cauchy_boot) |>
  filter(abs(x) < 50) |>  # Remove extreme values for visualization
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  labs(title = "Bootstrap Means: Cauchy \n(truncated)", x = expression(bar(x)^"*"))

p1 + p2
```

For Cauchy, the bootstrap distribution is erratic and doesn't converge!

## Failure Mode 4: Extreme Quantiles

::: callout-warning
## Estimating Extreme Quantiles
Bootstrap has difficulty with extreme quantiles (e.g., 99th percentile) because:

1. Few observations near the tail
2. Bootstrap samples may not include extreme values
3. Standard bootstrap underestimates tail variability
:::

```{r}
#| echo: true
# Bootstrap for 95th percentile
set.seed(551)
sample_data <- rexp(50, rate = 0.1)
boot_q95 <- replicate(5000, quantile(sample(sample_data, 50, replace = TRUE), 0.95))

cat("Sample 95th percentile:", round(quantile(sample_data, 0.95), 2), "\n")
cat("Bootstrap SE:", round(sd(boot_q95), 2))
```

. . .

For extreme quantiles, consider **parametric bootstrap** or **extreme value theory**.

## Summary: When Bootstrap Works and Fails

```{r}
#| echo: false
tibble(
  `Works Well` = c("Sample mean", "Trimmed mean", "Correlation", "Regression coefficients", 
                   "Smooth functions of means"),
  `Use Caution` = c("Median (small n)", "Extreme quantiles", "Small samples (n < 15)", 
                    "Heavily skewed data", "Discrete distributions"),
  `May Fail` = c("Heavy tails (no variance)", "Dependent data", "Non-representative sample",
                 "Maximum/minimum", "Lattice distributions")
) |>
  knitr::kable()
```

# Summary and Looking Ahead {background-color="#2c3e50"}

## Key Takeaways: The Bootstrap Principle

::: callout-important
## The Bootstrap Principle
1. **The sample represents the population** — resample from the sample
2. **Sampling with replacement** mimics drawing new samples
3. **Bootstrap distribution approximates sampling distribution**
4. Use bootstrap to estimate **standard errors** and the **shape** of the sampling distribution
:::

. . .

**Main advantages:**

- No distributional assumptions required
- Works for **any statistic** (medians, ratios, etc.)
- Computationally straightforward

## Why Resampling WITH Replacement?

::: callout-note
## Key Points
1. **Without replacement:** Would get the same sample every time (size $n$)
2. **With replacement:** Creates variability that mimics sampling from population
3. **Empirical distribution:** Assigns probability $1/n$ to each observation
4. **Number of bootstrap samples:** $\binom{2n-1}{n}$ possible — astronomically large!
:::

## Asymptotic Properties Summary

::: callout-tip
## Bootstrap Consistency
Under regularity conditions:

- Bootstrap distribution **converges** to sampling distribution
- Bootstrap SE **consistently estimates** true SE
- Bootstrap t intervals have coverage error $O(n^{-1})$ — better than percentile!
:::

. . .

**The bootstrap accurately estimates:**

- Spread (standard error)
- Shape (including skewness)
- Bias

**But NOT the center** (stays at $\hat{\theta}$, not $\theta$)

## When Bootstrap Fails: Summary

| Failure Mode | Reason | Alternative |
|--------------|--------|-------------|
| Very small $n$ | Sample doesn't represent population | Classical methods |
| Median/quantiles | Discrete bootstrap distribution | Smoothed bootstrap |
| Heavy tails | Infinite variance | Subsampling |
| Extreme quantiles | Few observations in tails | Parametric bootstrap |
| Dependent data | Independence assumption violated | Block bootstrap |

## Looking Ahead

**Next Lesson: Bootstrap Confidence Intervals**

- Bootstrap standard errors
- Bootstrap percentile intervals
- Bootstrap t intervals
- Practical considerations and examples

. . .

**Coming Soon:**

- Hypothesis testing framework
- Type I and Type II errors
- Power and sample size

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 8.5
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapters 5
-   Efron, B. and Tibshirani, R. (1993). *An Introduction to the Bootstrap*. Chapman & Hall.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/09_Bootstrap_Introduction/09_Bootstrap_Introduction.html lessons/09_Bootstrap_Introduction/09_Bootstrap_Introduction.pdf
```
