---
title: "Lesson 9: Introduction to Bootstrap for Statistical Inference"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w4d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 9 - Bootstrap Introduction
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 9: The Bootstrap {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 8: One-Sample t-Intervals and Population Proportions**

-   t-intervals when $\sigma$ is unknown: $\bar{x} \pm t_{\alpha/2, n-1}\frac{s}{\sqrt{n}}$
-   Normality assumption for t-intervals
-   Confidence intervals for proportions (Wald, Score, Agresti-Coull)

. . .

**The Problem:** What if normality doesn't hold and the sample size is small?

. . .

**Today's Goals:**

1. Understand the bootstrap principle and its theoretical foundation
2. Learn why resampling **with replacement** is essential
3. Explore asymptotic properties of bootstrap estimates
4. Understand when bootstrap methods fail
5. Construct bootstrap percentile and t confidence intervals

## Motivating Example: Drug Response Times

In a clinical trial, 12 patients' response times (days) to a new medication:

```{r}
response_times <- c(8, 12, 15, 7, 22, 45, 11, 9, 14, 18, 10, 13)

tibble(x = response_times) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 8, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = mean(response_times), color = "red", linewidth = 1.2, linetype = "dashed") +
  labs(x = "Response Time (days)", y = "Count",
       title = "Distribution of Drug Response Times") +
  annotate("text", x = mean(response_times) + 5, y = 3, 
           label = paste0("Mean = ", round(mean(response_times), 1)), color = "red", size = 5)
```

## The Problem with Small Samples

```{r}
#| code-fold: true
#| fig-height: 5
# Q-Q plot to check normality
qqnorm(response_times, main = "Normal Q-Q Plot", pch = 19, col = "steelblue")
qqline(response_times, col = "red", lwd = 2)
```

. . .

**Issues:**

- Small sample (n = 12) 
- Evidence of right skew (outlier at 45 days)
- Normal-based t-interval may not have correct coverage

## The Key Question

How can we construct confidence intervals when:

1. The population distribution is **unknown or non-normal**?
2. The sample size is **too small** for CLT to apply reliably?
3. We want to estimate parameters **other than the mean** (median, trimmed mean, ratio)?

. . .

::: callout-tip
## The Bootstrap Solution
**Use the data itself** to approximate the sampling distribution of our statistic!
:::

# The Bootstrap Idea {background-color="#3498db"}

## What is the Bootstrap?

::: callout-important
## Definition (Devore, Berk, Carlton 8.5)
The **bootstrap** is a resampling method that uses the observed sample to approximate the sampling distribution of a statistic.

The method was developed by **Bradley Efron** in the late 1970s and substitutes heavy computation for theoretical derivation.
:::

. . .

**Key insight:** The sample is our best estimate of the population. If we resample *from the sample*, we can learn about sampling variability.

## Traditional vs. Bootstrap Approach

**Traditional Inference:**

1. Assume population distribution (e.g., Normal)
2. Derive sampling distribution theoretically
3. Use theoretical distribution to construct intervals

. . .

**Bootstrap Inference:**

1. Use the sample as a proxy for the population
2. Generate sampling distribution empirically via resampling
3. Use empirical distribution for inference

## The Plug-In Principle (Chihara & Hesterberg 5.2)

::: callout-important
## The Plug-In Principle
To estimate a parameter (a quantity that describes the **population**), use the statistic that is the corresponding quantity for the **sample**.
:::

. . .

**Examples:**

| Population Parameter | Sample Statistic |
|---------------------|------------------|
| Mean $\mu$ | $\bar{x}$ |
| Standard deviation $\sigma$ | $s$ |
| Median $\tilde{\mu}$ | Sample median |
| Standard error $\sigma/\sqrt{n}$ | $s/\sqrt{n}$ |

. . .

**The bootstrap extends this:** Use the **entire sample** as an estimate of the **entire population**.

## The Empirical Distribution Function

::: callout-important
## Definition: Empirical CDF
Given a sample $x_1, x_2, \ldots, x_n$, the **empirical cumulative distribution function** is:
$$\hat{F}_n(x) = \frac{1}{n}\#\{x_i \leq x\} = \frac{1}{n}\sum_{i=1}^{n} \mathbf{1}(x_i \leq x)$$

This assigns probability $1/n$ to each observed data point.
:::

. . .

**Key properties of $\hat{F}_n$:**

- $E_{\hat{F}}[X] = \bar{x}$ (mean under empirical distribution = sample mean)
- $\text{Var}_{\hat{F}}[X] = \frac{1}{n}\sum(x_i - \bar{x})^2$ (slightly different from $s^2$)

## Visualizing the Bootstrap Idea

```{r}
#| code-fold: true
#| fig-height: 7
#| fig-width: 14
library(patchwork)

set.seed(551)
# Simulated "population"
pop <- rgamma(10000, shape = 2, rate = 0.1)

# Take a sample
n <- 30
samp <- sample(pop, n)

# Generate bootstrap samples
boot_means <- replicate(10000, mean(sample(samp, n, replace = TRUE)))

p1 <- tibble(x = pop) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "darkgreen", alpha = 0.6) +
  geom_vline(xintercept = mean(pop), color = "darkgreen", linewidth = 1.5) +
  labs(title = "Population (Unknown in Practice)", 
       subtitle = bquote(mu == .(round(mean(pop), 2))),
       x = "Value", y = "Count") +
  theme(plot.title = element_text(size = 16))

p2 <- tibble(x = samp) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(samp), color = "steelblue", linewidth = 1.5) +
  labs(title = "Our Sample (What We Have)", 
       subtitle = bquote(bar(x) == .(round(mean(samp), 2))),
       x = "Value", y = "Count") +
  theme(plot.title = element_text(size = 16))

p3 <- tibble(x = boot_means) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  geom_vline(xintercept = mean(boot_means), color = "coral", linewidth = 1.5) +
  labs(title = "Bootstrap Distribution of Means", 
       subtitle = paste0("Mean = ", round(mean(boot_means), 2), ", SE = ", round(sd(boot_means), 2)),
       x = expression(bar(x)^"*"), y = "Count") +
  theme(plot.title = element_text(size = 16))

p1 + p2 + p3
```

## Sampling Distribution vs. Bootstrap Distribution

::: columns
::: {.column width="50%"}
**Sampling Distribution:**

- Draw **many samples** from the **population**
- Calculate statistic for each
- Distribution of these statistics

*In practice:* We can't do this (we only have one sample!)
:::

::: {.column width="50%"}
**Bootstrap Distribution:**

- Draw **many resamples** from the **sample**
- Calculate statistic for each
- Distribution of these statistics

*In practice:* We can do this computationally!
:::
:::

. . .

::: callout-note
## Key Point
The bootstrap distribution **approximates** the sampling distribution, allowing us to estimate standard errors and construct confidence intervals.
:::

# Why Sampling WITH Replacement? {background-color="#8e44ad"}

## The Critical Question

Why must we sample **with replacement** rather than **without replacement** (subsampling)?

. . .

::: callout-warning
## Without Replacement
If we sample **without replacement** from our sample of size $n$, we would get:

- A subsample of size $m < n$: Less information, different properties
- A sample of size $n$: **Exactly the same sample every time!**
:::

. . .

**Sampling with replacement allows:**

1. Observations to appear multiple times (or not at all)
2. Variability in the bootstrap samples
3. Approximation of the sampling process from the population

## Mathematical Justification

::: callout-important
## Why Replacement Matters
When sampling from a population of size $N$, we sample with replacement because:

1. The population is treated as **infinite** (or very large relative to $n$)
2. Each draw is **independent** of previous draws
3. Each observation has the **same probability** of selection

The empirical distribution $\hat{F}_n$ is a **discrete** distribution with $n$ equally likely values. Sampling from it **with replacement** mimics sampling from the true population.
:::

## Demonstrating: With vs. Without Replacement

```{r}
#| echo: true
#| code-fold: true
set.seed(551)
original_sample <- c(3, 5, 7, 9, 11)
n <- length(original_sample)

# With replacement: many possible samples
cat("With replacement (5 examples):\n")
for(i in 1:5) {
  boot_samp <- sample(original_sample, n, replace = TRUE)
  cat("  Sample", i, ":", boot_samp, "-> mean =", mean(boot_samp), "\n")
}

cat("\nWithout replacement:\n")
cat("  Always get:", sample(original_sample, n, replace = FALSE), "\n")
```

## Number of Possible Bootstrap Samples

::: callout-note
## Counting Bootstrap Samples
For a sample of size $n$, the number of possible bootstrap samples (sampling $n$ with replacement) is:
$$\binom{2n-1}{n-1} = \binom{2n-1}{n}$$

This is the number of ways to choose $n$ items from $n$ categories with repetition allowed.
:::

```{r}
#| echo: true
# Number of bootstrap samples for various n
tibble(n = c(5, 10, 15, 20, 30)) |>
  mutate(
    num_bootstrap_samples = choose(2*n - 1, n - 1),
    formatted = format(num_bootstrap_samples, big.mark = ",", scientific = FALSE)
  )
```

Even for $n = 15$, there are over **77 million** possible bootstrap samples!

## The Variance Connection

::: callout-important
## Variance of Bootstrap Mean
For the sample mean, the **theoretical** (exhaustive) bootstrap standard error is:
$$SE_{boot}(\bar{X}^*) = \frac{\hat{\sigma}}{\sqrt{n}} = \sqrt{\frac{1}{n^2}\sum_{i=1}^{n}(x_i - \bar{x})^2}$$

where $\hat{\sigma}^2 = \frac{1}{n}\sum(x_i - \bar{x})^2$ is the population variance under $\hat{F}_n$.
:::

. . .

**Note:** This differs slightly from $s/\sqrt{n}$ by a factor of $\sqrt{(n-1)/n}$:
$$SE_{boot} = \sqrt{\frac{n-1}{n}} \cdot \frac{s}{\sqrt{n}}$$

This is why bootstrap SEs are slightly **narrower** than classical SEs.

## Subsampling: An Alternative Approach

::: callout-note
## Subsampling (Without Replacement)
An alternative to bootstrap is **subsampling**: draw samples of size $m < n$ **without replacement**.

**Properties:**

- Requires choosing subsample size $m$
- Typically $m/n \to 0$ as $n \to \infty$
- Can work when bootstrap fails (heavy tails, dependent data)
- Generally less efficient than bootstrap
:::

. . .

**When to use subsampling:**

- Time series or dependent data
- Estimating extreme quantiles
- When bootstrap consistency fails

# The Bootstrap Method {background-color="#9b59b6"}

## Basic Bootstrap Algorithm (Devore 8.5)

::: callout-important
## Basic Bootstrap Method
Given a sample $x_1, x_2, \ldots, x_n$ and a statistic $\hat{\theta}$:

1. **Resample:** Draw a sample of size $n$ **with replacement** from $x_1, \ldots, x_n$
2. **Compute:** Calculate $\hat{\theta}^*$ from this bootstrap sample
3. **Repeat:** Repeat steps 1-2 many times (typically $B = 1000$ to $10000$)
4. **Analyze:** Use the distribution of $\hat{\theta}^*_1, \hat{\theta}^*_2, \ldots, \hat{\theta}^*_B$ for inference
:::

. . .

**Why sample with replacement?**

Without replacement, we'd get the same sample every time!

## Simple Example: Birth Weights (Chihara 5.1)

Consider just 3 birth weights: 3969, 3204, 2892 grams

```{r}
#| echo: true
small_sample <- c(3969, 3204, 2892)
cat("Original sample:", small_sample, "\n")
cat("Sample mean:", mean(small_sample))
```

. . .

**All possible bootstrap samples** (sampling with replacement):

```{r}
#| code-fold: true
# All 3^3 = 27 bootstrap samples
expand_grid(x1 = small_sample, x2 = small_sample, x3 = small_sample) |>
  mutate(
    sample_id = row_number(),
    mean = (x1 + x2 + x3) / 3
  ) |>
  head(10)
```

## Complete Bootstrap Distribution (Small Example)

```{r}
#| code-fold: true
#| fig-height: 5.5
all_boot_samples <- expand_grid(x1 = small_sample, x2 = small_sample, x3 = small_sample) |>
  mutate(boot_mean = (x1 + x2 + x3) / 3)

all_boot_samples |>
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = mean(small_sample), color = "red", linewidth = 1.2, linetype = "dashed") +
  scale_x_continuous(breaks = sort(unique(all_boot_samples$boot_mean))) +
  labs(x = expression(bar(x)^"*"), y = "Frequency",
       title = "Bootstrap Distribution of Mean (All 27 Samples)",
       subtitle = paste0("Original sample mean = ", round(mean(small_sample), 1))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Bootstrap standard error: `r round(sd(all_boot_samples$boot_mean), 1)` grams

## Bootstrap in R: Step by Step

```{r}
#| echo: true
# Original response time data
response_times <- c(8, 12, 15, 7, 22, 45, 11, 9, 14, 18, 10, 13)
n <- length(response_times)
B <- 10000  # Number of bootstrap samples

# Storage for bootstrap means
boot_means <- numeric(B)

# Bootstrap loop
for (i in 1:B) {
  # Step 1: Resample with replacement
  boot_sample <- sample(response_times, size = n, replace = TRUE)
  
  # Step 2: Calculate statistic
  boot_means[i] <- mean(boot_sample)
}

# Results
cat("Original sample mean:", round(mean(response_times), 2), "\n")
cat("Bootstrap mean:", round(mean(boot_means), 2), "\n")
cat("Bootstrap standard error:", round(sd(boot_means), 2))
```

## Bootstrap Distribution: Drug Response Times

```{r}
#| code-fold: true
#| fig-height: 6
tibble(x = boot_means) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", color = "white", alpha = 0.7) +
  geom_vline(xintercept = mean(response_times), color = "blue", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean(boot_means), color = "red", linewidth = 1, linetype = "dotted") +
  labs(x = expression(bar(x)^"*"), y = "Count",
       title = "Bootstrap Distribution of Mean Response Time",
       subtitle = paste0("B = ", B, " resamples; SE = ", round(sd(boot_means), 2), " days")) +
  annotate("text", x = 12, y = 800, 
           label = paste0("Original mean = ", round(mean(response_times), 1)), 
           color = "blue", size = 5)
```

## Using Tidyverse for Bootstrap

```{r}
#| echo: true
# More elegant approach using tidyverse
set.seed(551)
boot_results <- tibble(
  replicate = 1:B
) |>
  mutate(
    boot_sample = map(replicate, ~sample(response_times, n, replace = TRUE)),
    boot_mean = map_dbl(boot_sample, mean),
    boot_median = map_dbl(boot_sample, median),
    boot_sd = map_dbl(boot_sample, sd)
  )

boot_results |>
  select(-boot_sample) |>
  summarize(
    across(starts_with("boot_"), list(mean = mean, se = sd))
  ) |>
  pivot_longer(everything(), names_to = "statistic", values_to = "value")
```

# Asymptotic Properties of Bootstrap {background-color="#e67e22"}

## Bootstrap Consistency

::: callout-important
## Bootstrap Consistency Theorem
Under regularity conditions, the bootstrap distribution consistently estimates the sampling distribution. Specifically:

Let $\hat{\theta}_n$ be an estimator of $\theta$ based on a sample of size $n$. If:
$$\sqrt{n}(\hat{\theta}_n - \theta) \xrightarrow{d} N(0, \sigma^2)$$

Then under the bootstrap:
$$\sqrt{n}(\hat{\theta}_n^* - \hat{\theta}_n) \xrightarrow{d} N(0, \sigma^2)$$

where convergence is in probability.
:::

. . .

**In plain English:** The bootstrap distribution mimics the true sampling distribution as $n \to \infty$.

## What Bootstrap Estimates Well

::: callout-tip
## Bootstrap Accurately Estimates (Chihara 5.2.2)

1. **Spread:** The spread (standard error) of the sampling distribution
2. **Shape:** The shape (including skewness) of the sampling distribution
3. **Bias:** The bias of the estimator
:::

. . .

::: callout-warning
## Bootstrap Does NOT Estimate

**Center:** The bootstrap distribution is centered at $\hat{\theta}$ (the sample statistic), NOT at $\theta$ (the true parameter).

This means bootstrap **cannot improve point estimates**!
:::

## Visualizing What Bootstrap Captures

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 12
set.seed(551)
# True population: Gamma(2, 0.1)
true_mean <- 20
true_se <- 20 / sqrt(50)

# One sample and its bootstrap
samp <- rgamma(50, shape = 2, rate = 0.1)
sample_mean <- mean(samp)

# Bootstrap distribution
boot_dist <- replicate(10000, mean(sample(samp, 50, replace = TRUE)))

# Sampling distribution (simulated)
sampling_dist <- replicate(10000, mean(rgamma(50, shape = 2, rate = 0.1)))

p1 <- tibble(x = sampling_dist) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "darkgreen", alpha = 0.6) +
  geom_vline(xintercept = true_mean, color = "darkgreen", linewidth = 1.5) +
  labs(title = "True Sampling Distribution",
       subtitle = paste0("Centered at μ = ", true_mean, ", SE = ", round(sd(sampling_dist), 2)),
       x = expression(bar(X)), y = "Count")

p2 <- tibble(x = boot_dist) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.6) +
  geom_vline(xintercept = sample_mean, color = "coral", linewidth = 1.5) +
  labs(title = "Bootstrap Distribution",
       subtitle = paste0("Centered at x̄ = ", round(sample_mean, 2), ", SE = ", round(sd(boot_dist), 2)),
       x = expression(bar(X)^"*"), y = "Count")

p1 + p2
```

**Key observation:** Same spread and shape, but different centers!

## Comparison Table (Chihara Table 5.2)

```{r}
#| echo: false
tibble(
  Quantity = c("Population", "Sampling distribution of X̄", "Sample", "Bootstrap distribution"),
  Mean = c("μ", "μ", "x̄", "≈ x̄"),
  `Standard Deviation` = c("σ", "σ/√n", "s", "≈ s·√((n-1)/n)/√n")
) |>
  knitr::kable(align = "lcc")
```

. . .

**Key insights:**

- Bootstrap mean ≈ sample mean (not population mean)
- Bootstrap SE ≈ sampling distribution SE
- Bootstrap shape ≈ sampling distribution shape

## Rate of Convergence

::: callout-note
## Edgeworth Expansion Refinement
For the sample mean, classical confidence intervals have coverage error of order $O(n^{-1/2})$.

**Bootstrap percentile intervals:** Also $O(n^{-1/2})$ — same as classical

**Bootstrap t intervals:** $O(n^{-1})$ — **one order more accurate!**

This is because the bootstrap t automatically corrects for skewness.
:::

. . .

**Practical implication:** Bootstrap t intervals are preferred for skewed distributions, especially with moderate sample sizes.

# When Bootstrap Fails {background-color="#c0392b"}

## Bootstrap Failure Modes

The bootstrap is **not** a universal solution. It can fail when:

1. **Very small samples** ($n < 10$)
2. **Sample doesn't represent population**
3. **Estimating extreme quantiles**
4. **Discrete statistics** (e.g., median with few unique values)
5. **Heavy-tailed distributions** (infinite variance)
6. **Dependent data** (time series, spatial data)
7. **Lattice distributions** with small support

## Failure Mode 1: Very Small Samples

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
set.seed(551)
# Small sample from normal
small_n <- 6
small_samp <- rnorm(small_n, mean = 10, sd = 3)

# Multiple bootstrap distributions from same sample
boot_dists <- map(1:4, function(seed) {
  set.seed(seed)
  replicate(1000, mean(sample(small_samp, small_n, replace = TRUE)))
})

# True sampling distribution
true_sampling <- replicate(10000, mean(rnorm(small_n, 10, 3)))

plots <- map2(boot_dists, 1:4, function(bd, i) {
  tibble(x = bd) |>
    ggplot(aes(x = x)) +
    geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
    geom_vline(xintercept = mean(small_samp), color = "blue", linewidth = 1) +
    labs(title = paste("Bootstrap", i), 
         subtitle = paste0("SE = ", round(sd(bd), 2)),
         x = expression(bar(x)^"*"), y = "") +
    coord_cartesian(xlim = c(6, 14))
})

wrap_plots(plots, ncol = 4)
```

**Problem:** Bootstrap distributions vary substantially due to small sample size.

## Failure Mode 2: The Median

```{r}
#| code-fold: true
#| fig-height: 5.5
set.seed(551)
# Sample for median bootstrap
med_sample <- rnorm(15, mean = 0, sd = 1)

# Bootstrap the median
boot_medians <- replicate(10000, median(sample(med_sample, 15, replace = TRUE)))

tibble(x = boot_medians) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 30, fill = "purple", alpha = 0.7) +
  geom_vline(xintercept = median(med_sample), color = "blue", linewidth = 1.2) +
  labs(x = expression(tilde(x)^"*"), y = "Count",
       title = "Bootstrap Distribution of Median (n = 15)",
       subtitle = "Notice the discrete, irregular shape")
```

**Problem:** Bootstrap medians only take values from the original sample!

## Why Median Bootstrap Fails

::: callout-warning
## The Median Problem (Chihara 5.8.3)
For odd sample size $n$, the bootstrap median can only take values that are **observations in the original sample**.

**Consequences:**

- Bootstrap distribution is **discrete**, not continuous
- Shape depends heavily on **gaps between observations**
- May poorly approximate the continuous sampling distribution
:::

. . .

**Remedies:**

- **Smoothed bootstrap:** Add small noise to observations
- **Larger samples:** Problem diminishes as $n$ increases
- Use percentile intervals rather than bootstrap t

## Failure Mode 3: Heavy-Tailed Distributions

::: callout-important
## Cauchy Distribution Example
The Cauchy distribution has **no mean** (undefined). For such distributions:

- Sample mean is **not consistent**
- Bootstrap cannot "fix" an inconsistent estimator
- Bootstrap distribution doesn't converge properly
:::

```{r}
#| echo: true
# Cauchy has no mean - bootstrap can't help
set.seed(551)
cauchy_sample <- rcauchy(50)
boot_cauchy <- replicate(1000, mean(sample(cauchy_sample, 50, replace = TRUE)))

cat("Sample mean:", round(mean(cauchy_sample), 2), "\n")
cat("Bootstrap SE:", round(sd(boot_cauchy), 2), " (meaningless!)")
```

## Visualizing Heavy-Tail Failure

```{r}
#| code-fold: true
#| fig-height: 5
# Compare bootstrap for normal vs Cauchy
set.seed(551)
normal_boot <- replicate(5000, mean(sample(rnorm(30), 30, replace = TRUE)))
cauchy_boot <- replicate(5000, mean(sample(rcauchy(30), 30, replace = TRUE)))

p1 <- tibble(x = normal_boot) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Bootstrap Means: Normal(0,1)", x = expression(bar(x)^"*")) +
  coord_cartesian(xlim = c(-1, 1))

p2 <- tibble(x = cauchy_boot) |>
  filter(abs(x) < 50) |>  # Remove extreme values for visualization
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  labs(title = "Bootstrap Means: Cauchy (truncated)", x = expression(bar(x)^"*"))

p1 + p2
```

For Cauchy, the bootstrap distribution is erratic and doesn't converge!

## Failure Mode 4: Extreme Quantiles

::: callout-warning
## Estimating Extreme Quantiles
Bootstrap has difficulty with extreme quantiles (e.g., 99th percentile) because:

1. Few observations near the tail
2. Bootstrap samples may not include extreme values
3. Standard bootstrap underestimates tail variability
:::

```{r}
#| echo: true
# Bootstrap for 95th percentile
set.seed(551)
sample_data <- rexp(50, rate = 0.1)
boot_q95 <- replicate(5000, quantile(sample(sample_data, 50, replace = TRUE), 0.95))

cat("Sample 95th percentile:", round(quantile(sample_data, 0.95), 2), "\n")
cat("Bootstrap SE:", round(sd(boot_q95), 2))
```

. . .

For extreme quantiles, consider **parametric bootstrap** or **extreme value theory**.

## Summary: When Bootstrap Works and Fails

```{r}
#| echo: false
tibble(
  `Works Well` = c("Sample mean", "Trimmed mean", "Correlation", "Regression coefficients", 
                   "Smooth functions of means"),
  `Use Caution` = c("Median (small n)", "Extreme quantiles", "Small samples (n < 15)", 
                    "Heavily skewed data", "Discrete distributions"),
  `May Fail` = c("Heavy tails (no variance)", "Dependent data", "Non-representative sample",
                 "Maximum/minimum", "Lattice distributions")
) |>
  knitr::kable()
```

# Bootstrap Standard Errors {background-color="#e74c3c"}

## Bootstrap Standard Error

::: callout-important
## Definition
The **bootstrap standard error** of a statistic $\hat{\theta}$ is the standard deviation of its bootstrap distribution:

$$SE_{boot}(\hat{\theta}) = \sqrt{\frac{1}{B-1}\sum_{b=1}^{B}(\hat{\theta}^*_b - \bar{\hat{\theta}}^*)^2}$$

where $\bar{\hat{\theta}}^* = \frac{1}{B}\sum_{b=1}^{B}\hat{\theta}^*_b$
:::

. . .

This gives us an **empirical estimate** of the standard error without assuming any specific distribution!

## Medical Example: Blood Pressure Reduction

A pilot study measures systolic blood pressure reduction (mmHg) after 8 weeks on a new antihypertensive:

```{r}
#| echo: true
bp_reduction <- c(12, 8, 15, 22, 5, 18, 11, 14, 9, 25, 7, 16, 13, 10, 19)
n <- length(bp_reduction)

cat("Sample size:", n, "\n")
cat("Mean reduction:", round(mean(bp_reduction), 2), "mmHg\n")
cat("Sample SD:", round(sd(bp_reduction), 2), "mmHg\n")
cat("Classical SE:", round(sd(bp_reduction)/sqrt(n), 2), "mmHg")
```

## Bootstrap SE for Blood Pressure

```{r}
#| echo: true
set.seed(551)
B <- 10000
boot_means_bp <- replicate(B, {
  boot_sample <- sample(bp_reduction, n, replace = TRUE)
  mean(boot_sample)
})

# Compare standard errors
classical_se <- sd(bp_reduction) / sqrt(n)
boot_se <- sd(boot_means_bp)

cat("Classical SE (s/√n):", round(classical_se, 3), "\n")
cat("Bootstrap SE:", round(boot_se, 3), "\n")
cat("Ratio:", round(boot_se / classical_se, 3))
```

. . .

The bootstrap SE is very close to the classical SE when assumptions are met!

## When Bootstrap SE Shines: Trimmed Mean

For **trimmed means** (robust to outliers), we don't have simple SE formulas.

```{r}
#| echo: true
# Data with potential outlier
response_times <- c(8, 12, 15, 7, 22, 45, 11, 9, 14, 18, 10, 13)

# Bootstrap for 10% trimmed mean
set.seed(551)
boot_trimmed <- replicate(10000, {
  x <- sample(response_times, length(response_times), replace = TRUE)
  mean(x, trim = 0.10)  # Trim 10% from each end
})

cat("Sample mean:", round(mean(response_times), 2), "\n")
cat("10% Trimmed mean:", round(mean(response_times, trim = 0.10), 2), "\n")
cat("Bootstrap SE of trimmed mean:", round(sd(boot_trimmed), 2))
```

## Example: Database Administrator Salaries (Devore 8.18)

```{r}
#| echo: true
#| code-fold: true
# Sample salaries (in thousands) - simulated based on Devore example
set.seed(123)
salaries <- round(c(rlnorm(100, log(80), 0.3), rlnorm(15, log(150), 0.4))) * 1000
n <- length(salaries)

tibble(x = salaries) |>
  ggplot(aes(x = x/1000)) +
  geom_histogram(bins = 25, fill = "steelblue", alpha = 0.7) +
  labs(x = "Salary (thousands $)", y = "Count",
       title = "Distribution of Database Administrator Salaries",
       subtitle = paste0("n = ", n, ", with right skew"))
```

## Bootstrap for Trimmed Mean Salary

```{r}
#| echo: true
# 10% trimmed mean
xbar_tr <- mean(salaries, trim = 0.10)

# Bootstrap
set.seed(551)
B <- 10000
boot_trimmed_sal <- replicate(B, {
  x <- sample(salaries, n, replace = TRUE)
  mean(x, trim = 0.10)
})

cat("10% Trimmed mean:", round(xbar_tr, 0), "\n")
cat("Bootstrap SE:", round(sd(boot_trimmed_sal), 0), "\n")
cat("Bootstrap mean of trimmed means:", round(mean(boot_trimmed_sal), 0))
```

## Visualizing the Bootstrap Distribution

```{r}
#| code-fold: true
#| fig-height: 6
tibble(x = boot_trimmed_sal) |>
  ggplot(aes(x = x/1000)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  geom_vline(xintercept = xbar_tr/1000, color = "blue", linewidth = 1.2) +
  labs(x = expression(bar(x)[tr]^"*" ~ "(thousands $)"), y = "Count",
       title = "Bootstrap Distribution of 10% Trimmed Mean Salary",
       subtitle = paste0("SE = $", round(sd(boot_trimmed_sal)/1000, 1), "k"))
```

The bootstrap distribution appears approximately normal — good for t-based intervals!

## Bias in Bootstrap Distributions (Chihara 5.6)

::: callout-note
## Bootstrap Bias Estimate
The **bootstrap estimate of bias** is:
$$\widehat{\text{Bias}} = \bar{\hat{\theta}}^* - \hat{\theta}$$

where $\bar{\hat{\theta}}^*$ is the mean of the bootstrap distribution and $\hat{\theta}$ is the original sample statistic.
:::

```{r}
# Check bias for trimmed mean
boot_bias <- mean(boot_trimmed_sal) - xbar_tr
cat("Bootstrap bias estimate:", round(boot_bias, 0), "\n")
cat("Bias as % of estimate:", round(100 * boot_bias / xbar_tr, 2), "%")
```

. . .

Small bias suggests the bootstrap distribution is well-centered!

# Bootstrap Confidence Intervals {background-color="#27ae60"}

## Types of Bootstrap Confidence Intervals

Several methods exist for constructing bootstrap CIs:

1. **Bootstrap Percentile Interval** — Simple, intuitive
2. **Bootstrap t Interval** — Better theoretical properties
3. **BCa Interval** — Bias-corrected and accelerated (advanced)

. . .

::: callout-tip
## Practical Guidance
- **Percentile interval:** Good when bootstrap distribution is symmetric, low bias
- **Bootstrap t interval:** Better for small samples, handles skewness
- Both require the bootstrap distribution to be a reasonable approximation
:::

## Bootstrap Percentile Interval (Devore 8.5)

::: callout-important
## Definition
A **95% bootstrap percentile interval** for parameter $\theta$ has endpoints equal to the **2.5th and 97.5th percentiles** of the bootstrap distribution.

More generally, a $100(1-\alpha)\%$ bootstrap percentile interval uses the $\alpha/2$ and $1-\alpha/2$ quantiles.
:::

. . .

**Intuition:** If 95% of bootstrap statistics fall between $L$ and $U$, then $(L, U)$ is a plausible range for the true parameter.

## Computing Percentile Intervals in R

```{r}
#| echo: true
# Blood pressure reduction data
set.seed(551)
bp_reduction <- c(12, 8, 15, 22, 5, 18, 11, 14, 9, 25, 7, 16, 13, 10, 19)

# Bootstrap
boot_means_bp <- replicate(10000, mean(sample(bp_reduction, length(bp_reduction), replace = TRUE)))

# 95% Percentile interval
percentile_ci <- quantile(boot_means_bp, probs = c(0.025, 0.975))

cat("Sample mean:", round(mean(bp_reduction), 2), "mmHg\n")
cat("95% Bootstrap Percentile CI:", round(percentile_ci, 2), "mmHg\n")

# Compare to t-interval
t_ci <- t.test(bp_reduction)$conf.int
cat("95% t-interval:", round(t_ci, 2), "mmHg")
```

## Visualizing the Percentile Interval

```{r}
#| code-fold: true
#| fig-height: 6
tibble(x = boot_means_bp) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = percentile_ci, color = "red", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean(bp_reduction), color = "black", linewidth = 1) +
  annotate("rect", xmin = percentile_ci[1], xmax = percentile_ci[2], 
           ymin = 0, ymax = Inf, alpha = 0.2, fill = "red") +
  labs(x = expression(bar(x)^"*"), y = "Count",
       title = "Bootstrap Distribution with 95% Percentile Interval",
       subtitle = paste0("CI: (", round(percentile_ci[1], 1), ", ", round(percentile_ci[2], 1), ") mmHg"))
```

## Medical Example: Median Survival Time

For **median** survival, we don't have simple formulas for CIs. Bootstrap is ideal!

```{r}
#| echo: true
# Survival times (months) for 20 cancer patients
survival_times <- c(4, 8, 12, 15, 18, 6, 9, 24, 36, 11,
                    7, 14, 21, 10, 13, 5, 17, 22, 8, 19)

# Bootstrap for median
set.seed(551)
boot_medians <- replicate(10000, {
  median(sample(survival_times, length(survival_times), replace = TRUE))
})

# Results
cat("Sample median:", median(survival_times), "months\n")
cat("Bootstrap SE:", round(sd(boot_medians), 2), "months\n")
cat("95% Percentile CI:", quantile(boot_medians, c(0.025, 0.975)), "months")
```

## Bootstrap Distribution of Median

```{r}
#| code-fold: true
#| fig-height: 5.5
med_ci <- quantile(boot_medians, c(0.025, 0.975))

tibble(x = boot_medians) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7) +
  geom_vline(xintercept = med_ci, color = "red", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = median(survival_times), color = "black", linewidth = 1) +
  labs(x = expression(tilde(x)^"*"), y = "Count",
       title = "Bootstrap Distribution of Median Survival Time",
       subtitle = "Notice: Discrete nature of the median")
```

The bootstrap median distribution takes on **few unique values** — this is typical!

## Bootstrap t Confidence Interval (Devore 8.5)

::: callout-important
## Bootstrap t Interval
If the bootstrap distribution is approximately **normal** with small **bias**, use:
$$\hat{\theta} \pm t_{\alpha/2, n-1} \cdot SE_{boot}$$

where $SE_{boot}$ is the bootstrap standard error.
:::

. . .

```{r}
#| echo: true
# Bootstrap t interval for trimmed mean salary
xbar_tr <- mean(salaries, trim = 0.10)
se_boot <- sd(boot_trimmed_sal)
t_crit <- qt(0.975, df = length(salaries) - 1)

ci_lower <- xbar_tr - t_crit * se_boot
ci_upper <- xbar_tr + t_crit * se_boot

cat("Bootstrap t 95% CI for trimmed mean salary:\n")
cat("($", round(ci_lower, 0), ", $", round(ci_upper, 0), ")", sep = "")
```

## The Studentized Bootstrap t Interval (Chihara 7.5)

::: callout-important
## Bootstrap t Interval (More Sophisticated)
For each bootstrap sample, compute:
$$T^* = \frac{\bar{X}^* - \bar{x}}{S^*/\sqrt{n}}$$

Use quantiles $Q^*_1$ and $Q^*_2$ of the bootstrap t distribution:
$$\left(\bar{x} - Q^*_2 \cdot \frac{s}{\sqrt{n}}, \; \bar{x} - Q^*_1 \cdot \frac{s}{\sqrt{n}}\right)$$
:::

. . .

This method adapts to **skewness** in the data!

## Bootstrap t Interval: R Implementation

```{r}
#| echo: true
bootstrap_t_interval <- function(data, B = 10000, conf = 0.95) {
  n <- length(data)
  xbar <- mean(data)
  se <- sd(data) / sqrt(n)
  
  # Compute bootstrap t statistics
  t_star <- replicate(B, {
    boot_samp <- sample(data, n, replace = TRUE)
    (mean(boot_samp) - xbar) / (sd(boot_samp) / sqrt(n))
  })
  
  # Quantiles
  alpha <- 1 - conf
  q_lower <- quantile(t_star, alpha/2)
  q_upper <- quantile(t_star, 1 - alpha/2)
  
  # CI (note the reversal!)
  c(xbar - q_upper * se, xbar - q_lower * se)
}

# Apply to blood pressure data
boot_t_ci <- bootstrap_t_interval(bp_reduction)
cat("Bootstrap t CI:", round(boot_t_ci, 2), "mmHg")
```

## Comparing All Methods

```{r}
#| echo: true
#| code-fold: true
# Compare all CI methods for BP reduction data
set.seed(551)

# Classical t
t_result <- t.test(bp_reduction)$conf.int

# Bootstrap percentile
boot_stats <- replicate(10000, mean(sample(bp_reduction, length(bp_reduction), replace = TRUE)))
pct_ci <- quantile(boot_stats, c(0.025, 0.975))

# Bootstrap t
boot_t_ci <- bootstrap_t_interval(bp_reduction)

tibble(
  Method = c("Classical t", "Bootstrap Percentile", "Bootstrap t"),
  Lower = c(t_result[1], pct_ci[1], boot_t_ci[1]),
  Upper = c(t_result[2], pct_ci[2], boot_t_ci[2])
) |>
  mutate(
    Width = Upper - Lower,
    across(Lower:Width, ~round(.x, 2))
  )
```

## When Percentile May Be Suboptimal: Skewed Distributions

```{r}
#| code-fold: true
#| fig-height: 5.5
# Highly skewed data
set.seed(551)
skewed_data <- rexp(15, rate = 0.5)  # Exponential with mean 2

boot_skewed <- replicate(10000, mean(sample(skewed_data, 15, replace = TRUE)))
pct_ci <- quantile(boot_skewed, c(0.025, 0.975))

tibble(x = boot_skewed) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  geom_vline(xintercept = pct_ci, color = "blue", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean(skewed_data), color = "black", linewidth = 1) +
  labs(x = expression(bar(x)^"*"), y = "Count",
       title = "Bootstrap Distribution from Skewed Population",
       subtitle = "Percentile interval may have incorrect coverage")
```

# Practical Considerations {background-color="#16a085"}

## How Many Bootstrap Samples? (Chihara 5.9)

::: callout-note
## Guidelines
- **Standard errors:** $B = 1,000$ is often sufficient
- **Confidence intervals:** $B = 5,000$ to $10,000$ recommended
- **Percentile tails:** More samples needed for accurate tail quantiles
:::

```{r}
#| echo: true
# Demonstrating convergence of bootstrap SE
set.seed(551)
B_values <- c(100, 500, 1000, 5000, 10000)

convergence <- map_dfr(B_values, function(B) {
  boot_means <- replicate(B, mean(sample(bp_reduction, length(bp_reduction), replace = TRUE)))
  tibble(B = B, SE = sd(boot_means))
})

convergence
```

## Visualizing Bootstrap Convergence

```{r}
#| code-fold: true
#| fig-height: 5.5
set.seed(551)
# Run multiple bootstrap analyses with different B
running_se <- tibble(
  B = 1:10000,
  cumulative_se = accumulate(
    replicate(10000, mean(sample(bp_reduction, length(bp_reduction), replace = TRUE))),
    ~c(.x, .y),
    .init = numeric(0)
  )[-1] |> map_dbl(sd)
)

running_se |>
  ggplot(aes(x = B, y = cumulative_se)) +
  geom_line(color = "steelblue") +
  geom_hline(yintercept = tail(running_se$cumulative_se, 1), linetype = "dashed", color = "red") +
  labs(x = "Number of Bootstrap Samples (B)", y = "Bootstrap SE",
       title = "Convergence of Bootstrap Standard Error") +
  scale_x_log10()
```

## When Does Bootstrap Work Well?

::: callout-tip
## Bootstrap Works Well When:
1. The sample is **representative** of the population
2. The sample size is **not too small** ($n \geq 15$ typically)
3. The statistic is **smooth** (means, variances, regression coefficients)
4. The parameter is well-defined and has finite variance
:::

. . .

::: callout-warning
## Bootstrap May Struggle When:
1. Sample is very small ($n < 10$)
2. Data has extreme outliers not representative of population
3. Statistic is very discrete (e.g., median with few unique values)
4. Estimating extreme quantiles
5. Population has infinite variance (Cauchy, Pareto with $\alpha \leq 2$)
:::

## Classical vs. Bootstrap: Guidelines (Chihara 7.5.1)

```{r}
#| echo: false
tibble(
  Situation = c("Small n (≤10), normal data", 
                "Small n (≤10), skewed data",
                "Large n, normal data",
                "Large n, skewed data",
                "Complex statistic (trimmed mean, ratio)"),
  Recommendation = c("Classical t preferred",
                     "Bootstrap t (with caution)",
                     "Either works well",
                     "Bootstrap t preferred",
                     "Bootstrap (no classical alternative)")
) |>
  knitr::kable()
```

. . .

**Key insight:** The bootstrap **learns skewness from the sample**, but small samples may not accurately represent population skewness.

## Medical Application: Relative Risk (Chihara 5.5)

In a cardiovascular study:

- High BP group: 55 events in 3,338 men
- Low BP group: 21 events in 2,676 men

```{r}
#| echo: true
# Relative risk
p1 <- 55/3338  # High BP group
p2 <- 21/2676  # Low BP group
rr_hat <- p1/p2

cat("Risk (high BP):", round(p1 * 100, 2), "%\n")
cat("Risk (low BP):", round(p2 * 100, 2), "%\n")
cat("Relative Risk:", round(rr_hat, 2))
```

## Bootstrap for Relative Risk

```{r}
#| echo: true
set.seed(551)
n1 <- 3338; x1 <- 55
n2 <- 2676; x2 <- 21

# Bootstrap
boot_rr <- replicate(10000, {
  # Resample from each group (binary outcomes)
  boot_p1 <- sum(sample(c(rep(1, x1), rep(0, n1 - x1)), n1, replace = TRUE)) / n1
  boot_p2 <- sum(sample(c(rep(1, x2), rep(0, n2 - x2)), n2, replace = TRUE)) / n2
  
  # Handle division by zero
  if (boot_p2 == 0) return(NA)
  boot_p1 / boot_p2
})

boot_rr <- boot_rr[!is.na(boot_rr)]

cat("Bootstrap SE:", round(sd(boot_rr), 3), "\n")
cat("95% Percentile CI:", round(quantile(boot_rr, c(0.025, 0.975)), 2))
```

## Relative Risk: Bootstrap Distribution

```{r}
#| code-fold: true
#| fig-height: 5.5
rr_ci <- quantile(boot_rr, c(0.025, 0.975))

tibble(x = boot_rr) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  geom_vline(xintercept = rr_hat, color = "blue", linewidth = 1.2) +
  geom_vline(xintercept = rr_ci, color = "red", linewidth = 1, linetype = "dashed") +
  labs(x = "Relative Risk", y = "Count",
       title = "Bootstrap Distribution of Relative Risk",
       subtitle = paste0("RR = ", round(rr_hat, 2), ", 95% CI: (", 
                        round(rr_ci[1], 2), ", ", round(rr_ci[2], 2), ")"))
```

High BP patients have 1.3 to 3.7 times the cardiovascular risk (95% CI).

# Summary and Looking Ahead {background-color="#2c3e50"}

## Key Takeaways

::: callout-important
## The Bootstrap Principle
1. **The sample represents the population** — resample from the sample
2. **Sampling with replacement** mimics drawing new samples
3. **Bootstrap distribution approximates sampling distribution**
4. Use bootstrap to estimate **standard errors** and construct **confidence intervals**
:::

. . .

**Main advantages:**

- No distributional assumptions required
- Works for **any statistic** (medians, ratios, etc.)
- Computationally straightforward

## Why Resampling WITH Replacement?

::: callout-note
## Key Points
1. **Without replacement:** Would get the same sample every time (size $n$)
2. **With replacement:** Creates variability that mimics sampling from population
3. **Empirical distribution:** Assigns probability $1/n$ to each observation
4. **Number of bootstrap samples:** $\binom{2n-1}{n}$ possible — astronomically large!
:::

## Asymptotic Properties Summary

::: callout-tip
## Bootstrap Consistency
Under regularity conditions:

- Bootstrap distribution **converges** to sampling distribution
- Bootstrap SE **consistently estimates** true SE
- Bootstrap t intervals have coverage error $O(n^{-1})$ — better than percentile!
:::

. . .

**The bootstrap accurately estimates:**

- Spread (standard error)
- Shape (including skewness)
- Bias

**But NOT the center** (stays at $\hat{\theta}$, not $\theta$)

## When Bootstrap Fails: Summary

| Failure Mode | Reason | Alternative |
|--------------|--------|-------------|
| Very small $n$ | Sample doesn't represent population | Classical methods |
| Median/quantiles | Discrete bootstrap distribution | Smoothed bootstrap |
| Heavy tails | Infinite variance | Subsampling |
| Extreme quantiles | Few observations in tails | Parametric bootstrap |
| Dependent data | Independence assumption violated | Block bootstrap |

## Bootstrap CIs Summary

| Method | Formula | When to Use |
|--------|---------|-------------|
| **Percentile** | $(\hat{\theta}^*_{\alpha/2}, \hat{\theta}^*_{1-\alpha/2})$ | Symmetric bootstrap dist., low bias |
| **Bootstrap t** | $\hat{\theta} \pm t_{\alpha/2} \cdot SE_{boot}$ | Normal bootstrap dist. |
| **Bootstrap t (studentized)** | Uses bootstrap t-statistics | Skewed data, more accurate |

## What the Bootstrap Can and Cannot Do (Chihara 5.2.2)

::: columns
::: {.column width="50%"}
**Can estimate:**

- Standard error
- Bias
- Skewness
- Confidence intervals
:::

::: {.column width="50%"}
**Cannot improve:**

- Point estimates (still centered at $\hat{\theta}$, not $\theta$)
- A bad/unrepresentative sample
:::
:::

. . .

::: callout-warning
## Important Reminder
The bootstrap does **not create new information**. It uses the sample you have to understand the variability of your estimate.
:::

## Looking Ahead

**Next Topics:**

- **Two-sample bootstrap** — comparing groups
- **Bootstrap hypothesis testing** — computing p-values
- **Permutation tests** — alternative resampling approach

. . .

**Coming Soon:**

- Hypothesis testing framework (Lesson 10)
- Type I and Type II errors
- Power and sample size

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 8.5
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapters 5, 7
-   Efron, B. and Tibshirani, R. (1993). *An Introduction to the Bootstrap*. Chapman & Hall.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
quarto render 09_Bootstrap_Introduction.qmd
```
