---
title: "Lesson 21: Nonparametric Inference"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w10d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 21 - Nonparametric Inference
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
library(DiagrammeR)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 21: Nonparametric Inference {background-color="#2c3e50"}

## Review: Lessons 18–20

- **Lesson 18:** Parametric two-sample tests — Welch $t$, pooled $t$, $z$-test for proportions, $F$-test
- **Lesson 19:** Paired $t$-test — reduces to one-sample inference on $D_i = X_{1i} - X_{2i}$
- **Lesson 20:** Bootstrap CIs (resample separately) and permutation tests (relabel pooled data)

. . .

**Today's Goals:**

1. Understand what "nonparametric" means and when it is preferred over parametric tests
2. Derive and apply the **sign test** for paired data (DBC 14.1)
3. Derive and apply the **Wilcoxon signed-rank test** for paired data (DBC 14.2)
4. Derive and apply the **Wilcoxon rank-sum test** for two independent samples (DBC 14.3)
5. Understand **asymptotic relative efficiency (ARE)** and the power trade-offs it quantifies
6. Apply all three tests to medical examples and compare conclusions

## What Does "Nonparametric" Mean?

A **parametric** test specifies a distributional family (e.g., $D_i \sim N(\mu, \sigma^2)$) and makes inference about the **parameters** of that family. Validity depends on the assumed family being correct.

. . .

A **nonparametric** test makes inference about features of the distribution (median, location, equality of distributions) **without specifying a parametric family**. Validity holds over a broad class of distributions.

. . .

::: callout-note
## What "Nonparametric" Is NOT

Nonparametric $\neq$ assumption-free. These tests still require:

- **Independence** of observations (or independence of pairs)
- **Continuity** of the underlying distribution (for rank-based tests — ties deflate test power)
- **Symmetry** of $D_i$ about 0 (for the Wilcoxon signed-rank test)

The key relaxation is normality — not all assumptions.
:::

. . .

**Why does it matter in biomedical research?**

- **Biomarkers** (CRP, troponin, ALT) are right-skewed with occasional extreme values
- **Ordinal scales** (pain VAS, Likert, disability scores) — arithmetic means may be misleading
- **Small $n$** — CLT convergence too slow to trust normality-based tests

## Roadmap for Today

| Test | Data Structure | $H_0$ | Key Assumption | Reference |
|------|---------------|--------|---------------|-----------|
| Sign test | Paired / one-sample | Median $D = 0$ | None beyond independence | DBC 14.1 |
| Wilcoxon signed-rank | Paired / one-sample | Median $D = 0$ | Symmetric, continuous $D_i$ | DBC 14.2 |
| Wilcoxon rank-sum | Two independent samples | $F_1 = F_2$ | Continuous distributions | DBC 14.3 |

. . .

**Information used by each test:**

| Test | Uses direction of $D_i$? | Uses magnitude of $D_i$? |
|------|:---:|:---:|
| Sign test | ✓ | ✗ |
| Wilcoxon signed-rank | ✓ | ✓ (via ranks) |
| Paired $t$-test | ✓ | ✓ (exact values) |

Each test makes a different assumption about the data in exchange for the information it extracts.

# The Sign Test (DBC 14.1) {background-color="#3498db"}

## Sign Test: Setup and Motivation

**Setting:** Paired differences $D_i = X_{1i} - X_{2i}$, $i = 1, \ldots, n$.

**Goal:** Test whether the **population median** of the $D_i$ equals zero, without assuming any particular shape.

**Key insight:** If the true median of $D_i$ is zero, then $P(D_i > 0) = P(D_i < 0) = 1/2$ (for continuous distributions).

. . .

**Test statistic:** Count the positive differences.

$$S^+ = \#\{i : D_i > 0\}$$

Ties ($D_i = 0$): excluded. Let $n^* = \#\{i : D_i \neq 0\}$.

**Null distribution:** Under $H_0: P(D_i > 0) = 1/2$,

$$\boxed{S^+ \sim \text{Binomial}(n^*, 1/2)}$$

This follows because each non-zero $D_i$ is independently positive or negative with equal probability under $H_0$.

## Sign Test: Null Distribution Derivation

Why exactly $\text{Binomial}(n^*, 1/2)$?

Define $Z_i = \mathbf{1}(D_i > 0)$ for each $i$ with $D_i \neq 0$. Under $H_0$:

- $P(Z_i = 1) = P(D_i > 0) = 1/2$ (median is 0, distribution continuous)
- $P(Z_i = 0) = P(D_i < 0) = 1/2$
- $Z_1, \ldots, Z_{n^*}$ are **independent** (pairs are independent)

Therefore $Z_i \stackrel{iid}{\sim} \text{Bernoulli}(1/2)$ and

$$S^+ = \sum_{i=1}^{n^*} Z_i \sim \text{Binomial}(n^*, 1/2)$$

. . .

This is exact — no asymptotic approximation needed. The binomial distribution is completely tractable.

**Large-sample approximation:** For large $n^*$, by the CLT:

$$Z_S = \frac{S^+ - n^*/2}{\sqrt{n^*/4}} \;\dot{\sim}\; N(0,1)$$

Use the **continuity correction**: replace $S^+$ with $S^+ \pm 0.5$ to improve approximation to the discrete binomial.

## Sign Test: P-values

| Alternative | Exact p-value | Interpretation |
|-------------|---------|---|
| $H_a$: $\tilde{\mu}_D > 0$ | $P(S^+ \geq s^+)$ | More positives than expected |
| $H_a$: $\tilde{\mu}_D < 0$ | $P(S^+ \leq s^+)$ | Fewer positives than expected |
| $H_a$: $\tilde{\mu}_D \neq 0$ | $2\min\bigl(P(S^+ \geq s^+),\; P(S^+ \leq s^+)\bigr)$ | Either tail |

All probabilities from $\text{Bin}(n^*, 1/2)$ — exact, no normality required.

. . .

::: callout-tip
## Confidence Interval for the Median (Sign Test)

An exact CI for the **population median** is based on order statistics. The $100(1-\alpha)\%$ CI for the median is $(D_{(j)}, D_{(n^*-j+1)})$ where $j$ is chosen so $P(\text{Bin}(n^*, 1/2) \leq j - 1) \leq \alpha/2$.

In R: `wilcox.test(d, conf.int = TRUE)` gives the Wilcoxon CI (preferred when symmetry holds). For the sign-based CI: use `SignifReg::sign.ci()` or compute directly from the binomial.
:::

## Sign Test: Medical Example (Pain and Physical Therapy)

**Study:** 15 patients rate pain on a 0–10 VAS scale **before** and **after** a 4-week physical therapy program.

**Hypotheses:** $H_0$: median change $= 0$ vs. $H_a$: median change $< 0$ (pain **decreases**; $D_i = \text{after} - \text{before}$).

```{r}
#| echo: true
#| code-fold: false

pain_before <- c(8, 7, 9, 6, 8, 7, 9, 5, 8, 7, 6, 9, 8, 7, 6)
pain_after  <- c(5, 6, 6, 4, 5, 3, 7, 3, 5, 4, 4, 6, 5, 4, 3)
d_pain      <- pain_after - pain_before

cat("D = after - before:\n", d_pain, "\n\n")

n_star  <- sum(d_pain != 0)
s_plus  <- sum(d_pain > 0)

cat(sprintf("Non-zero differences: n* = %d\n", n_star))
cat(sprintf("Positive differences: S+ = %d\n", s_plus))
cat(sprintf("Negative differences: %d\n\n", sum(d_pain < 0)))

# One-sided p-value: H_a is median < 0 → few positives supports H_a
p_sign_one <- pbinom(s_plus, size = n_star, prob = 0.5)
p_sign_two <- 2 * min(pbinom(s_plus, n_star, 0.5),
                      pbinom(s_plus - 1, n_star, 0.5, lower.tail = FALSE))

cat(sprintf("One-sided p-value = %.5f\n", p_sign_one))
cat(sprintf("Two-sided p-value = %.5f\n", p_sign_two))
```

. . .

$S^+ = `r sum(d_pain > 0)`$ out of $n^* = `r sum(d_pain != 0)`$ non-zero differences.

Under $H_0$: $S^+ \sim \text{Bin}(14, 0.5)$, so $E[S^+] = 7$.

Observing only $S^+ = `r sum(d_pain > 0)`$ positive differences is very unlikely if the median were zero.

## Sign Test: Visualizing the Evidence

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5.5
#| fig-width: 14

# Panel 1: Dot chart of differences
p1 <- tibble(Patient = 1:15, D = d_pain,
             Direction = ifelse(d_pain > 0, "Positive", ifelse(d_pain < 0, "Negative", "Zero"))) |>
  ggplot(aes(x = Patient, y = D, color = Direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.8, color = "grey40") +
  geom_point(size = 5) +
  geom_segment(aes(xend = Patient, yend = 0), linewidth = 1) +
  scale_color_manual(values = c("Positive" = "#e74c3c",
                                 "Negative" = "#006a4e",
                                 "Zero"     = "grey60")) +
  labs(title = "Signed Differences (Pain After \u2212 Before)",
       x = "Patient", y = "D = After \u2212 Before", color = "Direction") +
  theme(legend.position = "bottom")

# Panel 2: Binomial null distribution
binom_df <- tibble(
  k    = 0:14,
  prob = dbinom(0:14, size = 14, prob = 0.5),
  tail = k <= s_plus
)
p2 <- binom_df |>
  ggplot(aes(x = k, y = prob, fill = tail)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = s_plus, color = "#e74c3c", linetype = "dashed", linewidth = 1.2) +
  scale_fill_manual(values = c("TRUE" = "#3498db", "FALSE" = "grey70"),
                    labels = c("TRUE" = "p-value region", "FALSE" = "Not in tail")) +
  annotate("text", x = s_plus + 0.5, y = 0.17,
           label = paste0("S+ = ", s_plus), color = "#e74c3c", size = 5, hjust = 0) +
  labs(title = expression("Null Distribution: " * S^"+" * " ~ Bin(14, 0.5)"),
       x = expression(S^"+"), y = "P(X = k)", fill = NULL) +
  theme(legend.position = "bottom")

p1 + p2
```

## Sign Test: Strengths and Limitations

**Strengths:**

- **No distributional assumptions** beyond independence — valid for any continuous or ordinal $D_i$
- Works when differences are measured on an **ordinal scale** (e.g., pain VAS, Likert), where arithmetic means are questionable
- The simplest nonparametric test to derive, compute, and explain

**Limitations:**

- Only uses the **sign** (direction) of each difference — discards all magnitude information
- **Lowest power** among the three paired tests when differences are continuous and symmetric
- Wide confidence intervals for the median, especially at small $n$

. . .

::: callout-note
## When to Prefer the Sign Test

- Data are **ordinal**: arithmetic differences aren't on an interval scale
- Differences are **highly asymmetric** with extreme outliers: sign is reliable, magnitude is not
- You need the most **assumption-free** result and are willing to sacrifice power
- Only the **direction** of each difference is observable (e.g., preferences without magnitudes)
:::

# Wilcoxon Signed-Rank Test (DBC 14.2) {background-color="#e74c3c"}

## Wilcoxon Signed-Rank: Motivation

The sign test throws away the **size** of each difference. Can we do better while still avoiding normality?

. . .

**Key idea:** Replace the raw differences with their **ranks** by absolute value.

- Ranks preserve the *relative ordering* of the magnitudes
- They are less sensitive to extreme outliers than raw values
- If the distribution is symmetric about 0, the ranks of positive and negative differences should look "balanced" under $H_0$

. . .

**Key assumption added over sign test:** The distribution of $D_i$ is **symmetric** about its median.

- Under $H_0$, symmetric about 0
- Under $H_a$, symmetric about $\theta \neq 0$ (a location shift)

This rules out heavily skewed $D_i$ — but covers most bell-shaped and even uniform-like distributions.

## Wilcoxon Signed-Rank: Algorithm

**Setup:** $D_1, \ldots, D_n \stackrel{iid}{\sim}$ a continuous, symmetric distribution. $H_0$: median $= 0$.

**Step-by-step:**

::: incremental
1. **Remove zeros**: drop any $D_i = 0$. Let $n^* = $ number remaining.
2. **Rank absolute values**: rank $|D_1|, \ldots, |D_{n^*}|$ from smallest (rank 1) to largest (rank $n^*$). Average ranks for ties.
3. **Compute $W^+$**: sum of ranks for observations with **positive** $D_i$:
$$W^+ = \sum_{\{i:\,D_i > 0\}} R_i$$
4. **Also define**: $W^- = \sum_{\{i:\,D_i < 0\}} R_i = \frac{n^*(n^*+1)}{2} - W^+$
:::

. . .

Under $H_0$ (symmetric about 0):

$$E[W^+] = \frac{n^*(n^*+1)}{4}, \qquad \text{Var}(W^+) = \frac{n^*(n^*+1)(2n^*+1)}{24}$$

Large $n^*$: $\quad Z = \dfrac{W^+ - E[W^+]}{\sqrt{\text{Var}(W^+)}} \;\dot{\sim}\; N(0,1)$

## Why Does $E[W^+] = \frac{n^*(n^*+1)}{4}$?

Under $H_0$ and symmetry, each rank $R_i$ is equally likely to be assigned to a positive or negative difference. Formally, let $\phi_i = \mathbf{1}(D_i > 0)$. Under $H_0$:

- $\phi_i \stackrel{iid}{\sim} \text{Bernoulli}(1/2)$, independent of the ranks $R_i$
- $W^+ = \sum_{i=1}^{n^*} \phi_i \cdot R_i$

Then:

$$E[W^+] = \sum_{i=1}^{n^*} E[\phi_i] \cdot R_i = \frac{1}{2} \sum_{i=1}^{n^*} R_i = \frac{1}{2} \cdot \frac{n^*(n^*+1)}{2} = \frac{n^*(n^*+1)}{4}$$

The variance follows from $\text{Var}(\phi_i) = 1/4$ and independence:

$$\text{Var}(W^+) = \sum_{i=1}^{n^*} R_i^2 \cdot \text{Var}(\phi_i) = \frac{1}{4} \sum_{i=1}^{n^*} i^2 = \frac{1}{4} \cdot \frac{n^*(n^*+1)(2n^*+1)}{6} = \frac{n^*(n^*+1)(2n^*+1)}{24}$$

## Wilcoxon Signed-Rank: Continuity Correction and Ties

**Continuity correction** for the normal approximation:

$$Z = \frac{W^+ - E[W^+] \pm 0.5}{\sqrt{\text{Var}(W^+)}}$$

Use $-0.5$ when $W^+ > E[W^+]$ (upper-tail test), $+0.5$ when $W^+ < E[W^+]$ (lower-tail test). R does this automatically.

. . .

**Correction for ties in ranking:**

When there are many ties among the $|D_i|$, the variance formula understates the true variance. The corrected formula is:

$$\text{Var}_{\text{ties}}(W^+) = \frac{1}{24}\left[n^*(n^*+1)(2n^*+1) - \frac{1}{2}\sum_{j} t_j(t_j-1)(t_j+1)\right]$$

where $t_j$ is the size of the $j$-th group of tied absolute differences.

- When there are no ties, this reduces to the standard formula
- Many ties $\Rightarrow$ reduced variance $\Rightarrow$ more conservative test  
- `wilcox.test()` in R applies this correction automatically when `exact = FALSE`

::: callout-important
## Practical Implication of Ties
Integer-valued outcomes (VAS pain, Likert scales) have many ties. The **exact** distribution is affected but R handles it. For data with many ties, the **sign test** or a permutation test may be preferred since rank-based tests assume continuous distributions.
:::

## Wilcoxon Signed-Rank: Medical Example

**Study:** 16 patients with Stage 1 hypertension. Systolic blood pressure (SBP, mmHg) measured before and 8 weeks after starting a new antihypertensive drug.

**Hypotheses:** $H_0$: median SBP change $= 0$ vs. $H_a$: median SBP change $< 0$ (drug lowers BP; $D_i = \text{after} - \text{before}$).

```{r}
#| echo: true
#| code-fold: false

sbp_before <- c(158, 142, 168, 135, 172, 145, 163, 138, 155, 149, 167, 141, 160, 153, 144, 170)
sbp_after  <- c(145, 148, 152, 131, 158, 139, 170, 135, 142, 143, 151, 150, 148, 140, 149, 155)
d_sbp      <- sbp_after - sbp_before

cat("D = after - before (mmHg):\n")
cat(d_sbp, "\n\n")
cat(sprintf("n = %d pairs | Positive (BP increased): %d | Negative (BP decreased): %d | Zero: %d\n",
            length(d_sbp), sum(d_sbp > 0), sum(d_sbp < 0), sum(d_sbp == 0)))
cat(sprintf("Mean change: %.1f mmHg | Median change: %.1f mmHg\n",
            mean(d_sbp), median(d_sbp)))
```

. . .

Most patients improved (SBP decreased), but **4 patients had higher SBP** after treatment — the kind of individual variability that makes nonparametric inference useful.

## Wilcoxon Signed-Rank: Step-by-Step Calculation

```{r}
#| echo: true
#| code-fold: false

# Step 1: Remove zeros (none here)
d_nz   <- d_sbp[d_sbp != 0]
n_star <- length(d_nz)

# Step 2: Rank absolute values (ties averaged automatically)
abs_ranks <- rank(abs(d_nz))

# Step 3: Compute W+
W_plus  <- sum(abs_ranks[d_nz > 0])
W_minus <- sum(abs_ranks[d_nz < 0])

cat(sprintf("n* = %d\n", n_star))
cat(sprintf("W+ = %.1f   (sum of ranks for %d positive differences)\n",
            W_plus, sum(d_nz > 0)))
cat(sprintf("W- = %.1f   (sum of ranks for %d negative differences)\n",
            W_minus, sum(d_nz < 0)))
cat(sprintf("Check: W+ + W- = %.1f  (should equal n*(n*+1)/2 = %.1f)\n\n",
            W_plus + W_minus, n_star*(n_star+1)/2))

# Normal approximation
mu_W  <- n_star * (n_star + 1) / 4
var_W <- n_star * (n_star + 1) * (2*n_star + 1) / 24
sd_W  <- sqrt(var_W)
z_W   <- (W_plus - mu_W) / sd_W  # one-sided: small W+ supports Ha

cat(sprintf("E[W+] = %.2f   Var(W+) = %.2f   SD(W+) = %.2f\n", mu_W, var_W, sd_W))
cat(sprintf("Z = (%.1f - %.2f) / %.2f = %.3f\n", W_plus, mu_W, sd_W, z_W))
cat(sprintf("One-sided p-value (normal approx): %.5f\n", pnorm(z_W)))
```

## Wilcoxon Signed-Rank: R Output and Hodges-Lehmann Estimate

```{r}
#| echo: true
#| code-fold: false

# R uses exact distribution for small n, Hodges-Lehmann estimate
wsr_result <- wilcox.test(sbp_after, sbp_before,
                           paired      = TRUE,
                           alternative = "less",
                           conf.int    = TRUE)
wsr_result |> tidy() |>
  select(statistic, p.value, conf.low, conf.high, method, alternative) |>
  print()
```

. . .

::: callout-note
## The Hodges-Lehmann Estimator

The **Hodges-Lehmann (HL) estimator** of location shift associated with the signed-rank test is:

$$\hat{\theta}_{HL} = \text{median}\left\{\frac{D_i + D_j}{2} : 1 \leq i \leq j \leq n^*\right\}$$

the median of all **Walsh averages** (pairwise averages of differences, including self-averages).

- This is a natural point estimate when using the signed-rank test
- The confidence interval from `wilcox.test(..., conf.int = TRUE)` is the **Hodges-Lehmann CI**
- More robust to outliers than the sample mean, but still uses magnitude information
:::

## Visualizing the Rankings for the SBP Data

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 15

sbp_df <- tibble(
  Patient = seq_along(d_sbp),
  Diff    = d_sbp,
  AbsDiff = abs(d_sbp),
  Sign    = ifelse(d_sbp > 0, "Positive: BP increased (D > 0)",
                              "Negative: BP decreased (D < 0)"),
  Rank    = rank(abs(d_sbp))
) |>
  filter(Diff != 0)

p_rank <- sbp_df |>
  ggplot(aes(x = reorder(factor(Patient), AbsDiff),
             y = AbsDiff,
             fill = Sign)) +
  geom_col(width = 0.65, color = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("R = %g", Rank), y = AbsDiff + 0.4),
            size = 4.5, vjust = 0, fontface = "bold") +
  scale_fill_manual(values = c(
    "Positive: BP increased (D > 0)" = "#e74c3c",
    "Negative: BP decreased (D < 0)" = "#006a4e"
  )) +
  labs(
    title    = "Wilcoxon Signed-Rank: Absolute SBP Changes with Ranks",
    subtitle = expression(W^"+" * " = sum of ranks for red bars (BP increased); small " *
                          W^"+" * " relative to E[" * W^"+" * "] = 68 supports H"[a]),
    x        = "Patient (ordered by |D|)",
    y        = "|SBP change| (mmHg)",
    fill     = "Direction of change"
  ) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 13))

p_rank
```

. . .

$W^+ = `r sum(rank(abs(d_sbp[d_sbp!=0]))[d_sbp[d_sbp!=0]>0])` \ll E[W^+] = `r 16*17/4`$ — the positive-difference ranks are concentrated at low values, indicating the drug tends to reduce SBP.

## Wilcoxon Signed-Rank: Visualizing the Paired Data

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 15

sbp_long <- tibble(
  Patient   = rep(1:16, 2),
  SBP       = c(sbp_before, sbp_after),
  Timepoint = rep(c("Before", "After"), each = 16),
  Direction = rep(ifelse(d_sbp < 0, "Decreased", "Increased"), 2)
) |>
  mutate(Timepoint = factor(Timepoint, levels = c("Before", "After")))

p_paired <- sbp_long |>
  ggplot(aes(x = Timepoint, y = SBP)) +
  geom_line(aes(group = Patient, color = Direction),
            linewidth = 0.8, alpha = 0.7) +
  geom_point(aes(color = Direction), size = 3, alpha = 0.9) +
  scale_color_manual(values = c("Decreased" = "#006a4e", "Increased" = "#e74c3c"),
                     labels = c("BP decreased", "BP increased")) +
  labs(title = "Systolic Blood Pressure Before and After Antihypertensive Treatment",
       subtitle = "Each line = one patient. Green = improved (BP decreased); Red = worsened (BP increased).",
       x = NULL, y = "SBP (mmHg)", color = "Change direction") +
  theme(legend.position = "bottom")

p_diff <- tibble(D = d_sbp, Patient = 1:16,
                 Direction = ifelse(d_sbp > 0, "Increased", "Decreased")) |>
  ggplot(aes(x = reorder(factor(Patient), D), y = D, fill = Direction)) +
  geom_col(width = 0.7, color = "white") +
  geom_hline(yintercept = 0, linewidth = 0.9) +
  geom_hline(yintercept = median(d_sbp), linetype = "dashed",
             color = "#3498db", linewidth = 1.2) +
  annotate("text", x = 1, y = median(d_sbp) - 1.5, hjust = 0,
           label = paste0("HL estimate = ", median(d_sbp), " mmHg"),
           color = "#3498db", size = 5) +
  scale_fill_manual(values = c("Increased" = "#e74c3c", "Decreased" = "#006a4e")) +
  labs(title = "Differences D = After \u2212 Before (mmHg)",
       x = "Patient (ordered by D)", y = "D (mmHg)", fill = "Direction") +
  theme(legend.position = "bottom")

p_paired + p_diff
```

## Comparing Sign Test, Wilcoxon Signed-Rank, and Paired *t* on SBP Data

```{r}
#| echo: true
#| code-fold: false

cat("=== SIGN TEST ===\n")
n_s  <- sum(d_sbp != 0)
sp   <- sum(d_sbp > 0)
p_sign <- pbinom(sp, n_s, 0.5)
cat(sprintf("S+ = %d   n* = %d   one-sided p = %.5f\n\n", sp, n_s, p_sign))

cat("=== WILCOXON SIGNED-RANK ===\n")
wilcox.test(sbp_after, sbp_before, paired = TRUE,
            alternative = "less", conf.int = TRUE) |>
  tidy() |>
  select(statistic, p.value, conf.low, conf.high) |>
  print()
cat("\n")

cat("=== PAIRED t-TEST ===\n")
t.test(sbp_after, sbp_before, paired = TRUE,
       alternative = "less") |>
  tidy() |>
  select(statistic, parameter, p.value, conf.low) |>
  print()
```

. . .

All three agree: strong evidence the drug reduces SBP. The paired $t$-test uses exact magnitudes; the Wilcoxon uses ranks; the sign test uses only directions.

# Wilcoxon Rank-Sum Test (DBC 14.3) {background-color="#27ae60"}

## Rank-Sum Test: Two Independent Samples

**Setting:** $X_{11}, \ldots, X_{1n_1} \stackrel{iid}{\sim} F_1$ and $X_{21}, \ldots, X_{2n_2} \stackrel{iid}{\sim} F_2$, independent samples.

**$H_0$:** $F_1 = F_2$ — the two samples come from the **same distribution**.

Under a **location-shift model** ($F_2(x) = F_1(x - \Delta)$), this is equivalent to $H_0: \Delta = 0$.

. . .

**Algorithm:**

1. Pool all $N = n_1 + n_2$ observations together; rank from 1 to $N$ (average ties)
2. Compute $W = \displaystyle\sum_{\text{group 1}} R_i$ = sum of ranks for group 1 observations
3. Under $H_0$: **every allocation of ranks to groups is equally likely**

$$E[W] = \frac{n_1(N+1)}{2}, \qquad \text{Var}(W) = \frac{n_1 n_2(N+1)}{12}$$

For large samples: $Z = \dfrac{W - E[W]}{\sqrt{\text{Var}(W)}} \;\dot{\sim}\; N(0,1)$

. . .

The exact null distribution of $W$ is computed by R for small $n_1, n_2$ (`wilcox.test` with `exact = TRUE`).

## Rank-Sum Test: Why This Null Distribution?

Under $H_0: F_1 = F_2$, the $N$ pooled observations are exchangeable. The group labels (group 1 or group 2) are equally likely to be assigned to any set of $n_1$ positions out of $N$.

There are $\binom{N}{n_1}$ equally-likely allocations.

$W$ = sum of group-1 ranks ranges from $\frac{n_1(n_1+1)}{2}$ (smallest $n_1$ ranks) to $\frac{n_1(2n_2+n_1+1)}{2}$ (largest $n_1$ ranks).

. . .

**Expected value:**

$$E[W] = n_1 \cdot E[\text{rank of a single obs.}] = n_1 \cdot \frac{N+1}{2}$$

since under $H_0$, each rank from 1 to $N$ is equally likely for any observation.

**Variance:**

$$\text{Var}(W) = \frac{n_1 n_2 (N+1)}{12}$$

derived from the variance of sampling without replacement from $\{1, \ldots, N\}$.

## The Mann–Whitney $U$ Statistic

The **Mann-Whitney $U$** is equivalent to $W$ and has a useful probabilistic interpretation:

$$U = W - \frac{n_1(n_1+1)}{2}$$

**$U$ counts the number of concordant pairs:**

$$U = \#\{(i,j) : X_{1i} > X_{2j}\}$$

out of all $n_1 \times n_2$ pairs. Under $H_0$: $E[U] = n_1 n_2 / 2$.

. . .

$U / (n_1 n_2)$ is a consistent estimator of $P(X_1 > X_2)$:

- $P(X_1 > X_2) = 1/2$ under $H_0$
- $P(X_1 > X_2) > 1/2$ when group 1 tends to be larger
- This is the **area under the ROC curve (AUC)** — a key measure in diagnostic medicine

::: callout-tip
## Mann-Whitney U and the ROC Curve
If you use group membership as a "predictor" and a continuous biomarker as the "score," the Mann-Whitney $U$ statistic normalized by $n_1 n_2$ equals the AUC of the ROC curve. This connects nonparametric testing directly to classification performance.
:::

## Rank-Sum Test: Ties Correction

When there are ties among the pooled observations, the variance is corrected:

$$\text{Var}_{\text{ties}}(W) = \frac{n_1 n_2}{12} \left[ N + 1 - \frac{\sum_j t_j(t_j^2-1)}{N(N-1)} \right]$$

where $t_j$ = size of the $j$-th tied group in the combined sample.

- Ties reduce the variance (less information in the ranks)
- The correction makes the $Z$-statistic slightly larger in absolute value
- R applies this automatically in `wilcox.test()`

. . .

**Hodges-Lehmann Estimator for Two Samples:**

$$\hat{\Delta}_{HL} = \text{median}\{X_{1i} - X_{2j} : i = 1,\ldots,n_1;\ j = 1,\ldots,n_2\}$$

The median of all $n_1 \times n_2$ pairwise differences. This is the natural point estimate of the location shift $\Delta$ under the rank-sum framework.

## Rank-Sum Test: Medical Example (Serum CRP)

**Study:** Compare serum CRP (mg/L), a right-skewed inflammatory biomarker, between patients **with** and **without** post-operative infections.

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
crp_infect  <- rgamma(18, shape = 2.0, rate = 0.08)  # mean ~25 mg/L
crp_control <- rgamma(20, shape = 1.2, rate = 0.12)  # mean ~10 mg/L

cat(sprintf("Infection:  n = %d   mean = %.1f   median = %.1f   sd = %.1f   range = [%.1f, %.1f]\n",
            length(crp_infect),  mean(crp_infect),
            median(crp_infect),  sd(crp_infect),
            min(crp_infect), max(crp_infect)))
cat(sprintf("Control:    n = %d   mean = %.1f   median = %.1f   sd = %.1f   range = [%.1f, %.1f]\n",
            length(crp_control), mean(crp_control),
            median(crp_control), sd(crp_control),
            min(crp_control), max(crp_control)))
```

```{r}
#| echo: true
#| code-fold: false

N  <- length(crp_infect) + length(crp_control)
n1 <- length(crp_infect)
n2 <- length(crp_control)

W_obs  <- sum(rank(c(crp_infect, crp_control))[1:n1])
E_W    <- n1 * (N + 1) / 2
Var_W  <- n1 * n2 * (N + 1) / 12
Z_W    <- (W_obs - E_W) / sqrt(Var_W)

cat(sprintf("N = %d   n1 = %d   n2 = %d\n", N, n1, n2))
cat(sprintf("W = %.1f   E[W] = %.1f   SD(W) = %.2f\n", W_obs, E_W, sqrt(Var_W)))
cat(sprintf("Z = %.3f   one-sided p (approx) = %.5f\n\n", Z_W, pnorm(Z_W, lower.tail = FALSE)))

cat("--- wilcox.test (exact + Hodges-Lehmann CI) ---\n")
wilcox.test(crp_infect, crp_control,
            alternative = "greater", conf.int = TRUE) |> tidy() |>
  select(statistic, p.value, conf.low, conf.high, method) |> print()
```

## Rank-Sum Test: Comparison with Welch *t*

```{r}
#| echo: true
#| code-fold: false

cat("--- Welch t-test ---\n")
t.test(crp_infect, crp_control, alternative = "greater") |>
  tidy() |>
  select(statistic, parameter, p.value, conf.low, method) |>
  print()
```

. . .

Both tests reject $H_0$, but note:

- CRP is strongly **right-skewed** with outliers → $t$-test p-value is less reliable
- Wilcoxon rank-sum p-value is based on a **distribution-free** null
- Hodges-Lehmann estimate of location shift: `conf.low` from `wilcox.test`
- With small $n$ and skewed biomarkers, the rank-sum test is the **preferred primary analysis**

## Visualizing the CRP Data

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6.5
#| fig-width: 15

p_vio <- tibble(
  CRP   = c(crp_infect, crp_control),
  Group = rep(c("Infection", "Control"),
              c(length(crp_infect), length(crp_control)))
) |>
  ggplot(aes(x = Group, y = CRP, fill = Group)) +
  geom_violin(alpha = 0.4, trim = FALSE) +
  geom_boxplot(width = 0.18, fill = "white",
               outlier.shape = 1, outlier.size = 2.5) +
  stat_summary(fun = median, geom = "point",
               shape = 18, size = 4, color = "black") +
  scale_fill_manual(values = c("Infection" = "#e74c3c",
                                "Control"   = "#006a4e")) +
  labs(
    title    = "Serum CRP by Post-Operative Infection Status",
    subtitle = "Right-skewed biomarker: Wilcoxon rank-sum preferred. Diamond = median.",
    x = NULL, y = "CRP (mg/L)"
  ) +
  theme(legend.position = "none")

# Show ranks
all_crp   <- c(crp_infect, crp_control)
all_group <- rep(c("Infection", "Control"), c(n1, n2))
p_ranks <- tibble(CRP = all_crp, Group = all_group,
                   Rank = rank(all_crp)) |>
  ggplot(aes(x = Rank, y = CRP, color = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("Infection" = "#e74c3c", "Control" = "#006a4e")) +
  geom_hline(yintercept = 0, linewidth = 0.2) +
  labs(title = "CRP Values vs. Pooled Ranks",
       subtitle = "Infection group tends to hold the higher ranks (rightward)",
       x = "Pooled rank", y = "CRP (mg/L)", color = NULL) +
  theme(legend.position = "bottom")

p_vio + p_ranks
```

# Asymptotic Relative Efficiency {background-color="#9b59b6"}

## What Is Asymptotic Relative Efficiency?

**Definition:** The ARE of test $T_1$ relative to test $T_2$ is:

$$\text{ARE}(T_1, T_2) = \lim_{n \to \infty} \frac{n_2^*}{n_1^*}$$

where $n_i^*$ = sample size needed by test $T_i$ to achieve the **same power** at the same significance level $\alpha$ and the same effect size.

. . .

**Interpretation:**

- $\text{ARE}(T_1, T_2) = 0.95$: $T_1$ needs $n_2 / 0.95 \approx 1.05 n_2$ observations — 5% more
- $\text{ARE}(T_1, T_2) = 1.50$: $T_1$ needs only $n_2 / 1.50 \approx 0.67 n_2$ — 33% fewer!
- $\text{ARE}(T_1, T_2) = 0.50$: $T_1$ needs twice as many observations

. . .

ARE is computed under a specific distributional family (the **Pitman local alternatives** framework) and tells us about asymptotic power, not finite-sample behavior — but it is a reliable guide in practice.

## ARE: Pitman Efficiency Results

The Pitman ARE for location-shift alternatives $H_a: \theta = \theta_n = c/\sqrt{n}$ (DBC 14.5, DBC p. 681):

| Distribution of $D_i$ | ARE(sign / $t$) | ARE(signed-rank / $t$) |
|-----------------------|:---:|:---:|
| Normal | $2/\pi \approx 0.637$ | $3/\pi \approx 0.955$ |
| Logistic | $\pi^2/12 \approx 0.822$ | $\pi^2/9 \approx 1.097$ |
| Double exponential (Laplace) | $2.000$ | $3/2 = 1.500$ |
| Uniform | $1/3 \approx 0.333$ | $1.000$ |
| Cauchy (very heavy tails) | $\infty$ | $\infty$ |
| $t_\nu$ ($\nu$ small) | $> 1$ | $> 1$ |

. . .

::: callout-important
## The Remarkable Result for the Signed-Rank Test

- At **normality**: ARE(signed-rank, $t$) = $3/\pi \approx 0.955$ — only **4.5% efficiency loss**
- The signed-rank test is **never catastrophically bad** — for all symmetric distributions, ARE $\geq 0.864$ (Hodges-Lehmann lower bound)
- For heavy-tailed or non-normal distributions, the signed-rank test can be **substantially more powerful**

This makes it an excellent "default" for small $n$ when normality is uncertain.
:::

## Where Do the ARE Formulas Come From?

For the **signed-rank test** vs. the $t$-test under the normal distribution:

The ARE can be computed using the **Pitman formula**:

$$\text{ARE}(W^+, t) = \frac{\left[E\left\{f(D)\right\}\right]^2 \cdot \text{Var}(D)}{\left[\frac{1}{2}f(0)\right]^2}$$

where $f$ is the density of $D_i$.

For $D_i \sim N(0, \sigma^2)$:

- $f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-x^2/(2\sigma^2)}$, so $f(0) = \frac{1}{\sigma\sqrt{2\pi}}$
- $E\{f(D)\} = \int_{-\infty}^{\infty} f(x)^2 \, dx = \frac{1}{2\sigma\sqrt{\pi}}$

This gives $\text{ARE} = \frac{3}{\pi} \approx 0.955$.

The analogous result for the sign test is $\text{ARE}(\text{sign}, t) = 4\sigma^2 f(0)^2 = 2/\pi$ under normality.

## ARE: Power Simulation Across Distributions

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 16

set.seed(551)
B_are <- 2000

sim_power <- function(n, effect, dist_fn) {
  replicate(B_are, {
    d <- dist_fn(n) + effect
    nt <- sum(d != 0)
    sp <- sum(d > 0)
    c(
      t_test   = t.test(d)$p.value,
      sign_p   = 2 * min(pbinom(sp, nt, 0.5),
                         pbinom(sp - 1, nt, 0.5, lower.tail = FALSE)),
      wilcoxon = wilcox.test(d, exact = FALSE)$p.value
    )
  }) |> t() |> as_tibble() |>
    summarise(across(everything(), \(x) mean(x < 0.05)))
}

dists <- list(
  list(fn = function(n) rnorm(n),                    label = "Normal"),
  list(fn = function(n) rt(n, df = 3),               label = expression(t[3] ~ "(heavy tails)")),
  list(fn = function(n) rexp(n) - log(2),            label = "Exponential\n(shifted, skewed)"),
  list(fn = function(n) rnorm(n, sd = 1) +
         (runif(n) < 0.2) * rnorm(n, mean = 4, sd = 2), label = "Contaminated Normal\n(outlier-prone)")
)

results_are <- map_dfr(dists, function(dd) {
  map_dfr(c(10, 20, 30, 50), function(n) {
    sim_power(n, effect = 0.5, dist_fn = dd$fn) |>
      mutate(n = n, Distribution = as.character(dd$label))
  })
})

results_are |>
  pivot_longer(c(t_test, sign_p, wilcoxon),
               names_to = "Test", values_to = "Power") |>
  mutate(Test = recode(Test,
                       t_test   = "Paired t-test",
                       sign_p   = "Sign test",
                       wilcoxon = "Wilcoxon signed-rank")) |>
  ggplot(aes(x = n, y = Power, color = Test, linetype = Test)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "grey50") +
  scale_color_manual(values = c("Paired t-test"        = "#006a4e",
                                 "Sign test"            = "#e74c3c",
                                 "Wilcoxon signed-rank" = "#3498db")) +
  scale_linetype_manual(values = c("Paired t-test"        = "solid",
                                    "Sign test"            = "dashed",
                                    "Wilcoxon signed-rank" = "solid")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  facet_wrap(~Distribution, scales = "free_x") +
  labs(
    title    = "Power vs. Sample Size: Sign Test, Wilcoxon Signed-Rank, Paired t-test",
    subtitle = "Effect size = 0.5. Signed-rank matches t-test at normality; both tests excel for heavy tails; sign test strongest for very skewed data",
    x = "Sample size n (pairs)", y = "Power",
    color = NULL, linetype = NULL
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12, face = "bold"))
```

## ARE: Interpreting the Simulation

The simulation confirms the Pitman ARE theory:

- **Normal distribution**: Signed-rank ≈ $t$-test power (ARE $\approx 0.955$). Sign test noticeably weaker (ARE $\approx 0.637$).
- **$t_3$ distribution (heavy tails)**: Both nonparametric tests **outperform** the $t$-test. Signed-rank especially strong.
- **Exponential (skewed)**: Signed-rank invalid in principle (asymmetric $D_i$), but often still competitive in practice. Sign test surprisingly strong.
- **Contaminated normal**: Even modest outlier contamination makes nonparametric tests superior.

. . .

::: callout-note
## Practical Takeaway for Biomedical Research

Many biomarkers (CRP, troponin, liver enzymes) follow right-skewed, heavy-tailed distributions.

For such data with small $n$:

- **Sign test**: if ordinal data or very extreme outliers
- **Wilcoxon signed-rank**: best default for continuous paired data with uncertain normality
- **Wilcoxon rank-sum**: best default for independent two-group comparisons
:::

# Comparing All Three Tests {background-color="#2c3e50"}

## Decision Flowchart: Which Test?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 14

grViz("
digraph test_selection {

  graph [layout = dot, rankdir = TB, fontname = 'Helvetica', nodesep = 0.6]

  node [shape = diamond, style = filled, fillcolor = '#f0f0f0',
        fontsize = 14, margin = '0.2,0.1', fontname = 'Helvetica']
  Q1 [label = 'Paired data or\\none-sample?']
  Q2 [label = 'Data ordinal or\\nhighly asymmetric?']
  Q3 [label = 'n large or\\nnormality plausible?']
  Q4 [label = 'n large or\\nnormality plausible?']

  node [shape = box, style = filled, fontsize = 14,
        margin = '0.2,0.1', fontname = 'Helvetica']
  T1 [label = 'Sign Test\\n(Binomial)', fillcolor = '#3498db', fontcolor = white]
  T2 [label = 'Wilcoxon\\nSigned-Rank', fillcolor = '#e74c3c', fontcolor = white]
  T3 [label = 'Paired t-test', fillcolor = '#006a4e', fontcolor = white]
  T4 [label = 'Wilcoxon\\nRank-Sum', fillcolor = '#27ae60', fontcolor = white]
  T5 [label = 'Welch t-test\\n(two independent)', fillcolor = '#9b59b6', fontcolor = white]

  Q1 -> Q2 [label = 'Yes']
  Q1 -> Q4 [label = 'No (two independent)']
  Q2 -> T1 [label = 'Yes / Ordinal']
  Q2 -> Q3 [label = 'No (continuous, symmetric)']
  Q3 -> T3 [label = 'Yes']
  Q3 -> T2 [label = 'No (small n, non-normal)']
  Q4 -> T5 [label = 'Yes']
  Q4 -> T4 [label = 'No (small n, non-normal)']
}
")
```

## Side-by-Side on the Pain Score Data (Sign Test Example)

```{r}
#| echo: true
#| code-fold: false

cat("=== SIGN TEST ===\n")
n_s <- sum(d_pain != 0); sp <- sum(d_pain > 0)
cat(sprintf("S+ = %d   n* = %d   one-sided p = %.5f\n\n",
            sp, n_s, pbinom(sp, n_s, 0.5)))

cat("=== WILCOXON SIGNED-RANK ===\n")
wilcox.test(pain_after, pain_before, paired = TRUE,
            alternative = "less", conf.int = TRUE) |>
  tidy() |>
  select(statistic, p.value, conf.low, conf.high, method) |>
  print()
cat("\n")

cat("=== PAIRED t-TEST ===\n")
t.test(pain_after, pain_before, paired = TRUE,
       alternative = "less") |>
  tidy() |>
  select(statistic, parameter, p.value, conf.low, method) |>
  print()
```

. . .

All three strongly reject $H_0$. The differences in $p$-values reflect what information each test uses.

## Summary Table: When to Use Each Test

| Test | Use When | Caution |
|------|----------|---------|
| **Sign test** | Ordinal data; only sign matters; extreme asymmetry of $D_i$ | Wastes magnitude information; lowest power |
| **Wilcoxon signed-rank** | Continuous paired data; small $n$; uncertain normality | Requires **symmetric** $D_i$; not for ordinal-only data |
| **Wilcoxon rank-sum** | Two independent continuous groups; small $n$; skewed | Requires **continuous** distributions; exact p-value affected by ties |
| **Paired $t$-test** | Paired; approximately normal $D_i$ or $n \geq 30$ | Sensitive to heavy tails and outliers for small $n$ |
| **Welch $t$-test** | Two independent groups; approximately normal or $n_i \geq 30$ | Heavy tails inflate Type I error for small $n$ |

. . .

::: callout-note
## Default Recommendation for Biomedical Research

For **paired** continuous $D_i$ with $n < 30$ and uncertain normality: use **Wilcoxon signed-rank**.
For **two independent** groups with small $n$ and skewed biomarkers: use **Wilcoxon rank-sum**.

If assumptions are met, the parametric test gains only ~5% efficiency — a small price for robustness.
:::

# Key Takeaways {background-color="#2c3e50"}

## Key Takeaways (DBC 14.1–14.3)

::: callout-important
## Nonparametric Inference: Main Points

1. **Sign test:** $S^+ \sim \text{Bin}(n^*, 1/2)$ under $H_0$; uses only direction of $D_i$; valid for ordinal data; no symmetry assumption; lowest power of the three

2. **Wilcoxon signed-rank:** ranks $|D_i|$, sums ranks of positive differences; requires symmetric (not necessarily normal) continuous $D_i$; ARE vs. $t$-test $= 3/\pi \approx 0.955$ at normality — minimal efficiency loss, with gains for non-normal data; Hodges-Lehmann estimator is the associated point estimate

3. **Wilcoxon rank-sum (Mann–Whitney $U$):** pools and ranks all $N$ observations; sums group-1 ranks; tests $H_0: F_1 = F_2$; $U/(n_1 n_2)$ estimates $P(X_1 > X_2)$ = AUC; Hodges-Lehmann estimate = median of all $n_1 \times n_2$ pairwise differences

4. **ARE quantifies the asymptotic power trade-off:** signed-rank ARE $\geq 0.864$ vs. $t$-test for all symmetric distributions (Hodges-Lehmann lower bound) — never much worse at normality, potentially much better for heavy tails

5. **Nonparametric $\neq$ assumption-free:** signed-rank requires **symmetric** $D_i$; rank-sum requires **continuous** distributions; both require **independent** observations; ties require variance corrections

6. **Variance corrections for ties:** both tests have modified variance formulas; R applies them automatically; heavy ties (e.g., ordinal data) reduce precision of rank-based tests — consider the sign test or permutation tests
:::

## Looking Ahead

**Next class (Lesson 22): Course Review**

- Complete roadmap: from point estimation to nonparametric inference
- Key theorems, formulas, and connections
- Method selection guide
- R function reference and practice problems

## Homework Problems

**From Devore, Berk, and Carlton (Chapter 14):**

::: callout-note
## Problems

1. **(DBC 14.1 / Sign Test)** A clinical trial measured resting heart rate (bpm) in 12 patients before and after beta-blocker therapy. The differences (after − before) were: $−8, −14, +3, −6, −11, −5, +2, −9, −7, −4, −12, −6$. 
   - (a) Carry out the sign test with $H_a$: median $< 0$ at $\alpha = 0.05$. Compute the exact p-value.
   - (b) What is $S^+$ and what is its null distribution? Why is this distribution exact?

2. **(DBC 14.2 / Wilcoxon Signed-Rank)** Using the heart rate data from Problem 1:
   - (a) Compute ranks of the absolute differences, noting any ties.
   - (b) Compute $W^+$, $E[W^+]$, and $\text{Var}(W^+)$. 
   - (c) Compute the large-sample $Z$ statistic and p-value. Apply the continuity correction.
   - (d) Verify with `wilcox.test()` in R. Why might the exact and asymptotic p-values differ?

3. **(DBC 14.3 / Rank-Sum)** Serum albumin levels (g/dL) were measured in critically ill patients with (Group 1, $n_1 = 10$) and without (Group 2, $n_2 = 12$) acute kidney injury. Use R to:
   - (a) Conduct the Wilcoxon rank-sum test ($H_a$: Group 1 median $<$ Group 2 median).
   - (b) Report the Hodges-Lehmann estimate and 95\% CI for the location shift.
   - (c) Compare conclusions to a Welch $t$-test. Which test is more appropriate here, and why?

4. **(ARE / Conceptual)** Suppose you have $n = 25$ paired differences known to follow a logistic distribution.
   - (a) The ARE of the Wilcoxon signed-rank test relative to the paired $t$-test is $\pi^2/9 \approx 1.097$. Interpret this: how many pairs would the $t$-test need to match the signed-rank test's power?
   - (b) Give an intuitive explanation for why the rank-based test outperforms the $t$-test for logistic data.
:::

## References

::: nonincremental
- Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Sections 14.1, 14.2, 14.3, 14.5.
- Chihara, L. M. and Hesterberg, T. C. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 3.
- Hodges, J. L. and Lehmann, E. L. (1963). Estimates of location based on rank tests. *Annals of Mathematical Statistics*, 34(2), 598–611.
- Wilcoxon, F. (1945). Individual comparisons by ranking methods. *Biometrics Bulletin*, 1(6), 80–83.
- Mann, H. B. and Whitney, D. R. (1947). On a test of whether one of two random variables is stochastically larger than the other. *Annals of Mathematical Statistics*, 18(1), 50–60.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/21_NonparametricInference/21_NonparametricInference.html lessons/21_NonparametricInference/21_NonparametricInference.pdf
```
