---
title: "Lesson 24: Course Review"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "2026-03-11"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 24 - Course Review
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 24: Course Review {background-color="#2c3e50"}

## The Big Picture: What This Course Was About

::: callout-note
## Central Question of Statistical Inference
Given a **random sample** $X_1, X_2, \ldots, X_n$ from a population with unknown parameter $\theta$, how do we:

1. **Estimate** $\theta$? (Point estimation: Lessons 1–6)
2. **Quantify uncertainty** about $\theta$? (Confidence intervals: Lessons 7–11)
3. **Test claims** about $\theta$? (Hypothesis testing: Lessons 12–21)
4. **Connect** to broader frameworks? (OLS, Bayesian: Lessons 22–23)
:::

. . .

**The three pillars share a foundation:** the sampling distribution of estimators.

## Course Roadmap

```{r}
#| echo: false
#| fig-width: 18
#| fig-height: 9
library(DiagrammeR)

grViz("
digraph course_map {
  graph [rankdir = TB, bgcolor = transparent, fontsize = 28]
  node [shape = box, style = filled, fontsize = 22, fontname = 'Helvetica']

  subgraph cluster_pe {
    label = 'Point Estimation (Lessons 1-6)'
    style = filled; color = '#3498db30'; fontsize = 24
    pe1 [label = 'Statistics &\\nSampling Distributions', fillcolor = '#85c1e9']
    pe2 [label = 'Bias, Variance,\\nMSE', fillcolor = '#85c1e9']
    pe3 [label = 'MVUE, CRLB,\\nFisher Information', fillcolor = '#85c1e9']
    pe4 [label = 'MOM & MLE', fillcolor = '#85c1e9']
    pe5 [label = 'Sufficiency,\\nRao-Blackwell', fillcolor = '#85c1e9']
  }

  subgraph cluster_ci {
    label = 'Confidence Intervals (Lessons 7-11)'
    style = filled; color = '#27ae6030'; fontsize = 24
    ci1 [label = 'Pivotal Quantities\\n(z, t)', fillcolor = '#82e0aa']
    ci2 [label = 'Wald, Score,\\nAgresti-Coull', fillcolor = '#82e0aa']
    ci3 [label = 'Bootstrap CIs\\n(percentile, t)', fillcolor = '#82e0aa']
    ci4 [label = 'Delta Method', fillcolor = '#82e0aa']
  }

  subgraph cluster_ht {
    label = 'Hypothesis Testing (Lessons 12-21)'
    style = filled; color = '#e74c3c30'; fontsize = 24
    ht1 [label = 'Null/Alt Hypotheses\\nType I/II Errors', fillcolor = '#f1948a']
    ht2 [label = 'z, t Tests\\nProportion Tests', fillcolor = '#f1948a']
    ht3 [label = 'P-values, Power,\\nSample Size', fillcolor = '#f1948a']
    ht4 [label = 'Neyman-Pearson\\nLikelihood Ratio', fillcolor = '#f1948a']
    ht5 [label = 'Bootstrap &\\nPermutation Tests', fillcolor = '#f1948a']
    ht6 [label = 'Two-Sample &\\nPaired Tests', fillcolor = '#f1948a']
    ht7 [label = 'Nonparametric:\\nSign, Wilcoxon', fillcolor = '#f1948a']
  }

  subgraph cluster_ext {
    label = 'Extensions (Lessons 22-23)'
    style = filled; color = '#9b59b630'; fontsize = 24
    ext1 [label = 'Gauss-Markov\\nOLS Theory', fillcolor = '#c39bd3']
    ext2 [label = 'Bayesian\\nInference', fillcolor = '#c39bd3']
  }

  pe4 -> ci2 [label = 'MLE → Wald CI', fontsize = 16]
  pe3 -> ci2 [label = 'Fisher Info → SE', fontsize = 16]
  ci1 -> ht2 [label = 'Duality', fontsize = 16]
  ht4 -> ext2 [label = 'Likelihood', fontsize = 16]
  pe5 -> ext1 [label = 'BLUE', fontsize = 16]
}
")
```


# I. Point Estimation (Lessons 1–6) {background-color="#3498db"}

## Key Concepts: Evaluating Estimators

Given data $X_1, \ldots, X_n \stackrel{iid}{\sim} f(x|\theta)$, a **statistic** $T = g(X_1, \ldots, X_n)$ is an **estimator** of $\theta$.

. . .

**Criteria for a good estimator:**

| Property | Definition | Measures |
|----------|-----------|----------|
| **Bias** | $\text{Bias}(\hat{\theta}) = E[\hat{\theta}] - \theta$ | Systematic error |
| **Variance** | $\text{Var}(\hat{\theta}) = E[(\hat{\theta} - E[\hat{\theta}])^2]$ | Precision / spread |
| **MSE** | $\text{MSE}(\hat{\theta}) = \text{Var}(\hat{\theta}) + [\text{Bias}(\hat{\theta})]^2$ | Overall accuracy |
| **Consistency** | $\hat{\theta}_n \xrightarrow{P} \theta$ as $n \to \infty$ | Large-sample reliability |

. . .

::: callout-tip
## The Bias-Variance Tradeoff
A biased estimator with small variance (low MSE) can be **preferred** over an unbiased estimator with large variance. MSE captures this tradeoff.
:::

## Finding Estimators: MOM and MLE

::: columns
::: {.column width="48%"}
**Method of Moments (MOM)**

Set sample moments equal to population moments:
$$\frac{1}{n}\sum_{i=1}^n X_i^k = E[X^k] \quad k = 1, 2, \ldots$$

- Simple to compute
- Not always efficient
- May not respect parameter constraints
:::

::: {.column width="48%"}
**Maximum Likelihood (MLE)**

Maximize the likelihood (or log-likelihood):
$$\hat{\theta}_{MLE} = \arg\max_\theta L(\theta | \mathbf{x}) = \arg\max_\theta \ell(\theta | \mathbf{x})$$

- Asymptotically efficient: $\text{Var}(\hat{\theta}_{MLE}) \to \frac{1}{I_n(\theta)}$
- Invariant: $g(\hat{\theta}_{MLE}) = \widehat{g(\theta)}_{MLE}$
- Asymptotically normal
:::
:::

## Optimality Theory: The Quest for the Best Estimator

**Goal:** Among all unbiased estimators, find the one with **minimum variance** (MVUE).

. . .

**Cramér-Rao Lower Bound (CRLB):**

For any unbiased estimator $\hat{\theta}$:
$$\text{Var}(\hat{\theta}) \geq \frac{1}{I_n(\theta)} = \frac{1}{nI_1(\theta)}$$

where the **Fisher Information** for a single observation is:
$$I_1(\theta) = E\left[-\frac{\partial^2}{\partial\theta^2}\ln f(X|\theta)\right] = \text{Var}\left(\frac{\partial}{\partial\theta}\ln f(X|\theta)\right)$$

. . .

**If** $\text{Var}(\hat{\theta}) = \frac{1}{I_n(\theta)}$, then $\hat{\theta}$ **is MVUE** (attains the CRLB).

## Finding MVUE: The Complete Toolkit

| Method | Approach | When to Use |
|--------|----------|-------------|
| **CRLB attainment** | Show $\text{Var}(\hat{\theta}) = 1/I_n(\theta)$ | If equality holds |
| **Rao-Blackwell** | Condition on sufficient statistic: $E[\hat{\theta}|T]$ | Improve an unbiased estimator |
| **Lehmann-Scheffé** | Unbiased + function of complete sufficient statistic | Most general approach |

. . .

**Sufficiency** (Fisher-Neyman Factorization): $T$ is sufficient for $\theta$ if
$$f(\mathbf{x}|\theta) = g(T(\mathbf{x}), \theta) \cdot h(\mathbf{x})$$

The sufficient statistic captures **all information** about $\theta$ in the data.

## Quick Example: Normal Mean

**Data:** $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$ with $\sigma^2$ known.

```{r}
#| code-fold: true
#| fig-height: 5.5
#| fig-width: 14
# Demonstrate CRLB: comparing estimators of normal mean
set.seed(551)
n <- 10
n_sim <- 10000
true_mu <- 5

sims <- tibble(sim = 1:n_sim) |>
  mutate(
    data = map(sim, ~rnorm(n, mean = true_mu, sd = 2)),
    xbar = map_dbl(data, mean),
    median = map_dbl(data, median),
    midrange = map_dbl(data, ~(max(.x) + min(.x)) / 2)
  )

sims_long <- sims |>
  select(sim, xbar, median, midrange) |>
  pivot_longer(-sim, names_to = "Estimator", values_to = "estimate") |>
  mutate(Estimator = fct_recode(Estimator,
    "Sample Mean" = "xbar",
    "Sample Median" = "median",
    "Midrange" = "midrange"
  ))

# CRLB for this problem: sigma^2 / n = 4/10 = 0.4
crlb_var <- 4 / n

var_summary <- sims_long |>
  group_by(Estimator) |>
  summarise(
    variance = var(estimate),
    bias = mean(estimate) - true_mu,
    .groups = "drop"
  )

ggplot(sims_long, aes(x = estimate, fill = Estimator)) +
  geom_histogram(bins = 60, alpha = 0.7, color = "white", position = "identity") +
  facet_wrap(~Estimator, nrow = 1) +
  geom_vline(xintercept = true_mu, linewidth = 1.2, linetype = "dashed") +
  geom_text(data = var_summary, 
            aes(x = 7.5, y = Inf, 
                label = paste0("Var = ", round(variance, 3), "\nBias = ", round(bias, 3))),
            vjust = 1.5, size = 5, inherit.aes = FALSE) +
  annotate("text", x = 7.5, y = 0, label = paste0("CRLB = ", crlb_var), 
           vjust = -0.5, size = 4.5, fontface = "italic", color = "red") +
  scale_fill_manual(values = c("#3498db", "#e74c3c", "#f39c12")) +
  labs(x = expression(hat(mu)), y = "Count",
       title = expression(paste("Comparing Estimators of ", mu, " (Normal data, n = 10, ", sigma^2, " = 4)")),
       subtitle = "Sample mean attains the CRLB — it is MVUE") +
  theme(legend.position = "none")
```


# II. Confidence Intervals (Lessons 7–11) {background-color="#27ae60"}

## CI Construction: Three Approaches

::: callout-important
## Core Idea
A $100(1-\alpha)\%$ confidence interval provides a range of **plausible values** for $\theta$ based on the observed data. In repeated sampling, $100(1-\alpha)\%$ of such intervals contain the true $\theta$.
:::

. . .

**Approach 1: Pivotal Quantities** — Find $Q(X_1, \ldots, X_n, \theta)$ whose distribution is known and free of $\theta$, then invert.

**Approach 2: Asymptotic (Wald / Score)** — Use the asymptotic normality of the MLE.

**Approach 3: Bootstrap** — Resample from the data to approximate the sampling distribution.


## CI Quick Reference

| Setting | Method | Formula |
|---------|--------|---------|
| $\mu$, $\sigma$ known | z-interval | $\bar{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$ |
| $\mu$, $\sigma$ unknown | t-interval | $\bar{x} \pm t_{\alpha/2, n-1}\frac{s}{\sqrt{n}}$ |
| Proportion $p$ | Wald | $\hat{p} \pm z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$ |
| Proportion $p$ | Score (Wilson) | $\tilde{p} \pm z_{\alpha/2}\frac{\sqrt{\hat{p}(1-\hat{p})/n + z^2/(4n^2)}}{1 + z^2/n}$ |
| Proportion $p$ | Agresti-Coull | Wald with $\tilde{p} = \frac{X + 2}{n + 4}$ |
| General $\theta$ | Wald | $\hat{\theta}_{MLE} \pm z_{\alpha/2}\sqrt{1/I_n(\hat{\theta})}$ |
| General $\theta$ | Bootstrap percentile | $(\hat{\theta}^*_{\alpha/2}, \; \hat{\theta}^*_{1-\alpha/2})$ |
| General $\theta$ | Bootstrap t | $\hat{\theta} \pm t_{\alpha/2, n-1} \cdot SE_{boot}$ |
| General $\theta$ | Studentized bootstrap | $\hat{\theta} - t^*_{1-\alpha/2} \cdot SE_{\hat{\theta}}, \;\; \hat{\theta} - t^*_{\alpha/2} \cdot SE_{\hat{\theta}}$ |

## The Delta Method

When we want a CI for $g(\theta)$ but have an estimator of $\theta$:

::: callout-note
## Delta Method
If $\hat{\theta} \stackrel{approx}{\sim} N(\theta, \sigma^2_{\hat{\theta}})$, then for differentiable $g$:
$$g(\hat{\theta}) \stackrel{approx}{\sim} N\left(g(\theta), \; [g'(\theta)]^2 \cdot \sigma^2_{\hat{\theta}}\right)$$

Equivalently: $SE[g(\hat{\theta})] \approx |g'(\hat{\theta})| \cdot SE(\hat{\theta})$
:::

. . .

**Example:** If $\hat{\lambda}_{MLE} = 1/\bar{X}$ for Exponential data and we want a CI for the **mean** $\mu = 1/\lambda$:

- $g(\lambda) = 1/\lambda$, so $g'(\lambda) = -1/\lambda^2$
- $SE(g(\hat{\lambda})) = \frac{1}{\hat{\lambda}^2} \cdot \frac{\hat{\lambda}}{\sqrt{n}} = \frac{1}{\hat{\lambda}\sqrt{n}} = \frac{\bar{X}}{\sqrt{n}}$

## Bootstrap: When and How

```{r}
#| echo: false
#| fig-height: 7
#| fig-width: 16
boot_flow <- tibble(
  step = factor(c("1. Original\nSample", "2. Resample\nwith Replacement", 
                   "3. Compute\nStatistic", "4. Repeat\nB times", 
                   "5. Bootstrap\nDistribution"),
                levels = c("1. Original\nSample", "2. Resample\nwith Replacement",
                           "3. Compute\nStatistic", "4. Repeat\nB times",
                           "5. Bootstrap\nDistribution")),
  x = 1:5, y = 1
)

ggplot(boot_flow, aes(x = x, y = y)) +
  geom_point(size = 70, color = c("#3498db", "#27ae60", "#e74c3c", "#f39c12", "#9b59b6")) +
  geom_text(aes(label = step), size = 7, fontface = "bold", color = "white") +
  geom_segment(aes(x = x + 0.35, xend = x + 0.65, y = 1, yend = 1),
               data = boot_flow |> filter(x < 5),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2, color = "gray40") +
  theme_void() +
  labs(title = "The Bootstrap Algorithm") +
  coord_cartesian(xlim = c(0.3, 5.7), ylim = c(0.5, 1.5))
```

. . .

**Use bootstrap when:** no analytical formula for SE, complex statistics (medians, ratios, correlations), or data are skewed.

**Bootstrap t intervals** have better coverage for skewed data than percentile intervals (second-order accuracy). **Percentile intervals** are preferred when the bootstrap distribution is discrete or violates normality assumptions.


# III. Hypothesis Testing (Lessons 12–17) {background-color="#e74c3c"}

## The Hypothesis Testing Framework

**Setup:**

- **Null hypothesis** $H_0$: the "status quo" or "no effect" claim
- **Alternative hypothesis** $H_a$: what we want to show evidence for
- **Test statistic** $T$: a summary of the data relevant to distinguishing $H_0$ from $H_a$
- **Decision rule**: reject $H_0$ if $T$ falls in the **rejection region**

. . .

**Two types of error:**

| | $H_0$ True | $H_0$ False |
|---|---|---|
| **Reject $H_0$** | Type I Error ($\alpha$) | Correct (Power = $1-\beta$) |
| **Fail to reject $H_0$** | Correct | Type II Error ($\beta$) |

. . .

::: callout-warning
## Key Distinction
- $\alpha$ = P(reject $H_0$ | $H_0$ true) — **significance level**, set in advance
- $\beta$ = P(fail to reject $H_0$ | $H_0$ false) — depends on true parameter value
- **Power** = $1 - \beta$ = P(reject $H_0$ | $H_0$ false) — ability to detect a real effect
:::

## Common One-Sample Tests

| Parameter | Hypotheses | Test Statistic | Distribution under $H_0$ |
|-----------|-----------|----------------|--------------------------|
| $\mu$ ($\sigma$ known) | $H_0: \mu = \mu_0$ | $Z = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}}$ | $N(0,1)$ |
| $\mu$ ($\sigma$ unknown) | $H_0: \mu = \mu_0$ | $T = \frac{\bar{X} - \mu_0}{S/\sqrt{n}}$ | $t_{n-1}$ |
| $p$ (proportion) | $H_0: p = p_0$ | $Z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$ | $N(0,1)$ |

. . .

**P-value:** The probability, under $H_0$, of observing a test statistic **as extreme or more extreme** than the observed value.

- Small p-value → strong evidence against $H_0$
- P-value is **not** the probability that $H_0$ is true

## P-values, Power, and Sample Size

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 16
# Visualize the relationship between alpha, beta, and power
mu_0 <- 0
mu_a <- 1.5
sigma <- 2
n <- 16
se <- sigma / sqrt(n)
alpha <- 0.05
z_crit <- qnorm(1 - alpha/2)
crit_val <- mu_0 + z_crit * se

x_seq <- seq(-2, 4, length.out = 500)
df_null <- tibble(x = x_seq, density = dnorm(x_seq, mu_0, se), dist = "Null")
df_alt  <- tibble(x = x_seq, density = dnorm(x_seq, mu_a, se), dist = "Alternative")

ggplot() +
  # Null distribution
  geom_area(data = df_null |> filter(x >= crit_val),
            aes(x = x, y = density), fill = "#e74c3c", alpha = 0.4) +
  geom_line(data = df_null, aes(x = x, y = density), 
            color = "#2c3e50", linewidth = 1.2) +
  # Alternative distribution
  geom_area(data = df_alt |> filter(x >= crit_val),
            aes(x = x, y = density), fill = "#27ae60", alpha = 0.4) +
  geom_area(data = df_alt |> filter(x < crit_val),
            aes(x = x, y = density), fill = "#f39c12", alpha = 0.3) +
  geom_line(data = df_alt, aes(x = x, y = density), 
            color = "#27ae60", linewidth = 1.2) +
  # Critical value
  geom_vline(xintercept = crit_val, linetype = "dashed", linewidth = 1) +
  # Labels
  annotate("text", x = mu_0, y = max(df_null$density) + 0.02, 
           label = expression(H[0]: ~ mu == mu[0]), size = 6, color = "#2c3e50") +
  annotate("text", x = mu_a, y = max(df_alt$density) + 0.02,
           label = expression(H[a]: ~ mu == mu[a]), size = 6, color = "#27ae60") +
  annotate("text", x = crit_val + 0.45, y = 0.15,
           label = expression(alpha ~ "(Type I)"), size = 5, color = "#e74c3c") +
  annotate("text", x = crit_val - 0.55, y = 0.1,
           label = expression(beta ~ "(Type II)"), size = 5, color = "#f39c12") +
  annotate("text", x = mu_a + 0.5, y = 0.25,
           label = expression("Power" == 1 - beta), size = 5, color = "#27ae60") +
  labs(x = expression(bar(X)), y = "Density",
       title = "Hypothesis Testing: Type I Error, Type II Error, and Power",
       subtitle = expression(paste("One-sided test: ", H[0]: mu == 0, " vs ", H[a]: mu == 1.5, 
                                   ", n = 16, ", sigma, " = 2"))) +
  theme(legend.position = "none")
```

## Sample Size Determination

To achieve power $1 - \beta$ at significance level $\alpha$ for detecting a difference $\delta = \mu_a - \mu_0$:

$$n = \left(\frac{(z_{\alpha/2} + z_\beta) \cdot \sigma}{\delta}\right)^2 \quad \text{(two-sided z-test)}$$

. . .

**Medical Example:** A trial needs 80% power ($\beta = 0.20$) to detect a 3 mmHg reduction in systolic blood pressure ($\sigma = 12$ mmHg) at $\alpha = 0.05$:

```{r}
#| echo: true
delta <- 3     # effect size (mmHg)
sigma <- 12    # population SD
alpha <- 0.05
beta <- 0.20

n_required <- ((qnorm(1 - alpha/2) + qnorm(1 - beta)) * sigma / delta)^2
cat("Required sample size per group:", ceiling(n_required))
```


## Neyman-Pearson Lemma

::: callout-important
## Neyman-Pearson Lemma (Most Powerful Tests)
For testing $H_0: \theta = \theta_0$ vs $H_a: \theta = \theta_1$ (simple vs. simple), the **most powerful** $\alpha$-level test rejects $H_0$ when:

$$\Lambda = \frac{L(\theta_1 | \mathbf{x})}{L(\theta_0 | \mathbf{x})} > k$$

where $k$ is chosen so that $P(\text{reject} | H_0) = \alpha$.
:::

. . .

**Uniformly Most Powerful (UMP) tests** exist for one-sided alternatives in **monotone likelihood ratio** families (e.g., exponential family with one parameter).

For **composite** hypotheses, we use the **Likelihood Ratio Test (LRT)**:

$$\Lambda = \frac{\sup_{\theta \in \Theta_0} L(\theta|\mathbf{x})}{\sup_{\theta \in \Theta} L(\theta|\mathbf{x})} \quad \text{Reject } H_0 \text{ when } -2\ln\Lambda > \chi^2_{\alpha, r}$$

where $r$ = number of restrictions imposed by $H_0$.

## Connection: Testing and Confidence Intervals

::: callout-tip
## The Duality Principle
A $100(1-\alpha)\%$ confidence interval for $\theta$ contains exactly those values $\theta_0$ for which a two-sided $\alpha$-level test would **fail to reject** $H_0: \theta = \theta_0$.

$$\theta_0 \in \text{CI} \iff \text{fail to reject } H_0: \theta = \theta_0$$
:::

. . .

This means:

- **Wald CI** ↔ **Wald test** (z-test or t-test)
- **Score CI** (Wilson) ↔ **Score test**
- **LRT CI** ↔ **Likelihood ratio test**

The CI tells you the same thing as the test — and gives you **more** information (the range of plausible values, not just reject/fail to reject).


# IV. Two-Sample & Paired Inference (Lessons 18–19) {background-color="#f39c12"}

## Two-Sample Methods

**Goal:** Compare parameters from two independent populations.

| Setting | Test Statistic | CI Formula |
|---------|---------------|------------|
| Means, $\sigma_1, \sigma_2$ known | $Z = \frac{\bar{X}_1 - \bar{X}_2 - \delta_0}{\sqrt{\sigma_1^2/n_1 + \sigma_2^2/n_2}}$ | $(\bar{X}_1 - \bar{X}_2) \pm z_{\alpha/2}\sqrt{\frac{\sigma_1^2}{n_1} + \frac{\sigma_2^2}{n_2}}$ |
| Means, $\sigma$ unknown | Welch t-test with $\hat{\nu}$ df | $(\bar{X}_1 - \bar{X}_2) \pm t_{\alpha/2, \hat{\nu}} \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$ |
| Proportions | $Z = \frac{\hat{p}_1 - \hat{p}_2}{\sqrt{\hat{p}(1-\hat{p})(1/n_1 + 1/n_2)}}$ | $(\hat{p}_1 - \hat{p}_2) \pm z_{\alpha/2}\sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$ |

. . .

::: callout-note
## Welch Degrees of Freedom
$$\hat{\nu} = \frac{\left(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}\right)^2}{\frac{(s_1^2/n_1)^2}{n_1 - 1} + \frac{(s_2^2/n_2)^2}{n_2 - 1}}$$
Use `t.test()` in R — it uses Welch by default (no equal-variance assumption needed).
:::

## Paired Data

**When observations are naturally paired** (before/after, matched subjects, same patient):

$$D_i = X_{1i} - X_{2i}, \quad i = 1, \ldots, n$$

Reduces to a **one-sample problem** on the differences $D_1, \ldots, D_n$.

. . .

**Paired t-test:**
$$T = \frac{\bar{D} - \delta_0}{S_D / \sqrt{n}} \sim t_{n-1} \quad \text{under } H_0$$

. . .

**Medical Example:** Blood pressure measured before and after treatment in 20 patients — pairing controls for patient-level variation (baseline health, age, genetics) that would add noise in an independent two-sample design.

```{r}
#| echo: true
# Simulated paired blood pressure data
set.seed(551)
n <- 20
bp_before <- rnorm(n, mean = 145, sd = 12)
bp_after <- bp_before + rnorm(n, mean = -5, sd = 6)  # treatment effect ~ -5 mmHg

t.test(bp_before, bp_after, paired = TRUE) |> tidy() |>
  select(estimate, statistic, p.value, conf.low, conf.high) |>
  mutate(across(everything(), ~round(.x, 3))) |>
  knitr::kable()
```


# V. Bootstrap & Permutation Methods (Lessons 17, 20) {background-color="#27ae60"}

## Bootstrap Hypothesis Testing

**Idea:** Under $H_0$, the null is true — create a bootstrap world that reflects this.

. . .

**One-sample bootstrap test** for $H_0: \mu = \mu_0$:

1. **Shift** the data: $X_i^{\text{shifted}} = X_i - \bar{X} + \mu_0$ (centers at $\mu_0$)
2. Resample from shifted data, compute $\bar{X}^*$ (or $t^*$) each time
3. P-value = proportion of bootstrap statistics as extreme as observed

. . .

**Two-sample permutation test** for $H_0: F_1 = F_2$ (same distribution):

1. Pool both samples
2. Randomly assign labels to groups (permute)
3. Compute test statistic under permutation
4. P-value = proportion of permutation statistics as extreme as observed

::: callout-warning
## Important `infer` Package Note
`generate(type = "bootstrap")` with `hypothesize(null = "independence")` does **not** implement the shift method for two-sample tests. Use `type = "permute"` for two-sample null hypothesis testing. The shift-bootstrap must be coded manually.
:::

## Bootstrap vs. Permutation: When to Use Each

| | Bootstrap | Permutation |
|---|---|---|
| **Tests** | One-sample (via shift), CIs | Two-sample equality |
| **$H_0$** | Specific parameter value | $F_1 = F_2$ (exchangeability) |
| **Resampling** | With replacement from (shifted) sample | Without replacement across pooled sample |
| **CIs** | Yes (percentile, t, studentized) | Not directly (test only) |
| **Assumption** | Data representative of population | Exchangeability under $H_0$ |


# VI. Nonparametric Methods (Lesson 21) {background-color="#8e44ad"}

## When Parametric Assumptions Fail

**Nonparametric tests** do not require a specific distributional form (e.g., normality).

| Test | Setting | $H_0$ | Based On |
|------|---------|--------|----------|
| **Sign test** | One-sample | $\text{median} = m_0$ | Count of $X_i > m_0$ |
| **Wilcoxon signed-rank** | One-sample / paired | Symmetric about $m_0$ | Signed ranks of $|X_i - m_0|$ |
| **Wilcoxon rank-sum** | Two independent samples | $F_1 = F_2$ | Ranks of combined sample |

. . .

**Asymptotic relative efficiency (ARE):**

- Sign test vs. t-test (normal data): ARE $= 2/\pi \approx 0.637$ — sign test needs ~57% more data
- Wilcoxon signed-rank vs. t-test (normal data): ARE $= 3/\pi \approx 0.955$ — nearly as efficient
- Wilcoxon signed-rank is **more** efficient than t for heavy-tailed distributions

## Nonparametric Example: Pain Relief Scores

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5
#| fig-width: 14
# Skewed pain relief data — nonparametric appropriate
set.seed(551)
pain_scores <- c(2.1, 3.5, 1.8, 4.2, 2.9, 6.1, 3.3, 2.5, 8.4, 3.0,
                 1.2, 5.7, 2.8, 3.6, 4.8)

# Null: median pain relief = 3.0 (clinically meaningful threshold)
m0 <- 3.0

# Sign test
n_above <- sum(pain_scores > m0)
n_below <- sum(pain_scores < m0)
n_nonzero <- n_above + n_below
sign_p <- 2 * pbinom(min(n_above, n_below), n_nonzero, 0.5)

# Wilcoxon signed-rank test
wsr <- wilcox.test(pain_scores, mu = m0, conf.int = TRUE)

# Visualization
p1 <- tibble(score = pain_scores) |>
  ggplot(aes(x = score)) +
  geom_histogram(bins = 8, fill = "#9b59b6", alpha = 0.7, color = "white") +
  geom_vline(xintercept = m0, linetype = "dashed", linewidth = 1.2, color = "red") +
  annotate("text", x = m0 + 0.3, y = 3.5, label = expression(m[0] == 3.0),
           size = 5, color = "red") +
  labs(x = "Pain Relief Score", y = "Count", title = "Distribution of Pain Scores") +
  theme_minimal(base_size = 16)

p2 <- tibble(
  Test = c("Sign Test", "Wilcoxon Signed-Rank", "t-test"),
  `P-value` = c(sign_p, wsr$p.value, t.test(pain_scores, mu = m0)$p.value)
) |>
  ggplot(aes(x = Test, y = `P-value`)) +
  geom_col(fill = c("#e74c3c", "#9b59b6", "#3498db"), alpha = 0.8, width = 0.6) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "gray40") +
  annotate("text", x = 3.4, y = 0.06, label = expression(alpha == 0.05), size = 5) +
  labs(y = "P-value", x = "", title = "Comparing Test Results") +
  theme_minimal(base_size = 16)

p1 + p2
```


# VII. Extensions (Lessons 22–23) {background-color="#9b59b6"}

## Gauss-Markov and OLS (Lesson 22)

::: callout-important
## Gauss-Markov Theorem
Under the linear model $\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}$ with:

1. $E[\boldsymbol{\varepsilon}] = \mathbf{0}$ (errors have mean zero)
2. $\text{Cov}(\boldsymbol{\varepsilon}) = \sigma^2 \mathbf{I}$ (constant variance, uncorrelated errors)

The OLS estimator $\hat{\boldsymbol{\beta}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y}$ is **BLUE**: Best (minimum variance) Linear Unbiased Estimator.
:::

. . .

**Connection to MVUE theory:**

- BLUE is the regression analog of MVUE — but restricted to **linear** estimators
- Adding normality → OLS estimator is MVUE (not just BLUE) and is the MLE
- Hypothesis tests for coefficients: $T = \frac{\hat{\beta}_j}{SE(\hat{\beta}_j)} \sim t_{n-p}$

## Bayesian Inference (Lesson 23)

**The Bayesian approach** treats $\theta$ as a random variable:

$$\underbrace{\pi(\theta | \mathbf{x})}_{\text{Posterior}} \propto \underbrace{L(\theta | \mathbf{x})}_{\text{Likelihood}} \times \underbrace{\pi(\theta)}_{\text{Prior}}$$

. . .

**Key conjugate families:**

| Likelihood | Prior | Posterior |
|-----------|-------|-----------|
| Binomial$(n, p)$ | Beta$(\alpha, \beta)$ | Beta$(\alpha + x, \; \beta + n - x)$ |
| Poisson$(\lambda)$ | Gamma$(\alpha, \beta)$ | Gamma$(\alpha + \sum x_i, \; \beta + n)$ |
| Normal$(\mu, \sigma^2)$ | Normal$(\mu_0, \tau^2)$ | Normal$\left(\frac{\mu_0/\tau^2 + n\bar{x}/\sigma^2}{1/\tau^2 + n/\sigma^2}, \; \frac{1}{1/\tau^2 + n/\sigma^2}\right)$ |

. . .

**Bayesian vs. Frequentist:**

- **Credible interval**: $P(\theta \in [a, b] | \mathbf{x}) = 1-\alpha$ — direct probability statement about $\theta$
- **Confidence interval**: $P(\theta \in [L(X), U(X)]) = 1-\alpha$ — coverage frequency across repeated samples
- As $n \to \infty$, Bayesian posterior concentrates at MLE; credible and confidence intervals converge


# VIII. Comprehensive Example {background-color="#2c3e50"}

## Tying It All Together: Hemoglobin Levels After Treatment

A clinical trial measures hemoglobin (g/dL) in $n = 25$ patients with chronic anemia after 12 weeks of a new erythropoiesis-stimulating agent. The clinically important threshold is $\mu_0 = 11$ g/dL.

```{r}
#| echo: true
set.seed(551)
# Simulated hemoglobin data (slightly skewed)
hgb <- c(rnorm(20, mean = 11.8, sd = 1.5), rlnorm(5, meanlog = log(12), sdlog = 0.15))
n <- length(hgb)

cat("n =", n, "\n")
cat("Mean:", round(mean(hgb), 3), "  SD:", round(sd(hgb), 3), "\n")
cat("Median:", round(median(hgb), 3), "\n")
```

## Step 1: Point Estimation

```{r}
#| echo: true
# MLE for normal mean (same as sample mean)
theta_hat <- mean(hgb)
se_hat <- sd(hgb) / sqrt(n)

cat("Point estimate (MLE):", round(theta_hat, 3), "g/dL\n")
cat("Standard error:", round(se_hat, 3), "g/dL\n")
cat("Fisher Information (estimated): I_n =", round(n / var(hgb), 3), "\n")
cat("CRLB-based SE: sqrt(1/I_n) =", round(sqrt(var(hgb) / n), 3), "g/dL")
```

## Step 2: Confidence Intervals (Multiple Methods)

```{r}
#| echo: true
# 1. t-interval (pivotal)
t_ci <- t.test(hgb)$conf.int

# 2. Bootstrap percentile CI
B <- 10000
boot_means <- replicate(B, mean(sample(hgb, replace = TRUE)))
pct_ci <- quantile(boot_means, c(0.025, 0.975))

# 3. Bootstrap t CI
boot_t_ci <- theta_hat + c(-1, 1) * qt(0.975, n - 1) * sd(boot_means)

tibble(
  Method = c("t-interval", "Bootstrap percentile", "Bootstrap t"),
  Lower = round(c(t_ci[1], pct_ci[1], boot_t_ci[1]), 3),
  Upper = round(c(t_ci[2], pct_ci[2], boot_t_ci[2]), 3),
  Width = round(c(diff(t_ci), diff(pct_ci), diff(boot_t_ci)), 3)
) |>
  knitr::kable()
```

## Step 3: Hypothesis Testing

$H_0: \mu = 11$ vs. $H_a: \mu > 11$ (is mean hemoglobin above threshold?)

```{r}
#| echo: true
# Classical t-test
t_result <- t.test(hgb, mu = 11, alternative = "greater")

# Wilcoxon signed-rank test
w_result <- wilcox.test(hgb, mu = 11, alternative = "greater")

# Bootstrap test (shift method)
hgb_shifted <- hgb - mean(hgb) + 11  # center at null
boot_null <- replicate(B, mean(sample(hgb_shifted, replace = TRUE)))
boot_pval <- mean(boot_null >= theta_hat)

tibble(
  Test = c("One-sample t-test", "Wilcoxon signed-rank", "Bootstrap (shift)"),
  Statistic = round(c(t_result$statistic, w_result$statistic, NA), 3),
  `P-value` = round(c(t_result$p.value, w_result$p.value, boot_pval), 4)
) |>
  knitr::kable()
```

## Step 4: Visualizing the Evidence

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 16
p_data <- tibble(hgb = hgb) |>
  ggplot(aes(x = hgb)) +
  geom_histogram(bins = 12, fill = "#3498db", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 11, color = "red", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean(hgb), color = "#2c3e50", linewidth = 1.2) +
  annotate("text", x = 11, y = Inf, label = expression(mu[0] == 11), 
           vjust = 1.5, color = "red", size = 5) +
  annotate("text", x = mean(hgb), y = Inf, label = paste0("x-bar = ", round(mean(hgb), 2)),
           vjust = 3, color = "#2c3e50", size = 5) +
  labs(x = "Hemoglobin (g/dL)", y = "Count", title = "Observed Data") +
  theme_minimal(base_size = 16)

p_boot <- tibble(stat = boot_null) |>
  ggplot(aes(x = stat)) +
  geom_histogram(bins = 50, fill = "#27ae60", alpha = 0.7, color = "white") +
  geom_vline(xintercept = theta_hat, color = "#e74c3c", linewidth = 1.2) +
  annotate("text", x = theta_hat + 0.05, y = Inf, 
           label = paste0("Observed = ", round(theta_hat, 2)),
           vjust = 1.5, hjust = 0, color = "#e74c3c", size = 5) +
  labs(x = expression(bar(X) * " under " * H[0]), y = "Count",
       title = "Bootstrap Null Distribution",
       subtitle = paste0("P-value = ", round(boot_pval, 4))) +
  theme_minimal(base_size = 16)

p_data + p_boot
```


# IX. Decision Guides {background-color="#2c3e50"}

## Choosing a Method: Master Flowchart

```{r}
#| echo: false
#| fig-height: 10
#| fig-width: 18
grViz("
digraph decision {
  graph [rankdir = TB, bgcolor = transparent]
  node [shape = box, style = filled, fontsize = 20, fontname = 'Helvetica']

  Q1 [label = 'How many samples?', fillcolor = '#ecf0f1']
  
  one [label = 'One Sample', fillcolor = '#d5f5e3']
  two [label = 'Two Samples', fillcolor = '#d6eaf8']
  
  paired [label = 'Paired or\\nIndependent?', fillcolor = '#ecf0f1']
  ind [label = 'Independent', fillcolor = '#d6eaf8']
  pair [label = 'Paired → Reduce\\nto One Sample', fillcolor = '#d5f5e3']
  
  norm1 [label = 'Data approx\\nnormal?', fillcolor = '#ecf0f1']
  norm2 [label = 'Data approx\\nnormal?', fillcolor = '#ecf0f1']
  
  t1 [label = 't-test / t-interval\\n(Lesson 8, 13)', fillcolor = '#85c1e9']
  nonpar1 [label = 'Sign test / Wilcoxon\\nSigned-Rank (21)\\nor Bootstrap (17, 20)', fillcolor = '#82e0aa']
  
  t2 [label = 'Welch t-test\\n(Lesson 18)', fillcolor = '#85c1e9']
  nonpar2 [label = 'Wilcoxon Rank-Sum (21)\\nor Permutation (20)', fillcolor = '#82e0aa']
  
  prop [label = 'Proportion?', fillcolor = '#fdebd0']
  proptest [label = 'z-test / Score CI\\n(Lessons 8, 13)', fillcolor = '#f9e79f']
  prop2test [label = 'Two-proportion\\nz-test (Lesson 18)', fillcolor = '#f9e79f']
  
  Q1 -> one
  Q1 -> two
  two -> paired
  paired -> ind
  paired -> pair
  pair -> norm1
  one -> prop [label = 'Categorical']
  one -> norm1 [label = 'Continuous']
  prop -> proptest
  ind -> norm2 [label = 'Continuous']
  ind -> prop2test [label = 'Categorical']
  norm1 -> t1 [label = 'Yes or n ≥ 30']
  norm1 -> nonpar1 [label = 'No, small n']
  norm2 -> t2 [label = 'Yes or n ≥ 30']
  norm2 -> nonpar2 [label = 'No, small n']
}
")
```

## CI Method Selection Guide

| Situation | Best Method | Why |
|-----------|------------|-----|
| Normal data, $\sigma$ known | z-interval | Exact pivotal quantity |
| Normal data, $\sigma$ unknown | t-interval | Exact pivotal quantity |
| Proportion, large n | Wald or Score | Wald is simpler; Score better near boundaries |
| Proportion, small n or $p$ near 0/1 | Score (Wilson) or Agresti-Coull | Avoids undercoverage |
| General parameter, large n | Wald (MLE-based) | Asymptotic normality of MLE |
| Complex statistic (median, ratio) | Bootstrap | No analytical SE formula |
| Skewed data, $n \geq 15$ | Studentized bootstrap t | Second-order accuracy |
| Very small n (< 10) | Classical t (if normal) | Bootstrap unreliable |

## Hypothesis Test Selection Guide

| Situation | Test | Assumption |
|-----------|------|------------|
| One mean, normal/large n | One-sample t | Normality or CLT |
| One mean, non-normal, small n | Wilcoxon signed-rank or bootstrap | Symmetry (Wilcoxon) |
| One median | Sign test | Minimal (none beyond iid) |
| Two means, independent | Welch t | Normality or CLT |
| Two means, paired | Paired t | Differences ~ normal |
| Two distributions equal | Permutation test | Exchangeability under $H_0$ |
| Two independent, non-normal | Wilcoxon rank-sum | Continuous distributions |
| Optimal test (simple $H_0$ vs $H_1$) | Neyman-Pearson (LR) | Parametric model specified |
| Composite $H_0$ | Likelihood ratio test | $-2\ln\Lambda \sim \chi^2$ (large n) |


# X. Key Formulas and Connections {background-color="#2c3e50"}

## The Likelihood Thread

The **likelihood function** $L(\theta|\mathbf{x})$ connects nearly every topic in this course:

```{r}
#| echo: false
#| fig-height: 8
#| fig-width: 16
grViz("
digraph likelihood {
  graph [rankdir = LR, bgcolor = transparent]
  node [shape = box, style = filled, fontsize = 22, fontname = 'Helvetica']
  
  L [label = 'Likelihood\\nL(θ|x)', fillcolor = '#f5b7b1', shape = ellipse, fontsize = 26]
  
  mle [label = 'MLE\\nmaximize L(θ|x)', fillcolor = '#85c1e9']
  fi [label = 'Fisher Information\\nI(θ) = -E[ℓ′′(θ)]', fillcolor = '#85c1e9']
  suff [label = 'Sufficiency\\nFactorization', fillcolor = '#85c1e9']
  
  wald [label = 'Wald CI & Test\\nθ̂ ± z·SE', fillcolor = '#82e0aa']
  score [label = 'Score CI & Test', fillcolor = '#82e0aa']
  lrt [label = 'Likelihood\\nRatio Test', fillcolor = '#82e0aa']
  
  np [label = 'Neyman-Pearson\\nLemma (LR > k)', fillcolor = '#f9e79f']
  bayes [label = 'Bayesian Posterior\\nπ(θ|x) ∝ L·π(θ)', fillcolor = '#d2b4de']
  
  L -> mle
  L -> fi
  L -> suff
  L -> lrt
  L -> np
  L -> bayes
  fi -> wald
  fi -> score
  mle -> wald
}
")
```

## Essential Formulas Reference

::: columns
::: {.column width="48%"}
**Point Estimation:**

- MLE: $\hat{\theta} = \arg\max \ell(\theta|\mathbf{x})$
- Fisher Info: $I_1(\theta) = E[-\ell''(\theta)]$
- CRLB: $\text{Var}(\hat{\theta}) \geq 1/I_n(\theta)$
- MSE: $\text{MSE} = \text{Var} + \text{Bias}^2$

**Confidence Intervals:**

- Wald: $\hat{\theta} \pm z_{\alpha/2}/\sqrt{I_n(\hat{\theta})}$
- t-interval: $\bar{X} \pm t_{\alpha/2, n-1} \cdot S/\sqrt{n}$
- Delta method: $SE[g(\hat{\theta})] \approx |g'(\hat{\theta})| \cdot SE(\hat{\theta})$
:::

::: {.column width="48%"}
**Hypothesis Testing:**

- z-statistic: $Z = (\bar{X} - \mu_0)/(\sigma/\sqrt{n})$
- t-statistic: $T = (\bar{X} - \mu_0)/(S/\sqrt{n})$
- LRT: $-2\ln\Lambda \stackrel{d}{\to} \chi^2_r$
- Sample size: $n = \left(\frac{(z_{\alpha/2} + z_\beta)\sigma}{\delta}\right)^2$

**Bayesian:**

- Posterior: $\pi(\theta|\mathbf{x}) \propto L(\theta|\mathbf{x}) \cdot \pi(\theta)$
- Posterior mean as Bayes estimator (squared error loss)
:::
:::


## Common Mistakes to Avoid

::: callout-warning
## Interpretation Errors
1. **P-value $\neq$ P($H_0$ true)** — it is P(data this extreme | $H_0$ true)
2. **CI does not mean** "95% probability $\theta$ is in the interval" — it's about the procedure's coverage rate
3. **"Fail to reject" $\neq$ "accept $H_0$"** — absence of evidence is not evidence of absence
4. **Statistical significance $\neq$ practical significance** — a tiny effect can be "significant" with large n
:::

. . .

::: callout-warning
## Technical Errors
5. Using pooled variance t-test when variances differ — use **Welch** (default in R)
6. Ignoring pairing — always use paired test when data are paired
7. Using Wald CI for proportions near 0 or 1 — use **Score/Wilson** or **Agresti-Coull**
8. Bootstrap with $n < 10$ — bootstrap distribution is a poor approximation
9. Applying Wilcoxon signed-rank without symmetry — use the **sign test** instead
:::


# Final Exam Preparation {background-color="#006a4e"}

## What to Expect

::: callout-note
## Final Exam Coverage
The final exam is **comprehensive**, covering Lessons 1–23. Emphasis is on:

- **Conceptual understanding**: What do the methods assume? When do they apply?
- **Mathematical derivations**: MLEs, Fisher information, sufficiency, CRLB, LRT
- **Choosing appropriate methods**: Given a scenario, identify the right analysis
- **Interpretation**: What do results mean in the context of the problem?
- **R output**: Read and interpret output from `t.test()`, `wilcox.test()`, bootstrap code
:::

. . .

**Study strategy:**

- Review the three review lessons (6, 11, and this one)
- Rework homework problems, especially derivations
- Practice choosing methods for new medical scenarios
- Review R code and output interpretation — know what each piece means

## Practice Problems

**Problem 1 (Point Estimation):** Let $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Gamma}(\alpha, \beta)$ with $\alpha$ known.

(a) Find the MLE of $\beta$.

(b) Find $I_1(\beta)$ and determine whether the MLE attains the CRLB.

(c) Is $T = \sum X_i$ sufficient for $\beta$? Prove using the factorization theorem.

. . .

**Problem 2 (CI and Testing):** In a clinical trial, 45 out of 200 patients experience symptom relief (vs. historical rate of 30%).

(a) Construct a 95% Wald CI and a 95% Score (Wilson) CI for the true relief rate.

(b) Test $H_0: p = 0.30$ vs $H_a: p \neq 0.30$. State conclusions at $\alpha = 0.05$.

(c) Explain why the Score CI is preferred here.


## Practice Problems (continued)

**Problem 3 (Two-sample and Nonparametric):** Researchers compare recovery times (days) for two surgical techniques. Group A ($n_1 = 15$): $\bar{x}_1 = 8.2$, $s_1 = 3.1$. Group B ($n_2 = 12$): $\bar{x}_2 = 6.5$, $s_2 = 2.4$.

(a) Perform a Welch t-test for $H_0: \mu_1 - \mu_2 = 0$ vs $H_a: \mu_1 - \mu_2 \neq 0$.

(b) If data were highly skewed, which nonparametric test would you use instead and why?

(c) Describe how you would perform a permutation test for this comparison.

. . .

**Problem 4 (Connections):** 

(a) Explain the duality between a 95% CI for $\mu$ and a two-sided $\alpha = 0.05$ t-test.

(b) For testing $H_0: \lambda = \lambda_0$ vs $H_a: \lambda = \lambda_1$ (where $\lambda_1 > \lambda_0$) in a Poisson model, use the Neyman-Pearson lemma to derive the most powerful test.

(c) Describe how the posterior distribution in Bayesian inference relates to the MLE as $n \to \infty$.

## Thank You! {.center background-color="#006a4e"}

**Good luck on the final exam!**

The exam is due **March 19 at 11pm.**

Office hours and review sessions will be available — check the course website for times.

## References

::: nonincremental
- Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapters 6–10, 15.
- Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapters 3, 5, 7, 8, 10.
:::
