---
title: "Lesson 23: Introduction to Bayesian Inference"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w10d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 23 - Introduction to Bayesian Inference
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 23: Introduction to Bayesian Inference {background-color="#2c3e50"}

## Review: The Frequentist Framework

Everything we have built this quarter rests on a single foundation:

- **Parameters** ($\theta$) are **fixed, unknown constants**
- **Data** ($X_1, \ldots, X_n$) are random
- Inference is based on the **sampling distribution** of estimators and test statistics

. . .

What we have accomplished:

| Framework | Tool | Key Idea |
|-----------|------|----------|
| Point estimation | MLE, MOM, MVUE | $\hat\theta$ chosen to summarize the data |
| Interval estimation | CI, bootstrap CI | Procedure covers $\theta$ with prob $1-\alpha$ |
| Hypothesis testing | NP, LRT, permutation | Control Type I error; reject $H_0$ if evidence extreme |

. . .

::: callout-note
## A Different Question
What if we treated $\theta$ as **random** — capturing our *uncertainty* about its value — and updated our beliefs when we observe data?

This is the **Bayesian approach**, and it underpins a large portion of modern biostatistical and machine learning methodology.
:::

## Today's Goals

1. Understand the **philosophical distinction** between Bayesian and frequentist inference
2. Formulate Bayes' theorem for continuous parameters: **prior → posterior**
3. Identify **conjugate prior families** for common likelihoods (Beta–Binomial, Normal–Normal, Gamma–Poisson)
4. Compute **credible intervals** and contrast them with confidence intervals
5. Derive **Bayes estimators** as minimizers of posterior expected loss under squared and absolute error
6. See these ideas applied in medical research contexts

## Roadmap for Today

| Topic | Key Idea |
|-------|----------|
| Bayesian philosophy | $\theta$ is random; prior encodes uncertainty before data |
| Bayes' theorem | $\pi(\theta \mid \mathbf{x}) \propto L(\theta) \cdot \pi(\theta)$ |
| Conjugate priors | Posterior is in the same family as the prior |
| Beta–Binomial | Disease prevalence from a clinical study |
| Normal–Normal | Blood pressure with informative historical data |
| Credible intervals | HPD and equal-tail intervals; comparison to CIs |
| Bayes estimators | Posterior mean minimizes squared error loss |
| Frequentist vs. Bayes | Key philosophical and practical contrasts |

# The Bayesian Philosophy {background-color="#3498db"}

## Parameters as Random Variables

In the frequentist framework, $\theta$ is a fixed constant. We ask:

> "What would happen if we repeated this experiment many times?"

. . .

In the **Bayesian framework**, $\theta$ is treated as a **random variable** with its own distribution. We ask:

> "Given the data I actually observed, what do I now believe about $\theta$?"

. . .

::: callout-important
## The Bayesian Ingredients
Three quantities determine everything:

1. **Prior distribution** $\pi(\theta)$: encodes beliefs about $\theta$ *before* observing data
2. **Likelihood** $L(\theta \mid \mathbf{x}) = f(\mathbf{x} \mid \theta)$: the probability of the data given $\theta$
3. **Posterior distribution** $\pi(\theta \mid \mathbf{x})$: updated beliefs about $\theta$ *after* observing data
:::

. . .

All Bayesian inference flows from the **posterior distribution**.

## Bayes' Theorem for Parameters

For a **continuous** parameter $\theta$ and observed data $\mathbf{x} = (x_1, \ldots, x_n)$:

$$\boxed{\pi(\theta \mid \mathbf{x}) = \frac{f(\mathbf{x} \mid \theta) \cdot \pi(\theta)}{f(\mathbf{x})} = \frac{f(\mathbf{x} \mid \theta) \cdot \pi(\theta)}{\int f(\mathbf{x} \mid \theta') \, \pi(\theta') \, d\theta'}}$$

. . .

Since the **marginal likelihood** $f(\mathbf{x}) = \int f(\mathbf{x} \mid \theta') \pi(\theta') d\theta'$ does not depend on $\theta$, we often write:

$$\pi(\theta \mid \mathbf{x}) \propto f(\mathbf{x} \mid \theta) \cdot \pi(\theta) = L(\theta) \cdot \pi(\theta)$$

. . .

In words:

$$\underbrace{\text{Posterior}}_{\text{updated beliefs}} \propto \underbrace{\text{Likelihood}}_{\text{evidence from data}} \times \underbrace{\text{Prior}}_{\text{initial beliefs}}$$

## Medical Motivation: Estimating Disease Prevalence

::: callout-tip
## Running Example
A public health researcher wants to estimate $p$, the true prevalence of a rare genetic condition in a population. A clinical study observes $X$ cases among $n$ patients.

- **Frequentist:** $\hat{p} = X/n$ with CI $\hat{p} \pm 1.96\sqrt{\hat{p}(1-\hat{p})/n}$
- **Bayesian:** Start with a prior $\pi(p)$ based on historical knowledge, observe data, and compute $\pi(p \mid X)$
:::

. . .

Why might Bayesian methods be attractive here?

- If the condition is **rare** ($n$ is small, $X$ may be 0 or 1), the MLE is unstable
- We often have **historical data** or **domain expertise** that should inform the estimate
- Regulatory agencies (e.g., FDA) increasingly accept Bayesian designs for clinical trials
- Bayesian estimates naturally **shrink** toward prior beliefs, reducing variance

# Prior Distributions {background-color="#8e44ad"}

## Choosing a Prior

The prior $\pi(\theta)$ encodes our uncertainty about $\theta$ before data are collected. Three common choices:

. . .

**Informative priors**: reflect genuine prior knowledge
$$\text{e.g., } p \sim \text{Beta}(3, 17) \text{ if historical data suggest prevalence} \approx 15\%$$

. . .

**Weakly informative priors**: constrain parameters to plausible ranges without dominating the data
$$\text{e.g., } p \sim \text{Beta}(1.5, 1.5) \text{ — slightly favors interior values}$$

. . .

**Non-informative (flat) priors**: minimal prior assumptions; let data dominate

$$\text{e.g., } p \sim \text{Beta}(1, 1) = \text{Uniform}(0,1) \text{ — all values of } p \text{ equally likely}$$

. . .

::: callout-note
## Jeffreys Prior
A principled non-informative prior is the **Jeffreys prior**: $\pi(\theta) \propto \sqrt{I(\theta)}$, where $I(\theta)$ is the Fisher information. For a Binomial proportion, the Jeffreys prior is $\text{Beta}(1/2, 1/2)$.
:::

## Visualizing Prior Beliefs

```{r}
#| label: fig-priors
#| fig-cap: "Different Beta priors for a disease prevalence parameter p"
#| code-fold: true
#| fig-height: 5

tibble(
  p = seq(0.001, 0.999, length.out = 500)
) |>
  mutate(
    `Beta(1,1) — Uniform`       = dbeta(p, 1, 1),
    `Beta(0.5,0.5) — Jeffreys`  = dbeta(p, 0.5, 0.5),
    `Beta(3,17) — Informative`  = dbeta(p, 3, 17),
    `Beta(1.5,1.5) — Weakly inf` = dbeta(p, 1.5, 1.5)
  ) |>
  pivot_longer(-p, names_to = "Prior", values_to = "density") |>
  ggplot(aes(x = p, y = density, color = Prior)) +
  geom_line(linewidth = 1.1) +
  labs(
    x = expression(italic(p)),
    y = "Prior density",
    title = expression("Prior distributions for disease prevalence " * italic(p)),
    color = NULL
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

# Conjugate Prior Families {background-color="#e74c3c"}

## What is a Conjugate Prior?

::: callout-important
## Definition: Conjugate Prior
A prior $\pi(\theta)$ is **conjugate** to the likelihood $f(\mathbf{x} \mid \theta)$ if the posterior $\pi(\theta \mid \mathbf{x})$ belongs to the **same parametric family** as the prior.
:::

. . .

Conjugate priors are computationally convenient:

- Posterior has a **closed-form** expression
- Posterior parameters have intuitive **data-update** interpretations
- Sequential Bayesian updating is trivial: the posterior after $n$ observations becomes the prior for observation $n+1$

. . .

We will study three conjugate families:

| Likelihood | Conjugate Prior | Posterior |
|------------|-----------------|-----------|
| Binomial$(n, p)$ | Beta$(\alpha, \beta)$ | Beta$(\alpha + x,\; \beta + n - x)$ |
| Normal$(\mu, \sigma^2)$, $\sigma^2$ known | Normal$(\mu_0, \tau_0^2)$ | Normal$(\mu_n, \tau_n^2)$ |
| Poisson$(\lambda)$ | Gamma$(\alpha, \beta)$ | Gamma$(\alpha + \sum x_i,\; \beta + n)$ |

## Beta–Binomial Conjugate Family

**Setup:** $X \mid p \sim \text{Binomial}(n, p)$, prior $p \sim \text{Beta}(\alpha, \beta)$

. . .

**Posterior derivation:**

$$\pi(p \mid x) \propto L(p) \cdot \pi(p) = \binom{n}{x} p^x (1-p)^{n-x} \cdot \frac{p^{\alpha-1}(1-p)^{\beta-1}}{B(\alpha,\beta)}$$

$$\propto p^{x + \alpha - 1}(1-p)^{n-x+\beta-1}$$

. . .

This is the kernel of a $\text{Beta}(x + \alpha,\; n - x + \beta)$ distribution:

$$\boxed{p \mid x \sim \text{Beta}(\alpha + x, \; \beta + n - x)}$$

. . .

::: callout-tip
## Intuition
The prior parameters $(\alpha, \beta)$ act like **pseudo-counts**: $\alpha - 1$ prior successes and $\beta - 1$ prior failures. Observing $x$ successes and $n-x$ failures simply adds to these counts.
:::

## Beta–Binomial: Prior-to-Posterior Updating

**Example:** Estimating prevalence of a rare genetic condition.

- Historical data suggests prevalence $\approx 15\%$, with uncertainty $\Rightarrow$ Prior: $p \sim \text{Beta}(3, 17)$
- New study: $n = 50$ patients, $x = 10$ test positive

```{r}
#| label: fig-beta-binomial
#| fig-cap: "Prior-to-posterior updating for disease prevalence (Beta-Binomial conjugate)"
#| code-fold: true
#| fig-height: 5

# Prior parameters
alpha0 <- 3; beta0 <- 17

# Data
n_obs <- 50; x_obs <- 10

# Posterior parameters
alpha1 <- alpha0 + x_obs
beta1  <- beta0 + n_obs - x_obs

# MLE
mle <- x_obs / n_obs

tibble(p = seq(0.001, 0.999, length.out = 600)) |>
  mutate(
    Prior     = dbeta(p, alpha0, beta0),
    Posterior = dbeta(p, alpha1, beta1),
    Likelihood = dbeta(p, x_obs + 1, n_obs - x_obs + 1) # proportional to likelihood
  ) |>
  # Normalize likelihood for display
  mutate(Likelihood = Likelihood / max(Likelihood) * max(Posterior)) |>
  pivot_longer(-p, names_to = "Distribution", values_to = "density") |>
  mutate(Distribution = factor(Distribution, levels = c("Prior", "Likelihood", "Posterior"))) |>
  ggplot(aes(x = p, y = density, color = Distribution, linetype = Distribution)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = mle, linetype = "dashed", color = "gray40") +
  annotate("text", x = mle + 0.01, y = 0.5, label = "MLE", hjust = 0, size = 5, color = "gray40") +
  geom_vline(xintercept = alpha1 / (alpha1 + beta1), linetype = "dotted", color = "#e74c3c") +
  annotate("text", x = alpha1/(alpha1+beta1) + 0.01, y = 2, 
           label = "Posterior\nmean", hjust = 0, size = 5, color = "#e74c3c") +
  scale_color_manual(values = c("Prior" = "#3498db", "Likelihood" = "#2ecc71", "Posterior" = "#e74c3c")) +
  scale_linetype_manual(values = c("Prior" = "dashed", "Likelihood" = "dotted", "Posterior" = "solid")) +
  labs(
    x = expression(italic(p)),
    y = "Density (likelihood normalized for display)",
    title = expression("Beta–Binomial updating: prevalence estimation"),
    subtitle = bquote("Prior: Beta("*.(alpha0)*","*.(beta0)*")   Data: x="*.(x_obs)*", n="*.(n_obs)*
                       "   Posterior: Beta("*.(alpha1)*","*.(beta1)*")"),
    color = NULL, linetype = NULL
  ) +
  theme(legend.position = "bottom")
```

## Beta–Binomial: Summary of Estimates

```{r}
#| label: beta-binomial-summary
#| code-fold: true

alpha0 <- 3; beta0 <- 17
n_obs  <- 50; x_obs <- 10
alpha1 <- alpha0 + x_obs
beta1  <- beta0 + n_obs - x_obs

# Posterior summaries
post_mean   <- alpha1 / (alpha1 + beta1)
post_median <- qbeta(0.5, alpha1, beta1)
post_mode   <- (alpha1 - 1) / (alpha1 + beta1 - 2)
mle_est     <- x_obs / n_obs

# 95% credible interval (equal-tail)
ci_lo <- qbeta(0.025, alpha1, beta1)
ci_hi <- qbeta(0.975, alpha1, beta1)

tibble(
  Quantity = c("Prior mean", "MLE (frequentist)", "Posterior mean", "Posterior median",
               "Posterior mode (MAP)", "95% Credible interval"),
  Value = c(
    round(alpha0/(alpha0+beta0), 3),
    round(mle_est, 3),
    round(post_mean, 3),
    round(post_median, 3),
    round(post_mode, 3),
    paste0("[", round(ci_lo, 3), ", ", round(ci_hi, 3), "]")
  )
) |>
  knitr::kable(caption = "Bayesian and frequentist estimates of disease prevalence p")
```

. . .

**Interpretation:**

- Posterior mean $= \frac{\alpha + x}{\alpha + \beta + n} = \frac{13}{63} \approx 0.206$ — a **weighted average** of prior mean ($0.150$) and MLE ($0.200$)
- With small samples, the prior has more influence; as $n \to \infty$, posterior concentrates around the MLE

## Beta–Binomial: Effect of Sample Size

```{r}
#| label: fig-sample-size-effect
#| fig-cap: "As sample size grows, the posterior concentrates around the MLE regardless of prior"
#| code-fold: true
#| fig-height: 5

alpha0 <- 3; beta0 <- 17
true_p <- 0.20  # true prevalence

crossing(
  n_obs = c(10, 50, 200, 1000),
  p = seq(0.001, 0.999, length.out = 500)
) |>
  mutate(
    x_obs = round(true_p * n_obs),
    alpha1 = alpha0 + x_obs,
    beta1  = beta0 + n_obs - x_obs,
    density = dbeta(p, alpha1, beta1),
    label = paste0("n = ", n_obs)
  ) |>
  ggplot(aes(x = p, y = density, color = factor(n_obs))) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = true_p, linetype = "dashed", color = "gray30") +
  annotate("text", x = true_p + 0.01, y = 25, 
           label = expression("True "*italic(p)*" = 0.20"), hjust = 0, size = 5) +
  labs(
    x = expression(italic(p)),
    y = "Posterior density",
    title = expression("Posterior Beta distributions for increasing sample sizes"),
    subtitle = bquote("Prior: Beta("*.(alpha0)*","*.(beta0)*"), true "*italic(p)*" = 0.20"),
    color = "Sample size"
  ) +
  scale_color_brewer(palette = "Reds", direction = 1) +
  theme(legend.position = "right")
```

## Normal–Normal Conjugate Family

**Setup:** $X_1, \ldots, X_n \mid \mu \stackrel{iid}{\sim} N(\mu, \sigma^2)$, $\sigma^2$ **known**

Prior: $\mu \sim N(\mu_0, \tau_0^2)$

. . .

**Posterior derivation** (completing the square in the exponent):

$$\pi(\mu \mid \mathbf{x}) \propto \exp\!\left(-\frac{n(\bar{x} - \mu)^2}{2\sigma^2}\right) \cdot \exp\!\left(-\frac{(\mu - \mu_0)^2}{2\tau_0^2}\right)$$

$$\propto \exp\!\left(-\frac{(\mu - \mu_n)^2}{2\tau_n^2}\right)$$

. . .

where the **posterior parameters** are:

$$\boxed{\frac{1}{\tau_n^2} = \frac{1}{\tau_0^2} + \frac{n}{\sigma^2}, \qquad \mu_n = \frac{\mu_0/\tau_0^2 + n\bar{x}/\sigma^2}{1/\tau_0^2 + n/\sigma^2}}$$

$$\mu \mid \mathbf{x} \sim N(\mu_n, \, \tau_n^2)$$

::: callout-tip
## Precision Weighting
Define **precision** = $1/\text{variance}$. The posterior precision is the **sum** of prior and data precisions. The posterior mean is a **precision-weighted average** of $\mu_0$ and $\bar{x}$.
:::

## Normal–Normal: Blood Pressure Example

::: callout-tip
## Medical Example
A cardiologist wants to estimate mean systolic blood pressure (SBP) for a new patient population.

- Historical data from similar populations: mean SBP $\approx 128$ mmHg, $\sigma_{\text{prior}} = 10$ mmHg
- Known measurement $\sigma = 15$ mmHg
- New study: $n = 30$, $\bar{x} = 135$ mmHg
:::

```{r}
#| label: fig-normal-normal
#| fig-cap: "Normal-Normal conjugate updating for mean systolic blood pressure"
#| code-fold: true
#| fig-height: 5

mu0   <- 128; tau0  <- 10
xbar  <- 135; sigma <- 15; n_obs <- 30

# Posterior parameters
prec0 <- 1/tau0^2
prec_data <- n_obs/sigma^2
prec_n    <- prec0 + prec_data
tau_n     <- sqrt(1/prec_n)
mu_n      <- (mu0 * prec0 + xbar * prec_data) / prec_n

tibble(mu = seq(100, 160, length.out = 600)) |>
  mutate(
    Prior     = dnorm(mu, mu0, tau0),
    Likelihood = dnorm(mu, xbar, sigma/sqrt(n_obs)),
    Posterior = dnorm(mu, mu_n, tau_n)
  ) |>
  pivot_longer(-mu, names_to = "Distribution", values_to = "density") |>
  mutate(Distribution = factor(Distribution, levels = c("Prior", "Likelihood", "Posterior"))) |>
  ggplot(aes(x = mu, y = density, color = Distribution, linetype = Distribution)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = mu_n, linetype = "solid", color = "#e74c3c", alpha = 0.4) +
  geom_vline(xintercept = mu0,  linetype = "dashed", color = "#3498db", alpha = 0.4) +
  geom_vline(xintercept = xbar, linetype = "dotted", color = "#2ecc71", alpha = 0.4) +
  scale_color_manual(values = c("Prior" = "#3498db", "Likelihood" = "#2ecc71", "Posterior" = "#e74c3c")) +
  scale_linetype_manual(values = c("Prior" = "dashed", "Likelihood" = "dotted", "Posterior" = "solid")) +
  labs(
    x = expression(mu ~ "(mmHg)"),
    y = "Density",
    title = expression("Normal–Normal updating: mean systolic blood pressure"),
    subtitle = bquote(
      "Prior: N("*.(mu0)*", "*.(tau0)^2*"),  "*
      bar(x)*" = "*.(xbar)*",  n = "*.(n_obs)*",  "*
      "Posterior: N("*.(round(mu_n,1))*", "*.(round(tau_n^2,2))*")"
    ),
    color = NULL, linetype = NULL
  ) +
  theme(legend.position = "bottom")
```

## Gamma–Poisson Conjugate Family

**Setup:** $X_1, \ldots, X_n \mid \lambda \stackrel{iid}{\sim} \text{Poisson}(\lambda)$

Prior: $\lambda \sim \text{Gamma}(\alpha, \beta)$ with mean $\alpha/\beta$

. . .

**Posterior derivation:**

$$\pi(\lambda \mid \mathbf{x}) \propto \prod_{i=1}^n \frac{e^{-\lambda}\lambda^{x_i}}{x_i!} \cdot \lambda^{\alpha-1} e^{-\beta\lambda}$$

$$\propto \lambda^{\alpha + \sum x_i - 1} e^{-(\beta + n)\lambda}$$

. . .

$$\boxed{\lambda \mid \mathbf{x} \sim \text{Gamma}\!\left(\alpha + \textstyle\sum_{i=1}^n x_i, \;\; \beta + n\right)}$$

. . .

::: callout-tip
## Medical Example
**Adverse event monitoring:** A clinical trial monitors rare adverse events. Prior knowledge suggests $\lambda \approx 0.5$ events/patient-month. Data from $n = 40$ patients: $\sum x_i = 25$ events.

$$\text{Prior: Gamma}(2, 4) \quad\Rightarrow\quad \text{Posterior: Gamma}(2 + 25, 4 + 40) = \text{Gamma}(27, 44)$$

Posterior mean $= 27/44 \approx 0.614$ events/patient-month
:::

# Credible Intervals {background-color="#f39c12"}

## What is a Credible Interval?

::: callout-important
## Definition: Credible Interval
A **$100(1-\alpha)\%$ credible interval** (also called a **Bayesian interval** or **posterior interval**) is a set $C(\mathbf{x})$ such that

$$P(\theta \in C(\mathbf{x}) \mid \mathbf{x}) = 1 - \alpha$$

The probability is with respect to the **posterior distribution** $\pi(\theta \mid \mathbf{x})$.
:::

. . .

This is the interval that students *think* a frequentist CI gives!

. . .

Two common choices:

**1. Equal-tail credible interval:** $\left[Q_{\alpha/2},\; Q_{1-\alpha/2}\right]$ where $Q_p$ is the $p$-th quantile of the posterior

**2. Highest posterior density (HPD) interval:** the shortest interval containing $1-\alpha$ posterior probability

$$C_{\text{HPD}} = \{\theta : \pi(\theta \mid \mathbf{x}) \geq k_\alpha\}$$

where $k_\alpha$ is chosen so that $P(\theta \in C_{\text{HPD}} \mid \mathbf{x}) = 1-\alpha$.

## Comparing Credible Intervals and Confidence Intervals

| Property | Frequentist CI | Bayesian Credible Interval |
|----------|----------------|---------------------------|
| What is random? | The interval (over repeated samples) | The parameter $\theta$ |
| Correct interpretation | "95% of such intervals contain $\theta$" | "95% probability that $\theta$ is in this interval" |
| Depends on prior? | No | Yes |
| Valid for small $n$? | Approximately | Yes (via posterior) |
| Can be asymmetric? | Typically symmetric | Naturally asymmetric for skewed posteriors |

. . .

::: callout-note
## The Interpretation Students Want
A 95% credible interval **literally means**: given the observed data, there is a 95% probability that $\theta$ lies in the interval. This is the statement frequentist CIs **cannot** make.
:::

## Computing Credible Intervals: Beta–Binomial

```{r}
#| label: fig-credible-interval
#| fig-cap: "Equal-tail and HPD 95% credible intervals for disease prevalence"
#| code-fold: true
#| fig-height: 5

alpha1 <- 13; beta1 <- 53

# Equal-tail CI
et_lo <- qbeta(0.025, alpha1, beta1)
et_hi <- qbeta(0.975, alpha1, beta1)

# HPD: search for shortest interval
# Use grid search for HPD
find_hpd <- function(a, b, level = 0.95) {
  grid <- seq(0.0001, 1 - level - 0.0001, length.out = 2000)
  widths <- qbeta(grid + level, a, b) - qbeta(grid, a, b)
  lo_prob <- grid[which.min(widths)]
  c(qbeta(lo_prob, a, b), qbeta(lo_prob + level, a, b))
}
hpd <- find_hpd(alpha1, beta1)

tibble(p = seq(0.001, 0.65, length.out = 600)) |>
  mutate(density = dbeta(p, alpha1, beta1)) |>
  ggplot(aes(x = p, y = density)) +
  # HPD region
  geom_area(
    data = ~ filter(.x, p >= hpd[1] & p <= hpd[2]),
    aes(fill = "HPD 95%"), alpha = 0.3
  ) +
  # Equal-tail region
  geom_ribbon(
    data = ~ filter(.x, p >= et_lo & p <= et_hi),
    aes(ymin = 0, ymax = density, fill = "Equal-tail 95%"), alpha = 0.2
  ) +
  geom_line(linewidth = 1.2, color = "#e74c3c") +
  geom_segment(aes(x = et_lo, xend = et_hi, y = -0.15, yend = -0.15),
               color = "#3498db", linewidth = 2, arrow = arrow(ends = "both", length = unit(0.1, "inches"))) +
  annotate("text", x = (et_lo + et_hi)/2, y = -0.6, 
           label = "Equal-tail 95% CI", hjust = 0.5, size = 4.5, color = "#3498db") +
  geom_segment(aes(x = hpd[1], xend = hpd[2], y = -1.0, yend = -1.0),
               color = "#2ecc71", linewidth = 2, arrow = arrow(ends = "both", length = unit(0.1, "inches"))) +
  annotate("text", x = (hpd[1] + hpd[2])/2, y = -1.45, 
           label = "HPD 95% CI", hjust = 0.5, size = 4.5, color = "#2ecc71") +
  scale_fill_manual(values = c("HPD 95%" = "#2ecc71", "Equal-tail 95%" = "#3498db")) +
  labs(
    x = expression(italic(p)),
    y = "Posterior density",
    title = expression("95% credible intervals: Beta("*alpha[1]*","*beta[1]*") posterior"),
    subtitle = bquote("Posterior: Beta("*.(alpha1)*", "*.(beta1)*
                       ")   ET: ["*.(round(et_lo,3))*", "*.(round(et_hi,3))*
                       "]   HPD: ["*.(round(hpd[1],3))*", "*.(round(hpd[2],3))*"]"),
    fill = NULL
  ) +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(-2, NA))
```

## Sequential Bayesian Updating

One of the most powerful features of the Bayesian framework: **today's posterior becomes tomorrow's prior**.

```{r}
#| label: fig-sequential-updating
#| fig-cap: "Sequential Bayesian updating as more patients are tested"
#| code-fold: true
#| fig-height: 5

# Start with Uniform prior; observe data sequentially
alpha0 <- 1; beta0 <- 1
# Simulated sequence of patients (true p = 0.15)
set.seed(551)
results <- rbinom(60, 1, prob = 0.15)  # 0 = negative, 1 = positive

check_points <- c(0, 5, 15, 30, 60)

plots_data <- map_dfr(check_points, function(k) {
  x_k <- sum(results[seq_len(k)])
  a_k <- alpha0 + x_k
  b_k <- beta0 + k - x_k
  tibble(
    p = seq(0.001, 0.6, length.out = 400),
    density = dbeta(p, a_k, b_k),
    n_obs = k,
    x_obs = x_k,
    post_mean = a_k / (a_k + b_k),
    label = paste0("n = ", k, ", x = ", x_k)
  )
})

ggplot(plots_data, aes(x = p, y = density, color = factor(n_obs))) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = 0.15, linetype = "dashed", color = "gray30") +
  annotate("text", x = 0.155, y = 12, label = expression("True "*italic(p)*" = 0.15"),
           hjust = 0, size = 4.5, color = "gray30") +
  labs(
    x = expression(italic(p)),
    y = "Posterior density",
    title = expression("Sequential updating: Beta(1,1) prior, "*italic(p)*" = 0.15"),
    color = "Patients seen"
  ) +
  scale_color_viridis_d(option = "C", end = 0.9,
                        labels = unique(plots_data$label)) +
  theme(legend.position = "right")
```

# Bayes Estimators and Decision Theory {background-color="#16a085"}

## Point Estimation from the Posterior

The posterior $\pi(\theta \mid \mathbf{x})$ is a full probability distribution. How do we extract a single **point estimate**?

. . .

Answer: it depends on your **loss function**. A **Bayes estimator** minimizes **posterior expected loss**:

$$\hat{\theta}_{\text{Bayes}} = \arg\min_{\delta} \, E[L(\theta, \delta) \mid \mathbf{x}] = \arg\min_{\delta} \int L(\theta, \delta) \, \pi(\theta \mid \mathbf{x}) \, d\theta$$

. . .

::: callout-important
## Three Key Results

| Loss function | $L(\theta, \delta)$ | Bayes estimator |
|---------------|---------------------|-----------------|
| **Squared error** | $(\theta - \delta)^2$ | Posterior **mean** $E[\theta \mid \mathbf{x}]$ |
| **Absolute error** | $|\theta - \delta|$ | Posterior **median** |
| **0–1 loss** | $\mathbf{1}(\theta \neq \delta)$ | Posterior **mode** (MAP) |
:::

## Derivation: Posterior Mean Minimizes Squared Error Loss

We want to find $\delta^*$ that minimizes:

$$E[(\theta - \delta)^2 \mid \mathbf{x}] = \int (\theta - \delta)^2 \pi(\theta \mid \mathbf{x}) \, d\theta$$

. . .

Add and subtract the posterior mean $\mu_{\text{post}} = E[\theta \mid \mathbf{x}]$:

$$E[(\theta - \delta)^2 \mid \mathbf{x}] = E[(\theta - \mu_{\text{post}} + \mu_{\text{post}} - \delta)^2 \mid \mathbf{x}]$$

$$= \underbrace{E[(\theta - \mu_{\text{post}})^2 \mid \mathbf{x}]}_{\text{Var}(\theta \mid \mathbf{x})} + (\mu_{\text{post}} - \delta)^2 + 2(\mu_{\text{post}} - \delta)\underbrace{E[\theta - \mu_{\text{post}} \mid \mathbf{x}]}_{=\, 0}$$

$$= \text{Var}(\theta \mid \mathbf{x}) + (\mu_{\text{post}} - \delta)^2$$

. . .

This is minimized when $\delta^* = \mu_{\text{post}} = E[\theta \mid \mathbf{x}]$. $\qquad \square$

## Bayes Estimator: Beta–Binomial Case

For $p \mid x \sim \text{Beta}(\alpha + x, \, \beta + n - x)$:

$$\hat{p}_{\text{Bayes}} = E[p \mid x] = \frac{\alpha + x}{\alpha + \beta + n}$$

. . .

Let $n_0 = \alpha + \beta$ (the "prior sample size") and $p_0 = \alpha/(\alpha+\beta)$ (the prior mean). Then:

$$\hat{p}_{\text{Bayes}} = \frac{n_0}{n_0 + n} \cdot p_0 + \frac{n}{n_0 + n} \cdot \hat{p}_{\text{MLE}}$$

. . .

This is a **convex combination** (weighted average) of the prior mean and the MLE:

- Weight on prior $= n_0/(n_0 + n)$ — decreases as $n$ grows
- Weight on MLE $= n/(n_0 + n)$ — increases as $n$ grows

. . .

::: callout-note
## Connection to Shrinkage Estimation
This "shrinkage toward the prior" is related to **James–Stein estimation** and **ridge regression** in linear models. Bayesian estimators systematically trade a little bias for reduced variance — often beneficial in finite samples.
:::

## Posterior Expected Loss: Visualization

```{r}
#| label: fig-posterior-loss
#| fig-cap: "Posterior expected loss under squared error for different point estimates"
#| code-fold: true
#| fig-height: 5

alpha1 <- 13; beta1 <- 53
post_mean   <- alpha1 / (alpha1 + beta1)
post_median <- qbeta(0.5, alpha1, beta1)
mle_val     <- 10 / 50  # x=10, n=50

# Compute posterior expected squared error for a range of delta values
delta_grid <- seq(0.05, 0.45, length.out = 500)

# E[(p - delta)^2 | x] = Var(p|x) + (E[p|x] - delta)^2
post_var <- alpha1 * beta1 / ((alpha1 + beta1)^2 * (alpha1 + beta1 + 1))
exp_loss <- post_var + (post_mean - delta_grid)^2

tibble(delta = delta_grid, expected_loss = exp_loss) |>
  ggplot(aes(x = delta, y = expected_loss)) +
  geom_line(linewidth = 1.2, color = "#2c3e50") +
  geom_vline(xintercept = post_mean,   color = "#e74c3c", linetype = "solid",  linewidth = 1) +
  geom_vline(xintercept = post_median, color = "#3498db", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = mle_val,     color = "#2ecc71", linetype = "dotted", linewidth = 1) +
  annotate("text", x = post_mean + 0.005,   y = max(exp_loss)*0.95,
           label = "Post. mean", color = "#e74c3c", hjust = 0, size = 4.5) +
  annotate("text", x = post_median - 0.005, y = max(exp_loss)*0.85,
           label = "Post. median", color = "#3498db", hjust = 1, size = 4.5) +
  annotate("text", x = mle_val + 0.005,     y = max(exp_loss)*0.75,
           label = "MLE", color = "#2ecc71", hjust = 0, size = 4.5) +
  labs(
    x = expression(delta ~ "(point estimate)"),
    y = expression("Posterior expected squared error loss"),
    title = expression("Posterior expected loss: "*E*"[("*theta-delta*")²|"*bold(x)*"]"),
    subtitle = bquote("Posterior: Beta("*.(alpha1)*", "*.(beta1)*")  —  minimum at posterior mean = "*.(round(post_mean,3)))
  )
```

# Frequentist vs. Bayesian: A Direct Comparison {background-color="#2c3e50"}

## Side-by-Side Comparison

|  | Frequentist | Bayesian |
|--|-------------|----------|
| **Nature of $\theta$** | Fixed, unknown constant | Random variable with prior distribution |
| **Source of probability** | Long-run frequency of events | Degree of belief / uncertainty |
| **Inference basis** | Sampling distribution of $\hat{\theta}$ | Posterior distribution $\pi(\theta \mid \mathbf{x})$ |
| **Point estimate** | MLE, MOM, UMVUE | Posterior mean, median, or mode |
| **Interval estimate** | CI: covers $\theta$ with prob $1-\alpha$ over repeated samples | Credible interval: $P(\theta \in C \mid \mathbf{x}) = 1-\alpha$ |
| **Prior needed?** | No | Yes |
| **Small samples** | May require asymptotic approximations | Can be exact if prior + likelihood are tractable |
| **Computation** | Often analytic or standard software | May require MCMC (Stan, JAGS, brms) |

## When Does the Prior Matter Most?

```{r}
#| label: fig-prior-influence
#| fig-cap: "Influence of prior vs. data as a function of sample size"
#| code-fold: true
#| fig-height: 5

# Three priors; true p = 0.30
priors <- tibble(
  name  = c("Strongly wrong prior\nBeta(2,18)", "Uniform\nBeta(1,1)", "Correct prior\nBeta(6,14)"),
  alpha = c(2, 1, 6),
  beta  = c(18, 1, 14)
)
true_p <- 0.30
n_grid <- c(5, 20, 100, 500)

results <- crossing(priors, n_obs = n_grid) |>
  mutate(
    x_obs = round(true_p * n_obs),
    alpha1 = alpha + x_obs,
    beta1  = beta  + n_obs - x_obs,
    post_mean = alpha1 / (alpha1 + beta1)
  )

ggplot(results, aes(x = n_obs, y = post_mean, color = name, group = name)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_hline(yintercept = true_p, linetype = "dashed", color = "gray30") +
  annotate("text", x = 450, y = true_p + 0.015,
           label = expression("True "*italic(p)*" = 0.30"), size = 4.5, color = "gray30") +
  scale_x_log10(breaks = n_grid, labels = n_grid) +
  labs(
    x = "Sample size (log scale)",
    y = "Posterior mean",
    title = "Prior influence diminishes with increasing sample size",
    subtitle = expression("All posteriors converge to true "*italic(p)*" as n grows"),
    color = "Prior"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "right")
```

## Key Philosophical Points

**Subjectivity vs. Objectivity**

- The prior is *subjective* — two analysts with different priors reach different posteriors
- Frequentist methods are also not fully "objective" (choice of test statistic, significance level, etc.)
- With large data, both approaches give similar answers

. . .

**The Bayesian Framework Naturally Handles:**

- **Sequential analyses**: posterior updates naturally as data accumulate (no multiple-testing correction needed for planned sequential designs)
- **Prediction**: posterior predictive distribution $f(\tilde{x} \mid \mathbf{x}) = \int f(\tilde{x} \mid \theta) \pi(\theta \mid \mathbf{x}) d\theta$
- **Complex models**: hierarchical models, latent variable models via MCMC

. . .

::: callout-note
## In Your Career
You will encounter Bayesian methods in:
- **Adaptive clinical trials** (FDA Bayesian guidance documents)
- **Hierarchical / mixed-effects models** (often estimated via MCMC)
- **Biomarker analysis, genomics, imaging** (high-dimensional priors)
- **Machine learning** (Bayesian optimization, Gaussian processes, probabilistic neural networks)
:::

## Frequentist Properties of Bayesian Estimators

::: callout-important
## Bernstein–von Mises Theorem
Under regularity conditions, as $n \to \infty$:

$$\pi(\theta \mid \mathbf{x}) \xrightarrow{d} N\!\left(\hat{\theta}_{\text{MLE}},\; [I_n(\hat\theta)]^{-1}\right)$$

That is, the posterior **concentrates around the MLE** and becomes approximately normal — regardless of the prior.
:::

. . .

**Consequences:**

- For large samples, Bayesian and frequentist estimates agree
- The $95\%$ credible interval is **asymptotically identical** to the $95\%$ Wald CI
- The prior only matters in small samples or with strongly informative priors

. . .

**Admissibility:** Under squared error loss, the Bayes estimator is **admissible** (cannot be uniformly improved). This connects to decision theory — Bayesian procedures often have excellent frequentist risk properties.

# Textbook Examples {background-color="#3498db"}

## Devore §12: Bayesian Estimation (Overview)

Devore–Berk–Carlton introduces Bayesian estimation in Chapter 12. The key formula for a **Beta–Binomial** model with $X \mid p \sim \text{Binomial}(n, p)$ and $p \sim \text{Beta}(\alpha, \beta)$:

- Posterior: $p \mid x \sim \text{Beta}(\alpha + x, \beta + n - x)$
- Bayes estimate (squared error): $\hat{p}_B = \dfrac{\alpha + x}{\alpha + \beta + n}$

. . .

**Example (DBC spirit):** A pharmaceutical company tests a new compound in 20 patients with $x = 7$ responders. Prior: $p \sim \text{Beta}(2, 2)$ (weakly centered at 0.5).

```{r}
#| label: dbc-example
#| code-fold: true

alpha0 <- 2; beta0 <- 2
x <- 7; n <- 20
alpha1 <- alpha0 + x; beta1 <- beta0 + n - x

tibble(
  Estimate = c("Prior mean", "MLE", "Bayes (posterior mean)", "95% Equal-tail CI"),
  Value = c(
    round(alpha0/(alpha0+beta0), 3),
    round(x/n, 3),
    round(alpha1/(alpha1+beta1), 3),
    paste0("[", round(qbeta(0.025, alpha1, beta1), 3), ", ", round(qbeta(0.975, alpha1, beta1), 3), "]")
  )
) |> knitr::kable()
```

## Chihara & Hesterberg Connection

Chihara & Hesterberg emphasize **computational** and **simulation-based** approaches. This connects beautifully to Bayesian inference:

```{r}
#| label: fig-posterior-simulation
#| fig-cap: "Posterior distribution for disease prevalence via direct simulation"
#| code-fold: true
#| fig-height: 5

# Simulate from Beta posterior directly (C&H computational spirit)
set.seed(551)
alpha1 <- 13; beta1 <- 53
n_sim <- 50000

posterior_draws <- tibble(p = rbeta(n_sim, alpha1, beta1))

# Summary
cred_int <- quantile(posterior_draws$p, c(0.025, 0.975))

ggplot(posterior_draws, aes(x = p)) +
  geom_histogram(aes(y = after_stat(density)), bins = 80,
                 fill = "#e74c3c", alpha = 0.7, color = "white") +
  geom_density(linewidth = 1.2, color = "#2c3e50") +
  stat_function(fun = dbeta, args = list(shape1 = alpha1, shape2 = beta1),
                color = "#3498db", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = cred_int, linetype = "dotted", color = "gray20", linewidth = 1) +
  annotate("text", x = cred_int[1] - 0.005, y = 4,
           label = paste0(round(cred_int[1], 3)), hjust = 1, size = 4.5) +
  annotate("text", x = cred_int[2] + 0.005, y = 4,
           label = paste0(round(cred_int[2], 3)), hjust = 0, size = 4.5) +
  labs(
    x = expression(italic(p)),
    y = "Density",
    title = expression("Posterior simulation: Beta("*.(13)*", "*.(53)*")"),
    subtitle = bquote("50,000 posterior draws — 95% credible interval: ["*
                        .(round(cred_int[1],3))*", "*.(round(cred_int[2],3))*"]")
  )
```

The dashed curve is the exact $\text{Beta}(13, 53)$ density — Monte Carlo is indistinguishable with 50,000 draws.

# Summary {background-color="#2c3e50"}

## Key Takeaways: Bayesian Inference

::: callout-important
## The Fundamental Workflow

$$\underbrace{\pi(\theta)}_{\text{prior}} \xrightarrow{\text{observe data}} \underbrace{\pi(\theta \mid \mathbf{x}) \propto L(\theta) \cdot \pi(\theta)}_{\text{posterior}}$$
:::

. . .

**Core concepts:**

- **Prior** encodes pre-data beliefs; can be informative, weakly informative, or flat
- **Posterior** is the complete inferential summary — it *is* the answer to "what do we know about $\theta$?"
- **Conjugate priors** give closed-form posteriors: Beta–Binomial, Normal–Normal, Gamma–Poisson
- **Credible intervals**: direct probability statements about $\theta$ — the interpretation frequentist CIs can't make
- **Bayes estimators**: minimize posterior expected loss; posterior mean (sq. error), median (abs. error), mode (0–1 loss)
- **Consistency**: by Bernstein–von Mises, posteriors concentrate around the MLE as $n \to \infty$

## Looking Forward

The tools from this quarter — MLE theory, sufficiency, sampling distributions, hypothesis testing — **all appear in the Bayesian framework** too:

- The likelihood $L(\theta)$ is the same object we have been maximizing
- Sufficient statistics reduce the posterior: $\pi(\theta \mid \mathbf{x}) = \pi(\theta \mid T(\mathbf{x}))$
- MCMC (Markov Chain Monte Carlo) generalizes bootstrap simulation for complex posteriors
- Bayesian adaptive designs, hierarchical models, and posterior predictive checks are standard tools in biostatistics

. . .

::: callout-tip
## Recommended Resources
- Gelman et al., *Bayesian Data Analysis* (3rd ed.) — the standard reference
- McElreath, *Statistical Rethinking* — intuitive introduction with R/Stan code
- Stan/brms: modern probabilistic programming for Bayesian models in R
:::

# Homework Problems {background-color="#006a4e"}

## Homework Problems

**Problem 1 (Beta–Binomial Updating)**

A clinical trial monitors the probability $p$ of a treatment-related serious adverse event. Historical data from similar drugs suggest $p \approx 0.08$, which you encode as a $\text{Beta}(2, 23)$ prior.

a. Verify that this prior has mean approximately 0.08 and compute its standard deviation.
b. In the current trial, 4 out of 75 patients experience the event. Derive the posterior distribution of $p$.
c. Compute the posterior mean, mode, and a 95% equal-tail credible interval.
d. An event rate above 0.15 would trigger a safety review. Compute the posterior probability $P(p > 0.15 \mid x = 4, n = 75)$.
e. How does the posterior probability in (d) change if you use a $\text{Uniform}(0,1)$ prior instead? Discuss the sensitivity of your conclusion to the prior choice.

---

**Problem 2 (Normal–Normal Conjugate)**

Systolic blood pressure measurements for a patient population are modeled as $X_i \mid \mu \stackrel{iid}{\sim} N(\mu, 20^2)$ (measurement $\sigma = 20$ mmHg is assumed known). A clinician's prior belief, based on patients from a similar clinic, is $\mu \sim N(130, 15^2)$.

a. A new dataset of $n = 25$ patients yields $\bar{x} = 142$ mmHg. Derive the posterior distribution of $\mu$.
b. Compute the posterior mean and show that it is a precision-weighted average of the prior mean and $\bar{x}$.
c. Construct a 95% credible interval for $\mu$. Compare it to the frequentist 95% CI for $\mu$ (using the known $\sigma$).
d. How large must $n$ be for the data precision to be at least 10 times the prior precision? At that sample size, how close are the Bayes and MLE estimates?

---

**Problem 3 (Bayes Estimators and Loss)**

Let $X_1, \ldots, X_n \mid p \stackrel{iid}{\sim} \text{Bernoulli}(p)$ and $p \sim \text{Beta}(\alpha, \beta)$.

a. Show that the Bayes estimator under squared error loss is $\hat{p}_B = (\alpha + \sum X_i)/(\alpha + \beta + n)$.
b. Write $\hat{p}_B$ as a weighted average of the prior mean $p_0 = \alpha/(\alpha + \beta)$ and the MLE $\bar{X}$. What are the weights?
c. Compute the **frequentist risk** $R(p, \hat{p}_B) = E_p[(\hat{p}_B - p)^2]$ and compare it to the risk of the MLE $\hat{p}_{\text{MLE}} = \bar{X}$.
d. Show that $\hat{p}_B$ has smaller MSE than the MLE when $p$ is near the prior mean, but larger MSE when $p$ is far from it. Illustrate with a plot in R using $\alpha = \beta = 2$, $n = 20$.

---

**Problem 4 (Gamma–Poisson; Devore-style)**

Hospital records show the number of patients presenting with a rare condition per month follows a Poisson distribution with unknown rate $\lambda$. The prior is $\lambda \sim \text{Gamma}(3, 1)$ (mean = 3 cases/month, $\beta = 1$ is the rate parameter).

a. In the past 6 months, the observed counts were 2, 5, 1, 4, 3, 2. Derive the posterior distribution of $\lambda$.
b. Compute the posterior mean and compare to the MLE.
c. Compute a 95% equal-tail credible interval using the `qgamma()` function in R.
d. The hospital wants to staff for a month where $\lambda > 6$. Compute the posterior probability of this event.
