---
title: "Lesson 13: Tests About a Population Mean and Proportion"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w6d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 13 - Tests About a Mean and Proportion
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 13: Tests About a Population Mean and Proportion {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 12: Introduction to Hypothesis Testing**

-   Null and alternative hypotheses ($H_0$ vs $H_a$)
-   Test statistics and rejection regions
-   Type I error ($\alpha$) and Type II error ($\beta$)
-   Power = $1 - \beta$
-   Connection between CIs and hypothesis tests

. . .

**Today's Goals:**

1. Apply the one-sample z test for $\mu$ ($\sigma$ known)
2. Develop the one-sample t test for $\mu$ ($\sigma$ unknown)
3. Understand when to use z vs t tests
4. Conduct tests about a population proportion $p$
5. Implement all tests in R with medical research examples

## Roadmap for Today

| Topic | Assumption | Test Statistic | Textbook |
|-------|------------|----------------|----------|
| z test for $\mu$ | $\sigma$ known | $Z = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}}$ | Devore 9.2 |
| t test for $\mu$ | $\sigma$ unknown | $T = \frac{\bar{X} - \mu_0}{S/\sqrt{n}}$ | Devore 9.2 |
| z test for $p$ | Large $n$ | $Z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$ | Devore 9.3 |

. . .

::: callout-tip
## Connection to CIs
Each test today has a corresponding confidence interval from Lessons 7--8. The CI contains all values of the parameter that would **not** be rejected by the corresponding test.
:::

# The One-Sample z Test (Devore 9.2) {background-color="#3498db"}

## Setup and Assumptions

**Assumptions:**

1. $X_1, X_2, \ldots, X_n$ is a random sample from a normal distribution
2. The population standard deviation $\sigma$ is **known**

. . .

**Null hypothesis:** $H_0: \mu = \mu_0$

**Test statistic:**

$$Z = \frac{\bar{X} - \mu_0}{\sigma / \sqrt{n}}$$

Under $H_0$: $Z \sim N(0, 1)$

## The One-Sample z Test: Rejection Regions

::: callout-important
## One-Sample z Test (Devore 9.2)

| Alternative | Rejection Region | Type |
|:---:|:---:|:---:|
| $H_a: \mu > \mu_0$ | $z \geq z_\alpha$ | Upper-tailed |
| $H_a: \mu < \mu_0$ | $z \leq -z_\alpha$ | Lower-tailed |
| $H_a: \mu \neq \mu_0$ | $|z| \geq z_{\alpha/2}$ | Two-tailed |
:::

. . .

**Key idea:** The standardized test statistic $Z$ measures how many standard errors $\bar{x}$ is from $\mu_0$. Rejection occurs when this distance is "too large" in the direction specified by $H_a$.

## Seven-Step Procedure (Devore 9.2)

::: callout-note
## Recommended Steps for Any Hypothesis Test
1. **Identify** the parameter of interest and describe it in context
2. **State** the null and alternative hypotheses
3. **Check** assumptions for the test procedure
4. **Write** the formula for the test statistic
5. **State** the rejection region for the chosen $\alpha$
6. **Compute** the test statistic value from sample data
7. **Decide** whether to reject $H_0$ and state the conclusion in context
:::

. . .

We will use this structured approach throughout the course.

## Example: Neonatal Birth Weight Screening

**Setting:** A neonatal unit monitors the average birth weight of infants. Historically, the mean birth weight in this hospital is 3300 g with known $\sigma = 450$ g. After a change in prenatal care protocol, a sample of $n = 36$ births yields $\bar{x} = 3180$ g.

**Question:** Is there evidence that the new protocol has changed the mean birth weight?

. . .

**Step 1:** $\mu$ = true mean birth weight (g) under the new protocol

**Step 2:** $H_0: \mu = 3300$ vs $H_a: \mu \neq 3300$ (two-tailed --- either direction is concerning)

**Step 3:** $\sigma = 450$ is known; $n = 36$ is large enough for CLT

## Example: Neonatal Birth Weight (continued)

**Step 4:** Test statistic:
$$Z = \frac{\bar{X} - 3300}{450/\sqrt{36}} = \frac{\bar{X} - 3300}{75}$$

**Step 5:** For $\alpha = 0.05$: Reject $H_0$ if $|z| \geq z_{0.025} = 1.96$

```{r}
#| echo: true
# Step 6: Compute test statistic
xbar <- 3180
mu0 <- 3300
sigma <- 450
n <- 36
se <- sigma / sqrt(n)

z_stat <- (xbar - mu0) / se
cat("Test statistic: z =", round(z_stat, 3), "\n")

# Critical value
z_crit <- qnorm(1 - 0.05/2)
cat("Critical value: z_{alpha/2} =", round(z_crit, 3), "\n")
cat("Rejection region: |z| >=", round(z_crit, 3))
```

## Example: Neonatal Birth Weight (decision)

```{r}
#| echo: true
# Step 7: Decision
cat("|z| =", round(abs(z_stat), 3), "\n")
cat("Critical value =", round(z_crit, 3), "\n\n")

if (abs(z_stat) >= z_crit) {
  cat("Decision: Reject H0 at alpha = 0.05\n\n")
  cat("Conclusion: There is statistically significant evidence that\n")
  cat("the true mean birth weight under the new protocol differs from 3300 g.")
} else {
  cat("Decision: Fail to reject H0 at alpha = 0.05\n\n")
  cat("Conclusion: There is insufficient evidence that the true mean\n")
  cat("birth weight differs from 3300 g under the new protocol.")
}
```

## Visualizing the Test

```{r}
#| code-fold: true
#| fig-height: 6
z_seq <- seq(-4, 4, length.out = 500)
y <- dnorm(z_seq)

tibble(z = z_seq, y = y) |>
  ggplot(aes(x = z, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(z = z_seq[z_seq >= z_crit], 
                          y = dnorm(z_seq[z_seq >= z_crit])),
            aes(x = z, y = y), fill = "coral", alpha = 0.5) +
  geom_area(data = tibble(z = z_seq[z_seq <= -z_crit], 
                          y = dnorm(z_seq[z_seq <= -z_crit])),
            aes(x = z, y = y), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = c(-z_crit, z_crit), linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = z_stat, color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = z_crit + 0.1, y = 0.35, hjust = 0, size = 5, color = "red",
           label = paste0("z_{0.025} = ", round(z_crit, 2))) +
  annotate("text", x = -z_crit - 0.1, y = 0.35, hjust = 1, size = 5, color = "red",
           label = paste0("-z_{0.025} = ", round(-z_crit, 2))) +
  annotate("text", x = z_stat - 0.1, y = 0.25, hjust = 1, size = 5, color = "darkgreen",
           label = paste0("z = ", round(z_stat, 2))) +
  annotate("text", x = 3, y = 0.05, label = expression(alpha/2 == 0.025), 
           color = "coral", size = 5) +
  annotate("text", x = -3, y = 0.05, label = expression(alpha/2 == 0.025), 
           color = "coral", size = 5) +
  labs(x = "z", y = "Density",
       title = "Two-Tailed z Test: Neonatal Birth Weight",
       subtitle = expression("Test statistic falls in rejection region" ~
                             symbol("\256") ~ "Reject" ~ H[0]))
```

## Confirming with a Confidence Interval

Recall the **CI-test duality**: the 95% CI contains all $\mu_0$ values that would not be rejected.

```{r}
#| echo: true
# 95% CI for mu
ci_lower <- xbar - z_crit * se
ci_upper <- xbar + z_crit * se
cat("95% CI for mu: (", round(ci_lower, 1), ",", round(ci_upper, 1), ")\n")
cat("Null value mu_0 = 3300\n\n")
cat("Since 3300 is NOT in the CI, we reject H0 -- consistent with our test!")
```

```{r}
#| code-fold: true
#| fig-height: 3.5
tibble(x = xbar, lower = ci_lower, upper = ci_upper) |>
  ggplot(aes(y = 1)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.1, linewidth = 2, color = "steelblue") +
  geom_point(aes(x = x), size = 5, color = "steelblue") +
  geom_vline(xintercept = 3300, linetype = "dashed", color = "red", linewidth = 1.5) +
  annotate("text", x = 3300, y = 1.15, label = expression(H[0]:~mu == 3300), 
           color = "red", size = 6) +
  annotate("text", x = xbar, y = 0.85, size = 5,
           label = bquote(bar(x) == .(xbar))) +
  coord_cartesian(xlim = c(3000, 3400), ylim = c(0.7, 1.3)) +
  labs(x = "Birth Weight (g)", y = "",
       title = "95% Confidence Interval vs Null Value") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

## Type II Error and Power for the z Test (Devore 9.2)

For the z test, we have closed-form expressions for $\beta(\mu')$:

| Alternative | $\beta(\mu')$ |
|:---:|:---:|
| $H_a: \mu > \mu_0$ | $\Phi\!\left(z_\alpha + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right)$ |
| $H_a: \mu < \mu_0$ | $1 - \Phi\!\left(-z_\alpha + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right)$ |
| $H_a: \mu \neq \mu_0$ | $\Phi\!\left(z_{\alpha/2} + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right) - \Phi\!\left(-z_{\alpha/2} + \frac{\mu_0 - \mu'}{\sigma/\sqrt{n}}\right)$ |

where $\Phi(z)$ is the standard normal CDF. Power = $1 - \beta(\mu')$.

## Power Curve: Birth Weight Example

```{r}
#| code-fold: true
#| fig-height: 6
alpha <- 0.05
mu_range <- seq(3000, 3600, by = 5)

# Two-tailed power
power_vals <- map_dbl(mu_range, \(mu_prime) {
  beta <- pnorm(z_crit + (mu0 - mu_prime) / se) - 
          pnorm(-z_crit + (mu0 - mu_prime) / se)
  1 - beta
})

tibble(mu = mu_range, power = power_vals) |>
  ggplot(aes(x = mu, y = power)) +
  geom_line(linewidth = 1.5, color = "darkgreen") +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "gray50") +
  geom_vline(xintercept = mu0, linetype = "dotted", alpha = 0.5) +
  annotate("text", x = 3550, y = 0.82, label = "80% Power", color = "red", size = 5) +
  annotate("text", x = 3300, y = 0.12, size = 5,
           label = expression(H[0]:~mu == 3300)) +
  annotate("text", x = 3550, y = 0.07, label = expression(alpha == 0.05), 
           color = "gray50", size = 4) +
  labs(x = expression("True " * mu * " (g)"), y = "Power",
       title = "Power Function: Birth Weight Two-Tailed z Test",
       subtitle = expression(n == 36 ~ ", " ~ sigma == 450 ~ ", " ~ alpha == 0.05)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2))
```

## Sample Size Determination for the z Test

The sample size $n$ for which a level $\alpha$ test also satisfies $\beta(\mu') = \beta$ is:

$$n = \begin{cases} \left[\frac{\sigma(z_\alpha + z_\beta)}{\mu_0 - \mu'}\right]^2 & \text{one-tailed test} \\[10pt] \left[\frac{\sigma(z_{\alpha/2} + z_\beta)}{\mu_0 - \mu'}\right]^2 & \text{two-tailed test (approx)} \end{cases}$$

```{r}
#| echo: true
# How many births to detect a 100g decrease with 80% power?
delta <- 100  # clinically meaningful difference
alpha <- 0.05
beta_target <- 0.20  # want power = 0.80
sigma <- 450

# Two-tailed test
n_needed <- (sigma * (qnorm(1 - alpha/2) + qnorm(1 - beta_target)) / delta)^2
cat("Sample size needed:", ceiling(n_needed), "births")
```

. . .

::: callout-tip
## Clinical Relevance
In medical research, we should determine the **clinically meaningful difference** before planning the study. A 100 g difference in birth weight may be clinically important; a 10 g difference likely is not.
:::

# The One-Sample t Test (Devore 9.2) {background-color="#9b59b6"}

## From z to t: The Realistic Case

**The Problem:** In practice, we **rarely know** $\sigma$.

. . .

**Solution:** Replace $\sigma$ with the sample standard deviation $S$:

$$Z = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}} \quad \longrightarrow \quad T = \frac{\bar{X} - \mu_0}{S/\sqrt{n}}$$

. . .

**Key result (from Lesson 8):** When sampling from a normal population,

$$T = \frac{\bar{X} - \mu_0}{S/\sqrt{n}} \sim t_{n-1} \quad \text{when } H_0 \text{ is true}$$

The $t_{n-1}$ distribution has **heavier tails** than the standard normal, reflecting the additional uncertainty from estimating $\sigma$.

## The One-Sample t Test

::: callout-important
## One-Sample t Test (Devore 9.2)

**Null hypothesis:** $H_0: \mu = \mu_0$

**Test statistic:** $\displaystyle t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}$

| Alternative | Rejection Region |
|:---:|:---:|
| $H_a: \mu > \mu_0$ | $t \geq t_{\alpha, n-1}$ |
| $H_a: \mu < \mu_0$ | $t \leq -t_{\alpha, n-1}$ |
| $H_a: \mu \neq \mu_0$ | $|t| \geq t_{\alpha/2, n-1}$ |
:::

. . .

**Assumptions:**

- The data are a random sample from a normal distribution
- For **large** $n$ (say $n > 40$), the test is robust to non-normality

## z vs t: Comparison of Critical Values

```{r}
#| code-fold: true
#| fig-height: 6
# Compare z and t critical values
df_vals <- c(5, 10, 20, 50, 100, Inf)
alpha <- 0.05

crit_table <- tibble(
  df = df_vals,
  `t critical value` = qt(1 - alpha/2, df = df_vals),
  `z critical value` = qnorm(1 - alpha/2)
) |>
  mutate(df_label = ifelse(is.infinite(df), "Inf (z)", as.character(df)))

crit_table |>
  select(-df_label) |>
  mutate(across(where(is.numeric) & !df, \(x) round(x, 4)))
```

. . .

```{r}
#| code-fold: true
#| fig-height: 5
x_seq <- seq(-4, 4, length.out = 500)

tibble(
  x = rep(x_seq, 3),
  density = c(dnorm(x_seq), dt(x_seq, df = 5), dt(x_seq, df = 15)),
  dist = rep(c("N(0,1)", "t (df=5)", "t (df=15)"), each = length(x_seq))
) |>
  mutate(dist = factor(dist, levels = c("N(0,1)", "t (df=15)", "t (df=5)"))) |>
  ggplot(aes(x = x, y = density, color = dist)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c("N(0,1)" = "steelblue", "t (df=15)" = "darkgreen", "t (df=5)" = "coral")) +
  labs(x = "x", y = "Density", color = "Distribution",
       title = "Standard Normal vs t Distributions",
       subtitle = "t distributions have heavier tails, especially with small df") +
  theme(legend.position = "right")
```

## Example: Hemoglobin A1c Monitoring (Devore 9.2 Style)

**Setting:** A diabetes clinic monitors glycated hemoglobin (HbA1c). The target for well-controlled diabetes is HbA1c $\leq$ 7.0%. A sample of $n = 15$ patients on a new medication yields:

```{r}
#| echo: true
hba1c <- c(6.8, 7.2, 6.5, 7.8, 6.9, 7.4, 6.3, 7.1, 6.7, 
            7.5, 6.6, 7.0, 7.3, 6.4, 7.6)

# Summary statistics
xbar <- mean(hba1c)
s <- sd(hba1c)
n <- length(hba1c)
cat("n =", n, "\n")
cat("x-bar =", round(xbar, 3), "\n")
cat("s =", round(s, 3))
```

**Question:** Is there evidence that the true mean HbA1c for patients on this medication exceeds the 7.0% target?

## HbA1c Example: Check Assumptions

**Step 3:** Check normality assumption.

```{r}
#| code-fold: true
#| fig-height: 5
#| fig-width: 14
p1 <- tibble(hba1c = hba1c) |>
  ggplot(aes(x = hba1c)) +
  geom_histogram(bins = 7, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = 7.0, linetype = "dashed", color = "red", linewidth = 1) +
  labs(x = "HbA1c (%)", y = "Count", title = "Distribution of HbA1c Values") +
  annotate("text", x = 7.0, y = 4.2, label = "Target = 7.0%", 
           color = "red", hjust = -0.1, size = 5)

p2 <- tibble(hba1c = hba1c) |>
  ggplot(aes(sample = hba1c)) +
  stat_qq(size = 3, color = "steelblue") +
  stat_qq_line(color = "red", linewidth = 1) +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles",
       title = "Normal Q-Q Plot")

p1 + p2
```

The Q-Q plot shows approximately linear pattern --- normality assumption is reasonable for $n = 15$.

## HbA1c Example: Conduct the t Test

```{r}
#| echo: true
# Step 1-2: mu = true mean HbA1c; H0: mu = 7.0 vs Ha: mu > 7.0
mu0 <- 7.0
alpha <- 0.05

# Step 4: Test statistic
se <- s / sqrt(n)
t_stat <- (xbar - mu0) / se
cat("Test statistic: t =", round(t_stat, 3), "\n")

# Step 5: Critical value (upper-tailed, df = n-1)
t_crit <- qt(1 - alpha, df = n - 1)
cat("Critical value: t_{0.05, 14} =", round(t_crit, 3), "\n")

# Step 7: Decision
cat("\nDecision:", 
    ifelse(t_stat >= t_crit, "Reject H0", "Fail to reject H0"))
```

. . .

**Step 7 Interpretation:** There is insufficient evidence at $\alpha = 0.05$ to conclude that the true mean HbA1c exceeds the 7.0% target.

## Using R's `t.test()` Function

```{r}
#| echo: true
# R's built-in t.test function
t_result <- t.test(hba1c, mu = 7.0, alternative = "greater")
t_result
```

. . .

```{r}
#| echo: true
# Extract tidy results
tidy(t_result)
```

## Visualizing the t Test Result

```{r}
#| code-fold: true
#| fig-height: 6
df_t <- n - 1
x_seq <- seq(-4, 4, length.out = 500)
y <- dt(x_seq, df = df_t)

tibble(x = x_seq, y = y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(x = x_seq[x_seq >= t_crit],
                          y = dt(x_seq[x_seq >= t_crit], df = df_t)),
            aes(x = x, y = y), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = t_crit, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = t_stat, color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = t_crit + 0.15, y = 0.33, hjust = 0, size = 5, color = "red",
           label = bquote(t[list(0.05, 14)] == .(round(t_crit, 3)))) +
  annotate("text", x = t_stat + 0.15, y = 0.25, hjust = 0, size = 5, color = "darkgreen",
           label = paste0("t = ", round(t_stat, 3))) +
  annotate("text", x = 2.8, y = 0.05, label = expression(alpha == 0.05), 
           color = "coral", size = 5) +
  labs(x = "t", y = "Density",
       title = expression("Upper-Tailed t Test: HbA1c (" * t[14] * " distribution)"),
       subtitle = expression("Test statistic does NOT fall in rejection region" ~
                             symbol("\256") ~ "Fail to reject" ~ H[0]))
```

## Example: Body Temperature Revisited (Two-Tailed t Test)

The widely cited "normal" body temperature of 98.6°F comes from a 19th century study. Mackowiak et al. (1992, *JAMA*) measured temperatures in 130 healthy adults.

```{r}
#| echo: true
# Simulated data consistent with the Mackowiak et al. findings
set.seed(1992)
body_temp <- rnorm(130, mean = 98.2, sd = 0.73)

cat("n =", length(body_temp), "\n")
cat("x-bar =", round(mean(body_temp), 3), "\n")
cat("s =", round(sd(body_temp), 3))
```

. . .

**Hypotheses:** $H_0: \mu = 98.6$ vs $H_a: \mu \neq 98.6$

## Body Temperature: t Test in R

```{r}
#| echo: true
# Two-tailed t test
temp_test <- t.test(body_temp, mu = 98.6, alternative = "two.sided")
temp_test
```

. . .

```{r}
#| echo: true
# Extract key quantities
cat("t statistic:", round(temp_test$statistic, 3), "\n")
cat("p-value:", format.pval(temp_test$p.value, digits = 3), "\n")
cat("95% CI: (", round(temp_test$conf.int[1], 3), ",", 
    round(temp_test$conf.int[2], 3), ")\n\n")
cat("Since 98.6 is NOT in the 95% CI, we reject H0.\n")
cat("Evidence strongly suggests true mean body temperature differs from 98.6 F.")
```

## When Can We Use the t Test?

::: callout-important
## Guidelines for the One-Sample t Test

**Normality is critical** when $n$ is small:

| Sample Size | Guideline |
|:---:|:---|
| $n < 15$ | Use t test only if data appear close to normal (check Q-Q plot) |
| $15 \leq n \leq 40$ | Mild skewness is acceptable; avoid if heavily skewed or outliers present |
| $n > 40$ | t test is robust to non-normality by CLT; can use for most distributions |
:::

. . .

::: callout-tip
## Practical Advice (Devore 9.2)
For moderate sample sizes, examine a **normal probability plot** (Q-Q plot) before using the t test. If the data show substantial departures from normality, consider a **bootstrap test** (Lesson 15) or a **nonparametric test** instead.
:::

## Simulation: Coverage of the t Test Under Non-Normality

```{r}
#| echo: true
#| code-fold: true
set.seed(551)
n_sims <- 10000
alpha <- 0.05
mu0 <- 5

# Simulate Type I error rate for different distributions
simulate_type1 <- function(rdist, n, ...) {
  rejections <- replicate(n_sims, {
    x <- rdist(n, ...)
    test <- t.test(x, mu = mu0)
    test$p.value < alpha
  })
  mean(rejections)
}

results <- tibble(
  Distribution = c("Normal", "Normal", "Exp (shifted)", "Exp (shifted)", 
                    "t(3) (shifted)", "t(3) (shifted)"),
  n = rep(c(10, 50), 3),
  `Type I Error Rate` = c(
    simulate_type1(rnorm, 10, mean = 5, sd = 1),
    simulate_type1(rnorm, 50, mean = 5, sd = 1),
    simulate_type1(\(n, ...) rexp(n, rate = 1) + 4, 10),
    simulate_type1(\(n, ...) rexp(n, rate = 1) + 4, 50),
    simulate_type1(\(n, ...) rt(n, df = 3) + 5, 10),
    simulate_type1(\(n, ...) rt(n, df = 3) + 5, 50)
  )
)

results |> mutate(`Type I Error Rate` = round(`Type I Error Rate`, 4))
```

. . .

The t test maintains $\alpha \approx 0.05$ well for normal and heavy-tailed symmetric distributions, but may be anticonservative for skewed distributions with small $n$.

# Textbook Example: Particle Size (Devore Example 9.12) {background-color="#2c3e50"}

## Devore Example 9.12: Particulate Matter

Roadside measurements of particle sizes ($d_{50}$ in microns) at $n = 9$ assays near a North Carolina highway:

```{r}
#| echo: true
d50 <- c(82.9, 56.8, 66.5, 49.4, 105.4, 79.5, 82.5, 50.7, 43.0)

# Summary statistics
cat("n =", length(d50), "\n")
cat("x-bar =", round(mean(d50), 2), "\n")
cat("s =", round(sd(d50), 2))
```

Previous studies indicated that the typical $d_{50}$ value alongside similar roads is 44 microns.

**Hypotheses:** $H_0: \mu = 44$ vs $H_a: \mu \neq 44$ at $\alpha = 0.01$.

## Devore Example 9.12: Solution

```{r}
#| echo: true
# Check normality
```

```{r}
#| code-fold: true
#| fig-height: 7
tibble(d50 = d50) |>
  ggplot(aes(sample = d50)) +
  stat_qq(size = 3, color = "steelblue") +
  stat_qq_line(color = "red", linewidth = 1) +
  labs(title = "Q-Q Plot: Particle Size Data", 
       x = "Theoretical Quantiles", y = "Sample Quantiles")
```

## Devore Example 9.12: Solution

```{r}
#| echo: true
# t test
t.test(d50, mu = 44, alternative = "two.sided", conf.level = 0.99)
```

. . .

Since $|t| = 3.59 > t_{0.005, 8} = 3.355$, we reject $H_0$ at $\alpha = 0.01$. The data provides strong evidence that the true mean $d_{50}$ at this site differs from 44 microns.

# Textbook Example: Tip Percentages (Devore Example 9.13) {background-color="#2c3e50"}

## Devore Example 9.13: Large-Sample t Test

A sample of $n = 70$ tip percentages at a restaurant. Summary statistics: $\bar{x} = 17.986$, $s = 5.937$.

**Question:** Is there evidence that the true average tip percentage exceeds 15%?

```{r}
#| echo: true
# Given summary statistics (large n, don't need raw data)
xbar_tip <- 17.986
s_tip <- 5.937
n_tip <- 70
mu0_tip <- 15
alpha <- 0.05

# t test statistic
se_tip <- s_tip / sqrt(n_tip)
t_stat_tip <- (xbar_tip - mu0_tip) / se_tip
t_crit_tip <- qt(1 - alpha, df = n_tip - 1)

cat("t =", round(t_stat_tip, 2), "\n")
cat("t_{0.05, 69} =", round(t_crit_tip, 3), "\n")
cat("Decision:", ifelse(t_stat_tip >= t_crit_tip, "Reject H0", "Fail to reject H0"))
```

. . .

**Note:** With $n = 70 > 40$, the t test does not require a normal population distribution. The data are positively skewed, but the large sample makes the test valid by the CLT.

# Tests About a Population Proportion (Devore 9.3) {background-color="#e74c3c"}

## Testing a Population Proportion

**Setting:** We want to test a claim about the population proportion $p$ of "successes."

**Data:** $X$ = number of successes in a random sample of size $n$, where $X \sim \text{Binomial}(n, p)$.

. . .

**Estimator:** $\hat{p} = X/n$ (sample proportion)

. . .

When $H_0: p = p_0$ is true and $n$ is large:

$$\hat{P} \approx N\!\left(p_0, \frac{p_0(1-p_0)}{n}\right)$$

Standardizing gives the test statistic:

$$Z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$$

## The One-Proportion z Test

::: callout-important
## One-Proportion z Test (Devore 9.3)

**Null hypothesis:** $H_0: p = p_0$

**Test statistic:** $\displaystyle z = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$

| Alternative | Rejection Region |
|:---:|:---:|
| $H_a: p > p_0$ | $z \geq z_\alpha$ |
| $H_a: p < p_0$ | $z \leq -z_\alpha$ |
| $H_a: p \neq p_0$ | $|z| \geq z_{\alpha/2}$ |

**Validity condition:** $np_0 \geq 10$ and $n(1-p_0) \geq 10$
:::

. . .

**Key point:** Under $H_0$, we use $p_0$ (not $\hat{p}$) in the standard error! This differs from the CI formula where we used $\hat{p}$ (Wald) or the score approach.

## Why $p_0$ in the Standard Error?

```{r}
#| code-fold: true
#| fig-height: 5
#| fig-width: 14
# Compare SE under H0 vs using p-hat
p0 <- 0.30
n <- 200
phat_vals <- seq(0.15, 0.45, by = 0.001)

se_h0 <- sqrt(p0 * (1 - p0) / n)
se_wald <- sqrt(phat_vals * (1 - phat_vals) / n)

p1 <- tibble(phat = phat_vals, se = se_wald) |>
  ggplot(aes(x = phat, y = se)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_hline(yintercept = se_h0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = p0, linetype = "dotted", alpha = 0.5) +
  annotate("text", x = 0.42, y = se_h0 + 0.001, color = "red", size = 5,
           label = bquote("SE under" ~ H[0] ~ "=" ~ .(round(se_h0, 4)))) +
  labs(x = expression(hat(p)), y = "Standard Error",
       title = "Standard Error: Under H0 vs Using Sample Proportion",
       subtitle = expression(p[0] == 0.30 ~ ", " ~ n == 200))

p1
```

. . .

**Reasoning:** When **testing**, we assume $H_0$ is true and ask "how unlikely is the data?" So we use $p_0$. When **estimating** (CI), we don't assume a specific value for $p$, so we use $\hat{p}$.

## Example: Post-Surgical Infection Rate

**Setting:** The national rate of post-surgical site infections is 2.5% for a certain procedure. A hospital implements a new sterilization protocol and observes 8 infections among 500 surgeries.

**Question:** Has the new protocol reduced the infection rate?

. . .

```{r}
#| echo: true
# Data
x <- 8          # infections observed
n <- 500        # total surgeries
p_hat <- x / n
p0 <- 0.025     # national rate

cat("Sample proportion: p-hat =", p_hat, "\n")
cat("Expected under H0:", n * p0, "infections\n")

# Check validity: np0 and n(1-p0)
cat("np0 =", n * p0, ">= 10?", n * p0 >= 10, "\n")
cat("n(1-p0) =", n * (1 - p0), ">= 10?", n * (1 - p0) >= 10)
```

## Infection Rate: Conduct the Test

```{r}
#| echo: true
# H0: p = 0.025 vs Ha: p < 0.025 (lower-tailed)
alpha <- 0.05

# Test statistic
se_h0 <- sqrt(p0 * (1 - p0) / n)
z_stat <- (p_hat - p0) / se_h0
z_crit <- qnorm(alpha)  # lower-tailed critical value

cat("z =", round(z_stat, 3), "\n")
cat("Critical value: -z_{0.05} =", round(z_crit, 3), "\n\n")

if (z_stat <= z_crit) {
  cat("Decision: Reject H0\n")
  cat("Conclusion: Evidence that the new protocol reduces infection rate.")
} else {
  cat("Decision: Fail to reject H0\n")
  cat("Conclusion: Insufficient evidence that the new protocol\n")
  cat("reduces infection rate below the national 2.5%.")
}
```

## Visualizing the Proportion Test

```{r}
#| code-fold: true
#| fig-height: 6
z_seq <- seq(-4, 4, length.out = 500)
y <- dnorm(z_seq)

tibble(z = z_seq, y = y) |>
  ggplot(aes(x = z, y = y)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_area(data = tibble(z = z_seq[z_seq <= qnorm(alpha)],
                          y = dnorm(z_seq[z_seq <= qnorm(alpha)])),
            aes(x = z, y = y), fill = "coral", alpha = 0.5) +
  geom_vline(xintercept = z_crit, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = z_stat, color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = z_crit - 0.1, y = 0.33, hjust = 1, size = 5, color = "red",
           label = bquote(-z[0.05] == .(round(z_crit, 3)))) +
  annotate("text", x = z_stat + 0.1, y = 0.25, hjust = 0, size = 5, color = "darkgreen",
           label = paste0("z = ", round(z_stat, 3))) +
  labs(x = "z", y = "Density",
       title = "Lower-Tailed z Test: Post-Surgical Infection Rate",
       subtitle = expression("Test statistic does NOT fall in rejection region" ~
                             symbol("\256") ~ "Fail to reject" ~ H[0]))
```

## What Does `prop.test()` Actually Do?

R's `prop.test()` performs the **score test** for a population proportion --- a test based on the score pivot used in the score interval in Lesson 8. Recall that the pivot:
$$Z_{score} = \frac{U(\theta)}{\sqrt{I_n(\theta)}} \xrightarrow{d} N(0,1)$$

. . .

and for a Bernoulli sample:
$$Z_{score} = \frac{\hat{p} - p}{\sqrt{p(1-p)/n}}$$
This is exactly the z statistic we computed by hand above. The function `prop.test()` reports the **squared** version under $H_0$:

$$X^2 = Z_{score}^2 = \frac{(\hat{p} - p_0)^2}{p_0(1-p_0)/n} \sim \chi^2_1$$

. . .

::: callout-important
## Why $\chi^2_1$?
If $Z \sim N(0,1)$, then $Z^2 \sim \chi^2_1$ (chi-squared with 1 df). The score test for proportions can equivalently be expressed as a z test or a $\chi^2$ test --- they give exactly the same p-value.
:::

## `prop.test()` is the Score Test

```{r}
#| echo: true
# Our z test by hand
z_stat_manual <- (p_hat - p0) / sqrt(p0 * (1 - p0) / n)
cat("Manual z statistic:", round(z_stat_manual, 3), "\n")
cat("Manual z^2:        ", round(z_stat_manual^2, 3), "\n\n")

# prop.test computes the same thing as chi-squared
prop_result <- prop.test(x = 8, n = 500, p = 0.025, 
                          alternative = "less", correct = FALSE)

cat("prop.test chi-squared:", round(prop_result$statistic, 3), "\n")
cat("sqrt(chi-squared):    ", round(sqrt(prop_result$statistic), 3), "\n")
cat("p-value (prop.test):  ", round(prop_result$p.value, 4))
```

. . .

The $X^2$ statistic from `prop.test()` equals $z^2$ from our manual calculation!

## Connection to the Score Interval (Lesson 8)

In Lesson 8, we derived the **Score (Wilson) confidence interval** by inverting the score test. Recall:

The score interval collects all values $p$ satisfying:
$$\left|\frac{\hat{p} - p}{\sqrt{p(1-p)/n}}\right| \leq z_{\alpha/2}$$

. . .

The key difference from the Wald interval: the **score** approach keeps the parameter $p$ in the standard error, rather than plugging in $\hat{p}$, and when we are "inverting the test statistic" we no longer use the null $p_0$ but a general $p$ that we need to solve for.

## Connection to the Score Interval (Lesson 8 `prop.test`)

**The CI returned by `prop.test()` is exactly the Score (Wilson) interval:**

```{r}
#| echo: true
# prop.test returns the Score (Wilson) CI
cat("Score CI from prop.test():\n")
cat("  (", round(prop_result$conf.int[1], 5), ",", 
         round(prop_result$conf.int[2], 5), ")\n\n")

# Verify: this is the Wilson interval from Lesson 8
z_alpha2 <- qnorm(0.975)
p_hat_val <- 8/500
n_val <- 500

# Wilson formula: (p_hat + z^2/(2n) +/- z*sqrt(p_hat*(1-p_hat)/n + z^2/(4n^2))) / (1 + z^2/n)
center <- (p_hat_val + z_alpha2^2 / (2 * n_val)) / (1 + z_alpha2^2 / n_val)
margin <- z_alpha2 * sqrt(p_hat_val * (1 - p_hat_val) / n_val + z_alpha2^2 / (4 * n_val^2)) / 
          (1 + z_alpha2^2 / n_val)

cat("Wilson formula by hand:\n")
cat("  (", round(center - margin, 5), ",", round(center + margin, 5), ")")
```

## Three Tests, One Framework

The relationship between the Wald test, Score test, and Likelihood Ratio test is a fundamental concept in statistical inference (we'll see the LRT in Lesson 15):

| Approach | SE evaluated at | Test Statistic | CI |
|:---|:---:|:---:|:---:|
| **Wald** | MLE $\hat{p}$ | $\frac{\hat{p} - p_0}{\sqrt{\hat{p}(1-\hat{p})/n}}$ | Wald CI |
| **Score** | Null $p_0$ | $\frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$ | Wilson/Score CI |

. . .

::: callout-tip
## Key Insight
When **testing** $H_0: p = p_0$, we use $p_0$ in the SE (score test). The z test for proportions from Devore 9.3 **is** the score test. The CI from inverting this test is the Score (Wilson) interval from Lesson 8 --- not the Wald interval.
:::

## `prop.test()` Options

```{r}
#| echo: true
# Full output from prop.test (using broom for tidy results)
prop.test(x = 8, n = 500, p = 0.025, 
          alternative = "less", correct = FALSE) |>
  tidy() |>
  select(estimate, statistic, p.value, conf.low, conf.high, method)
```

. . .

**Important arguments:**

- `correct = FALSE`: Do **not** apply Yates' continuity correction (matches our z test formula)
- `correct = TRUE` (default): Applies the correction $|\hat{p} - p_0| - \frac{1}{2n}$; slightly conservative
- `alternative`: "two.sided" (default), "less", or "greater"
- The `conf.int` returned is always the **Score (Wilson) CI**, regardless of the `alternative`

# The Exact Binomial Test {background-color="#e74c3c"}

## When the Normal Approximation Fails

The large-sample z test relies on:
$$\frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} \;\dot{\sim}\; N(0, 1)$$

This approximation requires $np_0 \geq 10$ and $n(1-p_0) \geq 10$.

. . .

**What do we do when these conditions fail?** For example:

- Testing if a rare adverse event rate differs from 1% with only 50 patients ($np_0 = 0.5$)
- Evaluating a new surgical technique on 15 patients

. . .

**Answer:** Use the **exact binomial test**, which works directly with the binomial distribution --- no approximation needed!

## Deriving the Exact Test from First Principles

**Setup:** Let $X$ = number of successes in $n$ independent Bernoulli trials.

Under $H_0: p = p_0$, we know exactly:
$$X \sim \text{Binomial}(n, p_0)$$

$$P(X = k \mid p_0) = \binom{n}{k} p_0^k (1 - p_0)^{n-k}, \quad k = 0, 1, \ldots, n$$

. . .

The test statistic is simply $X$ itself --- the count of successes. No standardization needed!

. . .

::: callout-important
## Why "Exact"?
The z test is *approximate* because it relies on the CLT approximation $\hat{p} \dot{\sim} N(p, p(1-p)/n)$. The binomial test is *exact* because it uses the actual, known distribution of $X$ under $H_0$ --- no asymptotics required.
:::

## Rejection Regions for the Exact Binomial Test

For the **upper-tailed** test $H_a: p > p_0$, we reject for large $X$:

The rejection region is $\{c, c+1, \ldots, n\}$ where $c$ is chosen so that:
$$P(X \geq c \mid p = p_0) = \sum_{k=c}^{n} \binom{n}{k} p_0^k(1-p_0)^{n-k} \leq \alpha$$

. . .

**Complication with discrete distributions:** Because $X$ is discrete, we generally **cannot** find $c$ such that $P(X \geq c) = \alpha$ exactly.

. . .

Instead, we choose $c$ as the **smallest** value such that $P(X \geq c \mid p_0) \leq \alpha$. This means the actual significance level is often *less than* $\alpha$ --- the test is **conservative** (Devore 9.3).

## Computing the Exact P-Value

The p-value for the exact binomial test is the probability, under $H_0$, of observing a result as extreme or more extreme than the observed data:

| Alternative | Exact P-value |
|:---:|:---:|
| $H_a: p > p_0$ | $P(X \geq x_{obs} \mid p_0) = \sum_{k=x_{obs}}^{n}\binom{n}{k}p_0^k(1-p_0)^{n-k}$ |
| $H_a: p < p_0$ | $P(X \leq x_{obs} \mid p_0) = \sum_{k=0}^{x_{obs}}\binom{n}{k}p_0^k(1-p_0)^{n-k}$ |
| $H_a: p \neq p_0$ | $2 \cdot \min\left\{P(X \leq x_{obs}),\; P(X \geq x_{obs})\right\}$ |

. . .

In R, these binomial probabilities are straightforward to compute using `pbinom()`:

```{r}
#| echo: true
# Example: x = 3 successes out of n = 20, testing p0 = 0.30
x_ex <- 3; n_ex <- 20; p0_ex <- 0.30

# P(X <= 3 | p = 0.30) for lower-tailed test
p_lower <- pbinom(x_ex, size = n_ex, prob = p0_ex)

# P(X >= 3 | p = 0.30) for upper-tailed test
p_upper <- 1 - pbinom(x_ex - 1, size = n_ex, prob = p0_ex)

cat("P(X <= 3 | p=0.30) =", round(p_lower, 4), "\n")
cat("P(X >= 3 | p=0.30) =", round(p_upper, 4))
```

## Visualizing the Exact Binomial Distribution

```{r}
#| code-fold: true
#| fig-height: 6.5
# Show the binomial distribution under H0 and highlight the p-value region
x_vals <- 0:n_ex
probs <- dbinom(x_vals, size = n_ex, prob = p0_ex)

tibble(x = x_vals, prob = probs) |>
  mutate(region = case_when(
    x <= x_ex ~ "P-value region (X <= 3)",
    TRUE ~ "Not in p-value region"
  )) |>
  ggplot(aes(x = x, y = prob, fill = region)) +
  geom_col(width = 0.7, color = "black") +
  scale_fill_manual(values = c("P-value region (X <= 3)" = "coral", 
                                "Not in p-value region" = "steelblue")) +
  geom_vline(xintercept = n_ex * p0_ex, linetype = "dashed", linewidth = 1) +
  annotate("text", x = n_ex * p0_ex + 0.3, y = max(probs) * 0.9, hjust = 0, size = 5,
           label = bquote("E[X] =" ~ np[0] ~ "=" ~ .(n_ex * p0_ex))) +
  labs(x = "X (number of successes)", y = "P(X = x)",
       title = expression("Binomial(n = 20, " * p[0] * " = 0.30) under " * H[0]),
       subtitle = paste0("Exact p-value: P(X \u2264 3 | p = 0.30) = ", round(p_lower, 4)),
       fill = NULL) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "bottom")
```

## Step-by-Step Exact Calculation

Let's compute $P(X \leq 3 \mid p_0 = 0.30, n = 20)$ term by term:

```{r}
#| echo: true
# Each individual probability
for (k in 0:x_ex) {
  prob_k <- dbinom(k, size = n_ex, prob = p0_ex)
  cat(sprintf("P(X = %d) = C(20,%d) × 0.30^%d × 0.70^%d = %.6f\n", 
              k, k, k, n_ex - k, prob_k))
}

cat("\n--- Sum ---\n")
cat("P(X <= 3) =", round(sum(dbinom(0:x_ex, n_ex, p0_ex)), 6), "\n\n")
cat("Since", round(p_lower, 4), "> 0.05, we fail to reject H0")
```

## Using `binom.test()` in R

```{r}
#| echo: true
# The easy way: binom.test() does all this for us
binom.test(x = 3, n = 20, p = 0.30, alternative = "less")
```

## Understanding `binom.test()` Output

```{r}
#| echo: true
# Tidy output
binom_result <- binom.test(x = 3, n = 20, p = 0.30, alternative = "less")

cat("Test statistic (X):      ", binom_result$statistic, "\n")
cat("Exact p-value:           ", round(binom_result$p.value, 4), "\n")
cat("95% CI for p:            (", round(binom_result$conf.int[1], 4), ",", 
    round(binom_result$conf.int[2], 4), ")\n")
cat("Sample proportion (p-hat):", binom_result$estimate, "\n\n")

# Verify the p-value matches our manual calculation
cat("Manual P(X <= 3 | p=0.30):", round(pbinom(3, 20, 0.30), 4))
```

. . .

::: callout-note
## CI from `binom.test()`
The confidence interval from `binom.test()` is the **Clopper-Pearson** exact interval, which inverts the exact binomial test. It guarantees at least $100(1-\alpha)\%$ coverage but can be conservative (wider than necessary). This is different from the Score (Wilson) interval returned by `prop.test()`.
:::

## Medical Example: Rare Adverse Drug Reaction

A pharmaceutical company is testing whether a new drug's rate of a serious adverse reaction exceeds the known background rate of 2%. In a Phase II trial, 4 out of 80 patients experience the reaction.

```{r}
#| echo: true
# Data
x_adr <- 4; n_adr <- 80; p0_adr <- 0.02

cat("p-hat =", x_adr / n_adr, "\n")
cat("Check np0 =", n_adr * p0_adr, "< 10: Normal approx NOT valid!\n\n")

# Exact binomial test: H0: p = 0.02 vs Ha: p > 0.02
exact_result <- binom.test(x_adr, n_adr, p = p0_adr, alternative = "greater")
cat("Exact p-value:", round(exact_result$p.value, 4), "\n\n")

# Compare with (invalid) z test for illustration
z_invalid <- (x_adr/n_adr - p0_adr) / sqrt(p0_adr * (1 - p0_adr) / n_adr)
p_z_invalid <- 1 - pnorm(z_invalid)
cat("Z test p-value (INVALID, for comparison):", round(p_z_invalid, 4), "\n")
cat("Note the difference: exact and approximate p-values diverge\n")
cat("when the normal approximation conditions are not met.")
```

## Exact vs Approximate: When Does It Matter?

```{r}
#| code-fold: true
#| fig-height: 6
#| fig-width: 14
# Compare exact and approximate p-values across different x values
n_comp <- 80
p0_comp <- 0.02

comp_data <- tibble(x = 0:10) |>
  mutate(
    exact_pval = map_dbl(x, ~1 - pbinom(.x - 1, n_comp, p0_comp)),
    approx_pval = map_dbl(x, ~1 - pnorm((.x/n_comp - p0_comp) / 
                                          sqrt(p0_comp * (1 - p0_comp) / n_comp)))
  )

comp_data |>
  pivot_longer(c(exact_pval, approx_pval), 
               names_to = "method", values_to = "pvalue") |>
  mutate(method = ifelse(method == "exact_pval", "Exact (Binomial)", 
                         "Approximate (z test)")) |>
  ggplot(aes(x = x, y = pvalue, color = method, shape = method)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.05, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 8, y = 0.07, label = expression(alpha == 0.05), size = 5) +
  scale_color_manual(values = c("Exact (Binomial)" = "steelblue", 
                                 "Approximate (z test)" = "coral")) +
  labs(x = "X (number of adverse events)", 
       y = "Upper-tail P-value",
       title = bquote("Exact vs Approximate P-values: " ~ n == 80 * ", " ~ p[0] == 0.02),
       subtitle = bquote("Normal approximation unreliable when" ~ np[0] ~ "=" ~ 
                          .(n_comp * p0_comp) < 10),
       color = NULL, shape = NULL) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "bottom")
```

## When to Use Each Test

::: callout-tip
## Practical Guidelines for Proportion Tests

**Use `prop.test()` (score/chi-squared test) when:**

- $np_0 \geq 10$ **and** $n(1-p_0) \geq 10$
- You want the Score (Wilson) confidence interval
- Consistency with Lesson 8's interval estimation framework

**Use `binom.test()` (exact binomial) when:**

- $np_0 < 10$ or $n(1-p_0) < 10$ (small sample or rare event)
- You want guaranteed Type I error control
- The proportion is very close to 0 or 1
- You need the Clopper-Pearson exact CI
:::

## Devore Example 9.15: Obesity Prevalence

An earlier survey found 20% of Americans considered themselves obese. A recent study observed 686 obese individuals among 2014 adult men. Is the true obesity rate more than 1.5 times the self-assessment rate (i.e., more than 30%)?

```{r}
#| echo: true
# H0: p = 0.30 vs Ha: p > 0.30, alpha = 0.10
x_obs <- 686
n_obs <- 2014
p0_obs <- 0.30
alpha_obs <- 0.10

phat_obs <- x_obs / n_obs
se_obs <- sqrt(p0_obs * (1 - p0_obs) / n_obs)
z_obs <- (phat_obs - p0_obs) / se_obs

cat("p-hat =", round(phat_obs, 3), "\n")
cat("z =", round(z_obs, 2), "\n")
cat("z_{0.10} =", round(qnorm(1 - alpha_obs), 3), "\n")
cat("Decision:", ifelse(z_obs >= qnorm(1 - alpha_obs), 
                         "Reject H0", "Fail to reject H0"), "\n\n")
cat("Conclusion: At alpha = 0.10, more than 30% of American\n")
cat("adult men are obese.")
```

## Example: Clinical Trial 30-Day Readmission (C&H 8.3 Style)

**Setting:** A hospital targets a 30-day readmission rate at or below 15%. Among 180 discharged patients, 35 were readmitted within 30 days.

```{r}
#| echo: true
# Data
x_readmit <- 35
n_readmit <- 180
p0_readmit <- 0.15
phat_readmit <- x_readmit / n_readmit

cat("p-hat =", round(phat_readmit, 4), "\n")
cat("Check: np0 =", n_readmit * p0_readmit, 
    ", n(1-p0) =", n_readmit * (1 - p0_readmit), "\n\n")

# H0: p = 0.15 vs Ha: p > 0.15
# Manual z test (score test)
se_readmit <- sqrt(p0_readmit * (1 - p0_readmit) / n_readmit)
z_readmit <- (phat_readmit - p0_readmit) / se_readmit
cat("Manual z =", round(z_readmit, 3), "\n")
cat("Manual p-value =", round(1 - pnorm(z_readmit), 4), "\n")
```

## Readmission: Comparing All Three Approaches

```{r}
#| echo: true
# Score test via prop.test()
score_result <- prop.test(x_readmit, n_readmit, p = p0_readmit, 
                           alternative = "greater", correct = FALSE)

# Exact binomial test
exact_result <- binom.test(x_readmit, n_readmit, p = p0_readmit, 
                            alternative = "greater")

cat("=== Comparison of Methods ===\n\n")
cat("Manual z test:        p-value =", round(1 - pnorm(z_readmit), 4), "\n")
cat("prop.test (score):    p-value =", round(score_result$p.value, 4), "\n")
cat("binom.test (exact):   p-value =", round(exact_result$p.value, 4), "\n\n")

cat("Score (Wilson) CI:     (", round(score_result$conf.int[1], 4), ",", 
    round(score_result$conf.int[2], 4), ")\n")
cat("Clopper-Pearson CI:    (", round(exact_result$conf.int[1], 4), ",", 
    round(exact_result$conf.int[2], 4), ")\n\n")
cat("With large n, all three p-values agree closely.")
```

## Power and Sample Size for Proportion Tests

The formulas for $\beta(p')$ and required $n$ parallel the z test for means:

$$n = \begin{cases} \left[\frac{z_\alpha\sqrt{p_0 q_0} + z_\beta\sqrt{p' q'}}{p' - p_0}\right]^2 & \text{one-tailed} \\[10pt] \left[\frac{z_{\alpha/2}\sqrt{p_0 q_0} + z_\beta\sqrt{p' q'}}{p' - p_0}\right]^2 & \text{two-tailed (approx)} \end{cases}$$

where $q_0 = 1 - p_0$ and $q' = 1 - p'$.

```{r}
#| echo: true
# How many patients to detect a readmission rate of 0.20 with 80% power?
p0 <- 0.15; p_alt <- 0.20; alpha <- 0.05; beta_target <- 0.20

n_prop <- ((qnorm(1 - alpha) * sqrt(p0 * (1 - p0)) + 
             qnorm(1 - beta_target) * sqrt(p_alt * (1 - p_alt))) / 
            (p_alt - p0))^2

cat("Required sample size:", ceiling(n_prop), "patients")
```

## `power.prop.test()` in R

```{r}
#| echo: true
# R has a built-in function for proportion test sample size
power.prop.test(p1 = 0.15, p2 = 0.20, sig.level = 0.05, 
                power = 0.80, alternative = "one.sided")
```

. . .

::: callout-note
## Note
`power.prop.test()` is designed for **two-sample** tests by default. For one-sample tests, the sample size from our formula is a good approximation. For exact calculations, use the `pwr` package.
:::

# Putting It All Together {background-color="#16a085"}

## Decision Tree: Which Test to Use?

```{r}
#| code-fold: true
#| fig-height: 7
#| fig-width: 14
# Decision tree as a visualization
tree_data <- tibble(
  x = c(5, 2.5, 7.5, 1.25, 3.75, 6.25, 8.75),
  y = c(3, 2, 2, 1, 1, 1, 1),
  label = c(
    "What parameter\nare you testing?",
    "Testing\nthe mean (mu)",
    "Testing a\nproportion (p)",
    "sigma known?\nz test",
    "sigma unknown?\nt test",
    "Large sample?\nz test for p",
    "Small sample?\nExact binomial"
  ),
  fill_color = c("gray90", "lightblue", "lightyellow", 
                  "steelblue", "coral", "steelblue", "coral")
)

ggplot(tree_data, aes(x = x, y = y)) +
  geom_label(aes(label = label, fill = fill_color), 
             size = 5, label.padding = unit(0.6, "lines"),
             label.size = 0.5) +
  scale_fill_identity() +
  geom_segment(aes(x = 5, y = 2.75, xend = 2.5, yend = 2.25), 
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 5, y = 2.75, xend = 7.5, yend = 2.25),
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 2.5, y = 1.75, xend = 1.25, yend = 1.25),
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 2.5, y = 1.75, xend = 3.75, yend = 1.25),
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 7.5, y = 1.75, xend = 6.25, yend = 1.25),
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 7.5, y = 1.75, xend = 8.75, yend = 1.25),
               arrow = arrow(length = unit(0.2, "cm"))) +
  coord_cartesian(xlim = c(0, 10), ylim = c(0.5, 3.5)) +
  theme_void(base_size = 16) +
  labs(title = "Decision Tree: Choosing the Right Test")
```

## Summary of Test Procedures

| Test | Statistic | Null Distribution | R Function | Conditions |
|:---|:---:|:---:|:---:|:---|
| z test for $\mu$ | $\frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}}$ | $N(0,1)$ | (manual) | $\sigma$ known; normal pop. or large $n$ |
| t test for $\mu$ | $\frac{\bar{x} - \mu_0}{s/\sqrt{n}}$ | $t_{n-1}$ | `t.test()` | $\sigma$ unknown; normal pop. or large $n$ |
| Score test for $p$ | $\frac{\hat{p} - p_0}{\sqrt{p_0 q_0 / n}}$ | $N(0,1)$ or $\chi^2_1$ | `prop.test()` | $np_0 \geq 10$, $n q_0 \geq 10$ |
| Exact binomial | $X$ (count) | $\text{Binom}(n, p_0)$ | `binom.test()` | Always valid |

. . .

::: callout-tip
## Practical Advice

- In real medical research, $\sigma$ is almost **never** known --- use the **t test**
- The z test for means is primarily pedagogical (understanding the framework)
- For proportions, verify the validity condition before using the z test
- Always check assumptions (normality for small-sample t tests)
:::

## Complete Medical Example: Serum Creatinine Levels

A nephrology study examines whether a new drug changes serum creatinine levels. The normal reference is 1.0 mg/dL.

```{r}
#| echo: true
# Patient data
creatinine <- c(0.92, 1.15, 1.03, 0.88, 1.21, 0.97, 1.08, 1.14,
                0.95, 1.06, 1.18, 0.99, 1.12, 0.91, 1.07, 1.01,
                1.10, 0.94, 1.16, 1.03)

# Comprehensive analysis
cat("--- Summary Statistics ---\n")
cat("n =", length(creatinine), "\n")
cat("x-bar =", round(mean(creatinine), 4), "\n")
cat("s =", round(sd(creatinine), 4), "\n")
cat("se =", round(sd(creatinine)/sqrt(length(creatinine)), 4), "\n")
```

## Serum Creatinine: Full Analysis

```{r}
#| echo: true
# Two-tailed t test: H0: mu = 1.0 vs Ha: mu != 1.0
creat_test <- t.test(creatinine, mu = 1.0, alternative = "two.sided")
creat_test
```

```{r}
#| code-fold: true
#| fig-height: 5
#| fig-width: 14
p1 <- tibble(x = creatinine) |>
  ggplot(aes(x = x)) +
  geom_histogram(bins = 8, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = mean(creatinine), color = "darkgreen", linewidth = 1) +
  labs(x = "Serum Creatinine (mg/dL)", y = "Count",
       title = "Distribution of Creatinine Values") +
  annotate("text", x = 1.0, y = 5, label = expression(mu[0] == 1.0), 
           color = "red", hjust = 1.1, size = 5) +
  annotate("text", x = mean(creatinine), y = 4.5, hjust = -0.1, size = 5,
           color = "darkgreen",
           label = bquote(bar(x) == .(round(mean(creatinine), 3))))

p2 <- tibble(x = creatinine) |>
  ggplot(aes(sample = x)) +
  stat_qq(size = 3, color = "steelblue") +
  stat_qq_line(color = "red", linewidth = 1) +
  labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles")

p1 + p2
```

## Serum Creatinine: Interpreting Results

```{r}
#| echo: true
# Summary of results
cat("=== Test Results ===\n")
cat("t =", round(creat_test$statistic, 3), "\n")
cat("df =", creat_test$parameter, "\n")
cat("p-value =", round(creat_test$p.value, 4), "\n")
cat("95% CI: (", round(creat_test$conf.int[1], 4), ",", 
    round(creat_test$conf.int[2], 4), ")\n\n")

if (creat_test$p.value < 0.05) {
  cat("Interpretation: At alpha = 0.05, there IS statistically\n")
  cat("significant evidence that the mean serum creatinine level\n")
  cat("differs from the reference value of 1.0 mg/dL.\n\n")
  cat("The 95% CI does NOT contain 1.0 -- consistent with rejection.")
} else {
  cat("Interpretation: At alpha = 0.05, there is INSUFFICIENT\n")
  cat("evidence that the mean serum creatinine level differs\n")
  cat("from the reference value of 1.0 mg/dL.\n\n")
  cat("The 95% CI CONTAINS 1.0 -- consistent with failure to reject.")
}
```

## Simulation: Comparing z and t Tests

```{r}
#| echo: true
#| code-fold: true
set.seed(551)
n_sims <- 10000
n <- 12
mu0 <- 100
sigma <- 15

# When sigma is truly known
z_reject <- replicate(n_sims, {
  x <- rnorm(n, mean = mu0, sd = sigma)
  z <- (mean(x) - mu0) / (sigma / sqrt(n))
  abs(z) >= qnorm(0.975)
})

# Using t test (sigma estimated)
t_reject <- replicate(n_sims, {
  x <- rnorm(n, mean = mu0, sd = sigma)
  t <- (mean(x) - mu0) / (sd(x) / sqrt(n))
  abs(t) >= qt(0.975, df = n - 1)
})

cat("Simulated Type I error rates (alpha = 0.05):\n")
cat("z test (sigma known):   ", round(mean(z_reject), 4), "\n")
cat("t test (sigma estimated):", round(mean(t_reject), 4), "\n\n")
cat("Both correctly control Type I error at approximately 0.05")
```

# Key Takeaways and Looking Ahead {background-color="#2c3e50"}

## Key Takeaways

::: callout-important
## Today's Main Points

1. **z test for $\mu$:** Used when $\sigma$ is known (rare in practice); test statistic $Z \sim N(0,1)$ under $H_0$
2. **t test for $\mu$:** The workhorse test when $\sigma$ is unknown; test statistic $T \sim t_{n-1}$ under $H_0$
3. **Score test for $p$:** The z test for proportions **is** the score test; `prop.test()` computes $Z^2 \sim \chi^2_1$ and returns the Wilson/Score CI from Lesson 8
4. **Exact binomial:** Uses $P(X \leq x \mid p_0)$ directly from $\text{Binom}(n, p_0)$; always valid, especially when $np_0 < 10$
5. **CI-test duality:** A $100(1-\alpha)\%$ CI contains all parameter values not rejected at level $\alpha$
:::

. . .

**Common mistakes to avoid:**

- Using $\hat{p}$ instead of $p_0$ in the SE for the proportion test (this would be Wald, not score!)
- Using `prop.test()` when $np_0 < 10$ --- use `binom.test()` instead
- Applying the t test to very small samples from non-normal populations
- Saying "accept $H_0$" instead of "fail to reject $H_0$"

## Looking Ahead

**Next class (Lesson 14):** P-Values; Power and Sample Size Determination (Devore 9.2, 9.24)

- P-values as a continuous measure of evidence
- Power curves for z and t tests
- Sample size determination for planning clinical trials
- The noncentral t distribution

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Sections 9.2--9.3
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Sections 8.1--8.3
-   Mackowiak, P. A., Wasserman, S. S., & Levine, M. M. (1992). A critical appraisal of 98.6°F, the upper limit of the normal body temperature, and other legacies of Carl Reinhold August Wunderlich. *JAMA*, 268(12), 1578-1580.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
# For PDF export
decktape reveal --fragments docs/lessons/13_Tests_Mean_Proportion/13_Tests_Mean_Proportion.html lessons/13_Tests_Mean_Proportion/13_Tests_Mean_Proportion.pdf
```
