---
title: "Lesson 8: One-Sample t-Intervals and Population Proportions"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w4d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 8 - t-Intervals and Proportions
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 8: The Realistic Case {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 7: Confidence Intervals (Introduction)**

-   CI definition and correct interpretation
-   Pivotal quantities for constructing CIs
-   Z-intervals when $\sigma$ is known: $\bar{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$

. . .

**The Problem:** In practice, we **rarely know** the population standard deviation $\sigma$.

. . .

**Today's Goals:**

1. Construct t-intervals when $\sigma$ is unknown
2. Check normality assumptions
3. Build confidence intervals for population proportions
4. Compare Wald, Score, and Agresti-Coull intervals

# CI for Mean: $\sigma$ Unknown {background-color="#9b59b6"}

## The Realistic Case

In practice, we **rarely know** the population standard deviation $\sigma$.

. . .

**Natural idea:** Replace $\sigma$ with the sample standard deviation $s$.

. . .

But what happens to our pivotal quantity?

$$Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim N(0, 1)$$

becomes

$$T = \frac{\bar{X} - \mu}{S/\sqrt{n}} \sim \; ???$$

## The Source of Extra Variability

```{r}
#| echo: false
#| fig-height: 6
set.seed(551)
n_sim <- 10000
n <- 10
true_mu <- 0
true_sigma <- 1

# Simulate both Z and T statistics
sim_data <- tibble(
  Z = replicate(n_sim, {
    x <- rnorm(n, true_mu, true_sigma)
    (mean(x) - true_mu) / (true_sigma / sqrt(n))
  }),
  T = replicate(n_sim, {
    x <- rnorm(n, true_mu, true_sigma)
    (mean(x) - true_mu) / (sd(x) / sqrt(n))
  })
) |>
  pivot_longer(everything(), names_to = "statistic", values_to = "value")

sim_data |>
  ggplot(aes(x = value, fill = statistic)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("Z" = "steelblue", "T" = "coral"),
                    labels = c("T = (X̄ - μ)/(S/√n)", "Z = (X̄ - μ)/(σ/√n)")) +
  labs(x = "Value", y = "Count", 
       title = "Comparing Z and T Statistics (n = 10)",
       fill = "Statistic") +
  theme_minimal(base_size = 18)
```

**Key insight:** T has more variability because both $\bar{X}$ and $S$ are random!

## The t Distribution

```{r}
#| echo: false
#| fig-height: 6
tibble(x = seq(-4, 4, by = 0.01)) |>
  mutate(
    normal = dnorm(x),
    t_5 = dt(x, df = 5),
    t_15 = dt(x, df = 15),
    t_30 = dt(x, df = 30)
  ) |>
  pivot_longer(-x, names_to = "distribution", values_to = "density") |>
  mutate(distribution = factor(distribution, 
                               levels = c("normal", "t_30", "t_15", "t_5"),
                               labels = c("N(0,1)", "t(df=30)", "t(df=15)", "t(df=5)"))) |>
  ggplot(aes(x = x, y = density, color = distribution)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("black", "#3498db", "#e74c3c", "#27ae60")) +
  labs(x = "x", y = "Density", color = "Distribution",
       title = "Student's t Distribution vs. Standard Normal") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "right")
```

The t distribution has **heavier tails** than the normal — accounting for extra uncertainty from estimating $\sigma$.

## Student's t Distribution

::: callout-important
## Gosset's Theorem
If $X_1, \ldots, X_n \stackrel{iid}{\sim} N(\mu, \sigma^2)$, then:
$$T = \frac{\bar{X} - \mu}{S/\sqrt{n}} \sim t_{n-1}$$
where $t_{n-1}$ denotes the t distribution with $n-1$ **degrees of freedom**.
:::

. . .

**Key properties of $t_\nu$:**

-   Symmetric around 0
-   Heavier tails than $N(0, 1)$
-   As $\nu \to \infty$, $t_\nu \to N(0, 1)$

## Simulating the t Distribution (Chihara 7.1.2)

```{r}
#| echo: true
#| fig-height: 4.5
# Verify that (X̄ - μ)/(S/√n) follows t distribution
set.seed(551)
n <- 15
t_stats <- replicate(10000, {
  x <- rnorm(n, mean = 25, sd = 7)
  (mean(x) - 25) / (sd(x) / sqrt(n))
})

# Compare to theoretical t distribution
qqplot(qt(ppoints(10000), df = n - 1), t_stats, 
       main = "Q-Q Plot: Simulated vs. t(14)", 
       xlab = "Theoretical t(14)", ylab = "Simulated")
abline(0, 1, col = "red", lwd = 2)
```

## Why $(n-1)$ Degrees of Freedom?

The sample variance involves:
$$S^2 = \frac{1}{n-1}\sum_{i=1}^n(X_i - \bar{X})^2$$

. . .

Since $\sum(X_i - \bar{X}) = 0$, knowing any $n-1$ deviations determines the last one.

. . .

**Intuition:** We "use up" one degree of freedom by estimating $\mu$ with $\bar{X}$.

## The One-Sample t Interval

::: callout-important
## t Confidence Interval for Normal Mean ($\sigma$ Unknown)
If $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$, a $100(1-\alpha)\%$ CI for $\mu$ is:
$$\bar{x} \pm t_{\alpha/2, n-1} \cdot \frac{s}{\sqrt{n}}$$
where $t_{\alpha/2, n-1}$ is the critical value with $P(T > t_{\alpha/2, n-1}) = \alpha/2$ for $T \sim t_{n-1}$.
:::

. . .

The quantity $s/\sqrt{n}$ is called the **standard error of the mean** (SEM or SE).

## t Critical Values

```{r}
# Compare z and t critical values for 95% CI
z_crit <- qnorm(0.975)

df_values <- c(5, 10, 20, 30, 50, 100, Inf)
t_crits <- qt(0.975, df = df_values)

tibble(
  df = c("5", "10", "20", "30", "50", "100", "∞ (z)"),
  t_critical = round(c(head(t_crits, -1), z_crit), 3)
)
```

**Key insight:** For small samples, $t$ critical values are substantially larger than $z$ critical values.

## Visualizing z vs. t Critical Values

```{r}
#| echo: false
#| fig-height: 6
tibble(df = 2:100) |>
  mutate(t_crit = qt(0.975, df)) |>
  ggplot(aes(x = df, y = t_crit)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_hline(yintercept = qnorm(0.975), linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 80, y = 2.1, label = "z = 1.96", color = "red", size = 6) +
  labs(x = "Degrees of Freedom", y = "Critical Value (α = 0.05)",
       title = "t Critical Values Approach z as df → ∞") +
  theme_minimal(base_size = 18)
```

## Example: Porsche Odometer Readings (Devore Example 8.8)

A sample of $n = 16$ used Porsche Boxsters has odometer readings (miles):

```{r}
odometer <- c(80000, 30100, 97500, 58551, 73787, 51800, 69267, 44530,
              42192, 104920, 41442, 27418, 43436, 77219, 5991, 14362)

n <- length(odometer)
xbar <- mean(odometer)
s <- sd(odometer)

cat("Sample size:", n, "\n")
cat("Sample mean:", round(xbar, 1), "miles\n")
cat("Sample SD:", round(s, 1), "miles\n")
```

## Example: Porsche (continued)

**95% Confidence Interval:**

```{r}
alpha <- 0.05
t_crit <- qt(1 - alpha/2, df = n - 1)

cat("t critical value (df=15):", round(t_crit, 3), "\n")

se <- s / sqrt(n)
margin <- t_crit * se

cat("Standard error:", round(se, 1), "\n")
cat("Margin of error:", round(margin, 1), "\n\n")

cat("95% CI: (", round(xbar - margin, 1), ",", round(xbar + margin, 1), ")")
```

## Checking the Normality Assumption

The t interval assumes normally distributed data. **Always check!**

```{r}
#| fig-height: 5
par(mfrow = c(1, 2))
hist(odometer, main = "Histogram", xlab = "Odometer (miles)", col = "lightblue")
qqnorm(odometer, main = "Normal Q-Q Plot", pch = 19)
qqline(odometer, col = "red", lwd = 2)
```

The Q-Q plot looks reasonably linear — normality assumption is plausible.

## Medical Example: Systolic Blood Pressure

A clinical study measures systolic blood pressure in $n = 25$ patients after a new treatment:

```{r}
# Simulated BP data (mmHg)
bp <- c(118, 124, 131, 127, 122, 119, 128, 135, 121, 126,
        130, 124, 129, 133, 120, 125, 127, 132, 123, 128,
        126, 131, 122, 129, 125)

n <- length(bp)
xbar <- mean(bp)
s <- sd(bp)

# 95% CI
t_crit <- qt(0.975, df = n - 1)
se <- s / sqrt(n)
ci <- c(xbar - t_crit * se, xbar + t_crit * se)

cat("Sample mean:", round(xbar, 1), "mmHg\n")
cat("Sample SD:", round(s, 1), "mmHg\n")
cat("95% CI: (", round(ci[1], 1), ",", round(ci[2], 1), ") mmHg")
```

## Using t.test() in R

```{r}
# The easy way to compute t confidence intervals
t.test(bp, conf.level = 0.95)
```

. . .

**Note:** R's `t.test()` also performs a hypothesis test (which we'll cover later).

## Medical Example: Stress and Cortisol (Devore 8.37)

Medical students' serum cortisol levels were measured at the beginning and end of semester. For $n = 41$ students, the mean difference (end - beginning) was $\bar{d} = 2.08$ with $s_d = 7.88$.

**Find a 95% lower confidence bound for the population mean change:**

```{r}
n <- 41
dbar <- 2.08
s <- 7.88

t_alpha <- qt(0.95, df = n - 1)  # One-sided!
se <- s / sqrt(n)

lower_bound <- dbar - t_alpha * se

cat("95% lower confidence bound:", round(lower_bound, 2), "\n")
cat("Is the mean change necessarily positive? ", 
    ifelse(lower_bound > 0, "Yes", "No"))
```

## When Can We Trust the t-Interval?

::: callout-tip
## Guidelines for t-Interval Validity
1. **Large samples (n ≥ 30-40):** t-interval is robust to non-normality (by CLT)
2. **Moderate samples (15-30):** Check for severe skewness or outliers
3. **Small samples (n < 15):** Normality is crucial; use Q-Q plots
4. **Outliers:** Can severely affect $\bar{x}$ and $s$; consider alternatives
:::

. . .

**Alternatives for non-normal data:**

-   Bootstrap confidence intervals (Lesson 9)
-   Non-parametric methods (median-based)
-   Transformations (log, square root)

## Your Turn: Practice Problem

**Exercise:** A study of $n = 20$ hospital ICU stays yields $\bar{x} = 4.2$ days with $s = 1.8$ days.

1. Calculate a 95% CI for mean ICU stay
2. Calculate a 90% CI — how does it compare?
3. What assumptions are we making?

. . .

```{r}
n <- 20; xbar <- 4.2; s <- 1.8

# 95% CI
t95 <- qt(0.975, df = n - 1)
ci95 <- xbar + c(-1, 1) * t95 * s / sqrt(n)

# 90% CI  
t90 <- qt(0.95, df = n - 1)
ci90 <- xbar + c(-1, 1) * t90 * s / sqrt(n)

cat("95% CI:", round(ci95, 2), "\n")
cat("90% CI:", round(ci90, 2))
```

# Confidence Intervals for Proportions {background-color="#e67e22"}

## The Setup

**Situation:** We observe $X$ successes in $n$ independent Bernoulli trials.

-   $X \sim \text{Binomial}(n, p)$
-   Point estimate: $\hat{p} = X/n$

. . .

**Examples in medical research:**

-   Proportion of patients responding to treatment
-   Proportion experiencing adverse events
-   Disease prevalence in a population
-   Sensitivity/specificity of diagnostic tests

## The Wald Interval (Traditional)

For large $n$, $\hat{P} = X/n$ is approximately normal:
$$\hat{P} \approx N\left(p, \frac{p(1-p)}{n}\right)$$

. . .

**The Wald (traditional) 95% CI:**
$$\hat{p} \pm z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

. . .

::: callout-warning
## Problems with the Wald Interval
-   Poor coverage probability for $p$ near 0 or 1
-   Can produce bounds outside [0, 1]
-   Not recommended for small samples
:::

## Coverage Problems with Wald

```{r}
#| echo: false
#| fig-height: 6
# Simulate coverage probability for Wald interval
set.seed(551)
n_sim <- 10000
n <- 100

coverage_wald <- function(p, n, n_sim = 10000) {
  mean(replicate(n_sim, {
    x <- rbinom(1, n, p)
    p_hat <- x/n
    se <- sqrt(p_hat * (1 - p_hat) / n)
    lower <- p_hat - 1.96 * se
    upper <- p_hat + 1.96 * se
    lower < p & p < upper
  }))
}

p_values <- seq(0.05, 0.95, by = 0.05)
coverage <- sapply(p_values, coverage_wald, n = 100)

tibble(p = p_values, coverage = coverage) |>
  ggplot(aes(x = p, y = coverage)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red", linewidth = 1) +
  labs(x = "True Proportion p", y = "Coverage Probability",
       title = "Coverage Probability of 95% Wald Interval (n = 100)",
       subtitle = "Dashed line shows nominal 95% level") +
  theme_minimal(base_size = 18) +
  ylim(0.85, 1)
```

The Wald interval often has coverage **below** the nominal level!

## The Score (Wilson) Interval

A better approach: solve for $p$ in the pivotal quantity directly.

::: callout-important
## Score Confidence Interval for a Proportion
$$\frac{\hat{p} + \frac{z^2}{2n} \pm z\sqrt{\frac{\hat{p}(1-\hat{p})}{n} + \frac{z^2}{4n^2}}}{1 + \frac{z^2}{n}}$$
where $z = z_{\alpha/2}$.
:::

. . .

**Advantages:**

-   Coverage probability much closer to nominal level
-   Handles extreme proportions well
-   Always produces bounds in [0, 1]

## Deriving the Score Interval (Chihara 7.4)

Starting from the pivotal quantity:
$$Z = \frac{\hat{p} - p}{\sqrt{p(1-p)/n}} \approx N(0,1)$$

. . .

Set $|Z| = z_{\alpha/2}$ and solve for $p$:
$$\left(\frac{\hat{p} - p}{\sqrt{p(1-p)/n}}\right)^2 = z_{\alpha/2}^2$$

. . .

This leads to a **quadratic equation** in $p$, giving the score interval endpoints.

## The Agresti-Coull Interval

A simpler approximation that works well in practice:

::: callout-tip
## Agresti-Coull ("Add Two Successes and Two Failures")
Let $\tilde{n} = n + 4$ and $\tilde{p} = \frac{X + 2}{n + 4}$

Then use:
$$\tilde{p} \pm z_{\alpha/2}\sqrt{\frac{\tilde{p}(1-\tilde{p})}{\tilde{n}}}$$
:::

. . .

**Why it works:** Adding 2 successes and 2 failures "shrinks" extreme estimates toward 0.5, improving coverage.

## Comparing Intervals: Clinical Trial Example

In a Phase II trial, 23 of 40 patients respond to treatment.

```{r}
x <- 23; n <- 40; p_hat <- x/n
cat("Sample proportion:", p_hat, "\n\n")

# Wald interval
se_wald <- sqrt(p_hat * (1 - p_hat) / n)
wald_ci <- p_hat + c(-1, 1) * 1.96 * se_wald

# Agresti-Coull interval
n_tilde <- n + 4
p_tilde <- (x + 2) / n_tilde
se_ac <- sqrt(p_tilde * (1 - p_tilde) / n_tilde)
ac_ci <- p_tilde + c(-1, 1) * 1.96 * se_ac

cat("Wald 95% CI:          (", round(wald_ci[1], 3), ",", round(wald_ci[2], 3), ")\n")
cat("Agresti-Coull 95% CI: (", round(ac_ci[1], 3), ",", round(ac_ci[2], 3), ")")
```

## Using prop.test() for Score Interval

```{r}
# Score (Wilson) interval in R
prop.test(x = 23, n = 40, conf.level = 0.95, correct = FALSE)
```

## Comparing All Three Methods

```{r}
#| echo: true
compare_proportion_cis <- function(x, n, conf = 0.95) {
  p_hat <- x/n
  z <- qnorm(1 - (1-conf)/2)
  
  # Wald
  se_w <- sqrt(p_hat * (1-p_hat) / n)
  wald <- p_hat + c(-1, 1) * z * se_w
  
  # Agresti-Coull
  p_ac <- (x + 2) / (n + 4)
  se_ac <- sqrt(p_ac * (1-p_ac) / (n + 4))
  agresti <- p_ac + c(-1, 1) * z * se_ac
  
  # Score (Wilson)
  score <- prop.test(x, n, conf.level = conf, correct = FALSE)$conf.int
  
  tibble(Method = c("Wald", "Agresti-Coull", "Score"),
         Lower = c(wald[1], agresti[1], score[1]),
         Upper = c(wald[2], agresti[2], score[2]))
}

compare_proportion_cis(23, 40)
```

## Medical Example: Adverse Event Rate

In a safety study, 8 out of 200 patients experience a serious adverse event.

```{r}
x <- 8; n <- 200

# Score interval
result <- prop.test(x, n, conf.level = 0.95, correct = FALSE)
cat("Point estimate:", x/n, "=", round(x/n * 100, 1), "%\n")
cat("95% Score CI:", round(result$conf.int * 100, 2), "%")
```

. . .

**Clinical interpretation:** We are 95% confident that the true rate of serious adverse events is between 1.8% and 7.7%.

## Extreme Proportions: Why Wald Fails

```{r}
# What if only 2 of 50 patients have adverse events?
x <- 2; n <- 50; p_hat <- x/n

# Wald can give negative lower bound!
se_wald <- sqrt(p_hat * (1-p_hat) / n)
wald_ci <- p_hat + c(-1, 1) * 1.96 * se_wald

cat("Wald 95% CI:  (", round(wald_ci[1], 4), ",", round(wald_ci[2], 4), ")\n")

# Score interval stays in [0,1]
score_ci <- prop.test(x, n, conf.level = 0.95, correct = FALSE)$conf.int
cat("Score 95% CI: (", round(score_ci[1], 4), ",", round(score_ci[2], 4), ")")
```

The Wald interval gives a **negative** lower bound — impossible for a proportion!

## Example: Super Bowl Viewing (Devore 8.15)

In a sample of 25 students, **all 25** said they watched the Super Bowl. What's a 95% CI for the true proportion?

```{r}
x <- 25; n <- 25

# Wald gives silly result
p_hat <- x/n
se_wald <- sqrt(p_hat * (1-p_hat) / n)  # This is 0!
cat("Wald 95% CI: (1, 1) — just the point estimate!\n\n")

# Score interval is much better
score_ci <- prop.test(x, n, conf.level = 0.95, correct = FALSE)$conf.int
cat("Score 95% CI: (", round(score_ci[1], 3), ",", round(score_ci[2], 3), ")")
```

## Sample Size for Proportions

To achieve a margin of error $b$ with confidence level $1 - \alpha$:

$$n = \frac{z_{\alpha/2}^2 \cdot p(1-p)}{b^2}$$

. . .

**Problem:** We don't know $p$ before the study!

. . .

**Conservative approach:** Use $p = 0.5$ (maximizes $p(1-p) = 0.25$):
$$n = \frac{z_{\alpha/2}^2 \cdot 0.25}{b^2}$$

## Sample Size Example: Disease Prevalence

An epidemiologist wants to estimate disease prevalence with a margin of error ≤ 3% with 95% confidence.

```{r}
b <- 0.03  # margin of error
z <- qnorm(0.975)

# Conservative (p = 0.5)
n_conservative <- z^2 * 0.25 / b^2
cat("Conservative sample size:", ceiling(n_conservative), "\n")

# If we believe p ≈ 0.10 from prior studies
n_informed <- z^2 * 0.10 * 0.90 / b^2
cat("Informed sample size (p ≈ 0.10):", ceiling(n_informed))
```

. . .

Using prior information can **substantially** reduce required sample size!

## Sample Size: Effect of Assumptions

```{r}
#| echo: false
#| fig-height: 5
tibble(p = seq(0.05, 0.95, by = 0.01)) |>
  mutate(n = qnorm(0.975)^2 * p * (1-p) / 0.03^2) |>
  ggplot(aes(x = p, y = n)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_hline(yintercept = qnorm(0.975)^2 * 0.25 / 0.03^2, 
             linetype = "dashed", color = "red") +
  annotate("text", x = 0.8, y = 1100, label = "Conservative (p = 0.5)", color = "red", size = 5) +
  labs(x = "Assumed Proportion p", y = "Required Sample Size",
       title = "Sample Size for 95% CI with Margin ≤ 3%") +
  theme_minimal(base_size = 18)
```

## One-Sided Bounds for Proportions

**Upper confidence bound** (e.g., maximum adverse event rate):

```{r}
# 12 adverse events in 150 patients
x <- 12; n <- 150

# 95% upper bound using Score method
result <- prop.test(x, n, conf.level = 0.95, 
                    alternative = "less", correct = FALSE)
cat("Point estimate:", round(x/n * 100, 1), "%\n")
cat("95% upper bound:", round(result$conf.int[2] * 100, 1), "%")
```

. . .

**Interpretation:** We are 95% confident that the true adverse event rate is **at most** 12.9%.

# Practical Considerations {background-color="#16a085"}

## Summary: CI Methods Comparison

| Method | When to Use | Notes |
|--------|-------------|-------|
| Z-interval | Normal data, $\sigma$ known | Rarely applicable |
| t-interval | Normal data, $\sigma$ unknown | Check normality |
| Wald (proportion) | Large $n$, $p$ not extreme | Poor coverage |
| Score/Wilson | General | Best for proportions |
| Agresti-Coull | Quick calculation | Good approximation |

## CI Interpretation in Medical Research

::: callout-tip
## Valid Interpretations
-   "We are 95% confident that $\mu$ lies between L and U"
-   "The interval (L, U) is a plausible range for $\mu$"
-   "In repeated sampling, 95% of such intervals would contain $\mu$"
:::

. . .

::: callout-warning
## Invalid Interpretations
-   "There is a 95% probability that $\mu$ is in (L, U)" ✗
-   "95% of the data falls between L and U" ✗
-   "The true mean varies between L and U" ✗
:::

## Clinical Decision Making with CIs

**Example:** A drug reduces blood pressure by an estimated 12 mmHg with 95% CI (8, 16).

. . .

**Questions we can answer:**

-   Is the effect clinically meaningful? (If minimum important difference is 5 mmHg, yes!)
-   Is zero a plausible value? (No — CI excludes 0)
-   What's the worst-case scenario? (Lower bound suggests at least 8 mmHg effect)

## Relationship to Hypothesis Testing

::: callout-note
## Preview of Coming Material
A 95% CI contains all parameter values that would **not** be rejected by a two-sided test at $\alpha = 0.05$.

-   If CI excludes the null value → reject $H_0$
-   If CI includes the null value → fail to reject $H_0$
:::

. . .

**Example:** 95% CI for mean difference is (8, 16).

-   Test $H_0: \mu = 0$ → Reject (0 not in CI)
-   Test $H_0: \mu = 10$ → Fail to reject (10 in CI)

# Summary and Looking Ahead {background-color="#2c3e50"}

## Lesson 8 Summary

**Key Concepts:**

1.  **t-intervals** handle unknown $\sigma$: $\bar{x} \pm t_{\alpha/2, n-1}\frac{s}{\sqrt{n}}$

2.  **t distribution** has heavier tails than normal; approaches normal as df → ∞

3.  **Always check normality** for t-intervals (Q-Q plots, histograms)

4.  **Proportion intervals:** Score/Wilson preferred over Wald

5.  **Agresti-Coull** ("add 2 and 2") is a good simple approximation

## Key Formulas

| Parameter | CI Formula | Conditions |
|-----------|------------|------------|
| $\mu$ ($\sigma$ known) | $\bar{x} \pm z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$ | Normal data |
| $\mu$ ($\sigma$ unknown) | $\bar{x} \pm t_{\alpha/2, n-1}\frac{s}{\sqrt{n}}$ | Normal data |
| $p$ (Wald) | $\hat{p} \pm z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$ | Large $n$ |
| $p$ (Agresti-Coull) | $\tilde{p} \pm z_{\alpha/2}\sqrt{\frac{\tilde{p}(1-\tilde{p})}{\tilde{n}}}$ | General |

## Lesson 8 Practice Problems

1.  **(Devore 8.17)** For a sample with $n = 20$, $\bar{x} = 13.71$, $s = 3.55$:
    a. Compute a 95% CI for the population mean.
    b. What assumptions are required?

2.  **(Devore 8.37)** Medical students' cortisol levels changed by $\bar{d} = 2.08$ ($s = 7.88$, $n = 41$). Find a 95% lower bound for the mean change.

3.  **(Devore 8.43)** In a survey, 859 of 1044 adults support mandatory GMO labeling. Construct a 95% CI for the true proportion.

4.  **(Devore 8.48)** In 48 trials, 16 resulted in ignition.
    a. Compute the Score interval.
    b. Compute the Wald interval.
    c. Compare and discuss which is preferable.

## Lesson 8 Practice Problems (continued)

5.  **(Medical Application)** A clinical trial tests a new cholesterol medication. In $n = 48$ patients, the mean LDL reduction is $\bar{x} = 35$ mg/dL with $s = 18$ mg/dL.
    a. Construct a 95% CI for the mean LDL reduction.
    b. Construct a 99% CI. How does it compare?
    c. The company claims a mean reduction of at least 30 mg/dL. Is this claim supported?

6.  **(Sample Size)** How many subjects are needed to estimate the proportion of patients responding to a treatment within $\pm 5\%$ with 95% confidence, using the conservative approach?

7.  **(Chihara style)** Simulate 1000 Wald and Score intervals for $p = 0.1$ with $n = 50$. Compare the actual coverage probabilities.

## Looking Ahead

**Next Lesson: Bootstrap Confidence Intervals**

-   What if normality doesn't hold?
-   Bootstrap percentile intervals
-   Bootstrap t intervals
-   When to use bootstrap vs. classical methods

. . .

**Coming Soon:**

-   Hypothesis testing (Lessons 9-12)
-   Two-sample inference (Lesson 13)
-   Paired data analysis (Lesson 14)

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 8.1-8.4
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 7.
-   Agresti, A. and Coull, B. (1998). Approximate is better than "exact" for interval estimation of binomial proportions. *The American Statistician*, 52:119-126.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/08_t_Intervals_Proportions/08_t_Intervals_Proportions.html lessons/08_t_Intervals_Proportions/08_t_Intervals_Proportions.pdf
```
