---
title: "Lesson 17: Bootstrap Hypothesis Testing"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w8d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 17 - Bootstrap Hypothesis Testing
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 17: Bootstrap Hypothesis Testing {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 16: Further Aspects of Hypothesis Testing; CI-Test Duality**

- Statistical vs. clinical significance: always report effect sizes + CIs
- **Duality Theorem:** $\theta_0 \in$ CI $\iff$ fail to reject $H_0: \theta = \theta_0$
- Inverting a test gives a general CI construction method
- Equivalence and non-inferiority testing via TOST / one-sided bounds

. . .

**Today's Goals:**

1. Understand the conceptual framework for **bootstrap hypothesis testing** (C&H Ch. 8)
2. Implement the **one-sample bootstrap test** by resampling under $H_0$
3. Implement the **two-sample bootstrap test**
4. Understand **permutation tests** as an exact alternative to parametric tests
5. Compare parametric, bootstrap, and permutation tests: when does each apply?
6. Apply these methods to real medical data scenarios

## Roadmap

| Topic | Key Idea | Reference |
|-------|----------|-----------|
| Why bootstrap testing? | When parametric assumptions fail | C&H Ch. 8 |
| One-sample bootstrap test | Resample after shifting to $H_0$ | C&H 8.1 |
| The naive bootstrap (wrong) | Resampling without shifting gives invalid null | C&H 8.1 |
| Bootstrap p-value | Fraction of bootstrap statistics exceeding observed | C&H 8.1 |
| Bootstrap vs. t-test | Robustness for non-normal, small samples | C&H 8.1 |

. . .

::: callout-note
## Why This Matters
Parametric tests (z, t) require distributional assumptions that may not hold in practice. The bootstrap test provides a **distribution-free** alternative that is valid under much weaker conditions — essential for small samples, skewed outcomes, and novel test statistics in biomedical research.
:::

# The Bootstrap Hypothesis Test Framework (C&H Ch. 8) {background-color="#3498db"}

## Bootstrap CI vs. Bootstrap Test: Different Goals

We have used the bootstrap to **construct confidence intervals** (Lessons 9–10). Now we use it for **hypothesis testing**. The ideas are related but not identical.

. . .

| Bootstrap CI | Bootstrap Hypothesis Test |
|-------------|--------------------------|
| Resample from estimated distribution | Resample from distribution **under $H_0$** |
| Goal: capture sampling variability of $\hat{\theta}$ | Goal: approximate null distribution of test statistic |
| CI covers $\theta$ with $\approx 1-\alpha$ probability | Null distribution used to compute p-value |

. . .

::: callout-important
## Core Principle (C&H 8.1)
A bootstrap hypothesis test approximates the **null distribution** of the test statistic by resampling from a population that **satisfies $H_0$**. The p-value is the fraction of bootstrap statistics at least as extreme as the observed value.
:::

## The Bootstrap Test: General Algorithm

**Setup:** Observe $x_1, \ldots, x_n$. Test $H_0: \theta = \theta_0$ vs. $H_a: \theta \neq \theta_0$. Let $T = T(x_1,\ldots,x_n)$ be a test statistic.

. . .

**Algorithm:**

1. Compute the observed test statistic $T_{\text{obs}} = T(x_1, \ldots, x_n)$

2. **Shift the data** to satisfy $H_0$: define $x_i^* = x_i - \bar{x} + \theta_0$ (so the shifted sample has mean exactly $\theta_0$)

3. For $b = 1, \ldots, B$:
   - Draw a bootstrap sample $x_1^{*b}, \ldots, x_n^{*b}$ **with replacement** from $(x_1^*, \ldots, x_n^*)$
   - Compute $T^{*b} = T(x_1^{*b}, \ldots, x_n^{*b})$

4. Compute p-value:
$$\hat{p} = \frac{1}{B}\sum_{b=1}^B \mathbf{1}(|T^{*b}| \geq |T_{\text{obs}}|)$$

## Why Shift the Data?

The fundamental challenge: we observe data from the **true distribution** (with $\theta \neq \theta_0$ possibly), but we need to approximate the distribution of $T$ **under $H_0$**.

. . .

**Shifting:** By replacing $x_i$ with $x_i^* = x_i - \bar{x} + \theta_0$, we create a dataset that:

- Has the **same shape and variability** as the original
- Has **mean exactly equal to $\theta_0$** (satisfying $H_0: \mu = \theta_0$)

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14

set.seed(42)
n_shift     <- 20
x_obs_shift <- rexp(n_shift, rate = 0.2)  # right-skewed; true mean = 5
theta_0     <- 4  # null hypothesis value
x_shifted   <- x_obs_shift - mean(x_obs_shift) + theta_0

# Shared x-axis range across both panels
x_range <- range(c(x_obs_shift, x_shifted))
x_pad   <- diff(x_range) * 0.07
x_lims  <- x_range + c(-x_pad, x_pad)

# Compute shared density y-range for annotation placement
dens_orig  <- density(x_obs_shift)
dens_shift <- density(x_shifted)
y_ann_dens <- max(dens_orig$y, dens_shift$y) * 0.88

p_orig <- ggplot(tibble(x = x_obs_shift), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, fill = "#3498db", color = "white", alpha = 0.6) +
  geom_density(color = "#3498db", linewidth = 1.4) +
  geom_vline(xintercept = mean(x_obs_shift), color = "#e74c3c", linewidth = 1.5) +
  annotate("text", x = mean(x_obs_shift) + 0.5, y = y_ann_dens,
           label = sprintf("bar(x) == %.1f", round(mean(x_obs_shift), 1)),
           parse = TRUE, color = "#e74c3c", size = 5.5) +
  coord_cartesian(xlim = x_lims) +
  labs(title = "Original Data",
       subtitle = bquote("Mean = " * .(round(mean(x_obs_shift), 1)) *
                           "  \u2260  " * theta[0] * " = " * .(theta_0) *
                           "  (not satisfying " * H[0] * ")"),
       x = "x", y = "Density")

p_shift <- ggplot(tibble(x = x_shifted), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, fill = "#27ae60", color = "white", alpha = 0.6) +
  geom_density(color = "#27ae60", linewidth = 1.4) +
  geom_vline(xintercept = theta_0, color = "#e74c3c", linewidth = 1.5) +
  annotate("text", x = theta_0 + 0.5, y = y_ann_dens,
           label = sprintf("theta[0] == %.0f", theta_0),
           parse = TRUE, color = "#e74c3c", size = 5.5) +
  coord_cartesian(xlim = x_lims) +
  labs(title = expression("Shifted Data (mean = " * theta[0] * ")"),
       subtitle = bquote("Same shape & spread" * " \u2014 " *
                           "mean moved to " * theta[0] * " = " * .(theta_0)),
       x = expression(x^"*"), y = "Density")

p_orig + p_shift
```

# One-Sample Bootstrap Test (C&H 8.1) {background-color="#e74c3c"}

## Full Implementation: One-Sample Bootstrap Test

**Medical Context:** A hospital monitors C-reactive protein (CRP, mg/L) as an inflammation marker. The healthy adult population norm is mean CRP = 1.5 mg/L. We observe a small sample of post-surgical patients and test whether their mean CRP differs from this norm.

$$H_0: \mu = 1.5 \quad \text{vs} \quad H_a: \mu \neq 1.5$$

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

set.seed(551)
n_crp <- 22
# Gamma(shape=2, rate=1): mean = 2, SD ≈ 1.41 — right-skewed, moderate elevation
crp_data <- rgamma(n_crp, shape = 2, rate = 1)

mu_0 <- 1.5
B    <- 5000

# Observed test statistic (t-statistic)
T_obs <- (mean(crp_data) - mu_0) / (sd(crp_data) / sqrt(n_crp))

# Shift data to satisfy H0
crp_shifted <- crp_data - mean(crp_data) + mu_0

# Bootstrap
T_boot <- replicate(B, {
  boot_samp <- sample(crp_shifted, size = n_crp, replace = TRUE)
  (mean(boot_samp) - mu_0) / (sd(boot_samp) / sqrt(n_crp))
})

# Bootstrap p-value (two-sided)
p_boot <- mean(abs(T_boot) >= abs(T_obs))

# Parametric t-test p-value
p_param <- t.test(crp_data, mu = mu_0)$p.value

cat("Sample mean CRP:", round(mean(crp_data), 3), "mg/L\n")
cat("Sample SD:      ", round(sd(crp_data), 3),  "mg/L\n")
cat("Observed t-statistic:", round(T_obs, 4), "\n")
cat("Bootstrap p-value:   ", round(p_boot, 4), "\n")
cat("Parametric t-test p: ", round(p_param, 4), "\n")
```

## Visualizing the Bootstrap Null Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

t_df <- n_crp - 1
t_grid <- seq(-6, 6, by = 0.05)

p_boot_null <- ggplot(tibble(T = T_boot), aes(x = T)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#3498db", color = "white", alpha = 0.7) +
  geom_line(data = tibble(x = t_grid, y = dt(t_grid, df = t_df)),
            aes(x = x, y = y), color = "coral", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = c(-abs(T_obs), abs(T_obs)),
             color = "#e74c3c", linewidth = 1.5) +
  annotate("rect", xmin = abs(T_obs), xmax = Inf, ymin = 0, ymax = Inf,
           fill = "#e74c3c", alpha = 0.15) +
  annotate("rect", xmin = -Inf, xmax = -abs(T_obs), ymin = 0, ymax = Inf,
           fill = "#e74c3c", alpha = 0.15) +
  annotate("text", x = 4.5, y = 0.32,
           label = sprintf("p[boot] == %.4f", p_boot), parse = TRUE,
           color = "#e74c3c", size = 5) +
  scale_color_manual(values = c("Bootstrap" = "#3498db", "t(n-1)" = "coral")) +
  labs(
    title = "Bootstrap Null Distribution vs. t Distribution",
    subtitle = "Blue histogram = bootstrap; dashed red = t(n-1).\nRed shading = p-value region.",
    x = "t-statistic", y = "Density"
  )

# Also show original data vs shifted
p_data_compare <- bind_rows(
  tibble(x = crp_data, type = "Original data"),
  tibble(x = crp_shifted, type = "Shifted data (H0)")
) |>
  ggplot(aes(x = x, fill = type)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = mu_0, linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Original data" = "#3498db",
                               "Shifted data (H0)" = "#27ae60")) +
  labs(title = "Original vs. Shifted Data",
       subtitle = expression("Vertical line at " * mu[0] * " = 1.5 mg/L"),
       x = "CRP (mg/L)", y = "Density", fill = NULL) +
  theme(legend.position = "bottom")

p_data_compare + p_boot_null
```

## When Does the Bootstrap Test Outperform the t-Test?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 14

# Compare rejection rates under H0 (alpha control) for normal vs skewed distributions
sim_compare <- function(n, dist = "normal", B_inner = 999, reps = 500) {
  replicate(reps, {
    if (dist == "normal") {
      x <- rnorm(n, mean = 0)
    } else {
      x <- rexp(n, rate = 1) - 1  # centered exponential (mean=0, skewed)
    }
    
    # t-test p-value
    p_t <- t.test(x, mu = 0)$p.value
    
    # Bootstrap p-value
    x_sh <- x - mean(x)  # shift to H0: mu = 0
    T_ob <- (mean(x)) / (sd(x) / sqrt(n))
    T_bs <- replicate(B_inner, {
      bs <- sample(x_sh, n, replace = TRUE)
      (mean(bs)) / (sd(bs) / sqrt(n))
    })
    p_bs <- mean(abs(T_bs) >= abs(T_ob))
    
    c(p_t = p_t, p_bs = p_bs)
  }) |>
    t() |>
    as_tibble() |>
    summarise(
      alpha_t  = mean(p_t < 0.05),
      alpha_bs = mean(p_bs < 0.05),
      .groups = "drop"
    ) |>
    mutate(n = n, dist = dist)
}

set.seed(42)
grid <- expand_grid(n = c(10, 25, 50), dist = c("normal", "skewed"))
res <- map2_dfr(grid$n, grid$dist, sim_compare)

res |>
  pivot_longer(starts_with("alpha"), names_to = "method", values_to = "type1") |>
  mutate(
    method = recode(method, alpha_t = "t-test", alpha_bs = "Bootstrap test"),
    dist   = recode(dist, normal = "Normal population", skewed = "Skewed population (Exp-1)")
  ) |>
  ggplot(aes(x = factor(n), y = type1, fill = method)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black", linewidth = 1) +
  facet_wrap(~dist) +
  scale_fill_manual(values = c("t-test" = "steelblue", "Bootstrap test" = "#27ae60")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.12)) +
  labs(
    title = expression("Type I Error Rate Under " * H[0] * " at " * alpha * " = 0.05"),
    subtitle = "Both methods well-calibrated for normal; bootstrap slightly better for small skewed samples",
    x = "Sample size (n)", y = "Empirical Type I error rate",
    fill = NULL
  ) +
  theme(legend.position = "bottom")
```

## What Happens If You Don't Shift? (A Common Mistake)

It is tempting to simply bootstrap from the **original** data and compare to $\theta_0$. Why is this wrong?

. . .

::: callout-warning
## The Naive Bootstrap Test — Incorrect
Resample from the original $x_1, \ldots, x_n$ and compute $\bar{x}^{*b}$.  
Compare the fraction of $\bar{x}^{*b}$ farther from $\bar{x}$ than $|\bar{x} - \theta_0|$.  

**Problem:** The bootstrap distribution is centered at $\bar{x}$, not at $\theta_0$.  
It approximates the sampling distribution of $\hat{\theta}$, **not** the null distribution.
:::

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14

set.seed(551)
n_demo <- 22
x_demo <- rgamma(n_demo, shape = 2, rate = 1)   # true mean ≈ 2
mu_demo <- 1.5
B_demo  <- 4000

xbar_demo <- mean(x_demo)

# WRONG: resample from original, compare distance from xbar
T_naive <- replicate(B_demo, mean(sample(x_demo, n_demo, replace = TRUE)))
p_naive <- mean(abs(T_naive - xbar_demo) >= abs(xbar_demo - mu_demo))

# CORRECT: shift then resample
x_sh_demo <- x_demo - xbar_demo + mu_demo
T_correct <- replicate(B_demo, {
  bs <- sample(x_sh_demo, n_demo, replace = TRUE)
  (mean(bs) - mu_demo) / (sd(bs) / sqrt(n_demo))
})
T_obs_demo <- (xbar_demo - mu_demo) / (sd(x_demo) / sqrt(n_demo))
p_correct <- mean(abs(T_correct) >= abs(T_obs_demo))
p_ttest_demo <- t.test(x_demo, mu = mu_demo)$p.value

p_nv <- ggplot(tibble(T = T_naive), aes(x = T)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#e74c3c", color = "white", alpha = 0.8) +
  geom_vline(xintercept = xbar_demo, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = mu_demo,   color = "#3498db", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = xbar_demo + 0.05, y = 2.5,
           label = "bar(x)", parse = TRUE, size = 5, color = "black") +
  annotate("text", x = mu_demo - 0.05, y = 2.5,
           label = "mu[0]", parse = TRUE, size = 5, color = "#3498db", hjust = 1) +
  labs(title = expression(bold("WRONG") * ": Naive bootstrap"),
       subtitle = sprintf("Centered at x-bar, not theta0 \u2014 p = %.3f (not a valid null distribution)",
                          p_naive),
       x = expression(bar(x)^"*"), y = "Density")

p_cr <- ggplot(tibble(T = T_correct), aes(x = T)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#27ae60", color = "white", alpha = 0.8) +
  geom_vline(xintercept = c(-abs(T_obs_demo), abs(T_obs_demo)),
             color = "#e74c3c", linewidth = 1.2) +
  annotate("text", x = abs(T_obs_demo) + 0.15, y = 0.32,
           label = bquote(T[obs] == .(round(T_obs_demo, 2))),
           color = "#e74c3c", size = 4.5) +
  labs(title = expression(bold("CORRECT") * ": Bootstrap from shifted data"),
       subtitle = sprintf("Centered at theta0 \u2014 Bootstrap p = %.3f,  t-test p = %.3f",
                          p_correct, p_ttest_demo),
       x = "t-statistic", y = "Density")

p_nv + p_cr
```

# Choosing Between Methods: One-Sample Tests {background-color="#f39c12"}

## Decision Framework

For a **one-sample** test $H_0: \mu = \mu_0$, which method should you use?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 13
library(DiagrammeR)

grViz("
digraph test_choice {
  graph [rankdir=TB, bgcolor=transparent, fontsize=18]
  node [shape=box, style=filled, fontsize=14, fontname='Helvetica']

  A [label='One-Sample Test\nH0: mu = mu0', fillcolor='#3498db', fontcolor=white, width=4]

  B [label='Normal population\nor large n (CLT)?', fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3.5]

  C [label='One-sample t-test\n(Most powerful for normal data)', fillcolor='#27ae60', fontcolor=white, width=3.5]

  D [label='iid observations?', fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3]

  E [label='Bootstrap Test\nShift to mu0, resample', fillcolor='#e74c3c', fontcolor=white, width=3.5]

  F [label='Non-iid structure present:\nConsider GEE, mixed models,\nor time series methods', fillcolor='#95a5a6', fontcolor=white, width=4]

  A -> B
  B -> C [label='  Yes', fontsize=13]
  B -> D [label='  No', fontsize=13]
  D -> E [label='  Yes', fontsize=13]
  D -> F [label='  No', fontsize=12]
}
")
```

## Assumptions Compared

| Assumption | t-test | Bootstrap test |
|-----------|--------|----------------|
| Normal population (or large $n$) | Required | Not required |
| iid observations | Required | Required |
| Distribution-free validity | No | Approximately |
| Works for any test statistic | No | **Yes** |
| Computationally intensive | No | Moderate |

. . .

::: callout-tip
## Key Message
The **bootstrap test** is the most flexible choice for one-sample problems with non-normal, skewed data — it only requires iid sampling and makes no symmetry assumption. The **t-test** is optimal when normality holds. Permutation methods will be covered after two-sample inference.
:::

## Simulation Study: Comparing t-test and Bootstrap

How well does each method control Type I error, and what is the power, across different population shapes?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

run_onesample_tests <- function(n = 20, delta = 0, dist = "normal", B_inner = 499) {
  x <- switch(dist,
    normal  = rnorm(n, mean = delta),
    skewed  = rexp(n, rate = 1) - 1 + delta,
    heavyt  = rt(n, df = 3) + delta
  )
  mu_0 <- 0

  # 1. Parametric t-test
  p_t <- t.test(x, mu = mu_0)$p.value

  # 2. Bootstrap test (shift-resample)
  x_sh <- x - mean(x) + mu_0
  T_ob <- (mean(x) - mu_0) / (sd(x) / sqrt(n))
  T_bs <- replicate(B_inner, {
    bs <- sample(x_sh, n, replace = TRUE)
    (mean(bs) - mu_0) / (sd(bs) / sqrt(n))
  })
  p_b <- mean(abs(T_bs) >= abs(T_ob))

  c(p_t = p_t, p_boot = p_b)
}

set.seed(99)
sim_grid <- expand_grid(
  n     = c(15, 40),
  delta = c(0, 0.6),
  dist  = c("normal", "skewed", "heavyt")
)

sim_res <- sim_grid |>
  rowwise() |>
  mutate(
    sims = list(
      replicate(500, run_onesample_tests(n = n, delta = delta, dist = dist,
                                        B_inner = 299)) |>
        t() |> as_tibble()
    )
  ) |>
  unnest(sims) |>
  pivot_longer(starts_with("p_"), names_to = "method", values_to = "pval") |>
  group_by(n, delta, dist, method) |>
  summarise(rejection = mean(pval < 0.05), .groups = "drop") |>
  mutate(
    method   = recode(method, p_t = "t-test", p_boot = "Bootstrap"),
    dist_lab = recode(dist,
                      normal = "Normal",
                      skewed = "Skewed (Exp-1)",
                      heavyt = "Heavy-tailed (t3)"),
    quantity = ifelse(delta == 0,
                      "Type I Error  (H0 true, target = 5%)",
                      "Power  (Ha true, delta = 0.6)")
  )

ggplot(sim_res |> filter(n == 15),
       aes(x = method, y = rejection, fill = method)) +
  geom_col(alpha = 0.85, width = 0.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed", linewidth = 0.8) +
  facet_grid(quantity ~ dist_lab, scales = "free_y") +
  scale_fill_manual(values = c("t-test" = "steelblue", "Bootstrap" = "#e74c3c")) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "t-test vs. Bootstrap: Type I Error and Power  (n = 15)",
    subtitle = "Bootstrap maintains better Type I error control for skewed/heavy-tailed data at small n",
    x = NULL, y = "Rejection rate", fill = NULL
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12),
        strip.text.y = element_text(size = 11),
        plot.margin = margin(5, 20, 5, 5))
```

## Simulation: Larger Sample (n = 40)

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

ggplot(sim_res |> filter(n == 40),
       aes(x = method, y = rejection, fill = method)) +
  geom_col(alpha = 0.85, width = 0.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed", linewidth = 0.8) +
  facet_grid(quantity ~ dist_lab, scales = "free_y") +
  scale_fill_manual(values = c("t-test" = "steelblue", "Bootstrap" = "#e74c3c")) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "t-test vs. Bootstrap: Type I Error and Power  (n = 40)",
    subtitle = "By n = 40, both methods converge — CLT smooths out distributional differences",
    x = NULL, y = "Rejection rate", fill = NULL
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12),
        strip.text.y = element_text(size = 11),
        plot.margin = margin(5, 20, 5, 5))
```

## Key Takeaways from Simulation

- **Normal data:** Both methods control Type I error well at $n = 15$ and $n = 40$; t-test has slightly higher power (it's optimal here)

. . .

- **Skewed data, small $n$:** The t-test can be slightly anti-conservative (Type I error > 5%); the bootstrap test maintains better control

. . .

- **Heavy-tailed data ($t_3$):** Similar pattern — bootstrap is more robust for small samples

. . .

- **Large $n$:** Differences wash out — the CLT rescues the t-test regardless of population shape

. . .

::: callout-note
## Practical Implication
For small samples ($n < 30$) with non-normal data, the bootstrap test is preferable. For $n \geq 40$, the t-test is nearly equivalent in practice.
:::

# Bootstrap Tests and the Duality with CIs {background-color="#16a085"}

## Bootstrap Tests and Bootstrap CIs: Are They Dual?

In the parametric setting, confidence intervals and tests are **exactly dual**. For bootstrap methods this is only **approximate** — and only when the CI and test use the **same pivot**.

::: callout-important
## The Key Condition
- **Percentile CI** resamples from the original data and uses $\bar{X}^*$ as the pivot
- **Bootstrap test** resamples from the shifted data and uses some test statistic $T^*$
- If both use the **same pivot** (e.g., raw mean), duality is approximate with only boundary disagreements from Monte Carlo noise
- If they use **different pivots** (e.g., percentile CI with raw mean vs. test with standardised $t$), they can disagree systematically over a wide range
:::

## Demonstrating: Same Pivot vs. Different Pivots

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 15

set.seed(42)
# Right-skewed small sample: CRP values (mg/L)
crp_small <- c(1.2, 1.5, 2.3, 1.8, 4.7, 3.1, 1.4, 6.2, 2.0, 1.6)
n_s   <- length(crp_small)
B_s   <- 9999
xbar_s <- mean(crp_small)

# Percentile CI (pivot: raw mean from original data)
boot_means <- replicate(B_s, mean(sample(crp_small, n_s, replace = TRUE)))
ci_lo <- quantile(boot_means, 0.025)
ci_hi <- quantile(boot_means, 0.975)

mu0_grid <- seq(0.8, 4.8, by = 0.05)

scan_results <- map_dfr(mu0_grid, function(mu0) {
  x_sh <- crp_small - xbar_s + mu0
  in_ci <- mu0 >= ci_lo & mu0 <= ci_hi

  # Same pivot: raw mean difference (matches what percentile CI inverts)
  T_ob_raw <- xbar_s - mu0
  T_bs_raw <- replicate(B_s, mean(sample(x_sh, n_s, replace = TRUE)) - mu0)
  p_raw <- mean(abs(T_bs_raw) >= abs(T_ob_raw))

  # Different pivot: standardised t-statistic
  T_ob_t <- (xbar_s - mu0) / (sd(crp_small) / sqrt(n_s))
  T_bs_t <- replicate(B_s, {
    bs <- sample(x_sh, n_s, replace = TRUE)
    (mean(bs) - mu0) / (sd(bs) / sqrt(n_s))
  })
  p_t <- mean(abs(T_bs_t) >= abs(T_ob_t))

  tibble(mu0      = mu0,
         in_ci    = in_ci,
         p_raw    = p_raw,
         p_t      = p_t,
         agree_raw = in_ci == (p_raw >= 0.05),
         agree_t   = in_ci == (p_t   >= 0.05))
})

make_panel <- function(data, p_col, agree_col, pivot_label) {
  ggplot(data, aes(x = mu0, y = .data[[p_col]])) +
    annotate("rect", xmin = ci_lo, xmax = ci_hi, ymin = -Inf, ymax = Inf,
             fill = "#16a085", alpha = 0.12) +
    geom_vline(xintercept = c(ci_lo, ci_hi),
               color = "#16a085", linewidth = 1.0, linetype = "dotted") +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "gray40") +
    geom_line(aes(color = .data[[agree_col]]), linewidth = 1.6) +
    geom_point(data = filter(data, !.data[[agree_col]]),
               aes(x = mu0, y = .data[[p_col]]),
               color = "#e74c3c", size = 3) +
    scale_color_manual(values = c("TRUE" = "#006a4e", "FALSE" = "#e74c3c"),
                       labels = c("TRUE" = "Agree", "FALSE" = "Disagree")) +
    annotate("text", x = mean(c(ci_lo, ci_hi)), y = 0.88,
             label = sprintf("95%% CI [%.2f, %.2f]", ci_lo, ci_hi),
             color = "#16a085", size = 5) +
    labs(title = pivot_label,
         x = expression(mu[0]),
         y = "Bootstrap test p-value",
         color = NULL) +
    theme(legend.position = "bottom",
          plot.title  = element_text(size = 15, lineheight = 1.2),
          axis.title  = element_text(size = 14),
          axis.text   = element_text(size = 13),
          legend.text = element_text(size = 13))
}

p_raw_panel <- make_panel(scan_results, "p_raw", "agree_raw",
  "Same pivot: raw mean\n(percentile CI approx. dual to bootstrap test)")
p_t_panel   <- make_panel(scan_results, "p_t", "agree_t",
  "Different pivots: percentile CI (raw mean)\nvs. bootstrap test (standardised t)")

p_raw_panel + p_t_panel +
  plot_annotation(
    title    = "Bootstrap CI-Test Duality: Matching vs. Mismatched Pivots",
    subtitle = "Green region = 95% percentile CI. Red points = mu0 where CI and test give contradictory decisions.",
    theme = theme(plot.title    = element_text(size = 17),
                  plot.subtitle = element_text(size = 14))
  )
```

. . .

::: callout-warning
## Key Lesson
**Left panel:** Using the same pivot (raw mean), disagreements are rare and only occur right at the CI boundary due to Monte Carlo variability — duality is approximately satisfied.  
**Right panel:** Using different pivots (percentile CI vs. standardised $t$-test), disagreements are systematic and spread over a range — this is a methodological mismatch, not boundary noise.
:::

# Key Takeaways {background-color="#2c3e50"}

## Key Takeaways

::: callout-important
## Today's Main Points (C&H Chapter 8)

1. **Bootstrap hypothesis tests** approximate the null distribution by resampling from data **shifted to satisfy $H_0$** — this is fundamentally different from bootstrap CIs (which resample from the original data)

2. **The naive bootstrap is wrong:** resampling from the original data without shifting produces a distribution centered at $\bar{x}$, not at $\mu_0$ — it does not represent the null distribution

3. **One-sample bootstrap test:** shift data so $\bar{x}^* = \theta_0$, resample, compute the t-statistic; the p-value is the fraction of bootstrap statistics as extreme as observed

4. **Bootstrap vs. t-test:** the bootstrap requires only iid sampling and no distributional assumptions; benefits are largest at small $n$ with skewed or heavy-tailed data; both converge for large $n$

5. **Duality is approximate, not exact:** the bootstrap percentile CI and bootstrap t-test use different pivots, so near CI boundaries with small, skewed samples they can give contradictory decisions — reject by one criterion but not the other
:::

## Looking Ahead

**Next class: Two-Sample z and t Tests; Confidence Intervals**

- Two-sample z-test for means ($\sigma$ known)
- Welch t-test for means ($\sigma$ unknown)
- Pooled t-test (equal variance assumption)
- Two-sample inference for proportions

. . .

**Later: Permutation and Bootstrap Tests for Two Samples (C&H Ch. 8)**

- Label-permutation tests for two independent groups
- One-sample sign-flip permutation test
- Bootstrap two-sample test

## References

::: nonincremental
-   Chihara, L. M. and Hesterberg, T. C. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 8: Bootstrap Hypothesis Testing.
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 9.6.
-   Efron, B. and Hastie, T. (2016). *Computer Age Statistical Inference*. Cambridge University Press. Chapter 15.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
# For PDF export
decktape reveal --fragments docs/lessons/17_BootstrapHypothesisTesting/17_BootstrapHypothesisTesting.html lessons/17_BootstrapHypothesisTesting/17_BootstrapHypothesisTesting.pdf
```
