---
title: "Lesson 17: Bootstrap Hypothesis Testing; Permutation Tests"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w9d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 17 - Bootstrap Hypothesis Testing; Permutation Tests
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 17: Bootstrap Hypothesis Testing {background-color="#2c3e50"}

## Review: Where We Left Off

**Lesson 16: Further Aspects of Hypothesis Testing; CI-Test Duality**

- Statistical vs. clinical significance: always report effect sizes + CIs
- **Duality Theorem:** $\theta_0 \in$ CI $\iff$ fail to reject $H_0: \theta = \theta_0$
- Inverting a test gives a general CI construction method
- Equivalence and non-inferiority testing via TOST / one-sided bounds

. . .

**Today's Goals:**

1. Understand the conceptual framework for **bootstrap hypothesis testing** (C&H Ch. 8)
2. Implement the **one-sample bootstrap test** by resampling under $H_0$
3. Implement the **two-sample bootstrap test**
4. Understand **permutation tests** as an exact alternative to parametric tests
5. Compare parametric, bootstrap, and permutation tests: when does each apply?
6. Apply these methods to real medical data scenarios

## Roadmap

| Topic | Key Idea | Reference |
|-------|----------|-----------|
| Why bootstrap testing? | When parametric assumptions fail | C&H Ch. 8 |
| One-sample bootstrap test | Resample after shifting to $H_0$ | C&H 8.1 |
| Two-sample bootstrap test | Pool, then resample under $H_0$ | C&H 8.2 |
| Permutation tests | Permute group labels; exact under exchangeability | C&H 8.3 |
| Bootstrap p-value | Fraction of bootstrap statistics exceeding observed | C&H 8.1 |
| Comparison of methods | Validity and power across situations | C&H 8.5 |

. . .

::: callout-note
## Why This Matters
Parametric tests (z, t, F) require distributional assumptions that may not hold in practice. Bootstrap and permutation tests provide **distribution-free** alternatives that are valid under much weaker conditions — essential for small samples, skewed outcomes, and novel statistics in biomedical research.
:::

# The Bootstrap Hypothesis Test Framework (C&H Ch. 8) {background-color="#3498db"}

## Bootstrap CI vs. Bootstrap Test: Different Goals

We have used the bootstrap to **construct confidence intervals** (Lessons 9–10). Now we use it for **hypothesis testing**. The ideas are related but not identical.

. . .

| Bootstrap CI | Bootstrap Hypothesis Test |
|-------------|--------------------------|
| Resample from estimated distribution | Resample from distribution **under $H_0$** |
| Goal: capture sampling variability of $\hat{\theta}$ | Goal: approximate null distribution of test statistic |
| CI covers $\theta$ with $\approx 1-\alpha$ probability | Null distribution used to compute p-value |

. . .

::: callout-important
## Core Principle (C&H 8.1)
A bootstrap hypothesis test approximates the **null distribution** of the test statistic by resampling from a population that **satisfies $H_0$**. The p-value is the fraction of bootstrap statistics at least as extreme as the observed value.
:::

## The Bootstrap Test: General Algorithm

**Setup:** Observe $x_1, \ldots, x_n$. Test $H_0: \theta = \theta_0$ vs. $H_a: \theta \neq \theta_0$. Let $T = T(x_1,\ldots,x_n)$ be a test statistic.

. . .

**Algorithm:**

1. Compute the observed test statistic $T_{\text{obs}} = T(x_1, \ldots, x_n)$

2. **Shift the data** to satisfy $H_0$: define $x_i^* = x_i - \bar{x} + \theta_0$ (so the shifted sample has mean exactly $\theta_0$)

3. For $b = 1, \ldots, B$:
   - Draw a bootstrap sample $x_1^{*b}, \ldots, x_n^{*b}$ **with replacement** from $(x_1^*, \ldots, x_n^*)$
   - Compute $T^{*b} = T(x_1^{*b}, \ldots, x_n^{*b})$

4. Compute p-value:
$$\hat{p} = \frac{1}{B}\sum_{b=1}^B \mathbf{1}(|T^{*b}| \geq |T_{\text{obs}}|)$$

## Why Shift the Data?

The fundamental challenge: we observe data from the **true distribution** (with $\theta \neq \theta_0$ possibly), but we need to approximate the distribution of $T$ **under $H_0$**.

. . .

**Shifting:** By replacing $x_i$ with $x_i^* = x_i - \bar{x} + \theta_0$, we create a dataset that:

- Has the **same shape and variability** as the original
- Has **mean exactly equal to $\theta_0$** (satisfying $H_0: \mu = \theta_0$)

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14

set.seed(42)
n_shift     <- 20
x_obs_shift <- rexp(n_shift, rate = 0.2)  # right-skewed; true mean = 5
theta_0     <- 4  # null hypothesis value
x_shifted   <- x_obs_shift - mean(x_obs_shift) + theta_0

# Shared x-axis range across both panels
x_range <- range(c(x_obs_shift, x_shifted))
x_pad   <- diff(x_range) * 0.07
x_lims  <- x_range + c(-x_pad, x_pad)

# Compute shared density y-range for annotation placement
dens_orig  <- density(x_obs_shift)
dens_shift <- density(x_shifted)
y_ann_dens <- max(dens_orig$y, dens_shift$y) * 0.88

p_orig <- ggplot(tibble(x = x_obs_shift), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, fill = "#3498db", color = "white", alpha = 0.6) +
  geom_density(color = "#3498db", linewidth = 1.4) +
  geom_vline(xintercept = mean(x_obs_shift), color = "#e74c3c", linewidth = 1.5) +
  annotate("text", x = mean(x_obs_shift) + 0.5, y = y_ann_dens,
           label = sprintf("bar(x) == %.1f", round(mean(x_obs_shift), 1)),
           parse = TRUE, color = "#e74c3c", size = 5.5) +
  coord_cartesian(xlim = x_lims) +
  labs(title = "Original Data",
       subtitle = bquote("Mean = " * .(round(mean(x_obs_shift), 1)) *
                           "  \u2260  " * theta[0] * " = " * .(theta_0) *
                           "  (not satisfying " * H[0] * ")"),
       x = "x", y = "Density")

p_shift <- ggplot(tibble(x = x_shifted), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10, fill = "#27ae60", color = "white", alpha = 0.6) +
  geom_density(color = "#27ae60", linewidth = 1.4) +
  geom_vline(xintercept = theta_0, color = "#e74c3c", linewidth = 1.5) +
  annotate("text", x = theta_0 + 0.5, y = y_ann_dens,
           label = sprintf("theta[0] == %.0f", theta_0),
           parse = TRUE, color = "#e74c3c", size = 5.5) +
  coord_cartesian(xlim = x_lims) +
  labs(title = expression("Shifted Data (mean = " * theta[0] * ")"),
       subtitle = bquote("Same shape & spread" * " \u2014 " *
                           "mean moved to " * theta[0] * " = " * .(theta_0)),
       x = expression(x^"*"), y = "Density")

p_orig + p_shift
```

# One-Sample Bootstrap Test (C&H 8.1) {background-color="#e74c3c"}

## Full Implementation: One-Sample Bootstrap Test

**Medical Context:** A hospital monitors C-reactive protein (CRP, mg/L) as an inflammation marker. The healthy adult population norm is mean CRP = 1.5 mg/L. We observe a small sample of post-surgical patients and test whether their mean CRP differs from this norm.

$$H_0: \mu = 1.5 \quad \text{vs} \quad H_a: \mu \neq 1.5$$

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

set.seed(551)
n_crp <- 22
# Gamma(shape=2, rate=1): mean = 2, SD ≈ 1.41 — right-skewed, moderate elevation
crp_data <- rgamma(n_crp, shape = 2, rate = 1)

mu_0 <- 1.5
B    <- 5000

# Observed test statistic (t-statistic)
T_obs <- (mean(crp_data) - mu_0) / (sd(crp_data) / sqrt(n_crp))

# Shift data to satisfy H0
crp_shifted <- crp_data - mean(crp_data) + mu_0

# Bootstrap
T_boot <- replicate(B, {
  boot_samp <- sample(crp_shifted, size = n_crp, replace = TRUE)
  (mean(boot_samp) - mu_0) / (sd(boot_samp) / sqrt(n_crp))
})

# Bootstrap p-value (two-sided)
p_boot <- mean(abs(T_boot) >= abs(T_obs))

# Parametric t-test p-value
p_param <- t.test(crp_data, mu = mu_0)$p.value

cat("Sample mean CRP:", round(mean(crp_data), 3), "mg/L\n")
cat("Sample SD:      ", round(sd(crp_data), 3),  "mg/L\n")
cat("Observed t-statistic:", round(T_obs, 4), "\n")
cat("Bootstrap p-value:   ", round(p_boot, 4), "\n")
cat("Parametric t-test p: ", round(p_param, 4), "\n")
```

## Visualizing the Bootstrap Null Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

t_df <- n_crp - 1
t_grid <- seq(-6, 6, by = 0.05)

p_boot_null <- ggplot(tibble(T = T_boot), aes(x = T)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#3498db", color = "white", alpha = 0.7) +
  geom_line(data = tibble(x = t_grid, y = dt(t_grid, df = t_df)),
            aes(x = x, y = y), color = "coral", linewidth = 1.3, linetype = "dashed") +
  geom_vline(xintercept = c(-abs(T_obs), abs(T_obs)),
             color = "#e74c3c", linewidth = 1.5) +
  annotate("rect", xmin = abs(T_obs), xmax = Inf, ymin = 0, ymax = Inf,
           fill = "#e74c3c", alpha = 0.15) +
  annotate("rect", xmin = -Inf, xmax = -abs(T_obs), ymin = 0, ymax = Inf,
           fill = "#e74c3c", alpha = 0.15) +
  annotate("text", x = 4.5, y = 0.32,
           label = sprintf("p[boot] == %.4f", p_boot), parse = TRUE,
           color = "#e74c3c", size = 5) +
  scale_color_manual(values = c("Bootstrap" = "#3498db", "t(n-1)" = "coral")) +
  labs(
    title = "Bootstrap Null Distribution vs. t Distribution",
    subtitle = "Blue histogram = bootstrap; dashed red = t(n-1).\nRed shading = p-value region.",
    x = "t-statistic", y = "Density"
  )

# Also show original data vs shifted
p_data_compare <- bind_rows(
  tibble(x = crp_data, type = "Original data"),
  tibble(x = crp_shifted, type = "Shifted data (H\u2080)")
) |>
  ggplot(aes(x = x, fill = type)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = mu_0, linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Original data" = "#3498db",
                               "Shifted data (H\u2080)" = "#27ae60")) +
  labs(title = "Original vs. Shifted Data",
       subtitle = expression("Vertical line at " * mu[0] * " = 1.5 mg/L"),
       x = "CRP (mg/L)", y = "Density", fill = NULL) +
  theme(legend.position = "bottom")

p_data_compare + p_boot_null
```

## When Does the Bootstrap Test Outperform the t-Test?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 7
#| fig-width: 14

# Compare rejection rates under H0 (alpha control) for normal vs skewed distributions
sim_compare <- function(n, dist = "normal", B_inner = 999, reps = 500) {
  replicate(reps, {
    if (dist == "normal") {
      x <- rnorm(n, mean = 0)
    } else {
      x <- rexp(n, rate = 1) - 1  # centered exponential (mean=0, skewed)
    }
    
    # t-test p-value
    p_t <- t.test(x, mu = 0)$p.value
    
    # Bootstrap p-value
    x_sh <- x - mean(x)  # shift to H0: mu = 0
    T_ob <- (mean(x)) / (sd(x) / sqrt(n))
    T_bs <- replicate(B_inner, {
      bs <- sample(x_sh, n, replace = TRUE)
      (mean(bs)) / (sd(bs) / sqrt(n))
    })
    p_bs <- mean(abs(T_bs) >= abs(T_ob))
    
    c(p_t = p_t, p_bs = p_bs)
  }) |>
    t() |>
    as_tibble() |>
    summarise(
      alpha_t  = mean(p_t < 0.05),
      alpha_bs = mean(p_bs < 0.05),
      .groups = "drop"
    ) |>
    mutate(n = n, dist = dist)
}

set.seed(42)
grid <- expand_grid(n = c(10, 25, 50), dist = c("normal", "skewed"))
res <- map2_dfr(grid$n, grid$dist, sim_compare)

res |>
  pivot_longer(starts_with("alpha"), names_to = "method", values_to = "type1") |>
  mutate(
    method = recode(method, alpha_t = "t-test", alpha_bs = "Bootstrap test"),
    dist   = recode(dist, normal = "Normal population", skewed = "Skewed population (Exp-1)")
  ) |>
  ggplot(aes(x = factor(n), y = type1, fill = method)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black", linewidth = 1) +
  facet_wrap(~dist) +
  scale_fill_manual(values = c("t-test" = "steelblue", "Bootstrap test" = "#27ae60")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.12)) +
  labs(
    title = expression("Type I Error Rate Under " * H[0] * " at " * alpha * " = 0.05"),
    subtitle = "Both methods well-calibrated for normal; bootstrap slightly better for small skewed samples",
    x = "Sample size (n)", y = "Empirical Type I error rate",
    fill = NULL
  ) +
  theme(legend.position = "bottom")
```

## What Happens If You Don't Shift? (A Common Mistake)

It is tempting to simply bootstrap from the **original** data and compare to $\theta_0$. Why is this wrong?

. . .

::: callout-warning
## The Naive Bootstrap Test — Incorrect
Resample from the original $x_1, \ldots, x_n$ and compute $\bar{x}^{*b}$.  
Compare the fraction of $\bar{x}^{*b}$ farther from $\bar{x}$ than $|\bar{x} - \theta_0|$.  

**Problem:** The bootstrap distribution is centered at $\bar{x}$, not at $\theta_0$.  
It approximates the sampling distribution of $\hat{\theta}$, **not** the null distribution.
:::

. . .

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6
#| fig-width: 14

set.seed(551)
n_demo <- 22
x_demo <- rgamma(n_demo, shape = 2, rate = 1)   # true mean ≈ 2
mu_demo <- 1.5
B_demo  <- 4000

xbar_demo <- mean(x_demo)

# WRONG: resample from original, compare distance from xbar
T_naive <- replicate(B_demo, mean(sample(x_demo, n_demo, replace = TRUE)))
p_naive <- mean(abs(T_naive - xbar_demo) >= abs(xbar_demo - mu_demo))

# CORRECT: shift then resample
x_sh_demo <- x_demo - xbar_demo + mu_demo
T_correct <- replicate(B_demo, {
  bs <- sample(x_sh_demo, n_demo, replace = TRUE)
  (mean(bs) - mu_demo) / (sd(bs) / sqrt(n_demo))
})
T_obs_demo <- (xbar_demo - mu_demo) / (sd(x_demo) / sqrt(n_demo))
p_correct <- mean(abs(T_correct) >= abs(T_obs_demo))
p_ttest_demo <- t.test(x_demo, mu = mu_demo)$p.value

p_nv <- ggplot(tibble(T = T_naive), aes(x = T)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#e74c3c", color = "white", alpha = 0.8) +
  geom_vline(xintercept = xbar_demo, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = mu_demo,   color = "#3498db", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = xbar_demo + 0.05, y = 2.5,
           label = "bar(x)", parse = TRUE, size = 5, color = "black") +
  annotate("text", x = mu_demo - 0.05, y = 2.5,
           label = "mu[0]", parse = TRUE, size = 5, color = "#3498db", hjust = 1) +
  labs(title = expression(bold("WRONG") * ": Naive bootstrap"),
       subtitle = sprintf("Centered at \u0078\u0305, not \u03b8\u2080 \u2014 p = %.3f (not a valid null distribution)",
                          p_naive),
       x = expression(bar(x)^"*"), y = "Density")

p_cr <- ggplot(tibble(T = T_correct), aes(x = T)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#27ae60", color = "white", alpha = 0.8) +
  geom_vline(xintercept = c(-abs(T_obs_demo), abs(T_obs_demo)),
             color = "#e74c3c", linewidth = 1.2) +
  annotate("text", x = abs(T_obs_demo) + 0.15, y = 0.32,
           label = bquote(T[obs] == .(round(T_obs_demo, 2))),
           color = "#e74c3c", size = 4.5) +
  labs(title = expression(bold("CORRECT") * ": Bootstrap from shifted data"),
       subtitle = sprintf("Centered at \u03b8\u2080 \u2014 Bootstrap p = %.3f,  t-test p = %.3f",
                          p_correct, p_ttest_demo),
       x = "t-statistic", y = "Density")

p_nv + p_cr
```

# Permutation Tests (C&H 8.3) {background-color="#27ae60"}

## The Permutation Test Idea

::: callout-important
## Key Principle (C&H 8.3)
Under $H_0: \mu_1 = \mu_2$ (or more generally, $H_0$: the two groups are **exchangeable**), the group labels are **arbitrary** — any assignment of observations to groups is equally likely.

A **permutation test** generates the exact null distribution of the test statistic by considering **all possible** (or a large random sample of) permutations of the group labels.
:::

. . .

**Algorithm:**

1. Compute observed statistic $T_{\text{obs}} = \bar{x}_1 - \bar{x}_2$
2. For $b = 1, \ldots, B$:
   - Randomly permute the combined vector of $n_1 + n_2$ observations
   - Assign first $n_1$ to Group 1, remaining $n_2$ to Group 2
   - Compute $T^{*b} = \bar{x}_1^{*b} - \bar{x}_2^{*b}$
3. p-value: $\hat{p} = \frac{1}{B}\sum_{b=1}^B \mathbf{1}(|T^{*b}| \geq |T_{\text{obs}}|)$

. . .

::: callout-note
## Exact vs. Approximate
With total $n_1+n_2$ observations, there are $\binom{n_1+n_2}{n_1}$ permutations. If this is small, we can enumerate all of them for an **exact** test. Otherwise, random permutations give a **Monte Carlo approximation** of the exact test.
:::

## Permutation Test vs. Bootstrap Test: The Key Distinction

| Feature | Bootstrap Test | Permutation Test |
|---------|---------------|-----------------|
| How $H_0$ is imposed | Shift/center data to satisfy $H_0$ | Permute group labels |
| What is resampled | From modified data | From original data |
| Validity assumption | $H_0$ + iid resampling | Exchangeability under $H_0$ |
| When exact? | Never (always approximate) | Yes, if all permutations enumerated |
| Handles one-sample? | Yes (shift to $\theta_0$) | Not naturally |
| Computes standard errors? | Yes | No |

. . .

::: callout-tip
## When to Prefer Permutation Tests
Permutation tests are preferred when:

- You want an **exact** p-value under $H_0$
- Exchangeability is a natural assumption (randomized trials!)
- You want a valid test without distributional assumptions
:::

## Medical Example: Permutation Test for Survival Times

A randomized trial compares time-to-discharge (days) between two ICU protocols. Small sample, non-normal.

```{r}
#| echo: true
# ICU time to discharge (days)
protocol_A <- c(3, 5, 7, 9, 4, 6, 8, 12, 5, 7)
protocol_B <- c(6, 9, 11, 14, 8, 10, 7, 15, 9, 13, 11)

n_A <- length(protocol_A)
n_B <- length(protocol_B)
combined <- c(protocol_A, protocol_B)
T_obs_perm <- mean(protocol_A) - mean(protocol_B)

cat("Protocol A mean:", round(mean(protocol_A), 2), "days\n")
cat("Protocol B mean:", round(mean(protocol_B), 2), "days\n")
cat("Observed difference:", round(T_obs_perm, 2), "days\n")
```

```{r}
#| echo: true
# Permutation test
set.seed(551)
B_perm <- 9999
T_perm <- replicate(B_perm, {
  perm <- sample(combined)
  mean(perm[1:n_A]) - mean(perm[(n_A + 1):(n_A + n_B)])
})

p_perm <- (sum(abs(T_perm) >= abs(T_obs_perm)) + 1) / (B_perm + 1)  # +1 for observed
p_ttest_icu <- t.test(protocol_A, protocol_B)$p.value
cat("\nPermutation p-value:", round(p_perm, 4), "\n")
cat("Welch t-test p-value:", round(p_ttest_icu, 4), "\n")
```

## Visualizing the Permutation Null Distribution

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

tibble(T = T_perm) |>
  ggplot(aes(x = T)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60,
                 fill = "#27ae60", color = "white", alpha = 0.8) +
  geom_vline(xintercept = c(-abs(T_obs_perm), abs(T_obs_perm)),
             color = "#e74c3c", linewidth = 1.5) +
  annotate("rect", xmin = abs(T_obs_perm), xmax = Inf, ymin = 0, ymax = Inf,
           fill = "#e74c3c", alpha = 0.15) +
  annotate("rect", xmin = -Inf, xmax = -abs(T_obs_perm), ymin = 0, ymax = Inf,
           fill = "#e74c3c", alpha = 0.15) +
  annotate("text", x = 3.5, y = 0.22,
           label = sprintf("p[perm] == %.4f", p_perm), parse = TRUE,
           color = "#e74c3c", size = 5.5) +
  labs(
    title = "Permutation Null Distribution: Difference in Mean ICU Days",
    subtitle = "Under H\u2080: exchangeability. Red lines = observed difference \u00b13.64 days.",
    x = "Permuted difference in means (days)",
    y = "Density"
  )
```

## Exact Permutation: Small Sample

With small samples, we can enumerate **all** permutations exactly.

```{r}
#| echo: true
#| code-fold: true
# Small example: 5 observations total, n1 = 2, n2 = 3
# All C(5,2) = 10 permutations

small_A <- c(10, 15)
small_B <- c(12, 18, 20)
combined_small <- c(small_A, small_B)
T_obs_small <- mean(small_A) - mean(small_B)

# Enumerate all C(5,2) = 10 permutations
library(combinat)
all_combs <- combn(length(combined_small), length(small_A), simplify = FALSE)

all_diffs <- map_dbl(all_combs, function(idx) {
  g1 <- combined_small[idx]
  g2 <- combined_small[-idx]
  mean(g1) - mean(g2)
})

exact_p <- mean(abs(all_diffs) >= abs(T_obs_small))

tibble(
  Permutation = seq_along(all_diffs),
  Difference  = all_diffs,
  As_extreme  = abs(all_diffs) >= abs(T_obs_small)
) |>
  mutate(Difference = round(Difference, 2)) |>
  knitr::kable(caption = sprintf("All C(5,2)=10 permutations. Observed diff = %.1f. Exact p = %.2f",
                                 T_obs_small, exact_p))
```

# Choosing Between Methods: One-Sample Tests {background-color="#f39c12"}

## Decision Framework

For a **one-sample** test $H_0: \mu = \mu_0$, which method should you use?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 15
library(DiagrammeR)

grViz("
digraph test_choice {
  graph [rankdir=TB, bgcolor=transparent, fontsize=18]
  node [shape=box, style=filled, fontsize=14, fontname='Helvetica']

  A [label='One-Sample Test\nH0: mu = mu0', fillcolor='#3498db', fontcolor=white, width=4]

  B [label='Normal population\nor large n (CLT)?', fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3.5]

  C [label='One-sample t-test\n(Most powerful for normal data)', fillcolor='#27ae60', fontcolor=white, width=3.5]

  D [label='iid observations?', fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3]

  E [label='Symmetric around mu0\nunder H0?', fillcolor='#f39c12', fontcolor=white, shape=diamond, width=3.5]

  F [label='Bootstrap Test\nShift to mu0, resample', fillcolor='#e74c3c', fontcolor=white, width=3.5]

  G [label='Sign-Flip Permutation Test\nFlip signs of (xi - mu0)', fillcolor='#8e44ad', fontcolor=white, width=3.5]

  H [label='Consult a statistician!\n(Dependent / clustered data)', fillcolor='#95a5a6', fontcolor=white, width=3.5]

  A -> B
  B -> C [label='  Yes', fontsize=13]
  B -> D [label='  No', fontsize=13]
  D -> E [label='  Yes', fontsize=13]
  D -> H [label='  No', fontsize=12]
  E -> G [label='  Yes', fontsize=13]
  E -> F [label='  No (or unsure)', fontsize=12]
}
")
```

## Assumptions Compared

| Assumption | t-test | Bootstrap test | Sign-flip permutation |
|-----------|--------|---------------|-----------------------|
| Normal population (or large $n$) | Required | Not required | Not required |
| iid observations | Required | Required | Required |
| Symmetric around $\mu_0$ under $H_0$ | Not required | Not required | **Required** |
| Distribution-free validity | No | Approximately | Yes (exact) |
| Works for any test statistic | No | **Yes** | **Yes** |
| Computationally intensive | No | Moderate | Moderate |

. . .

::: callout-tip
## Key Message
The **bootstrap test** is the most flexible choice for one-sample problems with non-normal, skewed data — it only requires iid sampling and makes no symmetry assumption. The **sign-flip permutation test** is exact under symmetry. The **t-test** is optimal when normality holds.
:::

## Simulation Study: Comparing the Three Methods

How well does each method control the Type I error rate, and what is the power, for different population shapes?

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

run_onesample_tests <- function(n = 20, delta = 0, dist = "normal", B_inner = 499) {
  # Generate data under true mean = delta (H0: mu = 0)
  x <- switch(dist,
    normal  = rnorm(n, mean = delta),
    skewed  = rexp(n, rate = 1) - 1 + delta,   # centered Exp, mean = delta
    heavyt  = rt(n, df = 3) + delta             # heavy-tailed t(3)
  )

  mu_0 <- 0

  # 1. Parametric t-test
  p_t <- t.test(x, mu = mu_0)$p.value

  # 2. Bootstrap test (shift-resample)
  x_sh  <- x - mean(x) + mu_0
  T_ob  <- (mean(x) - mu_0) / (sd(x) / sqrt(n))
  T_bs  <- replicate(B_inner, {
    bs <- sample(x_sh, n, replace = TRUE)
    (mean(bs) - mu_0) / (sd(bs) / sqrt(n))
  })
  p_b <- mean(abs(T_bs) >= abs(T_ob))

  # 3. Sign-flip permutation test
  d      <- x - mu_0
  T_sign <- replicate(B_inner, {
    signs <- sample(c(-1, 1), n, replace = TRUE)
    mean(signs * d)
  })
  p_s <- (sum(abs(T_sign) >= abs(mean(d))) + 1) / (B_inner + 1)

  c(p_t = p_t, p_boot = p_b, p_sign = p_s)
}

set.seed(99)
sim_grid <- expand_grid(
  n     = c(15, 40),
  delta = c(0, 0.6),
  dist  = c("normal", "skewed", "heavyt")
)

sim_res <- sim_grid |>
  rowwise() |>
  mutate(
    sims = list(
      replicate(500, run_onesample_tests(n = n, delta = delta, dist = dist,
                                        B_inner = 299)) |>
        t() |> as_tibble()
    )
  ) |>
  unnest(sims) |>
  pivot_longer(starts_with("p_"), names_to = "method", values_to = "pval") |>
  group_by(n, delta, dist, method) |>
  summarise(rejection = mean(pval < 0.05), .groups = "drop") |>
  mutate(
    method   = recode(method, p_t = "t-test", p_boot = "Bootstrap", p_sign = "Sign-flip"),
    dist_lab = recode(dist, normal = "Normal", skewed = "Skewed (Exp-1)", heavyt = "Heavy-tailed (t\u2083)"),
    quantity = ifelse(delta == 0,
                      "Type I error rate (H\u2080 true, target = 5%)",
                      "Power (H\u2090 true, \u03b4 = 0.6)")
  )

ggplot(sim_res |> filter(n == 15),
       aes(x = method, y = rejection, fill = method)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_hline(yintercept = 0.05, linetype = "dashed", linewidth = 0.8) +
  facet_grid(quantity ~ dist_lab, scales = "free_y") +
  scale_fill_manual(values = c("t-test" = "steelblue",
                               "Bootstrap" = "#e74c3c",
                               "Sign-flip" = "#8e44ad")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "One-Sample Tests: Type I Error and Power  (n = 15)",
    subtitle = "All methods well-calibrated; bootstrap and sign-flip more robust for skewed/heavy-tailed data",
    x = NULL, y = "Rejection rate", fill = NULL
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 13),
        axis.text.x = element_text(angle = 20, hjust = 1))
```

## Simulation: Larger Sample (n = 40)

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 8
#| fig-width: 14

ggplot(sim_res |> filter(n == 40),
       aes(x = method, y = rejection, fill = method)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_hline(yintercept = 0.05, linetype = "dashed", linewidth = 0.8) +
  facet_grid(quantity ~ dist_lab, scales = "free_y") +
  scale_fill_manual(values = c("t-test" = "steelblue",
                               "Bootstrap" = "#e74c3c",
                               "Sign-flip" = "#8e44ad")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "One-Sample Tests: Type I Error and Power  (n = 40)",
    subtitle = "By n = 40, all three methods converge — CLT smooths out distributional differences",
    x = NULL, y = "Rejection rate", fill = NULL
  ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 13),
        axis.text.x = element_text(angle = 20, hjust = 1))
```

## Key Takeaways from Simulation

- **Normal data:** All three methods control Type I error well at both $n = 15$ and $n = 40$; t-test has the highest power (it's optimal here)

. . .

- **Skewed data, small $n$:** The t-test can be slightly anti-conservative (Type I error > 5%); the bootstrap and sign-flip tests maintain better control

. . .

- **Heavy-tailed data ($t_3$):** Similar pattern — bootstrap and sign-flip are more robust for small samples

. . .

- **Large $n$:** All differences wash out — the CLT rescues the t-test regardless of the population shape

. . .

::: callout-note
## Practical Implication
For small samples ($n < 30$) with non-normal data, bootstrap or sign-flip tests are preferable. For $n \geq 40$, the t-test is nearly equivalent in practice. The computational cost of bootstrap/permutation is rarely a concern in modern software.
:::

# Bootstrap Tests and the Duality with CIs {background-color="#16a085"}

## Bootstrap Tests and Bootstrap CIs: Are They Dual?

In the parametric setting, we showed that confidence intervals and tests are **exactly dual**. Does this hold for bootstrap tests and bootstrap CIs?

. . .

::: callout-important
## Answer: Approximately, but the method matters

- **Bootstrap percentile CI** ↔ **bootstrap test based on $\hat{\theta}$**: approximately dual
- The duality is exact only if the bootstrap CI and test use the same pivot/statistic
- In practice, they may give slightly different inferences for small $n$
:::

. . .

```{r}
#| echo: true
#| code-fold: true
set.seed(551)
# One-sample example: check duality
crp_small <- c(2.1, 3.4, 1.8, 4.2, 2.9, 3.7, 1.5, 5.1, 2.6, 3.8)
n_s  <- length(crp_small)
B_s  <- 4999

# Bootstrap CI for mu (percentile method)
boot_means <- replicate(B_s, mean(sample(crp_small, n_s, replace = TRUE)))
ci_boot <- quantile(boot_means, c(0.025, 0.975))

# Bootstrap test: H0: mu = 2.5 vs Ha: mu != 2.5
mu0_test <- 2.5
x_sh     <- crp_small - mean(crp_small) + mu0_test
T_ob_s   <- mean(crp_small) - mu0_test
T_bs_s   <- replicate(B_s, mean(sample(x_sh, n_s, replace = TRUE)) - mu0_test)
p_boot_s <- mean(abs(T_bs_s) >= abs(T_ob_s))

cat("Bootstrap 95% CI: [", round(ci_boot[1], 3), ",", round(ci_boot[2], 3), "]\n")
cat("mu_0 = 2.5 in CI?", ci_boot[1] <= 2.5 & 2.5 <= ci_boot[2], "\n")
cat("Bootstrap test p-value for mu_0 = 2.5:", round(p_boot_s, 4), "\n")
cat("Consistent? (p>=0.05 iff in CI):", (p_boot_s >= 0.05) == (ci_boot[1] <= 2.5 & 2.5 <= ci_boot[2]))
```

# One-Sample Permutation Test (C&H 8.3) {background-color="#2c3e50"}

## One-Sample Permutation Test: Sign Test Perspective

For a one-sample test $H_0: \mu = \mu_0$, permutation tests work via **sign flipping** (not label permutation).

. . .

Under $H_0: \text{median} = \mu_0$ (or symmetry about $\mu_0$):

The deviations $d_i = x_i - \mu_0$ are symmetric about 0, so **each sign is equally likely to be + or −**.

. . .

**Algorithm:**

1. Compute $d_i = x_i - \mu_0$ and observed $T_{\text{obs}} = \bar{d}$
2. For $b = 1, \ldots, B$: randomly assign $\pm 1$ to each $d_i$; compute $T^{*b} = \bar{d}^{*b}$
3. p-value: fraction with $|T^{*b}| \geq |T_{\text{obs}}|$

```{r}
#| echo: true
# Sign-flip permutation test for mean CRP
crp_data_perm <- c(2.1, 3.4, 1.8, 4.2, 2.9, 3.7, 1.5, 5.1, 2.6, 3.8, 4.0, 2.2)
mu_0_crp <- 2.5
d <- crp_data_perm - mu_0_crp
T_obs_crp <- mean(d)

set.seed(551)
B_sign <- 9999
T_sign <- replicate(B_sign, {
  signs <- sample(c(-1, 1), length(d), replace = TRUE)
  mean(signs * d)
})

p_sign <- mean(abs(T_sign) >= abs(T_obs_crp))
cat("Observed mean deviation:", round(T_obs_crp, 4), "\n")
cat("Sign-flip permutation p-value:", round(p_sign, 4), "\n")
cat("Parametric t-test p-value:", round(t.test(crp_data_perm, mu = mu_0_crp)$p.value, 4))
```

# Key Takeaways {background-color="#2c3e50"}

## Key Takeaways

::: callout-important
## Today's Main Points (C&H Chapter 8)

1. **Bootstrap hypothesis tests** approximate the null distribution by resampling from data **shifted to satisfy $H_0$** — this is fundamentally different from bootstrap CIs (which resample from the original data)

2. **The naive bootstrap is wrong:** resampling from the original data without shifting produces a distribution centered at $\bar{x}$, not at $\mu_0$ — it does not represent the null distribution

3. **One-sample bootstrap test:** shift data so $\bar{x}^* = \theta_0$, resample, compute the t-statistic, obtain p-value as the fraction of bootstrap statistics as extreme as observed

4. **Sign-flip permutation test:** under $H_0$ and symmetry, the signs of $d_i = x_i - \mu_0$ are exchangeable — randomly flipping them generates the exact null distribution

5. **Permutation $\neq$ Bootstrap:** sign-flip permutation requires symmetry around $\mu_0$; bootstrap requires only iid resampling. Both avoid parametric distributional assumptions

6. **All three methods** converge for large $n$ (CLT); differences matter most for small, skewed, or heavy-tailed samples

7. **Duality:** bootstrap tests and bootstrap CIs are approximately (but not exactly) dual — the CI is not simply the set of values where the bootstrap test fails to reject
:::

## Looking Ahead

**Next class (Lesson 18): Two-Sample z and t Confidence Intervals and Tests (Devore 9.5)**

- Two-sample z-test for means ($\sigma$ known)
- Welch t-test for means ($\sigma$ unknown)
- Pooled t-test (equal variance assumption)
- Two-sample inference for proportions
- Inferences about two variances (F-test)
- Power and sample size for two-sample tests

## References

::: nonincremental
-   Chihara, L. M. and Hesterberg, T. C. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 8: Bootstrap Hypothesis Testing.
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 9.6.
-   Efron, B. and Hastie, T. (2016). *Computer Age Statistical Inference*. Cambridge University Press. Chapter 15.
-   Good, P. I. (2005). *Permutation, Parametric, and Bootstrap Tests of Hypotheses* (3rd ed.). Springer.
-   Ernst, M. D. (2004). Permutation methods: a basis for exact inference. *Statistical Science*, 19(4), 676–685.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
# For PDF export
decktape reveal --fragments docs/lessons/17_BootstrapHypothesisTesting/17_BootstrapHypothesisTesting.html lessons/17_BootstrapHypothesisTesting/17_BootstrapHypothesisTesting.pdf
```
