---
title: "Lesson 3: Unbiased Estimators, MVUE, Likelihood Theory"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w1d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 3
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```


# Lesson 3: Finding the Best Estimator

## Review: Where We Left Off

**Key concepts from Lesson 2:**

-   Point estimators vs. estimates
-   Bias: $\text{Bias}(\hat{\theta}) = E(\hat{\theta}) - \theta$
-   Unbiasedness: $E(\hat{\theta}) = \theta$
-   MSE = Variance + Bias²

. . .

**Today's Goals:**

-   Understand what makes one unbiased estimator better than another
-   Learn about Minimum Variance Unbiased Estimators (MVUE)
-   Introduce likelihood functions (preview of Lesson 4)
-   Define Fisher Information and its role in estimation
-   Learn the Cramér-Rao Lower Bound for variance


## Motivating Example: Estimating Drug Efficacy

A clinical trial tests the effectiveness of a new antibiotic. Two researchers propose different estimators for the true response rate $p$:

. . .

**Researcher A:** Use $\hat{p}_A = X/n$ (sample proportion)

**Researcher B:** Use $\hat{p}_B = (X + 1)/(n + 2)$ (adjusted proportion)

. . .

Both estimators are functions of the same data. **Which should we use?**

```{r}
# Suppose 15 out of 20 patients respond
n <- 20; x <- 15
p_hat_A <- x / n
p_hat_B <- (x + 1) / (n + 2)
cat("Researcher A's estimate:", p_hat_A, "\n")
cat("Researcher B's estimate:", p_hat_B)
```

# Choosing Among Unbiased Estimators {background-color="#27ae60"}

## The Problem of Multiple Unbiased Estimators

Recall from Lesson 2: There can be **many** unbiased estimators for the same parameter!

. . .

**Example:** For a symmetric distribution with mean $\mu$:

-   Sample mean $\bar{X}$ is unbiased for $\mu$
-   Sample median $\tilde{X}$ is unbiased for $\mu$
-   Trimmed means are unbiased for $\mu$

. . .

::: callout-note
## The Key Question
If two estimators are both unbiased, how do we choose between them?
:::

## Comparing Two Unbiased Estimators: Intuition

```{r}
#| echo: false
#| fig-height: 5
set.seed(123)
true_mu <- 50

comparison_data <- tibble(
  estimator = rep(c("Low Variance (Better)", "High Variance (Worse)"), each = 1000),
  estimate = c(
    rnorm(1000, true_mu, 2),
    rnorm(1000, true_mu, 5)
  )
)

comparison_data |> 
  ggplot(aes(x = estimate, fill = estimator)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = true_mu, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Two Unbiased Estimators: Which is Better?",
       subtitle = "Both centered on true value (red line), but with different spread",
       x = "Estimate", y = "Frequency", fill = "Estimator") +
  scale_fill_manual(values = c("steelblue", "orange")) +
  theme(legend.position = "top")
```

**Answer:** Among unbiased estimators, prefer the one with **smaller variance**.

## The Principle: Minimum Variance Unbiased Estimator

::: callout-important
## Definition (Devore 7.1)

The **Minimum Variance Unbiased Estimator (MVUE)** of a parameter $\theta$ is the estimator $\hat{\theta}$ such that:

1.  $\hat{\theta}$ is unbiased: $E(\hat{\theta}) = \theta$
2.  $\hat{\theta}$ has smaller or equal variance than any other unbiased estimator

If $\hat{\theta}_1$ and $\hat{\theta}_2$ are both unbiased for $\theta$, we prefer the one with smaller variance.
:::

. . .

**Intuition:** Among estimators that are "correct on average," we want the most precise one.

## Example: Uniform Distribution Upper Bound

**Setup:** $X_1, \ldots, X_n \sim \text{Uniform}[0, \theta]$

Two unbiased estimators for $\theta$:

. . .

1.  $\hat{\theta}_1 = 2\bar{X}$ (since $E(X_i) = \theta/2$, so $E(2\bar{X}) = \theta$)

2.  $\hat{\theta}_2 = \frac{n+1}{n} \cdot \max(X_1, \ldots, X_n)$ (corrected maximum)

. . .

**Question:** Which has smaller variance?

## Calculating Variances

For $\hat{\theta}_1 = 2\bar{X}$:
$$\text{Var}(\hat{\theta}_1) = \text{Var}(2\bar{X}) = 4 \cdot \text{Var}(\bar{X}) = 4 \cdot \frac{\sigma^2}{n} = 4 \cdot \frac{\theta^2/12}{n} = \frac{\theta^2}{3n}$$

. . .

For $\hat{\theta}_2 = \frac{n+1}{n} \cdot \max(X_i)$, it can be shown that:
$$\text{Var}(\hat{\theta}_2) = \frac{\theta^2}{n(n+2)}$$

. . .

**Comparison:** For $n > 1$:
$$\frac{\theta^2}{n(n+2)} < \frac{\theta^2}{3n}$$

The corrected maximum $\hat{\theta}_2$ is more efficient!

## Simulation: Comparing the Two Estimators

```{r}
#| code-line-numbers: "|1-3|5-12|14-19"
theta <- 10  # True upper bound
n <- 20
n_sims <- 5000

# Simulate both estimators
uniform_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) runif(n, 0, theta)),
    theta_hat_1 = map_dbl(data, \(d) 2 * mean(d)),
    theta_hat_2 = map_dbl(data, \(d) (n + 1) / n * max(d))
  )

# Compare variances
uniform_sim |> 
  summarize(
    `Var(2*X̄)` = var(theta_hat_1),
    `Var(corrected max)` = var(theta_hat_2),
    `Theoretical Var(2*X̄)` = theta^2 / (3 * n),
    `Theoretical Var(corrected max)` = theta^2 / (n * (n + 2))
  ) |> 
  pivot_longer(everything()) |> 
  mutate(value = round(value, 4))
```

## Visualizing the Comparison

```{r}
#| echo: false
#| fig-height: 5
uniform_sim |> 
  select(theta_hat_1, theta_hat_2) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = case_when(
    Estimator == "theta_hat_1" ~ "theta_hat_1",
    Estimator == "theta_hat_2" ~ "theta_hat_2"
  )) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = theta, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = expression(paste("Comparing Two Unbiased Estimators for ", theta)),
       subtitle = str_glue("True θ = {theta}, n = {n}"),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(
    values = c("theta_hat_1" = "darkgreen", "theta_hat_2" = "steelblue"),
    labels = c(
      "theta_hat_1" = expression(2*bar(X)~"(higher variance)"),
      "theta_hat_2" = expression("Corrected Maximum (MVUE)")
    )
  ) +
  theme(legend.position = "top")
```

The corrected maximum is **more concentrated** around the true value!


# Introduction to Likelihood {background-color="#27ae60"}

## The Likelihood Function: A Brief Introduction

::: callout-important
## Definition (Preview - Full treatment in Lesson 4)

The **likelihood function** $L(\theta)$ is the joint probability/density of the observed data, viewed as a *function of the parameter* $\theta$:

$$L(\theta) = \prod_{i=1}^n f(x_i; \theta)$$

where $f$ is the pmf (discrete) or pdf (continuous).
:::

. . .

**Key Idea:** The likelihood tells us how "likely" different parameter values are, given our observed data.

## The Log-Likelihood Function

Taking the logarithm converts products to sums:

$$\ell(\theta) = \ln L(\theta) = \sum_{i=1}^n \ln f(x_i; \theta)$$

. . .

**Why use log-likelihood?**

-   Easier to differentiate (sums vs. products)
-   Numerically more stable
-   Same maximum as likelihood (since log is monotonic)

. . .

The **score function** is the first derivative: $\ell'(\theta) = \frac{d}{d\theta}\ell(\theta)$

## Example: Bernoulli Likelihood

For $X_1, \ldots, X_n \sim \text{Bernoulli}(p)$ with $x = \sum x_i$ successes:

**Likelihood:**
$$L(p) = p^x(1-p)^{n-x}$$

**Log-likelihood:**
$$\ell(p) = x\ln(p) + (n-x)\ln(1-p)$$

**Score function:**
$$\ell'(p) = \frac{x}{p} - \frac{n-x}{1-p}$$

## Visualizing the Log-Likelihood

```{r}
#| echo: false
#| fig-height: 5
n <- 200
x <- 156

log_lik_data <- tibble(
  p = seq(0.5, 0.95, by = 0.001),
  log_likelihood = x * log(p) + (n - x) * log(1 - p)
)

mle_p <- x / n

# Calculate second derivative at MLE for curvature illustration
log_lik_data |> 
  ggplot(aes(x = p, y = log_likelihood)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = mle_p, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(x = mle_p, y = x * log(mle_p) + (n - x) * log(1 - mle_p)), 
             color = "red", size = 4) +
  labs(title = "Log-Likelihood for Bernoulli Data",
       subtitle = str_glue("n = {n}, x = {x} successes"),
       x = "Parameter (p)", 
       y = expression("Log-Likelihood "*ell(p))) +
  annotate("text", x = mle_p + 0.05, y = x * log(mle_p) + (n - x) * log(1 - mle_p) - 2,
           label = bquote(atop("Maximum at", hat(p) == .(mle_p))), 
           color = "red", size = 5, parse = FALSE)
```

The **curvature** of this function contains information about how precisely we can estimate $p$.

# Fisher Information {background-color="#8e44ad"}

## Curvature and Information

**Key Observation:** The more curved the log-likelihood is at its peak, the more precisely we can estimate the parameter.

```{r}
#| echo: false
# Compare curvature for different sample sizes
bind_rows(
  tibble(n = 25, y = 19, p = seq(0.5, 0.95, 0.01)) |> 
    mutate(log_lik = y * log(p) + (n - y) * log(1 - p) - (y * log(y/n) + (n-y) * log(1-y/n))),
  tibble(n = 100, y = 76, p = seq(0.5, 0.95, 0.01)) |> 
    mutate(log_lik = y * log(p) + (n - y) * log(1 - p) - (y * log(y/n) + (n-y) * log(1-y/n)))
) |> 
  ggplot(aes(x = p, y = log_lik, color = factor(n))) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0.76, linetype = "dashed", alpha = 0.5) +
  labs(title = "Log-Likelihood Curvature: More Data = More Information",
       subtitle = expression("Both have same MLE ("*hat(p)*" = 0.76), but n = 100 is more curved"),
       x = "Parameter (p)", y = "Centered Log-Likelihood",
       color = "Sample size") +
  scale_color_manual(values = c("steelblue", "darkgreen")) +
  theme(legend.position = "top",
        plot.title = element_text(size = 18),
        plot.subtitle = element_text(size = 16),
        plot.margin = margin(10, 15, 10, 10))
```

**More curvature = more information = more precise estimates**

## Fisher Information: Definition (Devore 7.4)

::: callout-important
## Definition (Devore 7.4)

Let $f(x; \theta)$ be a pmf or pdf. The **Fisher Information** in a single observation $X$ is:

$$I(\theta) = E\left[-\frac{\partial^2}{\partial\theta^2} \ln f(X; \theta)\right] = E[-\ell''(\theta)]$$

Equivalently:
$$I(\theta) = E\left[\left(\frac{\partial}{\partial\theta} \ln f(X; \theta)\right)^2\right] = E[(\ell'(\theta))^2] = \text{Var}(\ell'(\theta))$$
:::

. . .

**Intuition:** Fisher information measures the expected curvature of the log-likelihood.

## Fisher Information in a Sample

For a random sample $X_1, \ldots, X_n$:

$$I_n(\theta) = n \cdot I(\theta)$$

. . .

**The information is additive!** Each observation contributes the same amount of information.

. . .

::: callout-tip
## Key Properties
-   $I(\theta) > 0$ (always positive)
-   Larger sample size → more information
-   Information depends on the distribution, not the data
:::

## Example: Fisher Information for Bernoulli

For $X \sim \text{Bernoulli}(p)$: $f(x; p) = p^x(1-p)^{1-x}$

**Step 1:** Log-likelihood for single observation:
$$\ln f(x; p) = x\ln(p) + (1-x)\ln(1-p)$$

**Step 2:** Second derivative:
$$\frac{\partial^2}{\partial p^2}\ln f(x; p) = -\frac{x}{p^2} - \frac{1-x}{(1-p)^2}$$

## Bernoulli Fisher Information (continued)

**Step 3:** Take expectation (multiply by -1):
$$I(p) = E\left[\frac{X}{p^2} + \frac{1-X}{(1-p)^2}\right]$$

$$= \frac{E(X)}{p^2} + \frac{1 - E(X)}{(1-p)^2} = \frac{p}{p^2} + \frac{1-p}{(1-p)^2}$$

$$= \frac{1}{p} + \frac{1}{1-p} = \frac{1}{p(1-p)}$$

. . .

**For a sample of size $n$:**
$$I_n(p) = \frac{n}{p(1-p)}$$

## Example: Fisher Information for Normal (known $\sigma$)

For $X \sim N(\mu, \sigma^2)$ with $\sigma$ known:

$$f(x; \mu) = \frac{1}{\sqrt{2\pi}\sigma}e^{-(x-\mu)^2/(2\sigma^2)}$$

. . .

$$\ln f(x; \mu) = -\frac{1}{2}\ln(2\pi\sigma^2) - \frac{(x-\mu)^2}{2\sigma^2}$$

. . .

$$\frac{\partial}{\partial\mu}\ln f = \frac{x-\mu}{\sigma^2}, \quad \frac{\partial^2}{\partial\mu^2}\ln f = -\frac{1}{\sigma^2}$$

. . .

$$I(\mu) = E\left[\frac{1}{\sigma^2}\right] = \frac{1}{\sigma^2}$$

**For a sample:** $I_n(\mu) = \frac{n}{\sigma^2}$

## Example: Fisher Information for Exponential

For $X \sim \text{Exponential}(\lambda)$: $f(x; \lambda) = \lambda e^{-\lambda x}$

**Log-likelihood:** $\ln f(x; \lambda) = \ln(\lambda) - \lambda x$

. . .

**Derivatives:**
$$\frac{\partial}{\partial\lambda}\ln f = \frac{1}{\lambda} - x$$
$$\frac{\partial^2}{\partial\lambda^2}\ln f = -\frac{1}{\lambda^2}$$

. . .

**Fisher Information:**
$$I(\lambda) = E\left[\frac{1}{\lambda^2}\right] = \frac{1}{\lambda^2}$$

**For a sample:** $I_n(\lambda) = \frac{n}{\lambda^2}$

## Your Turn: Calculate Fisher Information

**Exercise:** For the Poisson distribution with parameter $\mu$, where $f(x; \mu) = \frac{e^{-\mu}\mu^x}{x!}$:

1.  Find $\ln f(x; \mu)$
2.  Calculate $\frac{\partial^2}{\partial\mu^2}\ln f(x; \mu)$
3.  Find $I(\mu)$

. . .

**Solution:**

1.  $\ln f = -\mu + x\ln(\mu) - \ln(x!)$
2.  $\frac{\partial}{\partial\mu}\ln f = -1 + \frac{x}{\mu}$, $\frac{\partial^2}{\partial\mu^2}\ln f = -\frac{x}{\mu^2}$
3.  $I(\mu) = E\left[\frac{X}{\mu^2}\right] = \frac{E(X)}{\mu^2} = \frac{\mu}{\mu^2} = \frac{1}{\mu}$

# The Cramér-Rao Lower Bound {background-color="#e74c3c"}

## The Fundamental Inequality

::: callout-important
## Cramér-Rao Inequality (Devore 7.4)

Let $X_1, \ldots, X_n$ be a random sample from $f(x; \theta)$ whose support does not depend on $\theta$. If $T = t(X_1, \ldots, X_n)$ is an **unbiased** estimator of $\theta$, then:

$$\text{Var}(T) \geq \frac{1}{I_n(\theta)} = \frac{1}{n \cdot I(\theta)}$$

The right-hand side is the **Cramér-Rao Lower Bound (CRLB)**.
:::

. . .

**In words:** No unbiased estimator can have variance smaller than $1/I_n(\theta)$.

## Why the CRLB Matters

The Cramér-Rao bound tells us:

1.  **The best possible precision** for any unbiased estimator
2.  **Whether an estimator is efficient** (achieves the bound)
3.  **How much we lose** by using a suboptimal estimator

. . .

::: callout-tip
## Efficiency Definition
An unbiased estimator $T$ is **efficient** if:
$$\text{Var}(T) = \frac{1}{I_n(\theta)}$$

An efficient estimator is the MVUE!
:::

## Example: Is $\hat{p}$ Efficient for Bernoulli?

For $X_1, \ldots, X_n \sim \text{Bernoulli}(p)$:

**Estimator:** $\hat{p} = \bar{X} = \frac{\sum X_i}{n}$

**Variance:** $\text{Var}(\hat{p}) = \frac{p(1-p)}{n}$

. . .

**Fisher Information:** $I_n(p) = \frac{n}{p(1-p)}$

**CRLB:** $\frac{1}{I_n(p)} = \frac{p(1-p)}{n}$

. . .

$$\text{Var}(\hat{p}) = \frac{p(1-p)}{n} = \text{CRLB} \checkmark$$

**$\hat{p}$ is efficient and is the MVUE of $p$!**

## Example: Is $\bar{X}$ Efficient for Normal Mean?

For $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$ with known $\sigma$:

**Estimator:** $\bar{X}$

**Variance:** $\text{Var}(\bar{X}) = \frac{\sigma^2}{n}$

. . .

**Fisher Information:** $I_n(\mu) = \frac{n}{\sigma^2}$

**CRLB:** $\frac{1}{I_n(\mu)} = \frac{\sigma^2}{n}$

. . .

$$\text{Var}(\bar{X}) = \frac{\sigma^2}{n} = \text{CRLB} \checkmark$$

**$\bar{X}$ is efficient and is the MVUE of $\mu$!**

## Visualizing Efficiency

```{r}
#| echo: false
#| fig-height: 5
set.seed(123)
true_mu <- 50
sigma <- 10
n <- 25
n_sims <- 3000

# Compare sample mean (efficient) to sample median (inefficient)
efficiency_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) rnorm(n, true_mu, sigma)),
    sample_mean = map_dbl(data, mean),
    sample_median = map_dbl(data, median)
  )

# Theoretical variances
var_mean <- sigma^2 / n  # = CRLB
var_median <- (pi/2) * sigma^2 / n  # approximately

efficiency_sim |> 
  select(sample_mean, sample_median) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = case_when(
    Estimator == "sample_mean" ~ str_glue("Sample Mean (Var = {round(var_mean, 2)}, Efficient)"),
    Estimator == "sample_median" ~ str_glue("Sample Median (Var ≈ {round(var_median, 2)}, Not Efficient)")
  )) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = true_mu, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Efficient vs. Inefficient Estimators for Normal Mean",
       subtitle = str_glue("CRLB = σ²/n = {round(var_mean, 2)}; Sample mean achieves it"),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("steelblue", "darkgreen")) +
  theme(legend.position = "top")
```

## Efficiency of the Sample Median

For normal data, the sample median is unbiased for $\mu$ but has:

$$\text{Var}(\tilde{X}) \approx \frac{\pi}{2} \cdot \frac{\sigma^2}{n} \approx 1.57 \cdot \frac{\sigma^2}{n}$$

. . .

**Efficiency of median:**
$$\text{Efficiency} = \frac{\text{CRLB}}{\text{Var}(\tilde{X})} = \frac{\sigma^2/n}{1.57 \cdot \sigma^2/n} \approx 0.64$$

. . .

The median is only about **64% as efficient** as the mean for normal data.

We need about **1.57 times as many observations** with the median to achieve the same precision as the mean!

## Medical Application: Estimating Treatment Effect

A clinical trial measures systolic BP reduction (mmHg). Assume $N(\mu, \sigma^2 = 100)$.

```{r}
n <- 30
sigma_sq <- 100

# CRLB for μ
crlb <- sigma_sq / n

# Variance of sample mean
var_xbar <- sigma_sq / n

# Efficiency check
cat("CRLB for μ:", round(crlb, 3), "\n")
cat("Var(X̄):", round(var_xbar, 3), "\n")
cat("Is X̄ efficient?", var_xbar == crlb, "\n\n")

# Sample data
bp_data <- c(12.3, 8.7, 15.2, 10.1, 7.4, 14.8, 11.2, 9.5, 13.1, 16.2,
             8.1, 12.5, 10.8, 14.3, 11.7, 9.2, 15.0, 13.4, 7.9, 12.1,
             10.5, 8.3, 14.1, 11.5, 13.8, 9.8, 11.9, 12.7, 10.3, 14.6)

cat("MVUE of μ (sample mean):", round(mean(bp_data), 2), "mmHg\n")
cat("Standard Error:", round(sqrt(crlb), 2), "mmHg")
```

## When is the CRLB Not Attained?

Not all unbiased estimators achieve the CRLB!

**Example:** Uniform[0, $\theta$] distribution

-   Fisher Information: $I(\theta) = 1/\theta^2$
-   CRLB: $\theta^2/n$

. . .

Two unbiased estimators:

-   $\hat{\theta}_1 = 2\bar{X}$ has variance $\theta^2/(3n)$
-   $\hat{\theta}_2 = \frac{n+1}{n}\max(X_i)$ has variance $\theta^2/[n(n+2)]$

. . .

Neither achieves CRLB = $\theta^2/n$! Why?

**The support of Uniform[0, $\theta$] depends on $\theta$**, violating a regularity condition.

## When Does the CRLB Not Apply?

The Cramér-Rao inequality requires:

1.  The **support** of $f(x; \theta)$ doesn't depend on $\theta$
2.  Certain **regularity conditions** on derivatives

. . .

**Examples where CRLB doesn't apply:**

-   Uniform[0, $\theta$] - support depends on $\theta$
-   Uniform[$\theta$, $\theta + 1$] - support depends on $\theta$
-   Some discrete distributions with restricted parameter spaces

. . .

::: callout-warning
Always check regularity conditions before applying the CRLB!
:::

## MVUE Summary

::: callout-important
## Finding the MVUE

1.  Calculate $I(\theta)$ for the distribution
2.  Compute CRLB = $1/[n \cdot I(\theta)]$
3.  Find an unbiased estimator
4.  Check if its variance equals the CRLB
5.  If yes → it's the MVUE!
:::

. . .

**Key Results:**

| Distribution | Parameter | MVUE | CRLB |
|--------------|-----------|------|------|
| Bernoulli($p$) | $p$ | $\bar{X}$ | $p(1-p)/n$ |
| Normal($\mu, \sigma^2$) | $\mu$ | $\bar{X}$ | $\sigma^2/n$ |
| Exponential($\lambda$) | $\lambda$ | See note | $\lambda^2/n$ |
| Poisson($\mu$) | $\mu$ | $\bar{X}$ | $\mu/n$ |

# Summary and Looking Ahead {background-color="#2c3e50"}

## Lesson 3 Summary

**Key Concepts:**

1.  **MVUE:** The unbiased estimator with minimum variance
    -   Among unbiased estimators, prefer smaller variance

2.  **Likelihood & Log-Likelihood:**
    -   $L(\theta) = \prod f(x_i; \theta)$, $\ell(\theta) = \sum \ln f(x_i; \theta)$
    -   Curvature relates to estimation precision

3.  **Fisher Information:** $I(\theta) = E[-\ell''(\theta)]$
    -   Measures curvature of log-likelihood
    -   Additive across observations: $I_n(\theta) = n \cdot I(\theta)$

4.  **Cramér-Rao Lower Bound:**
    -   $\text{Var}(T) \geq 1/I_n(\theta)$ for unbiased $T$
    -   Efficient estimators achieve the bound → MVUE

## Key Formulas to Remember

| Distribution | Parameter | $I(\theta)$ | CRLB |
|--------------|-----------|-------------|------|
| Bernoulli | $p$ | $\frac{1}{p(1-p)}$ | $\frac{p(1-p)}{n}$ |
| Normal (known $\sigma$) | $\mu$ | $\frac{1}{\sigma^2}$ | $\frac{\sigma^2}{n}$ |
| Exponential | $\lambda$ | $\frac{1}{\lambda^2}$ | $\frac{\lambda^2}{n}$ |
| Poisson | $\mu$ | $\frac{1}{\mu}$ | $\frac{\mu}{n}$ |

## Lesson 3 Practice Problems

1.  **(Devore 7.51)** For the geometric distribution $P(X = x) = (1-p)^{x-1}p$, find Fisher information and the CRLB.

2.  **(Devore 7.52)** For Poisson($\mu$), show that $\bar{X}$ is an efficient estimator.

3.  **(Devore 7.53)** For Uniform[0, $\theta$], explain why neither $2\bar{X}$ nor the corrected maximum achieves the CRLB.

4.  **Medical Application:** In a trial, 72 of 90 patients respond. Calculate Fisher information, CRLB, and verify that $\hat{p} = 72/90$ is efficient.



## Next Lesson Preview

**Lesson 4: Method of Moments and Maximum Likelihood Estimation**

-   Method of Moments: Equate population and sample moments
-   Maximum Likelihood: Maximize the likelihood function
-   Properties: invariance, consistency, asymptotic normality
-   MLEs are asymptotically efficient!

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapters 7.1, 7.4
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 6.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/03_MVUE_Likelihood/03_MVUE.html lessons/03_MVUE/03_MVUE.pdf
```
