---
title: "Lesson 3: Unbiased Estimators, MVUE, and Method of Moments"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w2d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 3
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 3: Finding the Best Estimator

## Review: Where We Left Off

**Key concepts from Lesson 2:**

-   Point estimators vs. estimates
-   Bias: $\text{Bias}(\hat{\theta}) = E(\hat{\theta}) - \theta$
-   Unbiasedness: $E(\hat{\theta}) = \theta$
-   MSE = Variance + Bias²

. . .

**Today's Goals:**

-   Understand what makes one unbiased estimator better than another
-   Learn about Minimum Variance Unbiased Estimators (MVUE)
-   Introduce the Method of Moments for constructing estimators
-   Apply these methods to medical research examples

## Motivating Example: Estimating Drug Efficacy

A clinical trial tests the effectiveness of a new antibiotic. Two researchers propose different estimators for the true response rate $p$:

. . .

**Researcher A:** Use $\hat{p}_A = X/n$ (sample proportion)

**Researcher B:** Use $\hat{p}_B = (X + 1)/(n + 2)$ (adjusted proportion)

. . .

Both estimators are functions of the same data. **Which should we use?**

```{r}
# Suppose 15 out of 20 patients respond
n <- 20; x <- 15
p_hat_A <- x / n
p_hat_B <- (x + 1) / (n + 2)
cat("Researcher A's estimate:", p_hat_A, "\n")
cat("Researcher B's estimate:", p_hat_B)
```

# Choosing Among Unbiased Estimators {background-color="#27ae60"}

## The Problem of Multiple Unbiased Estimators

Recall from Lesson 2: There can be **many** unbiased estimators for the same parameter!

. . .

**Example:** For a symmetric distribution with mean $\mu$:

-   Sample mean $\bar{X}$ is unbiased for $\mu$
-   Sample median $\tilde{X}$ is unbiased for $\mu$
-   Trimmed means are unbiased for $\mu$

. . .

::: callout-note
## The Key Question
If two estimators are both unbiased, how do we choose between them?
:::

## Comparing Two Unbiased Estimators: Intuition

```{r}
#| echo: false
#| fig-height: 5
set.seed(123)
true_mu <- 50

comparison_data <- tibble(
  estimator = rep(c("Low Variance (Better)", "High Variance (Worse)"), each = 1000),
  estimate = c(
    rnorm(1000, true_mu, 2),
    rnorm(1000, true_mu, 5)
  )
)

comparison_data |> 
  ggplot(aes(x = estimate, fill = estimator)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = true_mu, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Two Unbiased Estimators: Which is Better?",
       subtitle = "Both centered on true value (red line), but with different spread",
       x = "Estimate", y = "Frequency", fill = "Estimator") +
  scale_fill_manual(values = c("steelblue", "orange")) +
  theme(legend.position = "top")
```

**Answer:** Among unbiased estimators, prefer the one with **smaller variance**.

## The Principle: Minimum Variance Unbiased Estimator

::: callout-important
## Definition (Devore 7.1)

The **Minimum Variance Unbiased Estimator (MVUE)** of a parameter $\theta$ is the estimator $\hat{\theta}$ such that:

1.  $\hat{\theta}$ is unbiased: $E(\hat{\theta}) = \theta$
2.  $\hat{\theta}$ has smaller or equal variance than any other unbiased estimator

If $\hat{\theta}_1$ and $\hat{\theta}_2$ are both unbiased for $\theta$, we prefer the one with smaller variance.
:::

. . .

**Intuition:** Among estimators that are "correct on average," we want the most precise one.

## Example: Uniform Distribution Upper Bound

**Setup:** $X_1, \ldots, X_n \sim \text{Uniform}[0, \theta]$

Two unbiased estimators for $\theta$:

. . .

1.  $\hat{\theta}_1 = 2\bar{X}$ (since $E(X_i) = \theta/2$, so $E(2\bar{X}) = \theta$)

2.  $\hat{\theta}_2 = \frac{n+1}{n} \cdot \max(X_1, \ldots, X_n)$ (corrected maximum)

. . .

**Question:** Which has smaller variance?

## Calculating Variances

For $\hat{\theta}_1 = 2\bar{X}$:
$$\text{Var}(\hat{\theta}_1) = \text{Var}(2\bar{X}) = 4 \cdot \text{Var}(\bar{X}) = 4 \cdot \frac{\sigma^2}{n} = 4 \cdot \frac{\theta^2/12}{n} = \frac{\theta^2}{3n}$$

. . .

For $\hat{\theta}_2 = \frac{n+1}{n} \cdot \max(X_i)$, it can be shown that:
$$\text{Var}(\hat{\theta}_2) = \frac{\theta^2}{n(n+2)}$$

. . .

**Comparison:** For $n > 1$:
$$\frac{\theta^2}{n(n+2)} < \frac{\theta^2}{3n}$$

The corrected maximum $\hat{\theta}_2$ is more efficient!

## Simulation: Comparing the Two Estimators

```{r}
#| code-line-numbers: "|1-3|5-12|14-19"
theta <- 10  # True upper bound
n <- 20
n_sims <- 5000

# Simulate both estimators
uniform_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) runif(n, 0, theta)),
    theta_hat_1 = map_dbl(data, \(d) 2 * mean(d)),
    theta_hat_2 = map_dbl(data, \(d) (n + 1) / n * max(d))
  )

# Compare variances
uniform_sim |> 
  summarize(
    `Var(2*X̄)` = var(theta_hat_1),
    `Var(corrected max)` = var(theta_hat_2),
    `Theoretical Var(2*X̄)` = theta^2 / (3 * n),
    `Theoretical Var(corrected max)` = theta^2 / (n * (n + 2))
  ) |> 
  pivot_longer(everything()) |> 
  mutate(value = round(value, 4))
```

## Visualizing the Comparison

```{r}
#| echo: false
#| fig-height: 5
uniform_sim |> 
  select(theta_hat_1, theta_hat_2) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = case_when(
    Estimator == "theta_hat_1" ~ "2X̄ (higher variance)",
    Estimator == "theta_hat_2" ~ "Corrected Maximum (MVUE)"
  )) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = theta, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Comparing Two Unbiased Estimators for θ",
       subtitle = str_glue("True θ = {theta}, n = {n}"),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("darkgreen", "steelblue")) +
  theme(legend.position = "top")
```

The corrected maximum is **more concentrated** around the true value!

## The MVUE Theorem for Normal Distributions

::: callout-important
## Theorem (Devore 7.1)

Let $X_1, \ldots, X_n$ be a random sample from a **normal distribution** with parameters $\mu$ and $\sigma$. Then:

The sample mean $\bar{X}$ is the **MVUE** for $\mu$.
:::

. . .

**Why this matters:** When sampling from a normal population, no other unbiased estimator of $\mu$ can have smaller variance than $\bar{X}$.

. . .

This is a powerful result! It tells us that $\bar{X}$ is optimal (in the MVUE sense) for normal data.

## Medical Application: Blood Pressure Reduction

A pharmaceutical trial measures systolic blood pressure reduction (mmHg) in patients taking a new medication. Assume reductions are normally distributed.

```{r}
# Blood pressure reductions from 30 patients
bp_data <- c(12.3, 8.7, 15.2, 10.1, 7.4, 14.8, 11.2, 9.5, 13.1, 16.2,
             8.1, 12.5, 10.8, 14.3, 11.7, 9.2, 15.0, 13.4, 7.9, 12.1,
             10.5, 8.3, 14.1, 11.5, 13.8, 9.8, 11.9, 12.7, 10.3, 14.6)

# MVUE for mean reduction
x_bar <- mean(bp_data)
se_x_bar <- sd(bp_data) / sqrt(length(bp_data))

cat("MVUE of mean reduction:", round(x_bar, 2), "mmHg\n")
cat("Standard error:", round(se_x_bar, 2), "mmHg")
```

Since the data is assumed normal, $\bar{X}$ is not just unbiased—it's the **best** unbiased estimator.

## Your Turn: Comparing Estimators

**Exercise:** For a normal distribution with known $\sigma = 10$, which estimator of $\mu$ has smaller variance?

-   $\hat{\mu}_1 = \bar{X}$ (sample mean)
-   $\hat{\mu}_2 = \tilde{X}$ (sample median)

For a normal distribution with $n = 20$:

-   $\text{Var}(\bar{X}) = \sigma^2/n = 100/20 = 5$
-   $\text{Var}(\tilde{X}) \approx \frac{\pi}{2} \cdot \frac{\sigma^2}{n} \approx 1.57 \cdot 5 = 7.85$

. . .

**The sample mean wins!** For normal data, $\bar{X}$ is the MVUE.

## When Might the Median Be Better?

Although $\bar{X}$ is MVUE for normal data, the median can be better for:

-   **Heavy-tailed distributions** (where outliers are more common)
-   **Contaminated data** (mixture of two distributions)

. . .

```{r}
#| echo: false
#| fig-height: 4.5
set.seed(42)
n_sim <- 2000
n <- 30

# Normal vs. contaminated normal
normal_comparison <- tibble(sim = 1:n_sim) |> 
  mutate(
    normal_data = map(sim, \(s) rnorm(n, 100, 15)),
    contaminated_data = map(sim, \(s) {
      # 90% normal, 10% outliers
      x <- rnorm(n, 100, 15)
      outlier_idx <- sample(1:n, 3)
      x[outlier_idx] <- x[outlier_idx] + rnorm(3, 30, 10)
      x
    }),
    mean_normal = map_dbl(normal_data, mean),
    median_normal = map_dbl(normal_data, median),
    mean_contaminated = map_dbl(contaminated_data, mean),
    median_contaminated = map_dbl(contaminated_data, median)
  )

normal_comparison |> 
  select(mean_contaminated, median_contaminated) |> 
  pivot_longer(everything()) |> 
  mutate(name = ifelse(name == "mean_contaminated", "Sample Mean", "Sample Median")) |> 
  ggplot(aes(x = value, fill = name)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = 100, color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "With Contaminated Data: Median Can Be Better",
       subtitle = "10% outliers added to normal data",
       x = "Estimate", y = "Frequency", fill = "Estimator") +
  scale_fill_manual(values = c("steelblue", "darkgreen")) +
  theme(legend.position = "top")
```

# Method of Moments {background-color="#8e44ad"}

## Motivation: How Do We Find Estimators?

So far, we've evaluated estimators. But **how do we construct them** in the first place?

. . .

Two major constructive methods:

1.  **Method of Moments** (today)
2.  **Maximum Likelihood** (Lesson 4)

. . .

::: callout-note
## Key Idea
Both methods give us systematic recipes for deriving estimators from any specified distribution.
:::

## Population vs. Sample Moments

::: callout-important
## Definitions (Devore 7.2)

**Population Moments:**

-   1st moment: $E(X) = \mu$
-   2nd moment: $E(X^2)$
-   kth moment: $E(X^k)$

**Sample Moments:**

-   1st sample moment: $\frac{1}{n}\sum_{i=1}^n X_i = \bar{X}$
-   2nd sample moment: $\frac{1}{n}\sum_{i=1}^n X_i^2$
-   kth sample moment: $\frac{1}{n}\sum_{i=1}^n X_i^k$
:::

## The Method of Moments Principle

::: callout-important
## Method of Moments Estimators (MMEs)

To estimate $m$ unknown parameters $\theta_1, \ldots, \theta_m$:

1.  Express the first $m$ population moments as functions of the parameters
2.  Set each population moment equal to its sample counterpart
3.  Solve the resulting system of equations for $\theta_1, \ldots, \theta_m$
:::

. . .

**The idea:** If the population and sample moments should be "close," we can estimate parameters by equating them.

## Example 1: Exponential Distribution (One Parameter)

**Setup:** $X_1, \ldots, X_n \sim \text{Exponential}(\lambda)$ where $\lambda$ is the rate parameter.

**Goal:** Find the MME for $\lambda$.

. . .

**Step 1:** Population moment: $E(X) = 1/\lambda$

**Step 2:** Set equal to sample moment: $\frac{1}{\lambda} = \bar{X}$

**Step 3:** Solve: $\hat{\lambda} = \frac{1}{\bar{X}}$

. . .

::: callout-tip
## Intuition
If the population mean is $1/\lambda$, and we estimate the population mean with $\bar{X}$, then we estimate $\lambda$ with $1/\bar{X}$.
:::

## Medical Application: Hospital Wait Times

Emergency room wait times are often modeled with an exponential distribution. The parameter $\lambda$ represents the rate (patients served per hour).

```{r}
# Wait times (minutes) for 40 patients
wait_times <- c(12, 45, 8, 23, 67, 15, 31, 42, 19, 56,
                28, 11, 38, 22, 49, 17, 33, 25, 41, 14,
                52, 9, 36, 27, 44, 18, 29, 47, 21, 35,
                13, 39, 24, 51, 16, 32, 46, 20, 37, 26)

# Method of moments estimate for λ (per minute)
lambda_hat <- 1 / mean(wait_times)

cat("Sample mean wait time:", round(mean(wait_times), 2), "minutes\n")
cat("MME of λ:", round(lambda_hat, 4), "per minute\n")
cat("MME of λ:", round(lambda_hat * 60, 2), "per hour")
```

## Verifying with Simulation

```{r}
#| fig-height: 4.5
# Compare empirical distribution to fitted exponential
tibble(wait_time = wait_times) |> 
  ggplot(aes(x = wait_time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, 
                 fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dexp, args = list(rate = lambda_hat),
                color = "red", linewidth = 1.2) +
  labs(title = "ER Wait Times with Fitted Exponential Distribution",
       subtitle = str_glue("MME: λ̂ = {round(lambda_hat, 4)}"),
       x = "Wait Time (minutes)", y = "Density") +
  annotate("text", x = 50, y = 0.025, 
           label = str_glue("Mean = {round(mean(wait_times), 1)} min"),
           color = "red", size = 5)
```

## Example 2: Normal Distribution (Two Parameters)

**Setup:** $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$

**Goal:** Find MMEs for $\mu$ and $\sigma^2$.

. . .

**Population moments:**

-   $E(X) = \mu$
-   $E(X^2) = \mu^2 + \sigma^2$ (variance shortcut formula!)

. . .

**Set equal to sample moments:**

-   $\mu = \bar{X}$
-   $\mu^2 + \sigma^2 = \frac{1}{n}\sum X_i^2$

## Solving for the Normal MMEs

From the two equations:

$$\hat{\mu} = \bar{X}$$

. . .

$$\hat{\sigma}^2 = \frac{1}{n}\sum X_i^2 - \bar{X}^2 = \frac{1}{n}\sum(X_i - \bar{X})^2$$

. . .

::: callout-warning
## Note on Bias
The MME for $\sigma^2$ divides by $n$, not $n-1$. This means the MME is **biased**! 

$$E(\hat{\sigma}^2_{MME}) = \frac{n-1}{n}\sigma^2$$

The usual sample variance $S^2$ with $n-1$ in the denominator is unbiased.
:::

## Example 3: Gamma Distribution (Two Parameters)

**Setup:** $X_1, \ldots, X_n \sim \text{Gamma}(\alpha, \beta)$

**Moments:**

-   $E(X) = \alpha\beta$
-   $E(X^2) = \alpha(\alpha+1)\beta^2$

. . .

**MME equations:**
$$\bar{X} = \alpha\beta$$
$$\frac{1}{n}\sum X_i^2 = \alpha(\alpha+1)\beta^2$$

. . .

**Solutions (after algebra):**
$$\hat{\alpha} = \frac{\bar{X}^2}{\frac{1}{n}\sum X_i^2 - \bar{X}^2} \quad \quad \hat{\beta} = \frac{\frac{1}{n}\sum X_i^2 - \bar{X}^2}{\bar{X}}$$

## Medical Application: Surgery Recovery Times

Recovery times after a particular surgery often follow a gamma distribution. Let's estimate the parameters.

```{r}
# Recovery times (days) for 25 patients
recovery_times <- c(4.2, 6.8, 5.1, 8.3, 7.2, 4.9, 6.1, 9.5, 5.7, 7.8,
                    6.4, 8.1, 5.3, 7.5, 6.9, 4.5, 8.7, 5.9, 7.1, 6.2,
                    9.2, 5.5, 7.4, 6.6, 8.4)

# Calculate sample moments
x_bar <- mean(recovery_times)
second_moment <- mean(recovery_times^2)

# MME for gamma parameters
alpha_hat <- x_bar^2 / (second_moment - x_bar^2)
beta_hat <- (second_moment - x_bar^2) / x_bar

cat("Sample mean:", round(x_bar, 2), "days\n")
cat("MME of α:", round(alpha_hat, 2), "\n")
cat("MME of β:", round(beta_hat, 2), "days")
```

## Visualizing the Gamma Fit

```{r}
#| fig-height: 4.5
tibble(recovery_time = recovery_times) |> 
  ggplot(aes(x = recovery_time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, 
                 fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dgamma, 
                args = list(shape = alpha_hat, scale = beta_hat),
                color = "red", linewidth = 1.2) +
  labs(title = "Surgery Recovery Times with Fitted Gamma Distribution",
       subtitle = str_glue("MME: α̂ = {round(alpha_hat, 2)}, β̂ = {round(beta_hat, 2)}"),
       x = "Recovery Time (days)", y = "Density")
```

## Your Turn: Method of Moments

**Exercise:** Drug concentrations in blood plasma follow a Uniform[0, $\theta$] distribution. From $n = 15$ measurements, you observe $\bar{x} = 4.2$ mg/L.

Find the MME of $\theta$.

. . .

**Solution:**

For Uniform[0, $\theta$]: $E(X) = \theta/2$

Setting equal to sample mean: $\bar{x} = \theta/2$

Therefore: $\hat{\theta} = 2\bar{x} = 2(4.2) = 8.4$ mg/L

## Properties of Method of Moments Estimators

**Advantages:**

-   Simple to compute (usually just algebra)
-   Often has closed-form solutions
-   Usually consistent (converges to true value as $n \to \infty$)

. . .

**Disadvantages:**

-   May not be the most efficient (highest variance)
-   Can sometimes give impossible values
-   May be biased

. . .

::: callout-note
## Coming Up
Maximum likelihood estimators (Lesson 4) generally have better efficiency properties but require more computation.
:::

## Example: When MME Gives Impossible Values

**Problem:** $X_1, \ldots, X_n \sim \text{Uniform}[0, \beta]$

The MME is $\hat{\beta} = 2\bar{X}$.

. . .

**Potential issue:** What if some $X_i > 2\bar{X}$?

```{r}
# Example data where MME fails
problematic_data <- c(1, 1, 1, 9)  # max = 9
beta_mme <- 2 * mean(problematic_data)  # = 6

cat("Data:", problematic_data, "\n")
cat("MME of β:", beta_mme, "\n")
cat("But max observation is:", max(problematic_data), "\n")
cat("This is impossible if β =", beta_mme, "!")
```

The MME suggests $\beta = 6$, but we observed a value of 9!

# Practical Considerations {background-color="#e74c3c"}

## Computing MMEs in R: A Workflow

```{r}
#| code-line-numbers: "|1-3|5-17|19-23"
# General workflow for method of moments
# Step 1: Collect data
data <- c(2.3, 4.1, 3.8, 5.2, 2.9, 4.5, 3.1, 4.8, 3.5, 4.2)

# Step 2: Compute sample moments
compute_sample_moments <- function(x, k = 2) {
  moments <- map_dbl(1:k, \(j) mean(x^j))
  names(moments) <- paste0("m", 1:k)
  moments
}

sample_moments <- compute_sample_moments(data, k = 2)
cat("First moment (mean):", round(sample_moments["m1"], 3), "\n")
cat("Second moment:", round(sample_moments["m2"], 3), "\n")

# Step 3: Solve for parameters (depends on distribution)
# Example: Normal distribution
mu_hat <- sample_moments["m1"]
sigma2_hat <- sample_moments["m2"] - sample_moments["m1"]^2
cat("\nNormal MMEs:\n")
cat("μ̂ =", round(mu_hat, 3), "\n")
cat("σ̂² =", round(sigma2_hat, 3))
```

## Medical Research Application: Drug Clearance

Drug clearance from the body often follows an exponential decay model. A pharmacokinetics study measures the half-life of a new medication.

```{r}
# Half-lives (hours) from 35 patients
half_lives <- c(3.2, 4.1, 2.8, 3.5, 4.8, 3.1, 3.9, 2.6, 4.3, 3.7,
                3.4, 4.0, 2.9, 3.8, 4.5, 3.3, 3.6, 4.2, 3.0, 3.9,
                2.7, 4.1, 3.5, 3.8, 4.0, 3.2, 3.7, 4.4, 3.1, 3.6,
                2.8, 4.0, 3.4, 3.9, 3.5)

# Assume gamma distribution; compute MMEs
x_bar <- mean(half_lives)
var_x <- var(half_lives) * (length(half_lives) - 1) / length(half_lives)

alpha_mme <- x_bar^2 / var_x
beta_mme <- var_x / x_bar

cat("Sample statistics:\n")
cat("  Mean half-life:", round(x_bar, 2), "hours\n")
cat("  Sample variance:", round(var_x, 3), "\n\n")
cat("Gamma MMEs:\n")
cat("  α̂ =", round(alpha_mme, 2), "\n")
cat("  β̂ =", round(beta_mme, 2), "hours")
```

## Connecting Moments to Variance

Recall the variance formula: $\text{Var}(X) = E(X^2) - [E(X)]^2$

. . .

This connects moments to variance!

$$\sigma^2 = \mu_2' - (\mu_1')^2$$

where $\mu_k' = E(X^k)$ is the kth raw moment.

. . .

**For sample moments:**
$$\hat{\sigma}^2_{MME} = \frac{1}{n}\sum X_i^2 - \bar{X}^2 = \frac{1}{n}\sum(X_i - \bar{X})^2$$

This is why the MME for variance divides by $n$, not $n-1$.

## Summary Table: Common MMEs

| Distribution | Parameters | MME Formula |
|--------------|------------|-------------|
| Normal | $\mu$, $\sigma^2$ | $\hat{\mu} = \bar{X}$, $\hat{\sigma}^2 = \frac{1}{n}\sum(X_i-\bar{X})^2$ |
| Exponential | $\lambda$ | $\hat{\lambda} = 1/\bar{X}$ |
| Gamma | $\alpha$, $\beta$ | $\hat{\alpha} = \bar{X}^2/\hat{\sigma}^2$, $\hat{\beta} = \hat{\sigma}^2/\bar{X}$ |
| Uniform[0,$\theta$] | $\theta$ | $\hat{\theta} = 2\bar{X}$ |
| Poisson | $\lambda$ | $\hat{\lambda} = \bar{X}$ |

# Summary and Looking Ahead {background-color="#3498db"}

## Lesson 3 Summary

**Key Concepts:**

1.  **MVUE:** Among unbiased estimators, prefer the one with smallest variance
    -   $\bar{X}$ is MVUE for $\mu$ in normal distributions
    -   The corrected maximum is MVUE for uniform upper bound

2.  **Method of Moments:**
    -   Equate population moments to sample moments
    -   Solve for unknown parameters
    -   Simple but may not be most efficient

3.  **Moments:**
    -   kth population moment: $E(X^k)$
    -   kth sample moment: $\frac{1}{n}\sum X_i^k$

## Comparing What We've Learned

| Criterion | What It Measures | Ideal Value |
|-----------|------------------|-------------|
| Bias | Systematic error | 0 |
| Variance | Random fluctuation | Small |
| MSE | Total error | Small |
| MVUE | Best unbiased | Smallest variance among unbiased |

. . .

**Next lesson:** Maximum Likelihood Estimation—often more efficient than MME!

## Lesson 3 Practice Problems

1.  **(Devore 7.1)** For Uniform[0, $\theta$], compare the MSE of $\hat{\theta}_1 = 2\bar{X}$ and $\hat{\theta}_2 = \frac{n+1}{n}\max(X_i)$ for $n = 5, 10, 20$.

2.  **(Devore 7.2)** Derive the MME for the parameter $\theta$ of the distribution with pdf $f(x; \theta) = (\theta+1)x^\theta$ for $0 \le x \le 1$.

3.  **Medical Application:** Hospital length-of-stay data (days) for 20 patients: 2, 3, 1, 5, 4, 2, 7, 3, 2, 6, 4, 3, 2, 5, 3, 4, 2, 3, 6, 4. Assuming an exponential distribution, find the MME of the rate parameter $\lambda$. What is the estimated mean length of stay?

4.  **Simulation:** Generate 1000 samples of size $n = 25$ from a Gamma(3, 2) distribution. For each sample, compute the MME of $\alpha$ and $\beta$. Compute the bias and variance of each estimator.

## Homework Problems (Devore et al.)

::: nonincremental
**From Chapter 7:**

-   Exercise 7.5: Bias calculations
-   Exercise 7.8: MVUE comparison
-   Exercise 7.25: Method of moments for Weibull
-   Exercise 7.26: Method of moments for general pdf
-   Exercise 7.27: MME derivation and calculation
:::

## Next Lesson Preview

**Lesson 4: Likelihood Theory and Maximum Likelihood Estimation**

-   The likelihood function: $L(\theta) = \prod f(x_i; \theta)$
-   Maximum likelihood principle
-   Finding MLEs analytically
-   Properties of MLEs
-   Comparison with MMEs

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapters 7.1, 7.2
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 6.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape docs/lessons/03_MVUE_Method_of_Moments/03_MVUE_Method_of_Moments.html lessons/03_MVUE_Method_of_Moments/03_MVUE_Method_of_Moments.pdf
```
