---
title: "BSTA 551: Statistical Inference"
subtitle: "Lesson 4: Likelihood Theory and Maximum Likelihood Estimation"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w2d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 4
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 4: Maximum Likelihood Estimation

## Review: Where We Left Off

**Key concepts from Lesson 3:**

-   MVUE: Best unbiased estimator (smallest variance)
-   Method of Moments: Equate population moments to sample moments
-   MMEs are simple but not always most efficient

. . .

**Today's Goals:**

-   Understand the likelihood function and its interpretation
-   Learn to derive Maximum Likelihood Estimators (MLEs)
-   Explore properties of MLEs: invariance, consistency, efficiency
-   Compare MLEs to MMEs

## Motivating Example: Vaccine Efficacy

A clinical trial for a new vaccine enrolls 200 participants. After exposure to the pathogen, 156 remain disease-free.

. . .

**Question:** What is the true efficacy rate $p$ of this vaccine?

. . .

```{r}
n <- 200
successes <- 156
# What value of p makes this outcome most "likely"?
```

**Intuition:** We should choose the value of $p$ that makes our observed data "most likely" to have occurred.

. . .

This is the essence of **Maximum Likelihood Estimation**.

# The Likelihood Function {background-color="#27ae60"}

## From Probability to Likelihood

**Probability vs. Likelihood: A Key Distinction**

. . .

**Probability** (parameters fixed, data varies):

-   Given a known parameter value $\theta$, how likely are different data outcomes?
-   Example: If $p = 0.8$, what's $P(X = 156)$ out of 200?

. . .

**Likelihood** (data fixed, parameters vary):

-   Given our observed data, how "likely" are different parameter values?
-   Example: We observed 156 successes. Which value of $p$ best explains this?

## The Likelihood Function: Definition

::: callout-important
## Definition (Devore 7.2)

Let $X_1, \ldots, X_n$ have a joint distribution that depends on parameter $\theta$. The **likelihood function** $L(\theta)$ is this joint distribution, regarded as a function of $\theta$ (with observed data held fixed).

For a random sample:
$$L(\theta) = \prod_{i=1}^n f(x_i; \theta)$$

where $f$ is the pmf (discrete) or pdf (continuous).
:::

## Visualizing Likelihood: Binomial Example

```{r}
#| echo: false
#| fig-height: 5
# Likelihood function for vaccine trial
n <- 200
x <- 156

likelihood_data <- tibble(
  p = seq(0.01, 0.99, by = 0.001),
  likelihood = dbinom(x, n, p)
)

mle_p <- x / n

likelihood_data |> 
  ggplot(aes(x = p, y = likelihood)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = mle_p, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(x = mle_p, y = dbinom(x, n, mle_p)), color = "red", size = 4) +
  labs(title = "Likelihood Function for Vaccine Efficacy",
       subtitle = str_glue("Observed: {x} successes out of {n} participants"),
       x = "Efficacy Parameter (p)", y = "Likelihood L(p)") +
  annotate("text", x = mle_p + 0.08, y = dbinom(x, n, mle_p) * 0.9,
           label = str_glue("MLE: p̂ = {mle_p}"), color = "red", size = 5)
```

The MLE is the value of $p$ at the **peak** of the likelihood function.

## Maximum Likelihood Estimate: Definition

::: callout-important
## Definition

The **Maximum Likelihood Estimate (MLE)** $\hat{\theta}$ is the value of $\theta$ that maximizes the likelihood function:

$$\hat{\theta} = \arg\max_\theta L(\theta)$$

The MLE is the parameter value that makes the observed data "most likely."
:::

. . .

**Notation:**

-   $\hat{\theta}$ is the MLE (estimator, a random variable)
-   $\hat{\theta}$ is also the mle (estimate, a specific number from data)

## Why Maximum Likelihood?

**Intuitive Appeal:**

-   Choose the parameter that "best explains" our data
-   The data we observed should be highly probable under the true parameter

. . .

**Theoretical Properties** (more details later):

1.  **Consistent:** $\hat{\theta} \to \theta$ as $n \to \infty$
2.  **Asymptotically unbiased:** $E(\hat{\theta}) \approx \theta$ for large $n$
3.  **Asymptotically efficient:** Achieves smallest possible variance for large $n$
4.  **Invariance:** If $\hat{\theta}$ is MLE of $\theta$, then $g(\hat{\theta})$ is MLE of $g(\theta)$

# The Log-Likelihood Function {background-color="#8e44ad"}

## Why Use Log-Likelihood?

The likelihood function is often a **product** of many terms:
$$L(\theta) = \prod_{i=1}^n f(x_i; \theta)$$

. . .

Taking the logarithm converts products to **sums**:
$$\ell(\theta) = \ln L(\theta) = \sum_{i=1}^n \ln f(x_i; \theta)$$

. . .

**Key property:** Since $\ln$ is monotonically increasing:
$$\arg\max_\theta L(\theta) = \arg\max_\theta \ell(\theta)$$

Maximizing the log-likelihood gives the same answer!

## Advantages of Log-Likelihood

**Computational:**

-   Sums are easier to differentiate than products
-   Avoids numerical underflow (very small products)

. . .

**Mathematical:**

-   $\ell'(\theta) = 0$ is often easier to solve
-   Second derivative test is more tractable

. . .

```{r}
#| echo: false
#| fig-height: 4
# Compare likelihood and log-likelihood
likelihood_data |> 
  mutate(log_likelihood = log(likelihood)) |> 
  select(p, likelihood, log_likelihood) |> 
  pivot_longer(-p, names_to = "type", values_to = "value") |> 
  mutate(type = case_when(
    type == "likelihood" ~ "Likelihood L(p)",
    type == "log_likelihood" ~ "Log-Likelihood ℓ(p)"
  )) |> 
  ggplot(aes(x = p, y = value)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = mle_p, color = "red", linetype = "dashed") +
  facet_wrap(~type, scales = "free_y") +
  labs(title = "Likelihood vs. Log-Likelihood", 
       subtitle = "Both maximize at the same point",
       x = "Parameter (p)", y = "Value")
```

## The General MLE Recipe

::: callout-tip
## Finding the MLE

1.  **Write down the likelihood:** $L(\theta) = \prod f(x_i; \theta)$

2.  **Take the logarithm:** $\ell(\theta) = \sum \ln f(x_i; \theta)$

3.  **Differentiate:** Find $\ell'(\theta) = \frac{d\ell}{d\theta}$

4.  **Solve:** Set $\ell'(\theta) = 0$ and solve for $\hat{\theta}$

5.  **Verify:** Check that $\ell''(\hat{\theta}) < 0$ (maximum, not minimum)
:::

# MLE Examples {background-color="#e74c3c"}

## Example 1: Bernoulli Distribution

**Setup:** $X_1, \ldots, X_n \sim \text{Bernoulli}(p)$ (iid)

Let $x = \sum x_i$ be the number of successes.

. . .

**Step 1 - Likelihood:**
$$L(p) = \prod_{i=1}^n p^{x_i}(1-p)^{1-x_i} = p^x(1-p)^{n-x}$$

. . .

**Step 2 - Log-likelihood:**
$$\ell(p) = x\ln(p) + (n-x)\ln(1-p)$$

## Example 1: Bernoulli (continued)

**Step 3 - Differentiate:**
$$\ell'(p) = \frac{x}{p} - \frac{n-x}{1-p}$$

. . .

**Step 4 - Solve:**
$$\frac{x}{p} = \frac{n-x}{1-p}$$
$$x(1-p) = p(n-x)$$
$$x - xp = np - xp$$
$$x = np$$
$$\hat{p} = \frac{x}{n}$$

. . .

**The MLE is the sample proportion!** This matches our intuition.

## Simulation: Verifying the Bernoulli MLE

```{r}
#| code-line-numbers: "|1-3|5-12|14-18"
true_p <- 0.65
n <- 50
n_sims <- 5000

# Simulate many samples and compute MLE
mle_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    successes = rbinom(n_sims, size = n, prob = true_p),
    mle = successes / n
  )

# Check unbiasedness and variance
mle_sim |> 
  summarize(
    `True p` = true_p,
    `Mean of MLEs` = mean(mle),
    `Bias` = mean(mle) - true_p,
    `Variance of MLEs` = var(mle),
    `Theoretical Var` = true_p * (1 - true_p) / n
  ) |> 
  pivot_longer(everything()) |> 
  mutate(value = round(value, 5))
```

## Example 2: Exponential Distribution

**Setup:** $X_1, \ldots, X_n \sim \text{Exponential}(\lambda)$ where $\lambda$ is the rate parameter.

**Medical Context:** Time until symptom relief (hours) for patients taking a new pain medication.

. . .

**Step 1 - Likelihood:**
$$L(\lambda) = \prod_{i=1}^n \lambda e^{-\lambda x_i} = \lambda^n e^{-\lambda \sum x_i}$$

**Step 2 - Log-likelihood:**
$$\ell(\lambda) = n\ln(\lambda) - \lambda \sum x_i$$

## Example 2: Exponential (continued)

**Step 3 - Differentiate:**
$$\ell'(\lambda) = \frac{n}{\lambda} - \sum x_i$$

. . .

**Step 4 - Solve:**
$$\frac{n}{\lambda} = \sum x_i$$
$$\hat{\lambda} = \frac{n}{\sum x_i} = \frac{1}{\bar{X}}$$

. . .

**Note:** This is the same as the MME! (Not always the case.)

## Medical Application: Pain Relief Duration

```{r}
# Time to pain relief (hours) for 30 patients
relief_times <- c(2.1, 3.5, 1.8, 4.2, 2.7, 1.5, 3.8, 2.3, 4.1, 2.9,
                  1.9, 3.2, 2.5, 3.9, 2.1, 4.5, 1.7, 3.4, 2.8, 3.1,
                  2.4, 3.7, 1.6, 4.0, 2.6, 3.3, 2.0, 3.6, 2.2, 3.0)

# MLE of rate parameter
lambda_mle <- 1 / mean(relief_times)

cat("Sample mean relief time:", round(mean(relief_times), 2), "hours\n")
cat("MLE of λ:", round(lambda_mle, 4), "per hour\n")
cat("Estimated median relief time:", round(log(2) / lambda_mle, 2), "hours")
```

## Visualizing the Exponential MLE

```{r}
#| fig-height: 4.5
tibble(time = relief_times) |> 
  ggplot(aes(x = time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 12,
                 fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dexp, args = list(rate = lambda_mle),
                color = "red", linewidth = 1.2) +
  labs(title = "Time to Pain Relief with Fitted Exponential",
       subtitle = str_glue("MLE: λ̂ = {round(lambda_mle, 4)}"),
       x = "Time to Relief (hours)", y = "Density") +
  annotate("text", x = 4, y = 0.3, 
           label = str_glue("Mean = {round(1/lambda_mle, 2)} hours"),
           color = "red", size = 5)
```

## Example 3: Normal Distribution (Two Parameters)

**Setup:** $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$

. . .

**Likelihood:**
$$L(\mu, \sigma) = \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}} e^{-(x_i-\mu)^2/(2\sigma^2)}$$

**Log-likelihood:**
$$\ell(\mu, \sigma) = -\frac{n}{2}\ln(2\pi) - n\ln(\sigma) - \frac{1}{2\sigma^2}\sum(x_i - \mu)^2$$

## Normal MLEs: The Result

Taking partial derivatives and solving:

$$\hat{\mu} = \bar{X}$$

$$\hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^n (X_i - \bar{X})^2$$

. . .

::: callout-warning
## Important Note
The MLE of $\sigma^2$ divides by $n$, not $n-1$!

This means the MLE is **biased**: $E(\hat{\sigma}^2_{MLE}) = \frac{n-1}{n}\sigma^2$

The bias is small for large $n$, and we can correct it if desired.
:::

## Your Turn: Deriving the Normal MLE

**Exercise:** Verify that $\hat{\mu} = \bar{X}$ is the MLE for $\mu$ (treating $\sigma$ as known).

**Steps:**

1.  Write $\ell(\mu) = -\frac{n}{2}\ln(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum(x_i - \mu)^2$

2.  Differentiate: $\ell'(\mu) = ?$

3.  Set equal to zero and solve for $\hat{\mu}$

. . .

**Solution:**
$$\ell'(\mu) = \frac{1}{\sigma^2}\sum(x_i - \mu) = \frac{n\bar{x} - n\mu}{\sigma^2}$$
Setting to zero: $n\bar{x} = n\mu$, so $\hat{\mu} = \bar{X}$ ✓

## Example 4: Uniform Distribution (A Complication)

**Setup:** $X_1, \ldots, X_n \sim \text{Uniform}[0, \theta]$

**Likelihood:**
$$L(\theta) = \prod_{i=1}^n \frac{1}{\theta} \cdot \mathbf{1}(0 \le x_i \le \theta) = \frac{1}{\theta^n} \cdot \mathbf{1}(\max(x_i) \le \theta)$$

. . .

**The problem:** We can't use calculus! The likelihood is:

-   $L(\theta) = 0$ if $\theta < \max(x_i)$
-   $L(\theta) = 1/\theta^n$ if $\theta \ge \max(x_i)$

The function decreases as $\theta$ increases (when positive).

## Uniform MLE: Graphical Solution

```{r}
#| echo: false
#| fig-height: 5
# Uniform likelihood
set.seed(42)
sample_data <- runif(10, 0, 5)
x_max <- max(sample_data)

uniform_lik <- tibble(
  theta = seq(0, 8, by = 0.01),
  likelihood = ifelse(theta < x_max, 0, 1/theta^10)
)

uniform_lik |> 
  ggplot(aes(x = theta, y = likelihood)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = x_max, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(x = x_max, y = 1/x_max^10), color = "red", size = 4) +
  labs(title = "Likelihood Function for Uniform[0, θ]",
       subtitle = "Maximum occurs at the boundary!",
       x = "θ", y = "Likelihood L(θ)") +
  annotate("text", x = x_max + 0.5, y = 1/x_max^10 * 0.8,
           label = str_glue("MLE: θ̂ = max(Xᵢ) = {round(x_max, 2)}"),
           color = "red", size = 5)
```

**The MLE is** $\hat{\theta} = \max(X_1, \ldots, X_n)$

# Properties of MLEs {background-color="#3498db"}

## Property 1: Invariance Principle

::: callout-important
## MLE Invariance Principle (Devore 7.2)

If $\hat{\theta}$ is the MLE of $\theta$, then for any function $g(\cdot)$:

$$\widehat{g(\theta)} = g(\hat{\theta})$$

The MLE of a function of a parameter is that function of the MLE.
:::

. . .

**Example:** If $\hat{\mu} = \bar{X}$ and $\hat{\sigma} = \sqrt{\frac{1}{n}\sum(X_i-\bar{X})^2}$ are MLEs for normal parameters, then:

-   MLE of $\sigma^2$ is $\hat{\sigma}^2$
-   MLE of the coefficient of variation $\sigma/\mu$ is $\hat{\sigma}/\hat{\mu}$

## Applying Invariance: Medical Example

A clinical trial measures cholesterol levels (mg/dL). Assume normal distribution.

```{r}
cholesterol <- c(195, 210, 185, 220, 175, 230, 200, 188, 215, 205,
                 192, 218, 182, 225, 198, 212, 190, 208, 178, 222)

# MLEs for μ and σ
mu_mle <- mean(cholesterol)
sigma_mle <- sqrt(mean((cholesterol - mu_mle)^2))  # Note: dividing by n

# By invariance, MLE of P(X > 240) is:
prob_high <- 1 - pnorm(240, mean = mu_mle, sd = sigma_mle)

cat("MLE of μ:", round(mu_mle, 2), "mg/dL\n")
cat("MLE of σ:", round(sigma_mle, 2), "mg/dL\n")
cat("MLE of P(cholesterol > 240):", round(prob_high, 4))
```

## Property 2: Consistency

::: callout-important
## Theorem
Under general regularity conditions, MLEs are **consistent**:

$$\hat{\theta} \xrightarrow{P} \theta \quad \text{as } n \to \infty$$

The MLE converges in probability to the true parameter value.
:::

. . .

**Intuition:** With more data, our estimate gets closer to the truth.

## Demonstrating Consistency

```{r}
#| fig-height: 4.5
true_lambda <- 0.5
sample_sizes <- c(10, 30, 100, 300, 1000)

consistency_sim <- map_dfr(sample_sizes, function(n) {
  tibble(
    n = n,
    sim = 1:1000,
    mle = map_dbl(sim, \(s) 1 / mean(rexp(n, rate = true_lambda)))
  )
})

consistency_sim |> 
  ggplot(aes(x = mle, fill = factor(n))) +
  geom_histogram(bins = 40, alpha = 0.7) +
  geom_vline(xintercept = true_lambda, color = "red", linewidth = 1.2) +
  facet_wrap(~n, scales = "free_y", labeller = label_both) +
  labs(title = "Consistency of Exponential MLE",
       subtitle = "Distribution concentrates around true λ = 0.5 as n increases",
       x = "MLE of λ", y = "Frequency") +
  theme(legend.position = "none")
```

## Property 3: Asymptotic Normality & Efficiency

::: callout-important
## Theorem (Devore 7.4)
Under regularity conditions, for large $n$:

1.  $\hat{\theta}$ is **approximately unbiased**: $E(\hat{\theta}) \approx \theta$

2.  $\hat{\theta}$ is **approximately normal**: $\hat{\theta} \dot{\sim} N(\theta, \text{Var}(\hat{\theta}))$

3.  $\hat{\theta}$ is **asymptotically efficient**: It achieves (approximately) the smallest possible variance among all unbiased estimators.
:::

. . .

**In practice:** For large samples, the MLE is approximately the MVUE!

## Comparing MLE and MME

When do they agree? When do they differ?

| Distribution | MLE | MME | Same? |
|--------------|-----|-----|-------|
| Bernoulli($p$) | $\bar{X}$ | $\bar{X}$ | ✓ |
| Exponential($\lambda$) | $1/\bar{X}$ | $1/\bar{X}$ | ✓ |
| Normal($\mu$, $\sigma^2$) | $\bar{X}$, $\frac{1}{n}\sum(X_i-\bar{X})^2$ | $\bar{X}$, $\frac{1}{n}\sum(X_i-\bar{X})^2$ | ✓ |
| Uniform[0, $\theta$] | $\max(X_i)$ | $2\bar{X}$ | ✗ |
| Weibull | Numerical | Different | ✗ |

## When MLE ≠ MME: Which is Better?

For Uniform[0, $\theta$]:

```{r}
theta <- 10
n <- 20
n_sims <- 5000

uniform_comparison <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) runif(n, 0, theta)),
    mle = map_dbl(data, max),
    mme = map_dbl(data, \(d) 2 * mean(d))
  )

uniform_comparison |> 
  summarize(
    `MLE Bias` = mean(mle) - theta,
    `MME Bias` = mean(mme) - theta,
    `MLE Variance` = var(mle),
    `MME Variance` = var(mme),
    `MLE MSE` = mean((mle - theta)^2),
    `MME MSE` = mean((mme - theta)^2)
  ) |> 
  pivot_longer(everything()) |> 
  mutate(value = round(value, 4))
```

**The MLE has smaller MSE** despite being biased!

## Visualizing MLE vs MME

```{r}
#| echo: false
#| fig-height: 5
uniform_comparison |> 
  select(mle, mme) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = toupper(Estimator)) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = theta, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "MLE vs MME for Uniform[0, θ]",
       subtitle = str_glue("True θ = {theta}; MLE is biased but has lower variance"),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("steelblue", "darkgreen")) +
  theme(legend.position = "top")
```

## Computing MLEs in R

**Using built-in functions:**

```{r}
# For distributions with known MLEs
normal_data <- rnorm(100, mean = 50, sd = 10)
cat("Normal MLE for μ:", mean(normal_data), "\n")
cat("Normal MLE for σ:", sd(normal_data) * sqrt((100-1)/100))
```

. . .

**Using optimization for complex cases:**

```{r}
# General approach: maximize log-likelihood
neg_log_lik <- function(lambda, data) {
  -sum(dexp(data, rate = lambda, log = TRUE))
}

exp_data <- rexp(50, rate = 0.5)
result <- optimize(neg_log_lik, interval = c(0.01, 5), data = exp_data)
cat("\nExponential MLE for λ:", round(result$minimum, 4))
```

## Medical Application: Survival Analysis

Cancer survival times are often modeled with exponential or Weibull distributions.

```{r}
# Survival times (months) for 40 patients
survival_times <- c(8.2, 15.3, 12.1, 6.8, 22.5, 9.4, 18.7, 11.3, 7.5, 25.1,
                    14.2, 10.8, 5.3, 19.6, 13.4, 8.9, 16.2, 11.7, 6.1, 21.3,
                    12.8, 9.1, 17.5, 7.8, 24.2, 10.3, 14.9, 8.4, 20.1, 11.9,
                    6.5, 18.3, 13.1, 9.7, 15.8, 7.2, 22.8, 12.4, 8.1, 19.2)

# Fit exponential distribution
lambda_mle <- 1 / mean(survival_times)

# Estimated survival probability at 12 months
surv_12 <- exp(-lambda_mle * 12)

cat("MLE of λ:", round(lambda_mle, 4), "per month\n")
cat("Estimated median survival:", round(log(2)/lambda_mle, 1), "months\n")
cat("Estimated P(survive > 12 months):", round(surv_12, 3))
```

## Fitting More Complex Models

```{r}
#| message: false
# Using MASS package for Weibull MLE
library(MASS)

# Fit Weibull distribution to survival data
weibull_fit <- fitdistr(survival_times, "weibull")
print(weibull_fit)

# Compare fits
cat("\nExponential mean:", round(1/lambda_mle, 2), "months\n")
cat("Weibull mean:", round(weibull_fit$estimate["scale"] * 
    gamma(1 + 1/weibull_fit$estimate["shape"]), 2), "months")
```

## Your Turn: MLE Practice

**Exercise:** A diagnostic test is administered to 80 patients. 64 test positive. Assuming the number positive follows a Binomial distribution, find:

1.  The MLE of the sensitivity $p$
2.  The MLE of the odds of a positive result, $p/(1-p)$
3.  The MLE of $P(\text{at least 60 positive in next 80 patients})$

. . .

**Solutions:**

1.  $\hat{p} = 64/80 = 0.80$
2.  By invariance: $\widehat{p/(1-p)} = 0.80/0.20 = 4.0$
3.  By invariance: $\widehat{P(X \ge 60)} = P(X \ge 60 | p = 0.80)$

```{r}
1 - pbinom(59, size = 80, prob = 0.80)  # P(X >= 60)
```

# Summary and Looking Ahead {background-color="#2c3e50"}

## MLE Summary Table

| Aspect | Description |
|--------|-------------|
| **Definition** | $\hat{\theta} = \arg\max L(\theta)$ |
| **Method** | Maximize likelihood (or log-likelihood) |
| **Invariance** | MLE of $g(\theta)$ is $g(\hat{\theta})$ |
| **Consistency** | $\hat{\theta} \to \theta$ as $n \to \infty$ |
| **Efficiency** | Approximately MVUE for large $n$ |
| **Bias** | May be biased for small $n$ |

## Lesson 4 Summary

**Key Concepts:**

1.  **Likelihood Function:** $L(\theta) = \prod f(x_i; \theta)$
    -   Probability of observed data as function of parameters

2.  **Log-Likelihood:** $\ell(\theta) = \sum \ln f(x_i; \theta)$
    -   Easier to maximize; same solution

3.  **MLE Properties:**
    -   Invariance, consistency, asymptotic normality
    -   Approximately unbiased and efficient for large $n$

4.  **MLE vs MME:**
    -   Often the same; MLE usually more efficient when different

## Lesson 4 Practice Problems

1.  **(Devore 7.25)** Derive the MLE for the parameter $\theta$ of the pdf $f(x; \theta) = (\theta+1)x^\theta$, $0 \le x \le 1$.

2.  **(Devore 7.31)** For normal measurement errors with known mean 0, derive the MLE of the variance $\sigma^2$.

3.  **(Devore 7.34)** For the shifted exponential $f(x; \lambda, \theta) = \lambda e^{-\lambda(x-\theta)}$, $x \ge \theta$, find the MLEs of both parameters.

4.  **Medical Application:** In a vaccine trial, 45 out of 50 participants develop antibodies. Find the MLE of the seroconversion rate and construct an approximate 95% CI using the asymptotic normality of the MLE.

## Homework Problems (Devore et al.)

::: nonincremental
**From Chapter 7:**

-   Exercise 7.27: MLE derivation and application
-   Exercise 7.29: Poisson MLE
-   Exercise 7.31: Normal variance MLE
-   Exercise 7.32: Gamma MLE equations
-   Exercise 7.35: Exponential with censoring
:::

## Looking Ahead

**Upcoming Topics:**

-   Sufficient statistics (Chapter 7.3)
-   Fisher Information (Chapter 7.4)
-   Large-sample properties of MLEs
-   Confidence intervals (Chapter 8)

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 7.2
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 6.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape docs/lessons/04_MLE/04_MLE.html lessons/04_MLE/04_MLE.pdf
```
