---
title: "Lesson 4: Method of Moments and Maximum Likelihood Estimation"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w2d1`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 4
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 4: Constructing Estimators

## Review: Where We Left Off

**Key concepts from Lesson 3:**

-   MVUE: Best unbiased estimator (smallest variance)
-   Likelihood function and log-likelihood
-   Fisher Information: $I(\theta) = E[-\ell''(\theta)]$
-   Cramér-Rao Lower Bound: $\text{Var}(T) \geq 1/I_n(\theta)$
-   Efficient estimators achieve the CRLB

. . .

**Today's Goals:**

-   Learn to construct estimators systematically
-   Method of Moments: Simple, intuitive approach
-   Maximum Likelihood: Powerful, principled approach
-   Compare the two methods

## Motivating Question

We've learned how to **evaluate** estimators (bias, variance, efficiency).

But how do we **construct** them in the first place?

. . .

**Two major methods:**

1.  **Method of Moments (MME):** Match population and sample moments
2.  **Maximum Likelihood (MLE):** Find parameter that makes data most "likely"

. . .

::: callout-note
## Key Insight
Both methods give us systematic recipes for deriving estimators from any specified distribution.
:::

# Method of Moments {background-color="#3498db"}

## Population vs. Sample Moments

::: callout-important
## Definitions (Devore 7.2)

**Population Moments:**

-   1st moment: $E(X) = \mu_1'$
-   2nd moment: $E(X^2) = \mu_2'$
-   kth moment: $E(X^k) = \mu_k'$

**Sample Moments:**

-   1st sample moment: $\frac{1}{n}\sum_{i=1}^n X_i = \bar{X}$
-   2nd sample moment: $\frac{1}{n}\sum_{i=1}^n X_i^2$
-   kth sample moment: $\frac{1}{n}\sum_{i=1}^n X_i^k$
:::

## The Method of Moments Principle

::: callout-important
## Method of Moments Estimators (MMEs)

To estimate $m$ unknown parameters $\theta_1, \ldots, \theta_m$:

1.  Express the first $m$ population moments as functions of the parameters
2.  Set each population moment equal to its sample counterpart
3.  Solve the resulting system of equations for $\theta_1, \ldots, \theta_m$
:::

. . .

**The idea:** If the population and sample moments should be "close," we can estimate parameters by equating them.

## Example 1: Exponential Distribution (One Parameter)

**Setup:** $X_1, \ldots, X_n \sim \text{Exponential}(\lambda)$ where $\lambda$ is the rate parameter.

**Goal:** Find the MME for $\lambda$.

. . .

**Step 1:** Population moment: $E(X) = 1/\lambda$

**Step 2:** Set equal to sample moment: $\frac{1}{\lambda} = \bar{X}$

**Step 3:** Solve: $\hat{\lambda}_{MME} = \frac{1}{\bar{X}}$

. . .

::: callout-tip
## Intuition
If the population mean is $1/\lambda$, and we estimate the population mean with $\bar{X}$, then we estimate $\lambda$ with $1/\bar{X}$.
:::

## Medical Application: Hospital Wait Times

Emergency room wait times are often modeled with an exponential distribution.

```{r}
# Wait times (minutes) for 40 patients
wait_times <- c(12, 45, 8, 23, 67, 15, 31, 42, 19, 56,
                28, 11, 38, 22, 49, 17, 33, 25, 41, 14,
                52, 9, 36, 27, 44, 18, 29, 47, 21, 35,
                13, 39, 24, 51, 16, 32, 46, 20, 37, 26)

# Method of moments estimate for λ (per minute)
lambda_hat <- 1 / mean(wait_times)

cat("Sample mean wait time:", round(mean(wait_times), 2), "minutes\n")
cat("MME of λ:", round(lambda_hat, 4), "per minute\n")
cat("MME of λ:", round(lambda_hat * 60, 2), "per hour")
```

## Visualizing the Exponential Fit: Code

```{r}
#| label: exp-fit-code
#| eval: false
# Compare empirical distribution to fitted exponential
tibble(wait_time = wait_times) |> 
  ggplot(aes(x = wait_time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, 
                 fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dexp, args = list(rate = lambda_hat),
                color = "red", linewidth = 1.2) +
  labs(title = "ER Wait Times with Fitted Exponential Distribution",
       subtitle = str_glue("MME: λ̂ = {round(lambda_hat, 4)}"),
       x = "Wait Time (minutes)", y = "Density") +
  annotate("text", x = 50, y = 0.025, 
           label = str_glue("Mean = {round(mean(wait_times), 1)} min"),
           color = "red", size = 5)
```

## Visualizing the Exponential Fit: Result

```{r}
#| label: exp-fit-plot
#| echo: false
#| fig-height: 6
# Compare empirical distribution to fitted exponential
tibble(wait_time = wait_times) |> 
  ggplot(aes(x = wait_time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, 
                 fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dexp, args = list(rate = lambda_hat),
                color = "red", linewidth = 1.2) +
  labs(title = "ER Wait Times with Fitted Exponential Distribution",
       subtitle = str_glue("MME: λ̂ = {round(lambda_hat, 4)}"),
       x = "Wait Time (minutes)", y = "Density") +
  annotate("text", x = 50, y = 0.025, 
           label = str_glue("Mean = {round(mean(wait_times), 1)} min"),
           color = "red", size = 5)
```

## Example 2: Normal Distribution (Two Parameters)

**Setup:** $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$

**Goal:** Find MMEs for $\mu$ and $\sigma^2$.

. . .

**Population moments:**

-   $E(X) = \mu$
-   $E(X^2) = \mu^2 + \sigma^2$ (variance shortcut formula!)

. . .

**Set equal to sample moments:**

-   $\mu = \bar{X}$
-   $\mu^2 + \sigma^2 = \frac{1}{n}\sum X_i^2$

## Solving for the Normal MMEs

From the two equations:

$$\hat{\mu}_{MME} = \bar{X}$$

. . .

$$\hat{\sigma}^2_{MME} = \frac{1}{n}\sum X_i^2 - \bar{X}^2 = \frac{1}{n}\sum(X_i - \bar{X})^2$$

. . .

::: callout-warning
## Note on Bias
The MME for $\sigma^2$ divides by $n$, not $n-1$. This means the MME is **biased**! 

$$E(\hat{\sigma}^2_{MME}) = \frac{n-1}{n}\sigma^2$$

The usual sample variance $S^2$ with $n-1$ in the denominator is unbiased.
:::

## Example 3: Gamma Distribution (Two Parameters)

**Setup:** $X_1, \ldots, X_n \sim \text{Gamma}(\alpha, \beta)$

**Moments:**

-   $E(X) = \alpha\beta$
-   $E(X^2) = \alpha(\alpha+1)\beta^2$

. . .

**MME equations:**
$$\bar{X} = \alpha\beta$$
$$\frac{1}{n}\sum X_i^2 = \alpha(\alpha+1)\beta^2$$

. . .

**Solutions (after algebra):**
$$\hat{\alpha} = \frac{\bar{X}^2}{\frac{1}{n}\sum X_i^2 - \bar{X}^2} = \frac{\bar{X}^2}{\hat{\sigma}^2} \quad \quad \hat{\beta} = \frac{\frac{1}{n}\sum X_i^2 - \bar{X}^2}{\bar{X}} = \frac{\hat{\sigma}^2}{\bar{X}}$$

## Medical Application: Surgery Recovery Times

Recovery times after surgery often follow a gamma distribution.

```{r}
# Recovery times (days) for 25 patients
recovery_times <- c(4.2, 6.8, 5.1, 8.3, 7.2, 4.9, 6.1, 9.5, 5.7, 7.8,
                    6.4, 8.1, 5.3, 7.5, 6.9, 4.5, 8.7, 5.9, 7.1, 6.2,
                    9.2, 5.5, 7.4, 6.6, 8.4)

# Calculate sample moments
x_bar <- mean(recovery_times)
second_moment <- mean(recovery_times^2)

# MME for gamma parameters
alpha_hat <- x_bar^2 / (second_moment - x_bar^2)
beta_hat <- (second_moment - x_bar^2) / x_bar

cat("Sample mean:", round(x_bar, 2), "days\n")
cat("MME of α:", round(alpha_hat, 2), "\n")
cat("MME of β:", round(beta_hat, 2), "days")
```

## Visualizing the Gamma Fit: Code

```{r}
#| label: gamma-fit-code
#| eval: false
tibble(recovery_time = recovery_times) |> 
  ggplot(aes(x = recovery_time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, 
                 fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dgamma, 
                args = list(shape = alpha_hat, scale = beta_hat),
                color = "red", linewidth = 1.2) +
  labs(title = "Surgery Recovery Times with Fitted Gamma Distribution",
       subtitle = str_glue("MME: α̂ = {round(alpha_hat, 2)}, β̂ = {round(beta_hat, 2)}"),
       x = "Recovery Time (days)", y = "Density")
```

## Visualizing the Gamma Fit: Result

```{r}
#| label: gamma-fit-plot
#| echo: false
#| fig-height: 6
tibble(recovery_time = recovery_times) |> 
  ggplot(aes(x = recovery_time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, 
                 fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dgamma, 
                args = list(shape = alpha_hat, scale = beta_hat),
                color = "red", linewidth = 1.2) +
  labs(title = "Surgery Recovery Times with Fitted Gamma Distribution",
       subtitle = str_glue("MME: α̂ = {round(alpha_hat, 2)}, β̂ = {round(beta_hat, 2)}"),
       x = "Recovery Time (days)", y = "Density")
```

## Your Turn: Method of Moments

**Exercise:** Drug concentrations in blood plasma follow a Uniform[0, $\theta$] distribution. From $n = 15$ measurements, you observe $\bar{x} = 4.2$ mg/L.

Find the MME of $\theta$.

. . .

**Solution:**

For Uniform[0, $\theta$]: $E(X) = \theta/2$

Setting equal to sample mean: $\bar{x} = \theta/2$

Therefore: $\hat{\theta}_{MME} = 2\bar{x} = 2(4.2) = 8.4$ mg/L

## Properties of Method of Moments Estimators

**Advantages:**

-   Simple to compute (usually just algebra)
-   Often has closed-form solutions
-   Usually consistent (converges to true value as $n \to \infty$)

. . .

**Disadvantages:**

-   May not be efficient (doesn't achieve CRLB)
-   Can sometimes give impossible values
-   May be biased

. . .

::: callout-note
## Coming Up Next
Maximum likelihood estimators are often more efficient than MMEs.
:::

# Maximum Likelihood Estimation {background-color="#27ae60"}

## Motivating Example: Vaccine Efficacy

A clinical trial for a new vaccine enrolls 200 participants. After exposure to the pathogen, 156 remain disease-free.

. . .

**Question:** What is the true efficacy rate $p$ of this vaccine?

. . .

```{r}
n <- 200
successes <- 156
# What value of p makes this outcome most "likely"?
```

**Intuition:** We should choose the value of $p$ that makes our observed data "most likely" to have occurred.

. . .

This is the essence of **Maximum Likelihood Estimation**.

## From Probability to Likelihood

**Probability vs. Likelihood: A Key Distinction**

. . .

**Probability** (parameters fixed, data varies):

-   Given a known parameter value $\theta$, how likely are different data outcomes?
-   Example: If $\theta = 0.8$ for $X~$Binomial(200,$\theta$), what's $P(X = 156)$?

. . .

**Likelihood** (data fixed, parameters vary):

-   Given our observed data, how "likely" are different parameter values?
-   Example: We observed 156 successes. Which value of $p$ best explains this?

## The Likelihood Function: Definition

::: callout-important
## Definition (Devore 7.2)

Let $X_1, \ldots, X_n$ have a joint distribution that depends on parameter $\theta$. The **likelihood function** $L(\theta)$ is this joint distribution, regarded as a function of $\theta$ (with observed data held fixed).

For a random sample:
$$L(\theta) = \prod_{i=1}^n f(x_i; \theta)$$

where $f$ is the pmf (discrete) or pdf (continuous).
:::

## Visualizing Likelihood: Code

```{r}
#| label: likelihood-code
#| eval: false
# Likelihood function for vaccine trial
n <- 200
x <- 156

likelihood_data <- tibble(
  p = seq(0.01, 0.99, by = 0.001),
  likelihood = dbinom(x, n, p)
)

mle_p <- x / n

likelihood_data |> 
  ggplot(aes(x = p, y = likelihood)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = mle_p, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(x = mle_p, y = dbinom(x, n, mle_p)), color = "red", size = 4) +
  labs(title = "Likelihood Function for Vaccine Efficacy",
       subtitle = str_glue("Observed: {x} successes out of {n} participants"),
       x = "Efficacy Parameter (p)", y = "Likelihood L(p)") +
  annotate("text", x = mle_p + 0.08, y = dbinom(x, n, mle_p) * 0.9,
           label = str_glue("MLE: p̂ = {mle_p}"), color = "red", size = 5)
```

## Visualizing Likelihood: Result

```{r}
#| label: likelihood-plot
#| echo: false
#| fig-height: 6
# Likelihood function for vaccine trial
n <- 200
x <- 156

likelihood_data <- tibble(
  p = seq(0.01, 0.99, by = 0.001),
  likelihood = dbinom(x, n, p)
)

mle_p <- x / n

likelihood_data |> 
  ggplot(aes(x = p, y = likelihood)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = mle_p, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(x = mle_p, y = dbinom(x, n, mle_p)), color = "red", size = 4) +
  labs(title = "Likelihood Function for Vaccine Efficacy",
       subtitle = str_glue("Observed: {x} successes out of {n} participants"),
       x = "Efficacy Parameter (p)", y = "Likelihood L(p)") +
  annotate("text", x = mle_p + 0.08, y = dbinom(x, n, mle_p) * 0.9,
           label = str_glue("MLE: p̂ = {mle_p}"), color = "red", size = 5)
```

The MLE is the value of $p$ at the **peak** of the likelihood function.

## Maximum Likelihood Estimate: Definition

::: callout-important
## Definition

The **Maximum Likelihood Estimate (MLE)** $\hat{\theta}$ is the value of $\theta$ that maximizes the likelihood function:

$$\hat{\theta} = \arg\max_\theta L(\theta)$$

The MLE is the parameter value that makes the observed data "most likely."
:::

. . .

**Notation:**

-   $\hat{\theta}$ is the MLE (estimator, a random variable)
-   $\hat{\theta}$ is also the observed mle (estimate, a specific number from data)

## Why Maximum Likelihood?

**Intuitive Appeal:**

-   Choose the parameter that "best explains" our data
-   The data we observed should be highly probable under the true parameter

. . .

**Theoretical Properties** (more details later):

1.  **Consistent:** $\hat{\theta} \to \theta$ as $n \to \infty$
2.  **Asymptotically unbiased:** $E(\hat{\theta}) \approx \theta$ for large $n$
3.  **Asymptotically efficient:** Achieves smallest possible variance for large $n$
4.  **Invariance:** If $\hat{\theta}$ is MLE of $\theta$, then $g(\hat{\theta})$ is MLE of $g(\theta)$

# The Log-Likelihood Function {background-color="#8e44ad"}

## Why Use Log-Likelihood?

The likelihood function is often a **product** of many terms:
$$L(\theta) = \prod_{i=1}^n f(x_i; \theta)$$

. . .

Taking the logarithm converts products to **sums**:
$$\ell(\theta) = \ln L(\theta) = \sum_{i=1}^n \ln f(x_i; \theta)$$

. . .

**Key property:** Since $\ln$ is monotonically increasing:
$$\arg\max_\theta L(\theta) = \arg\max_\theta \ell(\theta)$$

Maximizing the log-likelihood gives the same answer!

## Advantages of Log-Likelihood

**Computational:**

-   Sums are easier to differentiate than products
-   Avoids numerical underflow (very small products)

. . .

**Mathematical:**

-   $\ell'(\theta) = 0$ is often easier to solve
-   Second derivative test is more tractable

## Likelihood vs. Log-Likelihood: Code

```{r}
#| label: loglik-compare-code
#| eval: false
# Compare likelihood and log-likelihood
likelihood_data |> 
  mutate(log_likelihood = log(likelihood)) |> 
  select(p, likelihood, log_likelihood) |> 
  pivot_longer(-p, names_to = "type", values_to = "value") |> 
  mutate(type = case_when(
    type == "likelihood" ~ "Likelihood L(p)",
    type == "log_likelihood" ~ "Log-Likelihood ℓ(p)"
  )) |> 
  ggplot(aes(x = p, y = value)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = mle_p, color = "red", linetype = "dashed") +
  facet_wrap(~type, scales = "free_y") +
  labs(title = "Likelihood vs. Log-Likelihood", 
       subtitle = "Both maximize at the same point",
       x = "Parameter (p)", y = "Value")
```

## Likelihood vs. Log-Likelihood: Result

```{r}
#| label: loglik-compare-plot
#| echo: false
#| fig-height: 5.5
# Compare likelihood and log-likelihood
likelihood_data |> 
  mutate(log_likelihood = log(likelihood)) |> 
  select(p, likelihood, log_likelihood) |> 
  pivot_longer(-p, names_to = "type", values_to = "value") |> 
  mutate(type = case_when(
    type == "likelihood" ~ "Likelihood L(p)",
    type == "log_likelihood" ~ "Log-Likelihood ℓ(p)"
  )) |> 
  ggplot(aes(x = p, y = value)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = mle_p, color = "red", linetype = "dashed") +
  facet_wrap(~type, scales = "free_y") +
  labs(title = "Likelihood vs. Log-Likelihood", 
       subtitle = "Both maximize at the same point",
       x = "Parameter (p)", y = "Value")
```

## The General MLE Recipe

::: callout-tip
## Finding the MLE

1.  **Write down the likelihood:** $L(\theta) = \prod f(x_i; \theta)$

2.  **Take the logarithm:** $\ell(\theta) = \sum \ln f(x_i; \theta)$

3.  **Differentiate:** Find $\ell'(\theta) = \frac{d\ell}{d\theta}$

4.  **Solve:** Set $\ell'(\theta) = 0$ and solve for $\hat{\theta}$

5.  **Verify:** Check that $\ell''(\hat{\theta}) < 0$ (maximum, not minimum)
:::

# MLE Examples {background-color="#e74c3c"}

## Example 1: Bernoulli Distribution

**Setup:** $X_1, \ldots, X_n \sim \text{Bernoulli}(p)$ (iid)

Let $x = \sum x_i$ be the number of successes.

. . .

**Step 1 - Likelihood:**
$$L(p) = \prod_{i=1}^n p^{x_i}(1-p)^{1-x_i} = p^x(1-p)^{n-x}$$

. . .

**Step 2 - Log-likelihood:**
$$\ell(p) = x\ln(p) + (n-x)\ln(1-p)$$

## Example 1: Bernoulli (continued)

**Step 3 - Differentiate:**
$$\ell'(p) = \frac{x}{p} - \frac{n-x}{1-p}$$

. . .

**Step 4 - Solve:**
$$\frac{x}{p} = \frac{n-x}{1-p}$$
$$x(1-p) = p(n-x)$$
$$x - xp = np - xp$$
$$x = np$$
$$\hat{p} = \frac{x}{n}$$

. . .

**The MLE is the sample proportion!** This matches our intuition.

## Simulation: Verifying the Bernoulli MLE

```{r}
#| code-line-numbers: "|1-3|5-12|14-18"
true_p <- 0.65
n <- 50
n_sims <- 5000

# Simulate MLE for many samples
bernoulli_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) rbinom(n, 1, true_p)),
    mle = map_dbl(data, mean)
  )

# Check properties
bernoulli_sim |> 
  summarize(
    mean_mle = mean(mle),
    var_mle = var(mle),
    theoretical_var = true_p * (1 - true_p) / n
  )
```

## Example 2: Exponential Distribution

**Setup:** $X_1, \ldots, X_n \sim \text{Exponential}(\lambda)$

. . .

**Likelihood:**
$$L(\lambda) = \prod_{i=1}^n \lambda e^{-\lambda x_i} = \lambda^n e^{-\lambda \sum x_i}$$

. . .

**Log-likelihood:**
$$\ell(\lambda) = n\ln(\lambda) - \lambda \sum x_i$$

. . .

**Solve:**
$$\ell'(\lambda) = \frac{n}{\lambda} - \sum x_i = 0 \Rightarrow \hat{\lambda} = \frac{n}{\sum x_i} = \frac{1}{\bar{X}}$$

. . .

**Same as MME!** This is common for exponential family distributions.

## Example 3: Normal Distribution

**Setup:** $X_1, \ldots, X_n \sim N(\mu, \sigma^2)$, both unknown

. . .

**Log-likelihood:**
$$\ell(\mu, \sigma^2) = -\frac{n}{2}\ln(2\pi) - \frac{n}{2}\ln(\sigma^2) - \frac{1}{2\sigma^2}\sum(x_i - \mu)^2$$

. . .

**Partial derivatives:**
$$\frac{\partial \ell}{\partial \mu} = \frac{1}{\sigma^2}\sum(x_i - \mu) = 0 \Rightarrow \hat{\mu} = \bar{X}$$

$$\frac{\partial \ell}{\partial \sigma^2} = -\frac{n}{2\sigma^2} + \frac{1}{2\sigma^4}\sum(x_i - \mu)^2 = 0 \Rightarrow \hat{\sigma}^2 = \frac{1}{n}\sum(x_i - \bar{X})^2$$

## Important Note on Normal MLE

$$\hat{\sigma}^2_{MLE} = \frac{1}{n}\sum(x_i - \bar{X})^2$$

. . .

::: callout-warning
## Bias Alert!
The MLE divides by $n$, not $n-1$:

$$E(\hat{\sigma}^2_{MLE}) = \frac{n-1}{n}\sigma^2 \neq \sigma^2$$

The MLE is **biased** for $\sigma^2$!
:::

. . .

However, for large $n$, this bias becomes negligible:
$$\frac{n-1}{n} \to 1 \text{ as } n \to \infty$$

# Numerical Optimization for MLEs {background-color="#9b59b6"}

## When Calculus Isn't Enough

Not all MLEs have closed-form solutions!

. . .

**Examples requiring numerical optimization:**

-   Gamma distribution (shape and rate parameters)
-   Weibull distribution
-   Beta distribution
-   Mixed models, GLMs, many real-world applications

. . .

::: callout-tip
## The General Approach
1. Write down the log-likelihood function $\ell(\theta)$
2. Use numerical optimization to find $\hat{\theta} = \arg\max \ell(\theta)$
3. In R: minimize the **negative** log-likelihood (since `optim` minimizes)
:::

## Example: Exponential MLE — Calculus vs. Numerical

Let's verify our calculus result numerically:

```{r}
# Generate exponential data
set.seed(42)
exp_data <- rexp(100, rate = 0.5)  # True λ = 0.5

# Calculus solution: λ̂ = 1/X̄
mle_calculus <- 1 / mean(exp_data)

# Numerical solution: minimize negative log-likelihood
neg_log_lik_exp <- function(lambda, data) {
  if (lambda <= 0) return(Inf)  # λ must be positive
  -sum(dexp(data, rate = lambda, log = TRUE))
}

# Use optimize() for single-parameter optimization
result <- optimize(neg_log_lik_exp, interval = c(0.01, 5), data = exp_data)
mle_numerical <- result$minimum
```

## Example: Exponential MLE — Calculus vs. Numerical (solution)

```{r}
cat("Calculus MLE:  λ̂ =", round(mle_calculus, 5), "\n")
cat("Numerical MLE: λ̂ =", round(mle_numerical, 5), "\n")
cat("Difference:", round(abs(mle_calculus - mle_numerical), 8))
```

They match! ✓

## Example: Gamma MLE — Numerical Only

The Gamma distribution has **no closed-form MLE** for the shape parameter $\alpha$.

**Log-likelihood for Gamma($\alpha$, $\beta$):**
$$\ell(\alpha, \beta) = n\alpha\ln(\beta) - n\ln\Gamma(\alpha) + (\alpha-1)\sum\ln(x_i) - \beta\sum x_i$$

. . .

Setting $\partial\ell/\partial\alpha = 0$ involves the **digamma function** $\psi(\alpha)$ — no closed-form solution!

## Gamma MLE: Numerical Optimization in R (1/2)

```{r}
# Generate gamma data (true α = 3, β = 2)
set.seed(123)
gamma_data <- rgamma(150, shape = 3, rate = 0.5)  # rate = 1/scale

# Define negative log-likelihood for Gamma
neg_log_lik_gamma <- function(params, data) {
  alpha <- params[1]
  beta <- params[2]   # rate parameter
  if (alpha <= 0 | beta <= 0) return(Inf)
  -sum(dgamma(data, shape = alpha, rate = beta, log = TRUE))
}
```

## Gamma MLE: Numerical Optimization in R (2/2)

```{r}
# Use optim() for multi-parameter optimization
# Start with method of moments estimates as initial values
start_alpha <- mean(gamma_data)^2 / var(gamma_data)
start_beta <- mean(gamma_data) / var(gamma_data)

result <- optim(
  par = c(start_alpha, start_beta),
  fn = neg_log_lik_gamma,
  data = gamma_data,
  method = "L-BFGS-B",           # Allows bounds
  lower = c(0.001, 0.001)        # Parameters must be positive
)

cat("MLE of α:", round(result$par[1], 3), "(true: 3)\n")
cat("MLE of β:", round(result$par[2], 3), "(true: 0.5)")
```

## Visualizing the Gamma Log-Likelihood Surface

```{r}
#| echo: false
#| fig-height: 5
# Create grid for contour plot
alpha_seq <- seq(2, 5, length.out = 50)
beta_seq <- seq(0.3, 0.8, length.out = 50)

log_lik_surface <- expand_grid(alpha = alpha_seq, beta = beta_seq) |> 
  mutate(
    log_lik = map2_dbl(alpha, beta, \(a, b) {
      -neg_log_lik_gamma(c(a, b), gamma_data)
    })
  )

ggplot(log_lik_surface, aes(x = alpha, y = beta, z = log_lik)) +
  geom_contour_filled(bins = 20) +
  geom_point(aes(x = result$par[1], y = result$par[2]), 
             color = "red", size = 4, shape = 4, stroke = 2) +
  geom_point(aes(x = 3, y = 0.5), 
             color = "white", size = 3, shape = 1, stroke = 2) +
  labs(title = "Log-Likelihood Surface for Gamma Distribution",
       subtitle = "Red X = MLE, White circle = true parameters",
       x = "Shape (α)", y = "Rate (β)", fill = "Log-Lik") +
  theme(legend.position = "right")
```

## Using MASS::fitdistr() for Common Distributions

```{r}
#| message: false
library(MASS)
select <- dplyr::select

# fitdistr uses numerical optimization internally
gamma_fit <- fitdistr(gamma_data, "gamma")

# Compare our manual optimization with fitdistr
cat("Our optim() result:\n")
cat("  α =", round(result$par[1], 4), ", β =", round(result$par[2], 4), "\n\n")

cat("MASS::fitdistr() result:\n")
print(gamma_fit)
```

Same answer — `fitdistr()` is just a convenient wrapper!

## Numerical Optimization: Practical Tips

::: callout-tip
## Best Practices

1. **Good starting values matter!** Use MME estimates as initial guesses

2. **Check convergence:** Look at `result$convergence` (0 = success)

3. **Use appropriate bounds:** For parameters that must be positive, use `method = "L-BFGS-B"` with `lower` bounds

4. **Multiple starting points:** Try different initial values to avoid local maxima

5. **Verify your result:** Compare with `fitdistr()` or plot the log-likelihood
:::

. . .

```{r}
# Always check convergence!
cat("Convergence code:", result$convergence, 
    ifelse(result$convergence == 0, "(success)", "(failed)"))
```

## Example 4: Uniform Distribution (Boundary Solution)

**Setup:** $X_1, \ldots, X_n \sim \text{Uniform}[0, \theta]$

**Likelihood:**
$$L(\theta) = \prod_{i=1}^n \frac{1}{\theta} \cdot I(0 \le x_i \le \theta) = \frac{1}{\theta^n} \cdot I(\max(x_i) \le \theta)$$

. . .

**Key insight:** $L(\theta) = 0$ if $\theta < \max(x_i)$

. . .

For $\theta \ge \max(x_i)$: $L(\theta) = 1/\theta^n$ is **decreasing** in $\theta$

. . .

**MLE:** $\hat{\theta} = \max(X_1, \ldots, X_n)$

**Derivative method fails!** This is a boundary maximum.

## MLE vs MME for Uniform (1/2)

```{r}
theta <- 10
n <- 20
n_sims <- 5000

uniform_comparison <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) runif(n, 0, theta)),
    mle = map_dbl(data, max),
    mme = map_dbl(data, \(d) 2 * mean(d))
  )
```

## MLE vs MME for Uniform (2/2)


```{r}
uniform_comparison |> 
  summarize(
    `MLE Bias` = mean(mle) - theta,
    `MME Bias` = mean(mme) - theta,
    `MLE Variance` = var(mle), `MME Variance` = var(mme),
    `MLE MSE` = mean((mle - theta)^2),
    `MME MSE` = mean((mme - theta)^2)
  ) |> 
  pivot_longer(everything()) |> 
  mutate(value = round(value, 4))
```

**The MLE has smaller MSE** despite being biased!

## Visualizing MLE vs MME: Code

```{r}
#| label: mle-mme-code
#| eval: false
uniform_comparison |> 
  select(mle, mme) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = toupper(Estimator)) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = theta, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "MLE vs MME for Uniform[0, θ]",
       subtitle = str_glue("True θ = {theta}; MLE is biased but has lower variance"),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("steelblue", "darkgreen")) +
  theme(legend.position = "top")
```

## Visualizing MLE vs MME: Result

```{r}
#| label: mle-mme-plot
#| echo: false
#| fig-height: 6
uniform_comparison |> 
  select(mle, mme) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = toupper(Estimator)) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = theta, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "MLE vs MME for Uniform[0, θ]",
       subtitle = str_glue("True θ = {theta}; MLE is biased but has lower variance"),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("steelblue", "darkgreen")) +
  theme(legend.position = "top")
```

# Properties of MLEs {background-color="#2c3e50"}

## Property 1: Invariance

::: callout-important
## Invariance Property

If $\hat{\theta}$ is the MLE of $\theta$, then for any function $g$:

$$\widehat{g(\theta)} = g(\hat{\theta})$$

The MLE of $g(\theta)$ is simply $g$ applied to the MLE of $\theta$.
:::

. . .

**Example:** If $\hat{p} = 0.78$ is the MLE for a proportion, then:

-   MLE of odds $= p/(1-p)$: $\hat{\text{odds}} = 0.78/0.22 = 3.55$
-   MLE of $\log(p)$: $\widehat{\log p} = \log(0.78) = -0.248$

## Property 2: Consistency — The Big Picture

::: callout-important
## What Does Consistency Mean?

An estimator $\hat{\theta}$ is **consistent** for $\theta$ if, as we collect more and more data, our estimate gets closer and closer to the true value.

**In plain English:** With enough data, we can estimate $\theta$ as accurately as we want.
:::

. . .

**Think of it like this:**

-   With $n = 10$ observations, your estimate might be off by a lot
-   With $n = 100$ observations, your estimate is probably closer
-   With $n = 1000$ observations, your estimate is very close
-   As $n \to \infty$, your estimate essentially equals the truth

## Consistency: A Medical Analogy

Imagine estimating the true cure rate $p$ of a new drug:

| Sample Size | Your Estimate | How Close? |
|-------------|---------------|------------|
| $n = 10$ | $\hat{p} = 0.60$ | Could be quite far off |
| $n = 100$ | $\hat{p} = 0.72$ | Getting closer |
| $n = 1,000$ | $\hat{p} = 0.748$ | Very close |
| $n = 10,000$ | $\hat{p} = 0.7502$ | Almost exactly right |

. . .

If the true cure rate is $p = 0.75$, a **consistent** estimator will eventually "find" this value.

. . .

::: callout-tip
## Key Insight
Consistency is about what happens as sample size grows—it's a **long-run guarantee** about the behavior of an estimator.
:::

## Consistency: Formal Statement

::: callout-important
## Theorem: MLEs are Consistent

Under general regularity conditions, the MLE $\hat{\theta}$ satisfies:

$$\hat{\theta} \xrightarrow{P} \theta \quad \text{as } n \to \infty$$

**Translation:** For any tolerance $\epsilon > 0$, no matter how small:

$$P(|\hat{\theta} - \theta| < \epsilon) \to 1 \quad \text{as } n \to \infty$$

The probability that our estimate is within $\epsilon$ of the truth approaches 100%.
:::

. . .

**What this does NOT mean:**

-   It does NOT mean $\hat{\theta}$ is unbiased
-   It does NOT mean $\hat{\theta}$ is exactly right for any finite $n$
-   It's a statement about the **limiting behavior** as $n$ grows

## Consistency via MSE → 0

::: callout-important
## Alternative Characterization

An estimator $\hat{\theta}$ is consistent if its **Mean Squared Error** goes to zero:

$$\text{MSE}(\hat{\theta}) = E[(\hat{\theta} - \theta)^2] \to 0 \quad \text{as } n \to \infty$$
:::

. . .

**Why does this work?** Recall: $\text{MSE} = \text{Variance} + \text{Bias}^2$

For consistency via MSE → 0, we need both:

-   $\text{Var}(\hat{\theta}) \to 0$ (estimates become more precise)
-   $\text{Bias}(\hat{\theta}) \to 0$ (estimates become centered on truth)

. . .

::: callout-tip
This is often the easiest way to **prove** consistency — just show that MSE → 0!
:::

## Visualizing Consistency: Setup

Let's see consistency in action with the Exponential MLE.

**True parameter:** $\lambda = 0.5$

**MLE:** $\hat{\lambda} = 1/\bar{X}$

We'll simulate 1000 MLEs at each sample size and watch the distribution "shrink" around the truth.

```{r}
#| label: consistency-setup
true_lambda <- 0.5
sample_sizes <- c(10, 30, 100, 300, 1000)
set.seed(100)

consistency_sim <- map(sample_sizes, function(sample_n) {
  tibble(
    n = sample_n,
    sim = 1:1000,
    mle = map_dbl(sim, \(s) 1 / mean(rexp(sample_n, rate = true_lambda)))
  )
}) %>%
  list_rbind()
```

## Visualizing Consistency: Code

```{r}
#| label: consistency-code
#| eval: false
consistency_sim |> 
  ggplot(aes(x = mle, fill = factor(n))) +
  geom_histogram(bins = 40, alpha = 0.7) +
  geom_vline(xintercept = true_lambda, color = "red", linewidth = 1.2) +
  facet_wrap(~n, scales = "free_y", labeller = label_both) +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Consistency of Exponential MLE",
       subtitle = "Distribution concentrates around true λ = 0.5 as n increases",
       x = "MLE of λ", y = "Frequency") +
  theme(legend.position = "none")
```

## Visualizing Consistency: Result

```{r}
#| label: consistency-plot
#| echo: false
#| fig-height: 6
consistency_sim |> 
  ggplot(aes(x = mle, fill = factor(n))) +
  geom_histogram(bins = 40, alpha = 0.7) +
  geom_vline(xintercept = true_lambda, color = "red", linewidth = 1.2) +
  facet_wrap(~n, scales = "free_y", labeller = label_both) +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Consistency of Exponential MLE",
       subtitle = "Distribution concentrates around true λ = 0.5 as n increases",
       x = "MLE of λ", y = "Frequency") +
  theme(legend.position = "none")
```

## Consistency: MSE Decomposition (1/2)


```{r}
# Show bias, variance, and MSE at each sample size
consistency_sim |> 
  group_by(n) |> 
  summarize(
    Bias = mean(mle) - true_lambda,
    Variance = var(mle),
    `Bias²` = (mean(mle) - true_lambda)^2,
    MSE = mean((mle - true_lambda)^2)
  ) |> 
  mutate(across(where(is.numeric), \(x) round(x, 5)))
```

## Consistency: MSE Decomposition (2/2)

**Key Observations:**

-   **Bias** is small and shrinks slightly (MLE is approximately unbiased for large $n$)
-   **Variance** shrinks dramatically as $n$ increases  
-   **MSE → 0** as $n$ increases — this proves consistency!


## Visualizing MSE → 0: Code

```{r}
#| label: mse-plot-code
#| eval: false
# Calculate MSE at each sample size
mse_summary <- consistency_sim |> 
  group_by(n) |> 
  summarize(
    MSE = mean((mle - true_lambda)^2),
    Variance = var(mle),
    Bias_sq = (mean(mle) - true_lambda)^2
  )

mse_summary |> 
  pivot_longer(-n, names_to = "Component", values_to = "Value") |> 
  mutate(Component = factor(Component, levels = c("MSE", "Variance", "Bias_sq"))) |> 
  ggplot(aes(x = n, y = Value, color = Component)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Consistency: MSE → 0 as Sample Size Increases",
       subtitle = "Log-log scale shows MSE shrinking toward zero",
       x = "Sample Size (n)", y = "Value (log scale)") +
  theme(legend.position = "right")
```

## Visualizing MSE → 0: Result

```{r}
#| label: mse-plot-result
#| echo: false
#| fig-height: 5.5
# Calculate MSE at each sample size
mse_summary <- consistency_sim |> 
  group_by(n) |> 
  summarize(
    MSE = mean((mle - true_lambda)^2),
    Variance = var(mle),
    Bias_sq = (mean(mle) - true_lambda)^2
  )

mse_summary |> 
  pivot_longer(-n, names_to = "Component", values_to = "Value") |> 
  mutate(Component = factor(Component, levels = c("MSE", "Variance", "Bias_sq"),
                            labels = c("MSE", "Variance", "Bias²"))) |> 
  ggplot(aes(x = n, y = Value, color = Component)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Consistency: MSE → 0 as Sample Size Increases",
       subtitle = "Log-log scale; MSE ≈ Variance since bias is small",
       x = "Sample Size (n)", y = "Value (log scale)") +
  scale_color_manual(values = c("MSE" = "red", "Variance" = "steelblue", "Bias²" = "darkgreen")) +
  theme(legend.position = "right")
```

MSE is dominated by variance (bias is negligible), and both → 0 as n → ∞

## Property 3: Asymptotic Normality & Efficiency

::: callout-important
## Theorem (Devore 7.4)
Under regularity conditions, for large $n$:

1.  $\hat{\theta}$ is **approximately unbiased**: $E(\hat{\theta}) \approx \theta$

2.  $\hat{\theta}$ is **approximately normal**: $\hat{\theta} \dot{\sim} N\left(\theta, \frac{1}{nI(\theta)}\right)$

3.  $\hat{\theta}$ is **asymptotically efficient**: Achieves the Cramér-Rao lower bound
:::

. . .

**In practice:** For large samples, the MLE is approximately the MVUE!

## Comparing MLE and MME

When do they agree? When do they differ?

| Distribution | MLE | MME | Same? |
|--------------|-----|-----|-------|
| Bernoulli($p$) | $\bar{X}$ | $\bar{X}$ | ✓ |
| Exponential($\lambda$) | $1/\bar{X}$ | $1/\bar{X}$ | ✓ |
| Normal($\mu$, $\sigma^2$) | $\bar{X}$, $\frac{1}{n}\sum(X_i-\bar{X})^2$ | $\bar{X}$, $\frac{1}{n}\sum(X_i-\bar{X})^2$ | ✓ |
| Uniform[0, $\theta$] | $\max(X_i)$ | $2\bar{X}$ | ✗ |
| Weibull | Numerical | Different | ✗ |

## Comparing Estimator Properties

| Property | MME | MLE |
|----------|-----|-----|
| **Computation** | Usually simple algebra | May require optimization |
| **Bias** | Sometimes unbiased | Often biased for small $n$ |
| **Efficiency** | Rarely efficient | Asymptotically efficient |
| **Consistency** | Usually consistent | Always consistent |
| **Invariance** | No | Yes |

. . .

::: callout-tip
## General Guidance
-   Use MME for quick, rough estimates or simple distributions
-   Use MLE for final analysis and complex distributions
-   MLE is generally preferred when computationally feasible
:::

## Computing MLEs in R

**Using built-in functions:**

```{r}
# For distributions with known MLEs
normal_data <- rnorm(100, mean = 50, sd = 10)
cat("Normal MLE for μ:", mean(normal_data), "\n")
cat("Normal MLE for σ:", sd(normal_data) * sqrt((100-1)/100))
```

. . .

**Using optimization for complex cases:**

```{r}
# General approach: maximize log-likelihood
neg_log_lik <- function(lambda, data) {
  -sum(dexp(data, rate = lambda, log = TRUE))
}

exp_data <- rexp(50, rate = 0.5)
result <- optimize(neg_log_lik, interval = c(0.01, 5), data = exp_data)
cat("\nExponential MLE for λ:", round(result$minimum, 4))
```

## Using fitdistr for Common Distributions

```{r}
#| message: false
library(MASS)

# Fit various distributions
gamma_data <- rgamma(100, shape = 2, scale = 3)

# Method of moments (manual)
xbar <- mean(gamma_data)
s2 <- var(gamma_data) * (99/100)
alpha_mme <- xbar^2 / s2
beta_mme <- s2 / xbar

# MLE using fitdistr
gamma_mle <- fitdistr(gamma_data, "gamma")

cat("MME: α =", round(alpha_mme, 3), ", β =", round(beta_mme, 3), "\n")
cat("MLE: α =", round(gamma_mle$estimate["shape"], 3), 
    ", β =", round(1/gamma_mle$estimate["rate"], 3))
```

## Medical Application: Survival Analysis

Cancer survival times are often modeled with exponential or Weibull distributions.

```{r}
# Survival times (months) for 40 patients
survival_times <- c(8.2, 15.3, 12.1, 6.8, 22.5, 9.4, 18.7, 11.3, 7.5, 25.1,
                    14.2, 10.8, 5.3, 19.6, 13.4, 8.9, 16.2, 11.7, 6.1, 21.3,
                    12.8, 9.1, 17.5, 7.8, 24.2, 10.3, 14.9, 8.4, 20.1, 11.9,
                    6.5, 18.3, 13.1, 9.7, 15.8, 7.2, 22.8, 12.4, 8.1, 19.2)

# Fit exponential distribution
lambda_mle <- 1 / mean(survival_times)

# Estimated survival probability at 12 months
surv_12 <- exp(-lambda_mle * 12)

cat("MLE of λ:", round(lambda_mle, 4), "per month\n")
cat("Estimated median survival:", round(log(2)/lambda_mle, 1), "months\n")
cat("Estimated P(survive > 12 months):", round(surv_12, 3))
```

## Your Turn: MLE Practice

**Exercise:** A diagnostic test is administered to 80 patients. 64 test positive. Assuming the number positive follows a Binomial distribution, find:

1.  The MLE of the sensitivity $p$
2.  The MLE of the odds of a positive result, $p/(1-p)$
3.  The MLE of $P(\text{at least 60 positive in next 80 patients})$

. . .

**Solutions:**

1.  $\hat{p} = 64/80 = 0.80$
2.  By invariance: $\widehat{p/(1-p)} = 0.80/0.20 = 4.0$
3.  By invariance: $\widehat{P(X \ge 60)} = P(X \ge 60 | p = 0.80)$

```{r}
1 - pbinom(59, size = 80, prob = 0.80)  # P(X >= 60)
```

# Summary and Looking Ahead {background-color="#2c3e50"}

## Lesson 4 Summary

**Two Methods for Constructing Estimators:**

1.  **Method of Moments:**
    -   Equate population and sample moments
    -   Simple but not always efficient
    
2.  **Maximum Likelihood:**
    -   Maximize $L(\theta)$ or $\ell(\theta)$
    -   Properties: invariance, consistency, asymptotic efficiency

. . .

**Key Insight:** MLEs are asymptotically MVUE—they achieve the Cramér-Rao bound for large samples!

## MLE and MME Summary Table

| Distribution | Parameter | MLE | MME |
|--------------|-----------|-----|-----|
| Bernoulli | $p$ | $\bar{X}$ | $\bar{X}$ |
| Exponential | $\lambda$ | $1/\bar{X}$ | $1/\bar{X}$ |
| Normal | $\mu$, $\sigma^2$ | $\bar{X}$, $\frac{1}{n}\sum(X_i-\bar{X})^2$ | Same |
| Uniform[0,$\theta$] | $\theta$ | $\max(X_i)$ | $2\bar{X}$ |
| Poisson | $\mu$ | $\bar{X}$ | $\bar{X}$ |

## Lesson 4 Practice Problems

1.  **(Devore 7.25)** Derive the MLE for the parameter $\theta$ of the pdf $f(x; \theta) = (\theta+1)x^\theta$, $0 \le x \le 1$.

2.  **(Devore 7.22)** Use method of moments to estimate the parameters of a Beta($\alpha$, $\beta$) distribution.

3.  **(Devore 7.31)** For normal measurement errors with known mean 0, derive the MLE of the variance $\sigma^2$.


## Looking Ahead

**Upcoming Topics:**

-   Sufficient statistics (Section 7.3)
-   Confidence intervals (Chapter 8)
-   Hypothesis testing (Chapter 9)

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapter 7.2
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 6.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/04_MME_MLE/04_MME_MLE.html lessons/04_MME_MLE/04_MME_MLE.pdf
```
