---
title: "BSTA 551: Statistical Inference"
subtitle: "Review: Point Estimation (Lessons 1-5)"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "Week 3"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Review - Point Estimation
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
set.seed(42)
theme_set(theme_minimal(base_size = 18))
```

# Review: The Big Picture of Point Estimation {background-color="#2c3e50"}

## Where We've Been: The Journey So Far

```{r}
#| echo: false
#| fig-height: 4
library(DiagrammeR)

grViz("
digraph journey {
  graph [rankdir = LR, bgcolor = transparent]
  node [shape = box, style = filled, fontsize = 14]
  
  L1 [label = 'Lesson 1\\nStatistics &\\nSampling', fillcolor = lightblue]
  L2 [label = 'Lesson 2\\nBias, Variance,\\nMSE', fillcolor = lightyellow]
  L3 [label = 'Lesson 3\\nMVUE &\\nCRLB', fillcolor = palegreen1]
  L4 [label = 'Lesson 4\\nMME &\\nMLE', fillcolor = lightsalmon]
  L5 [label = 'Lesson 5\\nSufficient\\nStatistics', fillcolor = plum]
  
  L1 -> L2 -> L3 -> L4 -> L5
}
")
```

**Today:** Tie it all together and review key conceps.

## The Central Problem of Statistical Inference

We observe data $X_1, X_2, \ldots, X_n$ from some distribution with unknown parameter $\theta$. Usually they are identically and independently distributed (i.i.d.), so they make a "random sample."

**The Question:** How do we learn about $\theta$ from the data?

. . .

```{r}
#| echo: false
#| fig-height: 4
tibble(x = c(1, 2), y = c(1, 1), size = c(4, 2)) |> 
  ggplot(aes(x = x, y = y)) +
  geom_point(aes(size = size), color = "steelblue", alpha = 0.5) +
  annotate("text", x = 1, y = 1, label = "Population\n(Unknown θ)", size = 7) +
  annotate("text", x = 2, y = 1, label = "Sample\n(Compute estimate)", size = 7) +
  annotate("segment", x = 1.3, xend = 1.7, y = 1, yend = 1,
           arrow = arrow(length = unit(0.3, "cm")), linewidth = 1.5) +
  annotate("text", x = 1.5, y = 1.12, label = "Sample", size = 5) +
  annotate("segment", x = 1.7, xend = 1.3, y = 0.88, yend = 0.88,
           arrow = arrow(length = unit(0.3, "cm")), linewidth = 1.5, color = "darkred") +
  annotate("text", x = 1.5, y = 0.76, label = "Inference", size = 5, color = "darkred") +
  scale_size_identity() + xlim(0.5, 2.5) + ylim(0.5, 1.3) + theme_void()
```

## The Two Key Questions in Point Estimation

**Question 1:** How do we **construct** estimators?

-   Method of Moments (MME)
-   Maximum Likelihood (MLE)

. . .

**Question 2:** How do we **evaluate** estimators?

-   Bias (accuracy)
-   Variance (precision)
-   Mean Squared Error (overall quality)
-   Efficiency (comparison to optimal)

. . .

**Question 3:** How do we find the **best** estimator?

-   MVUE: Minimum Variance Unbiased Estimator
-   Sufficient statistics: Data reduction without information loss

# Part 1: Properties of Estimators {background-color="#27ae60"}

## The Fundamental Tradeoff: Bias vs. Variance

::: callout-important
## Key Definitions

-   **Bias:** $\text{Bias}(\hat{\theta}) = E(\hat{\theta}) - \theta$ (systematic error)
-   **Variance:** $\text{Var}(\hat{\theta}) = E[(\hat{\theta} - E(\hat{\theta}))^2]$ (random error)
-   **MSE:** $\text{MSE}(\hat{\theta}) = E[(\hat{\theta} - \theta)^2] = \text{Var}(\hat{\theta}) + \text{Bias}^2$
:::

. . .

**The key insight:** MSE = Variance + Bias²

This means a **biased** estimator can be better than an unbiased one if its variance is low enough!

## Visual Intuition: Bias and Variance

```{r}
#| echo: false
#| fig-height: 5.5
set.seed(42)
estimator_comparison <- bind_rows(
  tibble(type = "A: Low Bias, Low Variance (IDEAL)", estimate = rnorm(500, 10, 0.5)),
  tibble(type = "B: Low Bias, High Variance", estimate = rnorm(500, 10, 2)),
  tibble(type = "C: High Bias, Low Variance", estimate = rnorm(500, 12, 0.5)),
  tibble(type = "D: High Bias, High Variance", estimate = rnorm(500, 12, 2))
)

estimator_comparison |> 
  ggplot(aes(x = estimate, fill = type)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  geom_vline(xintercept = 10, color = "red", linetype = "dashed", linewidth = 1.2) +
  facet_wrap(~type) +
  labs(x = "Estimate", y = "Frequency", 
       title = "The Four Scenarios of Bias-Variance",
       subtitle = "Red dashed line = true parameter θ = 10") +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")
```

## Medical Example: Estimating Treatment Effect

**Scenario:** Estimating the true cure rate $p$ of a new cancer treatment.

```{r}
#| code-fold: true
# Compare standard estimator vs. "add-two" estimator
true_p <- 0.2  # True cure rate (rare disease)
n <- 25
n_sims <- 5000

prop_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    x = rbinom(n_sims, size = n, prob = true_p),
    p_standard = x / n,                    # Unbiased
    p_addtwo = (x + 2) / (n + 4)           # Biased toward 0.5
  )

# Compare properties
prop_sim |> 
  summarize(
    `Standard: Mean` = mean(p_standard),
    `Standard: Bias` = mean(p_standard) - true_p,
    `Standard: Variance` = var(p_standard),
    `Standard: MSE` = mean((p_standard - true_p)^2),
    `Add-Two: Mean` = mean(p_addtwo),
    `Add-Two: Bias` = mean(p_addtwo) - true_p,
    `Add-Two: Variance` = var(p_addtwo),
    `Add-Two: MSE` = mean((p_addtwo - true_p)^2)
  ) |> 
  pivot_longer(everything()) |> 
  separate(name, into = c("Estimator", "Metric"), sep = ": ") |> 
  pivot_wider(names_from = Estimator, values_from = value) |> 
  mutate(across(where(is.numeric), ~round(., 5)))
```

. . .

**Key insight:** The biased add-two estimator has **lower MSE** for extreme $p$ values!

## What Properties Do We Want in an Estimator?

| Property | Definition | Why It Matters |
|----------|------------|----------------|
| **Unbiased** | $E(\hat{\theta}) = \theta$ | No systematic error |
| **Low Variance** | $\text{Var}(\hat{\theta})$ small | Precise estimates |
| **Consistent** | $\hat{\theta} \xrightarrow{P} \theta$ | Works with enough data |
| **Efficient** | Achieves CRLB | Best among unbiased |

. . .

::: callout-tip
## The Hierarchy of Goals
1. **Consistency** is the bare minimum (sniff test)
2. **Unbiasedness** is nice but not essential
3. **Efficiency** (low variance) is what really matters for precision
:::

# Part 2: Constructing Estimators {background-color="#3498db"}

## Two Systematic Methods: MME and MLE

**Method of Moments (MME):**

-   Match population moments to sample moments
-   Simple algebra, often intuitive
-   Not always optimal

. . .

**Maximum Likelihood (MLE):**

-   Find $\theta$ that maximizes $P(\text{data} | \theta)$
-   More complex, but has optimal properties
-   Asymptotically efficient!

## The MME Recipe

::: callout-important
## Method of Moments: The Algorithm

1. Express the first $k$ population moments as functions of $\theta$: $E(X), E(X^2), \ldots$
2. Set each equal to its sample counterpart: $\bar{X}, \frac{1}{n}\sum X_i^2, \ldots$
3. Solve for $\theta$
:::

. . .

**Example: Exponential($\lambda$)**

-   Population moment: $E(X) = 1/\lambda$
-   Set equal: $1/\lambda = \bar{X}$
-   Solve: $\hat{\lambda}_{MME} = 1/\bar{X}$

## The MLE Recipe

::: callout-important
## Maximum Likelihood: The Algorithm

1. Write the likelihood: $L(\theta) = \prod_{i=1}^n f(x_i; \theta)$
2. Take the log: $\ell(\theta) = \sum_{i=1}^n \ln f(x_i; \theta)$
3. Differentiate and set to zero: $\ell'(\theta) = 0$
4. Solve for $\hat{\theta}$
:::

. . .

**Intuition:** The MLE asks "What parameter value makes our observed data most probable?"

## Visual Intuition: What Does the MLE Do?

```{r}
#| echo: false
#| fig-height: 5
# Likelihood for binomial: 7 successes in 10 trials
tibble(p = seq(0.01, 0.99, by = 0.01)) |> 
  mutate(likelihood = dbinom(7, 10, p)) |> 
  ggplot(aes(x = p, y = likelihood)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_vline(xintercept = 0.7, color = "red", linetype = "dashed", linewidth = 1.2) +
  geom_point(aes(x = 0.7, y = dbinom(7, 10, 0.7)), color = "red", size = 5) +
  labs(title = "Likelihood Function: 7 Successes in 10 Trials",
       subtitle = "MLE finds the peak at p = 0.7",
       x = "Parameter p", y = "L(p) = P(X = 7 | p)") +
  annotate("text", x = 0.8, y = 0.28, label = "MLE = 7/10 = 0.7", color = "red", size = 6)
```

The MLE finds where the likelihood is **highest** — the most "likely" parameter value.

## When Do MME and MLE Agree?

| Distribution | MLE | MME | Same? |
|--------------|-----|-----|-------|
| Bernoulli($p$) | $\bar{X}$ | $\bar{X}$ | ✓ |
| Normal($\mu$, known $\sigma^2$) | $\bar{X}$ | $\bar{X}$ | ✓ |
| Exponential($\lambda$) | $1/\bar{X}$ | $1/\bar{X}$ | ✓ |
| Poisson($\mu$) | $\bar{X}$ | $\bar{X}$ | ✓ |
| Uniform[0, $\theta$] | $\max(X_i)$ | $2\bar{X}$ | **✗** |

. . .

::: callout-warning
## Key Insight
When they differ, the MLE is usually preferred (especially for large $n$) because it's asymptotically efficient.
:::

## MLE Properties: Why It's Often the Best Choice

::: callout-important
## Three Key Properties of MLEs (Devore 7.2)

1. **Invariance:** If $\hat{\theta}$ is the MLE of $\theta$, then $g(\hat{\theta})$ is the MLE of $g(\theta)$

2. **Consistency:** $\hat{\theta}_{MLE} \xrightarrow{P} \theta$ as $n \to \infty$

3. **Asymptotic Efficiency:** For large $n$, $\hat{\theta}_{MLE} \dot\sim N\left(\theta, \frac{1}{nI(\theta)}\right)$
:::

. . .

**In plain English:**

-   You can transform MLEs freely (invariance)
-   MLEs "work" given enough data (consistency)
-   MLEs are approximately the **best** for large samples (efficiency)

## Consistency: What Does It Really Mean?

::: callout-important
## Definition: Consistency (Devore 7.1, Chihara 6.3.4)

An estimator $\hat{\theta}_n$ is **consistent** if:
$$P(|\hat{\theta}_n - \theta| < \epsilon) \to 1 \quad \text{as } n \to \infty$$

for any $\epsilon > 0$, no matter how small.
:::

. . .

**Intuition:** With enough data, your estimate will be arbitrarily close to the truth.

**Alternative criterion:** $\text{MSE}(\hat{\theta}_n) \to 0$ as $n \to \infty$

. . .

**This requires:**

-   Variance $\to 0$ (estimates get more precise)
-   Bias $\to 0$ (estimates get more accurate)

## Visualizing Consistency

```{r}
#| echo: false
#| fig-height: 5.5
set.seed(42)
true_lambda <- 0.5
sample_sizes <- c(10, 50, 200, 1000)

consistency_sim <- map(sample_sizes, function(n) {
  tibble(
    n = n,
    mle = map_dbl(1:1000, \(s) 1 / mean(rexp(n, rate = true_lambda)))
  )
}) |> list_rbind()

consistency_sim |> 
  ggplot(aes(x = mle)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = true_lambda, color = "red", linewidth = 1.2) +
  facet_wrap(~paste("n =", n), scales = "free_y") +
  coord_cartesian(xlim = c(0, 1.5)) +
  labs(title = "Consistency: MLE Concentrates Around True Value",
       subtitle = "Exponential MLE: λ̂ = 1/X̄, True λ = 0.5",
       x = "MLE Estimate", y = "Frequency")
```

As $n$ increases, the distribution "shrinks" around the truth!

# Part 3: Finding the Best Estimator {background-color="#8e44ad"}

## The MVUE: Minimum Variance Unbiased Estimator

::: callout-important
## Definition (Devore 7.1)

The **MVUE** of $\theta$ is the estimator that:

1. Is unbiased: $E(\hat{\theta}) = \theta$
2. Has smaller variance than **any** other unbiased estimator
:::

. . .

**Why not just minimize MSE?**

-   MSE depends on the true $\theta$ (which we don't know!)
-   A biased estimator might be better at some $\theta$ values, worse at others
-   MVUE provides a clean, universal optimality criterion

## The Cramér-Rao Lower Bound (CRLB)

::: callout-important
## Theorem: CRLB (Devore 7.4)

For any unbiased estimator $T$ of $\theta$:
$$\text{Var}(T) \geq \frac{1}{nI(\theta)}$$

where $I(\theta)$ is the **Fisher Information**:
$$I(\theta) = E\left[-\frac{\partial^2}{\partial \theta^2} \ln f(X; \theta)\right]$$
:::

. . .

**Interpretation:**

-   The CRLB is a **floor** — no unbiased estimator can do better
-   If an estimator achieves the CRLB, it's the **MVUE**
-   Fisher Information measures the "curvature" of the log-likelihood

## Fisher Information: The Intuition

**Key Insight:** Fisher Information measures how much the log-likelihood changes as $\theta$ changes.

. . .

**High Fisher Information:**

-   Sharp, peaked likelihood → easy to pinpoint $\theta$
-   Small variance possible

**Low Fisher Information:**

-   Flat likelihood → hard to distinguish $\theta$ values
-   Large variance unavoidable

## Efficiency: How Close to Optimal?

::: callout-important
## Definition: Efficiency

An unbiased estimator $T$ is **efficient** if:
$$\text{Var}(T) = \frac{1}{nI(\theta)} = \text{CRLB}$$

More generally, the **efficiency** of $T$ is:
$$\text{Efficiency}(T) = \frac{\text{CRLB}}{\text{Var}(T)} \leq 1$$
:::

. . .

**Example:** Sample median for normal data

-   Variance ≈ $1.57 \cdot \sigma^2/n$
-   CRLB = $\sigma^2/n$
-   Efficiency = $1/1.57 \approx 0.64$ (64% efficient)

## Key Results: Efficient Estimators

| Distribution | Parameter | MVUE | CRLB | Efficient? |
|--------------|-----------|------|------|------------|
| Bernoulli($p$) | $p$ | $\bar{X}$ | $\frac{p(1-p)}{n}$ | ✓ |
| Normal($\mu$, known $\sigma^2$) | $\mu$ | $\bar{X}$ | $\frac{\sigma^2}{n}$ | ✓ |
| Poisson($\mu$) | $\mu$ | $\bar{X}$ | $\frac{\mu}{n}$ | ✓ |
| Exponential($\lambda$) | $\lambda$ | See note | $\frac{\lambda^2}{n}$ | Complex |

. . .

::: callout-note
## Important
For Uniform[0, $\theta$], the CRLB doesn't apply (support depends on $\theta$), yet efficient estimation is still possible!
:::

# Part 4: Sufficient Statistics — The Key to Everything {background-color="#e74c3c"}

## Why Sufficiency Matters: The Central Concept

**Students often ask:** "Why do we need sufficient statistics?"

. . .

**Answer:** Sufficient statistics answer a fundamental question:

> **What is the minimum amount of information we need to keep from our data to estimate $\theta$?**

. . .

If a statistic $T$ is sufficient, then:

-   We can throw away the raw data and just keep $T$
-   We lose **no information** about $\theta$
-   All optimal estimators will be functions of $T$

## Intuition: What Does "Sufficient" Really Mean?

**Scenario:** You observe $n$ coin flips and want to estimate $p$ (probability of heads).

. . .

**Raw data:** H, T, H, H, T, H, T, T, H, H (10 flips)

**Summary:** 6 heads out of 10

. . .

**Question:** Does knowing the exact sequence (HTHHTHTTHH) tell you more about $p$ than just knowing "6 out of 10"?

. . .

**Answer:** **No!** The total number of heads captures all the information about $p$.

This is why $T = \sum X_i$ is **sufficient** for $p$ in the Bernoulli model.

## Formal Definition of Sufficiency

::: callout-important
## Definition (Devore 7.3)

A statistic $T = t(X_1, \ldots, X_n)$ is **sufficient for $\theta$** if:

$$P(X_1 = x_1, \ldots, X_n = x_n \mid T = t) \text{ does not depend on } \theta$$
:::

. . .

**In words:** Given the value of $T$, the conditional distribution of the data doesn't involve $\theta$.

**Intuition:** Once you know $T$, learning the raw data tells you nothing new about $\theta$.

## The Neyman Factorization Theorem

Computing conditional distributions is hard. The **factorization theorem** gives us a shortcut!

::: callout-important
## Theorem (Neyman-Fisher Factorization)

$T$ is sufficient for $\theta$ if and only if the joint pdf/pmf can be written as:
$$f(x_1, \ldots, x_n; \theta) = g(T; \theta) \cdot h(x_1, \ldots, x_n)$$

where:

-   $g$ involves $\theta$ and depends on data **only through $T$**
-   $h$ depends on data but **not on $\theta$**
:::

. . .

**Strategy:** Look for parts involving $\theta$ and group them with a function of the data.

## Example: Finding the Sufficient Statistic for Poisson

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Poisson}(\mu)$

**Joint pmf:**
$$f(x_1, \ldots, x_n; \mu) = \prod_{i=1}^n \frac{e^{-\mu}\mu^{x_i}}{x_i!} = \frac{e^{-n\mu}\mu^{\sum x_i}}{\prod x_i!}$$

. . .

**Factor it:**
$$= \underbrace{e^{-n\mu}\mu^T}_{g(T; \mu)} \cdot \underbrace{\frac{1}{\prod x_i!}}_{h(x_1, \ldots, x_n)}$$

where $T = \sum x_i$.

. . .

::: callout-tip
## Result
$T = \sum_{i=1}^n X_i$ is sufficient for $\mu$.
:::

## Example: Uniform[0, θ] — A Surprising Result

**Setup:** $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Uniform}[0, \theta]$

**Joint pdf:**
$$f(x_1, \ldots, x_n; \theta) = \frac{1}{\theta^n} \cdot I(0 \leq \min(x_i)) \cdot I(\max(x_i) \leq \theta)$$

. . .

**Factor it:**
$$= \underbrace{\frac{1}{\theta^n} I(\max(x_i) \leq \theta)}_{g(T; \theta)} \cdot \underbrace{I(\min(x_i) \geq 0)}_{h(\mathbf{x})}$$

. . .

::: callout-important
## Key Result
$T = \max(X_i)$ is sufficient for $\theta$.

Notice: $\bar{X}$ is **NOT** sufficient for the uniform distribution!
:::

## Why Isn't $\bar{X}$ Sufficient for Uniform?

```{r}
# Two datasets with same mean but different information about θ
data_A <- c(2, 4, 6, 8)   # mean = 5, max = 8
data_B <- c(1, 4, 5, 10)  # mean = 5, max = 10

cat("Dataset A: mean =", mean(data_A), ", max =", max(data_A), "\n")
cat("Dataset B: mean =", mean(data_B), ", max =", max(data_B), "\n")
```

. . .

Both have $\bar{X} = 5$, but:

-   Dataset A tells us $\theta \geq 8$
-   Dataset B tells us $\theta \geq 10$

**Knowing only $\bar{X}$ loses critical information about $\theta$!**

## Sufficient Statistics: Summary Table

| Distribution | Parameter(s) | Sufficient Statistic(s) |
|--------------|--------------|-------------------------|
| Bernoulli($p$) | $p$ | $\sum X_i$ |
| Poisson($\mu$) | $\mu$ | $\sum X_i$ |
| Exponential($\lambda$) | $\lambda$ | $\sum X_i$ |
| Normal($\mu$, known $\sigma^2$) | $\mu$ | $\bar{X}$ |
| Normal($\mu$, $\sigma^2$) | $(\mu, \sigma^2)$ | $(\bar{X}, S^2)$ or $(\sum X_i, \sum X_i^2)$ |
| Uniform[0, $\theta$] | $\theta$ | $\max(X_i)$ |

## Why Sufficiency Matters: The Big Picture

::: callout-important
## Three Key Connections

1. **MLEs are functions of sufficient statistics**
   - You don't lose anything by working with $T$ instead of raw data

2. **MVUEs are functions of sufficient statistics** (Rao-Blackwell)
   - The best unbiased estimator depends only on $T$

3. **If $T$ is complete and sufficient, any unbiased function of $T$ is the MVUE** (Lehmann-Scheffé)
   - Finding the MVUE becomes much easier!
:::

## The Rao-Blackwell Theorem: Improving Estimators

::: callout-important
## Theorem (Rao-Blackwell)

If $U$ is any unbiased estimator of $\tau(\theta)$ and $T$ is sufficient for $\theta$, then:
$$U^* = E(U \mid T)$$

is also unbiased, and $\text{Var}(U^*) \leq \text{Var}(U)$.
:::

. . .

**In plain English:** Conditioning on a sufficient statistic **never hurts** and usually **helps**.

**Practical implication:** If your estimator doesn't use all the information in $T$, you can always improve it!

## Medical Example: Rao-Blackwell in Action

**Problem:** Estimate $e^{-\mu} = P(\text{no defects})$ for Poisson($\mu$) data.

**Crude estimator:** $U = I(X_1 = 0)$ (is the first observation zero?)

```{r}
#| code-fold: true
set.seed(42)
mu_true <- 1.5
n <- 10
n_sims <- 5000

# Simulation
rb_sim <- tibble(sim = 1:n_sims) |> 
  mutate(
    data = map(sim, \(s) rpois(n, mu_true)),
    T_sum = map_dbl(data, sum),
    crude_U = map_dbl(data, \(d) as.numeric(d[1] == 0)),
    improved_U = (1 - 1/n)^T_sum
  )

rb_sim |> 
  summarize(
    `True e^(-μ)` = exp(-mu_true),
    `Crude: E[U]` = mean(crude_U),
    `Crude: Var` = var(crude_U),
    `Improved: E[U*]` = mean(improved_U),
    `Improved: Var` = var(improved_U)
  ) |> 
  pivot_longer(everything()) |> 
  mutate(value = round(value, 4))
```

The Rao-Blackwell estimator has **much lower variance**!

## Visualization: Rao-Blackwell Improvement

```{r}
#| echo: false
#| fig-height: 5
rb_sim |> 
  select(crude_U, improved_U) |> 
  pivot_longer(everything(), names_to = "Estimator", values_to = "Estimate") |> 
  mutate(Estimator = case_when(
    Estimator == "crude_U" ~ "Crude: I(X₁ = 0)",
    Estimator == "improved_U" ~ "Improved: (1 - 1/n)^T"
  )) |> 
  ggplot(aes(x = Estimate, fill = Estimator)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = exp(-mu_true), color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(title = "Rao-Blackwell: Variance Reduction Through Sufficiency",
       subtitle = paste0("True e^(-μ) = ", round(exp(-mu_true), 3)),
       x = "Estimate", y = "Frequency") +
  scale_fill_manual(values = c("orange", "steelblue")) +
  theme(legend.position = "top")
```

## Completeness: The Final Piece

::: callout-important
## Definition: Complete Statistic

$T$ is **complete** if for any function $g$:
$$E_\theta[g(T)] = 0 \text{ for all } \theta \implies P_\theta(g(T) = 0) = 1 \text{ for all } \theta$$
:::

. . .

**Intuition:** The only unbiased estimator of zero based on $T$ is the zero function itself.

. . .

::: callout-tip
## Lehmann-Scheffé Theorem
If $T$ is **complete** and **sufficient**, and $h(T)$ is unbiased for $\tau(\theta)$, then $h(T)$ is the **unique MVUE** of $\tau(\theta)$.
:::

# Part 5: Putting It All Together {background-color="#2c3e50"}

## The Complete Picture: How Everything Connects

```{r}
#| echo: false
#| fig-height: 5
grViz("
digraph connections {
  graph [rankdir = TB, bgcolor = transparent]
  node [shape = box, style = filled, fontsize = 14]
  
  Data [label = 'Raw Data\\nX₁, ..., Xₙ', fillcolor = lightgray]
  Suff [label = 'Sufficient Statistic T\\n(Data Reduction)', fillcolor = lightyellow]
  MLE [label = 'MLE\\n(Asymptotically Optimal)', fillcolor = lightblue]
  MVUE [label = 'MVUE\\n(Best Unbiased)', fillcolor = lightgreen]
  CRLB [label = 'CRLB\\n(Lower Bound)', fillcolor = lightsalmon]
  
  Data -> Suff [label = 'Factorization\\nTheorem']
  Suff -> MLE [label = 'MLE is function of T']
  Suff -> MVUE [label = 'Rao-Blackwell &\\nLehmann-Scheffé']
  CRLB -> MVUE [label = 'Efficiency:\\nVar(MVUE) = CRLB']
  MLE -> MVUE [label = 'For large n', style = dashed]
}
")
```

## Decision Tree: Finding the Best Estimator

```{r}
#| echo: false
#| fig-height: 6
grViz("
digraph decision {
  graph [rankdir = TB, bgcolor = transparent]
  node [shape = box, style = filled, fontsize = 13]
  
  Q1 [label = 'Need an estimator for θ', fillcolor = lightblue]
  Q2 [label = 'Can you find a sufficient\\nstatistic T?', fillcolor = lightyellow, shape = diamond]
  Q3 [label = 'Is T complete?', fillcolor = lightyellow, shape = diamond]
  Q4 [label = 'Can you find an unbiased\\nfunction h(T)?', fillcolor = lightyellow, shape = diamond]
  
  A1 [label = 'Use MLE\\n(asymptotically optimal)', fillcolor = lightgreen]
  A2 [label = 'h(T) is the MVUE!', fillcolor = palegreen]
  A3 [label = 'Use Rao-Blackwell\\nto improve', fillcolor = lightgreen]
  A4 [label = 'Consider biased\\nestimators via MSE', fillcolor = lightsalmon]
  
  Q1 -> Q2
  Q2 -> Q3 [label = 'Yes']
  Q2 -> A1 [label = 'No or Hard']
  Q3 -> Q4 [label = 'Yes']
  Q3 -> A3 [label = 'No']
  Q4 -> A2 [label = 'Yes']
  Q4 -> A4 [label = 'No']
}
")
```

## Summary: Key Formulas to Remember

| Concept | Formula |
|---------|---------|
| Bias | $E(\hat{\theta}) - \theta$ |
| MSE | $\text{Var}(\hat{\theta}) + \text{Bias}^2$ |
| Standard Error | $\sqrt{\text{Var}(\hat{\theta})}$ |
| Fisher Information | $I(\theta) = E[-\ell''(\theta)]$ |
| CRLB | $\frac{1}{nI(\theta)}$ |
| Efficiency | $\frac{\text{CRLB}}{\text{Var}(\hat{\theta})}$ |
| Consistency | $\text{MSE} \to 0$ as $n \to \infty$ |

## Summary: Key Results by Distribution

| Distribution | MLE | MME | Sufficient Stat | MVUE of Mean |
|--------------|-----|-----|-----------------|--------------|
| Bernoulli($p$) | $\bar{X}$ | $\bar{X}$ | $\sum X_i$ | $\bar{X}$ |
| Poisson($\mu$) | $\bar{X}$ | $\bar{X}$ | $\sum X_i$ | $\bar{X}$ |
| Exponential($\lambda$) | $1/\bar{X}$ | $1/\bar{X}$ | $\sum X_i$ | $\bar{X}$ (for mean) |
| Normal($\mu$, $\sigma^2$) | $\bar{X}$, $\frac{1}{n}\sum(X_i-\bar{X})^2$ | Same | $(\bar{X}, S^2)$ | $\bar{X}$ |
| Uniform[0, $\theta$] | $\max(X_i)$ | $2\bar{X}$ | $\max(X_i)$ | $\frac{n+1}{n}\max(X_i)$ |

## Common Misconceptions to Avoid

::: callout-warning
## Pitfall 1: "Unbiased is always better"
**Reality:** A biased estimator with low variance can have lower MSE!
:::

. . .

::: callout-warning
## Pitfall 2: "MLE = Sufficient Statistic"
**Reality:** MLE is a **function of** sufficient statistics, not the sufficient statistic itself.
:::

. . .

::: callout-warning
## Pitfall 3: "$\bar{X}$ is always sufficient"
**Reality:** $\bar{X}$ is sufficient for location parameters (normal, exponential) but NOT for scale parameters like in Uniform[0, $\theta$].
:::

# Practice Problems {background-color="#9b59b6"}

## Practice Problem 1: Bias-Variance Tradeoff

**Problem (Devore 7.3):** For estimating the mean $\mu$ of a normal distribution with known $\sigma^2 = 4$, compare:

-   $\hat{\mu}_1 = \bar{X}$ (sample mean)
-   $\hat{\mu}_2 = 0.9\bar{X}$ (shrunken mean)

For $n = 16$:

(a) Calculate the bias of each estimator
(b) Calculate the variance of each estimator  
(c) Calculate the MSE of each estimator
(d) For what values of $\mu$ is $\hat{\mu}_2$ better?

## Practice Problem 2: Finding Sufficient Statistics

**Problem (Devore 7.42):** For $X_1, \ldots, X_n \stackrel{iid}{\sim} \text{Gamma}(\alpha, \beta)$ with known $\alpha$:

$$f(x; \beta) = \frac{1}{\Gamma(\alpha)\beta^\alpha}x^{\alpha-1}e^{-x/\beta}$$

(a) Find the sufficient statistic for $\beta$
(b) Find the MLE of $\beta$
(c) Show that the MLE is a function of the sufficient statistic

## Practice Problem 3: CRLB and Efficiency

**Problem (Devore 7.51):** For the geometric distribution $P(X = x) = (1-p)^{x-1}p$:

(a) Find the Fisher Information $I(p)$
(b) Find the CRLB for estimating $p$
(c) Show that $\hat{p} = 1/\bar{X}$ is the MLE
(d) Is $\hat{p}$ efficient? (Hint: Find $\text{Var}(\hat{p})$ using the delta method)

## Practice Problem 4: Medical Application

**Problem:** In a clinical trial, the time (in months) until disease progression for 20 cancer patients follows an Exponential($\lambda$) distribution. The observed times are:

```{r}
progression_times <- c(3.2, 8.1, 5.6, 12.4, 2.1, 7.8, 4.3, 9.5, 6.2, 11.3,
                       3.8, 7.1, 5.9, 10.2, 4.7, 8.6, 6.5, 9.1, 3.5, 7.4)
cat("Sample mean:", mean(progression_times), "months")
```

(a) Find the MLE of $\lambda$
(b) Find the MLE of median survival time (using invariance)
(c) Find the MLE of $P(\text{survive} > 12 \text{ months})$
(d) What is the sufficient statistic for $\lambda$?

## Practice Problem 5: Sufficiency (Conceptual)

**Problem:** Explain in your own words:

(a) Why is $T = \sum X_i$ sufficient for $p$ in Bernoulli data, but not $T = X_1 + 2X_2$?

(b) Why does the CRLB not apply to Uniform[0, $\theta$]?

(c) A colleague proposes using the sample median instead of the sample mean to estimate the mean of a normal distribution. Use the concepts of sufficiency and efficiency to explain why this is not optimal.

## Homework Problems

From **Devore, Berk, Carlton**:

1. **Section 7.1:** Problems 7, 10, 14, 17
2. **Section 7.2:** Problems 22, 25, 31
3. **Section 7.3:** Problems 42, 44, 48
4. **Section 7.4:** Problems 51, 53

. . .

::: callout-tip
## Study Strategy
For each problem:
1. Identify what you're asked to find (bias? variance? sufficient statistic?)
2. Write down the relevant formula
3. Show your work step-by-step
4. Interpret your result in context
:::

## References

::: nonincremental
-   Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Chapters 6-7.
-   Chihara and Hesterberg. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 6.
:::

## Questions? {.center background-color="#3498db"}

**Key Takeaways:**

1. MSE = Variance + Bias² (total error)
2. MLEs are consistent and asymptotically efficient
3. Sufficient statistics capture ALL information about $\theta$
4. MVUE = Best unbiased estimator = achieves CRLB (when it exists)
5. Complete + Sufficient → Unique MVUE

Thank you!
