---
title: "Lesson 19: Analysis of Paired Data"
subtitle: "BSTA 551: Statistical Inference"
author: "Jessica Minnier"
title-slide-attributes:
    data-background-color: "#006a4e"
date: "`r library(here); source(here('class_dates.R')); w9d2`"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    scrollable: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Lesson 19 - Analysis of Paired Data
    html-math-method: mathjax
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(patchwork)
set.seed(551)
theme_set(theme_minimal(base_size = 18))
```

# Lesson 19: Analysis of Paired Data {background-color="#2c3e50"}

## Review: Lesson 18

**Two independent samples (Lesson 18):**

- Two-sample $z$: $\sigma_1^2, \sigma_2^2$ known $\Rightarrow$ pivot $Z \sim N(0,1)$
- **Welch $t$-test** (default): $\sigma_1^2, \sigma_2^2$ unknown $\Rightarrow$ $T_W \approx t_{\hat\nu}$
- **Pooled $t$-test**: equal variances assumed $\Rightarrow$ $T_P \sim t_{n_1+n_2-2}$ (exact)
- $z$-test for $p_1 - p_2$; $F$-test for $\sigma_1^2/\sigma_2^2$

. . .

**Today's Goals:**

1. Recognize when data are **paired** vs. independent and why it matters
2. Show how pairing **reduces variance** when within-pair correlation is positive
3. Derive and apply the **paired $t$-test** and **paired CI** for $\mu_D$ (DBC 10.3)
4. Understand the link between paired analysis and one-sample inference
5. Apply to crossover trials, pre/post studies, and matched case-control designs

## Roadmap

| Design | Key Operation | Method | Reference |
|--------|--------------|--------|-----------|
| Pre/post (same subjects) | $D_i = X_{1i} - X_{2i}$ | One-sample $t$ on $\{D_i\}$ | DBC 10.3 |
| Crossover trial | Same | Paired $t$ | DBC 10.3 |
| Matched case-control | Same | Paired $t$ | DBC 10.3 |
| Bilateral measurements | Same | Paired $t$ | DBC 10.3 |

. . .

::: callout-note
## Motivating Example
**Crossover trial:** 20 hypertensive patients each receive an antihypertensive drug for 4 weeks, then (after washout) placebo for 4 weeks, in random order. Outcome: systolic blood pressure reduction (mmHg). Each patient is their own control so the two measurements are **not independent**; they are paired within each patient.
:::

# The Paired Design {background-color="#3498db"}

## When Are Data Paired?

Data are paired when each observation in group 1 is **naturally linked** to exactly one observation in group 2:

| Design | Example |
|--------|---------|
| Pre/post (same subject) | SBP before vs. after 12 weeks of treatment |
| Crossover trial | Same patient receives drug and placebo in random order |
| Matched case-control | Cases matched 1:1 to controls on age and sex |
| Repeated measurement | Lab value at baseline and 6-month follow-up |
| Bilateral organs | Right vs. left eye intraocular pressure |

. . .

::: callout-important
## The Critical Distinction
**Paired:** $n$ pairs $\to$ $n$ differences $\to$ **one-sample problem on** $D_i$

**Independent:** $n_1 + n_2$ separate observations $\to$ **two-sample problem**

Applying a two-sample test to paired data **ignores within-pair correlation**, discards information, and loses power — sometimes substantially.
:::

## The Paired Reduction

**Setup:** $(X_{1i}, X_{2i})$, $i = 1, \ldots, n$ are $n$ independent pairs (but within each pair, $X_{1i}$ and $X_{2i}$ are correlated).

Define the **within-pair difference**: $D_i = X_{1i} - X_{2i}$

Then $D_1, \ldots, D_n \stackrel{iid}{\sim}$ some distribution with:
$$\mu_D = E[D_i] = \mu_1 - \mu_2 \qquad \sigma_D^2 = \text{Var}(D_i)$$

**Everything reduces to one-sample inference on $D_1, \ldots, D_n$.**

. . .

**Why the variance is smaller:** Let $\rho = \text{Cor}(X_{1i}, X_{2i})$. Then:
$$\sigma_D^2 = \text{Var}(X_{1i} - X_{2i}) = \sigma_1^2 + \sigma_2^2 - 2\rho\sigma_1\sigma_2$$

When $\rho > 0$ (the typical case in biological paired designs):
$$\sigma_D^2 = \sigma_1^2 + \sigma_2^2 - 2\rho\sigma_1\sigma_2 \;<\; \sigma_1^2 + \sigma_2^2$$

Pairing **removes between-subject noise** by differencing within each pair.

## Paired $t$-Test (DBC 10.3)

Let $\bar{D} = \frac{1}{n}\sum_{i=1}^n D_i$ and $S_D^2 = \frac{1}{n-1}\sum_{i=1}^n (D_i - \bar{D})^2$.

**Paired $t$-statistic** for $H_0: \mu_D = \mu_0$ (usually $\mu_0 = 0$):
$$T = \frac{\bar{D} - \mu_0}{S_D / \sqrt{n}} \sim t_{n-1}$$

Exact under normality of $D_i$; approximately valid by CLT for large $n$.

**Paired $100(1-\alpha)\%$ CI for $\mu_D$:**
$$\bar{D} \;\pm\; t_{\alpha/2,\,n-1} \cdot \frac{S_D}{\sqrt{n}}$$

. . .

In R:
```r
t.test(x1, x2, paired = TRUE)   # from two measurement vectors
t.test(d)                        # from pre-computed differences
```

::: callout-note
The paired $t$-test is **identical** to the one-sample $t$-test on the differences — all the theory from Lessons 7–8 applies directly.
:::

## Variance Reduction: How Much Does Pairing Help?

When $\sigma_1 = \sigma_2 = \sigma$, the SE of $\bar{D}$ in the paired analysis is:
$$\text{SE}_\text{paired} = \frac{\sigma_D}{\sqrt{n}} = \frac{\sigma\sqrt{2(1-\rho)}}{\sqrt{n}}$$

The SE from an unpaired analysis with the same total $n$ subjects per group would be:
$$\text{SE}_\text{unpaired} = \sigma\sqrt{\frac{2}{n}}$$

The ratio of CI widths (paired to unpaired) is $\sqrt{1 - \rho}$.

```{r}
#| echo: true
#| code-fold: false

tibble(rho = c(0, 0.3, 0.5, 0.7, 0.8, 0.9)) |>
  mutate(
    width_ratio     = round(sqrt(1 - rho), 3),
    pct_narrower    = scales::percent(1 - sqrt(1 - rho), accuracy = 1)
  ) |>
  knitr::kable(
    col.names = c("rho", "CI width ratio (paired / unpaired)",
                  "% CI width reduction"),
    caption   = "Benefit of pairing when sigma_1 = sigma_2"
  )
```

When $\rho = 0.8$: the paired CI is only $\sqrt{0.2} \approx 45\%$ as wide as the unpaired -- a lot of precision gained!

## Visualizing Variance Reduction by Simulation

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6.5
#| fig-width: 14

set.seed(551)

sim_ci_width <- function(rho, n = 25, sigma = 10, B = 2000) {
  map_dfr(seq_len(B), function(i) {
    L  <- chol(sigma^2 * matrix(c(1, rho, rho, 1), 2))
    xy <- matrix(rnorm(2 * n), n, 2) %*% L
    d  <- xy[, 1] - xy[, 2]
    tibble(
      paired   = 2 * qt(0.975, n - 1) * sd(d) / sqrt(n),
      unpaired = diff(t.test(xy[, 1], xy[, 2])$conf.int)
    )
  }) |>
    summarise(across(everything(), \(x) mean(x, na.rm = TRUE))) |>
    mutate(rho = rho)
}

widths <- map_dfr(c(0, 0.3, 0.5, 0.7, 0.8, 0.9), sim_ci_width)

widths |>
  pivot_longer(c(paired, unpaired), names_to = "Method", values_to = "Width") |>
  mutate(Method = str_to_title(Method)) |>
  ggplot(aes(x = factor(rho), y = Width, fill = Method)) +
  geom_col(position = "dodge", width = 0.55) +
  scale_fill_manual(values = c("Paired" = "#006a4e", "Unpaired" = "#e74c3c")) +
  labs(
    title    = "Mean 95% CI Width: Paired vs. Unpaired t-test",
    subtitle = bquote("Benefit grows with within-pair correlation" ~ rho ~
                      " (n = 25," ~ sigma ~ "= 10 per group, 2000 simulations)"),
    x        = expression("Within-pair correlation " * rho),
    y        = "Mean 95% CI Width",
    fill     = NULL
  ) +
  theme(legend.position = "bottom")
```

# Medical Example: Crossover Trial {background-color="#e74c3c"}

## Crossover Trial: Antihypertensive Drug vs. Placebo

**Study (DBC 10.3 style):** 20 hypertensive patients complete a crossover trial. Each receives both an antihypertensive drug and placebo (random order, 4-week washout). Outcome: systolic BP reduction (mmHg). Test $H_0: \mu_D = 0$ vs. $H_a: \mu_D < 0$ (drug lowers BP more than placebo) at $\alpha = 0.05$.

```{r}
#| echo: true
#| code-fold: false

set.seed(551)
n   <- 20
rho <- 0.75
L   <- chol(10^2 * matrix(c(1, rho, rho, 1), 2))
# Drug arm lowers BP ~5 mmHg more than placebo
xy  <- matrix(rnorm(2 * n), n, 2) %*% L +
       matrix(c(0, -5), n, 2, byrow = TRUE)

sbp_placebo <- xy[, 1]
sbp_drug    <- xy[, 2]
d           <- sbp_drug - sbp_placebo  # negative = drug better

cat(sprintf("Mean D (drug - placebo):  %.2f mmHg\n", mean(d)))
cat(sprintf("SD(D):                    %.2f mmHg\n", sd(d)))

result <- t.test(sbp_drug, sbp_placebo, paired = TRUE, alternative = "less")
tidy(result)
```

```{r}
#| echo: true
#| code-fold: false

cat(sprintf("95%% CI for mu_D:  (%.2f, Inf)\n", result$conf.int[2]))
cat(sprintf("p-value:            %.5f\n", result$p.value))
```

## Visualizing the Paired Data

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 6.5
#| fig-width: 14

p1 <- tibble(
  Patient = factor(seq_len(n)),
  Placebo = sbp_placebo,
  Drug    = sbp_drug
) |>
  ggplot(aes(x = 1, xend = 2, y = Placebo, yend = Drug)) +
  geom_segment(color = "grey65", linewidth = 0.7) +
  geom_point(aes(x = 1, y = Placebo), color = "#e74c3c", size = 3) +
  geom_point(aes(x = 2, y = Drug),    color = "#006a4e", size = 3) +
  scale_x_continuous(breaks = c(1, 2), labels = c("Placebo", "Drug"),
                     limits = c(0.7, 2.3)) +
  labs(title = "SBP Reduction: Each Patient's Profile",
       x = NULL, y = "SBP Reduction (mmHg)")

p2 <- tibble(d = d) |>
  ggplot(aes(x = d)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10,
                 fill = "#3498db", color = "white", alpha = 0.7) +
  geom_density(color = "#2c3e50", linewidth = 1.2) +
  geom_vline(xintercept = 0,       linetype = "dashed",
             color = "grey40", linewidth = 1.0) +
  geom_vline(xintercept = mean(d), color = "#e74c3c", linewidth = 1.4) +
  annotate("text", x = mean(d) - 0.5, y = 0.045,
           label = bquote(bar(D) == .(round(mean(d), 1))),
           color = "#e74c3c", size = 5.5, hjust = 1) +
  labs(title = "Distribution of Differences  D = Drug \u2212 Placebo",
       x = "Difference (mmHg)", y = "Density")

p1 + p2 +
  plot_annotation(
    title    = "Crossover Trial: Antihypertensive Drug vs. Placebo  (n = 20 pairs)",
    subtitle = "Left: spaghetti plot of individual patient profiles   |   Right: distribution of within-pair differences"
  )
```

## What If We Incorrectly Used a Two-Sample Test?

```{r}
#| echo: true
#| code-fold: false

paired_res   <- t.test(sbp_drug, sbp_placebo, paired = TRUE, alternative = "less")
unpaired_res <- t.test(sbp_drug, sbp_placebo, paired = FALSE, alternative = "less")

cat("--- Paired t-test (CORRECT) ---\n")
cat(sprintf("  t = %.3f,  df = %d,  p = %.5f\n",
            paired_res$statistic,
            paired_res$parameter,
            paired_res$p.value))
cat(sprintf("  95%% CI: (%.2f, %.2f)\n\n",
            paired_res$conf.int[1], paired_res$conf.int[2]))

cat("--- Welch two-sample t-test (INCORRECT for paired data) ---\n")
cat(sprintf("  t = %.3f,  df = %.1f,  p = %.5f\n",
            unpaired_res$statistic,
            unpaired_res$parameter,
            unpaired_res$p.value))
cat(sprintf("  95%% CI: (%.2f, %.2f)\n",
            unpaired_res$conf.int[1], unpaired_res$conf.int[2]))
```

. . .

The paired test has a much narrower CI and smaller p-value because it removes between-patient variability ($\rho = `r rho`$). The unpaired test wastes information by treating the two arms as unrelated samples.

# Pre/Post Study Example {background-color="#27ae60"}

## Pre/Post: Ferritin After Iron Supplementation

**Study (DBC 10.3):** 12 iron-deficient patients have serum ferritin (ng/mL) measured before and after 8 weeks of oral iron supplementation. Test for a significant increase ($H_a: \mu_D > 0$).

```{r}
#| echo: true
#| code-fold: false

ferritin_pre  <- c( 8, 11,  9, 14,  7, 12,  6, 15, 10,  9, 13,  8)
ferritin_post <- ferritin_pre + rnorm(length(ferritin_pre), 4, 3)
d_iron        <- ferritin_post - ferritin_pre

cat(sprintf("Mean pre:   %.1f   |   Mean post:  %.1f\n",
            mean(ferritin_pre), mean(ferritin_post)))
cat(sprintf("Mean D:     %.1f   |   SD(D):      %.1f   |   n = %d\n",
            mean(d_iron), sd(d_iron), length(d_iron)))

res_iron <- t.test(ferritin_post, ferritin_pre,
                   paired = TRUE, alternative = "greater")
tidy(res_iron)
```

```{r}
#| echo: true
#| code-fold: false

cat(sprintf("One-sided 95%% CI: (%.2f, Inf)\n", res_iron$conf.int[1]))
cat(sprintf("p-value: %.2e\n", res_iron$p.value))
```

## Checking the Normality Assumption for $D_i$

```{r}
#| echo: true
#| code-fold: true
#| fig-height: 5.5
#| fig-width: 13

p_hist <- tibble(d = d_iron) |>
  ggplot(aes(x = d)) +
  geom_histogram(bins = 8, fill = "#006a4e", color = "white", alpha = 0.75) +
  geom_vline(xintercept = mean(d_iron), color = "#e74c3c",
             linewidth = 1.3, linetype = "dashed") +
  annotate("text", x = mean(d_iron) + 0.4, y = 2.7,
           label = bquote(bar(D) == .(round(mean(d_iron), 1))),
           color = "#e74c3c", size = 5.5, hjust = 0) +
  labs(title = "Histogram of Differences",
       x = "Ferritin increase (ng/mL)", y = "Count")

p_qq <- tibble(d = d_iron) |>
  ggplot(aes(sample = d)) +
  stat_qq(color = "#006a4e", size = 3) +
  stat_qq_line(color = "#e74c3c", linewidth = 1.1) +
  labs(title = "Normal Q-Q Plot of Differences",
       x = "Theoretical quantiles", y = "Sample quantiles")

p_hist + p_qq +
  plot_annotation(
    title    = "Normality Check for Differences  (n = 12)",
    subtitle = "The paired t-test requires approximate normality of D_i, not of the raw pre/post values"
  )
```

. . .

With $n = 12$, normality of $D_i$ matters. The Q-Q plot looks reasonable here. For small $n$ with visibly non-normal differences, use the **Wilcoxon signed-rank test** or **sign test** (Lesson 21).

# Matched Case-Control Design {background-color="#9b59b6"}

## Matched Pairs: Smoking and Lung Cancer

**Study:** 15 lung cancer cases are each individually matched to a healthy control on age ($\pm 2$ yr) and sex. Exposure: pack-years of smoking. Within each matched pair, compute the case–control difference.

```{r}
#| echo: true
#| code-fold: false

set.seed(7)
cases    <- round(pmax(0, rnorm(15, mean = 35, sd = 15)))
controls <- round(pmax(0, cases - rnorm(15, mean = 20, sd = 12)))
d_match  <- cases - controls

cat("Pack-years — Cases:   ", cases,    "\n")
cat("Pack-years — Controls:", controls, "\n")
cat(sprintf("\nMean difference (case - control): %.1f  |  SD: %.1f\n",
            mean(d_match), sd(d_match)))

t.test(cases, controls, paired = TRUE) |> tidy()
```

. . .

`t.test(cases, controls, paired = TRUE)` is equivalent to `t.test(d_match)`. The 1:1 matching creates the pairing structure, and the analysis **must** respect it — ignoring the matching (unpaired test) would discard the correlation induced by matching on age and sex.

## Paired Analysis Checklist

Before choosing paired vs. two-sample analysis, confirm:

1. **1-to-1 link exists**: same subject measured twice, or two subjects deliberately matched
2. **Within-pair correlation is positive**: almost always true for biological measurements on the same individual
3. **$n$ = number of pairs**, not total observations — paired $t$-test has $n - 1$ df, not $2n - 2$
4. **$D_i$ approximately normal** (or $n$ large enough for CLT)

. . .

::: callout-warning
## Common Mistake
Reporting "$n_1 = n_2 = 20$" for a pre/post study and running `t.test(pre, post)` without `paired = TRUE`. This treats $2n = 40$ observations as independent, underestimates the SE, and produces an invalid test. The correct $n$ is 20 pairs, df = 19.
:::

# Key Takeaways {background-color="#2c3e50"}

## Key Takeaways (DBC 10.3)

::: callout-important
## Paired Data: Main Points

1. **Paired designs** arise from pre/post measurements, crossover trials, bilateral organs, and matched studies; each pair yields one difference $D_i = X_{1i} - X_{2i}$

2. **Variance reduction:** $\text{Var}(D_i) = \sigma_1^2 + \sigma_2^2 - 2\rho\sigma_1\sigma_2$; with $\sigma_1 = \sigma_2 = \sigma$, the CI width ratio (paired/unpaired) is $\sqrt{1 - \rho}$ — pairing eliminates between-subject noise when $\rho > 0$

3. **Paired $t$-test:** $T = \bar{D}/(S_D/\sqrt{n}) \sim t_{n-1}$; identical to one-sample $t$ on the $D_i$; exact under normality of $D_i$, approximately valid by CLT for moderate $n$

4. **Always use `paired = TRUE`** in R when the design is paired — omitting it ignores correlation, widens CIs, and reduces power

5. **Check normality of $D_i$** (not of $X_{1i}$ or $X_{2i}$) with a histogram and Q-Q plot; for small $n$ with non-normal $D_i$, use nonparametric alternatives (Lesson 21)
:::

## Looking Ahead

**Next class (Lesson 20): Bootstrap and Permutation Methods for Two Samples (C&H Ch. 3, 5; DBC 10.6)**

- Bootstrap CI for $\mu_1 - \mu_2$: resample each group independently
- Permutation test: shuffle labels under $H_0: F_1 = F_2$
- Paired bootstrap: resample differences
- When do resampling methods outperform parametric tests?

. . .

**Lesson 21:** Nonparametric inference — sign test, Wilcoxon signed-rank, Wilcoxon rank-sum, asymptotic relative efficiency (DBC Ch. 14)

**Lesson 22:** Course Review

## References

::: nonincremental
- Devore, Berk, and Carlton. *Modern Mathematical Statistics with Applications* (Springer). Section 10.3.
- Chihara, L. M. and Hesterberg, T. C. *Mathematical Statistics with Resampling and R* (Wiley). Chapter 3.
:::

## Questions? {.center background-color="#3498db"}

Thank you!

```{bash}
#| eval: false
#| include: false
decktape reveal --fragments docs/lessons/19_PairedData/19_PairedData.html lessons/19_PairedData/19_PairedData.pdf
```
